### 2023-11-28 ootavacuous.tex
Initial commit
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\documentclass[10]{article}

% standard packages

% A more pleasant font
\usepackage[T1]{fontenc} % use postscript type 1 fonts
\usepackage{textcomp} % use symbols in TS1 encoding
\usepackage{mathptmx,helvet,courier} % use nice, standard fonts for roman, sans and monospace respectively

% Improves the text layout
\usepackage{microtype}

\usepackage{lscape}
\usepackage{fancyhdr}
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage{url}
\usepackage{graphics}
\usepackage{enumerate}
\usepackage{ifthen}
\usepackage{float}
\usepackage{listings}
\lstset{basicstyle=\ttfamily}
% \usepackage[strings]{underscore}
% \usepackage{underscore}
\usepackage[bookmarks=true,bookmarksnumbered=true,pdfborder={0 0 0}]{hyperref}

\lstset{
  literate={\_}{}{0\discretionary{\_}{}{\_}}%
}

\usepackage[table]{xcolor}
\usepackage{booktabs}

\DeclareUrlCommand\email{}

\pagestyle{fancy}
\rhead{}

\newfloat{listing}{tbp}{lol}
\floatname{listing}{Listing}

\begin{document}
\title{DNNNNR0: OOTA Execution is Provably Vacuous}

\newcommand{\co}[1]{\lstinline[breaklines=yes,breakatwhitespace=yes]{#1}}

\author{
Paul E.~McKenney\\\email{paulmck@kernel.org} \and
The Indefatigible TBD
}
\date{November 27, 2023 (Pre-Tokyo)}
\maketitle{}

Audience: SG1

\begin{abstract}
	The out-of-thin-air (OOTA) and read-from-untaken-branch (RFUB)
	properties of the specification of \co{memory_order_relaxed}
	have resulted in considerable consternation over the years.
	Many argue that RFUB should be allowed.
	In addition, there are no known instances of full-blown OOTA
	behavior in real C++ implementations.

	This paper goes further, providing an informal proof that OOTA
	cannot occur in real C++ programs running in the real world.
\end{abstract}

\section{Background}
\label{sec:Background}

\subsection{Brief OOTA Overview}
\label{sec:Brief OOTA Overview}

OOTA occurs when a pair of threads loads from each others' stores,
but where each thread's store depends on the value returned by that
thread's load.

@@@ diagram and explanation @@@

Note that C++ implementations are permitted to evaluate to OOTA values
in cases involving undefined behavior, such as use of uninitialized
objects.
However, these situations do not involve OOTA cycles, but rather user
errors and other issues that can lead to unpredictable output.
This paper focuses instead on OOTA cycles.

The result of an OOTA cycle can be any value.

\subsection{Prior Work}
\label{sec:Prior Work}

Early executable C++ memory models~\cite{MarkBatty2011cppmem}
correctly flagged executions involving OOTA.
P0442R0 (``Out-of-Thin-Air Execution is Vacuous'')~\cite{PaulEMcKenney2016OOTA}
provided a decision procedure for classifying behaviors as permitted
misordering on the one hand or disallowed OOTA on the other, using
a perturbation method based on the insight that all OOTA behaviors are
fixed-point computations.

@@@ list of work attempting to avoid OOTA.

All this work focused on either identifying OOTA or on how C++
implementations could avoid it.
Except that no one has proposed a viable C++ implementation that results
in OOTA.

This paper therefore drops the question of how OOTA can be avoided and
instead focuses on proving that OOTA cannot occur.

\subsection{Real-World Constraints}
\label{sec:Real-World Constraints}

The speed of light is finite and atoms are of non-zero size.
These laws of physics mean that if one thread loads the value
stored by some other thread, that load must have occurred later
in global time than did the store.

Although accesses to atomic objects are not observable behavior,\footnote{
	Working Draft, Standard for Programming Language C++
	\co{[intro.abstract]}.}
in practice, any thread's store to a global variable must be as if the
correct value was actually stored.
In contrast, in C, atomic operations are volatile and thus observable
behavior, while in Rust atomic operations currently cannot be volatile.

\subsection{How to Form an OOTA Cycle?}
\label{sec:How to Form an OOTA Cycle?}

An OOTA cycle consists of loads from other threads' stores combined
with stores whose addresses or values depend on the value returned
by the same thread's prior load.
Courtesy of the laws of physics, a given store must execute earlier
in global time than any load from that stored-to object that returns
the value stored.

Therefore, any OOTA cycle must have arcs that go backwards in global
time, and the only arcs that can possibly do so are those arcs that
lie entirely with a given thread.
This in turn requires that a store that depends on an earlier load
be executed earlier in time than that load.
The only two ways that this can happen are when the implementation:

\begin{enumerate}
\item	Is able to determing (prove) the location and value of the store.
\item	Guesses the location and value of the store and has some means
	to take corrective action should any guess prove incorrect.
\end{enumerate}

\subsubsection{Proof}
\label{sec:Proof}

Implementation can sometimes prove the location and value of a given
store.
For example, if the value loaded is multiplied by zero, the result
is known to be zero, so that the store can proceed prior to the load.
But in this case, there is no perturbation of the value returned from
the load that can affect the value stored.
Therefore, by P0442R0, that store cannot possibly take part in an
OOTA cycle.

\subsubsection{Guess}
\label{sec:Guess}

Hardware speculative execution is commonplace, and permits the
hardware to guess at values, squashing the speculation if any
of the guesses prove incorrect.
But in this case, a given store is not committed until all guesses that
this store depends on have been confirmed, an no uncommitted store
is visible to any other thread.
Therefore, hardware speculation cannot result in OOTA cycles.
@@@ cite hardware manuals @@@
This same constraint applies to software, which might use compare-and-swap
loops or hardware transactional memory to confirm guesses.

Either way, the need to wait for guesses to be confirmed before committing
stores prevents those stores from becoming visible prior to the
completion of any loads that those stores depend on.

\subsubsection{What About Flattening?}
\label{sec:What About Flattening?}

The concept of flattening multiple threads into a single thread is
tantalizing, especially given the large cache-miss penalties inherent
to large multicore systems, which can have thousands of CPUs.
One might object to the whole concept of flattening as a compiler
optimization given the difficulty inherent in avoiding deadlocks
and other hazards.

But for the purposes of OOTA cycles, the flattening situation is
much simpler.
Any fully flattened program will be single threaded, which implies
that all of its atomic operations will be sequentially consistent,
which in turn guarantees that OOTA cycles cannot form.

\section{Conclusion}
\label{sec:Conclusion}

Therefore, correct C++ implementations cannot produce OOTA cycles.

% \section{History}
% \label{sec:History}

\bibliographystyle{plain}
\bibliography{bib/RCU,bib/WFS,bib/hw,bib/os,bib/parallelsys,bib/patterns,bib/perfmeas,bib/refs,bib/syncrefs,bib/search,bib/swtools,bib/realtime,bib/TM,bib/standards,bib/memorymodel.bib}

\end{document}

### 2023-11-29 ootavacuous.tex
Functionally complete
At least until further notice!

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{abstract}
-	The out-of-thin-air (OOTA) and read-from-untaken-branch (RFUB)
-	properties of the specification of \co{memory_order_relaxed}
-	have resulted in considerable consternation over the years.
-	Many argue that RFUB should be allowed.
-	In addition, there are no known instances of full-blown OOTA
+	The out-of-thin-air (OOTA) properties of the specification
+	of \co{memory_order_relaxed} have resulted in considerable
+	consternation over the years.
+	Attempts to create memory models that rule out OOTA behaviors
+	have either been non-executable, complex, or unloved by compiler
+	writers.
+	But at the same time, there are no known instances of OOTA
 	behavior in real C++ implementations.
 
-	This paper goes further, providing an informal proof that OOTA
-	cannot occur in real C++ programs running in the real world.
+	This paper goes further, adding temporal constraints from
+	long-standing laws of physics, thereby providing an informal
+	proof that OOTA cannot occur in correctly constructed C++
+	programs running on real-world systems.
+	In other words, a correct C++ implementation necessarily
+	preserves the semantic dependencies that prevent OOTA
+	cycles from forming.
 \end{abstract}
@@
 
-OOTA occurs when a pair of threads loads from each others' stores,
+OOTA occurs when a group of threads load from each others' stores,
 but where each thread's store depends on the value returned by that
 thread's load.
+A given out-of-thin-air value passes around the resulting cycle.
+Because there is a cycle, on a real system, at least one of the links
+must be counter-temporal, that is, either a load returns a value before
+that value is stored, or a store's address or value is computed before
+a prior load returns a value used in that computation.
 
-@@@ diagram and explanation @@@
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 int main() {
+ 2   atomic_int x=0;
+ 3   atomic_int y=0;
+ 4   int r1;
+ 5   int r2;
+ 6   {{{ { r1=x.load(memory_order_relaxed);
+ 7         y.store(r1,memory_order_relaxed); }
+ 8   ||| { r2=y.load(memory_order_relaxed);
+ 9         x.store(r2,memory_order_relaxed); }  }}}
+10   return 0;
+11 }
+\end{verbatim}
+\caption{OOTA Cycle}
+\label{lst:OOTA Cycle}
+\end{listing}
 
-Note that C++ implementations are permitted to evaluate to OOTA values
-in cases involving undefined behavior, such as use of uninitialized
-objects.
+Listing~\ref{lst:OOTA Cycle}
+shows an OOTA cycle, where the \co{\{\{\{} on line~5 begins a set of
+threads, the \co{|||} on line~8 separates a pair of threads, and
+the \co{\}\}\}} on line~9 ends a set of threads.
+Lines~6 and~7 do a relaxed copy from \co{x} to \co{y}, and
+lines~8 and~9 to a relaxed copy from \co{y} to \co{x}.
+Even though both \co{x} and \co{y} are initialized to zero on
+lines~2 and~3, respectively, according to the mathematical core
+of the C++ memory model, the only constraint on the final values
+of \co{x} and \co{y} is that they be equal.
+Counterintuitivily, their final values could be any member of their type.
+
+Note that C++ implementations are permitted to evaluate to a more
+general notion of OOTA values in cases involving unspecified or undefined
+behavior, such as use of uninitialized objects.
 However, these situations do not involve OOTA cycles, but rather user
-errors and other issues that can lead to unpredictable output.
+errors and other issues that can lead to unfriendly optimizations that
+in turn result in unpredictable output.
 This paper focuses instead on OOTA cycles.
 
-The result of an OOTA cycle can be any value.
+C++ implementations may of course avail themselves of the as-if rule.
+However, correct use of the as-if rule requires that the observable
+behaviors of the program be appropriately preserved.
 
@@
 
-Early executable C++ memory models~\cite{MarkBatty2011cppmem}
-correctly flagged executions involving OOTA.
+Some executable C++ memory models correctly flagged executions involving
+OOTA~\cite{JadeAlglave2014HerdingCats}.\footnote{
+	Others cleverly avoid this issue by forbidding atomic
+	stores of non-constant values~\cite{MarkBatty2011cppmem}.}
+% @@@ Better herd7 C11 citation?
+However, because these models are atemporal, they cannot reject
+OOTA executions other than by flagging the OOTA value as arbitrary
+(which they in fact do).
+
 P0442R0 (``Out-of-Thin-Air Execution is Vacuous'')~\cite{PaulEMcKenney2016OOTA}
@@
 
-@@@ list of work attempting to avoid OOTA.
+Some workers recommend avoiding OOTA by ordering prior relaxed
+loads before subsequent relaxed
+stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
+but this requires real instructions be executed, consuming real
+time and real electrical power to solve a strictly theoretical
+problem.
+
+Other workers recommend various procedures to identify and avoid OOTA
+cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
+but none of these have been looked upon favorably by compiler implementers.
 
@@
 implementations could avoid it.
-Except that no one has proposed a viable C++ implementation that results
-in OOTA.
+None of this work applied real-world temporal constraints to the OOTA
+problem.
+Which might explain why no known real-world C++ implementation results
+in OOTA executions.
 
@@
 
-\subsection{Real-World Constraints}
+\section{Real-World Constraints}
 \label{sec:Real-World Constraints}
 
-The speed of light is finite and atoms are of non-zero size.
+Real-world constraints are imposed by the standard, which have been
+considered in prior work, and by the laws of physics, which have
+not.
+
+\subsection{Constraints of the Standard}
+\label{sec:Constraints of the Standard}
+
+Because C++ volatile atomics constitute observable behavior,\footnote{
+	Working Draft, Standard for Programming Language C++
+	\co{[intro.abstract]}.}
+they must be executed in strict accordance with the rules of the abstract
+machine.
+This means that any C++ volatile atomic operation involving a store
+must execute as if all loads whose return values are used to compute
+that store's address or value have already returned.
+OOTA afficienados will recognize this as a special case of ordering
+relaxed loads before relaxed stores, albeit one not requiring
+expensive memory-fence instructions on weakly ordered architectures.
+
+Although non-volatile accesses to atomic objects are not observable behavior,
+they are still subject to the memory model.
+In particular, implementations cannot:
+
+\begin{enumerate}
+\item	Invent atomic stores.
+\item	Duplicate atomic stores.
+\end{enumerate}
+
+\paragraph{Invented Atomic Stores}
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 int main() {
+ 2   atomic_int x = 2;
+ 3   volatile int r1;
+ 4   {{{ { x.store(42, memory_order_relaxed); // invented store
+ 5         x.store(3, memory_order_relaxed); }
+ 6   ||| r1 = x.load(memory_order_relaxed);
+ 7   }}};
+ 8   return 0;
+ 9 }
+\end{verbatim}
+\caption{CPPMEM Invented Store}
+\label{lst:CPPMEM Invented Store}
+\end{listing}
+
+The reason that atomic stores cannot be invented is that doing so can
+introduce new (and almost certainly undesireable) behaviors.
+To see this, consider Listing~\ref{lst:CPPMEM Invented Store},
+a CPPMEM\footnote{
+	\url{http://svr-pes20-cppmem.cl.cam.ac.uk/cppmem/index.html}.}
+litmus test that demonstrates such a behavior.
+Without line~4, only the values~2 and~3 can be stored to \co{r1}.
+With that line, the additional value~42 can also be stored to \co{r1}.
+The compiler is therefore forbidded from inventing that store of 42
+unless it can prove that doing so does not negatively affect the
+program's observable behaviors, independently of OOTA cycles.
+For example, the implementation might be able to prove that there
+are no other accesses to \co{x} at the time of the invented store.
+
+\paragraph{Duplicated Atomic Stores}
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 int main() {
+ 2   atomic_int x = 0;
+ 3   volatile int r1;
+ 4   {{{ { x.store(1, memory_order_relaxed); // duplicated store
+ 5         x.store(1, memory_order_relaxed); }
+ 6   ||| { r1 = x.load(memory_order_relaxed);
+ 7         x.store(2, memory_order_relaxed); }
+ 8   }}};
+ 9   return 0;
+10 }
+\end{verbatim}
+\caption{CPPMEM Duplicated Store}
+\label{lst:CPPMEM Duplicated Store}
+\end{listing}
+
+Duplicating atomic stores can also introduce new and undesireable
+behaviors.
+To see this, consider Listing~\ref{lst:CPPMEM Duplicated Store}, a CPPMEM
+litmus test that demonstrates such a behavior.
+Without line~4, if the final value of \co{r1} is 1, then the final value
+of \co{x} must be 2.
+With that line, the final value of \co{x} can be 1 even when the
+final value of \co{r1} is 1.
+The compiler is therefore forbidded from duplicating that store of 1
+unless it can prove that doing so does not negatively affect the
+program's observable behaviors, independently of OOTA cycles.
+Again, the implementation might be able to prove that there
+are no other accesses to \co{x} at the time of the invented store.
+
+\paragraph{Omitted Atomic Stores}
+In contrast, an implementation is permitted to omit atomic stores
+under less restricted circumstances.
+For example, a pair of back-to-back stores to \co{x} might always be
+executed such that no other thread accesses \co{x} during the time
+between those two stores.
+This means that, short of inspecting the assembly code, the user has no
+way of proving that the store was in fact omitted.
+The user can always resort to volatile atomic stores or inline
+assembly\footnote{
+	For example, by placing the Linux-kernel \co{barrier()} macro
+	between the two stores.
+	This macro is an empty GCC \co{asm} that specifies the \co{memory}
+	clobber.}
+to prevent the compiler from omitting a store.
+However, omitting a store cannot create an OOTA cycle.
+
+\subsection{Laws of Physics}
+\label{sec:Laws of Physics}
+
+The speed of light is finite~\cite{OleRoemer1671SpeedOfLight}
+and atoms are of non-zero size~\cite{JeanBaptistePerrin1923AtomSize}.
 These laws of physics mean that if one thread loads the value
@@
 in global time than did the store.
+Current memory models do not capture these temporal constraints,
+which makes it more difficult for them to efficiently rule out
+OOTA behavior.
 
-Although accesses to atomic objects are not observable behavior,\footnote{
-	Working Draft, Standard for Programming Language C++
-	\co{[intro.abstract]}.}
-in practice, any thread's store to a global variable must be as if the
-correct value was actually stored.
-In contrast, in C, atomic operations are volatile and thus observable
-behavior, while in Rust atomic operations currently cannot be volatile.
-
-\subsection{How to Form an OOTA Cycle?}
+\section{How to Form an OOTA Cycle?}
 \label{sec:How to Form an OOTA Cycle?}
 
-An OOTA cycle consists of loads from other threads' stores combined
-with stores whose addresses or values depend on the value returned
-by the same thread's prior load.
+Rather than look at how to prevent an implementation from forming
+an OOTA cycle, this section instead looks at how to form one.
+This will lead to the conclusion that a correct C++ implementation
+cannot form an OOTA cycle.
+
+Again, an OOTA cycle consists of loads from other threads' stores where
+a given store's address or value depends on the value returned by
+the corresponding thread's prior load.
 Courtesy of the laws of physics, a given store must execute earlier
@@
 
-Therefore, any OOTA cycle must have arcs that go backwards in global
-time, and the only arcs that can possibly do so are those arcs that
-lie entirely with a given thread.
+And again, an OOTA cycle must have at least one link that goes backwards
+in time.
+The only possible candidate counter-temporal links so are those
+confined to a given thread.
 This in turn requires that a store that depends on an earlier load
 be executed earlier in time than that load.
-The only two ways that this can happen are when the implementation:
+The two ways that this can happen are when the implementation:
 
 \begin{enumerate}
-\item	Is able to determing (prove) the location and value of the store.
+\item	Is able to determine (prove) the location and value of the store.
 \item	Guesses the location and value of the store and has some means
@@
 
-\subsubsection{Proof}
-\label{sec:Proof}
+Each of these possibilities is discussed in the next sections.
+followed by a section on the possibility of flattening a multithreaded
+program into a single thread.
+
+\subsection{Implementation Proves Value}
+\label{sec:Implementation Proves Value}
 
@@
 
-\subsubsection{Guess}
-\label{sec:Guess}
+\subsection{Implementation Guesses at Value}
+\label{sec:Implementation Guesses at Value}
 
@@
 But in this case, a given store is not committed until all guesses that
-this store depends on have been confirmed, an no uncommitted store
+this store depends on have been confirmed, and thus no uncommitted store
 is visible to any other thread.
-Therefore, hardware speculation cannot result in OOTA cycles.
-@@@ cite hardware manuals @@@
+Therefore, hardware speculation cannot result in OOTA cycles, and
+by design, even on weakly ordered
+systems~\cite{ARMv7A:2010,ARMv8A:2017,PowerISA2.07-2013}.
 This same constraint applies to software, which might use compare-and-swap
@@
 
-\subsubsection{What About Flattening?}
-\label{sec:What About Flattening?}
+\subsection{But What About Flattening?}
+\label{sec:But What About Flattening?}
 
@@
 tantalizing, especially given the large cache-miss penalties inherent
-to large multicore systems, which can have thousands of CPUs.
+to large multicore systems, which can have many sockets containing
+thousands of CPUs.
 One might object to the whole concept of flattening as a compiler
@@
 and other hazards.
+Nevertheless, it is only reasonable to ask whether flattening can
+result in OOTA cycles.
+As we will see, the answer is ``no''.
 
@@
 much simpler.
-Any fully flattened program will be single threaded, which implies
-that all of its atomic operations will be sequentially consistent,
-which in turn guarantees that OOTA cycles cannot form.
+Any fully flattened program will be single threaded, which implies that
+each of its atomic load operations will return the last value stored to
+that same object in sequenced-before order (or the initial value if there
+is no such store).
+Thus, any sequence of loads and stores must have a first operation and
+a last operation, and the last operation cannot affect the first
+operation.
+Therefore, cycles cannot possibly form in a flattened program,
+which in turn implies that OOTA cycles cannot form.
+
+\section{Correct Implementations Cannot Form OOTA Cycles}
+\label{sec:Correct Implementations Cannot Form OOTA Cycles}
+
+This section has shown that an implementation that correctly
+handles the memory model, observable behavior, and the as-if rule
+cannot form OOTA cycles.
 
@@
 
-Therefore, correct C++ implementations cannot produce OOTA cycles.
+This paper has added temporal reasoning to OOTA analysis of the C++
+memory model.
+This work shows that correct C++ implementations are inherently incapable
+of producing OOTA cycles.
 

### 2023-11-29 ootavacuous.tex
Apply paper number and self-review
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{document}
-\title{DNNNNR0: OOTA Execution is Provably Vacuous}
+\title{D3064R0: OOTA Execution is Provably Vacuous}
 
@@
 \label{sec:Background}
+
+This section provides a brief overview of the OOTA problem followed
+by an equally brief summary of prior work in this area.
 
@@
 
+All OOTA workers owe a debt to the foundational work in the infamous
+``Causality Test Cases''.\footnote{
+	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
+
 Some executable C++ memory models correctly flagged executions involving
@@
 OOTA executions other than by flagging the OOTA value as arbitrary
-(which they in fact do).
+(which some in fact do).
 
@@
 
-\section{Correct Implementations Cannot Form OOTA Cycles}
+\subsection{Correct Implementations Cannot Form OOTA Cycles}
 \label{sec:Correct Implementations Cannot Form OOTA Cycles}

### 2023-11-29 ootavacuous.tex
Apply David Goldblatt feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Paul E.~McKenney\\\email{paulmck@kernel.org} \and
+Michael Wong\\\email{fraggamuffin@gmail.com} \and
+Maged Michael\\\email{maged.michael@gmail.com} \and
 The Indefatigible TBD
@@
 threads, the \co{|||} on line~8 separates a pair of threads, and
-the \co{\}\}\}} on line~9 ends a set of threads.
+the \co{\}\}\}} on line~9 ends a set of threads.\footnote{
+	This is pseudocode in the format of CPPMEM,
+	which does not permit writes from non-local variables.}
 Lines~6 and~7 do a relaxed copy from \co{x} to \co{y}, and
@@
 Counterintuitivily, their final values could be any member of their type.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 int main() {
+ 2   atomic_int x = 0;
+ 3   atomic_int y = 0;
+ 4   volatile int r1;
+ 5   volatile int r2;
+ 6   {{{ { r1 = x.load(memory_order_relaxed);
+ 7         if (r1 == 1)
+ 8           y.store(1, memory_order_relaxed);
+ 9         else
+10           y.store(r1, memory_order_relaxed); }
+11   ||| { r2=y.load(memory_order_relaxed);
+12         x.store(r2,memory_order_relaxed); }  }}}
+13   }}};
+14   return 0;
+15 }
+\end{verbatim}
+\caption{OOTA Cycle Based on Conditionals}
+\label{lst:OOTA Cycle Based on Conditionals}
+\end{listing}
+
+A similar effect can be achieved using conditional statements, as
+shown in Listing~\ref{lst:OOTA Cycle Based on Conditionals}.
 
@@
 not.
+Additional constraints are imposed by hardware architecture and
+design, which have been partially accounted for in the standard,
+one limitation being that accurate and executable formal descriptions
+of hardware memory models did not appear until after the standard
+was released.
 
@@
 
-Because C++ volatile atomics constitute observable behavior,\footnote{
-	Working Draft, Standard for Programming Language C++
-	\co{[intro.abstract]}.}
+Because C++ volatile atomics constitute observable
+behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2022N4910},
 they must be executed in strict accordance with the rules of the abstract
@@
 
+\subsection{Hardware Architecture and Design}
+\label{sec:Hardware Architecture and Design}
+
+Heavily used CPU families now have accurate and executable formal
+memory models, but even more important, there is much more clarity
+on exactly what hardware will and will not do in the course of
+speculative execution.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 int foo(int i) {
+ 2   atomic_int x = 0;
+ 3   int y[3];
+ 4   volatile int r1;
+ 5   {{{ { y[i].store(42, memory_order_relaxed);
+ 6         x.store(1, memory_order_relaxed); }
+ 7   ||| r1 = x.load(memory_order_relaxed);
+ 8   }}};
+ 9   return 0;
+10 }
+\end{verbatim}
+\caption{Speculated Store and Non-Speculated Load}
+\label{lst:Speculated Store and Non-Speculated Load}
+\end{listing}
+
+For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
+suppose that the storage for \co{x} is located immediately before that
+for \co{a[]}.
+Suppose further that hardware speculation incorrectly guesses the value
+of \co{i} to be -1, so that the speculated store of 42 on line~5 uses
+the address of \co{x}.\footnote{
+	Yes, \co{y[-1]} is undefined behavior, but the CPU neither knows
+	nor cares.}
+Then line~7 can cause the final value of \co{r1} to be 42, when
+by the rules of the C++ abstract machine the value must instead be
+either zero or one.
+
+Therefore, non-speculative loads are prohibited from loading values from
+a speculative stores.
+
+In addition, in order to avoid hardware-based OOTA, a value stored
+speculatively will not be exposed to loads from threads other than the
+one executing that store.
+
 \section{How to Form an OOTA Cycle?}
@@
 
+\subsection{Illustrative Examples}
+\label{sec:Illustrative Examples}
+
+This section works through OOTA examples, showing the effects of
+the constraints called out in this paper.
+
+\subsubsection{OOTA Cycle}
+\label{sec:OOTA Cycle}
+
+Listing~\ref{lst:OOTA Cycle}
+shows the canonical example of an OOTA cycle.
+Because values from speculative stores cannot be visible to other threads,
+line~7 cannot be committed until after line~6 completes.
+Similarly, line~9 cannot be committed until after line~8 completes.
+
+Because values stored take non-zero time to propagate from one thread
+to another, no cycle can form, and thus no OOTA cycle can form.
+
+\subsubsection{OOTA Cycle Based on Conditionals}
+\label{sec:OOTA Cycle Based on Conditionals}
+
+Listing~\ref{lst:OOTA Cycle Based on Conditionals}
+shows an if-condition-based example of an OOTA cycle.
+Because values from speculative stores cannot be visible to other threads,
+lines~8 and~10 cannot be committed until after the load in line~6 completes
+and the conditional on line~7 is evaluated non-speculatively.
+Similarly, line~12 cannot be committed until after line~11 completes.
+
+Because values stored take non-zero time to propagate from one thread
+to another, no cycle can form, and thus no OOTA cycle can form.
+
+Note that if both legs of the \co{if} statement spanning lines~7-10,
+the compiler would be within its rights to hoist the store out of
+that \co{if} statement.
+However, in this case, the two stores on lines~8 and~10 are storing
+distinct values, preventing the compiler from hoisting.
+
+Note also that some architectures provide conditional-move instructions,
+which allows line~7 to avoid emitting a conditional branch.
+However, such architectures would still treat a store of the
+target of the conditional-move instructions as speculative until
+such time as the conditional moves (and any loads that they depend
+on) complete.
+
 \subsection{Correct Implementations Cannot Form OOTA Cycles}
@@
 
+\section{Acknowledgments}
+\label{sec:Acknowledgments}
+
+We are grateful to David Goldblatt for his careful review of an early
+draft of this paper.
+Nonetheless, all errors and omissions in this paper are the sole property
+of the authors.
+
 % \section{History}

### 2023-11-30 ootavacuous.tex
Provide prototype definition of OOTA cycle
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 instead focuses on proving that OOTA cannot occur.
+
+\subsection{OOTA Definition}
+\label{sec:OOTA Definition}
+
+Prior work defines ``OOTA cycle'' by example, without a precise
+definition.
+Sometimes ``causal cycle'' is used as if it was a definition, but
+without a clear decision process for what does and does not
+constitute a causal cycle..
+P0442R0 defines an OOTA cycle as a fixed-point computation that is
+sensitive to perturbations, which makes perfect sense to that paper's
+authors, but has left others unsatisfied.
+
+We could follow this long-established tradition of proposing
+solutions to the OOTA problem without a definition of this problem.
+However, we (perhaps foolishly) felt the need to provide a clear
+definition.
+
+An OOTA cycle (also known as a causal cycle) is a cycle in the
+memory-model sense involving the load-buffering (LB) pattern.
+Within each thread in the cycle, the value stored depends on
+the value loaded.
+This dependency might be a data dependency
+(see Listing~\ref{lst:OOTA Cycle})
+or it might be a control dependency
+(see Listing~\ref{lst:OOTA Cycle Based on Conditionals}).\footnote{
+	Sufficiently motivated readers should have no problem constructing
+	examples involving address dependencies.}
+There must be at least two distinct values such that if a given load
+returns a given value, that value forces the cycle to execute so as to
+cause the corresponding store's value to be equal to that given value, and
+for that store to occur early enough for the load to return that value.
 

### 2023-11-30 ootavacuous.tex
Differentiate temporal and atemporal links
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 in global time than did the store.
-Current memory models do not capture these temporal constraints,
-which makes it more difficult for them to efficiently rule out
-OOTA behavior.
+In other words, the reads-from relation is temporal in nature.
+
+In contrast, the C++11 modification-order relation is atemporal.
+This means that the store that executed earlier in global time
+might appear later in the stored-to object's modification
+order~\cite{McKenney20xxParallelProgramming}.
+The atemporal nature of C++11 modification order is due to the hardware
+store-buffer optimizations used by modern multicore systems.
+
+Current programming-language abstract machines  do not capture these
+temporal constraints, which makes it more difficult for them to
+efficiently rule out OOTA behavior.
 
@@
 the corresponding thread's prior load.
-Courtesy of the laws of physics, a given store must execute earlier
-in global time than any load from that stored-to object that returns
-the value stored.
+All of the links from ont thread to the next are reads-from links, which
+are temporal in nature, that is, a given store must execute earlier in
+global time than any load from that stored-to object that returns the
+value stored.
 

### 2023-12-01 ootavacuous.tex
Self-review plus publicity plan
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	This paper goes further, adding temporal constraints from
-	long-standing laws of physics, thereby providing an informal
+	long-standing laws of physics and constraints on hardware
+	architecture and design, thereby providing an informal
 	proof that OOTA cannot occur in correctly constructed C++
 	programs running on real-world systems.
-	In other words, a correct C++ implementation necessarily
+
+	The result is that a correct C++ implementation necessarily
 	preserves the semantic dependencies that prevent OOTA
@@
 P0442R0 defines an OOTA cycle as a fixed-point computation that is
-sensitive to perturbations, which makes perfect sense to that paper's
+destroyed by perturbations, which makes perfect sense to that paper's
 authors, but has left others unsatisfied.
@@
 However, we (perhaps foolishly) felt the need to provide a clear
-definition.
+definition, which is a more detailed version of the wording in 33.5.4p8
+(\co{[atomics.order]}):
 
-An OOTA cycle (also known as a causal cycle) is a cycle in the
-memory-model sense involving the load-buffering (LB) pattern.
+\begin{quote}
+	Implementations should ensure that no “out-of-thin-air” values
+	are computed that circularly depend on their own computation.
+\end{quote}
+
+We therefore define an OOTA cycle (also known as a causal cycle) as a
+cycle in the memory-model sense involving the load-buffering (LB) pattern.
 Within each thread in the cycle, the value stored depends on
@@
 store-buffer optimizations used by modern multicore systems.
+
+This atemporality is also why cycles are not considered to have
+OOTA cycles when at least one link from one thread to the next
+involves both threads storing to the same object.
+Similarly, cycles having at least link where one thread loads from an
+object and the next thread stores to that object are also not considered
+to be OOTA cycles.
 
@@
 
+\section{Wording}
+\label{sec:Wording}
+
+Add the following note in \co{[atomics.order]} after 33.5.4p9:
+
+\begin{quote}
+	[ \emph{Note:} Implementations that respect the following
+	constraints are incapable of computing out-of-thin-air value
+	that depend circularly on their own computation:
+
+	\begin{itemize}
+	\item	Correctly implement the C++ abstract machine, and
+	\item	Run on correctly implemented hardware
+		(specifically, hardware that forbids non-speculative
+		loads from returning values from speculative stores
+		and that forbids exposing values from speculative
+		stores to other threads), and
+	\item	Reside in a universe having a finite speed of light
+		and atomic matter.
+	\end{itemize}
+
+	All three constraints are subject to the as-if rule.
+	For example, those creating an alternate universe in which
+	C++ implementations are to run correctly should ensure
+	that their universe behaves as if the speed of light was
+	finite and the nature of matter was atomic.
+	--- \emph{end note} ]
+\end{quote}
+
 \section{Conclusion}
@@
 We are grateful to David Goldblatt for his careful review of an early
-draft of this paper.
+draft of this paper and to John Wikerson for asking Paul for a rant and
+taking the proffered rant seriously.
+We also owe David Goldblatt a debt of gratitude for his having asked
+an insightful question at the right time.
 Nonetheless, all errors and omissions in this paper are the sole property
 of the authors.
+
+% To do:
+% Luc Maranget: Memory models.
+% Richard Gristenthwaite: ARM hardware.
+% Mark Rutland: ARM memory ordering.
+% Will Deacon: ARM hardware, C++, and OOTA.
+% Greg Kroah-Hartman and Linus Torvalds: FYI, trouble being caused.
+% Christoph Hellwig: Why slow on BPF memory model.
+% Miguel Ojeda: LKMM and Rust.
+% Alice Ryhl: LKMM and Rust.
+% Ralf Jung: Rust.
+% Catalin Marinas: ARM and formal methods (qspinlock).
+% Segher Boessenkool:  GCC.
+% Dan Lustig: NVIDIA memory ordering.
+% Andrea Parri: Memory models, RISC-V hardware.
+% Jonas Oberhauser: Memory models.
+% Hernan Ponce de Leon: Memory models.
+% Palmer Dabbelt: RISC-V hardware.
+% Ali Sezgin: P0422R0.
+% Tony Tye: P0422R0.
+% Other memory-model maintainers, including Peter Zijlstra.
+% Derek Williams: PowerPC hardware.
+% Hans Boehm: Skeptic, memory models, OOTA.  Suggestions for others?
+% Olivier Giroux: Skeptic, memory models, NVIDIA.
+% Luke Geeson: Whatever...  Review.
+% Ori Lahav: Prior OOTA.
+% Viktor Vafeiadis:  Prior OOTA.
+% Brian Demsky: Prior OOTA.
+
+% @@@ Inter-CPU speculation if inter-CPU squashing of speculation.
+
+% Done:
+% David Goldblatt
+% Michael Wong
+% Maged Michael
+% John Wickerson: Thank you.  Peter Sewell?  Mark Batty? Other TFWM authors?
+% Akira Yokosawa: Professional courtesy, perfbook memory models.
+% Dan Kelley, Alexei Starovoitov, Mykola Lysenko: FYI.  OOTA dropped
+% Alan Jeffrey: Fixed-point insight P0422R0.  Reached out via LinkedIn.
+% Jade Alglave: C++ herd model?  Memory models and variety of hardware.
+% Alan Stern: Memory models, mathematical logic, and variety of hardware.
+
+% \appendix
+
+% \section{Demonstration of OOTA-Cycle Definition}
+% \label{app:Demonstration of OOTA-Cycle Definition}
 

### 2023-12-01 ootavacuous.tex
Clarify that OOTA cycles imply non-OOTA bugs
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+Therefore, if your C++ implementation produces OOTA cycles, you have
+bugs that would be considered bugs even if OOTA cycles were allowed.
+And you need to fix those bugs, regardless of whether they are in your
+compiler, your hardware, or your alternative universe.
+
 \section{Acknowledgments}
@@
 
-We are grateful to David Goldblatt for his careful review of an early
-draft of this paper and to John Wikerson for asking Paul for a rant and
-taking the proffered rant seriously.
-We also owe David Goldblatt a debt of gratitude for his having asked
-an insightful question at the right time.
+We are grateful to David Goldblatt and Jade Alglave for their careful
+review of an early draft of this paper and to John Wikerson for asking
+Paul for a rant and taking the proffered rant seriously.
+We also owe David Goldblatt a debt of gratitude for his having asked an
+insightful question at the right time, and for his ``Deathstation 9000''
+demonic CPU.
 Nonetheless, all errors and omissions in this paper are the sole property
@@
 % Viktor Vafeiadis:  Prior OOTA.
+% Derek Dreyer: Prior OOTA.
 % Brian Demsky: Prior OOTA.
@@
 % David Goldblatt
+% Boqun Feng?
 % Michael Wong

### 2023-12-01 ootavacuous.tex
Rework real-world constraints, mainly from Jade's questions
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+The following sections discuss these constraints, starting with the
+laws of physics, continuing with hardware constraints, and ending
+with constraints imposed by the standard.
+
+\subsection{Laws of Physics}
+\label{sec:Laws of Physics}
+
+Evidence to date suggests that the universe is
+causal~\cite{Plato360BC-causality},
+the speed of light is finite~\cite{OleRoemer1671SpeedOfLight}, and
+atoms are of non-zero size~\cite{JeanBaptistePerrin1923AtomSize}.
+These fundamental laws of physics are empirical in nature, but
+are backed by a great many observations extending back to a time
+preceding electronic computers, let alone computer languages that
+support concurrency.
+
+The relationship between these three laws of physics and OOTA
+cycles is described in the following sections.
+
+\subsubsection{Causality}
+\label{sec:Causality}
+
+Plato's ``Timaeus'' notwithstanding, there has been much recent debate
+as to whether the universe is in fact causal with cause necessarily
+preceding effect.
+However, no one has managed to construct a real time-travel machine
+or any other apparatus that might result in effect preceding cause,
+as discussed at length in the Wikipedia article on retrocausality and
+its many references.\footnote{
+	\url{https://en.wikipedia.org/wiki/Retrocausality}.}
+
+This paper will therefore assume that effect cannot precede cause in
+any real-world computing system.
+
+\subsubsection{Finite Speed of Light}
+\label{sec:Finite Speed of Light}
+
+If the speed of light was infinite, then information could be transmitted
+at infinite speed, arriving at any destination, no matter how remote,
+in zero time.
+This situation would permit a load to return the value stored by some
+other thread at the exact instant that the store executed, which
+would in turn remove propagation delay as a factor preventing
+causal loops such as OOTA cycles from forming.
+
+However, all macroscopic measurements to date have shown the speed of
+light to be finite, thus preventing causal loops based on infinite-speed
+propagation of information.
+
+This paper will therefore assume that the speed of light is finite, and
+thus the maximum speed at which information can propagate is also finite.
+
+\subsubsection{Non-Zero Sized Atoms}
+\label{sec:Non-Zero Sized Atoms}
+
+If atoms were of zero size, a zero-sized computer might be constructed.
+In such a computer, even give finite speed of light, information could
+propagate among computing elements in zero time.
+This situation would also permit a load to return the value stored by
+some other thread at the exact instant that the store executed, which
+would again in turn remove propagation delay as a factor preventing
+causal loops such as OOTA cycles from forming.
+
+However, there has to date been no successful attempt to construct a
+zero-sized computing device capable of running non-trivial C++ programs.
+
+This paper will therefore assume that any computing device capable of
+running C++ will be of non-zero size.
+
+\subsubsection{Laws of Physics: Consequences}
+\label{sec:Laws of Physics: Consequences}
+
+These three laws of physics mean that if one thread loads the value
+stored by some other thread, that load must have occurred later
+in global time than did the store.
+In other words, the reads-from relation is temporal in nature.
+
+In contrast, the C++11 modification-order relation is atemporal,
+because computing systems can and observably do determine the
+modification order of a set of concurrent stores long after the fact.
+This means that the store that executed earlier in global time
+might appear later in the stored-to object's modification
+order~\cite{McKenney20xxParallelProgramming}.
+The atemporal nature of C++11 modification order is due to the hardware
+store-buffer optimizations used by modern multicore systems.
+
+This atemporality is also why cycles are not considered to have
+OOTA cycles when at least one link from one thread to the next
+involves both threads storing to the same object.
+Similarly, cycles having at least link where one thread loads from an
+object and the next thread stores to that object are also not considered
+to be OOTA cycles.
+
+Current programming-language abstract machines do not capture these
+temporal constraints, which makes it more difficult for them to
+efficiently rule out OOTA behavior.
+
+\subsection{Hardware Architecture and Design}
+\label{sec:Hardware Architecture and Design}
+
+In happy contrast to the situation at the start of the C++11 memory-model
+effort, heavily used CPU families now have accurate and executable formal
+memory models, all of which prohibit the causal cycles required to form
+OOTA cycles.
+Even more important, there is clarity on exactly what hardware can and
+can not do in the course of speculative execution.
+Finally, computation is based on instructions, and instructions require
+finite time to execute, even when executed speculatively.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 int foo(int i) {
+ 2   atomic_int x = 0;
+ 3   int y[3];
+ 4   volatile int r1;
+ 5   {{{ { y[i].store(42, memory_order_relaxed);
+ 6         x.store(1, memory_order_relaxed); }
+ 7   ||| r1 = x.load(memory_order_relaxed);
+ 8   }}};
+ 9   return 0;
+10 }
+\end{verbatim}
+\caption{Speculated Store and Non-Speculated Load}
+\label{lst:Speculated Store and Non-Speculated Load}
+\end{listing}
+
+For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
+suppose that the storage for \co{x} is located immediately before that
+for \co{a[]}.
+Suppose further that hardware speculation incorrectly guesses the value
+of \co{i} to be -1, so that the speculated store of 42 on line~5 uses
+the address of \co{x}.\footnote{
+	Yes, \co{y[-1]} is undefined behavior, but the CPU neither knows
+	nor cares, nor should it.}
+Then line~7 can cause the final value of \co{r1} to be 42, when
+by the rules of the C++ abstract machine the value must instead be
+either zero or one.
+
+Therefore, non-speculative loads are prohibited from loading values from
+speculative stores.
+
+As a special case of this prohibitino, a value stored speculatively must
+not be exposed to non-speculative loads from other threads.
+In the further special case of multiple hardware threads sharing a core,
+one might imagine speculation involving multiple threads.
+However, in this case, any squashing of speculation must squash the
+full extent of that speculation, across all of the hardware threads
+that were involved.
+
+Finally, the prohibition of hardware OOTA applies within a single
+multi-threaded core as well as between cores.
+This means that for any set of speculative accesses to be committed,
+all loads in that set must have been confirmed by corresponding
+non-speculative stores.
+
 \subsection{Constraints of the Standard}
@@
 The reason that atomic stores cannot be invented is that doing so can
-introduce new (and almost certainly undesireable) behaviors.
+introduce new (and almost certainly undesirable) behaviors.
 To see this, consider Listing~\ref{lst:CPPMEM Invented Store},
@@
 However, omitting a store cannot create an OOTA cycle.
-
-\subsection{Laws of Physics}
-\label{sec:Laws of Physics}
-
-The speed of light is finite~\cite{OleRoemer1671SpeedOfLight}
-and atoms are of non-zero size~\cite{JeanBaptistePerrin1923AtomSize}.
-These laws of physics mean that if one thread loads the value
-stored by some other thread, that load must have occurred later
-in global time than did the store.
-In other words, the reads-from relation is temporal in nature.
-
-In contrast, the C++11 modification-order relation is atemporal.
-This means that the store that executed earlier in global time
-might appear later in the stored-to object's modification
-order~\cite{McKenney20xxParallelProgramming}.
-The atemporal nature of C++11 modification order is due to the hardware
-store-buffer optimizations used by modern multicore systems.
-
-This atemporality is also why cycles are not considered to have
-OOTA cycles when at least one link from one thread to the next
-involves both threads storing to the same object.
-Similarly, cycles having at least link where one thread loads from an
-object and the next thread stores to that object are also not considered
-to be OOTA cycles.
-
-Current programming-language abstract machines do not capture these
-temporal constraints, which makes it more difficult for them to
-efficiently rule out OOTA behavior.
-
-\subsection{Hardware Architecture and Design}
-\label{sec:Hardware Architecture and Design}
-
-Heavily used CPU families now have accurate and executable formal
-memory models, but even more important, there is much more clarity
-on exactly what hardware will and will not do in the course of
-speculative execution.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 int foo(int i) {
- 2   atomic_int x = 0;
- 3   int y[3];
- 4   volatile int r1;
- 5   {{{ { y[i].store(42, memory_order_relaxed);
- 6         x.store(1, memory_order_relaxed); }
- 7   ||| r1 = x.load(memory_order_relaxed);
- 8   }}};
- 9   return 0;
-10 }
-\end{verbatim}
-\caption{Speculated Store and Non-Speculated Load}
-\label{lst:Speculated Store and Non-Speculated Load}
-\end{listing}
-
-For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
-suppose that the storage for \co{x} is located immediately before that
-for \co{a[]}.
-Suppose further that hardware speculation incorrectly guesses the value
-of \co{i} to be -1, so that the speculated store of 42 on line~5 uses
-the address of \co{x}.\footnote{
-	Yes, \co{y[-1]} is undefined behavior, but the CPU neither knows
-	nor cares.}
-Then line~7 can cause the final value of \co{r1} to be 42, when
-by the rules of the C++ abstract machine the value must instead be
-either zero or one.
-
-Therefore, non-speculative loads are prohibited from loading values from
-a speculative stores.
-
-In addition, in order to avoid hardware-based OOTA, a value stored
-speculatively will not be exposed to loads from threads other than the
-one executing that store.
 
@@
 
-\section{Wording}
-\label{sec:Wording}
+\section{Suggested Changes to Wording}
+\label{sec:Suggested Changes to Wording}
 
@@
 
-% Done:
+% Done as of November 30, 2023:
 % David Goldblatt
-% Boqun Feng?
+% Boqun Feng, Uladzislau Rezki, Neeraj Upadhyay, Joel Fernandes, and
+%	Frederic Weisbecker (RCU proteges)
 % Michael Wong
 % Maged Michael
-% John Wickerson: Thank you.  Peter Sewell?  Mark Batty? Other TFWM authors?
+% John Wickerson: Thank you.  OK for Peter Sewell, Mark Batty, etc.
 % Akira Yokosawa: Professional courtesy, perfbook memory models.

### 2023-12-01 ootavacuous.tex
Clarify that invented atomic stores forbidden by abstract machine
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 The reason that atomic stores cannot be invented is that doing so can
-introduce new (and almost certainly undesirable) behaviors.
+introduce new (and almost certainly undesirable) behaviors that are
+forbidden by the abstract machine.
 To see this, consider Listing~\ref{lst:CPPMEM Invented Store},

### 2023-12-01 ootavacuous.tex
Add "as if" qualifier to excluded atomic behavior
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 without a clear decision process for what does and does not
-constitute a causal cycle..
+constitute a causal cycle.
 P0442R0 defines an OOTA cycle as a fixed-point computation that is
@@
 they are still subject to the memory model.
-In particular, implementations cannot:
+In particular, implementations must not behave as if they have:
 
 \begin{enumerate}
-\item	Invent atomic stores.
-\item	Duplicate atomic stores.
+\item	Invented atomic stores.
+\item	Duplicated atomic stores.
 \end{enumerate}

### 2023-12-01 ootavacuous.tex
Note that tools do not currently flag all OOTA cycles
Reported-by: Jade Alglave <Jade.Alglave@arm.com>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 OOTA executions other than by flagging the OOTA value as arbitrary
-(which some in fact do).
+(which some in fact do in at least some cases).
 

### 2023-12-02 ootavacuous.tex
Add the first five causality tests
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 they are still subject to the memory model.
-In particular, implementations must not behave as if they have:
+In particular, implementations must not behave as if the source code has
+additional:
 
@@
 
-% \appendix
+\appendix
 
-% \section{Demonstration of OOTA-Cycle Definition}
-% \label{app:Demonstration of OOTA-Cycle Definition}
+\section{Demonstration of OOTA-Cycle Definition}
+\label{app:Demonstration of OOTA-Cycle Definition}
+
+\subsection{Litmus Tests from “Causality Test Cases"}
+\label{app:Litmus Tests from “Causality Test Cases"}
+
+\subsubsection{Causality Test Case 1}
+\label{app:Causality Test Case 1}
+
+Listing~\ref{lst:Causality Test Case 1}
+shows causality test case 1, for which the \co{r1 == r2 == 1} result
+is to be allowed.
+However, the herd7 tool does not find this litmus test's OOTA cycle.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 C oota-causality-1
+ 2 {
+ 3   [x] = 0;
+ 4   [y] = 0;
+ 5 }
+ 6
+ 7 P0(volatile atomic_int *x, volatile atomic_int *y) {
+ 8   int r1 = atomic_load_explicit(x, memory_order_relaxed);
+ 9   if (r1 >= 0)
+10     atomic_store_explicit(y, 1, memory_order_relaxed);
+11 }
+12
+13 P1(volatile atomic_int *x, volatile atomic_int *y) {
+14   int r2 = atomic_load_explicit(y, memory_order_relaxed);
+15   atomic_store_explicit(x, r1, memory_order_relaxed);
+16 }
+17
+18 exists(0:r1=1 /\ 1:r2=1)
+\end{verbatim}
+\caption{Causality Test Case 1}
+\label{lst:Causality Test Case 1}
+\end{listing}
+
+\subsubsection{Causality Test Case 2}
+\label{app:Causality Test Case 2}
+
+Listing~\ref{lst:Causality Test Case 2}
+shows causality test case 2, for which the \co{r1 == r2 == r3 == 1} result
+is to be allowed.
+The herd7 tool finds this OOTA cycle.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 C oota-causality-2
+ 2 {
+ 3   [x] = 0;
+ 4   [y] = 0;
+ 5 }
+ 6
+ 7 P0(volatile atomic_int *x, volatile atomic_int *y) {
+ 8   int r1 = atomic_load_explicit(x, memory_order_relaxed);
+ 9   int r2 = atomic_load_explicit(x, memory_order_relaxed);
+10   if (r1 == r2)
+11     atomic_store_explicit(y, 1, memory_order_relaxed);
+12 }
+13
+14 P1(volatile atomic_int *x, volatile atomic_int *y) {
+15   int r3 = atomic_load_explicit(y, memory_order_relaxed);
+16   atomic_store_explicit(x, r3, memory_order_relaxed);
+17 }
+18
+19 exists(0:r1=1 /\ 0:r2=1 /\ 1:r3=1)
+\end{verbatim}
+\caption{Causality Test Case 2}
+\label{lst:Causality Test Case 2}
+\end{listing}
+
+\subsubsection{Causality Test Case 3}
+\label{app:Causality Test Case 3}
+
+Listing~\ref{lst:Causality Test Case 3}
+shows causality test case 3, for which the \co{r1 == r2 == r3 == 1} result
+is to be allowed.
+The herd7 tool finds this OOTA cycle.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 C oota-causality-3
+ 2 {
+ 3   [x] = 0;
+ 4   [y] = 0;
+ 5 }
+ 6
+ 7 P0(volatile atomic_int *x, volatile atomic_int *y) {
+ 8   int r1 = atomic_load_explicit(x, memory_order_relaxed);
+ 9   int r2 = atomic_load_explicit(x, memory_order_relaxed);
+10   if (r1 == r2)
+11     atomic_store_explicit(y, 1, memory_order_relaxed);
+12 }
+13
+14 P1(volatile atomic_int *x, volatile atomic_int *y) {
+15   int r3 = atomic_load_explicit(y, memory_order_relaxed);
+16   atomic_store_explicit(x, r3, memory_order_relaxed);
+17 }
+18
+19 P2(volatile atomic_int *x, volatile atomic_int *y) {
+20   atomic_store_explicit(x, 2, memory_order_relaxed);
+21 }
+22
+23 exists(0:r1=1 /\ 0:r2=1 /\ 1:r3=1)
+\end{verbatim}
+\caption{Causality Test Case 3}
+\label{lst:Causality Test Case 3}
+\end{listing}
+
+\subsubsection{Causality Test Case 4}
+\label{app:Causality Test Case 4}
+
+Listing~\ref{lst:Causality Test Case 4}
+shows causality test case 4, for which the \co{r1 == r2 == 1} result
+is to be forbidden.
+The herd7 tool finds this OOTA cycle, but flags it as such using \co{S08}.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 C oota-causality-4
+ 2 {
+ 3   [x] = 0;
+ 4   [y] = 0;
+ 5 }
+ 6
+ 7 P0(atomic_int *x, atomic_int *y) {
+ 8   int r1 = atomic_load_explicit(x, memory_order_relaxed);
+ 9   atomic_store_explicit(y, r1, memory_order_relaxed);
+10 }
+11
+12 P1(atomic_int *x, atomic_int *y) {
+13   int r2 = atomic_load_explicit(y, memory_order_relaxed);
+14   atomic_store_explicit(x, r2, memory_order_relaxed);
+15 }
+16
+17 exists(0:r1=1 /\ 1:r2=1)
+\end{verbatim}
+\caption{Causality Test Case 4}
+\label{lst:Causality Test Case 4}
+\end{listing}
+
+\subsubsection{Causality Test Case 5}
+\label{app:Causality Test Case 5}
+
+Listing~\ref{lst:Causality Test Case 5}
+shows causality test case 5, for which the \co{r1 == r2 == 1 && r3 == 0}
+result is to be forbidden.
+The herd7 tool finds this OOTA cycle, but flags it as such using \co{S08}.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 C oota-causality-5
+ 2 {
+ 3   [x] = 0;
+ 4   [y] = 0;
+ 5   [z] = 0;
+ 6 }
+ 7
+ 8 P0(atomic_int *x, atomic_int *y, atomic_int *z) {
+ 9   int r1 = atomic_load_explicit(x, memory_order_relaxed);
+10   atomic_store_explicit(y, r1, memory_order_relaxed);
+11 }
+12
+13 P1(atomic_int *x, atomic_int *y, atomic_int *z) {
+14   int r2 = atomic_load_explicit(y, memory_order_relaxed);
+15   atomic_store_explicit(x, r2, memory_order_relaxed);
+16 }
+17
+18 P2(atomic_int *x, atomic_int *y, atomic_int *z) {
+19   atomic_store_explicit(z, 1, memory_order_relaxed);
+20 }
+21
+22 P3(atomic_int *x, atomic_int *y, atomic_int *z) {
+23   int r3 = atomic_load_explicit(z, memory_order_relaxed);
+24   atomic_store_explicit(x, r3, memory_order_relaxed);
+25 }
+26
+27 exists(0:r1=1 /\ 1:r2=1 /\ 3:r3=0)
+\end{verbatim}
+\caption{Causality Test Case 5}
+\label{lst:Causality Test Case 5}
+\end{listing}
+
+\subsection{Additional Litmus Tests}
+\label{app:Additional Litmus Tests}
+
+\subsubsection{Non-LB Litmus Tests}
+\label{app:Non-LB Litmus Tests}
+
+Listing~\ref{lst:Non-LB Litmus Test}
+shows a litmus test that can produce an OOTA cycle, but that is not
+a member of the load-buffering (LB) class of litmus tests.
+Suppose that lines~9 and~10 both somehow return the value 21.
+Then line~12 will store the value 42 to \co{x}, which will cause
+line~17 to store 21 to \co{y} and line~22 to store 21 to \co{z},
+thus justifying the values initially loaded by lines~9 and~10.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 C oota-non-lb
+ 2 {
+ 3   [x] = 2;
+ 4   [y] = 1;
+ 5   [z] = 1;
+ 6 }
+ 7
+ 8 P0(atomic_int *x, atomic_int *y, atomic_int *z) {
+ 9   int r0 = atomic_load_explicit(y, memory_order_relaxed);
+10   int r1 = atomic_load_explicit(z, memory_order_relaxed);
+11   int r2 = r0 + r1;
+12   atomic_store_explicit(x, r2, memory_order_relaxed);
+13 }
+14
+15 P1(atomic_int *x, atomic_int *y, atomic_int *z) {
+16   int r3 = atomic_load_explicit(x, memory_order_relaxed);
+17   atomic_store_explicit(y, r3 / 2, memory_order_relaxed);
+18 }
+19
+20 P2(atomic_int *x, atomic_int *y, atomic_int *z) {
+21   int r4 = atomic_load_explicit(x, memory_order_relaxed);
+22   atomic_store_explicit(z, r4 / 2, memory_order_relaxed);
+23 }
+24
+25 locations [0:r0;0:r1;0:r2;1:r3;2:r4]
+26 exists(x=42 /\ y=21 /\ z=21)
+\end{verbatim}
+\caption{Non-LB Litmus Test}
+\label{lst:Non-LB Litmus Test}
+\end{listing}
 

### 2023-12-02 litmus/checkalllitmus.sh

@@ -1,0 +1,17 @@
+#!/bin/bash
+#
+# Usage: checkalllitmus.sh
+#
+# Runs all .litmus tests, placing the output in .litmus.out files.
+
+failed=0
+for i in *.litmus
+do
+	herd7 -c11 $i > $i.out 2>&1
+	ret=$?
+	if test "$ret" -ne 0
+	then
+		failed=1
+	fi
+done
+exit $failed

### 2023-12-02 ootavacuous.tex
More Alan Stern feedback (December 2nd)
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-	The result is that a correct C++ implementation necessarily
-	preserves the semantic dependencies that prevent OOTA
-	cycles from forming.
+	The result is that a correct per-execution-path C++ implementation
+	necessarily preserves the semantic dependencies that prevent
+	OOTA cycles from forming.
+	Implementations that undertake complex cross-thread analysis
+	still need to be careful.
 \end{abstract}
@@
 We therefore define an OOTA cycle (also known as a causal cycle) as a
-cycle in the memory-model sense involving the load-buffering (LB) pattern.
+cycle in the memory-model sense involving only reads-from links, that
+is, the value stored in on thread being read by the next thread.
 Within each thread in the cycle, the value stored depends on
@@
 	examples involving address dependencies.}
-There must be at least two distinct values such that if a given load
+@@@ There must be at least two distinct values such that if a given load
 returns a given value, that value forces the cycle to execute so as to
@@
 
-Because C++ volatile atomics constitute observable
+Because volatile atomics constitute observable
 behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2022N4910},
@@
 machine.
-This means that any C++ volatile atomic operation involving a store
+This means that any volatile atomic operation involving a store
 must execute as if all loads whose return values are used to compute
@@
 
-Although non-volatile accesses to atomic objects are not observable behavior,
-they are still subject to the memory model.
+Although C++ non-volatile accesses to atomic objects are not observable
+behavior, they are still subject to the memory model.
 In particular, implementations must not behave as if the source code has
@@
 \end{enumerate}
+
+The discussion of these two constraints is followed by a discussion of
+when redundant atomic stores may safely be omitted.
 
@@
 
-\paragraph{Omitted Atomic Stores}
-In contrast, an implementation is permitted to omit atomic stores
-under less restricted circumstances.
-For example, a pair of back-to-back stores to \co{x} might always be
-executed such that no other thread accesses \co{x} during the time
-between those two stores.
+\paragraph{Omitted Redundant Atomic Stores}
+In contrast, a pair of back-to-back non-volatile atomic stores to \co{x}
+might always be executed such that no other thread accesses \co{x}
+during the time between those two stores.
 This means that, short of inspecting the assembly code, the user has no
-way of proving that the store was in fact omitted.
-The user can always resort to volatile atomic stores or inline
-assembly\footnote{
+way of proving that the C++ implementation acted as if that store was
+omitted from the source code.
+In cases where this is undesirable, the user can resort to volatile
+atomic stores or to inline assembly\footnote{
 	For example, by placing the Linux-kernel \co{barrier()} macro
@@
 to prevent the compiler from omitting a store.
-However, omitting a store cannot create an OOTA cycle.
+However, acting as if such a redundant store was omitted from the source
+code cannot create an OOTA cycle.
+After all, the implementation is already within its as-if rights to
+assume that the first of a back-to-back pair of non-volatile relaxed
+atomic stores was omitted from the source code.
 
@@
 
-We are grateful to David Goldblatt and Jade Alglave for their careful
-review of an early draft of this paper and to John Wikerson for asking
-Paul for a rant and taking the proffered rant seriously.
+We are grateful to David Goldblatt, Jade Alglave, and Alan Stern for their
+careful review of an early draft of this paper and to John Wikerson for
+asking Paul for a rant and taking the proffered rant seriously.
 We also owe David Goldblatt a debt of gratitude for his having asked an

### 2023-12-03 ootavacuous.tex
Add script to auto-format/include litmus tests
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{verbatim}
- 1 C oota-causality-5
- 2 {
- 3   [x] = 0;
- 4   [y] = 0;
- 5   [z] = 0;
- 6 }
- 7
- 8 P0(atomic_int *x, atomic_int *y, atomic_int *z) {
- 9   int r1 = atomic_load_explicit(x, memory_order_relaxed);
-10   atomic_store_explicit(y, r1, memory_order_relaxed);
-11 }
-12
-13 P1(atomic_int *x, atomic_int *y, atomic_int *z) {
-14   int r2 = atomic_load_explicit(y, memory_order_relaxed);
-15   atomic_store_explicit(x, r2, memory_order_relaxed);
-16 }
-17
-18 P2(atomic_int *x, atomic_int *y, atomic_int *z) {
-19   atomic_store_explicit(z, 1, memory_order_relaxed);
-20 }
-21
-22 P3(atomic_int *x, atomic_int *y, atomic_int *z) {
-23   int r3 = atomic_load_explicit(z, memory_order_relaxed);
-24   atomic_store_explicit(x, r3, memory_order_relaxed);
-25 }
-26
-27 exists(0:r1=1 /\ 1:r2=1 /\ 3:r3=0)
+@@ DisplayLitmus litmus/oota-causality-5.litmus @@
 \end{verbatim}
@@
 \begin{verbatim}
- 1 C oota-non-lb
- 2 {
- 3   [x] = 2;
- 4   [y] = 1;
- 5   [z] = 1;
- 6 }
- 7
- 8 P0(atomic_int *x, atomic_int *y, atomic_int *z) {
- 9   int r0 = atomic_load_explicit(y, memory_order_relaxed);
-10   int r1 = atomic_load_explicit(z, memory_order_relaxed);
-11   int r2 = r0 + r1;
-12   atomic_store_explicit(x, r2, memory_order_relaxed);
-13 }
-14
-15 P1(atomic_int *x, atomic_int *y, atomic_int *z) {
-16   int r3 = atomic_load_explicit(x, memory_order_relaxed);
-17   atomic_store_explicit(y, r3 / 2, memory_order_relaxed);
-18 }
-19
-20 P2(atomic_int *x, atomic_int *y, atomic_int *z) {
-21   int r4 = atomic_load_explicit(x, memory_order_relaxed);
-22   atomic_store_explicit(z, r4 / 2, memory_order_relaxed);
-23 }
-24
-25 locations [0:r0;0:r1;0:r2;1:r3;2:r4]
-26 exists(x=42 /\ y=21 /\ z=21)
+@@ DisplayLitmus litmus/oota-non-lb.litmus @@
 \end{verbatim}
@@
 \label{lst:Non-LB Litmus Test}
+\end{listing}
+
+\subsubsection{Multi-Cycle Litmus Tests}
+\label{app:Multi-Cycle Litmus Tests}
+
+\begin{listing}[tbp]
+\begin{verbatim}
+@@ DisplayLitmus litmus/oota-two-source.litmus @@
+\end{verbatim}
+\caption{Multi-Cycle Litmus Test}
+\label{lst:Multi-Cycle Litmus Test}
 \end{listing}

### 2023-12-03 ootavacuous.sh

@@ -1,0 +1,52 @@
+#!/bin/bash
+#
+# Usage: sh ootavacuous.sh
+#
+# Commands:
+#
+# @@ DisplayLitmus path.litmus @@
+#	Use c2latex.sh to squish the tabs and number the lines.
+#	Note that "path.litmus" must not contain whitespace.
+# @@ RunLitmus path.litmus @@
+#	Use "herd7 -c11" to run the test, placing the output in
+#	path.litmus.out, then use c2latex.sh to number the lines.
+#	Note that "path.litmus" must not contain whitespace.
+
+T="`mktemp -d ${TMPDIR-/tmp}/ootavacuous.sh.XXXXXX`"
+trap 'rm -rf $T' 0
+
+awk -v q="'" -v bs='\\' < ootavacuous.tex '
+BEGIN {
+	print "cat << '___EOF___'";
+	print "% Automatically built, do not edit";
+	print "%";
+}
+
+/^@@ DisplayLitmus .* @@$/ {
+	path = $3;
+	print "'___EOF___'";
+	print "grep -v " q "^[( ]" bs "*" q " < " path " | c2latex.sh";
+	print "cat << '___EOF___'";
+	next;
+}
+
+/^@@ RunLitmus .* @@$/ {
+	path = $3;
+	print "'___EOF___'";
+	print "herd7 -c11 " path " | c2latex.sh ";
+	print "cat << '___EOF___'";
+	next;
+}
+
+{
+	print $0;
+}
+
+END {
+	print "___EOF___";
+}' | sh > ootavacuous-flat.tex
+
+pdflatex ootavacuous-flat.tex
+bibtex ootavacuous-flat.aux
+pdflatex ootavacuous-flat.tex
+pdflatex ootavacuous-flat.tex

### 2023-12-03 ootavacuous.tex
Use directives for litmus-test listings December 3rd
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{verbatim}
- 1 C oota-causality-1
- 2 {
- 3   [x] = 0;
- 4   [y] = 0;
- 5 }
- 6
- 7 P0(volatile atomic_int *x, volatile atomic_int *y) {
- 8   int r1 = atomic_load_explicit(x, memory_order_relaxed);
- 9   if (r1 >= 0)
-10     atomic_store_explicit(y, 1, memory_order_relaxed);
-11 }
-12
-13 P1(volatile atomic_int *x, volatile atomic_int *y) {
-14   int r2 = atomic_load_explicit(y, memory_order_relaxed);
-15   atomic_store_explicit(x, r1, memory_order_relaxed);
-16 }
-17
-18 exists(0:r1=1 /\ 1:r2=1)
+@@ DisplayLitmus litmus/oota-causality-1.litmus @@
 \end{verbatim}
@@
 \begin{verbatim}
- 1 C oota-causality-2
- 2 {
- 3   [x] = 0;
- 4   [y] = 0;
- 5 }
- 6
- 7 P0(volatile atomic_int *x, volatile atomic_int *y) {
- 8   int r1 = atomic_load_explicit(x, memory_order_relaxed);
- 9   int r2 = atomic_load_explicit(x, memory_order_relaxed);
-10   if (r1 == r2)
-11     atomic_store_explicit(y, 1, memory_order_relaxed);
-12 }
-13
-14 P1(volatile atomic_int *x, volatile atomic_int *y) {
-15   int r3 = atomic_load_explicit(y, memory_order_relaxed);
-16   atomic_store_explicit(x, r3, memory_order_relaxed);
-17 }
-18
-19 exists(0:r1=1 /\ 0:r2=1 /\ 1:r3=1)
+@@ DisplayLitmus litmus/oota-causality-2.litmus @@
 \end{verbatim}
@@
 \begin{verbatim}
- 1 C oota-causality-3
- 2 {
- 3   [x] = 0;
- 4   [y] = 0;
- 5 }
- 6
- 7 P0(volatile atomic_int *x, volatile atomic_int *y) {
- 8   int r1 = atomic_load_explicit(x, memory_order_relaxed);
- 9   int r2 = atomic_load_explicit(x, memory_order_relaxed);
-10   if (r1 == r2)
-11     atomic_store_explicit(y, 1, memory_order_relaxed);
-12 }
-13
-14 P1(volatile atomic_int *x, volatile atomic_int *y) {
-15   int r3 = atomic_load_explicit(y, memory_order_relaxed);
-16   atomic_store_explicit(x, r3, memory_order_relaxed);
-17 }
-18
-19 P2(volatile atomic_int *x, volatile atomic_int *y) {
-20   atomic_store_explicit(x, 2, memory_order_relaxed);
-21 }
-22
-23 exists(0:r1=1 /\ 0:r2=1 /\ 1:r3=1)
+@@ DisplayLitmus litmus/oota-causality-3.litmus @@
 \end{verbatim}
@@
 \begin{verbatim}
- 1 C oota-causality-4
- 2 {
- 3   [x] = 0;
- 4   [y] = 0;
- 5 }
- 6
- 7 P0(atomic_int *x, atomic_int *y) {
- 8   int r1 = atomic_load_explicit(x, memory_order_relaxed);
- 9   atomic_store_explicit(y, r1, memory_order_relaxed);
-10 }
-11
-12 P1(atomic_int *x, atomic_int *y) {
-13   int r2 = atomic_load_explicit(y, memory_order_relaxed);
-14   atomic_store_explicit(x, r2, memory_order_relaxed);
-15 }
-16
-17 exists(0:r1=1 /\ 1:r2=1)
+@@ DisplayLitmus litmus/oota-causality-4.litmus @@
 \end{verbatim}

### 2023-12-03 ootavacuous.tex
Alan Stern review December 3rd
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 omitted from the source code.
-In cases where this is undesirable, the user can resort to volatile
-atomic stores or to inline assembly\footnote{
+In cases where this omission is undesirable, the user can resort to
+volatile atomic stores or to inline assembly\footnote{
 	For example, by placing the Linux-kernel \co{barrier()} macro
@@
 However, acting as if such a redundant store was omitted from the source
-code cannot create an OOTA cycle.
-After all, the implementation is already within its as-if rights to
-assume that the first of a back-to-back pair of non-volatile relaxed
-atomic stores was omitted from the source code.
+code cannot create an OOTA cycle because any valid execution (OOTA or not)
+of the program with the first of a back-to-back pair of non-volatile
+relaxed atomic stores omitted is essentially a valid execution of the
+program with the store present.
 

### 2023-12-03 cbmc/mult.c
Add a couple of cbmc sdep detectors
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -1,0 +1,23 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include <assert.h>
+
+int main(int argc, char *argv[])
+{
+	int i;
+	int y[2] = { 0, 0 };
+
+	for (i = 0; i < 2; i++) {
+		char r1;
+		char r2;
+		int *yp = &y[i];
+
+		// Uncomment this statement for non-zero case.
+		r1 = argv[i + 1][0] - '0';
+		// Uncomment this statement for zero case.
+		// r1 = 0;
+		r2 = argv[i + 1][0] - '0';
+		*yp = r1 * r2;
+	}
+	assert(y[0] == y[1]);
+}

### 2023-12-03 cbmc/oota-causality-2-p0.c

@@ -1,0 +1,18 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include <assert.h>
+
+int main(int argc, char *argv[])
+{
+	int i;
+	int y[2] = { 0, 0 };
+
+	for (i = 0; i < 2; i++) {
+		char r1;
+		int *yp = &y[i];
+
+		r1 = argv[i + 1][0];
+		*yp = r1;
+	}
+	assert(y[0] == y[1]);
+}

### 2023-12-03 ootavacuous.tex
Add JMM Causality Test Cases 6-8.
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 is to be allowed.
-However, the herd7 tool does not find this litmus test's OOTA cycle.
+However, the herd7 tool does not find this litmus test's cycle, presumably
+because this tool does not do the deep inter-thread compiler analysis
+required to find it.
 
@@
 is to be allowed.
-The herd7 tool finds this OOTA cycle.
+The herd7 tool finds this cycle, and does not flag it as an OOTA cycle.
 
@@
 is to be forbidden.
-The herd7 tool finds this OOTA cycle, but flags it as such using \co{S08}.
+The herd7 tool finds this OOTA cycle, and flags it as such using \co{S8}.
 
@@
 result is to be forbidden.
-The herd7 tool finds this OOTA cycle, but flags it as such using \co{S08}.
+The herd7 tool finds this OOTA cycle, and flags it as such using \co{S8}.
 
@@
 \label{lst:Causality Test Case 5}
+\end{listing}
+
+\subsubsection{Causality Test Case 6}
+\label{app:Causality Test Case 6}
+
+Listing~\ref{lst:Causality Test Case 6}
+shows causality test case 6, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The herd7 tool finds this cycle, and does not flag it as an OOTA cycle.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+@@ DisplayLitmus litmus/oota-causality-6.litmus @@
+\end{verbatim}
+\caption{Causality Test Case 6}
+\label{lst:Causality Test Case 6}
+\end{listing}
+
+\subsubsection{Causality Test Case 7}
+\label{app:Causality Test Case 7}
+
+Listing~\ref{lst:Causality Test Case 7}
+shows causality test case 7, for which the \co{r1 == r2 == r3 == 1}
+result is to be allowed.
+The herd7 tool finds this cycle, and does not flag it as an OOTA cycle.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+@@ DisplayLitmus litmus/oota-causality-7.litmus @@
+\end{verbatim}
+\caption{Causality Test Case 7}
+\label{lst:Causality Test Case 7}
+\end{listing}
+
+\subsubsection{Causality Test Case 8}
+\label{app:Causality Test Case 8}
+
+Listing~\ref{lst:Causality Test Case 8}
+shows causality test case 8, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The herd7 tool does not find this cycle, but when \co{r2} is explicitly
+initialized to the constant \co{1}, it does find the cycle and refrains
+from flagging it as an OOTA cycle.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+@@ DisplayLitmus litmus/oota-causality-8.litmus @@
+\end{verbatim}
+\caption{Causality Test Case 8}
+\label{lst:Causality Test Case 8}
 \end{listing}

### 2023-12-04 ootavacuous.tex
Add JMM Causality Test Cases 9 and 9a
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+\subsubsection{Causality Test Case 9}
+\label{app:Causality Test Case 9}
+
+Listing~\ref{lst:Causality Test Case 9}
+shows causality test case 9, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The herd7 tool does not find this cycle, but when \co{r2} is explicitly
+initialized to the constant \co{1}, it does find the cycle and refrains
+from flagging it as an OOTA cycle.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+@@ DisplayLitmus litmus/oota-causality-9.litmus @@
+\end{verbatim}
+\caption{Causality Test Case 9}
+\label{lst:Causality Test Case 9}
+\end{listing}
+
+\subsubsection{Causality Test Case 9a}
+\label{app:Causality Test Case 9a}
+
+Listing~\ref{lst:Causality Test Case 9a}
+shows causality test case 9a, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The herd7 tool does not find this cycle, but when \co{r2} is explicitly
+initialized to the constant \co{1}, it does find the cycle and refrains
+from flagging it as an OOTA cycle.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+@@ DisplayLitmus litmus/oota-causality-9a.litmus @@
+\end{verbatim}
+\caption{Causality Test Case 9a}
+\label{lst:Causality Test Case 9a}
+\end{listing}
+
 \subsection{Additional Litmus Tests}

### 2023-12-04 ootavacuous.tex
Add JMM Causality Test Case 10
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Add and use a DisplayRunLitmus directive
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{listing}[tbp]
-\begin{verbatim}
-@@ DisplayLitmus litmus/oota-causality-1.litmus @@
-\end{verbatim}
+@@ DisplayRunLitmus litmus/oota-causality-1.litmus @@
 \caption{Causality Test Case 1}
@@
 \begin{listing}[tbp]
-\begin{verbatim}
-@@ DisplayLitmus litmus/oota-causality-2.litmus @@
-\end{verbatim}
+@@ DisplayRunLitmus litmus/oota-causality-2.litmus @@
 \caption{Causality Test Case 2}
@@
 \begin{listing}[tbp]
-\begin{verbatim}
-@@ DisplayLitmus litmus/oota-causality-3.litmus @@
-\end{verbatim}
+@@ DisplayRunLitmus litmus/oota-causality-3.litmus @@
 \caption{Causality Test Case 3}
@@
 \begin{listing}[tbp]
-\begin{verbatim}
-@@ DisplayLitmus litmus/oota-causality-4.litmus @@
-\end{verbatim}
+@@ DisplayRunLitmus litmus/oota-causality-4.litmus @@
 \caption{Causality Test Case 4}
@@
 \begin{listing}[tbp]
-\begin{verbatim}
-@@ DisplayLitmus litmus/oota-causality-5.litmus @@
-\end{verbatim}
+@@ DisplayRunLitmus litmus/oota-causality-5.litmus @@
 \caption{Causality Test Case 5}
@@
 \begin{listing}[tbp]
-\begin{verbatim}
-@@ DisplayLitmus litmus/oota-causality-6.litmus @@
-\end{verbatim}
+@@ DisplayRunLitmus litmus/oota-causality-6.litmus @@
 \caption{Causality Test Case 6}
@@
 \begin{listing}[tbp]
-\begin{verbatim}
-@@ DisplayLitmus litmus/oota-causality-7.litmus @@
-\end{verbatim}
+@@ DisplayRunLitmus litmus/oota-causality-7.litmus @@
 \caption{Causality Test Case 7}
@@
 \begin{listing}[tbp]
-\begin{verbatim}
-@@ DisplayLitmus litmus/oota-causality-8.litmus @@
-\end{verbatim}
+@@ DisplayRunLitmus litmus/oota-causality-8.litmus @@
 \caption{Causality Test Case 8}
@@
 \begin{listing}[tbp]
-\begin{verbatim}
-@@ DisplayLitmus litmus/oota-causality-9.litmus @@
-\end{verbatim}
+@@ DisplayRunLitmus litmus/oota-causality-9.litmus @@
 \caption{Causality Test Case 9}
@@
 \begin{listing}[tbp]
-\begin{verbatim}
-@@ DisplayLitmus litmus/oota-causality-9a.litmus @@
-\end{verbatim}
+@@ DisplayRunLitmus litmus/oota-causality-9a.litmus @@
 \caption{Causality Test Case 9a}
 \label{lst:Causality Test Case 9a}
+\end{listing}
+
+\subsubsection{Causality Test Case 10}
+\label{app:Causality Test Case 10}
+
+Listing~\ref{lst:Causality Test Case 10}
+shows causality test case 10, for which the \co{r1 == r2 == 1 && r3 == 0}
+result is to be forbidden.
+The herd7 tool nevertheless reports this result and does not flag it as
+an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-10.litmus @@
+\caption{Causality Test Case 10}
+\label{lst:Causality Test Case 10}
 \end{listing}
@@
 \begin{listing}[tbp]
-\begin{verbatim}
-@@ DisplayLitmus litmus/oota-non-lb.litmus @@
-\end{verbatim}
+@@ DisplayRunLitmus litmus/oota-non-lb.litmus @@
 \caption{Non-LB Litmus Test}
@@
 \begin{listing}[tbp]
-\begin{verbatim}
-@@ DisplayLitmus litmus/oota-two-source.litmus @@
-\end{verbatim}
+@@ DisplayRunLitmus litmus/oota-two-source.litmus @@
 \caption{Multi-Cycle Litmus Test}

### 2023-12-04 ootavacuous.sh

@@ -15,26 +15,46 @@
 T="`mktemp -d ${TMPDIR-/tmp}/ootavacuous.sh.XXXXXX`"
 trap 'rm -rf $T' 0
 
-awk -v q="'" -v bs='\\' < ootavacuous.tex '
+awk -v q="'" -v dq='"' -v bs='\\' < ootavacuous.tex '
 BEGIN {
-	print "cat << '___EOF___'";
+	print "cat << " sq "___EOF___" sq;
 	print "% Automatically built, do not edit";
 	print "%";
 }
 
 /^@@ DisplayLitmus .* @@$/ {
 	path = $3;
-	print "'___EOF___'";
+	print "___EOF___";
 	print "grep -v " q "^[( ]" bs "*" q " < " path " | c2latex.sh";
-	print "cat << '___EOF___'";
+	print "cat << " sq "___EOF___" sq;
+	next;
+}
+
+/^@@ DisplayRunLitmus .* @@$/ {
+	path = $3;
+	print "{";
+	print bs "scriptsize";
+	print bs "begin{verbatim}";
+	print "___EOF___";
+	print "grep -v " q "^[( ]" bs "*" q " < " path " | c2latex.sh";
+	print "cat << " sq "___EOF___" sq;
+	print bs "end{verbatim}";
+	print "";
+	print "Analysis by " dq bs "co{herd7 -c11 " path dq "}:";
+	print bs "begin{verbatim}";
+	print "___EOF___";
+	print "herd7 -c11 " path " | grep -v " sq "^$" sq " | c2latex.sh ";
+	print "cat << " sq "___EOF___" sq;
+	print bs "end{verbatim}";
+	print "}";
 	next;
 }
 
 /^@@ RunLitmus .* @@$/ {
 	path = $3;
-	print "'___EOF___'";
+	print sq "___EOF___" sq;
 	print "herd7 -c11 " path " | c2latex.sh ";
-	print "cat << '___EOF___'";
+	print "cat << ___EOF___";
 	next;
 }
 

### 2023-12-05 ootavacuous.tex
Add JMM Causality Test Cases 11-20
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 is to be allowed.
-However, the herd7 tool does not find this litmus test's cycle, presumably
+However, the \co{herd7} tool does not find this litmus test's cycle, presumably
 because this tool does not do the deep inter-thread compiler analysis
@@
 is to be allowed.
-The herd7 tool finds this cycle, and does not flag it as an OOTA cycle.
+The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
 
@@
 is to be forbidden.
-The herd7 tool finds this OOTA cycle, and flags it as such using \co{S8}.
+The \co{herd7} tool finds this OOTA cycle, and flags it as such using \co{S8}.
 
@@
 result is to be forbidden.
-The herd7 tool finds this OOTA cycle, and flags it as such using \co{S8}.
+The \co{herd7} tool finds this OOTA cycle, and flags it as such using \co{S8}.
 
@@
 result is to be allowed.
-The herd7 tool finds this cycle, and does not flag it as an OOTA cycle.
+The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
 
@@
 result is to be allowed.
-The herd7 tool does not find this cycle, but when \co{r2} is explicitly
+The \co{herd7} tool does not find this cycle, but when \co{r2} is explicitly
 initialized to the constant \co{1}, it does find the cycle and refrains
@@
 result is to be forbidden.
-The herd7 tool nevertheless reports this result and does not flag it as
+The \co{herd7} tool nevertheless reports this result and does not flag it as
 an OOTA cycle.
@@
 \label{lst:Causality Test Case 10}
+\end{listing}
+
+\subsubsection{Causality Test Case 11}
+\label{app:Causality Test Case 11}
+
+Listing~\ref{lst:Causality Test Case 11}
+shows causality test case 11, for which the \co{r1 == r2 == r3 == r4 == 1}
+result is to be allowed.
+The \co{herd7} tool reports this result and does not flag it as an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-11.litmus @@
+\caption{Causality Test Case 11}
+\label{lst:Causality Test Case 11}
+\end{listing}
+
+\subsubsection{Causality Test Case 12}
+\label{app:Causality Test Case 12}
+
+JMM Causality Test Case 12 uses arrays, which are not yet supported
+by the \co{herd7} tool.
+
+\subsubsection{Causality Test Case 13}
+\label{app:Causality Test Case 13}
+
+Listing~\ref{lst:Causality Test Case 13}
+shows causality test case 13, for which the \co{r1 == r2 == 1}
+result is to be forbidden.
+The \co{herd7} tool nevertheless reports this result and does not flag it as
+an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-13.litmus @@
+\caption{Causality Test Case 13}
+\label{lst:Causality Test Case 13}
+\end{listing}
+
+\subsubsection{Causality Test Case 14}
+\label{app:Causality Test Case 14}
+
+Listing~\ref{lst:Causality Test Case 14}
+shows causality test case 14, for which the \co{r1 == r3 == 1 && r2 == 0}
+result is to be forbidden.
+The \co{herd7} tool does not report this result.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-14.litmus @@
+\caption{Causality Test Case 14}
+\label{lst:Causality Test Case 14}
+\end{listing}
+
+\subsubsection{Causality Test Case 15}
+\label{app:Causality Test Case 15}
+
+Listing~\ref{lst:Causality Test Case 15}
+shows causality test case 15, for which the \co{r0 == r1 == r3 == 1 && r2 == 0}
+result is to be forbidden.
+The \co{herd7} tool does not report this result.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-15.litmus @@
+\caption{Causality Test Case 15}
+\label{lst:Causality Test Case 15}
+\end{listing}
+
+\subsubsection{Causality Test Case 16}
+\label{app:Causality Test Case 16}
+
+Listing~\ref{lst:Causality Test Case 16}
+shows causality test case 16, for which the \co{r1 == 2 && r2 == 1}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be expected
+given the differences between Java on the one hand and C and C++
+on the other.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-16.litmus @@
+\caption{Causality Test Case 16}
+\label{lst:Causality Test Case 16}
+\end{listing}
+
+\subsubsection{Causality Test Case 17}
+\label{app:Causality Test Case 17}
+
+Listing~\ref{lst:Causality Test Case 17}
+shows causality test case 17, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be expected
+given the differences between Java on the one hand and C and C++
+on the other.
+But note that the \co{herd7} tool correctly notes the presence of an
+OOTA cycle with \co{S17}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-17.litmus @@
+\caption{Causality Test Case 17}
+\label{lst:Causality Test Case 17}
+\end{listing}
+
+\subsubsection{Causality Test Case 18}
+\label{app:Causality Test Case 18}
+
+Listing~\ref{lst:Causality Test Case 18}
+shows causality test case 18, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be expected
+given the differences between Java on the one hand and C and C++
+on the other.
+But note that the \co{herd7} tool correctly notes the presence of an
+OOTA cycle with \co{S17}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-18.litmus @@
+\caption{Causality Test Case 18}
+\label{lst:Causality Test Case 18}
+\end{listing}
+
+\subsubsection{Causality Test Case 19}
+\label{app:Causality Test Case 19}
+
+Listing~\ref{lst:Causality Test Case 19}
+shows causality test case 19, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be expected
+given the differences between Java on the one hand and C and C++
+on the other.
+But note that the \co{herd7} tool correctly notes the presence of an
+OOTA cycle with \co{S8}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-19.litmus @@
+\caption{Causality Test Case 19}
+\label{lst:Causality Test Case 19}
+\end{listing}
+
+\subsubsection{Causality Test Case 20}
+\label{app:Causality Test Case 20}
+
+Listing~\ref{lst:Causality Test Case 20}
+shows causality test case 20, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be expected
+given the differences between Java on the one hand and C and C++
+on the other.
+But note that the \co{herd7} tool correctly notes the presence of an
+OOTA cycle with \co{S8}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-20.litmus @@
+\caption{Causality Test Case 20}
+\label{lst:Causality Test Case 20}
 \end{listing}

### 2023-12-05 ootavacuous.tex
Alan Stern December 4th feedback
Plus revert to old .pdf filename.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 during the time between those two stores.
-This means that, short of inspecting the assembly code, the user has no
-way of proving that the C++ implementation acted as if that store was
-omitted from the source code.
-In cases where this omission is undesirable, the user can resort to
+This means that, if the implementation acted as if the store was omitted
+from the source, the user would have no way to prove that fact short of
+inspecting the assembly code.
+In cases where such omissions are undesirable, the user can resort to
 volatile atomic stores or to inline assembly\footnote{

### 2023-12-05 ootavacuous.sh

@@ -70,3 +70,4 @@ pdflatex ootavacuous-flat.tex
 bibtex ootavacuous-flat.aux
 pdflatex ootavacuous-flat.tex
 pdflatex ootavacuous-flat.tex
+cp ootavacuous-flat.pdf ootavacuous.pdf

### 2023-12-05 ootavacuous.sh
Fix single-quote bugs for Deathstation 9000.
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -15,7 +15,7 @@
 T="`mktemp -d ${TMPDIR-/tmp}/ootavacuous.sh.XXXXXX`"
 trap 'rm -rf $T' 0
 
-awk -v q="'" -v dq='"' -v bs='\\' < ootavacuous.tex '
+awk -v sq="'" -v dq='"' -v bs='\\' < ootavacuous.tex '
 BEGIN {
 	print "cat << " sq "___EOF___" sq;
 	print "% Automatically built, do not edit";
@@ -25,7 +25,7 @@ BEGIN {
 /^@@ DisplayLitmus .* @@$/ {
 	path = $3;
 	print "___EOF___";
-	print "grep -v " q "^[( ]" bs "*" q " < " path " | c2latex.sh";
+	print "grep -v " sq "^[( ]" bs "*" sq " < " path " | c2latex.sh";
 	print "cat << " sq "___EOF___" sq;
 	next;
 }
@@ -36,7 +36,7 @@ BEGIN {
 	print bs "scriptsize";
 	print bs "begin{verbatim}";
 	print "___EOF___";
-	print "grep -v " q "^[( ]" bs "*" q " < " path " | c2latex.sh";
+	print "grep -v " sq "^[( ]" bs "*" sq " < " path " | c2latex.sh";
 	print "cat << " sq "___EOF___" sq;
 	print bs "end{verbatim}";
 	print "";
@@ -54,7 +54,7 @@ BEGIN {
 	path = $3;
 	print sq "___EOF___" sq;
 	print "herd7 -c11 " path " | c2latex.sh ";
-	print "cat << ___EOF___";
+	print "cat << " sq "___EOF___" sq;
 	next;
 }
 

### 2023-12-06 ootavacuous.tex
Add implementation-constraints section
While in the area, upgrade to the most recent C++ standard, remove
all references to compilers, and start the rework of the section
looking at OOTA definitions.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	Attempts to create memory models that rule out OOTA behaviors
-	have either been non-executable, complex, or unloved by compiler
-	writers.
+	have either been non-executable, complex, or unloved by C++
+	implementers.
 	But at the same time, there are no known instances of OOTA
@@
 All OOTA workers owe a debt to the foundational work in the infamous
-``Causality Test Cases''.\footnote{
+``Causality Test Cases'',\footnote{
 	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
+which may be found in
+Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}.
 
@@
 cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
-but none of these have been looked upon favorably by compiler implementers.
+but none of these have been looked upon favorably by C++ implementers.
 
@@
 
-We could follow this long-established tradition of proposing
-solutions to the OOTA problem without a definition of this problem.
-However, we (perhaps foolishly) felt the need to provide a clear
-definition, which is a more detailed version of the wording in 33.5.4p8
-(\co{[atomics.order]}):
+The C++ standard says this in 33.5.4p8
+(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
 
@@
 \end{quote}
+
+Others, current authors included, have define an OOTA cycle as a
+cycle in sdep $\cup$ rf, where sdep is the set of semantic
+dependencies within each thread and rf is the set of store-to-load
+links, whether intra-thread or between threads.
+This is a fine definition, and is consistent with the words in the C++
+standard, except that it simply shunts all the complexity into the
+term ``semantic dependency''.
+
+We could follow this long-established tradition of proposing
+solutions to the OOTA problem without a definition of this problem.
+However, we (perhaps foolishly) felt the need to provide a clear
+definition,
 
@@
 was released.
+A final set of constraints is imposed by the C++ implementations
+themselves.
 
 The following sections discuss these constraints, starting with the
-laws of physics, continuing with hardware constraints, and ending
-with constraints imposed by the standard.
+laws of physics, continuing with hardware constraints, continuing
+further with constraints imposed by the standard, and culminating
+with constraints imposed by the C++ implementations themselves.
 
@@
 
-As a special case of this prohibitino, a value stored speculatively must
+As a special case of this prohibition, a value stored speculatively must
 not be exposed to non-speculative loads from other threads.
@@
 Because volatile atomics constitute observable
-behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2022N4910},
+behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2023N4950},
 they must be executed in strict accordance with the rules of the abstract
@@
 With that line, the additional value~42 can also be stored to \co{r1}.
-The compiler is therefore forbidded from inventing that store of 42
+The implementation is therefore forbidded from inventing that store of 42
 unless it can prove that doing so does not negatively affect the
@@
 final value of \co{r1} is 1.
-The compiler is therefore forbidded from duplicating that store of 1
+The implementation is therefore forbidden from duplicating that store of 1
 unless it can prove that doing so does not negatively affect the
@@
 	clobber.}
-to prevent the compiler from omitting a store.
+to prevent the implementation from omitting a store.
 However, acting as if such a redundant store was omitted from the source
@@
 program with the store present.
+
+\subsection{C++ Implementations}
+\label{sec:C++ Implementations}
+
+C++ implementations execute C++ programs, and use a range of techniques.
+For example, many implementations take steps to optimize the code to
+be executed.
+This optimization presents a trade-off between the time spent optimizing
+the code and the time spent actually executing it.
+One type of optimization is the identification of non-semantic dependencies
+and the replacement of them with simpler computations, up to and including
+replacement of arbitrarily complex passages of code with constants.\footnote{
+	There are a great many other optimizations, but this paper
+	necessarily focuses on semantic dependencies.}
+
+This optimization process may, roughly speaking, be divided into two
+activities, analysis of the code and classification of dependencies.
+
+\subsubsection{Code Analysis}
+\label{sec:Code Analysis}
+
+As would be expected given the various optimization/execution tradeoffs,
+different C++ implementations carry out different degrees of analysis.
+
+First, at one extreme, ``simple implementations'' do absolutely no
+analysis beyond that absolutely required to correctly execute the program.
+
+Second, other implementations do local analysis.
+This analysis might be on a per-code-path basis, within a single function,
+within a single function after inlining, and so forth.
+
+Third, one could imagine implementations whose analysis is restricted to
+a thread.
+In this case, the analysis stops at loads to objects that were last
+stored to by other threads (``rfe''), stores to objects where the next
+and/or previous stores in that object's modification order were executed
+by other threads (``coe'')), and loads from objects for which the value
+returned by a given load was overwritten by some other thread (``fre'').
+
+Fourth, one could also imagine implementations whose analysis is
+completely unrestricted, up to and including the limiting case where
+execution is oracular in nature.
+Of course, if we are going to imagine omniscient analysis and oracles,
+we can also rely on them to locate any OOTA cycles that might appear
+in environments not constrained by the laws of physics, hardware
+limitations, and C++ implementation concerns.
+
+However, for more realistic implementations in the third category,
+analysis relevant to OOTA cycles begins with loads from objects that were
+last stored to by some other thread and ends with stores to objects that
+might be loaded by other threads.
+
+If additional inter-thread information is available, the corresponding
+constraints may be applied.
+For example, if an implementation determines that the values of a pair
+of objects are always equal at any time that a given thread might access
+them, one could imagine a sufficiently brave implementation substituting
+a load from one for a load from the other.
+
+The next section focuses on the key analysis step for OOTA cycles,
+namely dependency classification.
+
+\subsubsection{Dependency Classification}
+\label{sec:Dependency Classification}
+
+Carrying out this type of optimization requires that the implementation
+classify dependencies as semantic or not.
+Given the limited resources available, almost all implementations will
+choose to approximate sdep, which entails some misclassifications, with
+the consequences summarized in
+Table~\ref{tab:Semantic-Dependency Classification}.
+The first row of the table depicts the case where the implementation
+correctly identifies a semantic dependency, and thus exhibits temporal
+behavior for the corresponding code.
+The second row depicts the case where the implementation misclassifies
+the dependency as non-semantic, which will result in an incorrect
+execution.
+In this case, the proper course of action is to fix that bug.
+The third row depicts the case where the implementation misclassifies
+the dependency as semantic, thus missing a chance to optimize.
+The resulting execution is nevertheless correct, and is also temporal.
+The fourth and final row depicts the case where the implementation
+correctly classifies the dependency as non-semantic, in which case
+the execution is both correct and atemporal.
+
+\begin{table}
+\centering
+\begin{tabular}{c|c|l}
+Actual Code	& Implementation	& Result \\
+\hline
+sdep		& sdep			& Temporal \\
+\cline{2-3}
+		& $\neg$sdep		& BUG!!! \\
+\hline
+$\neg$sdep	& sdep			& Temporal \\
+\cline{2-3}
+		& $\neg$sdep		& Atemporal \\
+\end{tabular}
+\caption{Semantic-Dependency Classification}
+\label{tab:Semantic-Dependency Classification}
+\end{table}
+
+Note well that in the absence of bugs in the implementation (which
+again should be fixed), it is the implementation's classification of
+the dependency that dictates whether or not the corresponding execution
+is temporal.
+
+Therefore, in the limiting case, analyses of OOTA cycles that focus on a
+given implementation can simply use that implementation's classification
+of dependencies as semantic on the one hand or non-semantic on the other.
+While taking this approach sacrifices significant generality, it has the
+benefit of greatly reduces the cost of dependency analysis, potentially
+to zero.
 
@@
 thousands of CPUs.
-One might object to the whole concept of flattening as a compiler
+One might object to the whole concept of flattening as an
 optimization given the difficulty inherent in avoiding deadlocks
@@
 Note that if both legs of the \co{if} statement spanning lines~7-10,
-the compiler would be within its rights to hoist the store out of
+the implementation would be within its rights to hoist the store out of
 that \co{if} statement.
 However, in this case, the two stores on lines~8 and~10 are storing
-distinct values, preventing the compiler from hoisting.
+distinct values, preventing the hoisting.
 
@@
 And you need to fix those bugs, regardless of whether they are in your
-compiler, your hardware, or your alternative universe.
+C++ implementation, your hardware, or your alternative universe.
 
@@
 However, the \co{herd7} tool does not find this litmus test's cycle, presumably
-because this tool does not do the deep inter-thread compiler analysis
+because this tool does not do the deep inter-thread program analysis
 required to find it.

### 2023-12-06 ootavacuous.sh
Fix RunLitmus directive
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -8,9 +8,11 @@
 #	Use c2latex.sh to squish the tabs and number the lines.
 #	Note that "path.litmus" must not contain whitespace.
 # @@ RunLitmus path.litmus @@
-#	Use "herd7 -c11" to run the test, placing the output in
-#	path.litmus.out, then use c2latex.sh to number the lines.
-#	Note that "path.litmus" must not contain whitespace.
+#	Use "herd7 -c11" to run the test, then use c2latex.sh
+#	to number the lines.  Note that "path.litmus" must not
+#	contain whitespace.
+# @@ DisplayRunLitmus path.litmus @@
+#	Why not both?  ;-)
 
 T="`mktemp -d ${TMPDIR-/tmp}/ootavacuous.sh.XXXXXX`"
 trap 'rm -rf $T' 0
@@ -24,9 +26,14 @@ BEGIN {
 
 /^@@ DisplayLitmus .* @@$/ {
 	path = $3;
+	print "{";
+	print bs "scriptsize";
+	print bs "begin{verbatim}";
 	print "___EOF___";
 	print "grep -v " sq "^[( ]" bs "*" sq " < " path " | c2latex.sh";
 	print "cat << " sq "___EOF___" sq;
+	print bs "end{verbatim}";
+	print "}";
 	next;
 }
 
@@ -52,9 +59,14 @@ BEGIN {
 
 /^@@ RunLitmus .* @@$/ {
 	path = $3;
-	print sq "___EOF___" sq;
-	print "herd7 -c11 " path " | c2latex.sh ";
+	print "{";
+	print bs "scriptsize";
+	print bs "begin{verbatim}";
+	print "___EOF___";
+	print "herd7 -c11 " path " | grep -v " sq "^$" sq " | c2latex.sh ";
 	print "cat << " sq "___EOF___" sq;
+	print bs "end{verbatim}";
+	print "}";
 	next;
 }
 

### 2023-12-06 ootavacuous.tex
Relocate the Illustrative Examples section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{listing}[tbp]
-\begin{verbatim}
- 1 int main() {
- 2   atomic_int x=0;
- 3   atomic_int y=0;
- 4   int r1;
- 5   int r2;
- 6   {{{ { r1=x.load(memory_order_relaxed);
- 7         y.store(r1,memory_order_relaxed); }
- 8   ||| { r2=y.load(memory_order_relaxed);
- 9         x.store(r2,memory_order_relaxed); }  }}}
-10   return 0;
-11 }
-\end{verbatim}
+@@ DisplayLitmus litmus/oota-ctrl.litmus @@
 \caption{OOTA Cycle}
@@
 Listing~\ref{lst:OOTA Cycle}
-shows an OOTA cycle, where the \co{\{\{\{} on line~5 begins a set of
-threads, the \co{|||} on line~8 separates a pair of threads, and
-the \co{\}\}\}} on line~9 ends a set of threads.\footnote{
-	This is pseudocode in the format of CPPMEM,
-	which does not permit writes from non-local variables.}
-Lines~6 and~7 do a relaxed copy from \co{x} to \co{y}, and
-lines~8 and~9 to a relaxed copy from \co{y} to \co{x}.
-Even though both \co{x} and \co{y} are initialized to zero on
-lines~2 and~3, respectively, according to the mathematical core
-of the C++ memory model, the only constraint on the final values
-of \co{x} and \co{y} is that they be equal.
-Counterintuitivily, their final values could be any member of their type.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 int main() {
- 2   atomic_int x = 0;
- 3   atomic_int y = 0;
- 4   volatile int r1;
- 5   volatile int r2;
- 6   {{{ { r1 = x.load(memory_order_relaxed);
- 7         if (r1 == 1)
- 8           y.store(1, memory_order_relaxed);
- 9         else
-10           y.store(r1, memory_order_relaxed); }
-11   ||| { r2=y.load(memory_order_relaxed);
-12         x.store(r2,memory_order_relaxed); }  }}}
-13   }}};
-14   return 0;
-15 }
-\end{verbatim}
-\caption{OOTA Cycle Based on Conditionals}
-\label{lst:OOTA Cycle Based on Conditionals}
-\end{listing}
-
-A similar effect can be achieved using conditional statements, as
-shown in Listing~\ref{lst:OOTA Cycle Based on Conditionals}.
+shows an OOTA cycle.
+The first line identifies it as a C-language litmus test and gives it
+a name.
+Lines~2-5 initialize variables, in this case setting the initial
+values of the global shared variables \co{x} and \co{y} to zero.
+Lines~7-13 define the first process, \co{P0()}, and lines~15-21
+define the second process, \co{P1()}.
+The arguments to both \co{P0()} and \co{P1()} specify which of
+the global shared variables each process may access.
+In this case, both processes may access both \co{x} and \co{y}.
+The body of each process specifies a limited subset of C++ code,
+which also happens to be a somewhat less limited subset of C code.
+
+Finally, line~23 specifies an \co{exists} clause, which gives a
+condition to check for the final value of the specified local
+variables.
+The \co{0:} prefix specifies a variable local to \co{P0()} and
+the \co{1:} prefix specifies a variable local to \co{P1()}.
+The \co{/\\} specifies a boolean AND, and the \co{=} specifies
+an equality comparison.
+
+\begin{listing}[tbp]
+@@ RunLitmus litmus/oota-ctrl.litmus @@
+\caption{OOTA Cycle, \co{herd7} Output}
+\label{lst:OOTA Cycle, herd7 Output}
+\end{listing}
+
+Line~4 of the corresponding output of the \co{herd7} tool shows the
+counterintuitive outcome where both processes load the value 42.
+As we will see, this outcome is not possible in real-world implementations
+due to physical constraints of which \co{herd7} is (by design) unaware.
 
@@
 This paper focuses instead on OOTA cycles.
-
-C++ implementations may of course avail themselves of the as-if rule.
-However, correct use of the as-if rule requires that the observable
-behaviors of the program be appropriately preserved.
 
@@
 the value loaded.
-This dependency might be a data dependency
+This dependency might be a control dependency
 (see Listing~\ref{lst:OOTA Cycle})
-or it might be a control dependency
-(see Listing~\ref{lst:OOTA Cycle Based on Conditionals}).\footnote{
+or it might be a data dependency
+(see Listing~\ref{lst:Causality Test Case 4} on
+page~\pageref{lst:Causality Test Case 4}).\footnote{
 	Sufficiently motivated readers should have no problem constructing
@@
 that were involved.
+
+Applying this to Listing~\ref{lst:OOTA Cycle}, because values from
+speculative stores cannot be visible to non-speculative loads, lines~10
+and~12 cannot be committed until after the load in line~8 completes
+and the conditional on line~9 is evaluated non-speculatively.\footnote{
+	Alternatively, one could imagine that both the speculation
+	and the squashing of failed speculation might span multiple
+	threads.}
+Similarly, lines~18 and~20 cannot be committed until after lines~16
+and~17 complete non-speculatively.
+
+Note that if both legs of the \co{if} statement spanning lines~9-12 were
+to store the same value, the implementation would be within its rights
+to hoist the store out of that \co{if} statement.
+However, in this case, the two stores on lines~10 and~12 are storing
+distinct values, preventing the hoisting.
+
+Note also that some architectures provide conditional-move instructions,
+which allows line~7 to avoid emitting a conditional branch.
+However, such architectures would still treat a store of the
+target of the conditional-move instructions as speculative until
+such time as the conditional moves (and any loads that they depend
+on) complete.
 
@@
 which in turn implies that OOTA cycles cannot form.
-
-\subsection{Illustrative Examples}
-\label{sec:Illustrative Examples}
-
-This section works through OOTA examples, showing the effects of
-the constraints called out in this paper.
-
-\subsubsection{OOTA Cycle}
-\label{sec:OOTA Cycle}
-
-Listing~\ref{lst:OOTA Cycle}
-shows the canonical example of an OOTA cycle.
-Because values from speculative stores cannot be visible to other threads,
-line~7 cannot be committed until after line~6 completes.
-Similarly, line~9 cannot be committed until after line~8 completes.
-
-Because values stored take non-zero time to propagate from one thread
-to another, no cycle can form, and thus no OOTA cycle can form.
-
-\subsubsection{OOTA Cycle Based on Conditionals}
-\label{sec:OOTA Cycle Based on Conditionals}
-
-Listing~\ref{lst:OOTA Cycle Based on Conditionals}
-shows an if-condition-based example of an OOTA cycle.
-Because values from speculative stores cannot be visible to other threads,
-lines~8 and~10 cannot be committed until after the load in line~6 completes
-and the conditional on line~7 is evaluated non-speculatively.
-Similarly, line~12 cannot be committed until after line~11 completes.
-
-Because values stored take non-zero time to propagate from one thread
-to another, no cycle can form, and thus no OOTA cycle can form.
-
-Note that if both legs of the \co{if} statement spanning lines~7-10,
-the implementation would be within its rights to hoist the store out of
-that \co{if} statement.
-However, in this case, the two stores on lines~8 and~10 are storing
-distinct values, preventing the hoisting.
-
-Note also that some architectures provide conditional-move instructions,
-which allows line~7 to avoid emitting a conditional branch.
-However, such architectures would still treat a store of the
-target of the conditional-move instructions as speculative until
-such time as the conditional moves (and any loads that they depend
-on) complete.
 

### 2023-12-06 ootavacuous.tex
Add verbiage and tools for dependency classification
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-We could follow this long-established tradition of proposing
-solutions to the OOTA problem without a definition of this problem.
-However, we (perhaps foolishly) felt the need to provide a clear
-definition,
-
-We therefore define an OOTA cycle (also known as a causal cycle) as a
-cycle in the memory-model sense involving only reads-from links, that
-is, the value stored in on thread being read by the next thread.
-Within each thread in the cycle, the value stored depends on
-the value loaded.
-This dependency might be a control dependency
-(see Listing~\ref{lst:OOTA Cycle})
-or it might be a data dependency
-(see Listing~\ref{lst:Causality Test Case 4} on
-page~\pageref{lst:Causality Test Case 4}).\footnote{
-	Sufficiently motivated readers should have no problem constructing
-	examples involving address dependencies.}
-@@@ There must be at least two distinct values such that if a given load
-returns a given value, that value forces the cycle to execute so as to
-cause the corresponding store's value to be equal to that given value, and
-for that store to occur early enough for the load to return that value.
+One aspect of this complexity is the fact that semantic dependencies
+are a function not strictly of the source code, but of a particular
+execution.
+Consider for example:
+
+\begin{quote}
+	\co{x = y * z;}
+\end{quote}
+
+There is a semantic dependency from \co{y} to \co{x} if and only if the
+value of \co{z} is non-zero.
+Otherwise, in the absence of observable behavior (such as volatile
+accesses), the implementation is within its rights to discard this
+statement and act as if later accesses to \co{x} were instead the
+constant zero, at least up to the next assignment to \co{x}.
+
+Similarly:
+
+\begin{quote}
+	\co{x = y + z;}
+\end{quote}
+
+There is a semantic dependency from both \co{y} and \co{z} if and only
+if these values do not cancel.
+However, if the implementation knows that the value of \co{y} is always
+the negative of \co{x}, there is no semantic dependency and the
+implementation can again act as if later accesses to \co{x} were
+instead the constant zero.
+
+And perhaps worse:
+
+\begin{quote}
+	\co{x = horribly_complex(y, z);}
+\end{quote}
+
+Here, the presence or absence of a semantic dependency from \co{y} to
+\co{x} depends not only on the value of \co{z}, but also the code in
+a horribly complex function.
+
+Fortunately, there are now tools that can help.
+The SAT-solver-based \co{cbmc} tool, which is readily available,\footnote{
+	On Debian-based distributions, via \co{sudo apt install cbmc}.}
+has been used to mechanically verify signicant portions of Linux-kernel
+RCU from the C-language source
+code~\cite{LihaoLiang2016VerifyTreeRCU,LanceRoy2017CBMC-SRCU}.
+Another tool, Nidhugg, which is based on partial-order
+reduction, has been used to carry out a similar
+verification~\cite{MichalisKokologiannakis2017NidhuggRCU,SMC-TreeRCU,MichalisKokologiannakis2019RCUstatelessModelCheck}.
+Both tools can easily check whether or not a given dependency is semantic.
+% @@@ More on cbmc in an appendix.
+
+However, as will be seen in later sections, in some special but
+commonly occurring circumstances, there are easier ways.
+
+% @@@ We could follow this long-established tradition of proposing
+% solutions to the OOTA problem without a definition of this problem.
+% However, we (perhaps foolishly) felt the need to provide a clear
+% definition,
+
+% We therefore define an OOTA cycle (also known as a causal cycle) as a
+% cycle in the memory-model sense involving only reads-from links, that
+% is, the value stored in on thread being read by the next thread.
+% Within each thread in the cycle, the value stored depends on
+% the value loaded.
+% This dependency might be a control dependency
+% (see Listing~\ref{lst:OOTA Cycle})
+% or it might be a data dependency
+% (see Listing~\ref{lst:Causality Test Case 4} on
+% page~\pageref{lst:Causality Test Case 4}).\footnote{
+%	Sufficiently motivated readers should have no problem constructing
+%	examples involving address dependencies.}
+% @@@ There must be at least two distinct values such that if a given load
+% returns a given value, that value forces the cycle to execute so as to
+% cause the corresponding store's value to be equal to that given value, and
+% for that store to occur early enough for the load to return that value.
 

### 2023-12-06 ootavacuous.tex
Self-review December 5 evening.
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Some executable C++ memory models correctly flagged executions involving
-OOTA~\cite{JadeAlglave2014HerdingCats}.\footnote{
+Some executable C++ memory models correctly flag at least some executions
+involving OOTA cycles~\cite{JadeAlglave2014HerdingCats}.\footnote{
 	Others cleverly avoid this issue by forbidding atomic
@@
 
+But suppose that the implementation can prove that at any time that
+the above statement might execute, \co{y} is equal to \co{z}?
+That would allow the implementation to act as if the source code
+was instead as follows:
+
+\begin{quote}
+	\co{x = 2 * y;}
+\end{quote}
+
+Or equally valid:
+
+\begin{quote}
+	\co{x = 2 * z;}
+\end{quote}
+
+In this case, is the semantic dependency from \co{y} to \co{x},
+from \co{z} to \co{x}, or both \co{y} and \co{z} to \co{x}?
+Or is this a free choice on the part of the implementation?
+
 And perhaps worse:
@@
 Fortunately, there are now tools that can help.
-The SAT-solver-based \co{cbmc} tool, which is readily available,\footnote{
+For example, the SAT-solver-based \co{cbmc} tool, which is readily
+available,\footnote{
 	On Debian-based distributions, via \co{sudo apt install cbmc}.}
@@
 
+There are also many other tools, but failing that, it is always possible
+to fall back to the oldest software-verification tool, namely manual
+code inspection.
+
 However, as will be seen in later sections, in some special but
-commonly occurring circumstances, there are easier ways.
+commonly occurring circumstances, much simpler methods exist.
 
@@
 In contrast, a pair of back-to-back non-volatile atomic stores to \co{x}
-might always be executed such that no other thread accesses \co{x}
+might be executed such that no other thread accesses \co{x}
 during the time between those two stores.
-This means that, if the implementation acted as if the store was omitted
+This means that, if the implementation acted as if the first store was omitted
 from the source, the user would have no way to prove that fact short of
@@
 a load from one for a load from the other.
+
+However, if all of the inter-thread relaxed accesses are volatile,
+as they are in the C language, then all of the inter-thread traffic is
+observed behavior.
+In this case, even omniscient oracular implementations are required
+to confine their analyses within a single thread.
 
@@
 
+\subsubsection{C++ Implementations: Consequences}
+\label{sec:C++ Implementations: Consequences}
+
+Given a correct C++ implementation, a given store will either be
+executed as if is was semantically dependent on a given prior load or
+not.
+In the former case on a real-world implementation, a non-negative interval
+of time will elapse between the load and the store, and in the latter
+case, that load-store pair cannot possibly be a member of an OOTA cycle.
+
+Of course, dependencies can be separately analyzed, and such analysis
+might show that one of the dependencies that the implementation treated
+as a semantic dependency was in fact non-semantic.
+And this is a belt-and-suspenders situation:
+The corresponding load-store pair cannot be a member of an OOTA cycle,
+and a non-negative interval of time will elapse between the load and
+the store.
+
 \section{How to Form an OOTA Cycle?}
@@
 the corresponding thread's prior load.
-All of the links from ont thread to the next are reads-from links, which
+All of the links from one thread to the next are reads-from links, which
 are temporal in nature, that is, a given store must execute earlier in
@@
 
-And again, an OOTA cycle must have at least one link that goes backwards
-in time.
-The only possible candidate counter-temporal links so are those
-confined to a given thread.
+But it is also the case that an OOTA cycle must have at least one link
+that goes backwards in time.
+Given that the inter-thread links are temporal, the only possible
+candidate atemporal links so are those confined to a given thread.
 This in turn requires that a store that depends on an earlier load
@@
 
-But for the purposes of OOTA cycles, the flattening situation is
-much simpler.
-Any fully flattened program will be single threaded, which implies that
-each of its atomic load operations will return the last value stored to
-that same object in sequenced-before order (or the initial value if there
-is no such store).
+The reason for this answer is that any fully flattened program will be
+single threaded, which implies that each of its atomic load operations
+will return the last value stored to that same object in sequenced-before
+order (or the initial value if there is no such store).
 Thus, any sequence of loads and stores must have a first operation and
@@
 And you need to fix those bugs, regardless of whether they are in your
-C++ implementation, your hardware, or your alternative universe.
+C++ implementation, your hardware, or your alternate universe.
 
@@
 We are grateful to David Goldblatt, Jade Alglave, and Alan Stern for their
-careful review of an early draft of this paper and to John Wikerson for
+careful review of an early draft of this paper and to John Wickerson for
 asking Paul for a rant and taking the proffered rant seriously.

### 2023-12-06 ootavacuous.tex
C++ implementations need not use cbmc/nidhugg/whatever
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-However, as will be seen in later sections, in some special but
-commonly occurring circumstances, much simpler methods exist.
+All of these tools can incur high overhead, especially manual code
+inspection.
+However, as we will see, C++ implementations need not use these tools.
+Their current analysis is guanteed to be sufficient.
+Furthermore, as we will also see, in some special but commonly occurring
+circumstances, much simpler methods exist that incur negligible overhead.
 

### 2023-12-06 ootavacuous.tex
Add example from P1916R0 and note per-execution sdep
Reported-by: David Goldblatt <davidtgoldblatt@gmail.com>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 general notion of OOTA values in cases involving unspecified or undefined
-behavior, such as use of uninitialized objects.
+behavior (UB), such as use of uninitialized objects.
 However, these situations do not involve OOTA cycles, but rather user
@@
 
+Goldblatt looked at interactions between OOTA cycles and
+UB~\cite{DavidGoldblatt2019NoElegantOOTAfix}.
+Appendix~\ref{app:Aside on Undefined Behavior}
+analyzes the examples and proposes a way to separate UB and OOTA-cycle
+concerns.
+The remainder of this document will focus on examples lacking UB.
+
 All this work focused on either identifying OOTA or on how C++
@@
 instead the constant zero.
+
+These examples demonstrate a key point: Although in some cases sdep can
+be a strict function of the source code, \emph{in general, sdep must be
+defined on a per-execution basis}.
 
@@
 the address of \co{x}.\footnote{
-	Yes, \co{y[-1]} is undefined behavior, but the CPU neither knows
+	Yes, \co{y[-1]} is UB, but the CPU neither knows
 	nor cares, nor should it.}
@@
 
+\section{Aside on Undefined Behavior}
+\label{app:Aside on Undefined Behavior}
+
+The combination of OOTA cycles and UB has proven especially
+challenging~\cite{DavidGoldblatt2019NoElegantOOTAfix}.
+
+\subsection{Alignment-Based Undefined Behavior}
+\label{sec:Alignment-Based Undefined Behavior}
+
+The example shown in
+Listing~\ref{lst:Alignment-Based Undefined Behavior (P1916R0)}
+(taken from P1916R0) is especially instructive.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 std::atomic<int> y;
+ 2 template <typename T>
+ 3 void t1() {
+ 4   long long r1 = x.load(std::memory_order_relaxed);
+ 5   if (r1 % 4 == 0)
+ 6     y.store(1, std::memory_order_relaxed);
+ 7   *(T*)r1 = *(T*)r1;
+ 8 }
+\end{verbatim}
+\caption{Alignment-Based Undefined Behavior (P1916R0)}
+\label{lst:Alignment-Based Undefined Behavior (P1916R0)}
+\end{listing}
+
+Because the compiler often assumes that UB cannot happen, it might
+look at line~7 and conclude that the value of \co{r1} necessarily
+corresponded to an address aligned per the requirements of type \co{T}.
+If that type's alignment was larger than four bytes, the compiler could
+then replace the condition in line~5 with \co{true}, which would mean
+that there was no semantic dependency from the load of \co{x} to the
+store of \co{y}.
+
+Except that the developer might have knowingly carried out a misaligned
+access, for example, to test an alignment trap handler or because the
+system on which this code was to run was known to gracefully handle
+misaligned accesses.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 std::atomic<int> y;
+ 2 template <typename T>
+ 3 void t1() {
+ 4   long long r1 = x.load(std::memory_order_relaxed);
+ 5   if (r1 % 4 == 0)
+ 6     y.store(1, std::memory_order_relaxed);
+ 7   __asm__ __volatile__("": : :"mightnotreturn");
+ 8   *(T*)r1 = *(T*)r1;
+ 9 }
+\end{verbatim}
+\caption{Alignment-Based Undefined Behavior Fix}
+\label{lst:Alignment-Based Undefined Behavior Fix}
+\end{listing}
+
+However, in that case, the developer is stepping outside the bounds
+of the standard.
+It would be good to have some way for the developer to get the job
+done without confusing the compiler, thus separating OOTA-cycle and
+UB concerns.
+One approach is to have a \co{mightnotreturn} clobber similar to the
+\co{memory} clobber for GCC asms.
+Then this example could be rewritten as shown in
+Listing~\ref{lst:Alignment-Based Undefined Behavior Fix},
+with the asm on line~7 preventing the compiler from assuming that line~8
+is guaranteed to be executed, thus preventing the compiler from
+applying any lessons that it might learn from line~8 to line~5.
+
+% Clean up <200b> in vim: ":%s/\%u200b//g"
+
+This \co{mightnotreturn} clobber is merely an example.
+There might well be better ways to prevent UB from propagating backwards.
+Other options being considered include a call to an empty function that
+is invisible to the compiler and a GCC goto asm that the compiler cannot
+prove falls through.
+
+\subsection{Unreachability-Based Undefined Behavior}
+\label{sec:Unreachability-Based Undefined Behavior}
+
+This section attempts to apply the lessons
+of Appendix~\ref{sec:Unreachability-Based Undefined Behavior} to
+Goldblatt's unreachability-UB example.  Inserting the asm at line~12
+as shown in Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
+avoids confusing the compiler into assuming that the value of \co{local}
+must necessarily be zero.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 // UB if this function is ever called.
+ 2 [[noreturn]]
+ 3 inline void unreachable() {
+ 4   return;
+ 5 }
+ 6 void t1() {
+ 7   r1 = x.load(std::memory_order_relaxed);
+ 8   int local = (r1 & 1);
+ 9   if (local == 0)
+10     y.store(1, std::memory_order_relaxed);
+11   if (local != 0) {
+12     __asm__ __volatile__("": : :"mightnotreturn");
+13     // If we get here, UB.
+14     unreachable();
+15   }
+16 }
+\end{verbatim}
+\caption{Unreachability-Based Undefined Behavior Fix}
+\label{lst:Unreachability-Based Undefined Behavior Fix}
+\end{listing}
+
+By the same reasoning, Goldblatt's \co{misbhave} example that
+also involves unreachability can also be addressed through insertion
+of the asm.
+
 % \section{History}

### 2023-12-07 cbmc/sdep-quadratic.c
Add sdep-quadratic.c for cbmc sdep demonstration
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -1,0 +1,23 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include <assert.h>
+
+int main(int argc, char *argv[])
+{
+	int i;
+	int y[2] = { 0, 0 };
+
+	for (i = 0; i < 2; i++) {
+		int r1;
+		int r2;
+		int r3;
+		int *yp = &y[i];
+
+		r1 = argv[2 * i + 1][0] & 0x7;
+		// r2 = 0; // Comment out for sdep
+		r2 = argv[2 * i + 2][0] & 0x7; // Comment out for !sdep
+		r3 = r1 * r1 + 2 * r1 + 2;
+		*yp = r2 <= r3 ? r2 : r3;
+	}
+	assert(y[0] == y[1]);
+}

### 2023-12-07 ootavacuous.tex
Move acknowledgments and get todo list electronic
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+\appendix
+
+\section{Demonstration of OOTA-Cycle Definition}
+\label{app:Demonstration of OOTA-Cycle Definition}
+
+\subsection{Litmus Tests from “Causality Test Cases"}
+\label{app:Litmus Tests from “Causality Test Cases"}
+
+\subsubsection{Causality Test Case 1}
+\label{app:Causality Test Case 1}
+
+Listing~\ref{lst:Causality Test Case 1}
+shows causality test case 1, for which the \co{r1 == r2 == 1} result
+is to be allowed.
+However, the \co{herd7} tool does not find this litmus test's cycle, presumably
+because this tool does not do the deep inter-thread program analysis
+required to find it.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-1.litmus @@
+\caption{Causality Test Case 1}
+\label{lst:Causality Test Case 1}
+\end{listing}
+
+\subsubsection{Causality Test Case 2}
+\label{app:Causality Test Case 2}
+
+Listing~\ref{lst:Causality Test Case 2}
+shows causality test case 2, for which the \co{r1 == r2 == r3 == 1} result
+is to be allowed.
+The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-2.litmus @@
+\caption{Causality Test Case 2}
+\label{lst:Causality Test Case 2}
+\end{listing}
+
+\subsubsection{Causality Test Case 3}
+\label{app:Causality Test Case 3}
+
+Listing~\ref{lst:Causality Test Case 3}
+shows causality test case 3, for which the \co{r1 == r2 == r3 == 1} result
+is to be allowed.
+The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-3.litmus @@
+\caption{Causality Test Case 3}
+\label{lst:Causality Test Case 3}
+\end{listing}
+
+\subsubsection{Causality Test Case 4}
+\label{app:Causality Test Case 4}
+
+Listing~\ref{lst:Causality Test Case 4}
+shows causality test case 4, for which the \co{r1 == r2 == 1} result
+is to be forbidden.
+The \co{herd7} tool finds this OOTA cycle, and flags it as such using \co{S8}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-4.litmus @@
+\caption{Causality Test Case 4}
+\label{lst:Causality Test Case 4}
+\end{listing}
+
+\subsubsection{Causality Test Case 5}
+\label{app:Causality Test Case 5}
+
+Listing~\ref{lst:Causality Test Case 5}
+shows causality test case 5, for which the \co{r1 == r2 == 1 && r3 == 0}
+result is to be forbidden.
+The \co{herd7} tool finds this OOTA cycle, and flags it as such using \co{S8}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-5.litmus @@
+\caption{Causality Test Case 5}
+\label{lst:Causality Test Case 5}
+\end{listing}
+
+\subsubsection{Causality Test Case 6}
+\label{app:Causality Test Case 6}
+
+Listing~\ref{lst:Causality Test Case 6}
+shows causality test case 6, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-6.litmus @@
+\caption{Causality Test Case 6}
+\label{lst:Causality Test Case 6}
+\end{listing}
+
+\subsubsection{Causality Test Case 7}
+\label{app:Causality Test Case 7}
+
+Listing~\ref{lst:Causality Test Case 7}
+shows causality test case 7, for which the \co{r1 == r2 == r3 == 1}
+result is to be allowed.
+The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-7.litmus @@
+\caption{Causality Test Case 7}
+\label{lst:Causality Test Case 7}
+\end{listing}
+
+\subsubsection{Causality Test Case 8}
+\label{app:Causality Test Case 8}
+
+Listing~\ref{lst:Causality Test Case 8}
+shows causality test case 8, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The \co{herd7} tool does not find this cycle, but when \co{r2} is explicitly
+initialized to the constant \co{1}, it does find the cycle and refrains
+from flagging it as an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-8.litmus @@
+\caption{Causality Test Case 8}
+\label{lst:Causality Test Case 8}
+\end{listing}
+
+\subsubsection{Causality Test Case 9}
+\label{app:Causality Test Case 9}
+
+Listing~\ref{lst:Causality Test Case 9}
+shows causality test case 9, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The \co{herd7} tool does not find this cycle, but when \co{r2} is explicitly
+initialized to the constant \co{1}, it does find the cycle and refrains
+from flagging it as an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-9.litmus @@
+\caption{Causality Test Case 9}
+\label{lst:Causality Test Case 9}
+\end{listing}
+
+\subsubsection{Causality Test Case 9a}
+\label{app:Causality Test Case 9a}
+
+Listing~\ref{lst:Causality Test Case 9a}
+shows causality test case 9a, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The \co{herd7} tool does not find this cycle, but when \co{r2} is explicitly
+initialized to the constant \co{1}, it does find the cycle and refrains
+from flagging it as an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-9a.litmus @@
+\caption{Causality Test Case 9a}
+\label{lst:Causality Test Case 9a}
+\end{listing}
+
+\subsubsection{Causality Test Case 10}
+\label{app:Causality Test Case 10}
+
+Listing~\ref{lst:Causality Test Case 10}
+shows causality test case 10, for which the \co{r1 == r2 == 1 && r3 == 0}
+result is to be forbidden.
+The \co{herd7} tool nevertheless reports this result and does not flag it as
+an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-10.litmus @@
+\caption{Causality Test Case 10}
+\label{lst:Causality Test Case 10}
+\end{listing}
+
+\subsubsection{Causality Test Case 11}
+\label{app:Causality Test Case 11}
+
+Listing~\ref{lst:Causality Test Case 11}
+shows causality test case 11, for which the \co{r1 == r2 == r3 == r4 == 1}
+result is to be allowed.
+The \co{herd7} tool reports this result and does not flag it as an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-11.litmus @@
+\caption{Causality Test Case 11}
+\label{lst:Causality Test Case 11}
+\end{listing}
+
+\subsubsection{Causality Test Case 12}
+\label{app:Causality Test Case 12}
+
+JMM Causality Test Case 12 uses arrays, which are not yet supported
+by the \co{herd7} tool.
+
+\subsubsection{Causality Test Case 13}
+\label{app:Causality Test Case 13}
+
+Listing~\ref{lst:Causality Test Case 13}
+shows causality test case 13, for which the \co{r1 == r2 == 1}
+result is to be forbidden.
+The \co{herd7} tool nevertheless reports this result and does not flag it as
+an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-13.litmus @@
+\caption{Causality Test Case 13}
+\label{lst:Causality Test Case 13}
+\end{listing}
+
+\subsubsection{Causality Test Case 14}
+\label{app:Causality Test Case 14}
+
+Listing~\ref{lst:Causality Test Case 14}
+shows causality test case 14, for which the \co{r1 == r3 == 1 && r2 == 0}
+result is to be forbidden.
+The \co{herd7} tool does not report this result.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-14.litmus @@
+\caption{Causality Test Case 14}
+\label{lst:Causality Test Case 14}
+\end{listing}
+
+\subsubsection{Causality Test Case 15}
+\label{app:Causality Test Case 15}
+
+Listing~\ref{lst:Causality Test Case 15}
+shows causality test case 15, for which the \co{r0 == r1 == r3 == 1 && r2 == 0}
+result is to be forbidden.
+The \co{herd7} tool does not report this result.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-15.litmus @@
+\caption{Causality Test Case 15}
+\label{lst:Causality Test Case 15}
+\end{listing}
+
+\subsubsection{Causality Test Case 16}
+\label{app:Causality Test Case 16}
+
+Listing~\ref{lst:Causality Test Case 16}
+shows causality test case 16, for which the \co{r1 == 2 && r2 == 1}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be expected
+given the differences between Java on the one hand and C and C++
+on the other.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-16.litmus @@
+\caption{Causality Test Case 16}
+\label{lst:Causality Test Case 16}
+\end{listing}
+
+\subsubsection{Causality Test Case 17}
+\label{app:Causality Test Case 17}
+
+Listing~\ref{lst:Causality Test Case 17}
+shows causality test case 17, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be expected
+given the differences between Java on the one hand and C and C++
+on the other.
+But note that the \co{herd7} tool correctly notes the presence of an
+OOTA cycle with \co{S17}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-17.litmus @@
+\caption{Causality Test Case 17}
+\label{lst:Causality Test Case 17}
+\end{listing}
+
+\subsubsection{Causality Test Case 18}
+\label{app:Causality Test Case 18}
+
+Listing~\ref{lst:Causality Test Case 18}
+shows causality test case 18, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be expected
+given the differences between Java on the one hand and C and C++
+on the other.
+But note that the \co{herd7} tool correctly notes the presence of an
+OOTA cycle with \co{S17}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-18.litmus @@
+\caption{Causality Test Case 18}
+\label{lst:Causality Test Case 18}
+\end{listing}
+
+\subsubsection{Causality Test Case 19}
+\label{app:Causality Test Case 19}
+
+Listing~\ref{lst:Causality Test Case 19}
+shows causality test case 19, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be expected
+given the differences between Java on the one hand and C and C++
+on the other.
+But note that the \co{herd7} tool correctly notes the presence of an
+OOTA cycle with \co{S8}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-19.litmus @@
+\caption{Causality Test Case 19}
+\label{lst:Causality Test Case 19}
+\end{listing}
+
+\subsubsection{Causality Test Case 20}
+\label{app:Causality Test Case 20}
+
+Listing~\ref{lst:Causality Test Case 20}
+shows causality test case 20, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be expected
+given the differences between Java on the one hand and C and C++
+on the other.
+But note that the \co{herd7} tool correctly notes the presence of an
+OOTA cycle with \co{S8}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-20.litmus @@
+\caption{Causality Test Case 20}
+\label{lst:Causality Test Case 20}
+\end{listing}
+
+\subsection{Additional Litmus Tests}
+\label{app:Additional Litmus Tests}
+
+\subsubsection{Non-LB Litmus Tests}
+\label{app:Non-LB Litmus Tests}
+
+Listing~\ref{lst:Non-LB Litmus Test}
+shows a litmus test that can produce an OOTA cycle, but that is not
+a member of the load-buffering (LB) class of litmus tests.
+Suppose that lines~9 and~10 both somehow return the value 21.
+Then line~12 will store the value 42 to \co{x}, which will cause
+line~17 to store 21 to \co{y} and line~22 to store 21 to \co{z},
+thus justifying the values initially loaded by lines~9 and~10.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-non-lb.litmus @@
+\caption{Non-LB Litmus Test}
+\label{lst:Non-LB Litmus Test}
+\end{listing}
+
+\subsubsection{Multi-Cycle Litmus Tests}
+\label{app:Multi-Cycle Litmus Tests}
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-two-source.litmus @@
+\caption{Multi-Cycle Litmus Test}
+\label{lst:Multi-Cycle Litmus Test}
+\end{listing}
+
+\section{Aside on Undefined Behavior}
+\label{app:Aside on Undefined Behavior}
+
+The combination of OOTA cycles and UB has proven especially
+challenging~\cite{DavidGoldblatt2019NoElegantOOTAfix}.
+
+\subsection{Alignment-Based Undefined Behavior}
+\label{sec:Alignment-Based Undefined Behavior}
+
+The example shown in
+Listing~\ref{lst:Alignment-Based Undefined Behavior (P1916R0)}
+(taken from P1916R0) is especially instructive.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 std::atomic<int> y;
+ 2 template <typename T>
+ 3 void t1() {
+ 4   long long r1 = x.load(std::memory_order_relaxed);
+ 5   if (r1 % 4 == 0)
+ 6     y.store(1, std::memory_order_relaxed);
+ 7   *(T*)r1 = *(T*)r1;
+ 8 }
+\end{verbatim}
+\caption{Alignment-Based Undefined Behavior (P1916R0)}
+\label{lst:Alignment-Based Undefined Behavior (P1916R0)}
+\end{listing}
+
+Because the compiler often assumes that UB cannot happen, it might
+look at line~7 and conclude that the value of \co{r1} necessarily
+corresponded to an address aligned per the requirements of type \co{T}.
+If that type's alignment was larger than four bytes, the compiler could
+then replace the condition in line~5 with \co{true}, which would mean
+that there was no semantic dependency from the load of \co{x} to the
+store of \co{y}.
+
+Except that the developer might have knowingly carried out a misaligned
+access, for example, to test an alignment trap handler or because the
+system on which this code was to run was known to gracefully handle
+misaligned accesses.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 std::atomic<int> y;
+ 2 template <typename T>
+ 3 void t1() {
+ 4   long long r1 = x.load(std::memory_order_relaxed);
+ 5   if (r1 % 4 == 0)
+ 6     y.store(1, std::memory_order_relaxed);
+ 7   __asm__ __volatile__("": : :"mightnotreturn");
+ 8   *(T*)r1 = *(T*)r1;
+ 9 }
+\end{verbatim}
+\caption{Alignment-Based Undefined Behavior Fix}
+\label{lst:Alignment-Based Undefined Behavior Fix}
+\end{listing}
+
+However, in that case, the developer is stepping outside the bounds
+of the standard.
+It would be good to have some way for the developer to get the job
+done without confusing the compiler, thus separating OOTA-cycle and
+UB concerns.
+One approach is to have a \co{mightnotreturn} clobber similar to the
+\co{memory} clobber for GCC asms.
+Then this example could be rewritten as shown in
+Listing~\ref{lst:Alignment-Based Undefined Behavior Fix},
+with the asm on line~7 preventing the compiler from assuming that line~8
+is guaranteed to be executed, thus preventing the compiler from
+applying any lessons that it might learn from line~8 to line~5.
+
+% Clean up <200b> in vim: ":%s/\%u200b//g"
+
+This \co{mightnotreturn} clobber is merely an example.
+There might well be better ways to prevent UB from propagating backwards.
+Other options being considered include a call to an empty function that
+is invisible to the compiler and a GCC goto asm that the compiler cannot
+prove falls through.
+
+\subsection{Unreachability-Based Undefined Behavior}
+\label{sec:Unreachability-Based Undefined Behavior}
+
+This section attempts to apply the lessons
+of Appendix~\ref{sec:Unreachability-Based Undefined Behavior} to
+Goldblatt's unreachability-UB example.  Inserting the asm at line~12
+as shown in Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
+avoids confusing the compiler into assuming that the value of \co{local}
+must necessarily be zero.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 // UB if this function is ever called.
+ 2 [[noreturn]]
+ 3 inline void unreachable() {
+ 4   return;
+ 5 }
+ 6 void t1() {
+ 7   r1 = x.load(std::memory_order_relaxed);
+ 8   int local = (r1 & 1);
+ 9   if (local == 0)
+10     y.store(1, std::memory_order_relaxed);
+11   if (local != 0) {
+12     __asm__ __volatile__("": : :"mightnotreturn");
+13     // If we get here, UB.
+14     unreachable();
+15   }
+16 }
+\end{verbatim}
+\caption{Unreachability-Based Undefined Behavior Fix}
+\label{lst:Unreachability-Based Undefined Behavior Fix}
+\end{listing}
+
+By the same reasoning, Goldblatt's \co{misbhave} example that
+also involves unreachability can also be addressed through insertion
+of the asm.
+
+% \section{History}
+% \label{sec:History}
+
 \section{Acknowledgments}
@@
 We also owe David Goldblatt a debt of gratitude for his having asked an
-insightful question at the right time, and for his ``Deathstation 9000''
-demonic CPU.
+insightful question at the right time, his insights on combinations of
+OOTA and UB, and for his ``Deathstation 9000'' demonic CPU.
+Alan Stern located many unclear statements and gaps in reasoning.
+Martin Uecker contributed valuable insights on backwards-propagating UB.
 Nonetheless, all errors and omissions in this paper are the sole property
@@
 
-% To do:
+% @@@ reviewer to do.
 % Luc Maranget: Memory models.
@@
 
-\appendix
-
-\section{Demonstration of OOTA-Cycle Definition}
-\label{app:Demonstration of OOTA-Cycle Definition}
-
-\subsection{Litmus Tests from “Causality Test Cases"}
-\label{app:Litmus Tests from “Causality Test Cases"}
-
-\subsubsection{Causality Test Case 1}
-\label{app:Causality Test Case 1}
-
-Listing~\ref{lst:Causality Test Case 1}
-shows causality test case 1, for which the \co{r1 == r2 == 1} result
-is to be allowed.
-However, the \co{herd7} tool does not find this litmus test's cycle, presumably
-because this tool does not do the deep inter-thread program analysis
-required to find it.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-1.litmus @@
-\caption{Causality Test Case 1}
-\label{lst:Causality Test Case 1}
-\end{listing}
-
-\subsubsection{Causality Test Case 2}
-\label{app:Causality Test Case 2}
-
-Listing~\ref{lst:Causality Test Case 2}
-shows causality test case 2, for which the \co{r1 == r2 == r3 == 1} result
-is to be allowed.
-The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-2.litmus @@
-\caption{Causality Test Case 2}
-\label{lst:Causality Test Case 2}
-\end{listing}
-
-\subsubsection{Causality Test Case 3}
-\label{app:Causality Test Case 3}
-
-Listing~\ref{lst:Causality Test Case 3}
-shows causality test case 3, for which the \co{r1 == r2 == r3 == 1} result
-is to be allowed.
-The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-3.litmus @@
-\caption{Causality Test Case 3}
-\label{lst:Causality Test Case 3}
-\end{listing}
-
-\subsubsection{Causality Test Case 4}
-\label{app:Causality Test Case 4}
-
-Listing~\ref{lst:Causality Test Case 4}
-shows causality test case 4, for which the \co{r1 == r2 == 1} result
-is to be forbidden.
-The \co{herd7} tool finds this OOTA cycle, and flags it as such using \co{S8}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-4.litmus @@
-\caption{Causality Test Case 4}
-\label{lst:Causality Test Case 4}
-\end{listing}
-
-\subsubsection{Causality Test Case 5}
-\label{app:Causality Test Case 5}
-
-Listing~\ref{lst:Causality Test Case 5}
-shows causality test case 5, for which the \co{r1 == r2 == 1 && r3 == 0}
-result is to be forbidden.
-The \co{herd7} tool finds this OOTA cycle, and flags it as such using \co{S8}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-5.litmus @@
-\caption{Causality Test Case 5}
-\label{lst:Causality Test Case 5}
-\end{listing}
-
-\subsubsection{Causality Test Case 6}
-\label{app:Causality Test Case 6}
-
-Listing~\ref{lst:Causality Test Case 6}
-shows causality test case 6, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-6.litmus @@
-\caption{Causality Test Case 6}
-\label{lst:Causality Test Case 6}
-\end{listing}
-
-\subsubsection{Causality Test Case 7}
-\label{app:Causality Test Case 7}
-
-Listing~\ref{lst:Causality Test Case 7}
-shows causality test case 7, for which the \co{r1 == r2 == r3 == 1}
-result is to be allowed.
-The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-7.litmus @@
-\caption{Causality Test Case 7}
-\label{lst:Causality Test Case 7}
-\end{listing}
-
-\subsubsection{Causality Test Case 8}
-\label{app:Causality Test Case 8}
-
-Listing~\ref{lst:Causality Test Case 8}
-shows causality test case 8, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not find this cycle, but when \co{r2} is explicitly
-initialized to the constant \co{1}, it does find the cycle and refrains
-from flagging it as an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-8.litmus @@
-\caption{Causality Test Case 8}
-\label{lst:Causality Test Case 8}
-\end{listing}
-
-\subsubsection{Causality Test Case 9}
-\label{app:Causality Test Case 9}
-
-Listing~\ref{lst:Causality Test Case 9}
-shows causality test case 9, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not find this cycle, but when \co{r2} is explicitly
-initialized to the constant \co{1}, it does find the cycle and refrains
-from flagging it as an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-9.litmus @@
-\caption{Causality Test Case 9}
-\label{lst:Causality Test Case 9}
-\end{listing}
-
-\subsubsection{Causality Test Case 9a}
-\label{app:Causality Test Case 9a}
-
-Listing~\ref{lst:Causality Test Case 9a}
-shows causality test case 9a, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not find this cycle, but when \co{r2} is explicitly
-initialized to the constant \co{1}, it does find the cycle and refrains
-from flagging it as an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-9a.litmus @@
-\caption{Causality Test Case 9a}
-\label{lst:Causality Test Case 9a}
-\end{listing}
-
-\subsubsection{Causality Test Case 10}
-\label{app:Causality Test Case 10}
-
-Listing~\ref{lst:Causality Test Case 10}
-shows causality test case 10, for which the \co{r1 == r2 == 1 && r3 == 0}
-result is to be forbidden.
-The \co{herd7} tool nevertheless reports this result and does not flag it as
-an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-10.litmus @@
-\caption{Causality Test Case 10}
-\label{lst:Causality Test Case 10}
-\end{listing}
-
-\subsubsection{Causality Test Case 11}
-\label{app:Causality Test Case 11}
-
-Listing~\ref{lst:Causality Test Case 11}
-shows causality test case 11, for which the \co{r1 == r2 == r3 == r4 == 1}
-result is to be allowed.
-The \co{herd7} tool reports this result and does not flag it as an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-11.litmus @@
-\caption{Causality Test Case 11}
-\label{lst:Causality Test Case 11}
-\end{listing}
-
-\subsubsection{Causality Test Case 12}
-\label{app:Causality Test Case 12}
-
-JMM Causality Test Case 12 uses arrays, which are not yet supported
-by the \co{herd7} tool.
-
-\subsubsection{Causality Test Case 13}
-\label{app:Causality Test Case 13}
-
-Listing~\ref{lst:Causality Test Case 13}
-shows causality test case 13, for which the \co{r1 == r2 == 1}
-result is to be forbidden.
-The \co{herd7} tool nevertheless reports this result and does not flag it as
-an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-13.litmus @@
-\caption{Causality Test Case 13}
-\label{lst:Causality Test Case 13}
-\end{listing}
-
-\subsubsection{Causality Test Case 14}
-\label{app:Causality Test Case 14}
-
-Listing~\ref{lst:Causality Test Case 14}
-shows causality test case 14, for which the \co{r1 == r3 == 1 && r2 == 0}
-result is to be forbidden.
-The \co{herd7} tool does not report this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-14.litmus @@
-\caption{Causality Test Case 14}
-\label{lst:Causality Test Case 14}
-\end{listing}
-
-\subsubsection{Causality Test Case 15}
-\label{app:Causality Test Case 15}
-
-Listing~\ref{lst:Causality Test Case 15}
-shows causality test case 15, for which the \co{r0 == r1 == r3 == 1 && r2 == 0}
-result is to be forbidden.
-The \co{herd7} tool does not report this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-15.litmus @@
-\caption{Causality Test Case 15}
-\label{lst:Causality Test Case 15}
-\end{listing}
-
-\subsubsection{Causality Test Case 16}
-\label{app:Causality Test Case 16}
-
-Listing~\ref{lst:Causality Test Case 16}
-shows causality test case 16, for which the \co{r1 == 2 && r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be expected
-given the differences between Java on the one hand and C and C++
-on the other.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-16.litmus @@
-\caption{Causality Test Case 16}
-\label{lst:Causality Test Case 16}
-\end{listing}
-
-\subsubsection{Causality Test Case 17}
-\label{app:Causality Test Case 17}
-
-Listing~\ref{lst:Causality Test Case 17}
-shows causality test case 17, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be expected
-given the differences between Java on the one hand and C and C++
-on the other.
-But note that the \co{herd7} tool correctly notes the presence of an
-OOTA cycle with \co{S17}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-17.litmus @@
-\caption{Causality Test Case 17}
-\label{lst:Causality Test Case 17}
-\end{listing}
-
-\subsubsection{Causality Test Case 18}
-\label{app:Causality Test Case 18}
-
-Listing~\ref{lst:Causality Test Case 18}
-shows causality test case 18, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be expected
-given the differences between Java on the one hand and C and C++
-on the other.
-But note that the \co{herd7} tool correctly notes the presence of an
-OOTA cycle with \co{S17}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-18.litmus @@
-\caption{Causality Test Case 18}
-\label{lst:Causality Test Case 18}
-\end{listing}
-
-\subsubsection{Causality Test Case 19}
-\label{app:Causality Test Case 19}
-
-Listing~\ref{lst:Causality Test Case 19}
-shows causality test case 19, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be expected
-given the differences between Java on the one hand and C and C++
-on the other.
-But note that the \co{herd7} tool correctly notes the presence of an
-OOTA cycle with \co{S8}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-19.litmus @@
-\caption{Causality Test Case 19}
-\label{lst:Causality Test Case 19}
-\end{listing}
-
-\subsubsection{Causality Test Case 20}
-\label{app:Causality Test Case 20}
-
-Listing~\ref{lst:Causality Test Case 20}
-shows causality test case 20, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be expected
-given the differences between Java on the one hand and C and C++
-on the other.
-But note that the \co{herd7} tool correctly notes the presence of an
-OOTA cycle with \co{S8}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-20.litmus @@
-\caption{Causality Test Case 20}
-\label{lst:Causality Test Case 20}
-\end{listing}
-
-\subsection{Additional Litmus Tests}
-\label{app:Additional Litmus Tests}
-
-\subsubsection{Non-LB Litmus Tests}
-\label{app:Non-LB Litmus Tests}
-
-Listing~\ref{lst:Non-LB Litmus Test}
-shows a litmus test that can produce an OOTA cycle, but that is not
-a member of the load-buffering (LB) class of litmus tests.
-Suppose that lines~9 and~10 both somehow return the value 21.
-Then line~12 will store the value 42 to \co{x}, which will cause
-line~17 to store 21 to \co{y} and line~22 to store 21 to \co{z},
-thus justifying the values initially loaded by lines~9 and~10.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-non-lb.litmus @@
-\caption{Non-LB Litmus Test}
-\label{lst:Non-LB Litmus Test}
-\end{listing}
-
-\subsubsection{Multi-Cycle Litmus Tests}
-\label{app:Multi-Cycle Litmus Tests}
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-two-source.litmus @@
-\caption{Multi-Cycle Litmus Test}
-\label{lst:Multi-Cycle Litmus Test}
-\end{listing}
-
-\section{Aside on Undefined Behavior}
-\label{app:Aside on Undefined Behavior}
-
-The combination of OOTA cycles and UB has proven especially
-challenging~\cite{DavidGoldblatt2019NoElegantOOTAfix}.
-
-\subsection{Alignment-Based Undefined Behavior}
-\label{sec:Alignment-Based Undefined Behavior}
-
-The example shown in
-Listing~\ref{lst:Alignment-Based Undefined Behavior (P1916R0)}
-(taken from P1916R0) is especially instructive.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 std::atomic<int> y;
- 2 template <typename T>
- 3 void t1() {
- 4   long long r1 = x.load(std::memory_order_relaxed);
- 5   if (r1 % 4 == 0)
- 6     y.store(1, std::memory_order_relaxed);
- 7   *(T*)r1 = *(T*)r1;
- 8 }
-\end{verbatim}
-\caption{Alignment-Based Undefined Behavior (P1916R0)}
-\label{lst:Alignment-Based Undefined Behavior (P1916R0)}
-\end{listing}
-
-Because the compiler often assumes that UB cannot happen, it might
-look at line~7 and conclude that the value of \co{r1} necessarily
-corresponded to an address aligned per the requirements of type \co{T}.
-If that type's alignment was larger than four bytes, the compiler could
-then replace the condition in line~5 with \co{true}, which would mean
-that there was no semantic dependency from the load of \co{x} to the
-store of \co{y}.
-
-Except that the developer might have knowingly carried out a misaligned
-access, for example, to test an alignment trap handler or because the
-system on which this code was to run was known to gracefully handle
-misaligned accesses.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 std::atomic<int> y;
- 2 template <typename T>
- 3 void t1() {
- 4   long long r1 = x.load(std::memory_order_relaxed);
- 5   if (r1 % 4 == 0)
- 6     y.store(1, std::memory_order_relaxed);
- 7   __asm__ __volatile__("": : :"mightnotreturn");
- 8   *(T*)r1 = *(T*)r1;
- 9 }
-\end{verbatim}
-\caption{Alignment-Based Undefined Behavior Fix}
-\label{lst:Alignment-Based Undefined Behavior Fix}
-\end{listing}
-
-However, in that case, the developer is stepping outside the bounds
-of the standard.
-It would be good to have some way for the developer to get the job
-done without confusing the compiler, thus separating OOTA-cycle and
-UB concerns.
-One approach is to have a \co{mightnotreturn} clobber similar to the
-\co{memory} clobber for GCC asms.
-Then this example could be rewritten as shown in
-Listing~\ref{lst:Alignment-Based Undefined Behavior Fix},
-with the asm on line~7 preventing the compiler from assuming that line~8
-is guaranteed to be executed, thus preventing the compiler from
-applying any lessons that it might learn from line~8 to line~5.
-
-% Clean up <200b> in vim: ":%s/\%u200b//g"
-
-This \co{mightnotreturn} clobber is merely an example.
-There might well be better ways to prevent UB from propagating backwards.
-Other options being considered include a call to an empty function that
-is invisible to the compiler and a GCC goto asm that the compiler cannot
-prove falls through.
-
-\subsection{Unreachability-Based Undefined Behavior}
-\label{sec:Unreachability-Based Undefined Behavior}
-
-This section attempts to apply the lessons
-of Appendix~\ref{sec:Unreachability-Based Undefined Behavior} to
-Goldblatt's unreachability-UB example.  Inserting the asm at line~12
-as shown in Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
-avoids confusing the compiler into assuming that the value of \co{local}
-must necessarily be zero.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 // UB if this function is ever called.
- 2 [[noreturn]]
- 3 inline void unreachable() {
- 4   return;
- 5 }
- 6 void t1() {
- 7   r1 = x.load(std::memory_order_relaxed);
- 8   int local = (r1 & 1);
- 9   if (local == 0)
-10     y.store(1, std::memory_order_relaxed);
-11   if (local != 0) {
-12     __asm__ __volatile__("": : :"mightnotreturn");
-13     // If we get here, UB.
-14     unreachable();
-15   }
-16 }
-\end{verbatim}
-\caption{Unreachability-Based Undefined Behavior Fix}
-\label{lst:Unreachability-Based Undefined Behavior Fix}
-\end{listing}
-
-By the same reasoning, Goldblatt's \co{misbhave} example that
-also involves unreachability can also be addressed through insertion
-of the asm.
-
-% \section{History}
-% \label{sec:History}
+% @@@ TODO:
+% Create nicer sdep-evaluation example based on cbmc.
+% Add a "Tools" section to the real-world constraints section.
+%	Note the inversion from compiler, and the ability to complain
+%	when analysis cannot be carried out accurately.
+% Update wording.
+% Add more detail on blocking UB to "Aside on Undefined Behavior".
+
+% Additional snark:
+% "This work reduces the problem of OOTA analysis to that of sequential
+%	program analysis.  If you expect better verification of concurrent
+%	code than of sequential code, you are in an intellectual state
+%	of sin."
+5 "Forcing relaxed loads to be ordered before subsequent relaxed stores
+%	is an expensive no-op."
+
+% One-pager:
+% Threats to validity:
+% Speed of light might not always be finite.
+% Zero-sized atoms might be discovered.
+% It might be possible to propagate information > c.
+% It might be possible to violate causality.
+% ML hallucination might convince people that OOTA is OK.
 

### 2023-12-07 cbmc/sdep-quadratic.c
Make sdep-quadratic.c more amenable to experimentation
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -1,6 +1,67 @@
+/*
+ * Demonstrate cbmc-based computation of sdep.
+ *
+ * Usage: cbmc -DCBMC sdep-quadratic.c
+ *
+ * Additional preprocessor variables:
+ *	-DLOAD_LIMITS=N limits the arguments from -N to N, defaulting
+ *		to 0..7.
+ *	-DLOAD_CONSTANT=N forces the r2 arguments to the value N.
+ *
+ * This can also be run as a stand-alone C program:
+ *	./sdep-quadratic r1_1 r2_1 r1_2 r2_2
+ *
+ * Where the command-line arguments give the values of r1 and r2
+ * for each iteration of the loop.  The same additional preprocessor
+ * variables can be used when compiling.  You can also specify -DCBMC
+ * to exercise CBMC-specific code paths, sort of, anyway.
+ */
+
 #include <stdlib.h>
 #include <stdio.h>
 #include <assert.h>
+
+#ifdef CBMC
+int get_arg(char *arg)
+{
+	return (int)(long)arg;
+}
+#else
+int get_arg(char *arg)
+{
+	return strtol(arg, NULL, 0) & 0x7;
+}
+#endif
+
+#ifndef LOAD_LIMITS
+int limited_arg(char *arg)
+{
+	return get_arg(arg);
+}
+#else
+int limited_arg(char *arg)
+{
+	int val = get_arg(arg);
+
+	if (val > LOAD_LIMITS)
+		val = LOAD_LIMITS;
+	else if (val < -LOAD_LIMITS)
+		val = -LOAD_LIMITS;
+	return val;
+}
+#endif
+
+#ifndef LOAD_CONSTANT
+int const_arg(int arg)
+{
+	return arg;
+}
+#else
+int const_arg(int arg)
+{
+	return LOAD_CONSTANT;
+}
+#endif
 
 int main(int argc, char *argv[])
 {
@@ -13,11 +74,11 @@ int main(int argc, char *argv[])
 		int r3;
 		int *yp = &y[i];
 
-		r1 = argv[2 * i + 1][0] & 0x7;
-		// r2 = 0; // Comment out for sdep
-		r2 = argv[2 * i + 2][0] & 0x7; // Comment out for !sdep
+		r1 = limited_arg(argv[2 * i + 1]);
+		r2 = const_arg(limited_arg(argv[2 * i + 2]));
[…]

### 2023-12-08 ootavacuous.tex
Add appendix on use of cbmc to compute sdep
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Fortunately, there are now tools that can help.
-For example, the SAT-solver-based \co{cbmc} tool, which is readily
-available,\footnote{
-	On Debian-based distributions, via \co{sudo apt install cbmc}.}
+For example, the SAT-solver-based \co{cbmc} tool
 has been used to mechanically verify signicant portions of Linux-kernel
@@
 Both tools can easily check whether or not a given dependency is semantic.
-% @@@ More on cbmc in an appendix.
+An example use of \co{cbmc} is illustrated in
+Appendix~\ref{sec:Evaluating sdep Using cbmc}.
 
@@
 prove falls through.
+@@@ show details.
+
+A \co{std::observable()} function has been proposed to the C++ standards
+committee~\cite{DavisHerring2021P1494R2}.
 
@@
 of the asm.
+
+\section{Evaluating sdep Using \co{cbmc}}
+\label{sec:Evaluating sdep Using cbmc}
+
+This section demonstrates the evaluation of sdep using
+the C Bounded Model Checker
+(\co{cbmc})~\cite{EdmundClarke2004CBMC}.
+This tool takes a smallish C program as input and constructs a logic
+expression representing that program.
+The logic expression's variables represent bits in the program's inputs
+and state, and the result of that logic expression is \co{true} if that
+program has a bug, for example, if it is possible for that C program to
+trigger an assertion or execute an out-of-bounds array reference.
+The tool then hands this expression to a SAT solver, and if this solver
+finds a satisfying set of values, the tool uses that set to provide
+the execution path to the bug.
+
+\begin{listing}[tbp]
+\scriptsize
+\begin{verbatim}
+ 1 P0(atomic_int *x, atomic_int *y, atomic_int *limit)
+ 2 {
+ 3   r1 = atomic_load_explicit(x, memory_order_relaxed);
+ 4   r2 = atomic_load_explicit(limit, memory_order_relaxed);
+ 5   r3 = r1 * r1 + 2 * r1 + 2;
+ 6   r4 = r2;
+ 7   if (r2 >= r3)
+ 8     r4 = r3;
+ 9   atomic_store_explicit(y, r4, memory_order_relaxed);
+10 }
+\end{verbatim}
+\caption{Example Code for Evaluation of sdep}
+\label{lst:Example Code for Evaluation of sdep}
+\end{listing}
+
+Listing~\ref{lst:Example Code for Evaluation of sdep}
+shows a \co{herd7} process \co{P0()}.
+Is there a semantic dependency between the load from \co{x} on
+line~3 to the store to \co{y} on line~9?
+
+\begin{listing}[tbp]
+\scriptsize
+\begin{verbatim}
+ 1 int main(int argc, char *argv[])
+ 2 {
+ 3   int i;
+ 4   int y[2] = { 0, 0 };
+ 5
+ 6   for (i = 0; i < 2; i++) {
+ 7     int r1;
+ 8     int r2;
+ 9     int r3;
+10     int *yp = &y[i];
+11
+12     r1 = limited_arg(argv[2 * i + 1]);
+13     r2 = const_arg(limited_arg(argv[2 * i + 2]));
+14     r3 = r1 * r1 + 2 * r1 + 2;
+15     *yp = r2 <= r3 ? r2 : r3;
+16     printf("i = %d, r1 = %d, r2 = %d, r3 = %d\n", i, r1, r2, r3);
+17   }
+18   assert(y[0] == y[1]);
+19 }
+\end{verbatim}
+\caption{Code for cbmc-Based Evaluation of sdep}
+\label{lst:Code for cbmc-Based Evaluation of sdep}
+\end{listing}
+
+One way to answer this question is to create a C program
+(see Listing~\ref{lst:Code for cbmc-Based Evaluation of sdep})
+that runs the \co{P0()} process's code twice with arbitrary inputs
+(the loop spanning lines~6-17) and asserts that the results of
+each iteration be equal (line~18).
+
+Running this with:
+
+\begin{quote}
+	\scriptsize
+	\co{cbmc -DCBMC cbmc/sdep-quadratic.c}
+\end{quote}
+
+gives the result \co{VERIFICATION FAILED}.
+This indicates that there are different values of \co{x} and \co{limit}
+that yield different values of \co{y}, that is, that there can be
+a semantic dependency.
+
+However, $x^2 + 2x + 2$ has a minimum value of 1 at $x=-1$, so
+forcing \co{limit} to (say) zero should eliminate this static
+dependency:
+
+\begin{quote}
+	\scriptsize
+	\co{cbmc -DCBMC -DLOAD_CONSTANT=0 cbmc/sdep-quadratic.c}
+\end{quote}
+
+But this also gives the result \co{VERIFICATION FAILED}.
+One reason for this is that this program does 32-bit computer arithmetic,
+and is thus subject to wraparound.
+Limiting the value of \co{x} to 46,339 prevents wraparound:
+
+\begin{quote}
+	\scriptsize
+	\co{cbmc -DCBMC -DLOAD_CONSTANT=0 -DLOAD_LIMITS=46339 cbmc/sdep-quadratic.c}
+\end{quote}
+
+And this yields \co{VERIFICATION SUCCESSFUL}, indicating that there
+is no semantic dependency in this case.
+This is to be expected because the specified value of zero for \co{limit}
+is smaller than the minimum of the quadratic expression, so that the
+value stored to \co{y} is always zero.
+Specifying \co{-DLOAD_CONSTANT=1} also yields \co{VERIFICATION SUCCESSFUL}
+because this value is equal to that minimum, so that the value stored to
+\co{y} is always 1.
+However, specifying any larger value (for example, \co{-DLOAD_CONSTANT=2})
+yields \co{VERIFICATION FAILED} because the value assigned to \co{y}
+might be either 1 or 2, indicating that a semantic dependency exists.
+
+This situation underscores the point that the existance of a semantic
+dependency can be a property of a particular execution rather than a
+universal property of the code itself.
+
+There are numerous other tools that can be used, up to and including
+manual inspection, but \co{cbmc} is readily available and easy to
+use.\footnote{
+	On Debian-based distributions, via \co{sudo apt install cbmc}.}
 
@@
 % Zero-sized atoms might be discovered.
-% It might be possible to propagate information > c.
+% It might be possible to propagate information at infinite velocity
+%	despite the finite speed of light.
 % It might be possible to violate causality.

### 2023-12-08 ootavacuous.tex
Update abstract
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	long-standing laws of physics and constraints on hardware
-	architecture and design, thereby providing an informal
-	proof that OOTA cannot occur in correctly constructed C++
-	programs running on real-world systems.
-
-	The result is that a correct per-execution-path C++ implementation
-	necessarily preserves the semantic dependencies that prevent
-	OOTA cycles from forming.
-	Implementations that undertake complex cross-thread analysis
-	still need to be careful.
+	architecture and design to demonstrate that OOTA cannot occur in
+	correctly constructed C++ programs and correctly constructed C++
+	implementations running on on real-world systems.
+	This paper also looks at some tooling implicatoins of OOTA.
 \end{abstract}
@@
 % @@@ TODO:
-% Create nicer sdep-evaluation example based on cbmc.
 % Add a "Tools" section to the real-world constraints section.

### 2023-12-08 ootavacuous.tex
Rename Appendix A and forward-reference it
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Add section on tooling
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+Other OOTA-related litmus tests may be found in
+Appendix~\ref{app:OOTA-Related Litmus Tests}.
+
 Note that C++ implementations are permitted to evaluate to a more
@@
 However, as we will see, C++ implementations need not use these tools.
-Their current analysis is guanteed to be sufficient.
+Their current analysis is guaranteed to be sufficient, courtesy of the
+real-world constraints discussed in the next section.
 Furthermore, as we will also see, in some special but commonly occurring
 circumstances, much simpler methods exist that incur negligible overhead.
-
-% @@@ We could follow this long-established tradition of proposing
-% solutions to the OOTA problem without a definition of this problem.
-% However, we (perhaps foolishly) felt the need to provide a clear
-% definition,
-
-% We therefore define an OOTA cycle (also known as a causal cycle) as a
-% cycle in the memory-model sense involving only reads-from links, that
-% is, the value stored in on thread being read by the next thread.
-% Within each thread in the cycle, the value stored depends on
-% the value loaded.
-% This dependency might be a control dependency
-% (see Listing~\ref{lst:OOTA Cycle})
-% or it might be a data dependency
-% (see Listing~\ref{lst:Causality Test Case 4} on
-% page~\pageref{lst:Causality Test Case 4}).\footnote{
-%	Sufficiently motivated readers should have no problem constructing
-%	examples involving address dependencies.}
-% @@@ There must be at least two distinct values such that if a given load
-% returns a given value, that value forces the cycle to execute so as to
-% cause the corresponding store's value to be equal to that given value, and
-% for that store to occur early enough for the load to return that value.
 
@@
 with constraints imposed by the C++ implementations themselves.
+An additional section discusses the OOTA-cycle implications for
+any tooling that exists separately for these C++ implementations.
 
@@
 correctly identifies a semantic dependency, and thus exhibits temporal
-behavior for the corresponding code.
+behavior for the execution at hand of the corresponding code.
 The second row depicts the case where the implementation misclassifies
@@
 
-Therefore, in the limiting case, analyses of OOTA cycles that focus on a
-given implementation can simply use that implementation's classification
-of dependencies as semantic on the one hand or non-semantic on the other.
-While taking this approach sacrifices significant generality, it has the
-benefit of greatly reduces the cost of dependency analysis, potentially
-to zero.
-
 \subsubsection{C++ Implementations: Consequences}
@@
 the store.
+
+\subsection{Semantic Dependencies and Tooling}
+\label{sec:Semantic Dependencies and Tooling}
+
+Where a C++ implementation's primary function is to correctly execute
+a C++ program, the function of tooling is often to evaluate properties
+of possible executions of that same program.
+This difference in function results in different sensitivities to
+sdep misclassifications, as shown in
+Table~\ref{tab:Semantic-Dependency Classification For Tools}.
+
+\begin{table}
+\centering
+\begin{tabular}{c|c|l}
+Actual Code	& Tooling		& Result \\
+\hline
+sdep		& sdep			& Temporal \\
+\cline{2-3}
+		& $\neg$sdep		& False Positive (Warning?) \\
+\hline
+$\neg$sdep	& sdep			& False Negative (Warning?) \\
+\cline{2-3}
+		& $\neg$sdep		& Atemporal \\
+\end{tabular}
+\caption{Semantic-Dependency Classification For Tools}
+\label{tab:Semantic-Dependency Classification For Tools}
+\end{table}
+
+Any misclassification will result in either a false positive or
+a false negative, except that many tools give some indication of a
+misclassification, for example, printing a warning message indicating
+that the code is too complex for it to analyze.
+Some projects would choose to restructure the code in such cases on the
+grounds that code that is difficult for an automated tool to understand
+is likely also being misunderstood by its all-too-human developers.
+
+Some other tools respond to complexity by consuming all available CPU
+and memory, which can also be interpreted as a good and sufficient
+warning message.
+
+However, there is an important special case for tools that are closely
+associated with a C++ implementation.
+Such tools can simply use that implementation's classification of
+dependencies as semantic on the one hand or non-semantic on the other.
+While taking this approach sacrifices significant generality, it has
+the not insignificant benefit of greatly reducing the cost of dependency
+analysis, potentially all the way to zero.
 
@@
 handles the memory model, observable behavior, and the as-if rule
-cannot form OOTA cycles.
+cannot form OOTA cycles when running on real systems in this
+universe.
 
@@
 
-\section{Demonstration of OOTA-Cycle Definition}
-\label{app:Demonstration of OOTA-Cycle Definition}
+\section{OOTA-Related Litmus Tests}
+\label{app:OOTA-Related Litmus Tests}
 
@@
 % @@@ TODO:
-% Add a "Tools" section to the real-world constraints section.
-%	Note the inversion from compiler, and the ability to complain
-%	when analysis cannot be carried out accurately.
 % Update wording.

### 2023-12-08 ootavacuous.tex
Remove prototype wording section
Perhaps revert this commit at a later time?

Reported-by: Maged Michael <maged.michael@gmail.com>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 universe.
-
-\section{Suggested Changes to Wording}
-\label{sec:Suggested Changes to Wording}
-
-Add the following note in \co{[atomics.order]} after 33.5.4p9:
-
-\begin{quote}
-	[ \emph{Note:} Implementations that respect the following
-	constraints are incapable of computing out-of-thin-air value
-	that depend circularly on their own computation:
-
-	\begin{itemize}
-	\item	Correctly implement the C++ abstract machine, and
-	\item	Run on correctly implemented hardware
-		(specifically, hardware that forbids non-speculative
-		loads from returning values from speculative stores
-		and that forbids exposing values from speculative
-		stores to other threads), and
-	\item	Reside in a universe having a finite speed of light
-		and atomic matter.
-	\end{itemize}
-
-	All three constraints are subject to the as-if rule.
-	For example, those creating an alternate universe in which
-	C++ implementations are to run correctly should ensure
-	that their universe behaves as if the speed of light was
-	finite and the nature of matter was atomic.
-	--- \emph{end note} ]
-\end{quote}
 
@@
 % @@@ TODO:
-% Update wording.
 % Add more detail on blocking UB to "Aside on Undefined Behavior".

### 2023-12-08 ootavacuous.tex
Wordsmith the tooling section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Some other tools respond to complexity by consuming all available CPU
-and memory, which can also be interpreted as a good and sufficient
-warning message.
+One advantage that tools often have over C++ implementations is the
+ability to devote much more computational power to the problem at hand.
+In fact, some tools respond to excessive complexity by consuming all
+available CPU and memory, which can also be interpreted as a good and
+sufficient warning message.
 
@@
 dependencies as semantic on the one hand or non-semantic on the other.
-While taking this approach sacrifices significant generality, it has
-the not insignificant benefit of greatly reducing the cost of dependency
-analysis, potentially all the way to zero.
+While taking this approach sacrifices significant generality, it has the
+significant benefit of greatly reducing the cost of dependency analysis,
+potentially all the way down to zero.
 

### 2023-12-09 ootavacuous.tex
Add subsections in OOTA Definition
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+\subsubsection{Semantic Dependencies Are Per Execution}
+\label{sec:Semantic Dependencies Are Per Execution}
+
 One aspect of this complexity is the fact that semantic dependencies
@@
 
+\subsubsection{Semantic Dependencies Can Be a Choice}
+\label{sec:Semantic Dependencies Can Be a Choice}
+
 But suppose that the implementation can prove that at any time that
@@
 from \co{z} to \co{x}, or both \co{y} and \co{z} to \co{x}?
-Or is this a free choice on the part of the implementation?
-
-And perhaps worse:
+The answer is that this is a free choice on the part of the
+implementation.
+
+\subsubsection{Semantic Dependencies Can Be Complex}
+\label{sec:Semantic Dependencies Can Be Complex}
+
+Consider this example:
 
@@
 Here, the presence or absence of a semantic dependency from \co{y} to
-\co{x} depends not only on the value of \co{z}, but also the code in
-a horribly complex function.
+\co{x} depends not only on the value of \co{z}, but also the code in a
+function that, although single threaded, is horribly complex.
 
@@
 
+\subsubsection{Complex Semantic Dependencies Are a Choice}
+\label{sec:Complex Semantic Dependencies Are a Choice}
+
 All of these tools can incur high overhead, especially manual code
 inspection.
-However, as we will see, C++ implementations need not use these tools.
+However, as we will see, the actual C++ implementations themselves need
+not use these tools.
 Their current analysis is guaranteed to be sufficient, courtesy of the
@@
 Furthermore, as we will also see, in some special but commonly occurring
-circumstances, much simpler methods exist that incur negligible overhead.
+circumstances, there are much simpler methods that incur negligible overhead.
 
@@
 %	of sin."
-5 "Forcing relaxed loads to be ordered before subsequent relaxed stores
+% "Forcing relaxed loads to be ordered before subsequent relaxed stores
 %	is an expensive no-op."
+% "So you don't want to learn concurrency?  Then I can only suggest
+%	that you either move further up the stack or take early
+%	retirement."
 

### 2023-12-09 ootavacuous.tex
Clean up UB-blocking discussion
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Peter Sewell feedback, initial installment
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-	This paper goes further, adding temporal constraints from
-	long-standing laws of physics and constraints on hardware
-	architecture and design to demonstrate that OOTA cannot occur in
-	correctly constructed C++ programs and correctly constructed C++
-	implementations running on on real-world systems.
-	This paper also looks at some tooling implicatoins of OOTA.
+	This paper adds constraints from long-standing laws of physics
+	and on hardware systems to demonstrate that OOTA cannot occur
+	in correctly constructed C++ programs running on correctly
+	constructed C++ implementations.
+
+	In this, OOTA behaviors are similar to the issues surrounding
+	undecidability and its real-world counterpart, high complexity.
+	So similar, in fact, that not only the same strategies apply
+	to OOTA, but also the same \emph{code} applies, both for C++
+	implementations and for associated tooling.
 \end{abstract}
@@
 
-This section provides a brief overview of the OOTA problem followed
-by an equally brief summary of prior work in this area.
+This section provides a brief overview of the OOTA problem, followed
+by an equally brief summary of prior work in this area, and ending
+with a quick review of a few of the challenges for general definitions of
+``OOTA cycle''.
 
@@
 prove falls through.
-@@@ show details.
-
-A \co{std::observable()} function has been proposed to the C++ standards
+
+A \co{std::observable()} function that would block backwards-propagating
+UB has been proposed to the C++ standards
 committee~\cite{DavisHerring2021P1494R2}.
@@
 % Luc Maranget: Memory models.
-% Richard Gristenthwaite: ARM hardware.
 % Mark Rutland: ARM memory ordering.
@@
 % Alan Stern: Memory models, mathematical logic, and variety of hardware.
+% Martin Uecker (offered)
+% Peter Sewell
+% Mark Batty
+% Richard Gristenthwaite: ARM hardware.  Posed HW speculation question.
 
 % @@@ TODO:
-% Add more detail on blocking UB to "Aside on Undefined Behavior".
+% Measure store-buffer delays on an x86 system.
 

### 2023-12-09 ootavacuous.tex
Peter Sewell December 9 feedback, second installment
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\subsection{OOTA Definition}
-\label{sec:OOTA Definition}
+\subsection{On OOTA Definitions}
+\label{sec:On OOTA Definitions}
 
@@
 authors, but has left others unsatisfied.
+Still others argue (perhaps correctly) that it is impossible to provide
+a general definition of ``OOTA cycle'', but even accepting this raises
+the question of whether a \emph{useful} definition can be formulated.
 
@@
 There are also many other tools, but failing that, it is always possible
-to fall back to the oldest software-verification tool, namely manual
+to fall back to older software-verification tools such as manual
 code inspection.

### 2023-12-11 ootavacuous.tex
Peter Sewell December 10 feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+\subsubsection{Semantic Dependencies and I/O}
+\label{sec:Semantic Dependencies and I/O}
+
+A computation might include I/O:
+
+\begin{quote}
+	\co{x = y * input_int();}
+\end{quote}
+
+Then there is a semantic dependency from \co{y} to \co{x} only if
+the number input is non-zero.
+Similar issues arise when reading timestamps.
+
+In addition, I/O operations are observed behavior, which greatly restricts
+optimizations involving them.
+
 \subsubsection{Semantic Dependencies Can Be Complex}
@@
 function that, although single threaded, is horribly complex.
+
+A special case is as follows:
+
+\begin{quote}
+	\co{x = y & 0x1 ? do_odd(y) : do_even(y);}
+\end{quote}
+
+The key point here is that diferent values might cause almost completely
+disjoint code to be executed.
 
@@
 \cline{2-3}
-		& $\neg$sdep		& BUG!!! \\
+		& $\neg$sdep		& Fix C++ bug! \\
 \hline
@@
 
+% Objections:
+% "Atomic operations are not observable behavior."
+%	They are for volatile atomic objects.  Plus an implementation
+%	whose analysis is confined to a thread would need to assume
+%	that any atomic store might lead to observable behavior in some
+%	other thread.  An implementation with a global view must preserve
+%	observable behavior regardless of OOTA cycles.	Plus if an OOTA
+%	cycle does not affect observable behavior, who cares?
+% "The C++ abstract machine is independent of the laws of physics."
+%	That might be, but any real implementation of that abstract
+%	machine will be governed by those laws.
+% "Real hardware can exhibit OOTA, as exemplified by audio feedback."
+%	Such hardware does not run C++.  Also, audio feedback is more
+%	closely approximated by a loop calculating sound intensity than
+%	by OOTA, especially given that you must include the audio
+%	properties of the enclosing space in analyzing the feedback loop.
+
 % Additional snark:

### 2023-12-11 ootavacuous.tex
December 10 self-review
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Because there is a cycle, on a real system, at least one of the links
-must be counter-temporal, that is, either a load returns a value before
+must be atemporal, that is, either a load returns a value before
 that value is stored, or a store's address or value is computed before
@@
 
-Line~4 of the corresponding output of the \co{herd7} tool shows the
+Listing~\ref{lst:OOTA Cycle, herd7 Output} shows the corresponding output
+of the \co{herd7} tool.
+Lines~3 and~4 show the possible states, with line~4 showing the
 counterintuitive outcome where both processes load the value 42.
@@
 due to physical constraints of which \co{herd7} is (by design) unaware.
+Line~9 contains \co{Sometimes} (as opposed to \co{Never} or \co{Always}),
+indicating that some executions satisfy the \co{exists} clause and
+others do not.
 
@@
 However, because these models are atemporal, they cannot reject
-OOTA executions other than by flagging the OOTA value as arbitrary
-(which some in fact do in at least some cases).
+OOTA executions other than by flagging the OOTA value as arbitrary,
+which some in fact do in at least some cases.
 
@@
 Appendix~\ref{app:Aside on Undefined Behavior}
-analyzes the examples and proposes a way to separate UB and OOTA-cycle
+analyzes the examples and notes ways to separate UB and OOTA-cycle
 concerns.
@@
 
-Others, current authors included, have define an OOTA cycle as a
+Others, current authors included, have defined an OOTA cycle as a
 cycle in sdep $\cup$ rf, where sdep is the set of semantic
@@
 This is a fine definition, and is consistent with the words in the C++
-standard, except that it simply shunts all the complexity into the
+standard, except that it simply shunts all the complexity onto the
 term ``semantic dependency''.
+
+The following sections discuss properties of and constraints on
+useful definitions of ``semantic dependency''.
 
@@
 
-Similarly:
+Consider also:
 
@@
 be a strict function of the source code, \emph{in general, sdep must be
-defined on a per-execution basis}.
+defined on a per-execution basis}.\footnote{
+	Thanks to Alan Stern for the examples in this section and
+	in the preceding one.}
+
+\subsubsection{Semantic Dependencies Can Be Nondeterministic}
+\label{sec:Semantic Dependencies Can Be Nondeterministic}
+
+Consider this example, with \co{i} initially zero:
+
+\begin{quote}
+	\co{x = ++i / ++i;}
+\end{quote}
+
+The standard does not specify the order of evaluation, so the resulting
+value might be zero ($\frac{1}{2}$ truncated) or two ($\frac{2}{1}$).
+This indeterminacy is why many coding guidelines prohibit side effects
+in expressions, but the above code nevertheless conforms to the standard.
 
@@
 verification~\cite{MichalisKokologiannakis2017NidhuggRCU,SMC-TreeRCU,MichalisKokologiannakis2019RCUstatelessModelCheck}.
-Both tools can easily check whether or not a given dependency is semantic.
+Both tools can easily check whether or not a given dependency is semantic
+for at least some approximate definitions of ``semantic dependency''.
 An example use of \co{cbmc} is illustrated in
@@
 Their current analysis is guaranteed to be sufficient, courtesy of the
-real-world constraints discussed in the next section.
+real-world constraints discussed in the next section, at least assuming
+that their analysis is in fact correct.
 Furthermore, as we will also see, in some special but commonly occurring
@@
 This paper will therefore assume that the speed of light is finite, and
-thus the maximum speed at which information can propagate is also finite.
+that the maximum speed at which information can propagate is also finite.
 
@@
 If atoms were of zero size, a zero-sized computer might be constructed.
-In such a computer, even give finite speed of light, information could
+In such a computer, even given finite speed of light, information could
 propagate among computing elements in zero time.
@@
 These three laws of physics mean that if one thread loads the value
-stored by some other thread, that load must have occurred later
+stored by some other thread, that load must have occurred earlier
 in global time than did the store.
@@
 	and the squashing of failed speculation might span multiple
-	threads.}
+	threads.
+	This might make sense for multiple hardware threads sharing
+	a core.}
 Similarly, lines~18 and~20 cannot be committed until after lines~16
@@
 In this case, even omniscient oracular implementations are required
-to confine their analyses within a single thread.
+to confine their analyses to within a single thread.
 
@@
 \begin{tabular}{c|c|l}
-Actual Code	& Implementation	& Result \\
+Actual Execution	& Implementation	& Result \\
 \hline
@@
 \hline
-$\neg$sdep	& sdep			& Temporal \\
+$\neg$sde		& sdep			& Temporal \\
 \cline{2-3}
@@
 \begin{tabular}{c|c|l}
-Actual Code	& Tooling		& Result \\
+Actual Execution	& Tooling		& Result \\
 \hline
@@
 
-We are grateful to David Goldblatt, Jade Alglave, and Alan Stern for their
-careful review of an early draft of this paper and to John Wickerson for
-asking Paul for a rant and taking the proffered rant seriously.
+We are grateful to David Goldblatt, Jade Alglave, Alan Stern, and Peter
+Sewell for their careful review of an early draft of this paper and
+to John Wickerson for asking Paul for a rant and taking the proffered
+rant seriously.
 We also owe David Goldblatt a debt of gratitude for his having asked an
@@
 Martin Uecker contributed valuable insights on backwards-propagating UB.
+
 Nonetheless, all errors and omissions in this paper are the sole property
-of the authors.
+of the authors, and the appearance of a name in this paragraph does
+not in any way constitute agreement with anything in this paper.
 
@@
 % Measure store-buffer delays on an x86 system.
+% Real-world compilers freeze nondeterministic choices.
 

### 2023-12-12 ootavacuous.tex
Add Memarian citations on C standard
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Because volatile atomics constitute observable
+The exact definition of a computer language is subject to some debate,
+with standards, implementations, and users all having
+influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.
+Nevertheless, because volatile atomics constitute observable
 behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2023N4950},
@@
 Martin Uecker contributed valuable insights on backwards-propagating UB.
+Richard Grisenthwaite patiently explained the architectural constraints
+that prevent hardware OOTA.
 
@@
 % It might be possible to violate causality.
+% Hardware and compiler bugs might result in OOTA.
 % ML hallucination might convince people that OOTA is OK.

### 2023-12-13 ootavacuous.tex
December 13 Michael Wong review
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 The exact definition of a computer language is subject to some debate,
-with standards, implementations, and users all having
-influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.
+with standards, implementations, and users all having some degree of
+influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
+and each of which is subject to change over time.
 Nevertheless, because volatile atomics constitute observable
@@
 With that line, the additional value~42 can also be stored to \co{r1}.
-The implementation is therefore forbidded from inventing that store of 42
+The implementation is therefore forbidden from inventing that store of 42
 unless it can prove that doing so does not negatively affect the
@@
 
-Duplicating atomic stores can also introduce new and undesireable
+Duplicating atomic stores can also introduce new and undesirable
 behaviors.

### 2023-12-15 ootavacuous.tex
Add appendix on coe, fre, and rfe IPC modes
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{document}
-\title{D3064R0: OOTA Execution is Provably Vacuous}
+\title{D3064R0: OOTA Execution on Real Implementations is Provably Vacuous}
 
@@
 by other threads (``coe'')), and loads from objects for which the value
-returned by a given load was overwritten by some other thread (``fre'').
+returned by a given load was overwritten by some other thread
+(``fre'').\footnote{
+	See Appendix~\ref{app:Inter-Process Communications} for more
+	information on these memory-reference interprocess communications
+	modes.}
 
@@
 Carrying out this type of optimization requires that the implementation
-classify dependencies as semantic or not.
+classify dependencies as maybe semantic or definitely not.
 Given the limited resources available, almost all implementations will
@@
 
+\section{Inter-Process Communications}
+\label{app:Inter-Process Communications}
+
+This section more clearly defines the ``coe'', ``fre'', and ``rfe''
+types of interprocess communication (also collectively called ``links'')
+and gives evidence on the temporal qualities of these different types.
+
+First, context.
+The final ``e'' in all three acronyms stands for ``external'' as opposed
+to within-thread communication (``coi'', ``fri'', and ``rfi'') and as
+also opposed to all communication, whether inter-thread or intra-thread
+(``co'', ``fr'', and ``rf'').
+
+Next, definitions:
+
+\begin{description}
+\item[coe]
+	Coherency (external), which connects a store to any other store
+	to that same object that overwrites either the first store's
+	value or that overwrites some later value.
+\item[fre]
+	From-read (external), which connects a load to any other store
+	object that overwrites either the value that the load returned
+	or that overwrites some later value.
+\item[rfe]
+	Reads-from (external), which connects a store to any load from
+	that same object that returns the value stored.
+\end{description}
+
+\begin{figure}[tb]
+\begin{center}
+\resizebox{4in}{!}{\includegraphics{ipc}}
+\caption{IPC Diagram for coe, fre, and rfe}
+\label{fig:IPC Diagram for coe, fre, and rfe}
+\end{center}
+\end{figure}
+
+Thirdly, Figure~\ref{fig:IPC Diagram for coe, fre, and rfe} illustrates
+the coe, fre, fri, and rfe links for four threads with time advancing
+from the top to the bottom of the figure.  Starting with coe, even though
+Thread~1's store was later in global time than that of Thread~0, that of
+Thread~1 came first in coe order, followed by those of Threads~2 and ~3.
+There is thus a coe link from Thread~1's store to every other store,
+from Thread~0's store to those of Threads~2 and~3, and from Thread~2's
+store to that of Thread~3.
+
+Thread~3's load reads the value stored by Thread~0's store, so there is
+an rfe link from that store to that load.  There is also an fre link from
+that load to Thread~2's store and an fri link to Thread~3's later store.
+Note that this fre link goes backwards in time.
+
+Finally, there might be a desire for hard evidence that coe and fre
+links really can go backwards in time.
+We provide this evidence on x86 to demonstrate that these effects
+are not confined to weakly ordered systems.
+This system is a laptop with Intel(R) Core(TM) i9-10885H CPU @ 2.40GHz,
+having a single socket with 8~cores, each having a pair of hardware
+threads.
+The code generating this data may be found in the \path{CodeSamples/cpu}
+directory of the
+perfbook~\cite{McKenney2018ParallelProgramming-2018-12-08a}.\footnote{
+	\co{git clone git://git.kernel.org/pub/scm/linux/kernel/git/paulmck/perfbook.git}}
+
+\begin{figure}[tb]
+\begin{center}
+\resizebox{3in}{!}{\includegraphics{coe-sh-out}}
+\caption{On x86, coe Links Are Atemporal}
+\label{fig:On x86, coe Links Are Atemporal}
+\end{center}
+\end{figure}
+
+Figure~\ref{fig:On x86, coe Links Are Atemporal}
+shows the results of 15 of the laptop's hardware threads simultaneously
+storing their identifying integer (ranging from 0 to 14) to a shared
+variable, then repeatedly doing timestamped reads from that variable.
+The x-axis displays time in TSC periods, each of which is about 0.417~ns
+in duration.
+The y-axis shows the number of distinct opinions that the hardware
+threads have as a function of time, a number that is frequently rather
+larger than one.
+The backwards-in-time coe link from Thread~1 to Thread~0 in
+Figure~\ref{fig:IPC Diagram for coe, fre, and rfe}
+is therefore entirely plausible.
+
+\begin{figure}[tb]
+\begin{center}
+\resizebox{5in}{!}{\includegraphics{coe-out-4}}
+\caption{On x86, coe Links Are Partially Ordered}
+\label{fig:On x86, coe Links Are Partially Ordered}
+\end{center}
+\end{figure}
+
+\begin{figure}[tb]
+\begin{center}
+\resizebox{3in}{!}{\rotatebox{90}{\includegraphics{store15tred}}}
+\caption{On Power5, coe Links Are Also Partially Ordered}
+\label{fig:On Power5, coe Links Are Also Partially Ordered}
+\end{center}
+\end{figure}
+
+However, Figure~\ref{fig:On x86, coe Links Are Partially Ordered}
+shows that the order of reads of each value at each thread are
+consistent with a number of global orders, as required.
+Figure~\ref{fig:On Power5, coe Links Are Also Partially Ordered}
+shows a rather more elaborate partial order from a Power5
+system~\cite{McKenney2018ParallelProgramming-2018-12-08a}.
+
+\begin{figure}[tb]
+\begin{center}
+\resizebox{3in}{!}{\includegraphics{fre-sh-out}}
+\caption{On x86, fre Links Are Atemporal}
+\label{fig:On x86, fre Links Are Atemporal}
+\end{center}
+\end{figure}
+
+Figure~\ref{fig:On x86, fre Links Are Atemporal}
+plots a histogram of the elapsed time from the last load to return the
+old value to the store that provided the new value.
+Note well that most of the data falls into negative time, that is to say,
+the last load of the old value executes about 50 timestamp periods (about
+20~nanoseconds) \emph{after} the store that overwrote that same old value.
+That is, fre links can and do go backwards in time.
+This corresponds to the fre link between Thread~3 an Thread~2 depicted in
+Figure~\ref{fig:IPC Diagram for coe, fre, and rfe}.
+
+\begin{figure}[tb]
+\begin{center}
+\resizebox{3in}{!}{\includegraphics{rfe-sh-out}}
+\caption{On x86, rfe Links Are Temporal}
+\label{fig:On x86, rfe Links Are Temporal}
+\end{center}
+\end{figure}
+
+Figure~\ref{fig:On x86, rfe Links Are Temporal}
+plots a histogram of the elapsed time from a store of a new value
+to the first load of that new value.
+Note well that all of the data falls into positive time, indicating
+that rfe links always go forward in time, as those familiar with
+computer hardware and the laws of physics would expect.
+
+And this is exactly why the C++ memory model guarantees ordering from
+rfe links but not the coe and fre links for relaxed accesses in the
+absence of other ordering from stronger atomic memory accesses or
+\co{atomic_thread_fence()}.
+
 % \section{History}
@@
 Nonetheless, all errors and omissions in this paper are the sole property
-of the authors, and the appearance of a name in this paragraph does
+of the authors, and the appearance of a name in this appendix does
 not in any way constitute agreement with anything in this paper.

### 2023-12-15 data/plots.sh

@@ -1,0 +1,52 @@
+#!/bin/sh
+
+gnuplot << ---EOF---
+set term png
+set output "../coe-sh-out.png"
+set xlabel "Time (Timestamp Periods)"
+set ylabel "Number of Simultaneous Values"
+#set logscale y
+set xrange [:3500]
+#set yrange [100:10000]
+set nokey
+# set label 1 "RCU" at 0.1,10 left
+# set label 2 "spinlock" at 0.5,3.0 left
+# set label 3 "brlock" at 0.4,0.6 left
+# set label 4 "rwlock" at 0.3,1.6 left
+# set label 5 "refcnt" at 0.15,2.8 left
+plot "coe.out-4.dat" w l
+---EOF---
+
+gnuplot << ---EOF---
+set term png
+set output "../fre-sh-out.png"
+set xlabel "Store-to-Load Latency (Timestamp Periods)"
+set ylabel "Number of Samples"
+#set logscale y
+set xrange [-200:100]
+#set yrange [100:10000]
+set nokey
+# set label 1 "RCU" at 0.1,10 left
+# set label 2 "spinlock" at 0.5,3.0 left
+# set label 3 "brlock" at 0.4,0.6 left
+# set label 4 "rwlock" at 0.3,1.6 left
+# set label 5 "refcnt" at 0.15,2.8 left
+plot "fre.sh.out.dat"
+---EOF---
+
+gnuplot << ---EOF---
+set term png
+set output "../rfe-sh-out.png"
+set xlabel "Store-to-Load Latency (Timestamp Periods)"
+set ylabel "Number of Samples"
+#set logscale y
+set xrange [0:200]
+#set yrange [100:10000]
+set nokey
+# set label 1 "RCU" at 0.1,10 left
+# set label 2 "spinlock" at 0.5,3.0 left
+# set label 3 "brlock" at 0.4,0.6 left
+# set label 4 "rwlock" at 0.3,1.6 left
+# set label 5 "refcnt" at 0.15,2.8 left
+plot "rfe.sh.out.dat"
+---EOF---

### 2023-12-15 ootavacuous.sh

@@ -78,6 +78,17 @@ END {
 	print "___EOF___";
 }' | sh > ootavacuous-flat.tex
 
+svgfiles=`find . -name '*.svg' -print`
+for i in $svgfiles
+do
+	basename="${i%.svg}"
+	if test ! -f "$basename.pdf" -o "$basename.svg" -nt "$basename.pdf"
+	then
+		echo "$basename.svg -> $basename.pdf"
+		inkscape --export-pdf=$basename.pdf $basename.svg
+	fi
+done
+
 pdflatex ootavacuous-flat.tex
 bibtex ootavacuous-flat.aux
 pdflatex ootavacuous-flat.tex

### 2023-12-15 ootavacuous.tex
December 14 self-review
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{quote}
-	\co{x = ++i / ++i;}
+	\co{y = ++i / ++i;} \\
+	\co{x = y * z;}
 \end{quote}
@@
 The standard does not specify the order of evaluation, so the resulting
-value might be zero ($\frac{1}{2}$ truncated) or two ($\frac{2}{1}$).
-This indeterminacy is why many coding guidelines prohibit side effects
-in expressions, but the above code nevertheless conforms to the standard.
-
-\subsubsection{Semantic Dependencies Can Be a Choice}
-\label{sec:Semantic Dependencies Can Be a Choice}
+value of \co{y} might be zero ($\frac{1}{2}$ truncated) or two
+($\frac{2}{1}$).
+In the former case, there is no semantic dependency from \co{z} to
+\co{x}, but in the latter case there is.
+Thus, semantic dependencies are not just a function of a particular
+execution through the source code, but also of arbitrary choices made
+by the C++ implementation.\footnote{
+	Thanks to Peter Sewell for pointing out this possibility.}
+
+Note well that C++ implementations can and do act as if portions of
+the source code was duplicated in order to apply specialization
+optimizations, in which case any nondeterminism might also be
+duplicated.
+
+Many projects' coding guidelines prohibit side effects in expressions,
+their goal being to obtain portable code whose behavior does not depend
+on arbitrary choices on the part of the implementation.
+However, the above code nevertheless conforms to the standard.
+
+\subsubsection{Semantic Dependencies And Optimization}
+\label{sec:Semantic Dependencies And Optimization}
 
@@
 the above statement might execute, \co{y} is equal to \co{z}?
-That would allow the implementation to act as if the source code
-was instead as follows:
+That would allow the implementation to optimize, that is, to act as if
+the source code was instead as follows:
 
@@
 
-\subsubsection{Complex Semantic Dependencies Are a Choice}
-\label{sec:Complex Semantic Dependencies Are a Choice}
+\subsubsection{Complex Semantic Dependencies Analysis is a Choice}
+\label{sec:Complex Semantic Dependencies Analysis is a Choice}
 
@@
 speculative stores.
+And yes, this does mean that hardware can produce OOTA cycles during
+speculative execution, but the hardware is required to prevent all such
+speculative exection from ever committing.
 
@@
 activities, analysis of the code and classification of dependencies.
+Two sections on these topics are followed by sections on optimizations
+and on consequences.
 
@@
 is temporal.
+
+\subsubsection{C++ Implementations: Optimization}
+\label{sec:C++ Implementations: Optimization}
+
+C++ implementations undertake a wide variety of optimizations, in
+some cases taking advantage of nondeterminism and in others applying
+the results of code analysis to produce faster code.
+The changes can be profound, for example, with large computations being
+replaced by simple constants.
+Either way, the implementation is behaving as if the original program
+had been written more optimally.
+
+In such cases, one option is to carry out analysis of semantic dependencies
+on the optimized code.
+The seL4 project took this approach to the extreme of analyzing
+the binary output~\cite{ThomasSewell2013L4binaryVerification}.
+
+In C++ implementations using compilers, non-determinism of a given
+(post-optimization) code path is ``frozen'' into the resulting binary,
+which greatly simplifies analysis of semantic dependencies.
+Other c++ implementations (for example, those using interpreters or JITs)
+might need to directly model any nondeterminism.
+This latter situation might motivate additional C++ projects to ban
+nondeterminism in favor of portable code.
 

### 2023-12-16 ootavacuous.tex
December 15 self-review plus narrate JVM litmus tests
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 }
-\date{November 27, 2023 (Pre-Tokyo)}
+\date{December 15, 2023 (Pre-Tokyo)}
 \maketitle{}
@@
 constraints may be applied.
-For example, if an implementation determines that the values of a pair
-of objects are always equal at any time that a given thread might access
-them, one could imagine a sufficiently brave implementation substituting
-a load from one for a load from the other.
-
-However, if all of the inter-thread relaxed accesses are volatile,
+For example, if an implementation proves that the values of a pair of
+objects are always equal at any time that a given thread might access
+them, that implementation could behave as if the that thread's code had
+substituted a load from one for a load from the other.
+
+Note that if all of the inter-thread relaxed accesses are volatile,
 as they are in the C language, then all of the inter-thread traffic is
@@
 
+Another question is how to prove that a given optimization is valid.
+This has been a hard question in the past, and is likely to remain so
+both for sequential and concurrent programs.
+The consequences of using an invalid optimization can be quite severe,
+and the evidence thus far indicates that OOTA cycles are not a bit issue.
+Nevertheless, a proof that the same value is always stored implies a
+proof that there is no semantic dependency to that relaxed atomic store,
+and thus no chance of that store participating in an OOTA cycle.
+
 \subsubsection{C++ Implementations: Consequences}
@@
 Such tools can simply use that implementation's classification of
-dependencies as semantic on the one hand or non-semantic on the other.
+dependencies as semantic on the one hand or non-semantic on the other,
+whether by working with that implementation's intermediate representations,
+binaries produced by that implementation, or by tracing the program's
+execution.
 While taking this approach sacrifices significant generality, it has the
@@
 an OOTA cycle, this section instead looks at how to form one.
-This will lead to the conclusion that a correct C++ implementation
-cannot form an OOTA cycle.
+This will lead to the conclusion that a correct real-world C++
+implementation cannot form an OOTA cycle.
 
@@
 that goes backwards in time.
-Given that the inter-thread links are temporal, the only possible
-candidate atemporal links so are those confined to a given thread.
+Given that the inter-thread links are temporal with a give store preceding
+any load returning the value stored, the only possible candidate atemporal
+links are those confined to a given thread.
 This in turn requires that a store that depends on an earlier load
@@
 \begin{enumerate}
-\item	Is able to determine (prove) the location and value of the store.
+\item	Is able to prove the location and value of the store.
 \item	Guesses the location and value of the store and has some means
@@
 is known to be zero, so that the store can proceed prior to the load.
-But in this case, there is no perturbation of the value returned from
+But in this case, there is no way that the value returned from
 the load that can affect the value stored.
-Therefore, by P0442R0, that store cannot possibly take part in an
-OOTA cycle.
+Therefore, that store cannot possibly take part in an OOTA cycle.
 
@@
 is visible to any other thread.
-Therefore, hardware speculation cannot result in OOTA cycles, and
-by design, even on weakly ordered
+Therefore, any hardware speculation that results in OOTA cycles is doomed
+to be squashed an discarded, and by design, even on weakly ordered
 systems~\cite{ARMv7A:2010,ARMv8A:2017,PowerISA2.07-2013}.
@@
 memory model.
-This work shows that correct C++ implementations are inherently incapable
-of producing OOTA cycles.
-
-Therefore, if your C++ implementation produces OOTA cycles, you have
-bugs that would be considered bugs even if OOTA cycles were allowed.
+This work shows that correct real-world C++ implementations are inherently
+incapable of producing OOTA cycles.
+
+Therefore, if your C++ implementation produces OOTA cycles, you have bugs.
 And you need to fix those bugs, regardless of whether they are in your
@@
 is to be allowed.
-However, the \co{herd7} tool does not find this litmus test's cycle, presumably
-because this tool does not do the deep inter-thread program analysis
-required to find it.
+The \co{herd7} tool finds this litmus test's cycle, even without the
+inter-thread analysis, because C++ does not respect control dependencies
+to relaxed atomic stores.
 
@@
 \end{listing}
+
+Implementations that can prove that the value of \co{x} is always
+non-negative can also prove that there is no semantic dependency between
+\co{P0()}'s load from \co{x} on line~8 to its store to \co{y} on line~10,
+that is, no matter which of the possible values is loaded from \co{x},
+the value 1 will always be stored to \co{y}.
+
+Less-omniscient implementations, including those whose code analysis
+is restricted to a thread, would need to act as if there was in fact a
+semantic dependency.
+
+Even omniscient implementations will observe a semantic dependency from
+\co{P1()}'s load from \co{y} on line~14 to its store to \co{x} on line~15.
+
+Either way, there is no OOTA cycle, with omniscient implementations
+having no cycle at all, and other implementations retaining the
+non-semantic dependency.
 
@@
 
+Because \co{P0()}'s loads from \co{x} on lines~8 and~9 are unordered,
+an implementation can act as if the source code loaded from \co{x} only
+once and assigned the returned value to both \co{r1} and \co{r2}.\footnote{
+	But not in C, where relaxed atomic operations are volatile,
+	and thus observable behavior.}
+The implementation could then prove that the condition on line~10
+would always evaluate to \co{true}, and act as if the source code had
+unconditionally executed the relaxed store to \co{y} on line~11.
+The implementation could then act as if the resulting relaxed accesses
+to \co{x} and \co{y} appeared in reverse order, at which point
+\co{P0()}'s store to \co{y} could be read by \co{P1()} load so that
+\co{P1()} stored 1 to \co{x}, thus resulting in \co{P1()} loading
+that 1 from \co{x}.
+
+Note that \co{P1()} has a trivial semantic dependency from its load
+from \co{y} on line~15 to its store to \co{x} on line~16.
+
+If \co{P0()}'s loads from \co{x} are fused, there is no cycle,
+otherwise the preserved non-semantic dependency prevents an
+OOTA cycle from forming on correct real-world implementations.
+
 \subsubsection{Causality Test Case 3}
@@
 
+The analysis from Section~\ref{app:Causality Test Case 2}
+applies here, including the difference in volatility between C and C++
+atomics.
+Although the addition of \co{P2} allows sequentially consistent executions
+where \co{P0()}'s two loads from \co{x} might return different values,
+it does not change the fact that those two loads can be merged.
+
 \begin{listing}[tbp]
@@
 is to be forbidden.
-The \co{herd7} tool finds this OOTA cycle, and flags it as such using \co{S8}.
+The \co{herd7} tool finds this OOTA cycle because C++ does not respect
+data dependencies, and flags it as such using \co{S8}.
 
@@
 
+Both \co{P0()} and \co{P1()} have straightforward semantic dependencies
+from their relaxed atomic loads to their relaxed atomic stores, which
+prevents this OOTA cycle on real-world C++ implementations.
+
 \subsubsection{Causality Test Case 5}
@@
 result is to be forbidden.
-The \co{herd7} tool finds this OOTA cycle, and flags it as such using \co{S8}.
+The \co{herd7} tool finds this OOTA cycle because C++ does not respect
+data dependencies, and flags it as such using \co{S8}.
 
@@
 
+Each of \co{P0()}, \co{P1()}, and \co{P3()} have trivial semantic
+dependencies from their relaxed atomic loads to their relaxed
+atomic stores, which prevents any OOTA cycle from forming in correct
+real-world C++ implementations.
+\co{P2()} has no loads, so cannot participate in an OOTA cycle in
+any case.
+
 \subsubsection{Causality Test Case 6}
@@
 result is to be allowed.
-The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
+The \co{herd7} tool finds this cycle, and does not flag it as an OOTA
+cycle because C++ does not respect control dependencies to relaxed
+atomic stores.
 
@@
 \end{listing}
+
+A C++ implementation that can prove that \co{y} is always either
+zero or one can also prove that there is no semantic dependency
+from \co{P1()}'s load on line~14 to its stores on lines~16 and~29.
+Such an implementation could act as if lines~15-20 had been replaced
+by a single store to \co{x}, allowing that store to execute first,
+followed by lines~8, 10, and~14, in that order, satisfying the
+\co{exists} clause on line~23.
+
+Less-omniscient implementations will preserve the non-semantic
+dependency, preventing any OOTA loop from forming in a correct real-world C++
+implementation.
 
@@
 \end{listing}
+
+Because all accesses are relaxed atomics, this test case has everything
+to do with reordering and nothing to do with OOTA cycles.
+The outcome happens when lines~17, 10, 11, 15, 16 and~9 execute in that
+order.
+Note that this ordering preserves the semantic dependencies from
+lines~10-11 and~15-16, and can thus happen on correct real-world
+C++ implementations.
 
@@
 
+Omniscient C++ implementations might note that the computation on line~9
+can only have the value one when the load from \co{x} returns either
+zero or one, and both \co{x} and \co{y} are initialized to zero.
+Such an implementation act as if line~10's \co{r2} had instead been
+the value~1,
+Then there would be no semantic dependency between \co{P0()}'s load
+and store, and thus no OOTA cycle in a correct real-world C++
+implementation.
+
+A less-capable implementation would leave the non-semantic dependency
+in place, also preventing any OOTA cycle in a correct real-world C++
+implementation.
+
 \subsubsection{Causality Test Case 9}
@@
 result is to be allowed.
-The \co{herd7} tool does not find this cycle, but when \co{r2} is explicitly
-initialized to the constant \co{1}, it does find the cycle and refrains
-from flagging it as an OOTA cycle.
+The \co{herd7} tool does not find this cycle because it depends on
+scheduling \co{P2()} after the other two processes, and C++ is not
+constrained to do this.
 
@@
 
+Both processes have straightforward semantic dependencies between their
+respective relaxed atomic loads and relaxed atomic stores, so an OOTA
+cycle cannot form in a correct real-world C++ implementation.
+
 \subsubsection{Causality Test Case 9a}
@@
 result is to be allowed.
-The \co{herd7} tool does not find this cycle, but when \co{r2} is explicitly
-initialized to the constant \co{1}, it does find the cycle and refrains
-from flagging it as an OOTA cycle.
+The \co{herd7} tool does not find this cycle because it depends on
+scheduling \co{P2()} before the other two processes, and C++ is not
+constrained to do this.
 
@@
 
+Both processes have straightforward semantic dependencies between their
+respective relaxed atomic loads and relaxed atomic stores, so an OOTA
+cycle cannot form in a correct real-world C++ implementation.
+
 \subsubsection{Causality Test Case 10}
@@
 result is to be forbidden.
-The \co{herd7} tool nevertheless reports this result and does not flag it as
-an OOTA cycle.
+Because C++ does not respect control dependencies, the \co{herd7} tool
+nevertheless reports this result and does not flag it as an OOTA cycle.
 
@@
 \end{listing}
+
+However, each of \co{P0()}, \co{P1()}, and \co{P3()} have straightforward
+semantic dependencies between their respective loads and stores, and
+thus cannot form an OOTA cycle when running on a correct real-world C++
+implementation.
+For its part, \co{P2()} has no semantic dependency, so it cannot be part
+of an OOTA cycle when running on a correct real-world C++ implementation.
 
@@
 
+This test case has nothing to do with OOTA cycles.
+The result may be obtained by executing the statements in the following
+order: line~20, line~12, line~13, line~18, line~19, line~10, line~11,
+and finally line~17.
+
 \subsubsection{Causality Test Case 12}
@@
 
-JMM Causality Test Case 12 uses arrays, which are not yet supported
-by the \co{herd7} tool.
+JMM Causality Test Case 12 in
+Figure~\ref{lst:Causality Test Case 12}
+uses arrays, which are not yet supported by the \co{herd7} tool.
+
+\begin{listing}[tbp]
+@@ DisplayLitmus litmus/oota-causality-12.litmus @@
+\caption{Causality Test Case 12}
+\label{lst:Causality Test Case 12}
+\end{listing}
+
+An omniscient C++ implementation could note that only \co{P0()} accesses
+array \co{a[]} and act as if the code was simpler, but there would
+still be semantic dependencies between \co{P0()}'s and \co{P1()}s
+relaxed atomic loads and stores involving \co{x} and \co{y}.
+Those dependencies prevent an OOTA cycle from forming on a correct
+real-world C++ implementation, omniscient or not.
 
@@
 result is to be forbidden.
-The \co{herd7} tool nevertheless reports this result and does not flag it as
-an OOTA cycle.
+But because C++ does not respect control dependencies, \co{herd7} tool
+reports this result and does not flag it as an OOTA cycle.
 
@@
 
+There are straightforward semantic dependencies between \co{P0()}'s and
+\co{P1()}s relaxed atomic loads and stores involving \co{x} and \co{y}.
+Those dependencies prevent an OOTA cycle from forming on a correct
+real-world C++ implementation.
+
 \subsubsection{Causality Test Case 14}
@@
 result is to be forbidden.
-The \co{herd7} tool does not report this result.
+The \co{herd7} tool nevertheless reports this result because C++ does not
+respect control dependencies for atomic stores.
 
@@
 
+In C++, this is not a matter of OOTA but rather reordering.
+Dependencies notwithstanding, according to the C++ memory model,
+the statements could execute in the following order:
+Line~17 (so \co{r2 == 0}), line~21, line~9 (so \co{r1 == 1}),
+line~13, line~18 (so \co{r3 == 1)}.
+The sum of \co{r2} and \co{r3} is non-zero, justifying line~21's
+store to \co{x}.
+
+However, correct real-world C++ implementations would have semantic
+dependencies between \co{P0()}'s line~9 and lines~11 and~13, and also
+between \co{P1()}'s lines~17 and~18 and line~21, preventing any
+OOTA cycle from forming.
+
 \subsubsection{Causality Test Case 15}
@@
 \end{listing}
+
+In C++, this is not a matter of OOTA but rather reordering.
+Dependencies notwithstanding, according to the C++ memory model,
+the statements could execute in the following order:
+Line~31, line~27, line~10 (so \co{r0 == 1}), line~13 (so \co{r1 == 1}),
+line~19 (because \co{r1 == 1}), line~23 (so \co{r2 == 0}),
+and finally line~24 (so \co{r3 == 1}).
+The sum of \co{r2} and \co{r3} is non-zero, justifying line~27's
+store to \co{u}.
+
+However, correct real-world C++ implementations would have semantic
+dependencies between \co{P0()}'s lines~10 and~13 and lines~17 and~19, and also
+between \co{P1()}'s lines~23 and~24 and line~27, preventing any
+OOTA cycle from forming.
+Having no load, \co{P2()} cannot participate in an OOTA cycle.
 
@@
 The \co{herd7} tool does not report this result, which is to be expected
-given the differences between Java on the one hand and C and C++
-on the other.
+given that C++ does not permit a thread to reorder accesses to a single
+atomic variable, even when those accesses are relaxed.
 
@@
 The \co{herd7} tool does not report this result, which is to be expected
-given the differences between Java on the one hand and C and C++
-on the other.
+given that C++ does not permit a thread to reorder accesses to a single
+atomic variable, even when those accesses are relaxed.
 But note that the \co{herd7} tool correctly notes the presence of an
@@
 \end{listing}
+
+Both processes have straightforward semantic dependencies between their
+respective relaxed atomic loads and relaxed atomic stores, so an OOTA
+cycle cannot form in a correct real-world C++ implementation.
 
@@
 The \co{herd7} tool does not report this result, which is to be expected
-given the differences between Java on the one hand and C and C++
-on the other.
+given that C++ does not permit a thread to reorder accesses to a single
+atomic variable, even when those accesses are relaxed.
 But note that the \co{herd7} tool correctly notes the presence of an
 OOTA cycle with \co{S8}.
+
+It is interesting that this \co{S8} value does not propagate to
+\co{r3}.
 
@@
 \end{listing}
+
+All processes have straightforward semantic dependencies between their
+respective relaxed atomic loads and relaxed atomic stores, so an OOTA
+cycle cannot form in a correct real-world C++ implementation.
 
@@
  6     y.store(1, std::memory_order_relaxed);
- 7   __asm__ __volatile__("": : :"mightnotreturn");
+ 7   std::observable();
  8   *(T*)r1 = *(T*)r1;
@@
 UB concerns.
-One approach is to have a \co{mightnotreturn} clobber similar to the
-\co{memory} clobber for GCC asms.
+One approach is to use the \co{std::observable()} function
+that has been proposed to the C++ standards
+committee~\cite{DavisHerring2021P1494R2}.
+This function would blocks backwards-propagating UB.
 Then this example could be rewritten as shown in
@@
 
-This \co{mightnotreturn} clobber is merely an example.
-There might well be better ways to prevent UB from propagating backwards.
-Other options being considered include a call to an empty function that
-is invisible to the compiler and a GCC goto asm that the compiler cannot
-prove falls through.
-
-A \co{std::observable()} function that would block backwards-propagating
-UB has been proposed to the C++ standards
-committee~\cite{DavisHerring2021P1494R2}.
+There are a number of ways to implement \co{std::observable()},
+for example, a new \co{"mightnotreturn"} GCC asm clobber, a call to an empty
+function that is invisible to the compiler, and a GCC goto asm that the
+compiler cannot prove falls through.
 
@@
 of Appendix~\ref{sec:Unreachability-Based Undefined Behavior} to
-Goldblatt's unreachability-UB example.  Inserting the asm at line~12
+Goldblatt's unreachability-UB example.
+
+Inserting \co{std::observable()} at line~12
 as shown in Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
@@
 11   if (local != 0) {
-12     __asm__ __volatile__("": : :"mightnotreturn");
+12     std::observable();
 13     // If we get here, UB.
@@
 also involves unreachability can also be addressed through insertion
-of the asm.
+of the proposed \co{std::observable()}.
 
@@
 %	that you either move further up the stack or take early
-%	retirement."
+%	retirement."  Note that that C++ committee's concurrency
+%	expertise has grown beyond recognition over the past 15 years.
 

### 2023-12-16 ootavacuous.tex
Document user influence over semantics
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+\subsubsection{Implementers and Users Influence ``Semantic Dependencies'' Definition}
+\label{sec:Implementers and Users Influence ``Semantic Dependencies'' Definition}
+
+The exact definition of a computer language is subject to some debate,
+with standards, implementations, and users all having some degree of
+influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
+and each of which is subject to change over time.
+Users  and implementers are of course wise to assume that their opinions
+might be overridden by those of the standard and, for the users,
+especially the implementation.
+However, consider the following code:
+
+\begin{quote}
+	\co{void foo(char c} \\
+	\co{\{} \\
+	\co{\ \ \ \ return c + 128;} \\
+	\co{\}}
+\end{quote}
+
+The standard does not specify whether or not type \co{char} is signed,
+leaving that choice to the implementations, which follow the lead
+of the architects of a given CPU family.
+Therefore, the code above is undefined behavior on systems choosing
+signed \co{char} and having 8-bit bytes (so that there is no semantic
+dependency from the \co{foo()} function's argument to its return value),
+but perfectly well-defined on systems having unsigned \co{char}
+(so that there is a semantic dependency).
+So score one for the implementers, albeit by conscious choice on the
+part of the standards committee.
+
+Except that GCC provides the \co{-funsigned-char} command-line
+argument that causes this implementation to treat variables of
+type \co{char} as unsigned, regardless of the architects' wishes.
+Given this command-line argument, there is always a semantic dependency
+from the \co{foo()} function's argument to its return value.
+So score one for at least some implementations' users.
+
+Please see Appendix~\ref{sec:User Influence Over Language Semantics}
+for more examples of user control over semantics, including semantic
+dependencies.
+
 \subsubsection{Complex Semantic Dependencies Analysis is a Choice}
@@
 
-The exact definition of a computer language is subject to some debate,
-with standards, implementations, and users all having some degree of
-influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
-and each of which is subject to change over time.
-Nevertheless, because volatile atomics constitute observable
+Because volatile atomics constitute observable
 behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2023N4950},
@@
 Although C++ non-volatile accesses to atomic objects are not observable
-behavior, they are still subject to the memory model.
+behavior, @@@
+
+In addition,such accesses are still subject to the memory model.
 In particular, implementations must not behave as if the source code has
@@
 \end{listing}
+
+\clearpage
 
@@
 \co{atomic_thread_fence()}.
+
+\clearpage
+
+\section{User Influence Over Language Semantics}
+\label{sec:User Influence Over Language Semantics}
+
+As noted earlier, the exact definition of a computer language is subject
+to some debate, with standards, implementations, and users all having
+some degree of
+influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
+and each of which is subject to change over time.
+It is natural to dismiss user influence when compared to the text of
+standards or the code in implementations, but both are subject to
+change and do change over time.
+The implementation is especially subject to change, for example,
+consider the following command line used to build the Linux kernel's
+\co{kernel/rcu/tree.c} C-language source file:
+
+~\\
+{
+	\scriptsize
+	\texttt{
+	gcc -Wp,-MMD,kernel/rcu/.tree.o.d -nostdinc
+	-I./arch/x86/include -I./arch/x86/include/generated  -I./include
+	-I./arch/x86/include/uapi -I./arch/x86/include/generated/uapi
+	-I./include/uapi -I./include/generated/uapi
+	-include ./include/linux/compiler-version.h
+	-include ./include/linux/kconfig.h -include
+	./include/linux/compiler\_types.h -D\_\_KERNEL\_\_
+	-fmacro-prefix-map=./= -Werror -std=gnu11 -fshort-wchar
+	-funsigned-char -fno-common -fno-PIE -fno-strict-aliasing -mno-sse
+	-mno-mmx -mno-sse2 -mno-3dnow -mno-avx -fcf-protection=branch
+	-fno-jump-tables -m64 -falign-jumps=1 -falign-loops=1
+	-mno-80387 -mno-fp-ret-in-387 -mpreferred-stack-boundary=3
+	-mskip-rax-setup -mtune=generic -mno-red-zone -mcmodel=kernel
+	-Wno-sign-compare -fno-asynchronous-unwind-tables
+	-mindirect-branch=thunk-extern -mindirect-branch-register
+	-mindirect-branch-cs-prefix -mfunction-return=thunk-extern
+	-fno-jump-tables -fpatchable-function-entry=16,16
+	-fno-delete-null-pointer-checks -O2 -fno-allow-store-data-races
+	-fstack-protector-strong -fomit-frame-pointer
+	-fno-stack-clash-protection -falign-functions=16
+	-fno-strict-overflow -fno-stack-check -fconserve-stack
+	-Wall -Wundef -Werror=implicit-function-declaration
+	-Werror=implicit-int -Werror=return-type -Werror=strict-prototypes
+	-Wno-format-security -Wno-trigraphs -Wno-frame-address
+	-Wno-address-of-packed-member -Wframe-larger-than=2048 -Wno-main
+	-Wno-unused-but-set-variable -Wno-unused-const-variable -Wvla
+	-Wno-pointer-sign -Wcast-function-type -Wno-array-bounds
+	-Wno-alloc-size-larger-than -Wimplicit-fallthrough=5
+	-Werror=date-time -Werror=incompatible-pointer-types
+	-Werror=designated-init -Wenum-conversion
+	-Wno-unused-but-set-variable -Wno-unused-const-variable
+	-Wno-restrict -Wno-packed-not-aligned -Wno-format-overflow
+	-Wno-format-truncation -Wno-stringop-overflow
+	-Wno-stringop-truncation -Wno-missing-field-initializers
+	-Wno-type-limits -Wno-shift-negative-value
+	-Wno-maybe-uninitialized -Wno-sign-compare
+	-DKBUILD\_MODFILE='"kernel/rcu/tree"' -DKBUILD\_BASENAME='"tree"'
+	-DKBUILD\_MODNAME='"tree"' -D\_\_KBUILD\_MODNAME=kmod\_tree -c
+	-o kernel/rcu/tree.o kernel/rcu/tree.c
+	}
+}
+
+We do not propose to explain all of these, and sufficiently motivated
+readers can avail themselves of the GCC documentation.
+We instead look at representative members of several categories.
+
+The \co{-funsigned-char} causes type \co{char} to be unsigned, which
+overrides per-architecture defaults, some of which treat \co{char}
+as signed and others as \co{unsigned}.
+This choice prevents a class of bugs, and also allows the kernel to make
+reliable use of the uppermost bit of variables of type \co{char}.
+It also affects the definition of ``semantic dependency'' by changing
+the arithmetic proporties of this type.
+In theory, the standard could specify the signedness of \co{char},
+but presumably the variety of existing practice in the 1980s prevented
+this from being done.
+
+The \co{-mno-sse} prevents GCC from making use of the SSE hardware.
+This is done for performance reasons, as it avoids the overhead of
+saving and restoring the state of this hardware when switching between
+user and kernel contexts.
+Similarly, the \co{-mcmodel=kernel} causes the kernel binary to be
+placed in the uppermost 2GB of the address space, again reducing
+the overhead of switching between user and kernel contexts.
+These are cases where the standard does not specify anything, nor
+should it.
+
+The \co{-fpatchable-function-entry=16,16} causes GCC to emit 16
+\co{nop} instructions at the beginning of each function, with
+the function's entry point being just after this string of \co{nop}
+instructions.
+The resulting buffers are used by the Linux-kernel tracing infrastructure
+which is in turn used for debugging, performance measurement, and
+monitoring.
+This is again clearly outside of the scope of the standard.
+
+The \co{-fstack-protector-strong} causes GCC to emit code that provides
+some protection against some clases of attacks based on buffer overflows.
+One could rightly argue code should simply avoid ever overflowing buffers,
+but things like memory allocators and userspace memory accesses must
+use code that can be difficult to distinguish from buffer overflows.
+It is not clear that the ever-increasing variety of attacks should
+affect the standard.
+
+The \co{-fno-strict-overflow} causes GCC to act as if signed integer
+overflow is defined behavior.
+Note well that this command-line option also affects the definition of
+``semantic dependency''.
+This might be a controversial choice, and another option would be to add
+a new set of signed integer types to the standard for which overflow is
+defined as wrapping, similar to the situation with unsigned integers.
+
+The \co{-Werror=strict-prototypes} causes GCC to warn if old-style
+non-ANSI function prototypes are used.
+This helps avoid certain classes of bugs.
+Warnings are by design outside of the scope of the standard.
+
+In other words, user preference can exert a non-trivial influence over
+language semantics, and in particular can affect some aspects of the
+definition of ``semantic dependency''.
 

### 2023-12-16 ootavacuous.tex
Add if-then-else example to per-execution sdep section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
From sdep U rf to sdep U rfe plus minor edits
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Others, current authors included, have defined an OOTA cycle as a
-cycle in sdep $\cup$ rf, where sdep is the set of semantic
-dependencies within each thread and rf is the set of store-to-load
-links, whether intra-thread or between threads.
+Others have defined an OOTA cycle as a cycle in sdep $\cup$ rf, where
+sdep is the set of semantic dependencies within each thread and rf is
+the set of store-to-load links, whether intra-thread (rfi) or between
+threads (rfe).\footnote{
+	See Appendix~\ref{app:Inter-Process Communications} for definitions
+	and properies of rf, rfe, and rfi.}
 This is a fine definition, and is consistent with the words in the C++
@@
 term ``semantic dependency''.
+
+The current authors deviate slightly from this tradition by instead
+defining an OOTA cycle as a cycle in sdep $\cup$ rfe, where rfe is the
+set of store-to-load links, but excluding intra-thread rfi links, which
+are instead considered to be part of sdep.
+Although this does shunt even more complexity onto the term
+``semantic dependency'', it also enables us to bring to bear
+laws of physics and properties of hardware onto correct real-world
+C++ implementations.\footnote{
+	Thanks to Alan Stern for providing a litmus test that demonstrated
+	that we had unwittingly (but productively!) shifted our definition
+	to sdep $\cup$ rfe.}
 
@@
 instead the constant zero.
+
+Finally, consider this:
+
+\begin{quote}
+	\co{r1 = x;} \\
+	\co{if (r1 > 0)} \\
+	\co{\ \ \ \ y = r1;} \\
+	\co{else} \\
+	\co{\ \ \ \ z = r1;}
+\end{quote}
+
+Here there is a semantic dependency from \co{x}, but in some executions
+it extends to \co{y} and in others to \co{z}.
 
@@
 
-\subsubsection{Implementers and Users Influence ``Semantic Dependencies'' Definition}
-\label{sec:Implementers and Users Influence ``Semantic Dependencies'' Definition}
+\subsubsection{Implementers and Users Influence Semantic-Dependencies Definition}
+\label{sec:Implementers and Users Influence Semantic-Dependencies Definition}
 
@@
 dependencies.
+
+\emph{A definition of sdep drawn strictly from the standard will be
+at best an approximation to a definition that is useful in practice.}
 
@@
 	information on these memory-reference interprocess communications
-	modes.}
+	links.}
 

### 2023-12-16 ootavacuous.sh
Improve ootavacuous.sh tempdir cleanup
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -15,7 +15,7 @@
 #	Why not both?  ;-)
 
 T="`mktemp -d ${TMPDIR-/tmp}/ootavacuous.sh.XXXXXX`"
-trap 'rm -rf $T' 0
+trap 'rm -rf $T' 0 2
 
 awk -v sq="'" -v dq='"' -v bs='\\' < ootavacuous.tex '
 BEGIN {

### 2023-12-16 ootavacuous.tex
Outline constraints from non-volatile relaxed atomic stores
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Although C++ non-volatile accesses to atomic objects are not observable
-behavior, @@@
-
-In addition,such accesses are still subject to the memory model.
+behavior, any implementation that restricts its code analysis to a
+single thread must assume (unless it can prove otherwise) that a given
+relaxed atomic store might affect observable behavior following any
+load that returns the value stored.
+This assumption does not constrain the implementation to anywhere near
+the extent that a volatile relaxed atomic store would, but it does add
+significant constrants over those related to non-volatile non-atomic
+stores.
+
+For example, non-volatile relaxed atomic accesses are subject to
+the memory model.
 In particular, implementations must not behave as if the source code has

### 2023-12-17 ootavacuous.tex
Alan Stern December 16 feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Maged Michael\\\email{maged.michael@gmail.com} \and
+Alan Stern\\\email{stern@rowland.harvard.edu} \and
 The Indefatigible TBD
@@
 
-	This paper adds constraints from long-standing laws of physics
-	and on hardware systems to demonstrate that OOTA cannot occur
-	in correctly constructed C++ programs running on correctly
-	constructed C++ implementations.
+	This paper refactors the commonly used condition for detecting
+	OOTA behavior, which permits bringing to bear constraints from
+	long-standing laws of physics and on hardware systems.
+	These constraints prevent OOTA from occurring in correctly
+	constructed C++ programs running on correctly constructed C++
+	implementations in our current universe.
 
@@
 The current authors deviate slightly from this tradition by instead
-defining an OOTA cycle as a cycle in sdep $\cup$ rfe, where rfe is the
+defining an OOTA cycle as a cycle in sdep $\cup$ rfe (or
+\co{acyclic(sdep ; rfe)} in \co{herd7} notation), where rfe is the
 set of store-to-load links, but excluding intra-thread rfi links, which
@@
 laws of physics and properties of hardware onto correct real-world
-C++ implementations.\footnote{
-	Thanks to Alan Stern for providing a litmus test that demonstrated
-	that we had unwittingly (but productively!) shifted our definition
-	to sdep $\cup$ rfe.}
+C++ implementations.% \footnote{
+%	Thanks to Alan Stern for providing a litmus test that demonstrated
+%	that we had unwittingly (but productively!) shifted our definition
+%	to sdep $\cup$ rfe.}
 
@@
 be a strict function of the source code, \emph{in general, sdep must be
-defined on a per-execution basis}.\footnote{
-	Thanks to Alan Stern for the examples in this section and
-	in the preceding one.}
+defined on a per-execution basis}.% \footnote{
+% 	Thanks to Alan Stern for the examples in this section and
+% 	in the preceding one.}
 
@@
 
-For example, non-volatile relaxed atomic accesses are subject to
-the memory model.
+For example, non-volatile relaxed atomic accesses are subject to the
+memory model.
 In particular, implementations must not behave as if the source code has
@@
 data dependencies, and flags it as such using \co{S8}.
+Interestingly enough, there is no state with \co{r3} also shown as
+\co{S8}, perhaps indicating that \co{herd7} does not propagate OOTA
+values.
 
@@
 Such an implementation act as if line~10's \co{r2} had instead been
-the value~1,
+the value~1, so that the value returned by \co{P0()}'s load could not
+affect the value stored.
 Then there would be no semantic dependency between \co{P0()}'s load
@@
 
-We are grateful to David Goldblatt, Jade Alglave, Alan Stern, and Peter
+We are grateful to David Goldblatt, Jade Alglave, and Peter
 Sewell for their careful review of an early draft of this paper and
@@
 OOTA and UB, and for his ``Deathstation 9000'' demonic CPU.
-Alan Stern located many unclear statements and gaps in reasoning.
+% Alan Stern located many unclear statements and gaps in reasoning.
 Martin Uecker contributed valuable insights on backwards-propagating UB.
@@
 % Luc Maranget: Memory models.
-% Mark Rutland: ARM memory ordering.
-% Will Deacon: ARM hardware, C++, and OOTA.
 % Greg Kroah-Hartman and Linus Torvalds: FYI, trouble being caused.
@@
 % Tony Tye: P0422R0.
-% Other memory-model maintainers, including Peter Zijlstra.
+% Other memory-model maintainers.
 % Derek Williams: PowerPC hardware.
@@
 % Brian Demsky: Prior OOTA.
-
-% @@@ Inter-CPU speculation if inter-CPU squashing of speculation.
 
@@
 % Richard Gristenthwaite: ARM hardware.  Posed HW speculation question.
+% Peter Zijlstra.
+% Mark Rutland: ARM memory ordering.
+% Will Deacon: ARM hardware, C++, and OOTA.
 

### 2023-12-17 ootavacuous.tex
Fill out text for "Additional Litmus Tests"
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+This appendix evaluates a number of OOTA-related litmus tests.
+The evaluations are carried out using the \co{herd7} tool and also
+using manual analysis.
+The manual analysis uses the informal rule that a semantic dependency
+exists between a given relaxed atomic load and store on any execution
+where a different value loaded would change either the address or the
+value of that store.
+Producing a precise formal statement of this rule is quite challenging.
+
+Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}
+evaluates the obligatory litmus tests from the JMM Causality Test
+Cases, and
+Appendix~\ref{app:Additional Litmus Tests}
+evaluates some additional litmus tests that affected our thinking or
+otherwise gave us pause.
+
 \subsection{Litmus Tests from “Causality Test Cases"}
 \label{app:Litmus Tests from “Causality Test Cases"}
+
+These tests might be two decades old, but they are still quite
+relevant.
 
@@
 
+These litmus tests helped illuminate important aspects of the problem
+of defining sdep and identifying OOTA cycles.
+
 \subsubsection{Non-LB Litmus Tests}
@@
 
-All processes have straightforward semantic dependencies between their
-respective relaxed atomic loads and relaxed atomic stores, so an OOTA
-cycle cannot form in a correct real-world C++ implementation.
+All processes's loads' return values affect the value stored, so all
+processes have semantic dependencies between their respective relaxed
+atomic loads and relaxed atomic stores.
+Therefore, an OOTA cycle cannot form in a correct real-world C++
+implementation.
 
@@
 \label{app:Multi-Cycle Litmus Tests}
+
+Figure~\ref{lst:Multi-Cycle Litmus Test}
+shows an expansion to four threads of Causality Test Case 4, which is shown in
+Listing~\ref{lst:Causality Test Case 4}
+and discussed in
+Appendix~\ref{app:Causality Test Case 4}.
+The \co{herd7} tool shows multiple different possible OOTA cycles,
+as expected.
+However, the value returned by each thread's load directly affects the
+value stored, so there is a semantic dependency in each thread.
+Therefore, none of these OOTA cycles can happen in a correct real-world
+C++ implementation.
 

### 2023-12-17 ootavacuous.tex
Add Alan Stern's example invalidating (sdep U rf)
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 from their relaxed atomic loads to their relaxed atomic stores, which
-prevents this OOTA cycle on real-world C++ implementations.
+prevents this OOTA cycle on correct real-world C++ implementations.
 
@@
 JMM Causality Test Case 12 in
-Figure~\ref{lst:Causality Test Case 12}
+Listing~\ref{lst:Causality Test Case 12}
 uses arrays, which are not yet supported by the \co{herd7} tool.
@@
 
-Figure~\ref{lst:Multi-Cycle Litmus Test}
+Listing~\ref{lst:Multi-Cycle Litmus Test}
 shows an expansion to four threads of Causality Test Case 4, which is shown in
@@
 \end{listing}
+
+\subsubsection{Three-Process Version of JMM Causality Test Case 4}
+\label{app:Three-Process Version of JMM Causality Test Case 4}
+
+Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
+shows a different way of expanding
+Listing~\ref{lst:Causality Test Case 4}
+(discussed in
+Appendix~\ref{app:Causality Test Case 4})
+from two processes to three processes.
+The \co{herd7} tool finds OOTA values, but in a rather odd manner.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-3proc.litmus @@
+\caption{Three-Process Version of JMM Causality Test Case 4}
+\label{lst:Three-Process Version of JMM Causality Test Case 4}
+\end{listing}
+
+All processes have straightforward semantic dependencies from their
+relaxed atomic loads to their relaxed atomic stores, which prevents this
+OOTA cycle on correct real-world C++ implementations.
+
+\subsubsection{Why rfe Instead of Tried-And-True rf?}
+\label{app:Why rfe Instead of Tried-And-True rf?}
+
+Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
+revises
+Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
+to flatten \co{PO()} and \co{P1()} into a single thread, so that the
+resulting \co{P0()} has an rfi link instead of an rfe link connecting
+the store to and load from \co{y}.
+In addition, the value stored to \co{y} is forced to be at least the
+value 17, and the value stored to \co{z} is forced to be at most the
+value 17.
+This of course means that the store to \co{z} must always be the value
+17, and that a C++ implementation could prove this by examining \co{P0()}
+in isolation given that it is also legal for a C++ implementation
+to replace the value loaded from \co{y} with the value just stored.\footnote{
+	This is not an option in C because atomic operations are volatile,
+	and thus observed behavior.}
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-whyrfe.litmus @@
+\caption{Why rfe Instead of Tried-And-True rf?}
+\label{lst:Why rfe Instead of Tried-And-True rf?}
+\end{listing}
+
+In the confines of the C++ abstract machine (omitting the no-OOTA
+restriction), this allows a value-17 OOTA cycle, courtesy of the fact
+that C++ respects neither control nor data dependencies.
+However, the \co{herd7} tool does not identify this cycle, perhaps due
+to its restrictions on its handling of arithmetic on OOTA values.
+
+In a real-world implementation that proves that the value 17 is always
+stored to \co{z}, there is no semantic dependency to
+\co{P0()}'s store to \co{z}, but there is to all of the other relaxed
+atomic stores.
+On the other hand, if the implementation does not prove the value
+stored to \co{z}, there will be a global-time delay between each of
+\co{P0()}'s loads and its store to \co{z}, preventing a value-17 OOTA
+cycle from forming.
+
+Either way, the value-17 OOTA cycle cannot occur in a correct real-world
+C++ implementation.
+Yes, an implementation that proves that the value 17 is always
+stored to \co{z} can satisfy the \co{exists} clause, but because
+there is no semantic dependency to \co{P0()}'s store to \co{z},
+this is not an OOTA cycle.
+
+But what if we chose to define an OOTA cycle as (sdep $\cup$ rf) instead
+of (sdep $\cup$ rfe)?
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-whyrfe-z17.litmus @@
+\caption{Why rfe Instead of Tried-And-True rf When z=17?}
+\label{lst:Why rfe Instead of Tried-And-True rf When z=17?}
+\end{listing}
+
+In that case, it would be necessary to separately consider the dependency
+from \co{P0()}'s load from \co{x} and \co{y} and that from its load
+from \co{y} and its store to \co{z}.
+The problem with this is shown in
+Listing~\ref{lst:Why rfe Instead of Tried-And-True rf When z=17?},
+which shows an equivalent program given the knowledge that the value~17
+will always be stored to \co{z}.
+Because all atomic operations are relaxed and because C++ does not
+respect control dependencies, the store to \co{z} can be moved up
+to the very beginning of \co{P0()}.
+This transformation causes the rfi link associated with \co{y} to
+go backwards in time, which is perfectly legitimate in this case,
+but which makes defining OOTA cycles in terms of (sdep $\cup$ rf)
+problematic.
+
+Again, in C, the load from \co{y} cannot be removed because it is
+volatile, and thus observable behavior.
+C++ could remove that load, but we left it in to demonstrate the all-17s
+outcome.
+
+This example demonstrates why OOTA cycles must be defined in terms
+of (sdep $\cup$ rfe), or in \co{herd7} nomenclature,
+\co{acyclic(sdep; rfe)}.
 

### 2023-12-17 ootavacuous.tex
Add Alan Stern's invented-load litmus tests
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Alan Stern December 17 feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 data dependencies, and flags it as such using \co{S8}.
-Interestingly enough, there is no state with \co{r3} also shown as
-\co{S8}, perhaps indicating that \co{herd7} does not propagate OOTA
-values.
 
@@
 
-Both processes have straightforward semantic dependencies between their
-respective relaxed atomic loads and relaxed atomic stores, so an OOTA
-cycle cannot form in a correct real-world C++ implementation.
+\co{P1()} has a straightforward semantic dependency between its
+relaxed atomic load and relaxed atomic store.
+However, suppose that \co{P0()}'s load from \co{x} on line~8 returns
+the value 42.
+In that case, the second load on line~11 can be fused with the first,
+so that it also returns the value 42.
+On the other hand, if the load on line~8 returns some other value, then
+line~10 will store the value 42 to \co{x}, and then the load on line~11
+can be elided\footnote{
+	In C++, but not in C due to atomics being volatile.}
+in favor of using the value 42 stored by line~10.
+
+Either way, the value of \co{r1} will be the constant 42.
+Therefore, the loads from \co{x} cannot affect the store on line~17,
+so there is no semantic dependency.
+This allows that store to \co{y} to be moved to the very beginning of
+\co{P0()}, thus allowing a correct real-world C++ implementation to
+satisfy the \co{exists} clause.
 
@@
 \co{acyclic(sdep; rfe)}.
+
+\subsubsection{Inventing Atomic Loads}
+\label{app:Inventing Atomic Loads}
+
+Listing~\ref{lst:Inventing Atomic Loads}
+and~\ref{lst:Atomic Loads Invented}
+shows a TBD point about inventing atomic loads.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-invented-load.litmus @@
+\caption{Inventing Atomic Loads}
+\label{lst:Inventing Atomic Loads}
+\end{listing}
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-load-invented.litmus @@
+\caption{Atomic Loads Invented}
+\label{lst:Atomic Loads Invented}
+\end{listing}
 

### 2023-12-17 ootavacuous.tex
Add Michael Wong's summary of debates on causality
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+\section{Debates on Causality}
+\label{sec:Debates on Causality}
+
+This Appendix is provided as a service to readers who want to know more
+about the ongoing debates on causality.
+References may be found here:
+
+{ \scriptsize
+\url{https://docs.google.com/document/d/1MJ6pQ4TcWchUQvoI-qhLjdhPVYF14AytbTW3gr4-xag/edit?usp=sharing}
+}
+
+The concept of causality, the idea that events are caused by preceding
+events in a predictable and consistent manner, is fundamental to many
+scientific theories and our understanding of the world.
+However, when it comes to the universe as a whole, particularly at the
+most fundamental levels such as in quantum physics or the origins of
+the universe, the idea of strict causality can become more complex.
+
+In classical physics, causality is often described in terms of
+deterministic laws, where events are determined by prior causes.
+But at the quantum level, the behavior of particles can be probabilistic
+and uncertain.
+Quantum mechanics introduces a level of randomness and uncertainty that
+challenges our classical notions of causality.
+
+Additionally, when considering the origin of the universe (the Big Bang),
+the laws of physics as we understand them break down at extremely high
+densities and temperatures.
+The conditions at the very beginning of the universe might not conform
+to our classical understanding of causality, and some theories suggest
+that time itself might have begun with the Big Bang, making it difficult
+to define causality in a traditional sense within such a framework.
+
+On the one hand, there is a strong case to be made that the universe
+is causal.
+Our understanding of physics is based on the principle of causality,
+which states that every event has a cause.
+This principle has been incredibly successful in explaining and predicting
+the behavior of the universe at large scales.
+Additionally, our own personal experience seems to confirm causality.
+We constantly observe events being caused by other events, and we have
+no reason to believe that this is an illusion.
+
+However, there are also some challenges to the idea of a perfectly
+causal universe.
+One challenge comes from quantum mechanics, which suggests that some
+events may be fundamentally non-causal.
+For example, the collapse of a wave function in quantum mechanics seems
+to be a random event, with no underlying cause.
+Additionally, our understanding of the universe's origins is still
+very limited.
+We don't know whether the Big Bang itself had a cause, or whether it
+was simply the beginning of everything.
+
+In summary, while causality is a fundamental concept in our understanding
+of the universe, its application and implications can become more complex,
+especially at the quantum level and regarding the origin of the universe.
+There are ongoing debates and explorations within physics and philosophy
+about the nature of causality in these contexts.
+
 % \section{History}

### 2023-12-17 data/plots.sh
Update x86 coe example to more typical scenario
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -6,7 +6,7 @@ set output "../coe-sh-out.png"
 set xlabel "Time (Timestamp Periods)"
 set ylabel "Number of Simultaneous Values"
 #set logscale y
-set xrange [:3500]
+set xrange [:750]
 #set yrange [100:10000]
 set nokey
 # set label 1 "RCU" at 0.1,10 left
@@ -14,7 +14,7 @@ set nokey
 # set label 3 "brlock" at 0.4,0.6 left
 # set label 4 "rwlock" at 0.3,1.6 left
 # set label 5 "refcnt" at 0.15,2.8 left
-plot "coe.out-4.dat" w l
+plot "coe.out-1.dat" w l
 ---EOF---
 
 gnuplot << ---EOF---

### 2023-12-17 ootavacuous.sh
Add diagrams explaining temporal properties of coe, fre, and rfe
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -85,7 +85,7 @@ do
 	if test ! -f "$basename.pdf" -o "$basename.svg" -nt "$basename.pdf"
 	then
 		echo "$basename.svg -> $basename.pdf"
-		inkscape --export-pdf=$basename.pdf $basename.svg
+		XDG_RUNTIME_DIR=na DBUS_SESSION_BUS_ADDRESS=na inkscape -o $basename.pdf < $basename.svg
 	fi
 done
 

### 2023-12-18 ootavacuous.tex
Start appendices on new page and transcribe todo lists
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Update x86 coe example to more typical scenario
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Add diagrams explaining temporal properties of coe, fre, and rfe
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Propagation delay and control dependencies
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 it extends to \co{y} and in others to \co{z}.
+This is an example of the load affecting not the value or address of
+a given store, but rather whether or not that store is executed at all.
 
@@
 
+\clearpage
 \appendix
@@
 exists between a given relaxed atomic load and store on any execution
-where a different value loaded would change either the address or the
-value of that store.
-Producing a precise formal statement of this rule is quite challenging.
+where a different value loaded would change the address stored to, the
+value stored, or the whether or not that store would be executed in
+the first place.
+Producing a precise formal statement of this rule turns out to be quite
+challenging.
 
@@
 of the proposed \co{std::observable()}.
+
+\clearpage
 
@@
 
+\clearpage
+
 \section{Inter-Process Communications}
@@
 \begin{center}
-\resizebox{5in}{!}{\includegraphics{coe-out-4}}
+\resizebox{5in}{!}{\includegraphics{coe-out-1}}
 \caption{On x86, coe Links Are Partially Ordered}
@@
 
+\begin{figure}[tb]
+\begin{center}
+\resizebox{4in}{!}{\includegraphics{coe}}
+\resizebox{4in}{!}{\includegraphics{fre}}
+\resizebox{4in}{!}{\includegraphics{rfe}}
+\caption{Propagation Delay and Temporal Properties of coe, fre, and rfe}
+\label{fig:Propagation Delay and Temporal Properties of coe, fre, and rfe}
+\end{center}
+\end{figure}
+
+Figure~\ref{fig:Propagation Delay and Temporal Properties of coe, fre, and rfe}
+shows how the temporal properties of coe, fre, and rfe are explained
+by propagation delay.
+In the case of coe, the two stores take time to propagate across the
+system, and their ordering can and is decided after the fact.
+In the case of fre, the store again takes time to propagate across
+the system, and so a load might execute later in global time than
+did the store, but not late enough to return that store's value.
+In the case of rfe, the load cannot return a given store's value until
+the value has propagated to that part of the system executing that load.
+
 And this is exactly why the C++ memory model guarantees ordering from
@@
 absence of other ordering from stronger atomic memory accesses or
-\co{atomic_thread_fence()}.
+\co{atomic_thread_fence(memory_order_seq_cst}.
 
@@
 
+\clearpage
+
 \section{Debates on Causality}
@@
 % \label{sec:History}
+
+\clearpage
 
@@
 % @@@ TODO:
-% Measure store-buffer delays on an x86 system.
-% Real-world compilers freeze nondeterministic choices.
+% Convert remaining PPCMEM diagrams to herd7
+% Emphasize that real-world implementations map abstract machine to matter
+%	and energy in the real world.
+% Look into why inkscape no longer likes generating PDFs.
 

### 2023-12-19 ootavacuous.tex
Credit where it is due for causality appendix
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Convert CPPMEM litmus tests to herd7
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{listing}[tbp]
-\begin{verbatim}
- 1 int main() {
- 2   atomic_int x = 2;
- 3   volatile int r1;
- 4   {{{ { x.store(42, memory_order_relaxed); // invented store
- 5         x.store(3, memory_order_relaxed); }
- 6   ||| r1 = x.load(memory_order_relaxed);
- 7   }}};
- 8   return 0;
- 9 }
-\end{verbatim}
-\caption{CPPMEM Invented Store}
-\label{lst:CPPMEM Invented Store}
+@@ DisplayRunLitmus litmus/invented-store.litmus @@
+\caption{Example Invented Store}
+\label{lst:Example Invented Store}
 \end{listing}
@@
 forbidden by the abstract machine.
-To see this, consider Listing~\ref{lst:CPPMEM Invented Store},
-a CPPMEM\footnote{
-	\url{http://svr-pes20-cppmem.cl.cam.ac.uk/cppmem/index.html}.}
-litmus test that demonstrates such a behavior.
-Without line~4, only the values~2 and~3 can be stored to \co{r1}.
-With that line, the additional value~42 can also be stored to \co{r1}.
+To see this, consider Listing~\ref{lst:Example Invented Store},
+a \co{herd7} litmus test that demonstrates such a behavior.
+Without line~8, only the values~0 and~3 can be loaded into \co{r1} on
+line~13.
+With line~8, the additional value~42 can also be loaded into \co{r1}.
 The implementation is therefore forbidden from inventing that store of 42
@@
 \begin{listing}[tbp]
-\begin{verbatim}
- 1 int main() {
- 2   atomic_int x = 0;
- 3   volatile int r1;
- 4   {{{ { x.store(1, memory_order_relaxed); // duplicated store
- 5         x.store(1, memory_order_relaxed); }
- 6   ||| { r1 = x.load(memory_order_relaxed);
- 7         x.store(2, memory_order_relaxed); }
- 8   }}};
- 9   return 0;
-10 }
-\end{verbatim}
-\caption{CPPMEM Duplicated Store}
-\label{lst:CPPMEM Duplicated Store}
+@@ DisplayRunLitmus litmus/duplicated-store.litmus @@
+\caption{Example Duplicated Store}
+\label{lst:Example Duplicated Store}
 \end{listing}
@@
 behaviors.
-To see this, consider Listing~\ref{lst:CPPMEM Duplicated Store}, a CPPMEM
+To see this, consider Listing~\ref{lst:Example Duplicated Store}, a
 litmus test that demonstrates such a behavior.
-Without line~4, if the final value of \co{r1} is 1, then the final value
+Without line~9, if the value loaded into \co{r1} is 1, then the final value
 of \co{x} must be 2.
-With that line, the final value of \co{x} can be 1 even when the
-final value of \co{r1} is 1.
+With line~9, the final value of \co{x} can be 1 even when the
+value loaded into \co{r1} is 1.
 The implementation is therefore forbidden from duplicating that store of 1
@@
 This Appendix is provided as a service to readers who want to know more
-about the ongoing debates on causality.
+about the ongoing debates on causality, provided by the combined efforts
+of Bard, Chat-GPT, and Bing Co-Pilot.
 References may be found here:
@@
 %	and energy in the real world.
-% Look into why inkscape no longer likes generating PDFs.
 

### 2023-12-19 ootavacuous.tex
Add examples and text for fungible stores
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 a given store, but rather whether or not that store is executed at all.
+
+But beware of this example:
+
+\begin{quote}
+	\co{r1 = x;} \\
+	\co{if (r1 > 0)} \\
+	\co{\ \ \ \ y = 42;} \\
+	\co{else} \\
+	\co{\ \ \ \ y = 42;}
+\end{quote}
+
+Because the stores executed on each branch of that "if" statement
+store identical values to identical addresses, they are for all intents
+and purposes the same store.
+Therefore, there is no semantic dependency~\cite{Howells2009membartxt}.
 
@@
 challenging.
+For but one example, identical stores are fungible, so that if all
+executions always execute one of several equivalently positioned stores
+having identical addresses and values, there is no semantic dependency
+from any load to any of those stores~\cite{Howells2009membartxt}.
 
@@
 % @@@ TODO:
-% Convert remaining PPCMEM diagrams to herd7
 % Emphasize that real-world implementations map abstract machine to matter

### 2023-12-19 ootavacuous.tex
Convert untagged litmus test to herd7
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{listing}[tbp]
-\begin{verbatim}
- 1 int foo(int i) {
- 2   atomic_int x = 0;
- 3   int y[3];
- 4   volatile int r1;
- 5   {{{ { y[i].store(42, memory_order_relaxed);
- 6         x.store(1, memory_order_relaxed); }
- 7   ||| r1 = x.load(memory_order_relaxed);
- 8   }}};
- 9   return 0;
-10 }
-\end{verbatim}
+@@ DisplayLitmus litmus/speculative-store.litmus @@
 \caption{Speculated Store and Non-Speculated Load}
@@
 suppose that the storage for \co{x} is located immediately before that
-for \co{a[]}.
+for \co{y[]}.
 Suppose further that hardware speculation incorrectly guesses the value
-of \co{i} to be -1, so that the speculated store of 42 on line~5 uses
+of \co{i} to be -1, so that the speculated store of 42 on line~9 uses
 the address of \co{x}.\footnote{
@@
 	nor cares, nor should it.}
-Then line~7 can cause the final value of \co{r1} to be 42, when
+Then line~11 can cause the value 42 to be loaded into \co{r1}, when
 by the rules of the C++ abstract machine the value must instead be

### 2023-12-19 ootavacuous.tex
Explicitly call out mapping of abstract machine onto matter/energy
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Real-world constraints are imposed by the standard, which have been
-considered in prior work, and by the laws of physics, which have
-not.
-Additional constraints are imposed by hardware architecture and
-design, which have been partially accounted for in the standard,
-one limitation being that accurate and executable formal descriptions
-of hardware memory models did not appear until after the standard
-was released.
+considered in prior work.
+However, in correct real-world C++ implementations, the standard's
+abstract machine must be mapped onto real matter and energy.
+This mapping has not been considered by prior work, and brings to
+bear additional constraints from the laws of physics and from hardware
+architecture and design, both of which have been only partially accounted
+for in the standard.
+One reason for the incomplete accounting for hardware constraints is
+that accurate and executable formal descriptions of hardware memory
+models did not appear until after the standard was released.
 A final set of constraints is imposed by the C++ implementations
-themselves.
+themselves, some of which have resulted in non-determinisms within
+the standard.\footnote{
+	For example, when pre-existing implementations could not agree
+	on any specific deterministic handling of side-effects within
+	expressions.
+	On the other hand, implementer freedom does have significant
+	value.}
 
@@
 % @@@ TODO:
-% Emphasize that real-world implementations map abstract machine to matter
-%	and energy in the real world.
 

### 2023-12-19 ootavacuous.tex
Reference Appendix instead of Retrocausality URL
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 or any other apparatus that might result in effect preceding cause,
-as discussed at length in the Wikipedia article on retrocausality and
-its many references.\footnote{
-	\url{https://en.wikipedia.org/wiki/Retrocausality}.}
+as discussed at length in Appendix~\ref{sec:Debates on Causality} and 
+the URL that it provides.
 

### 2023-12-19 ootavacuous.tex
December 18 self-review
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 However, all macroscopic measurements to date have shown the speed of
-light to be finite, thus preventing causal loops based on infinite-speed
-propagation of information.
-
+light to be finite.
 This paper will therefore assume that the speed of light is finite, and
 that the maximum speed at which information can propagate is also finite.
+
+If empirical evidence continues to validate this assumption, it is
+not possible to form causal loops based on infinite-speed propagation
+of information.
 
@@
 This paper will therefore assume that any computing device capable of
-running C++ will be of non-zero size.
+running C++ will be of non-zero size, which rules out formation of
+causal loops based on zero-size computing systems.
 
@@
 
-% Done as of November 30, 2023:
+% Done as of December 18, 2023:
 % David Goldblatt

### 2023-12-21 ootavacuous.tex
Self-review in sending initial December 19 email to Alan
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Call out "no invented load" in filename
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Add intermediate three-thread whyrfe litmus tests
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
+so that the value stored to \co{y} is forced to be at most the
+value 17, and the value stored to \co{z} is forced to be at least the
+value 17.
+This of course means that the store to \co{z} must always be the value
+17, as can be seen from the \co{herd7} output.
+However, a single-thread-at-a-time analysis could not prove this.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-whyrfe-3.litmus @@
+\caption{Three-Process Version of JMM Causality Test Case 4 With Max and Min}
+\label{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
+\end{listing}
+
+Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
+further revises
+Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
 to flatten \co{PO()} and \co{P1()} into a single thread, so that the
@@
 the store to and load from \co{y}.
-In addition, the value stored to \co{y} is forced to be at least the
-value 17, and the value stored to \co{z} is forced to be at most the
-value 17.
-This of course means that the store to \co{z} must always be the value
-17, and that a C++ implementation could prove this by examining \co{P0()}
-in isolation given that it is also legal for a C++ implementation
-to replace the value loaded from \co{y} with the value just stored.\footnote{
+This means that a C++ implementation could prove that the value 17 is
+always stored to \co{z} by examining \co{P0()} in isolation, given that
+it is also legal for a C++ implementation to replace the value loaded
+from \co{y} with the value just stored.\footnote{
 	This is not an option in C because atomic operations are volatile,
@@
 \begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-invented-load.litmus @@
+@@ DisplayRunLitmus litmus/oota-no-invented-load.litmus @@
 \caption{Inventing Atomic Loads}
@@
 % @@@ TODO:
+% Construct interleaved OOTA examples (see sheets of paper).
+% Construct nondeterminism example, for example, based on i++.
+% Check cmbc on nondeterministic i++.
+% Use LKMM as check for dependencies?  Check invented-load litmus tests.
 

### 2023-12-22 ootavacuous.tex
Add invented-load litmus test and commentary
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Listing~\ref{lst:Inventing Atomic Loads}
-and~\ref{lst:Atomic Loads Invented}
-shows a TBD point about inventing atomic loads.
+This section contains several example litmus tests from the active
+mind of Alan Stern.
 
@@
 
+First, consider Listing~\ref{lst:Inventing Atomic Loads}.
+At first glance, this might appear straightforward.
+\co{P0()}'s store to \co{z} clearly depends on its load from \co{y},
+and \co{P1()}'s store to \co{y} clearly depends on its load from \co{z}.
+If both dependencies are semantic dependency, the OOTA cycle cannot
+form.
+
 \begin{listing}[tbp]
@@
 \end{listing}
+
+But is \co{P0()}'s dependency a semantic dependency?
+For example, is a real-world C++ implementation permitted to
+duplicate \co{P0()}'s load from \co{x} and then make use of both loads'
+values?
+If this is permitted, then the implementation would be within its rights
+to transform
+Listing~\ref{lst:Inventing Atomic Loads}
+into
+Listing~\ref{lst:Atomic Loads Invented}.
+
+Because the values of \co{r1}, \co{r2}, and \co{r3} are all known to be
+either zero or one, if the values returned by the two loads from \co{x}
+differ, then one of those two values must be equal to that from \co{y}.
+If the implementation is further permitted to make a runtime choice
+between the two values loaded from \co{x}, it could always choose the
+one that was equal to the value loaded from \co{y}, as shown by the
+\co{if} statement on line~13.
+This would mean that the store to \co{z} was always the constant one,
+which would permit the store to be ordered before the load from \co{y},
+as shown in lines~14 and~15.
+Doing so permits the OOTA-like outcome of \co{r3 == r4 == 1}.
+
+The implementation might also drop the load from \co{y}, whose value
+is now unused except by the \co{exists} clause.
+However, we use the convention that the \co{exists} clause is observable
+behavior.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-unused-load.litmus @@
+\caption{Unused Extra Atomic Load}
+\label{lst:Unused Extra Atomic Load}
+\end{listing}
+
+One possible reaction to this situation is to declare that useful
+real-world C++ implementations should never duplicate non-volatile
+relaxed atomic loads.
+But suppose that the initial code was as shown in
+Listing~\ref{lst:Unused Extra Atomic Load}?\footnote{
+	Such unused loads can easily be generated by complex macros or
+	template metaprograms.}
+In this case, the transformation to
+Listing~\ref{lst:Atomic Loads Invented}
+would not require inventing a load, but instead merely making use of
+the previously discarded value in \co{r0}.
+
+There are a number of possible reactions to this situation:
+
+\begin{enumerate}
+\item	Attempts to use real-world constraints to prevent OOTA cycles
+	are futile.
+	What to do instead?
+\item	Implementations should not invent uses for values from
+	non-volatile relaxed atomic loads.
+	But are there any useful optimizations that invent such uses?
+	(One can argue that in all such cases, the loads should instead
+	be fused into a single load.)
+\item	Listing~\ref{lst:Inventing Atomic Loads}
+	looks like it has an OOTA cycle, but there is in fact no
+	semantic dependency.
+	But then how exactly does one identify a semantic dependency?
+\item	What other equally interesting situations might come to light?
+\item	Your reaction here!
+\end{enumerate}
 

### 2023-12-28 ootavacuous.tex
Add examples based on Collatz conjecture and on FLT
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 code inspection.
+
+But consider the following code fragment:
+
+\begin{quote}
+	\co{x = collatz_cycle_min(y);}
+\end{quote}
+
+Here, \co{collatz_cycle_min()}\footnote{
+	\url{https://en.wikipedia.org/wiki/Collatz_conjecture}}
+takes its arbitrary-precision-integer argument and repeatedly either
+divides it by two if it is even, or multiplies it by three and adds one
+if it is odd.
+Should the resulting sequence of arbitrary-precision integers reach
+a cycle, the function returns the minimum element of that cycle.
+Leaving aside the practical complication of a value of \co{y} that
+consumes more than half of the available memory, there is reason to
+believe that the above code fragment will always assign the value 1 to
+\co{x}.
+Except that this reason for belief is based on the Collatz conjecture,
+which is unproven and might never be proven.
+
+A similar example could be constructed from Fermat's Last Theorem
+(FLT),\footnote{
+\url{https://en.wikipedia.org/wiki/Fermat\%27s_Last_Theorem}}
+which was first stated in the year 1637 and finally proved more than
+300~years later, in 1994.
+This of course means that it was only after 1994 that code relying on
+FLT was known not to have a semantic dependency.
+
+One could quite reasonably argue that these two examples are irrelevant
+to any real-world C++ implementation.
+After all, evidence to date suggests that code relying on either the
+Collatz Conjecture or FLT is quite rare, and thus not worth optimizing.
+And the value of any given optimization is another real-world constraint
+on the problem of locating OOTA cycles, but one that will not be considered
+further in this document.
+
+\emph{A C++ implementation must assume that a given dependency is
+a semantic dependency unless and until it can prove otherwise.}
 

### 2023-12-29 ootavacuous.tex
Non-unitary semantic dependencies
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 constant zero, at least up to the next assignment to \co{x}.
+
+Worse yet, suppose that both \co{y} and \co{z} are zero.
+Then there is no semantic dependency from either to \co{x} unless both
+\co{y} and \co{z} change.
+Alternatively, when both \co{y} and \co{z} are zero, there is no semantic
+dependency from either \co{y} or \co{z} to \co{x}, but rather a semantic
+dependency from the combination of \co{y} and \co{z} to \co{x}.
+This means that precise definitions of sdep cannot always rely on
+single-variable perturbations.
 

### 2024-01-04 ootavacuous.tex
Add OOTA-definition examples for context and loads
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+A similar caution applies to loads:
+
+\begin{quote}
+	\co{r1 = x;} \\
+	\co{if (r1 > 42)} \\
+	\co{\ \ \ \ r2 = y;} \\
+	\co{else} \\
+	\co{\ \ \ \ r2 = y;} \\
+	\co{z = r2;}
+\end{quote}
+
+The two loads from \co{y} are fungible, and one should therefore expect
+a compiler to replace this \co{if} statement with a single load from \co{y}.
+
 These examples demonstrate a key point: Although in some cases sdep can
@@
 % 	in the preceding one.}
+
+\subsubsection{Semantic Dependencies Depend on Context}
+\label{sec:Semantic Dependencies Depend on Context}
+
+Consider the following:
+
+\begin{quote}
+	\co{r1 = x;} \\
+	\co{// r1 = 0;} \\
+	\co{z = y * r1;} \\
+	\co{// z = y / r1;}
+\end{quote}
+
+As written, there is a semantic dependency from \co{x} and \co{y}
+to \co{z}.
+But if the second line is uncommented, there is no semantic
+dependency.
+If the second line is left commented, but the last line is uncommented,
+then the compiler is free to backwards-propagate the potential
+divide-by-zero undefined behavior to precede the first assignment
+to \co{z}.
+This could prevent both assignments to \co{z} from executing, for example,
+due to a divide-by-zero exception.
+
+More subtle changes in context can also have an effect.
+For example:
+
+\begin{quote}
+	\co{r1 = x;} \\
+	\co{z = y * r1;} \\
+	\co{z = 42;}
+\end{quote}
+
+Given that there is no ordering, a C++ implementation could choose to
+act as if the first store to \co{z} was omitted, even if that store was
+a relaxed atomic store.
+But if \co{z} was marked volatile, both stores would need to be executed
+in strict accordance with the rules of the abstract machine.
+
+\emph{Any analysis of semantic dependencies must either be fully informed
+of the enclosing context, or much make conservative assumptions about
+that context.}
 
@@
 
-\emph{A C++ implementation must assume that a given dependency is
+\emph{Because of unsolved mathematical problems, precise definitions of
+``semantic dependency'' are problematic.
+Therefore, C++ implementations must assume that a given dependency is
 a semantic dependency unless and until it can prove otherwise.}

### 2024-01-06 ootavacuous.tex
OOTA definitions and I/O: Complete the thought
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Wordsmith "what about non-volatile" section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 In addition, I/O operations are observed behavior, which greatly restricts
-optimizations involving them.
+optimizations involving them, in turn reducing the need to evaluate
+potential semantic dependencies.
 
@@
 
+Many analysis tools would respond to these examples by failing, for example,
+by consuming all available CPU and memory.
+
 One could quite reasonably argue that these two examples are irrelevant
@@
 \emph{Because of unsolved mathematical problems, precise definitions of
-``semantic dependency'' are problematic.
+``semantic dependency'' have limitations.
 Therefore, C++ implementations must assume that a given dependency is
@@
 which in turn implies that OOTA cycles cannot form.
+
+\subsection{But What About Non-Volatile Atomics?}
+\label{sec:But What About Non-Volatile Atomics?}
+
+Volatile atomic accesses have some very attractive properties from
+the viewpoint of analyzing executions for OOTA behavior because C++
+compilers are prohibited from reordering, omitting, fusing, or inventing
+volatile accesses.
+Thus, if the compiler emits code that faithfully reflects the behaviors
+specified by the source code, the hardware memory model will prohibit
+OOTA cycles.
+
+However, there is some question as to what exactly a C++ implementation
+is permitted to do with non-volatile atomic accesses~\cite{JFBastien2015N4455}.
+For example,
+Appendix~\ref{app:Inventing Atomic Loads}
+shows a clever litmus test that results in OOTA-like behavior
+if a C++ implementation duplicates an atomic load, compares the two values
+loaded, and then, if the two loads return different values, chooses the
+one that eliminates a semantic dependency that would otherwise remain
+in place.
+Perhaps in the fullness of time, such examples will be considered to be
+simple optimizations rather than full-on OOTA cycles, but in the meantime
+C++ implementations might wish to prohibit such behavior.
+
+One way to accomplish this while still permitting optimization of code
+containing non-volatile atomics is to model non-volatile behavior on that
+of volatile atomics, but with these specific weakenings:
+
+\begin{itemize}
+\item	Loads that do not affect observable behavior can be omitted,
+	along with any computations based on their return values.
+\item	Given a pair of relaxed loads from the same object that can
+	be reordered to be adjacent to each other, one of the pair can
+	be omitted with uses of the omitted load instead using the value
+	returned by the surviving load.
+	This is termed \emph{load fusing}.
+\item	Given a pair of relaxed stores to the same object that can
+	be reordered to be adjacent to each other, the first of the pair
+	can be omitted.
+	This is termed \emph{store fusing}.
+\item	If an expression can be proven to always result in a constant,
+	that constant can be substituted for that expression.
+	Note well that combining this with specialization optimizations
+	can convert address and data dependencies to weaker control
+	dependencies.
+	Address and data dependencies order later loads and stores,
+	and control dependencies order only later stores.\footnote{
+		Yes, yes, the C++ standard does not mention address or
+		data dependencies, and mandates control dependencies
+		only indirectly.
+		Such dependencies nevertheless still exist.}
+\end{itemize}
+
+Note that load and store fusing can only be applied a finite number of
+times.
+In practice, this means that fusing can be used to unroll loops, but
+not to hoist atomic loads or stores out of loops than cannot be proven
+to be always finite.
+
+These optimizations are examples, not necessarily a full set.
+Additional optimizations might be added over time, but great care will
+of course be required.
 

### 2024-01-06 ootavacuous.tex
Wordsmith "Inventing Atomic Loads" appendix reactions
Also wordsmith point about "exists" clause being observable behavior.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 is now unused except by the \co{exists} clause.
-However, we use the convention that the \co{exists} clause is observable
-behavior.
+However, the \co{exists} clause is considered to be observable behavior,
+which means that the load from \co{y} must remain.
 
@@
 \begin{enumerate}
-\item	Attempts to use real-world constraints to prevent OOTA cycles
-	are futile.
-	What to do instead?
-\item	Implementations should not invent uses for values from
-	non-volatile relaxed atomic loads.
-	But are there any useful optimizations that invent such uses?
-	(One can argue that in all such cases, the loads should instead
-	be fused into a single load.)
 \item	Listing~\ref{lst:Inventing Atomic Loads}
@@
 	semantic dependency.
-	But then how exactly does one identify a semantic dependency?
-\item	What other equally interesting situations might come to light?
-\item	Your reaction here!
+\item	Attempts to use real-world constraints to prevent OOTA cycles
+	are futile.
+	(But see Section~\ref{sec:But What About Non-Volatile Atomics?}.)
+\item	Implementations should not invent uses for values from
+	non-volatile relaxed atomic loads.
+	If it turns out that there are useful optimizations that invent
+	such uses, then such optimizations must be applied only with
+	careful attention to the as-if rule.
+	Some might argue that inventing comparisons between a pair of
+	non-volatile relaxed atomic loads from the same object should
+	be permitted only if the two loads could be reordered to be
+	adjacent to each other.
+\item	Any time an impolementation might be tempted to invent uses from
+	values from non-volatile relaxed atomic loads, the loads should
+	instead be fused into a single load.
+	This reaction might be attractive to those who carefully consider
+	the pointlessness of keeping two loads that return the same value
+	on the one hand or the cache-miss overhead involved should two
+	closely spaced loads from the same object return different values
+	on the other.
+\item	What other interesting situations might come to light?
 \end{enumerate}

### 2024-01-06 ootavacuous.tex
Check and wordsmith non-JMM litmus tests
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 using manual analysis.
+
 The manual analysis uses the informal rule that a semantic dependency
-exists between a given relaxed atomic load and store on any execution
-where a different value loaded would change the address stored to, the
+exists between a set of relaxed atomic loads and a store on any execution
+where different values loaded would change the address stored to, the
 value stored, or the whether or not that store would be executed in
-the first place.
+the first place, with the proviso that stores of the same value to the
+same address having the same ordering properties are fungible.
 Producing a precise formal statement of this rule turns out to be quite
-challenging.
-For but one example, identical stores are fungible, so that if all
-executions always execute one of several equivalently positioned stores
-having identical addresses and values, there is no semantic dependency
-from any load to any of those stores~\cite{Howells2009membartxt}.
+challenging, but as noted earlier, there is considerable work along
+various
+lines~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA}.
+For examples of the challenges, please see
+Section~\ref{sec:On OOTA Definitions}.
 
@@
 from two processes to three processes.
-The \co{herd7} tool finds OOTA values, but in a rather odd manner.
+The \co{herd7} tool finds OOTA values, as expected.
 
@@
 Listing~\ref{lst:Atomic Loads Invented}
-would not require inventing a load, but instead merely making use of
+would not require inventing a load, but instead merely inventing a use of
 the previously discarded value in \co{r0}.

### 2024-01-06 ootavacuous.tex
Expand informal definition of "semantic dependency"
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 the source code was duplicated in order to apply specialization
-optimizations, in which case any nondeterminism might also be
-duplicated.
+optimizations, in which case any non-determinism might also be
+a function of a given execution.
 
@@
 
+Compiler-based C++ implementions make their non-deterministic choices
+at compile time, so that the object code is deterministic, at least
+on a per-execution basis.
+
 \subsubsection{Semantic Dependencies And Optimization}
@@
 
-But suppose that the implementation can prove that at any time that
-the above statement might execute, \co{y} is equal to \co{z}?
+Consider again the following:
+
+\begin{quote}
+	\co{x = y + z;}
+\end{quote}
+
+Suppose that the implementation can prove that at any time that
+the above statement might execute, \co{y} is equal to \co{z}.
 That would allow the implementation to optimize, that is, to act as if
@@
 
-Please see Appendix~\ref{sec:User Influence Over Language Semantics}
+Please see Appendix~\ref{app:User Influence Over Language Semantics}
 for more examples of user control over semantics, including semantic
@@
 C++ implementations undertake a wide variety of optimizations, in
-some cases taking advantage of nondeterminism and in others applying
+some cases taking advantage of non-determinism and in others applying
 the results of code analysis to produce faster code.
@@
 Other c++ implementations (for example, those using interpreters or JITs)
-might need to directly model any nondeterminism.
+might need to directly model any non-determinism.
 This latter situation might motivate additional C++ projects to ban
-nondeterminism in favor of portable code.
+non-determinism in favor of portable code.
 
@@
 
+Appendix~\ref{app:Informal Definition of Semantic Dependency}
+outlines the informal informal definition of ``semantic dependency''
+used to evaluate the litmus tests in this section,
+Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}
+evaluates the obligatory litmus tests from the JMM Causality Test
+Cases, and
+Appendix~\ref{app:Additional Litmus Tests}
+evaluates some additional litmus tests that affected our thinking or
+otherwise gave us pause.
+
+\subsection{Informal Definition of Semantic Dependency}
+\label{app:Informal Definition of Semantic Dependency}
+
 The manual analysis uses the informal rule that a semantic dependency
@@
 same address having the same ordering properties are fungible.
+
 Producing a precise formal statement of this rule turns out to be quite
@@
 Section~\ref{sec:On OOTA Definitions}.
-
-Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}
-evaluates the obligatory litmus tests from the JMM Causality Test
-Cases, and
-Appendix~\ref{app:Additional Litmus Tests}
-evaluates some additional litmus tests that affected our thinking or
-otherwise gave us pause.
+Our informal definition meets these challenges as follows:
+
+\begin{enumerate}
+\item	Section~\ref{sec:Semantic Dependencies Are Per Execution}
+	is responsible for the phrase ``on any execution'' as well
+	as the proviso regarding fungible loads.
+\item	Section~\ref{sec:Semantic Dependencies Depend on Context}'s
+	challenges are met by looking at the full context within
+	a given thread.
+\item	Section~\ref{sec:Semantic Dependencies Can Be Nondeterministic}'s
+	non-determinism challenges are avoided by avoiding litmus tests
+	exhibiting non-deterministic behavior.
+	Again, note that most production C++ code uses compiler-based C++
+	implementations, which make their non-deterministic choices at
+	compile time.
+\item	Section~\ref{sec:Semantic Dependencies And Optimization}'s
+	optimization-choice challenges are met by using the C++
+	implementation's choice of semantic dependency based on the
+	optimizations that the implementation actually applied.
+\item	Section~\ref{sec:Semantic Dependencies and I/O}'s
+	I/O challenges are met by again noting that I/O is observed
+	behavior.
+	And by avoiding litmus tests involving I/O, given that relatively
+	few of the available tools handle I/O.
+\item	Section~\ref{sec:Semantic Dependencies Can Be Complex}'s
+	complexity challenges are met by looking at what the code
+	actually does.
+	For example, if an as-yet unproven mathematical conjecture
+	holds, the code will act consistently with that conjecture,
+	and vice versa.
+\item	Section~\ref{sec:Implementers and Users Influence Semantic-Dependencies Definition}'s
+	implementer-/user-choice challenges are met by accounting for
+	those choices.
+	For example, if a user has specified \co{-funsigned-char},
+	then the actual executions will be consistent with that choice.
+	For more examples, please see
+	Appendix~\ref{app:User Influence Over Language Semantics}.
+\end{enumerate}
+
+In short, we are treating ``semantic dependency'' in a manner similar
+to our treatment of the laws of physics.
+We let the actual executions provide an empirical basis for the analysis.
+
+We do not expect this approach to please everyone.
+However, those who are displeased might do well to reflect on the fact
+that the C++ standard is not mathematical in nature, which implies that
+any mathematical model of C++ is inherently an approximation.
+We look forward to the day when the standard gains more mathematical
+rigor, and perhaps the C++ memory model is a first step in that
+direction.\footnote{
+	But if so, the OOTA situation should be a caution!}
+Nevertheless, until that day, we must live with the standard in its
+current decidedly non-mathematical form.
 
@@
 \section{User Influence Over Language Semantics}
-\label{sec:User Influence Over Language Semantics}
+\label{app:User Influence Over Language Semantics}
 
@@
 % Construct interleaved OOTA examples (see sheets of paper).
-% Construct nondeterminism example, for example, based on i++.
 % Check cmbc on nondeterministic i++.

### 2024-01-06 ootavacuous.tex
Cross-linking, formatting, and wordsmithing
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{document}
-\title{D3064R0: OOTA Execution on Real Implementations is Provably Vacuous}
+\title{D3064R0: Making OOTA Execution on Real Implementations be Vacuous}
 
@@
 laws of physics and properties of hardware onto correct real-world
-C++ implementations.% \footnote{
+C++ implementations.
+% \footnote{
 %	Thanks to Alan Stern for providing a litmus test that demonstrated
@@
 %	to sdep $\cup$ rfe.}
+Please see
+Appendix~\ref{app:Informal Definition of Semantic Dependency}
+for more a more detailed (albeit quite informal) definition.
 
@@
 
+It is also important to note that implementations and users have a
+surprising degree of influence over language
+semantics~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.
+Please see
+Appendix~\ref{app:User Influence Over Language Semantics}
+for examples.
+
 \paragraph{Invented Atomic Stores}
@@
 program with the store present.
+
+For more discussion along these lines, please see
+Section~\ref{sec:But What About Non-Volatile Atomics?}.
 
@@
 universal property of the code itself.
+The interested reader is encouraged to create more complex examples,
+for example, demonstrating semantic dependencies from multiple loads or
+exploring indeterminacy.
 
@@
 	gcc -Wp,-MMD,kernel/rcu/.tree.o.d -nostdinc
-	-I./arch/x86/include -I./arch/x86/include/generated  -I./include
-	-I./arch/x86/include/uapi -I./arch/x86/include/generated/uapi
-	-I./include/uapi -I./include/generated/uapi
+	-I./arch/x86/include \\
+	-I./arch/x86/include/generated  -I./include
+	-I./arch/x86/include/uapi \\
+	-I./arch/x86/include/generated/uapi
+	-I./include/uapi -I./include/generated/uapi \\
 	-include ./include/linux/compiler-version.h
-	-include ./include/linux/kconfig.h -include
-	./include/linux/compiler\_types.h -D\_\_KERNEL\_\_
-	-fmacro-prefix-map=./= -Werror -std=gnu11 -fshort-wchar
-	-funsigned-char -fno-common -fno-PIE -fno-strict-aliasing -mno-sse
-	-mno-mmx -mno-sse2 -mno-3dnow -mno-avx -fcf-protection=branch
-	-fno-jump-tables -m64 -falign-jumps=1 -falign-loops=1
+	-include ./include/linux/kconfig.h \\
+	-include ./include/linux/compiler\_types.h -D\_\_KERNEL\_\_
+	-fmacro-prefix-map=./= \\
+	-Werror -std=gnu11 -fshort-wchar
+	-funsigned-char -fno-common -fno-PIE \\
+	-fno-strict-aliasing -mno-sse
+	-mno-mmx -mno-sse2 -mno-3dnow -mno-avx \\
+	-fcf-protection=branch
+	-fno-jump-tables -m64 -falign-jumps=1 -falign-loops=1 \\
 	-mno-80387 -mno-fp-ret-in-387 -mpreferred-stack-boundary=3
-	-mskip-rax-setup -mtune=generic -mno-red-zone -mcmodel=kernel
-	-Wno-sign-compare -fno-asynchronous-unwind-tables
-	-mindirect-branch=thunk-extern -mindirect-branch-register
+	-mskip-rax-setup \\
+	-mtune=generic -mno-red-zone -mcmodel=kernel
+	-Wno-sign-compare \\
+	-fno-asynchronous-unwind-tables
+	-mindirect-branch=thunk-extern \\
+	-mindirect-branch-register
 	-mindirect-branch-cs-prefix -mfunction-return=thunk-extern
 	-fno-jump-tables -fpatchable-function-entry=16,16
-	-fno-delete-null-pointer-checks -O2 -fno-allow-store-data-races
-	-fstack-protector-strong -fomit-frame-pointer
+	-fno-delete-null-pointer-checks -O2 \\
+	-fno-allow-store-data-races
+	-fstack-protector-strong -fomit-frame-pointer \\
 	-fno-stack-clash-protection -falign-functions=16
-	-fno-strict-overflow -fno-stack-check -fconserve-stack
-	-Wall -Wundef -Werror=implicit-function-declaration
-	-Werror=implicit-int -Werror=return-type -Werror=strict-prototypes
+	-fno-strict-overflow -fno-stack-check \\
+	-fconserve-stack
+	-Wall -Wundef -Werror=implicit-function-declaration \\
+	-Werror=implicit-int -Werror=return-type -Werror=strict-prototypes \\
 	-Wno-format-security -Wno-trigraphs -Wno-frame-address
-	-Wno-address-of-packed-member -Wframe-larger-than=2048 -Wno-main
-	-Wno-unused-but-set-variable -Wno-unused-const-variable -Wvla
-	-Wno-pointer-sign -Wcast-function-type -Wno-array-bounds
-	-Wno-alloc-size-larger-than -Wimplicit-fallthrough=5
+	-Wno-address-of-packed-member \\
+	-Wframe-larger-than=2048 -Wno-main
+	-Wno-unused-but-set-variable \\
+	-Wno-unused-const-variable -Wvla
+	-Wno-pointer-sign -Wcast-function-type \\
+	-Wno-array-bounds
+	-Wno-alloc-size-larger-than -Wimplicit-fallthrough=5 \\
 	-Werror=date-time -Werror=incompatible-pointer-types
-	-Werror=designated-init -Wenum-conversion
-	-Wno-unused-but-set-variable -Wno-unused-const-variable
+	-Werror=designated-init \\
+	-Wenum-conversion
+	-Wno-unused-but-set-variable -Wno-unused-const-variable \\
 	-Wno-restrict -Wno-packed-not-aligned -Wno-format-overflow
-	-Wno-format-truncation -Wno-stringop-overflow
-	-Wno-stringop-truncation -Wno-missing-field-initializers
+	-Wno-format-truncation \\
+	-Wno-stringop-overflow
+	-Wno-stringop-truncation -Wno-missing-field-initializers \\
 	-Wno-type-limits -Wno-shift-negative-value
-	-Wno-maybe-uninitialized -Wno-sign-compare
-	-DKBUILD\_MODFILE='"kernel/rcu/tree"' -DKBUILD\_BASENAME='"tree"'
-	-DKBUILD\_MODNAME='"tree"' -D\_\_KBUILD\_MODNAME=kmod\_tree -c
-	-o kernel/rcu/tree.o kernel/rcu/tree.c
+	-Wno-maybe-uninitialized \\
+	-Wno-sign-compare
+	-DKBUILD\_MODFILE='"kernel/rcu/tree"'
+	-DKBUILD\_BASENAME='"tree"' \\
+	-DKBUILD\_MODNAME='"tree"' -D\_\_KBUILD\_MODNAME=kmod\_tree \\
+	-c -o kernel/rcu/tree.o kernel/rcu/tree.c
 	}
@@
 % Check cmbc on nondeterministic i++.
-% Use LKMM as check for dependencies?  Check invented-load litmus tests.
+% Use LKMM as check for dependencies?
 

### 2024-01-07 ootavacuous.tex
Improve "Inventing Atomic Loads" listing caption
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 @@ DisplayRunLitmus litmus/oota-no-invented-load.litmus @@
-\caption{Inventing Atomic Loads}
-\label{lst:Inventing Atomic Loads}
-\end{listing}
-
-First, consider Listing~\ref{lst:Inventing Atomic Loads}.
+\caption{No Invented Atomic Loads}
+\label{lst:No Invented Atomic Loads}
+\end{listing}
+
+First, consider Listing~\ref{lst:No Invented Atomic Loads}.
 At first glance, this might appear straightforward.
@@
 to transform
-Listing~\ref{lst:Inventing Atomic Loads}
+Listing~\ref{lst:No Invented Atomic Loads}
 into
@@
 \begin{enumerate}
-\item	Listing~\ref{lst:Inventing Atomic Loads}
+\item	Listing~\ref{lst:No Invented Atomic Loads}
 	looks like it has an OOTA cycle, but there is in fact no

### 2024-01-07 ootavacuous.tex
January 7 self-review
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 }
-\date{December 15, 2023 (Pre-Tokyo)}
+\date{January 7, 2024 (Pre-Tokyo)}
 \maketitle{}
@@
 	have either been non-executable, complex, or unloved by C++
-	implementers.
+	implementers, and this paper summarizes some of the challenges.
 	But at the same time, there are no known instances of OOTA
@@
 	long-standing laws of physics and on hardware systems.
-	These constraints prevent OOTA from occurring in correctly
-	constructed C++ programs running on correctly constructed C++
-	implementations in our current universe.
+	These constraints prevent OOTA cycles containing only volatile
+	atomics from occurring in correctly constructed C++ programs
+	running on correctly constructed C++ implementations in our
+	current universe.
+	Given current optimizations, these constraints also prevent
+	OOTA cycles involving non-volatile atomics.
 
@@
 \end{abstract}
+
+\pagebreak
 
@@
 The arguments to both \co{P0()} and \co{P1()} specify which of
-the global shared variables each process may access.
-In this case, both processes may access both \co{x} and \co{y}.
+the global shared variables each process may access, in this case, \co{x}
+and \co{y}.
 The body of each process specifies a limited subset of C++ code,
@@
 an equality comparison.
+If a variable appears in an \co{exists} clause, then the final value of
+that variable constitutes observable behavior.
 
@@
 	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
-which may be found in
+which also may be found in
 Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}.
@@
 problem.
+This might have been acceptable in the 1960s of my youth, but it is now
+the year 2024.
 
@@
 Goldblatt looked at interactions between OOTA cycles and
-UB~\cite{DavidGoldblatt2019NoElegantOOTAfix}.
+UB~\cite{DavidGoldblatt2019NoElegantOOTAfix}, and
 Appendix~\ref{app:Aside on Undefined Behavior}
@@
 
-Prior work defines ``OOTA cycle'' by example, without a precise
+The earliest work defines ``OOTA cycle'' by example, without a precise
 definition.
@@
 Still others argue (perhaps correctly) that it is impossible to provide
-a general definition of ``OOTA cycle'', but even accepting this raises
+a general definition of ``OOTA cycle'', but this simply raises
 the question of whether a \emph{useful} definition can be formulated.
@@
 
-The current authors deviate slightly from this tradition by instead
-defining an OOTA cycle as a cycle in sdep $\cup$ rfe (or
-\co{acyclic(sdep ; rfe)} in \co{herd7} notation), where rfe is the
-set of store-to-load links, but excluding intra-thread rfi links, which
-are instead considered to be part of sdep.
+We deviate slightly from this tradition by instead defining an OOTA cycle
+as a cycle in sdep $\cup$ rfe (or \co{acyclic(sdep ; rfe)} in \co{herd7}
+notation), where rfe is the set of store-to-load links, but excluding
+intra-thread rfi links, which are instead considered to be part of sdep.
 Although this does shunt even more complexity onto the term
@@
 Worse yet, suppose that both \co{y} and \co{z} are zero.
-Then there is no semantic dependency from either to \co{x} unless both
-\co{y} and \co{z} change.
+Then there is no semantic dependency from either \co{y} or \co{z} to
+\co{x} unless both \co{y} and \co{z} change.
 Alternatively, when both \co{y} and \co{z} are zero, there is no semantic
 dependency from either \co{y} or \co{z} to \co{x}, but rather a semantic
-dependency from the combination of \co{y} and \co{z} to \co{x}.
+dependency from the combination of \co{y} \emph{and} \co{z} to \co{x}.
 This means that precise definitions of sdep cannot always rely on
@@
 
-There is a semantic dependency from both \co{y} and \co{z} if and only
-if these values do not cancel.
+There are semantic dependencies from \co{y} to \co{x} and from \co{z}
+to \co{x}.
 However, if the implementation knows that the value of \co{y} is always
-the negative of \co{x}, there is no semantic dependency and the
+the negative of \co{z}, there is no semantic dependency and the
 implementation can again act as if later accesses to \co{x} were
-instead the constant zero.
-
-Finally, consider this:
+instead the constant zero, again, at least until the next assignment
+to \co{x}.
+
+Also, consider this:
 
@@
 
-The two loads from \co{y} are fungible, and one should therefore expect
-a compiler to replace this \co{if} statement with a single load from \co{y}.
+The two loads from \co{y} are fungible, and so a C++ implementation
+could behave as if this \co{if} statement had instead been a single load
+from \co{y}.
 
@@
 implementation.
+Those wishing to change the value of \co{y} or \co{z} externally
+(for example, using a debugger) would do well to avail themselves
+of the \co{volatile} keyword.
+
+\emph{Semantic dependencies can be a free choice on the part of the
+C++ implementation.}
 
@@
 Except that this reason for belief is based on the Collatz conjecture,
-which is unproven and might never be proven.
+which is currently unproven and might never be proven (or disproven).
 
@@
 300~years later, in 1994.
-This of course means that it was only after 1994 that code relying on
-FLT was known not to have a semantic dependency.
+Therefore, it was only after 1994 that code relying on FLT was known
+not to have a semantic dependency.
 
@@
 \begin{quote}
-	\co{void foo(char c} \\
+	\co{void foo(char c)} \\
 	\co{\{} \\
@@
 
-If empirical evidence continues to validate this assumption, it is
-not possible to form causal loops based on infinite-speed propagation
+As long as empirical evidence continues to validate this assumption, it
+is not possible to form causal loops based on infinite-speed propagation
 of information.
@@
 propagate among computing elements in zero time.
-This situation would also permit a load to return the value stored by
+This situation would again permit a load to return the value stored by
 some other thread at the exact instant that the store executed, which
@@
 because computing systems can and observably do determine the
-modification order of a set of concurrent stores long after the fact.
+modification order of a set of concurrent stores long after the fact
+(see Appendix~\ref{app:Inter-Process Communications}).
 This means that the store that executed earlier in global time
@@
 speculative execution, but the hardware is required to prevent all such
-speculative exection from ever committing.
+speculative OOTA cycles from ever committing.
 
@@
 and the conditional on line~9 is evaluated non-speculatively.\footnote{
-	Alternatively, one could imagine that both the speculation
-	and the squashing of failed speculation might span multiple
-	threads.
+	Alternatively, and as noted earlier, one could imagine that both
+	the speculation and the squashing of failed speculation might
+	span multiple threads.
 	This might make sense for multiple hardware threads sharing
@@
 Note also that some architectures provide conditional-move instructions,
-which allows line~7 to avoid emitting a conditional branch.
+which allows line~9 to avoid emitting a conditional branch.
 However, such architectures would still treat a store of the
@@
 non-speculative stores.
+
+The key point is that a store having a semantic dependency on a given
+set of loads must execute after the completion of the last of those
+loads in global time.
 
@@
 The discussion of these two constraints is followed by a discussion of
-when redundant atomic stores may safely be omitted.
+when redundant atomic stores may safely be omitted and on fusing of
+non-volatile atomic loads from adjacent objects.
 
@@
 
-For more discussion along these lines, please see
+\paragraph{Loads From Adjacent Non-Volatile Atomic Objects}
+
+Suppose there are a pair of adjacent non-volatile atomic objects,
+and that these objects taken together are machine-word aligned
+and machine-word sized.
+Suppose further that the source code contains a pair of atomic
+loads, one from each of these objects, with the last of the pair
+being a relaxed load.
+Then the C++ implementation could choose to act as if the source code
+had instead contained a single atomic load from the combined
+pair of objects, with the memory-ordering semantics of the first load.
+
+For more discussion of potential optimizations, please see
 Section~\ref{sec:But What About Non-Volatile Atomics?}.
@@
 and/or previous stores in that object's modification order were executed
-by other threads (``coe'')), and loads from objects for which the value
+by other threads (``coe''), and loads from objects for which the value
 returned by a given load was overwritten by some other thread
@@
 execution.
-In this case, the proper course of action is to fix that bug.
+In this case, the proper course of action is to fix that bug, which is
+most likely a bug even in the absence of OOTA cycles.
 The third row depicts the case where the implementation misclassifies
@@
 which greatly simplifies analysis of semantic dependencies.
-Other c++ implementations (for example, those using interpreters or JITs)
+Other C++ implementations (for example, those using interpreters or JITs)
 might need to directly model any non-determinism.
@@
 Such tools can simply use that implementation's classification of
-dependencies as semantic on the one hand or non-semantic on the other,
-whether by working with that implementation's intermediate representations,
-binaries produced by that implementation, or by tracing the program's
-execution.
+dependencies as semantic on the one hand or non-semantic on the
+other, whether by working with that implementation's intermediate
+representations, binaries produced by that implementation, by tracing
+the program's execution, or, in the case of dynamic tools, by actually
+executing the program.
 While taking this approach sacrifices significant generality, it has the
@@
 
-Rather than look at how to prevent an implementation from forming
-an OOTA cycle, this section instead looks at how to form one.
-This will lead to the conclusion that a correct real-world C++
-implementation cannot form an OOTA cycle.
+Rather than look at how to prevent an implementation from forming an
+OOTA cycle, this section instead looks at how to form one.
+This will lead to the conclusion that a correct real-world C
+implementation cannot form OOTA cycles,\footnote{
+	Keep in mind that C does not have non-volatile atomic
+	operations.}
+and provide guidance on how to avoid OOTA cycles for non-volatile relaxed
+atomics in C++ implementations.
 
@@
 
-One way to accomplish this while still permitting optimization of code
-containing non-volatile atomics is to model non-volatile behavior on that
-of volatile atomics, but with these specific weakenings:
+Keeping in mind that some early C++ implementations treated non-volatile
+atomic operations as if they were volatile, one way to accomplish this
+while still permitting optimization of code containing non-volatile
+atomics is to model non-volatile behavior on that of volatile atomics,
+but with some specific weakenings, for example:
 
@@
 	This is termed \emph{store fusing}.
+\item	Suppose the source code contains a pair of non-volatile atomic
+	loads from objects that are adjacent in memory.
+	Suppose further that these loads can be reordered to be adjacent
+	to each other, and that the last of the two is a relaxed load.
+	Then this pair of loads can be replaced with a single load
+	from both objects using the first load's memory order.
+\item	Suppose the source code contains a pair of non-volatile atomic
+	stores to objects that are adjacent in memory.
+	Suppose further that these stores can be reordered to be adjacent
+	to each other, and that the first of the two is a relaxed load.
+	Then this pair of loads can be replaced with a single store
+	to both objects using the last store's memory order.
 \item	If an expression can be proven to always result in a constant,
@@
 
+Note also that many users would consider it to be an extremely unfriendly
+act for the C++ implemenation to combine loads or stores to adjacent
+objects where each object \co{is_lock_free()}, but the combination is not.
+
 These optimizations are examples, not necessarily a full set.
@@
 
-This section has shown that an implementation that correctly
-handles the memory model, observable behavior, and the as-if rule
-cannot form OOTA cycles when running on real systems in this
+This section has shown that an implementation that correctly handles
+the memory model, observable behavior, and the as-if rule cannot form
+OOTA cycles from volatile atomics when running on real systems in this
 universe.
+It has also presented a substantial list of optimizations that avoid
+creation of OOTA cycles from non-volatile atomics.
 
@@
 This work shows that correct real-world C++ implementations are inherently
-incapable of producing OOTA cycles.
-
-Therefore, if your C++ implementation produces OOTA cycles, you have bugs.
-And you need to fix those bugs, regardless of whether they are in your
-C++ implementation, your hardware, or your alternate universe.
+incapable of producing OOTA cycles from volatile atomics.
+It has also presented a substantial list of optimizations that avoid
+OOTA cycles on non-volatile atomics.
 
@@
 	Again, note that most production C++ code uses compiler-based C++
-	implementations, which make their non-deterministic choices at
-	compile time.
+	implementations, which freeze their non-deterministic choices
+	into the binary at compile time.
 \item	Section~\ref{sec:Semantic Dependencies And Optimization}'s
@@
 
-There are a number of possible reactions to this situation:
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-invent-int-load.litmus @@
+\caption{Inventing Non-Boolean Atomic Loads}
+\label{lst:Unused Inventing Non-Boolean Atomic Loads}
+\end{listing}
+
+This is not restricted to booleans, as can be seen in
+Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}.
+This example is more straightforward in that it does not make quite so
+creative use of the multiple values loaded.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/inc.litmus @@
+\caption{Only Atomic Increment}
+\label{lst:Only Atomic Increment}
+\end{listing}
+
+But suppose that we drop the single-thread-analysis constraint for
+the moment, and suppose that the C++ implementation is able to prove
+that the only modification of a given atomic variable is to atomically
+increment it, as is the case for \co{y} in
+Listing~\ref{lst:Only Atomic Increment}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/inc-range.litmus @@
+\caption{Only Atomic Increment, Optimized}
+\label{lst:Only Atomic Increment, Optimized}
+\end{listing}
+
+Is a C++ implementation within its rights to behave as if the code
+shown in
+Listing~\ref{lst:Only Atomic Increment}
+was instead as shown in
+Listing~\ref{lst:Only Atomic Increment, Optimized}?
+One argument in favor is that the value of \co{y} must have
+passed through the value 42 in the case where \co{r0} and \co{r1}
+bracket that value.
+Arguments against might cite the added overhead of the invented load
+in Listing~\ref{lst:Only Atomic Increment, Optimized}, or that the
+possibility that the check for the value 42 is intended for statistical
+sampling.
+
+There are a number of possible reactions to these situations:
 
@@
 	adjacent to each other.
-\item	Any time an impolementation might be tempted to invent uses from
-	values from non-volatile relaxed atomic loads, the loads should
-	instead be fused into a single load.
+\item	Any time an implementation might be tempted to invent uses from
+	values from non-volatile relaxed atomic loads from the same
+	object, the loads should instead be fused into a single load.
 	This reaction might be attractive to those who carefully consider
 	the pointlessness of keeping two loads that return the same value
-	on the one hand or the cache-miss overhead involved should two
-	closely spaced loads from the same object return different values
-	on the other.
+	on the one hand or the cache-miss overhead incurred when closely
+	spaced loads return different values on the other.
 \item	What other interesting situations might come to light?
@@
 
-Inserting \co{std::observable()} at line~12
-as shown in Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
-avoids confusing the compiler into assuming that the value of \co{local}
-must necessarily be zero.
+The \co{std::observable()} at line~12 of
+Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
+avoids confusing the compiler, instead forcing it to assume that the
+value of \co{local} might not be zero.
 
@@
 
-By the same reasoning, Goldblatt's \co{misbhave} example that
+By the same reasoning, Goldblatt's \co{misbehave} example that
 also involves unreachability can also be addressed through insertion
@@
 The tool then hands this expression to a SAT solver, and if this solver
-finds a satisfying set of values, the tool uses that set to provide
+finds a satisfying set of values, the tool uses those values provide
 the execution path to the bug.
@@
 
+\begin{listing}[tbp]
+@@ DisplayLitmus cbmc/sdep-quadratic.c @@
+\caption{Full Code for cbmc-Based Evaluation of sdep}
+\label{lst:Full Code for cbmc-Based Evaluation of sdep}
+\end{listing}
+
 However, $x^2 + 2x + 2$ has a minimum value of 1 at $x=-1$, so
 forcing \co{limit} to (say) zero should eliminate this static
-dependency:
+dependency:\footnote{
+	See Listing~\ref{lst:Full Code for cbmc-Based Evaluation of sdep}
+	for the full \path{cbmc/sdep-quadratic.c} program.}
 
@@
 store to that of Thread~3.
+This situation can occur due to the initial placement and subsequent
+movement of cache lines.
 
@@
 That is, fre links can and do go backwards in time.
-This corresponds to the fre link between Thread~3 an Thread~2 depicted in
+This corresponds to the fre link between Thread~3 and Thread~2 depicted in
 Figure~\ref{fig:IPC Diagram for coe, fre, and rfe}.
@@
 In the case of coe, the two stores take time to propagate across the
-system, and their ordering can and is decided after the fact.
+system, and their ordering can be and often is decided after the fact.
 In the case of fre, the store again takes time to propagate across
@@
 the arithmetic proporties of this type.
-In theory, the standard could specify the signedness of \co{char},
-but presumably the variety of existing practice in the 1980s prevented
-this from being done.
+In theory, the standard could specify the signedness of \co{char}, but the
+variety of existing practice in the 1980s prevented this.
+Plus machine instructions were much more expensive back then, so the
+signedness of \co{char} was much more of a performance concern than
+it is toda than
+it is todayy.
 
@@
 % @@@ TODO:
+% Create nice bell graph for coe, or at least try!
 % Construct interleaved OOTA examples (see sheets of paper).

### 2024-01-07 ootavacuous.sh

@@ -30,7 +30,7 @@ BEGIN {
 	print bs "scriptsize";
 	print bs "begin{verbatim}";
 	print "___EOF___";
-	print "grep -v " sq "^[( ]" bs "*" sq " < " path " | c2latex.sh";
+	print "grep -v " sq "^[(/ ]" bs "*" sq " < " path " | c2latex.sh";
 	print "cat << " sq "___EOF___" sq;
 	print bs "end{verbatim}";
 	print "}";
@@ -43,7 +43,7 @@ BEGIN {
 	print bs "scriptsize";
 	print bs "begin{verbatim}";
 	print "___EOF___";
-	print "grep -v " sq "^[( ]" bs "*" sq " < " path " | c2latex.sh";
+	print "grep -v " sq "^[(/ ]" bs "*" sq " < " path " | c2latex.sh";
 	print "cat << " sq "___EOF___" sq;
 	print bs "end{verbatim}";
 	print "";

### 2024-01-08 ootavacuous.tex
Invented loads and uses for loads and statistical sampling
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Update conclusion per Alan Stern January 7 2024 feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-This paper has added temporal reasoning to OOTA analysis of the C++
-memory model.
-This work shows that correct real-world C++ implementations are inherently
-incapable of producing OOTA cycles from volatile atomics.
-It has also presented a substantial list of optimizations that avoid
-OOTA cycles on non-volatile atomics.
+This paper has refined the definition of ``OOTA cycle'' from
+\co{acyclic(sdep; rf)} to \co{acyclic(sdep; rfe)}, which enables the
+application of temporal reasoning to OOTA analysis of the C++ memory
+model.
+This reasoning shows that correct real-world C++ implementations relying
+on single-threaded analyses are inherently incapable of producing OOTA
+cycles from UB-free programs using volatile atomics.
+
+C++ implementations that additionally refrain from inventing atomic loads
+and that also ensure that each atomic load that is executed corresponds
+to a specific load in the abstract machine are incapable of producing
+OOTA cycles from UB-free programs using non-volatile as well as volatile
+atomics.
 
@@
 	spaced loads return different values on the other.
+\item	Implementations that invent non-volatile atomic loads or invent
+	new uses for non-volatile atomic loads can adversely affect
+	observed behavior, for example, by introducing error into code
+	attempting to do statistical sampling.
 \item	What other interesting situations might come to light?

### 2024-01-08 ootavacuous.tex
Upgrade title for January 8, 2024
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Update load/store fusing and load/store consolidation
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{document}
-\title{D3064R0: Making OOTA Execution on Real Implementations be Vacuous}
+\title{D3064R0: Is OOTA Execution Vacuous for Real-World C++ Implementations?}
 
@@
 	along with any computations based on their return values.
-\item	Given a pair of relaxed loads from the same object that can
-	be reordered to be adjacent to each other, one of the pair can
-	be omitted with uses of the omitted load instead using the value
-	returned by the surviving load.
+\item	Given a pair of loads from the same object that can be
+	reordered to be adjacent to each other, if one of the pair is a
+	relaxed load, it can be omitted, with uses of the omitted load
+	instead using the value returned by the surviving load.
 	This is termed \emph{load fusing}.
-\item	Given a pair of relaxed stores to the same object that can
-	be reordered to be adjacent to each other, the first of the pair
-	can be omitted.
+\item	Given a pair of stores to the same object that can be reordered
+	to be adjacent to each other, if the first of the pair is a
+	relaxed store, it can be omitted.
 	This is termed \emph{store fusing}.
@@
 	from both objects using the first load's memory order.
+	This is termed \emph{load consolidation}.
 \item	Suppose the source code contains a pair of non-volatile atomic
@@
 	to both objects using the last store's memory order.
+	This is termed \emph{store consolidation}.
 \item	If an expression can be proven to always result in a constant,

### 2024-01-09 ootavacuous.tex
Apply Alan Stern feedback on abstract and conclusion
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	These constraints prevent OOTA cycles containing only volatile
-	atomics from occurring in correctly constructed C++ programs
-	running on correctly constructed C++ implementations in our
-	current universe.
+	atomics from occurring in UB-free executions of C++ programs
+	running on correctly constructed C++ single-thread analysis
+	implementations in our current universe.
 	Given current optimizations, these constraints also prevent
 	OOTA cycles involving non-volatile atomics.
-
-	In this, OOTA behaviors are similar to the issues surrounding
-	undecidability and its real-world counterpart, high complexity.
-	So similar, in fact, that not only the same strategies apply
-	to OOTA, but also the same \emph{code} applies, both for C++
-	implementations and for associated tooling.
 \end{abstract}
@@
 on single-threaded analyses are inherently incapable of producing OOTA
-cycles from UB-free programs using volatile atomics.
+cycles in UB-free executions of programs using volatile atomics.
 
@@
 to a specific load in the abstract machine are incapable of producing
-OOTA cycles from UB-free programs using non-volatile as well as volatile
-atomics.
+OOTA cycles in UB-free executions of programs using non-volatile as well
+as volatile atomics.
 

### 2024-01-10 ootavacuous.tex
January 10, 2024 update to paper title
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Add appendix and links for non-trivial semantic dependencies
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{document}
-\title{D3064R0: Is OOTA Execution Vacuous for Real-World C++ Implementations?}
+\title{D3064R0: Bringing the Laws of Physics to Bear on the OOTA Problem}
 
@@
 single-variable perturbations.
+See Appendix~\ref{app:Non-Trivial Semantic Dependencies}
+for examples and additional discussion.
 
@@
 Note also that many users would consider it to be an extremely unfriendly
-act for the C++ implemenation to combine loads or stores to adjacent
+act for a C++ implementation to combine loads or stores to adjacent
 objects where each object \co{is_lock_free()}, but the combination is not.
@@
 the memory model, observable behavior, and the as-if rule cannot form
-OOTA cycles from volatile atomics when running on real systems in this
-universe.
-It has also presented a substantial list of optimizations that avoid
-creation of OOTA cycles from non-volatile atomics.
+OOTA cycles from volatile atomics when running UB-free programs on real
+systems in this universe.
+It has also presented a substantial list of optimizations that can be
+applied to non-volatile atomics without the risk of creating OOTA cycles.
 
@@
 \co{acyclic(sdep; rf)} to \co{acyclic(sdep; rfe)}, which enables the
-application of temporal reasoning to OOTA analysis of the C++ memory
-model.
+application the laws of physics (and thus of temporal reasoning) to OOTA
+analysis of the C++ memory model.
 This reasoning shows that correct real-world C++ implementations relying
@@
 as volatile atomics.
+
+This paper also uncovers some shortcomings in typical definitions of
+``semantic dependency'', pointing out examples where a single semantic
+dependency extends from multiple loads to a store, but not from any one
+of those loads to that same store.
+We therefore propose that the definition of semantic dependency be
+extended accordingly.
 
@@
 The manual analysis uses the informal rule that a semantic dependency
-exists between a set of relaxed atomic loads and a store on any execution
-where different values loaded would change the address stored to, the
-value stored, or the whether or not that store would be executed in
-the first place, with the proviso that stores of the same value to the
-same address having the same ordering properties are fungible.
+exists between a set of relaxed atomic loads\footnote{
+	Please see
+	Appendix~\ref{app:Non-Trivial Semantic Dependencies}
+	for examples showing why this is a set of loads rather than a
+	single load.}
+and a store on any execution where different values loaded would change
+the address stored to, the value stored, or the whether or not that store
+would be executed in the first place, with the proviso that stores of
+the same value to the same address having the same ordering properties
+are fungible.
 
@@
 \end{enumerate}
+
+\subsubsection{Non-Trivial Semantic Dependencies}
+\label{app:Non-Trivial Semantic Dependencies}
+
+Most of the examples in this paper involve simple semantic dependencies
+connecting a single load to a single store.
+One exception appears in
+Appendix~\ref{app:Non-LB Litmus Tests},
+but this section will present a more difficult example.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult-0.litmus @@
+\caption{OOTA Multiplication Example, Initial Value Zero}
+\label{lst:OOTA Multiplication Example, Initial Value Zero}
+\end{listing}
+
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+shows an example that builds on the multiplication-by-zero
+discussion in
+Section~\ref{sec:Semantic Dependencies Are Per Execution}.
+Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
+\co{x} and \co{y} on lines~9 and~10, respectively.
+Similarly, lines~16 and~17 store to \co{x} and \co{y}, respectively,
+the value that line~15 loads from \co{z}.
+Because all three variables are initialized to zero, changes to the value
+loaded from \co{x} cannot affect the value stored by line~11 to \co{z}
+unless there is also a change loaded from \co{y} and vice versa.
+However, if both \co{x} and \co{y} change, the value stored to \co{z}
+will change.
+
+Therefore, as noted in
+Section~\ref{sec:Semantic Dependencies Are Per Execution},
+there is no semantic dependency from the load from \co{x} to the
+store to \co{z}, nor is there a semantic dependency from the load
+from \co{y} to the store to \co{z}.
+However, there \emph{is} a semantic dependency from the \emph{combination}
+of the loads from \co{x} and \co{y} to the store to \co{z}.
+
+This situation poses a challenge to current work on semantic dependencies,
+which typically restricts such dependencies to a single load and store.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult-1.litmus @@
+\caption{OOTA Multiplication Example, Initial Value 1}
+\label{lst:OOTA Multiplication Example, Initial Value of One}
+\end{listing}
+
+In contrast, consider
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value of One},
+which differs from
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+only in the initial values (1 instead of 0) and the \co{exists}
+clause (0 instead of 1).
+
+Because the values loaded from \co{x} and \co{y} on lines~9 and~10
+are non-zero, changes to either \co{x} or \co{y} will affect the
+value stored to \co{z} by line~11.
+There is therefore a separate semantic dependency from the load from
+\co{x} to the store to \co{z} and from the load from \co{y} to the store
+to \co{z}, which matches typical definitions of semantic dependency.
+Yet
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value of One}
+and
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+can be thought of as being merely two different executions of the same
+program.
+
+One might ask why semantic dependencies can be non-trivial when hardware
+address, control, and data dependencies are much more straightforward.
+Answers include:
+(1)~Respecting dependencies even when they reduce to a constant,
+(2)~Less-aggressive optimizations, and
+(3)~Lack of undefined behavior.
+
+This situation further underscores the per-execution nature of semantic
+dependencies as well as the need for an improved definition of ``semantic
+dependency''.
+We propose expanding the traditional definition to include dependencies
+extending from groups of loads to a single store.
 

### 2024-01-10 ootavacuous.tex
Add appendix showing UB-induced OOTA cycles
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \co{if} statement on line~13.
-This would mean that the store to \co{z} was always the constant one,
+This would mean that the store to \co{z} was always the constant 1,
 which would permit the store to be ordered before the load from \co{y},
@@
 extending from groups of loads to a single store.
+
+\subsubsection{OOTA From Undefined-Behavior Optimizations}
+\label{app:OOTA From Undefined-Behavior Optimizations}
+
+Many C++ implementations carry out optimizations based on the assumption
+that UB cannot happen, and these optimizations can result in OOTA cycles,
+with a simple example shown in
+Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-div-ub.litmus @@
+\caption{OOTA Due To Divide-By-Zero UB}
+\label{lst:OOTA Due To Divide-By-Zero UB}
+\end{listing}
+
+On line~14 of this listing, the value of \co{r2} must be either zero
+or one due to the comparison on line~13.
+If it is the value one, then there is a divide-by-zero error, which
+enables the C++ implementation to assume that the value of \co{r1} is
+always zero.
+This in turn allows the implementation to behave as if the expression
+stored to \co{y} on line~14 was the constant 1, at which point there
+is no semantic dependency from the load from \co{x} on line~13 to the
+store to \co{y} on line~14.
+
+The implementation is then within its rights to act as if lines~13 and~14
+were interchanged.
+At which point, the order of execution could be as follows:
+
+\begin{enumerate}
+\item	Line~14 stores the constant 1 to \co{y}.
+\item	Line~8 loads the constant 1 from \co{y}, setting \co{r1} to 1.
+\item	Line~9 stores the constant 1 to \co{x}.
+\item	Line~8 loads the constant 1 from \co{x}, setting \co{r2} to 1.
+\end{enumerate}
+
+Please note that the OOTA-like cycle called out in the \co{exists}
+clause can happen even though the program never actually executes any
+code that would have resulted in division-by-zero UB.
+It is tempting to consider using the techniques described in
+Appendix~\ref{app:Aside on Undefined Behavior}
+that prevent back-propagation of UB, but these are ineffective in this
+case because the C++ implementation can detect and exploit UB when
+considering line~14 in isolation.
+Once the constant 1 is substituted for line~14's expression, the
+hardware is free to reorder lines~13 and~14 even if a compiler-based
+C++ implementation refrains from doing so.
+
+Therefore, this paper focuses on UB-free programs, rare though these
+beasts might be among large-scale C++ programs running in production.
 

### 2024-01-10 ootavacuous.tex
Add "conjecture" weasel words  ;-)
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	implementers, and this paper summarizes some of the challenges.
-	But at the same time, there are no known instances of OOTA
-	behavior in real C++ implementations.
-
-	This paper refactors the commonly used condition for detecting
+	But at the same time, we know of no instances of OOTA behavior
+	in real C++ implementations.
+
+	This paper refactors the commonly used condition for classifying
 	OOTA behavior, which permits bringing to bear constraints from
-	long-standing laws of physics and on hardware systems.
-	These constraints prevent OOTA cycles containing only volatile
-	atomics from occurring in UB-free executions of C++ programs
-	running on correctly constructed C++ single-thread analysis
-	implementations in our current universe.
-	Given current optimizations, these constraints also prevent
-	OOTA cycles involving non-volatile atomics.
+	long-standing laws of physics and from hardware systems.
+	We present informal arguments that these constraints prevent
+	OOTA cycles containing only volatile atomics from occurring
+	in UB-free C++ programs running on correctly constructed C++
+	implementations that are restricted to single-threaded analysis
+	in our current universe.
+	We further conjecture that, given current optimizations, these
+	constraints also prevent OOTA cycles involving non-volatile
+	atomics.
 \end{abstract}
@@
 on single-threaded analyses are inherently incapable of producing OOTA
-cycles in UB-free executions of programs using volatile atomics.
+cycles in UB-free programs using volatile atomics.
 
@@
 to a specific load in the abstract machine are incapable of producing
-OOTA cycles in UB-free executions of programs using non-volatile as well
+OOTA cycles in UB-free programs using non-volatile as well
 as volatile atomics.
@@
 C++ implementation refrains from doing so.
+Note that this OOTA-like cycle can appear even if all of the atomic
+operations in
+Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}
+were converted to be volatile.
 

### 2024-01-11 ootavacuous.tex
Add a table of contents
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Add more elaborate non-trivial sdep example
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \end{abstract}
+
+\pagebreak
+
+\tableofcontents
 
@@
 program.
+
+Much more elaborate examples can be constructed, for example,
+involving sums of products of values loaded.
+Reverting back to short-form code fragments, consider the following:
+
+\begin{quote}
+	\co{z = w * x + w * y + w;}
+\end{quote}
+
+This has the following semantic dependencies when all of \co{w},
+\co{x}, and \co{y} are zero:
+
+\begin{enumerate}
+\item	From the combination of \co{w} and \co{x} to \co{z}.
+\item	From the combination of \co{w} and \co{y} to \co{z}.
+\item	From \co{w} to \co{z}.
+\end{enumerate}
+
+Despite appearances, these are not redundant.
+The last semantic dependency does not imply the first, because given
+that \co{w} becomes non-zero, a change in \co{x} further changes
+the value stored to \co{z}.
+Similar reasoning shows that the last semantic dependency does not
+imply the second.
+Much more elaborate examples can easily be constructed.
+
+This raises the question of how to determine the set of semantic
+dependencies in a given execution of a given fragment of code.
+The short answer is that when running on a real-world system, 
+any C++ program is a finite-state machine, which, unlike Turing-complete
+systems, can be analyzed.
+% @@@ reference to cbmc appendix.
 

### 2024-01-11 ootavacuous.tex
Add per-code semantic dependencies as union over executions
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 % 	in the preceding one.}
+
+But what if you really need a definition of semantic dependency
+for a given code fragment that covers all executions?
+One approach is to take the union of semantic dependencies across
+all executions, which might be useful when carrying out static analysis.
+On the other hand, many C++ implementations carry out specialization
+optimizations, which consider groups of executions.
+For example, an implementation might special-case a \co{NULL} pointer when
+considering several inline functions that take early exits in that case.
+Tools working with such implementations might therefore choose to focus
+on executions corresponding to a given specialization when generating
+semantic dependencies for code corresponding to that specialization.
 

### 2024-01-11 ootavacuous.tex
Add section on multithreaded optimizations
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+\subsection{But What About Multi-Thread Optimizations?}
+\label{sec:But What About Multi-Thread Optimizations?}
+
+Although this paper mostly ignores multi-threaded implementations,
+JMM Causality Test Cases 1, 8, 9, and 9a discuss inter-thread
+analysis or constraints on thread scheduling.
+This topic is therefore worth a short discussion.
+
+In order to set context, it is important to note that an oracular C++
+implementation could in theory read a program and its input, and then
+perform a sequence of observable behaviors consistent with that program
+running on that input without actually executing that program's code.
+This admittedly extreme example shows that it is not unreasonable to
+impose some real-world constraints on a real C++ implementation, and
+limiting analysis to a single thread is one of the constraints that we
+have chosen.
+
+At the same time, it is worth calling out some optimizations based on
+multi-threaded analysis that can be introduced without risk of
+introducing OOTA cycles.
+
+First, given a variable that the implementation can prove is always equal
+to some constant, the implementation may act as if the program's uses of
+that variable's value instead used that constant.
+
+Second, given a pair of variables that the implementation can prove always
+have a given algebraic relation to each other, the implementation can
+act as if that program's uses of one of the variable's values instead
+used the corresponding function of the other variable's value.
+However, if the values of those variables are subject to change, there
+must be some form of synchronization in effect to prevent accesses
+to those variables' values during times when a given modification has
+changed the value of one of the variables but not yet that of the other.
+If there is no such synchronization, then this optimization cannot
+be applied.
+
+Third, given that the implementation can prove that one thread always
+terminates before another thread starts, and if one or another of the
+threads does nothing related to thread identity, the implementation
+may act as if these two threads had been fused into a single thread.
+
+Fourth and finally, if the implementation can prove that a given thread
+never executes, the implementation may act as if code specific to that
+thread had been omitted from the program.
+
+Although these last two examples might appear to be a bit utopian,
+the first two provide ample room for multi-threaded analysis.
+
 \subsection{Correct Implementations Cannot Form OOTA Cycles}
@@
 These tests might be two decades old, but they are still quite
-relevant.
+relevant.\footnote{
+	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
 

### 2024-01-11 ootavacuous.tex
Explicitly say how C++ accounts for some control dependencies
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 		data dependencies, and mandates control dependencies
-		only indirectly.
+		only indirectly (via the prohibition against introducing
+		data races).
 		Such dependencies nevertheless still exist.}
@@
 
-\subsection{But What About Multi-Thread Optimizations?}
-\label{sec:But What About Multi-Thread Optimizations?}
+\subsection{But What About Multi-Threaded Optimizations?}
+\label{sec:But What About Multi-Threaded Optimizations?}
 

### 2024-01-11 ootavacuous.tex
Add conditional-multiplication non-trivial sdep example
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult-0-cond.litmus @@
+\caption{OOTA Conditional Multiplication Example, Initial Value Zero}
+\label{lst:OOTA Conditional Multiplication Example, Initial Value Zero}
+\end{listing}
+
+For further evidence that there is no semantic dependency from \co{y}
+to \co{z} when \co{x} is zero, see the equivalent\footnote{
+	But equivalent only because the atomics are all non-volatile!}
+program shown in
+Listing~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero}.
+Here, there is not even a load from \co{y} when \co{x} is zero, so there
+cannot possibly be a semantic dependency involving this non-existent load.
+A similar transformation would load \co{x} only if \co{y} was non-zero,
+emphasizing the lack of any semantic dependency between \co{x} and \co{y}.
+
+In both
+Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+and~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero},
+an OOTA cycle would need to affect both the \co{x} and the \co{y} branches
+of this compound semantic dependency.
+
 This situation poses a challenge to current work on semantic dependencies,
@@
 \caption{OOTA Multiplication Example, Initial Value 1}
-\label{lst:OOTA Multiplication Example, Initial Value of One}
+\label{lst:OOTA Multiplication Example, Initial Value 1}
 \end{listing}
@@
 In contrast, consider
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value of One},
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1},
 which differs from
@@
 to \co{z}, which matches typical definitions of semantic dependency.
+Further
 Yet
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value of One}
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
 and
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-can be thought of as being merely two different executions of the same
-program.
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
+are merely two different executions of the same program.
 

### 2024-01-11 ootavacuous.tex
Add conjecture on tools and complex OOTA cycles
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+Note that in a C++ implementation that (erroneously) permitted an OOTA
+cycle to form, that cycle would involve both \co{r0} and \co{r1}.
+However, we conjecture that tools designed to find OOTA cycles could
+satisfy themselves with having found either of the two resulting cycles.
+
 \subsubsection{Multi-Cycle Litmus Tests}
@@
 of this compound semantic dependency.
+However, we again conjecture that tools designed to find OOTA cycles could
+satisfy themselves with having found either of the two resulting cycles.
 

### 2024-01-11 ootavacuous.tex
January 11, 2024 title change
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Yet another title change and sdep definition adjustment
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{document}
-\title{D3064R0: Bringing the Laws of Physics to Bear on the OOTA Problem}
+\title{D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem}
 
@@
 and a store on any execution where different values loaded would change
-the address stored to, the value stored, or the whether or not that store
-would be executed in the first place, with the proviso that stores of
-the same value to the same address having the same ordering properties
-are fungible.
+the address stored to, the value stored, or the number of times that
+store would be executed, with the proviso that stores of the same value
+to the same address having the same ordering properties are fungible.
 

### 2024-01-13 ootavacuous.tex
Tighten wording for non-trivial semantic dependencies
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 the value that line~15 loads from \co{z}.
-Because all three variables are initialized to zero, changes to the value
-loaded from \co{x} cannot affect the value stored by line~11 to \co{z}
-unless there is also a change loaded from \co{y} and vice versa.
-However, if both \co{x} and \co{y} change, the value stored to \co{z}
-will change.
+Because all three variables are initialized to zero, and because loads
+can be ordered before any store, there is an execution in which all
+three loads return the value zero, in which case all three stores will
+of course store the value zero.
+In any related execution where the value loaded from \co{y} remains zero,
+non-zero values loaded from \co{x} cannot affect the value stored by
+line~11 to \co{z}.
+Only executions in which differ in the values loaded from both \co{x} and
+\co{y} can cause the value stored to \co{z} to change to a non-zero value.
 
@@
 Section~\ref{sec:Semantic Dependencies Are Per Execution},
+in an execution where the loads from both \co{x} and \co{y} return zero,
 there is no semantic dependency from the load from \co{x} to the
@@
 For further evidence that there is no semantic dependency from \co{y}
-to \co{z} when \co{x} is zero, see the equivalent\footnote{
+to \co{z} in executions where the load from \co{x} is zero, see the
+equivalent\footnote{
 	But equivalent only because the atomics are all non-volatile!}
@@
 Listing~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero}.
-Here, there is not even a load from \co{y} when \co{x} is zero, so there
-cannot possibly be a semantic dependency involving this non-existent load.
-A similar transformation would load \co{x} only if \co{y} was non-zero,
-emphasizing the lack of any semantic dependency between \co{x} and \co{y}.
+Here, there is not even a load from \co{y} in executions where the
+load from \co{x} returns zero, so there cannot possibly be a semantic
+dependency involving this non-existent load.
+A similar transformation would load \co{x} only in executions where the
+load from \co{y} returned non-zero, emphasizing the lack of any semantic
+dependency between \co{x} and \co{z}.
 
@@
 and~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero},
-an OOTA cycle would need to affect both the \co{x} and the \co{y} branches
-of this compound semantic dependency.
+an OOTA cycle (such as the one indicated by their \co{exists} clauses)
+would need to affect both the \co{x} and the \co{y} components of this
+compound semantic dependency.
 However, we again conjecture that tools designed to find OOTA cycles could
@@
 
-Because the values loaded from \co{x} and \co{y} on lines~9 and~10
-are non-zero, changes to either \co{x} or \co{y} will affect the
+Starting with the intuitive execution where the values loaded from \co{x}
+and \co{y} on lines~9 and~10 are non-zero, executions that differ only
+in the value loaded from either \co{x} or \co{y} will still affect the
 value stored to \co{z} by line~11.
@@
 to \co{z}, which matches typical definitions of semantic dependency.
-Further
 Yet

### 2024-01-13 ootavacuous.tex
Add three-way multiplication OOTA litmus test
Also do more wordsmithing.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 line~11 to \co{z}.
-Only executions in which differ in the values loaded from both \co{x} and
-\co{y} can cause the value stored to \co{z} to change to a non-zero value.
+Only executions which load different (that is, non-zero) values from
+both \co{x} and \co{y} can cause the value stored to \co{z} to change
+from zero to a non-zero value.
 
@@
 of the loads from \co{x} and \co{y} to the store to \co{z}.
+For example, the (forbidden) execution that satisfies the \co{exists}
+clause, which would be an OOTA cycle.
 
@@
 A similar transformation would load \co{x} only in executions where the
-load from \co{y} returned non-zero, emphasizing the lack of any semantic
-dependency between \co{x} and \co{z}.
+load from \co{y} returned non-zero, which similarly emphasizes the lack
+of any semantic dependency between \co{x} and \co{z}.
 
@@
 satisfy themselves with having found either of the two resulting cycles.
-
 This situation poses a challenge to current work on semantic dependencies,
@@
 clause (0 instead of 1).
+Note that an execution satisfying the \co{exists} clause is again
+a (forbidden) OOTA cycle.
 
 Starting with the intuitive execution where the values loaded from \co{x}
-and \co{y} on lines~9 and~10 are non-zero, executions that differ only
-in the value loaded from either \co{x} or \co{y} will still affect the
-value stored to \co{z} by line~11.
-There is therefore a separate semantic dependency from the load from
-\co{x} to the store to \co{z} and from the load from \co{y} to the store
-to \co{z}, which matches typical definitions of semantic dependency.
+and \co{y} on lines~9 and~10 are the value 1 (that is, non-zero), executions
+that differ only in the value loaded from either \co{x} or \co{y} will
+still affect the value stored to \co{z} by line~11.
+There is therefore one semantic dependency from the load from \co{x}
+to the store to \co{z} and another separate semantic dependency from
+the load from \co{y} to the store to \co{z}, which matches typical
+definitions of semantic dependency.
 Yet
@@
 Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
-are merely two different executions of the same program.
+can be thought of as being merely two different executions of the same
+program.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult3-0.litmus @@
+\caption{OOTA Three-Factor Multiplication Example, Initial Value Zero}
+\label{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
+\end{listing}
+
+Because the only two solutions to $x^2 = x$ are the values zero and one,
+Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+and~\ref{lst:OOTA Multiplication Example, Initial Value 1}
+each have but one OOTA cycle, and an execution having non-zero non-one
+initial values could have at most two OOTA cycles, all of which are
+(again) forbidden by the C++ language.
+Listing~\ref{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
+addressses this limitation by multiplying the values obtained from three
+different loads.
+Because $x^3 = x$ has three solutions ($\left\{ -1, 0, 1 \right\}$),
+this execution could have up to three (yet again, forbidden) OOTA cycles.
 

### 2024-01-17 ootavacuous.tex
Fix signed/unsigned char example, add Gonzalo
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{quote}
-	\co{void foo(char c)} \\
+	\co{void foo(int x, char c)} \\
 	\co{\{} \\
-	\co{\ \ \ \ return c + 128;} \\
+	\co{\ \ \ \ return x * (c >= 0);} \\
 	\co{\}}
@@
 of the architects of a given CPU family.
-Therefore, the code above is undefined behavior on systems choosing
-signed \co{char} and having 8-bit bytes (so that there is no semantic
-dependency from the \co{foo()} function's argument to its return value),
-but perfectly well-defined on systems having unsigned \co{char}
-(so that there is a semantic dependency).
+Therefore, the code above has a a semantic dependency when run on systems
+choosing signed \co{char}, but not when run on systems choosing unsigned
+\co{char}.
 So score one for the implementers, albeit by conscious choice on the
@@
 Martin Uecker contributed valuable insights on backwards-propagating UB.
+Gonzalo Brito Gadeschi shared an alternative way of handling non-volatile
+atomic operations.
 Richard Grisenthwaite patiently explained the architectural constraints

### 2024-01-17 ootavacuous.tex
Fix side-effect issue
In expressions, multiple side effects is UB.  In parameter lists,
the order is unspecified.  So convert to parameter lists.

Reported-by: Stephen Dolan
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{quote}
-	\co{y = ++i / ++i;} \\
+	\co{int foo(int a, int b)} \\
+	\co{\{} \\
+	\co{\ \ \ \ return a / b;} \\
+	\co{\}} \\
+	\\
+	\co{y = foo(++i, ++i);} \\
 	\co{x = y * z;}
@@
 
-The standard does not specify the order of evaluation, so the resulting
-value of \co{y} might be zero ($\frac{1}{2}$ truncated) or two
+Because early C~implementers could not come to agreement, standard
+does not specify the order of evaluation of function arguments, so the
+resulting value of \co{y} might be zero ($\frac{1}{2}$ truncated) or two
 ($\frac{2}{1}$).

### 2024-01-17 ootavacuous.tex
Add "Load/Store Ordering" appendix
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+\section{Load/Store Ordering: Hardware View for Software Hackers}
+\label{sec:Load/Store Ordering: Hardware View for Software Hackers}
+
+% @@@ Get citations
+
+Both the ARM and PowerPC memory models permit reordering prior loads
+against subsequent stores, but actual tests on recent hardware fail to
+produce any evidence of any such reordering.
+Nevertheless, thus far neither ARM nor PowerPC hardware architects
+have been willing to strengthen their memory models so as to forbid
+such reordering.
+% @@@ Cite Gristenthwaite
+
+This appendix offers some possible reasons for this odd juxtaposition
+of negative test results and flat refusal.
+
+One key point is that some hardware (including ARM and PowerPC) provides
+precise exceptions.
+For example, if a given load results in a segmentation violation
+exception, that exception will occur before any instructions following
+that load have committed.
+This means that a given load instruction's execution must have proceeded
+beyond the point where an exception might occur before any subsequent
+store can be permitted to commit, even if that store is completely
+unrelated to that load.
+For example, the later store cannot commit until all prior loads' address
+translations have completed successfully.
+
+However, if that load suffers a cache miss, address translation will have
+completed long before the load returns its value, meaning that a later
+store might well commit before the load completes.
+In fact, that later store might commit before the store that supplies the
+value returned by that load!
+Which explains why some older hardware systems really do reorder prior
+loads and later stores.
+
+The question then becomes ``Why would weakly ordered systems fail to
+reorder prior loads with later stores?''
+
+One possible explanation is ECC errors.
+
+If correctable ECC errors were fixed up in hardware, then only uncorrectable
+ECC errors would be directly visible to software.
+If the only possible reaction to an uncorrectable ECC error was to terminate
+the program suffering that ECC error, there would be no problem.
+
+However, some code has ways of handling even uncorrectable errors.
+Operating-system kernels encountering an ECC error might note that the
+affected data was being used by only one process, and might react by
+killing that process.
+Some user applications might be note that the affected data affected only
+a particular computation, and might react by restarting that computation.
+Both the kernel and the user application would need to continue execution,
+and might thus desire an exact exception.
+
+Systems that offload processing of correctable ECC errors to software
+also require exact exceptions.
+After all, the software is going to need to be able to correct the
+error and then fix up state as if the load had returned the fixed-up
+value.
+
+ECC errors are detected near the end of their corresponding load
+instructions' execution, so that any need for exact exceptions rules
+out the commiting of any subsequent stores before the load has fetched
+its value.
+
+So why are hardware architects reluctant to tighten their memory models
+to forbid reordering of earlier loads and later stores?
+
+If you would like an authoratative answer to this question, you should of
+course ask your friendly local hardware architect.
+In the meantime, here is some semi-informed speculation on this topic:
+
+\begin{enumerate}
+\item	Some hardware might choose to forego ECC, for example, in order to
+	reduce cost or to improve energy efficiency.
+\item	ECC error correction might still be done in hardware, avoiding
+	the need for precise software-visible exceptions.
+\item	Some systems might prefer to immediately shut down, forbidding
+	software from continuing after an uncorrectable ECC error,
+	perhaps due to safety considerations.
+\item	Someone might come up with a clever way of correcting ECC errors
+	in firmware, avoiding the need for precise software-visible
+	exceptions.
+\item	As late as early 2024, some GPUs have been observed reordering
+	earlier loads against later stores.
+	Other hardware architects might therefore feel the need to keep
+	this option open for their own systems.
+\end{enumerate}
+
+So although it is not unreasonable to continue asking hardware vendors
+to tighten their memory models so as to prohibit reordering of earlier
+loads and later stores, it also would not be too surprising for them to
+continue to refuse.
+
+\clearpage
+
 \section{Debates on Causality}

### 2024-01-18 ootavacuous.tex
Fix three-laws-of-physics typo s/earlier/later/
Reported-by: Dan Kelley <djk@meta.com>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 These three laws of physics mean that if one thread loads the value
-stored by some other thread, that load must have occurred earlier
+stored by some other thread, that load must have occurred later
 in global time than did the store.

### 2024-01-23 ootavacuous.tex
Add links to C and C++ proposals for side-effect definition
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 However, the above code nevertheless conforms to the standard.
+Note that side effects in expressions lacking sequence points is
+undefined behavior, however, there are proposals to define it in
+the C++ and C languages, allowing implementations to nevertheless reorder
+evaluation when permitted by the as-if rule.\footnote{
+	\url{https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2016/p0145r3.pdf}
+	\url{https://www.open-std.org/jtc1/sc22/wg14/www/docs/n3203.htm}}
 

### 2024-01-25 ootavacuous.tex
January 25th reorganization, part 1
Based on Alan Stern's feedback up through moving from (sdep U rf) to
(sdep U rfe).

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-This section provides a brief overview of the OOTA problem, followed
+@@@ This section provides a brief overview of the OOTA problem, followed
 by an equally brief summary of prior work in this area, and ending
@@
 
-OOTA occurs when a group of threads load from each others' stores,
-but where each thread's store depends on the value returned by that
-thread's load.
-A given out-of-thin-air value passes around the resulting cycle.
-Because there is a cycle, on a real system, at least one of the links
-must be atemporal, that is, either a load returns a value before
-that value is stored, or a store's address or value is computed before
-a prior load returns a value used in that computation.
+OOTA occurs when a group of threads load from each others' stores to
+each of a group of objects, but where each thread's store depends on
+the value returned by that thread's load.
+A given out-of-thin-air value passes around the resulting \emph{OOTA cycle}.
+
+\subsubsection{Simple Reordering}
+\label{sec:Simple Reordering}
+
+However, it is important to distinguish OOTA cycles from the simple
+reordering exemplified by
+Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
+Both the C++ implementation\footnote{
+	This paper will use ``C++ implementation'' both excluding the
+	CPU, as here, and including the CPU.
+	Explicit disambiguation is added when necessary.}
+and the CPU are within their rights to reorder lines~12 and~13, which
+can result in all of \co{x}, \co{y}, \co{r1}, and \co{r2} having the
+value 42 as follows:
+
+\begin{enumerate}
+\item   Line~13 stores 42 to \co{x}.
+\item   Line~6 loads 42 from \co{x} into \co{r1}.
+\item   Line~7 stores \co{r1}, and thus 42, to \co{y}.
+\item   Line~12 loads 42 from \co{y} to \co{r2}.
+\end{enumerate}
+
+Current C++ implementations can and do exhibit this simple-reordering
+behavior.
+
+\subsubsection{Simple OOTA Cycle}
+\label{sec:Simple OOTA Cycle}
+
+In contrast,
+Listing~\ref{lst:Simple OOTA}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
+shows a simple example where an OOTA cycle might result in all of \co{x},
+\co{y}, \co{r1}, and \co{r2} have final values of 42, despite the fact
+that there is nothing in the code or in the initial values that would
+support such an outcome:
+
+\begin{enumerate}
+\item   Line~6 loads from \co{x} into \co{r1}, somehow obtaining the
+	value 42, never mind the \co{x}'s initial value of zero.
+\item   Line~7 stores \co{r1}, and thus 42, to \co{y}.
+\item   Line~12 loads 42 from \co{y} to \co{r2}.
+\item   Line~13 stores 42 to \co{x}, justifying the value loaded by
+	line~6 after the fact.
+\end{enumerate}
+
+Fortunately, the C++ standard prohibits such OOTA cycles in 33.5.4p8
+(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
+
+\begin{quote}
+	Implementations should ensure that no “out-of-thin-air” values
+	are computed that circularly depend on their own computation.
+\end{quote}
+
+This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
+in using the term \emph{full C++} to denote the standard including the
+above prohibition, and the term \emph{strict C++} to denote a hypothetical
+standard that excluded this prohibition, but was otherwise identical
+to full C++.
+Unqualified \emph{C++} means full C++.
+
+The next section looks at how prior work has refined this prohibition.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 atomic<int> x(0);
+ 2 atomic<int> y(0);
+ 3
+ 4 void thread1()
+ 5 {
+ 6   unsigned long r1 = x.load(memory_order_relaxed);
+ 7   y.store(r1, memory_order_relaxed);
+ 8 }
+ 9
+10 void thread2()
+11 {
+12   unsigned long r2 = y.load(memory_order_relaxed);
+13   x.store(42, memory_order_relaxed);
+14 }
+\end{verbatim}
+\caption{Simple Reordering}
+\label{lst:Simple Reordering}
+\end{listing}
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 atomic<int> x(0);
+ 2 atomic<int> y(0);
+ 3
+ 4 void thread1()
+ 5 {
+ 6   unsigned long r1 = x.load(memory_order_relaxed);
+ 7   y.store(r1, memory_order_relaxed);
+ 8 }
+ 9
+10 void thread2()
+11 {
+12   unsigned long r2 = y.load(memory_order_relaxed);
+13   x.store(r2, memory_order_relaxed);
+14 }
+\end{verbatim}
+\caption{Simple OOTA}
+\label{lst:Simple OOTA}
+\end{listing}
+
+\subsection{Prior Work}
+\label{sec:Prior Work}
+
+All OOTA workers owe a debt to the foundational work in the infamous
+``Causality Test Cases'',\footnote{
+	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
+which also may be found in
+Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}.
+
+Some executable C++ memory models correctly flag at least some executions
+involving OOTA cycles~\cite{JadeAlglave2014HerdingCats}.\footnote{
+	Others cleverly avoid this issue by forbidding atomic
+	stores of non-constant values~\cite{MarkBatty2011cppmem}.}
+% @@@ Better herd7 C11 citation?
+However, because these models are atemporal, they cannot reject
+OOTA executions other than by flagging the OOTA value as arbitrary,
+which some in fact do in at least some cases.
+
+P0442R0 (``Out-of-Thin-Air Execution is Vacuous'')~\cite{PaulEMcKenney2016OOTA}
+provided a decision procedure for classifying behaviors as permitted
+misordering on the one hand or disallowed OOTA on the other, using
+a perturbation method based on the insight that all OOTA behaviors are
+fixed-point computations.
+
+Some workers recommend avoiding OOTA by ordering prior relaxed
+loads before subsequent relaxed
+stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
+but this requires real instructions be executed, consuming real
+time and real electrical power to solve a strictly theoretical
+problem.
+This might have been acceptable in the 1960s of my youth, but it is now
+the year 2024.
+
+Other workers recommend various procedures to identify and avoid OOTA
+cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
+but none of these have been looked upon favorably by C++ implementers.
+
+Goldblatt looked at interactions between OOTA cycles and
+UB~\cite{DavidGoldblatt2019NoElegantOOTAfix}, and
+Appendix~\ref{app:Aside on Undefined Behavior}
+analyzes the examples and notes ways to separate UB and OOTA-cycle
+concerns.
+The remainder of this document will focus on examples lacking UB.
+
+All this work focused on either identifying OOTA or on how C++
+implementations could avoid it.
+None of this work applied real-world temporal constraints to the OOTA
+problem.
+Which might explain why no known real-world C++ implementation results
+in OOTA executions.
+
+This paper therefore digs more deeply into OOTA cycles in the light
+of these real-world constraints.
+
+\subsection{Code-Analysis Tool}
+\label{sec:Code-Analysis Tool}
+
+This paper will use the \co{herd7} tool to analyze fragments of C++ code.
+This tool carries out the moral equivalent of full state-space searches
+of concurrent code fragments, and further has some understanding of OOTA.
 
@@
 Listing~\ref{lst:OOTA Cycle}
-shows an OOTA cycle.
+shows a code fragment that in strict C++ would have an OOTA cycle.
+This section describes this fragment, thus giving an overview of
+the \co{herd7} tool.
+
 The first line identifies it as a C-language litmus test and gives it
@@
 
+The earlier examples shown in
+Listings~\ref{lst:Simple Reordering}
+and~\ref{lst:Simple OOTA}
+can be recast as similar \co{herd7} litmus tests, which areas shown in
+Listing~\ref{lst:Simple Reordering as Litmus Test}
+on page~\pageref{lst:Simple Reordering as Litmus Test}
+and Listing~\ref{lst:Causality Test Case 4}
+on page~\pageref{lst:Causality Test Case 4}, respectively.
 Other OOTA-related litmus tests may be found in
 Appendix~\ref{app:OOTA-Related Litmus Tests}.
+
+\section{Semantic Dependencies, Executions, and OOTA}
+\label{sec:Semantic Dependencies; Executions; and OOTA}
+
+@@@
+
+\subsection{Semantic Dependencies}
+\label{sec:Semantic Dependencies}
+
+The standard's prohibition of OOTA is of course important, but those of us
+writing code in the real world must rely on actual C++ implementations.
+And in these implementations, this prohibition is in fact enforced by TSO
+ordering in strongly ordered systems and by data dependency ordering in
+weakly ordered systems.
+Real-world implementations cannot store something until after they load
+it, and this \emph{semantic dependency} results in compilers emitting
+instructions such that the hardware enforcement applies.\footnote{
+	The need to prohibit simple OOTA is one reason why compiler-based
+	value speculation optimizations are frowned upon.}
+Roughly speaking, there is semantic dependency from a given load to
+a given store when a change in the value loaded that can change the
+value stored.
+
+\subsection{OOTA: rf vs. rfe}
+\label{sec:OOTA: rf vs. rfe}
+
+However, the ordering implied by the semantic dependency is only one
+type of link in an OOTA cycle.
+The other link extends from a given store to the load that returns the
+value stored.
+It is tempting to make the intuitive argument that this store must precede
+this load in global time, and then to combine this with the notion that
+any real C++ implementation must consume global time when computing a
+semantic dependency.
+This combination suggests that such implementations cannot exhibit
+OOTA cycles.
+This combination has been formalized by defining an OOTA cycle as a cycle
+in sdep $\cup$ rf, where sdep is the set of semantic dependencies within
+each thread and rf is the set of store-to-load links, whether intra-thread
+(rfi) or between threads (rfe).\footnote{
+	See Appendix~\ref{app:Inter-Process Communications} for definitions
+	and properies of rf, rfe, and rfi.}
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-3-2-proc.litmus @@
+\caption{OOTA With an Intra-Process Store-Load Link}
+\label{lst:OOTA With an Intra-Process Store-Load Link}
+\end{listing}
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-3-2-proc-opt.litmus @@
+\caption{OOTA With an Intra-Process Store-Load Link, Optimized}
+\label{lst:OOTA With an Intra-Process Store-Load Link, Optimized}
+\end{listing}
+
+This is a fine definition, and is consistent with the words in the C++
+standard, except for code fragments exemplified by
+Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link}.
+This is an elaboration of
+Listing~\ref{lst:Simple OOTA}
+that adds \co{z} and adds lines~15 and~16 to \co{thread2()}.
+The problem is that the C++ implementation is within its rights to
+note that line~16 might well execute immediately after line~15, so that
+no other thread had a chance to modify \co{z}.
+Such an implementation might therefore behave as if the source code
+had instead been as shown in
+Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized},
+with line~16 optimized away in favor of line~17 storing the value
+in \co{r2} that was loaded from \co{y}.
+And given that the load from \co{z} no longer exists, it cannot possibly
+act as a temporal constraint.
+
+In order to avoid these backwards-in-time rfi links, we substitute rfe
+for rf, thus defining an OOTA cycle as a cycle in sdep $\cup$ rfe (or
+\co{acyclic(sdep ; rfe)} in \co{herd7} notation), where rfe is the set
+of store-to-load links, but excluding intra-thread rfi links, which are
+instead considered to be part of sdep.
+Although this does shunt additional complexity onto the term
+``semantic dependency'', it also enables us to cleanly separate
+the inter-thread and intra-thread portions of any given OOTA cycle.
+% \footnote{
+%	Thanks to Alan Stern for providing a litmus test that demonstrated
+%	that we had unwittingly (but productively!) shifted our definition
+%	to sdep $\cup$ rfe.}
+@@@ Please see
+Appendix~\ref{app:Informal Definition of Semantic Dependency}
+for more a more detailed (albeit quite informal) definition.
+
+@@@ Because there is a cycle, on a real system, at least one of the links
+must be atemporal, that is, either a load returns a value before
+that value is stored, or a store's address or value is computed before
+a prior load returns a value used in that computation.
 
@@
 This paper focuses instead on OOTA cycles.
-
-\subsection{Prior Work}
-\label{sec:Prior Work}
-
-All OOTA workers owe a debt to the foundational work in the infamous
-``Causality Test Cases'',\footnote{
-	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
-which also may be found in
-Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}.
-
-Some executable C++ memory models correctly flag at least some executions
-involving OOTA cycles~\cite{JadeAlglave2014HerdingCats}.\footnote{
-	Others cleverly avoid this issue by forbidding atomic
-	stores of non-constant values~\cite{MarkBatty2011cppmem}.}
-% @@@ Better herd7 C11 citation?
-However, because these models are atemporal, they cannot reject
-OOTA executions other than by flagging the OOTA value as arbitrary,
-which some in fact do in at least some cases.
-
-P0442R0 (``Out-of-Thin-Air Execution is Vacuous'')~\cite{PaulEMcKenney2016OOTA}
-provided a decision procedure for classifying behaviors as permitted
-misordering on the one hand or disallowed OOTA on the other, using
-a perturbation method based on the insight that all OOTA behaviors are
-fixed-point computations.
-
-Some workers recommend avoiding OOTA by ordering prior relaxed
-loads before subsequent relaxed
-stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
-but this requires real instructions be executed, consuming real
-time and real electrical power to solve a strictly theoretical
-problem.
-This might have been acceptable in the 1960s of my youth, but it is now
-the year 2024.
-
-Other workers recommend various procedures to identify and avoid OOTA
-cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
-but none of these have been looked upon favorably by C++ implementers.
-
-Goldblatt looked at interactions between OOTA cycles and
-UB~\cite{DavidGoldblatt2019NoElegantOOTAfix}, and
-Appendix~\ref{app:Aside on Undefined Behavior}
-analyzes the examples and notes ways to separate UB and OOTA-cycle
-concerns.
-The remainder of this document will focus on examples lacking UB.
-
-All this work focused on either identifying OOTA or on how C++
-implementations could avoid it.
-None of this work applied real-world temporal constraints to the OOTA
-problem.
-Which might explain why no known real-world C++ implementation results
-in OOTA executions.
-
-This paper therefore drops the question of how OOTA can be avoided and
-instead focuses on proving that OOTA cannot occur.
 
@@
 
-The C++ standard says this in 33.5.4p8
+@@@ The C++ standard says this in 33.5.4p8
 (\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
@@
 \end{quote}
-
-Others have defined an OOTA cycle as a cycle in sdep $\cup$ rf, where
-sdep is the set of semantic dependencies within each thread and rf is
-the set of store-to-load links, whether intra-thread (rfi) or between
-threads (rfe).\footnote{
-	See Appendix~\ref{app:Inter-Process Communications} for definitions
-	and properies of rf, rfe, and rfi.}
-This is a fine definition, and is consistent with the words in the C++
-standard, except that it simply shunts all the complexity onto the
-term ``semantic dependency''.
-
-We deviate slightly from this tradition by instead defining an OOTA cycle
-as a cycle in sdep $\cup$ rfe (or \co{acyclic(sdep ; rfe)} in \co{herd7}
-notation), where rfe is the set of store-to-load links, but excluding
-intra-thread rfi links, which are instead considered to be part of sdep.
-Although this does shunt even more complexity onto the term
-``semantic dependency'', it also enables us to bring to bear
-laws of physics and properties of hardware onto correct real-world
-C++ implementations.
-% \footnote{
-%	Thanks to Alan Stern for providing a litmus test that demonstrated
-%	that we had unwittingly (but productively!) shifted our definition
-%	to sdep $\cup$ rfe.}
-Please see
-Appendix~\ref{app:Informal Definition of Semantic Dependency}
-for more a more detailed (albeit quite informal) definition.
 
@@
 
+\subsubsection{Simple Reordering as Litmus Test}
+\label{app:Simple Reordering as Litmus Test}
+
+Listing~\ref{lst:Simple Reordering as Litmus Test}
+shows how
+Listing~\ref{lst:Simple Reordering}
+can be recast into a \co{herd7} litmus test.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/simple-reordering.litmus @@
+\caption{Simple Reordering as Litmus Test}
+\label{lst:Simple Reordering as Litmus Test}
+\end{listing}
+
 \subsubsection{Non-LB Litmus Tests}
@@
 
+Although the examples presented in
+Section~\ref{sec:OOTA: rf vs. rfe}
+demonstrate that rf can be atemporal, they demonstrate neither
+OOTA cycles nor OOTA-like behavior in full C++ implementations.
+This appendix provides more involved examples that can be optimized
+so as to demonstrate OOTA-like behavior.
+
 Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
 revises
-Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
+Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
 so that the value stored to \co{y} is forced to be at most the

### 2024-01-25 ootavacuous.tex
January 25 self-review #1
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-@@@ This section provides a brief overview of the OOTA problem, followed
+This section provides a brief overview of the OOTA problem, followed
 by an equally brief summary of prior work in this area, and ending
-with a quick review of a few of the challenges for general definitions of
-``OOTA cycle''.
+with a quick overview of the \co{herd7} tool that will be used to
+evaluate litmus tests.
 
@@
 
-However, it is important to distinguish OOTA cycles from the simple
-reordering exemplified by
+However, it is important to distinguish OOTA cycles from the OOTA-like
+behavior that can be caused by simple reordering.
+An example of simple reordering is shown in
 Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
@@
 but none of these have been looked upon favorably by C++ implementers.
+Some of these workers appear to have abandoned this effort, but as of
+early 2024, Mark Batty is persisting with modular relaxed dependencies.
 
@@
 Appendix~\ref{app:Aside on Undefined Behavior}
-analyzes the examples and notes ways to separate UB and OOTA-cycle
+analyzes his examples and notes ways to separate UB and OOTA-cycle
 concerns.
@@
 Listing~\ref{lst:OOTA Cycle}
-shows a code fragment that in strict C++ would have an OOTA cycle.
+shows a code fragment that in strict C++ would have an OOTA cycle (but
+that is of course prohibited in full C++).
 This section describes this fragment, thus giving an overview of
@@
 
-Finally, line~23 specifies an \co{exists} clause, which gives a
+Line~23 gives an example of a \co{locations} clause, which causes
+\co{herd7} to dump out the final value of \co{x} and \co{y}.
+
+Finally, line~24 specifies an \co{exists} clause, which gives a
 condition to check for the final value of the specified local
@@
 counterintuitive outcome where both processes load the value 42.
+Normally these states would include only those variables mentioned
+in the \co{exists} clause, but because of the \co{locations} clause
+on line~23, the values of \co{x} and \co{y} is also included, which
+can be helpful when debugging.
 As we will see, this outcome is not possible in real-world implementations
@@
 	The need to prohibit simple OOTA is one reason why compiler-based
-	value speculation optimizations are frowned upon.}
+	value speculation optimizations require checks on such
+	speculation, and these checks must be based on actual values
+	loaded.}
 Roughly speaking, there is semantic dependency from a given load to
@@
 
-However, the ordering implied by the semantic dependency is only one
-type of link in an OOTA cycle.
+However, semantic dependencies are only one type of link in an OOTA cycle.
 The other link extends from a given store to the load that returns the
@@
 This is a fine definition, and is consistent with the words in the C++
-standard, except for code fragments exemplified by
+standard, but has an intra-thread problem exemplified by
 Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link}.
@@
 The problem is that the C++ implementation is within its rights to
-note that line~16 might well execute immediately after line~15, so that
-no other thread had a chance to modify \co{z}.
+note that line~16 might well execute immediately after line~15, giving
+threads no chance to modify \co{z}.
 Such an implementation might therefore behave as if the source code
@@
 
-In order to avoid these backwards-in-time rfi links, we substitute rfe
+In order to avoid these rfi links, we substitute rfe
 for rf, thus defining an OOTA cycle as a cycle in sdep $\cup$ rfe (or
@@
 %	to sdep $\cup$ rfe.}
+
 @@@ Please see
@@
 for more a more detailed (albeit quite informal) definition.
+Maybe eventually have an appendix that lists all the definitions
+of both ``OOTA cycle'' and ``semantic dependency''.
 

### 2024-01-26 ootavacuous.tex
January 25 self-review #2
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+The standard's prohibition of OOTA is of course important, but those of us
+writing code in the real world must rely on actual C++ implementations.
+And in these implementations, this prohibition is in fact enforced by TSO
+ordering in strongly ordered systems and by data dependency ordering in
+weakly ordered systems.
+Real-world implementations cannot store something until after they load
+it, so there is a \emph{semantic dependency} from line~6 to line~7 and
+another from line~12 to line~13 in
+Listing~\ref{lst:Simple OOTA}.
+These semantic dependencies cause compilers to emit instructions such
+that the hardware enforcement applies.\footnote{
+	The need to prohibit simple OOTA is one reason why compiler-based
+	value speculation optimizations require checks on such
+	speculation, and these checks must be based on actual values
+	loaded.}
+
+Roughly speaking, \emph{there is semantic dependency from a given load
+to a given store when a change in the value loaded that can change the
+value stored.}
+
 This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
@@
 above prohibition, and the term \emph{strict C++} to denote a hypothetical
-standard that excluded this prohibition, but was otherwise identical
+standard that excludes this prohibition, but is otherwise identical
 to full C++.
@@
 
-@@@
-
-\subsection{Semantic Dependencies}
-\label{sec:Semantic Dependencies}
-
-The standard's prohibition of OOTA is of course important, but those of us
-writing code in the real world must rely on actual C++ implementations.
-And in these implementations, this prohibition is in fact enforced by TSO
-ordering in strongly ordered systems and by data dependency ordering in
-weakly ordered systems.
-Real-world implementations cannot store something until after they load
-it, and this \emph{semantic dependency} results in compilers emitting
-instructions such that the hardware enforcement applies.\footnote{
-	The need to prohibit simple OOTA is one reason why compiler-based
-	value speculation optimizations require checks on such
-	speculation, and these checks must be based on actual values
-	loaded.}
-Roughly speaking, there is semantic dependency from a given load to
-a given store when a change in the value loaded that can change the
-value stored.
+@@@ Fill in roadmap when sectioning is finalized.
 
@@
 
+\subsection{Where Do Loaded Values Come From?}
+\label{sec:Where Do Loaded Values Come From?}
+
+Relying on any rf link, whether the intra-thread rfi links or the
+inter-thread rfe links, raises the question of what restrictions
+might apply to C++ implementations' choice of value loaded.
+For non-volatile atomic loads, this is question is answered by
+the C++ standard in 6.9.2p14
+\co{[intro.races]}~\cite{ThomasKoeppe2023N4950}:
+
+\begin{quote}
+	The value of an atomic object M, as determined by evaluation B,
+	shall be the value stored by some side effect A that modifies M,
+	where B does not happen before A.
+\end{quote}
+
+This wording presumably includes the initial value of M as one of the
+possible side effects.
+This means that an implementation might determine that a given atomic
+object always has a given value, and such an implementation might
+skip the load in favor of that value.\footnote{
+	There is some work starting to forbid the implementation
+	from using this assumption, for example, by requiring that
+	the implementation assume that there is at least one thread
+	that it does not know about.}
+
+Please note that this does not apply to volatile atomic objects,
+for which the implementation is forbidden from assuming that the
+value loaded has any relationship whatsoever to any initialization
+or any values stored.
+
+\subsection{Semantic Dependencies and Source Code}
+\label{sec:Semantic Dependencies and Source Code}
+
+Some discussions of semantic dependencies assume that semantic
+dependencies are a strict function of the source code.
+Although there are some ways of making this work, many uses of
+semantic dependencies require them to be a function of particular
+executions.
+Consider for example:
+
+\begin{quote}
+	\co{x = y * z;}
+\end{quote}
+
+There is a semantic dependency from \co{y} to \co{x} if and only if the
+value of \co{z} is non-zero.
+Otherwise, in the absence of observable behavior (such as volatile
+accesses), the implementation is within its rights to discard this
+statement and act as if later accesses to \co{x} were instead the
+constant zero, at least up to the next assignment to \co{x}.
+
+Worse yet, suppose that both \co{y} and \co{z} are zero.
+Then there is no semantic dependency from either \co{y} or \co{z} to
+\co{x} unless both \co{y} and \co{z} change.
+Alternatively, when both \co{y} and \co{z} are zero, there is no semantic
+dependency from either \co{y} or \co{z} to \co{x}, but rather a semantic
+dependency from the combination of \co{y} \emph{and} \co{z} to \co{x}.
+This means that precise definitions of sdep cannot always rely on
+single-variable perturbations.
+See Appendix~\ref{app:Non-Trivial Semantic Dependencies}
+for examples and additional discussion.
+
+Consider also:
+
+\begin{quote}
+	\co{x = y + z;}
+\end{quote}
+
+There are semantic dependencies from \co{y} to \co{x} and from \co{z}
+to \co{x}.
+However, if the implementation knows that the value of \co{y} is always
+the negative of \co{z}, there is no semantic dependency and the
+implementation can again act as if later accesses to \co{x} were
+instead the constant zero, again, at least until the next assignment
+to \co{x}.
+
+Also, consider this:
+
+\begin{quote}
+	\co{r1 = x;} \\
+	\co{if (r1 > 0)} \\
+	\co{\ \ \ \ y = r1;} \\
+	\co{else} \\
+	\co{\ \ \ \ z = r1;}
+\end{quote}
+
+Here there is a semantic dependency from \co{x}, but in some executions
+it extends to \co{y} and in others to \co{z}.
+This is an example of the load affecting not the value or address of
+a given store, but rather whether or not that store is executed at all.
+
+But beware of this example:
+
+\begin{quote}
+	\co{r1 = x;} \\
+	\co{if (r1 > 0)} \\
+	\co{\ \ \ \ y = 42;} \\
+	\co{else} \\
+	\co{\ \ \ \ y = 42;}
+\end{quote}
+
+Because the stores executed on each branch of that "if" statement
+store identical values to identical addresses, they are for all intents
+and purposes the same store.
+Therefore, there is no semantic dependency~\cite{Howells2009membartxt}.
+
+A similar caution applies to loads:
+
+\begin{quote}
+	\co{r1 = x;} \\
+	\co{if (r1 > 42)} \\
+	\co{\ \ \ \ r2 = y;} \\
+	\co{else} \\
+	\co{\ \ \ \ r2 = y;} \\
+	\co{z = r2;}
+\end{quote}
+
+The two loads from \co{y} are fungible, and so a C++ implementation
+could behave as if this \co{if} statement had instead been a single load
+from \co{y}.
+
+These examples demonstrate a key point: Although in some cases sdep can
+be a strict function of the source code, \emph{in general, sdep must be
+defined on a per-execution basis}.% \footnote{
+% 	Thanks to Alan Stern for the examples in this section and
+% 	in the preceding one.}
+
+But what if you really need a definition of semantic dependency
+for a given code fragment that covers all executions?
+One approach is to take the union of semantic dependencies across
+all executions, so that if any execution has a semantic dependency,
+the source code fragment has a semantic dependency.
+This approach might be useful when carrying out static analysis.
+
+On the other hand, many C++ implementations carry out specialization
+optimizations, which consider groups of executions.
+For example, an implementation might special-case a \co{NULL} pointer when
+considering several inline functions that take early exits in that case.
+Tools working with such implementations might therefore choose to focus
+on executions corresponding to a given specialization when generating
+semantic dependencies for code corresponding to that specialization.
+
+Other tools might need to take the intersection of semantic dependencies
+across all executions, so that the source code fragment has a semantic
+dependency only if all executions have a semantic dependency.
+This approach might be useful when analyzing ordering properties
+of that code fragment.
+
+Nevertheless, in general, sdep must be defined on a per-execution basis.
+
+\subsection{@@@ This marks the end of the reworked material}
+
 @@@ Because there is a cycle, on a real system, at least one of the links
@@
 
-Note that C++ implementations are permitted to evaluate to a more
+@@@ Note that C++ implementations are permitted to evaluate to a more
 general notion of OOTA values in cases involving unspecified or undefined
@@
 
-\subsubsection{Semantic Dependencies Are Per Execution}
+\subsubsection{@@@ Semantic Dependencies Are Per Execution}
 \label{sec:Semantic Dependencies Are Per Execution}
-
-One aspect of this complexity is the fact that semantic dependencies
-are a function not strictly of the source code, but of a particular
-execution.
-Consider for example:
-
-\begin{quote}
-	\co{x = y * z;}
-\end{quote}
-
-There is a semantic dependency from \co{y} to \co{x} if and only if the
-value of \co{z} is non-zero.
-Otherwise, in the absence of observable behavior (such as volatile
-accesses), the implementation is within its rights to discard this
-statement and act as if later accesses to \co{x} were instead the
-constant zero, at least up to the next assignment to \co{x}.
-
-Worse yet, suppose that both \co{y} and \co{z} are zero.
-Then there is no semantic dependency from either \co{y} or \co{z} to
-\co{x} unless both \co{y} and \co{z} change.
-Alternatively, when both \co{y} and \co{z} are zero, there is no semantic
-dependency from either \co{y} or \co{z} to \co{x}, but rather a semantic
-dependency from the combination of \co{y} \emph{and} \co{z} to \co{x}.
-This means that precise definitions of sdep cannot always rely on
-single-variable perturbations.
-See Appendix~\ref{app:Non-Trivial Semantic Dependencies}
-for examples and additional discussion.
-
-Consider also:
-
-\begin{quote}
-	\co{x = y + z;}
-\end{quote}
-
-There are semantic dependencies from \co{y} to \co{x} and from \co{z}
-to \co{x}.
-However, if the implementation knows that the value of \co{y} is always
-the negative of \co{z}, there is no semantic dependency and the
-implementation can again act as if later accesses to \co{x} were
-instead the constant zero, again, at least until the next assignment
-to \co{x}.
-
-Also, consider this:
-
-\begin{quote}
-	\co{r1 = x;} \\
-	\co{if (r1 > 0)} \\
-	\co{\ \ \ \ y = r1;} \\
-	\co{else} \\
-	\co{\ \ \ \ z = r1;}
-\end{quote}
-
-Here there is a semantic dependency from \co{x}, but in some executions
-it extends to \co{y} and in others to \co{z}.
-This is an example of the load affecting not the value or address of
-a given store, but rather whether or not that store is executed at all.
-
-But beware of this example:
-
-\begin{quote}
-	\co{r1 = x;} \\
-	\co{if (r1 > 0)} \\
-	\co{\ \ \ \ y = 42;} \\
-	\co{else} \\
-	\co{\ \ \ \ y = 42;}
-\end{quote}
-
-Because the stores executed on each branch of that "if" statement
-store identical values to identical addresses, they are for all intents
-and purposes the same store.
-Therefore, there is no semantic dependency~\cite{Howells2009membartxt}.
-
-A similar caution applies to loads:
-
-\begin{quote}
-	\co{r1 = x;} \\
-	\co{if (r1 > 42)} \\
-	\co{\ \ \ \ r2 = y;} \\
-	\co{else} \\
-	\co{\ \ \ \ r2 = y;} \\
-	\co{z = r2;}
-\end{quote}
-
-The two loads from \co{y} are fungible, and so a C++ implementation
-could behave as if this \co{if} statement had instead been a single load
-from \co{y}.
-
-These examples demonstrate a key point: Although in some cases sdep can
-be a strict function of the source code, \emph{in general, sdep must be
-defined on a per-execution basis}.% \footnote{
-% 	Thanks to Alan Stern for the examples in this section and
-% 	in the preceding one.}
-
-But what if you really need a definition of semantic dependency
-for a given code fragment that covers all executions?
-One approach is to take the union of semantic dependencies across
-all executions, which might be useful when carrying out static analysis.
-On the other hand, many C++ implementations carry out specialization
-optimizations, which consider groups of executions.
-For example, an implementation might special-case a \co{NULL} pointer when
-considering several inline functions that take early exits in that case.
-Tools working with such implementations might therefore choose to focus
-on executions corresponding to a given specialization when generating
-semantic dependencies for code corresponding to that specialization.
 
@@
 \begin{enumerate}
-\item	Section~\ref{sec:Semantic Dependencies Are Per Execution}
+\item	Section~\ref{sec:Semantic Dependencies and Source Code}
 	is responsible for the phrase ``on any execution'' as well
@@
 discussion in
-Section~\ref{sec:Semantic Dependencies Are Per Execution}.
+Section~\ref{sec:Semantic Dependencies and Source Code}.
 Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
@@
 Therefore, as noted in
-Section~\ref{sec:Semantic Dependencies Are Per Execution},
+Section~\ref{sec:Semantic Dependencies and Source Code},
 in an execution where the loads from both \co{x} and \co{y} return zero,

### 2024-01-26 ootavacuous.tex
January 26th reorganization, part 1
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	long-standing laws of physics and from hardware systems.
-	We present informal arguments that these constraints prevent
-	OOTA cycles containing only volatile atomics from occurring
-	in UB-free C++ programs running on correctly constructed C++
-	implementations that are restricted to single-threaded analysis
-	in our current universe.
+	In this initial work, we focus on C++ implementations based
+	on compilers and computing hardware (including CPUs and GPGPUs).
+	We present informal arguments that these constraints and
+	limitations prevent OOTA cycles containing only volatile atomics
+	from occurring in undefined-behavior-free C++ programs running on
+	correctly constructed C++ compiler-based implementations that are
+	restricted to single-threaded analysis in our current universe.
 	We further conjecture that, given current optimizations, these
@@
 Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
-Both the C++ implementation\footnote{
-	This paper will use ``C++ implementation'' both excluding the
-	CPU, as here, and including the CPU.
-	Explicit disambiguation is added when necessary.}
-and the CPU are within their rights to reorder lines~12 and~13, which
-can result in all of \co{x}, \co{y}, \co{r1}, and \co{r2} having the
-value 42 as follows:
+Both the C++ compiler and the CPU are within their rights to reorder
+lines~12 and~13, which can result in all of \co{x}, \co{y}, \co{r1},
+and \co{r2} having the value 42 as follows:
 
@@
 \co{y}, \co{r1}, and \co{r2} have final values of 42, despite the fact
-that there is nothing in the code or in the initial values that would
-support such an outcome:
+that there is nothing in the initial values or the executable code that
+would support such an outcome:
 
@@
 
-Roughly speaking, \emph{there is semantic dependency from a given load
-to a given store when a change in the value loaded that can change the
-value stored.}
+Roughly speaking and all other things being equal, \emph{there is semantic
+dependency from a given load to a given store when a change in the value
+loaded that can change the value stored.}
 
@@
  5 {
- 6   unsigned long r1 = x.load(memory_order_relaxed);
+ 6   int r1 = x.load(memory_order_relaxed);
  7   y.store(r1, memory_order_relaxed);
@@
 11 {
-12   unsigned long r2 = y.load(memory_order_relaxed);
+12   int r2 = y.load(memory_order_relaxed);
 13   x.store(42, memory_order_relaxed);
@@
 11 {
-12   unsigned long r2 = y.load(memory_order_relaxed);
+12   int r2 = y.load(memory_order_relaxed);
 13   x.store(r2, memory_order_relaxed);
@@
 Goldblatt looked at interactions between OOTA cycles and
-UB~\cite{DavidGoldblatt2019NoElegantOOTAfix}, and
+undefined behavior (UB)~\cite{DavidGoldblatt2019NoElegantOOTAfix}, and
 Appendix~\ref{app:Aside on Undefined Behavior}
@@
 implementations could avoid it.
-None of this work applied real-world temporal constraints to the OOTA
-problem.
+None of this work applied real-world hardware ordering to the problem
+of avoiding OOTA cycles.
 Which might explain why no known real-world C++ implementation results
@@
 This tool carries out the moral equivalent of full state-space searches
-of concurrent code fragments, and further has some understanding of OOTA.
+of concurrent code fragments, and further will in some cases identify
+OOTA cycles.
 
@@
 values of the global shared variables \co{x} and \co{y} to zero.
-Lines~7-13 define the first process, \co{P0()}, and lines~15-21
-define the second process, \co{P1()}.
+Lines~7-13 define the first thread, \co{P0()}, and lines~15-21
+define the second thread, \co{P1()}.\footnote{
+	The ``P'' stands for ``process'', which is \co{herd7}'s name
+	for ``thread''.}
 The arguments to both \co{P0()} and \co{P1()} specify which of
-the global shared variables each process may access, in this case, \co{x}
+the global shared variables each thread may access, in this case, \co{x}
 and \co{y}.
-The body of each process specifies a limited subset of C++ code,
+The body of each thread specifies a limited subset of C++ code,
 which also happens to be a somewhat less limited subset of C code.
@@
 Lines~3 and~4 show the possible states, with line~4 showing the
-counterintuitive outcome where both processes load the value 42.
+counterintuitive outcome where both threads load the value 42.
 Normally these states would include only those variables mentioned
@@
 can be helpful when debugging.
-As we will see, this outcome is not possible in real-world implementations
-due to physical constraints of which \co{herd7} is (by design) unaware.
+% @@@ As we will see, this outcome is not possible in real-world
+% @@@ implementations due to physical constraints of which \co{herd7}
+% @@@ is (by design) unaware.
 Line~9 contains \co{Sometimes} (as opposed to \co{Never} or \co{Always}),
@@
 
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-3-2-proc.litmus @@
-\caption{OOTA With an Intra-Process Store-Load Link}
-\label{lst:OOTA With an Intra-Process Store-Load Link}
-\end{listing}
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-3-2-proc-opt.litmus @@
-\caption{OOTA With an Intra-Process Store-Load Link, Optimized}
-\label{lst:OOTA With an Intra-Process Store-Load Link, Optimized}
-\end{listing}
-
 This is a fine definition, and is consistent with the words in the C++
-standard, but has an intra-thread problem exemplified by
-Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link}.
-This is an elaboration of
+standard, but has an intra-thread problem exemplified by the following
+code:
+
+\begin{quote}
+\scriptsize
+\begin{verbatim}
+ 1   int r2 = atomic_load_explicit(&x, memory_order_relaxed);
+ 2   atomic_store_explicit(&y, r2, memory_order_relaxed);
+ 3   int r3 = atomic_load_explicit(&y, memory_order_relaxed);
+ 4   atomic_store_explicit(&z, r3, memory_order_relaxed);
+\end{verbatim}
+\end{quote}
+
+This is an elaboration of \co{thread2()} from
 Listing~\ref{lst:Simple OOTA}
-that adds \co{z} and adds lines~15 and~16 to \co{thread2()}.
+that adds \co{z} along with lines~2 and~3.
 The problem is that the C++ implementation is within its rights to
-note that line~16 might well execute immediately after line~15, giving
-threads no chance to modify \co{z}.
+note that line~3 might well execute immediately after line~2, giving
+threads no chance to modify \co{y}.
 Such an implementation might therefore behave as if the source code
-had instead been as shown in
-Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized},
-with line~16 optimized away in favor of line~17 storing the value
-in \co{r2} that was loaded from \co{y}.
+had instead been as follows:
+
+\begin{quote}
+\scriptsize
+\begin{verbatim}
+ 1   int r2 = atomic_load_explicit(&x, memory_order_relaxed);
+ 2   atomic_store_explicit(&y, r2, memory_order_relaxed);
+ 3   // int r3 = atomic_load_explicit(&y, memory_order_relaxed);
+ 4   atomic_store_explicit(&z, r2, memory_order_relaxed);
+\end{verbatim}
+\end{quote}
+
+Here line~3 has been optimized away in favor of line~4 storing the value
+in \co{r2} that was loaded from \co{y} by line~2.
 And given that the load from \co{z} no longer exists, it cannot possibly
@@
 
-\subsection{Where Do Loaded Values Come From?}
-\label{sec:Where Do Loaded Values Come From?}
-
-Relying on any rf link, whether the intra-thread rfi links or the
-inter-thread rfe links, raises the question of what restrictions
-might apply to C++ implementations' choice of value loaded.
-For non-volatile atomic loads, this is question is answered by
-the C++ standard in 6.9.2p14
-\co{[intro.races]}~\cite{ThomasKoeppe2023N4950}:
-
-\begin{quote}
-	The value of an atomic object M, as determined by evaluation B,
-	shall be the value stored by some side effect A that modifies M,
-	where B does not happen before A.
-\end{quote}
-
-This wording presumably includes the initial value of M as one of the
-possible side effects.
-This means that an implementation might determine that a given atomic
-object always has a given value, and such an implementation might
-skip the load in favor of that value.\footnote{
-	There is some work starting to forbid the implementation
-	from using this assumption, for example, by requiring that
-	the implementation assume that there is at least one thread
-	that it does not know about.}
-
-Please note that this does not apply to volatile atomic objects,
-for which the implementation is forbidden from assuming that the
-value loaded has any relationship whatsoever to any initialization
-or any values stored.
-
 \subsection{Semantic Dependencies and Source Code}
@@
 \begin{quote}
-	\co{x = y * z;}
+\begin{verbatim}
+x = y * z;
+\end{verbatim}
 \end{quote}
@@
 \begin{quote}
-	\co{x = y + z;}
+\begin{verbatim}
+x = y + z;
+\end{verbatim}
 \end{quote}
@@
 to \co{x}.
-However, if the implementation knows that the value of \co{y} is always
-the negative of \co{z}, there is no semantic dependency and the
-implementation can again act as if later accesses to \co{x} were
-instead the constant zero, again, at least until the next assignment
-to \co{x}.
+However, if the implementation knows that the value of \co{y} is
+always the negative of \co{z} there is no semantic dependency and the
+implementation can again act as if later accesses to \co{x} were instead
+the constant zero, again, at least until the next assignment to \co{x}.
+Please note that this leaves aside the question of how the compiler
+would know this if \co{y} and \co{z} are atomic objects that cannot be
+updated pairwise atomically and are subject to change at any time by
+other threads.
 
@@
 \begin{quote}
-	\co{r1 = x;} \\
-	\co{if (r1 > 0)} \\
-	\co{\ \ \ \ y = r1;} \\
-	\co{else} \\
-	\co{\ \ \ \ z = r1;}
+\begin{verbatim}
+r1 = x;
+if (r1 > 0)
+    y = r1;
+else
+    z = r1;
+\end{verbatim}
 \end{quote}
@@
 
+Which raises the question:
+What exactly is an execution?
+
+\section{@@@ This marks the end of the reworked material}
+
+\subsection{Difficulties in Rigorously Defining ``Semantic Dependency''}
+
 But beware of this example:
@@
 \begin{quote}
-	\co{r1 = x;} \\
-	\co{if (r1 > 0)} \\
-	\co{\ \ \ \ y = 42;} \\
-	\co{else} \\
-	\co{\ \ \ \ y = 42;}
+\begin{verbatim}
+r1 = x;
+if (r1 > 0)
+    y = 42;
+else
+    y = 42;
+\end{verbatim}
 \end{quote}
 
-Because the stores executed on each branch of that "if" statement
-store identical values to identical addresses, they are for all intents
-and purposes the same store.
-Therefore, there is no semantic dependency~\cite{Howells2009membartxt}.
+Because the stores executed on each branch of that "if" statement store
+identical values to identical addresses, one could equally well regard
+the two statements as performing two different stores or as performing
+(for all intents and purposes) a single store.
+Although software projects relying on ordering from load-to-store control
+dependencies and using a variety of compilers should take the single-store
+viewpoint~\cite{Howells2009membartxt},
+the fact remains that these two viewpoints lead to differing conclusions
+about whether there is a semantic dependency.
 
@@
 \begin{quote}
-	\co{r1 = x;} \\
-	\co{if (r1 > 42)} \\
-	\co{\ \ \ \ r2 = y;} \\
-	\co{else} \\
-	\co{\ \ \ \ r2 = y;} \\
-	\co{z = r2;}
+\begin{verbatim}
+r1 = x;
+if (r1 > 42)
+    r2 = y;
+else
+    r2 = y;
+z = r2;
+\end{verbatim}
 \end{quote}
@@
 
-\subsection{@@@ This marks the end of the reworked material}
-
-@@@ Because there is a cycle, on a real system, at least one of the links
-must be atemporal, that is, either a load returns a value before
-that value is stored, or a store's address or value is computed before
-a prior load returns a value used in that computation.
-
-@@@ Note that C++ implementations are permitted to evaluate to a more
-general notion of OOTA values in cases involving unspecified or undefined
-behavior (UB), such as use of uninitialized objects.
-However, these situations do not involve OOTA cycles, but rather user
-errors and other issues that can lead to unfriendly optimizations that
-in turn result in unpredictable output.
-This paper focuses instead on OOTA cycles.
+\subsection{Where Do Loaded Values Come From?}
+\label{sec:Where Do Loaded Values Come From?}
+
+Relying on any rf link, whether the intra-thread rfi links or the
+inter-thread rfe links, raises the question of what restrictions
+might apply to C++ implementations' choice of value loaded.
+For non-volatile atomic loads, this is question is answered by
+the C++ standard in 6.9.2p14
+\co{[intro.races]}~\cite{ThomasKoeppe2023N4950}:
+
+\begin{quote}
+	The value of an atomic object M, as determined by evaluation B,
+	shall be the value stored by some side effect A that modifies M,
+	where B does not happen before A.
+\end{quote}
+
+This wording presumably includes the initial value of M as one of the
+possible side effects.
+This means that an implementation might determine that a given atomic
+object always has a given value, and such an implementation might
+skip the load in favor of that value.\footnote{
+	There is some work starting to forbid the implementation
+	from using this assumption, for example, by requiring that
+	the implementation assume that there is at least one thread
+	that it does not know about.
+	\url{https://hackmd.io/@gonzalob/Bk_ufPZ6o/edit}}
+
+Please note that this does not apply to volatile atomic objects,
+for which the implementation is prohibited from assuming that the
+value loaded has any relationship whatsoever to any initialization
+or any values stored.
+This prohibition may come as a surprise to those carefully reading the
+C++ standard~\cite{ThomasKoeppe2023N4950}
+and finding only this wording constraining volatile accessess in
+4.1.2p6
+\co{[intro.abstract]}:
+
+\begin{quote}
+	Accesses through volatile glvalues are evaluated strictly
+	according to the rules of the abstract machine.
+\end{quote}
+
+Unfortunately, volatile semantics are mostly folklore passed by word
+of mouth among some compiler writers and some device-driver developers.
+The key point is that volatile accesses must interact with MMIO
+registers in such a way as to permit device drivers to be written.
+And until such time as the semantics of all MMIO registers of all
+devices have been formalized, published, and incorporated into
+C++ implementations,\footnote{
+	Given that some CPU vendors have started publishing formal
+	memory models, this happy day might actually arrive.
+	But we strongly recommend against holding your breath while
+	waiting for it.}
+these implementations cannot asssume that they can predict the value
+returned by a volatile load.
 
@@
 Sometimes ``causal cycle'' is used as if it was a definition, but
-without a clear decision process for what does and does not
+without a clear decision thread for what does and does not
 constitute a causal cycle.
@@
 \begin{quote}
-	\co{r1 = x;} \\
-	\co{// r1 = 0;} \\
-	\co{z = y * r1;} \\
-	\co{// z = y / r1;}
+\begin{verbatim}
+r1 = x;
+// r1 = 0;
+z = y * r1;
+// z = y / r1;
+\end{verbatim}
 \end{quote}
@@
 \begin{quote}
-	\co{r1 = x;} \\
-	\co{z = y * r1;} \\
-	\co{z = 42;}
+\begin{verbatim}
+r1 = x;
+z = y * r1;
+z = 42;
+\end{verbatim}
 \end{quote}
@@
 \begin{quote}
-	\co{int foo(int a, int b)} \\
-	\co{\{} \\
-	\co{\ \ \ \ return a / b;} \\
-	\co{\}} \\
-	\\
-	\co{y = foo(++i, ++i);} \\
-	\co{x = y * z;}
+\begin{verbatim}
+int foo(int a, int b)
+{
+   return a / b;
+}
+
+y = foo(++i, ++i);
+x = y * z;
+\end{verbatim}
 \end{quote}
@@
 \begin{quote}
-	\co{x = 2 * y;}
+\begin{verbatim}
+x = 2 * y;
+\end{verbatim}
 \end{quote}
@@
 \begin{quote}
-	\co{x = 2 * z;}
+\begin{verbatim}
+x = 2 * z;
+\end{verbatim}
 \end{quote}
@@
 \begin{quote}
-	\co{x = y * input_int();}
+\begin{verbatim}
+x = y * input_int();
+\end{verbatim}
 \end{quote}
@@
 \begin{quote}
-	\co{x = horribly_complex(y, z);}
+\begin{verbatim}
+x = horribly_complex(y, z);
+\end{verbatim}
 \end{quote}
@@
 \begin{quote}
-	\co{x = y & 0x1 ? do_odd(y) : do_even(y);}
+\begin{verbatim}
+x = y & 0x1 ? do_odd(y) : do_even(y);
+\end{verbatim}
 \end{quote}
@@
 \begin{quote}
-	\co{x = collatz_cycle_min(y);}
+\begin{verbatim}
+x = collatz_cycle_min(y);
+\end{verbatim}
 \end{quote}
@@
 \begin{quote}
-	\co{void foo(int x, char c)} \\
-	\co{\{} \\
-	\co{\ \ \ \ return x * (c >= 0);} \\
-	\co{\}}
+\begin{verbatim}
+void foo(int x, char c)
+{
+    return x * (c >= 0);
+}
+\end{verbatim}
 \end{quote}
@@
 
+As noted earlier, this initial work focuses on C++ implementations based
+on compilers and computing hardware (including CPUs and GPGPUs).
+
 C++ implementations execute C++ programs, and use a range of techniques.
@@
 
-This optimization process may, roughly speaking, be divided into two
+This optimization thread may, roughly speaking, be divided into two
 activities, analysis of the code and classification of dependencies.
@@
 	See Appendix~\ref{app:Inter-Process Communications} for more
-	information on these memory-reference interprocess communications
+	information on these memory-reference inter-process communications
 	links.}
@@
 The \co{herd7} tool does not find this cycle because it depends on
-scheduling \co{P2()} after the other two processes, and C++ is not
+scheduling \co{P2()} after the other two threads, and C++ is not
 constrained to do this.
@@
 
-Both processes have straightforward semantic dependencies between their
+Both threads have straightforward semantic dependencies between their
 respective relaxed atomic loads and relaxed atomic stores, so an OOTA
@@
 The \co{herd7} tool does not find this cycle because it depends on
-scheduling \co{P2()} before the other two processes, and C++ is not
+scheduling \co{P2()} before the other two threads, and C++ is not
 constrained to do this.
@@
 
-All processes have straightforward semantic dependencies between their
+All threads have straightforward semantic dependencies between their
 respective relaxed atomic loads and relaxed atomic stores, so an OOTA
@@
 
-All processes's loads' return values affect the value stored, so all
-processes have semantic dependencies between their respective relaxed
+All thread's loads' return values affect the value stored, so all
+threads have semantic dependencies between their respective relaxed
 atomic loads and relaxed atomic stores.
@@
 Appendix~\ref{app:Causality Test Case 4})
-from two processes to three processes.
+from two threads to three threads.
 The \co{herd7} tool finds OOTA values, as expected.
@@
 
-All processes have straightforward semantic dependencies from their
+All threads have straightforward semantic dependencies from their
 relaxed atomic loads to their relaxed atomic stores, which prevents this
@@
 Section~\ref{sec:OOTA: rf vs. rfe}
+(and shown as litmus tests in
+Listings~\ref{lst:OOTA With an Intra-Process Store-Load Link}
+and~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized})
 demonstrate that rf can be atemporal, they demonstrate neither
@@
 so as to demonstrate OOTA-like behavior.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-3-2-proc.litmus @@
+\caption{OOTA With an Intra-Process Store-Load Link}
+\label{lst:OOTA With an Intra-Process Store-Load Link}
+\end{listing}
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-3-2-proc-opt.litmus @@
+\caption{OOTA With an Intra-Process Store-Load Link, Optimized}
+\label{lst:OOTA With an Intra-Process Store-Load Link, Optimized}
+\end{listing}
 
@@
 \begin{quote}
-	\co{z = w * x + w * y + w;}
+\begin{verbatim}
+z = w * x + w * y + w;
+\end{verbatim}
 \end{quote}
@@
 Listing~\ref{lst:Example Code for Evaluation of sdep}
-shows a \co{herd7} process \co{P0()}.
+shows a \co{herd7} thread \co{P0()}.
 Is there a semantic dependency between the load from \co{x} on
@@
 (see Listing~\ref{lst:Code for cbmc-Based Evaluation of sdep})
-that runs the \co{P0()} process's code twice with arbitrary inputs
+that runs the \co{P0()} thread's code twice with arbitrary inputs
 (the loop spanning lines~6-17) and asserts that the results of
@@
 \scriptsize
-	\co{cbmc -DCBMC cbmc/sdep-quadratic.c}
+\begin{verbatim}
+cbmc -DCBMC cbmc/sdep-quadratic.c
+\end{verbatim}
 \end{quote}
@@
 \scriptsize
-	\co{cbmc -DCBMC -DLOAD_CONSTANT=0 cbmc/sdep-quadratic.c}
+\begin{verbatim}
+cbmc -DCBMC -DLOAD_CONSTANT=0 cbmc/sdep-quadratic.c
+\end{verbatim}
 \end{quote}
@@
 \scriptsize
-	\co{cbmc -DCBMC -DLOAD_CONSTANT=0 -DLOAD_LIMITS=46339 cbmc/sdep-quadratic.c}
+\begin{verbatim}
+cbmc -DCBMC -DLOAD_CONSTANT=0 -DLOAD_LIMITS=46339 cbmc/sdep-quadratic.c
+\end{verbatim}
 \end{quote}
@@
 such reordering.
-% @@@ Cite Gristenthwaite
+% @@@ Cite Grisenthwaite
 
@@
 Operating-system kernels encountering an ECC error might note that the
-affected data was being used by only one process, and might react by
-killing that process.
+affected data was being used by only one thread, and might react by
+killing that thread.
 Some user applications might be note that the affected data affected only
@@
 about the nature of causality in these contexts.
+
+\section{Cutting-Room Floor}
+\label{sec:Cutting-Room Floor}
+
+This appendix includes material that was removed but which might later
+prove useful.
+
+@@@ Because there is a cycle, on a real system, at least one of the links
+must be atemporal, that is, either a load returns a value before
+that value is stored, or a store's address or value is computed before
+a prior load returns a value used in that computation.
+
+@@@ Note that C++ implementations are permitted to evaluate to a more
+general notion of OOTA values in cases involving unspecified or undefined
+behavior (UB), such as use of uninitialized objects.
+However, these situations do not involve OOTA cycles, but rather user
+errors and other issues that can lead to unfriendly optimizations that
+in turn result in unpredictable output.
+This paper focuses instead on OOTA cycles.
 

### 2024-01-27 ootavacuous.tex
January 26th reorganization, part 2
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 dependency from a given load to a given store when a change in the value
-loaded that can change the value stored.}
+loaded can change the value stored.}
 
@@
 
-There is a semantic dependency from \co{y} to \co{x} if and only if the
-value of \co{z} is non-zero.
-Otherwise, in the absence of observable behavior (such as volatile
-accesses), the implementation is within its rights to discard this
+As long as \co{z} is zero, and all else being equal, changes in the
+value of \co{y} will not cause changes to the value of \co{x}.
+In that case, the implementation is within its rights to discard this
 statement and act as if later accesses to \co{x} were instead the
@@
 
-Worse yet, suppose that both \co{y} and \co{z} are zero.
-Then there is no semantic dependency from either \co{y} or \co{z} to
-\co{x} unless both \co{y} and \co{z} change.
+\subsection{Semantic Dependencies Can Be Many To One}
+\label{sec:Semantic Dependencies Can Be Many To One}
+
+Suppose that both \co{y} and \co{z} are zero in the previous example.
+Then changes to either but not both of \co{y} or \co{z} (again, all
+else being equal) will not cause changes to the value of \co{x}.
+In other words, there is no semantic dependency from either \co{y}
+or \co{z} to \co{x} unless both \co{y} and \co{z} change.
 Alternatively, when both \co{y} and \co{z} are zero, there is no semantic
@@
 
-Consider also:
+Roughly speaking and all other things being equal, \emph{there is semantic
+dependency from a given set of loads to a given store when a change in
+the values loaded can change the value stored.}
+
+This definition allows the definition of an OOTA cycle to be expanded
+from \co{acyclic(sdep ; rfe)} to:
+
+\begin{quote}
+	There are stores
+	$W_i (i = 0, ..., m)$,
+	where $W_i$ semantically depends on loads
+	$R_{i,j} (j = 0, ..., n_i)$,
+	such that each $R_{i,j}$ reads from one of the $W_k$
+	stores in a different thread.
+\end{quote}
+
+\subsection{Semantic Dependencies Affected by Optimizations}
+\label{sec:Semantic Dependencies Affected by Optimizations}
+
+Consider the following:
 
@@
 
-Also, consider this:
+\subsection{Semantic Dependencies Affected by \co{if} Statements}
+\label{sec:Semantic Dependencies Affected by if Statements}
+
+Consider the following simple \co{if} statement:
 
@@
 
-\section{@@@ This marks the end of the reworked material}
-
-\subsection{Difficulties in Rigorously Defining ``Semantic Dependency''}
-
-But beware of this example:
+\subsection{What Exactly is an Execution?}
+\label{sec:What Exactly is an Execution?}
+
+@@@ Roadmap
+
+\subsubsection{Executions Can Be Merged}
+\label{sec:Executions Can Be Merged}
+
+Consider this example, which appears to be an \co{if} statement:
 
@@
 
-Because the stores executed on each branch of that "if" statement store
+Because the stores executed on each branch of that \co{if} statement store
 identical values to identical addresses, one could equally well regard
@@
 the fact remains that these two viewpoints lead to differing conclusions
-about whether there is a semantic dependency.
+about whether this code represents one execution or two.
+It is the compiler's choice.
 
@@
 
-These examples demonstrate a key point: Although in some cases sdep can
-be a strict function of the source code, \emph{in general, sdep must be
-defined on a per-execution basis}.% \footnote{
+These examples demonstrate a key point: 
+\emph{In general, executions do not necessarily map one-to-one onto
+paths through the source code}.% \footnote{
 % 	Thanks to Alan Stern for the examples in this section and
@@
 
-But what if you really need a definition of semantic dependency
-for a given code fragment that covers all executions?
-One approach is to take the union of semantic dependencies across
-all executions, so that if any execution has a semantic dependency,
-the source code fragment has a semantic dependency.
-This approach might be useful when carrying out static analysis.
-
-On the other hand, many C++ implementations carry out specialization
-optimizations, which consider groups of executions.
-For example, an implementation might special-case a \co{NULL} pointer when
-considering several inline functions that take early exits in that case.
-Tools working with such implementations might therefore choose to focus
-on executions corresponding to a given specialization when generating
-semantic dependencies for code corresponding to that specialization.
-
-Other tools might need to take the intersection of semantic dependencies
-across all executions, so that the source code fragment has a semantic
-dependency only if all executions have a semantic dependency.
-This approach might be useful when analyzing ordering properties
-of that code fragment.
-
-Nevertheless, in general, sdep must be defined on a per-execution basis.
-
-\subsection{Where Do Loaded Values Come From?}
-\label{sec:Where Do Loaded Values Come From?}
-
-Relying on any rf link, whether the intra-thread rfi links or the
-inter-thread rfe links, raises the question of what restrictions
-might apply to C++ implementations' choice of value loaded.
-For non-volatile atomic loads, this is question is answered by
-the C++ standard in 6.9.2p14
-\co{[intro.races]}~\cite{ThomasKoeppe2023N4950}:
-
-\begin{quote}
-	The value of an atomic object M, as determined by evaluation B,
-	shall be the value stored by some side effect A that modifies M,
-	where B does not happen before A.
-\end{quote}
-
-This wording presumably includes the initial value of M as one of the
-possible side effects.
-This means that an implementation might determine that a given atomic
-object always has a given value, and such an implementation might
-skip the load in favor of that value.\footnote{
-	There is some work starting to forbid the implementation
-	from using this assumption, for example, by requiring that
-	the implementation assume that there is at least one thread
-	that it does not know about.
-	\url{https://hackmd.io/@gonzalob/Bk_ufPZ6o/edit}}
-
-Please note that this does not apply to volatile atomic objects,
-for which the implementation is prohibited from assuming that the
-value loaded has any relationship whatsoever to any initialization
-or any values stored.
-This prohibition may come as a surprise to those carefully reading the
-C++ standard~\cite{ThomasKoeppe2023N4950}
-and finding only this wording constraining volatile accessess in
-4.1.2p6
-\co{[intro.abstract]}:
-
-\begin{quote}
-	Accesses through volatile glvalues are evaluated strictly
-	according to the rules of the abstract machine.
-\end{quote}
-
-Unfortunately, volatile semantics are mostly folklore passed by word
-of mouth among some compiler writers and some device-driver developers.
-The key point is that volatile accesses must interact with MMIO
-registers in such a way as to permit device drivers to be written.
-And until such time as the semantics of all MMIO registers of all
-devices have been formalized, published, and incorporated into
-C++ implementations,\footnote{
-	Given that some CPU vendors have started publishing formal
-	memory models, this happy day might actually arrive.
-	But we strongly recommend against holding your breath while
-	waiting for it.}
-these implementations cannot asssume that they can predict the value
-returned by a volatile load.
-
-\subsection{On OOTA Definitions}
-\label{sec:On OOTA Definitions}
-
-The earliest work defines ``OOTA cycle'' by example, without a precise
-definition.
-Sometimes ``causal cycle'' is used as if it was a definition, but
-without a clear decision thread for what does and does not
-constitute a causal cycle.
-P0442R0 defines an OOTA cycle as a fixed-point computation that is
-destroyed by perturbations, which makes perfect sense to that paper's
-authors, but has left others unsatisfied.
-Still others argue (perhaps correctly) that it is impossible to provide
-a general definition of ``OOTA cycle'', but this simply raises
-the question of whether a \emph{useful} definition can be formulated.
-
-@@@ The C++ standard says this in 33.5.4p8
-(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
-
-\begin{quote}
-	Implementations should ensure that no “out-of-thin-air” values
-	are computed that circularly depend on their own computation.
-\end{quote}
-
-The following sections discuss properties of and constraints on
-useful definitions of ``semantic dependency''.
-
-\subsubsection{@@@ Semantic Dependencies Are Per Execution}
-\label{sec:Semantic Dependencies Are Per Execution}
-
-\subsubsection{Semantic Dependencies Depend on Context}
-\label{sec:Semantic Dependencies Depend on Context}
+\subsubsection{Executions Depend on Context}
+\label{sec:Executions Depend on Context}
 
@@
 
-\emph{Any analysis of semantic dependencies must either be fully informed
+\emph{Any analysis of a given execution must either be fully informed
 of the enclosing context, or much make conservative assumptions about
@@
 
-\subsubsection{Semantic Dependencies Can Be Nondeterministic}
-\label{sec:Semantic Dependencies Can Be Nondeterministic}
+\subsubsection{Executions Can Be Nondeterministic}
+\label{sec:Executions Can Be Nondeterministic}
 
@@
 
-\subsubsection{Semantic Dependencies And Optimization}
-\label{sec:Semantic Dependencies And Optimization}
-
-Consider again the following:
-
-\begin{quote}
-\begin{verbatim}
-x = y + z;
-\end{verbatim}
-\end{quote}
-
-Suppose that the implementation can prove that at any time that
-the above statement might execute, \co{y} is equal to \co{z}.
-That would allow the implementation to optimize, that is, to act as if
-the source code was instead as follows:
-
-\begin{quote}
-\begin{verbatim}
-x = 2 * y;
-\end{verbatim}
-\end{quote}
-
-Or equally valid:
-
-\begin{quote}
-\begin{verbatim}
-x = 2 * z;
-\end{verbatim}
-\end{quote}
-
-In this case, is the semantic dependency from \co{y} to \co{x},
-from \co{z} to \co{x}, or both \co{y} and \co{z} to \co{x}?
-The answer is that this is a free choice on the part of the
-implementation.
-Those wishing to change the value of \co{y} or \co{z} externally
-(for example, using a debugger) would do well to avail themselves
-of the \co{volatile} keyword.
-
-\emph{Semantic dependencies can be a free choice on the part of the
-C++ implementation.}
-
-\subsubsection{Semantic Dependencies and I/O}
-\label{sec:Semantic Dependencies and I/O}
-
-A computation might include I/O:
-
-\begin{quote}
-\begin{verbatim}
-x = y * input_int();
-\end{verbatim}
-\end{quote}
-
-Then there is a semantic dependency from \co{y} to \co{x} only if
-the number input is non-zero.
-Similar issues arise when reading timestamps.
-
-In addition, I/O operations are observed behavior, which greatly restricts
-optimizations involving them, in turn reducing the need to evaluate
-potential semantic dependencies.
-
-\subsubsection{Semantic Dependencies Can Be Complex}
-\label{sec:Semantic Dependencies Can Be Complex}
+\subsubsection{Executions Can Be Complex}
+\label{sec:Executions Can Be Complex}
 
@@
 Here, the presence or absence of a semantic dependency from \co{y} to
-\co{x} depends not only on the value of \co{z}, but also the code in a
-function that, although single threaded, is horribly complex.
+\co{x} depends not only on the value of \co{z}, but also and execution
+that passes through the code in a function that, although single threaded,
+is horribly complex.
 
@@
 
-The key point here is that diferent values might cause almost completely
-disjoint code to be executed.
+The key point here is that diferent values might result in almost
+completely disjoint executions.
 
@@
 300~years later, in 1994.
-Therefore, it was only after 1994 that code relying on FLT was known
-not to have a semantic dependency.
+Therefore, it was only after 1994 that certain types of executions
+relying on FLT were known not to have a semantic dependency.
 
@@
 \emph{Because of unsolved mathematical problems, precise definitions of
-``semantic dependency'' have limitations.
-Therefore, C++ implementations must assume that a given dependency is
-a semantic dependency unless and until it can prove otherwise.}
-
-\subsubsection{Implementers and Users Influence Semantic-Dependencies Definition}
-\label{sec:Implementers and Users Influence Semantic-Dependencies Definition}
+``execution'' have limitations.
+Therefore, C++ implementations must assume that a dependency passing
+through a given execution is a semantic dependency unless and until it
+can prove otherwise.}
+
+\subsubsection{Implementers and Users Influence the Definition of Execution}
+\label{sec:Implementers and Users Influence the Definition of Execution}
 
@@
 of the architects of a given CPU family.
-Therefore, the code above has a a semantic dependency when run on systems
-choosing signed \co{char}, but not when run on systems choosing unsigned
-\co{char}.
+Therefore, the code above has a semantic dependency in executions running
+on systems choosing signed \co{char}, but not those running on systems
+choosing unsigned \co{char}.
 So score one for the implementers, albeit by conscious choice on the
@@
 Given this command-line argument, there is always a semantic dependency
-from the \co{foo()} function's argument to its return value.
+in the execution running from the \co{foo()} function's argument to its
+return value.
 So score one for at least some implementations' users.
@@
 
-\emph{A definition of sdep drawn strictly from the standard will be
-at best an approximation to a definition that is useful in practice.}
-
-\subsubsection{Complex Semantic Dependencies Analysis is a Choice}
-\label{sec:Complex Semantic Dependencies Analysis is a Choice}
-
-All of these tools can incur high overhead, especially manual code
-inspection.
-However, as we will see, the actual C++ implementations themselves need
-not use these tools.
-Their current analysis is guaranteed to be sufficient, courtesy of the
-real-world constraints discussed in the next section, at least assuming
-that their analysis is in fact correct.
-Furthermore, as we will also see, in some special but commonly occurring
-circumstances, there are much simpler methods that incur negligible overhead.
+\emph{A definition of ``execution'' drawn strictly from the standard will
+be at best an approximation to a definition that is useful in practice.}
+
+\subsubsection{So What Exactly is an Execution, Anyway?}
+\label{sec:So What Exactly is an Execution, Anyway?}
+
+Given all of these complications in C++ executions, and given that
+semantic dependencies must be defined in terms of executions, it makes
+sense to lower our sights from full generality.
+We therefore restrict ourselves to a specific context:
+traditional compilers.
+Not only that, we will take the notion of semantic dependence (and thus
+OOTA) to be per-compiler rather than absolute.
+
+For the special case of non-volatile atomic operations, we will further
+restrict ourselves to compilers that do single-threaded analysis, with
+the possible addition of specific constraints deduced from multi-threaded
+analysis.
+For volatile atomic operations, compilers are prohibited from inventing
+loads, and we will restrict ourselves to compilers that extend that
+prohibition to non-volatile atomic loads.
+
+@@@ LTO?
+
+These restrictions permit us to bring to bear the real-world
+constraints introduced by the next section.
+
+@@@  Need to add material on inventing loads, which is currently
+in an appendix.
 
@@
 potentially all the way down to zero.
+
+\section{@@@ This marks the end of the reworked material}
+
+\subsection{But what if you want source-code dependencies?}
+
+But what if you really need a definition of semantic dependency
+for a given code fragment that covers all executions?
+One approach is to take the union of semantic dependencies across
+all executions, so that if any execution has a semantic dependency,
+the source code fragment has a semantic dependency.
+This approach might be useful when carrying out static analysis.
+
+On the other hand, many C++ implementations carry out specialization
+optimizations, which consider groups of executions.
+For example, an implementation might special-case a \co{NULL} pointer when
+considering several inline functions that take early exits in that case.
+Tools working with such implementations might therefore choose to focus
+on executions corresponding to a given specialization when generating
+semantic dependencies for code corresponding to that specialization.
+
+Other tools might need to take the intersection of semantic dependencies
+across all executions, so that the source code fragment has a semantic
+dependency only if all executions have a semantic dependency.
+This approach might be useful when analyzing ordering properties
+of that code fragment.
+
+Nevertheless, in general, sdep must be defined on a per-execution basis.
+
+\subsection{Where Do Loaded Values Come From?}
+\label{sec:Where Do Loaded Values Come From?}
+
+Relying on any rf link, whether the intra-thread rfi links or the
+inter-thread rfe links, raises the question of what restrictions
+might apply to C++ implementations' choice of value loaded.
+For non-volatile atomic loads, this is question is answered by
+the C++ standard in 6.9.2p14
+\co{[intro.races]}~\cite{ThomasKoeppe2023N4950}:
+
+\begin{quote}
+	The value of an atomic object M, as determined by evaluation B,
+	shall be the value stored by some side effect A that modifies M,
+	where B does not happen before A.
+\end{quote}
+
+This wording presumably includes the initial value of M as one of the
+possible side effects.
+This means that an implementation might determine that a given atomic
+object always has a given value, and such an implementation might
+skip the load in favor of that value.\footnote{
+	There is some work starting to forbid the implementation
+	from using this assumption, for example, by requiring that
+	the implementation assume that there is at least one thread
+	that it does not know about.
+	\url{https://hackmd.io/@gonzalob/Bk_ufPZ6o/edit}}
+
+Please note that this does not apply to volatile atomic objects,
+for which the implementation is prohibited from assuming that the
+value loaded has any relationship whatsoever to any initialization
+or any values stored.
+This prohibition may come as a surprise to those carefully reading the
+C++ standard~\cite{ThomasKoeppe2023N4950}
+and finding only this wording constraining volatile accessess in
+4.1.2p6
+\co{[intro.abstract]}:
+
+\begin{quote}
+	Accesses through volatile glvalues are evaluated strictly
+	according to the rules of the abstract machine.
+\end{quote}
+
+Unfortunately, volatile semantics are mostly folklore passed by word
+of mouth among some compiler writers and some device-driver developers.
+The key point is that volatile accesses must interact with MMIO
+registers in such a way as to permit device drivers to be written.
+And until such time as the semantics of all MMIO registers of all
+devices have been formalized, published, and incorporated into
+C++ implementations,\footnote{
+	Given that some CPU vendors have started publishing formal
+	memory models, this happy day might actually arrive.
+	But we strongly recommend against holding your breath while
+	waiting for it.}
+these implementations cannot asssume that they can predict the value
+returned by a volatile load.
+
+\subsection{On OOTA Definitions}
+\label{sec:On OOTA Definitions}
+
+The earliest work defines ``OOTA cycle'' by example, without a precise
+definition.
+Sometimes ``causal cycle'' is used as if it was a definition, but
+without a clear decision thread for what does and does not
+constitute a causal cycle.
+P0442R0 defines an OOTA cycle as a fixed-point computation that is
+destroyed by perturbations, which makes perfect sense to that paper's
+authors, but has left others unsatisfied.
+Still others argue (perhaps correctly) that it is impossible to provide
+a general definition of ``OOTA cycle'', but this simply raises
+the question of whether a \emph{useful} definition can be formulated.
+
+@@@ The C++ standard says this in 33.5.4p8
+(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
+
+\begin{quote}
+	Implementations should ensure that no “out-of-thin-air” values
+	are computed that circularly depend on their own computation.
+\end{quote}
+
+The following sections discuss properties of and constraints on
+useful definitions of ``semantic dependency''.
+
+\subsubsection{@@@ Semantic Dependencies Are Per Execution}
+\label{sec:Semantic Dependencies Are Per Execution}
+
+\subsubsection{@@@ Semantic Dependencies Depend on Context}
+\label{sec:Semantic Dependencies Depend on Context}
+
+\subsubsection{@@@ Semantic Dependencies Can Be Nondeterministic}
+\label{sec:Semantic Dependencies Can Be Nondeterministic}
+
+\subsubsection{Semantic Dependencies And Optimization}
+\label{sec:Semantic Dependencies And Optimization}
+
+@@@ What do we do with this one???  Not sure we need it.
+
+Consider again the following:
+
+\begin{quote}
+\begin{verbatim}
+x = y + z;
+\end{verbatim}
+\end{quote}
+
+Suppose that the implementation can prove that at any time that
+the above statement might execute, \co{y} is equal to \co{z}.
+That would allow the implementation to optimize, that is, to act as if
+the source code was instead as follows:
+
+\begin{quote}
+\begin{verbatim}
+x = 2 * y;
+\end{verbatim}
+\end{quote}
+
+Or equally valid:
+
+\begin{quote}
+\begin{verbatim}
+x = 2 * z;
+\end{verbatim}
+\end{quote}
+
+In this case, is the semantic dependency from \co{y} to \co{x},
+from \co{z} to \co{x}, or both \co{y} and \co{z} to \co{x}?
+The answer is that this is a free choice on the part of the
+implementation.
+Those wishing to change the value of \co{y} or \co{z} externally
+(for example, using a debugger) would do well to avail themselves
+of the \co{volatile} keyword.
+
+\emph{Semantic dependencies can be a free choice on the part of the
+C++ implementation.}
+
+\subsubsection{Semantic Dependencies and I/O}
+\label{sec:Semantic Dependencies and I/O}
+
+@@@ Not sure we need this one.
+
+A computation might include I/O:
+
+\begin{quote}
+\begin{verbatim}
+x = y * input_int();
+\end{verbatim}
+\end{quote}
+
+Then there is a semantic dependency from \co{y} to \co{x} only if
+the number input is non-zero.
+Similar issues arise when reading timestamps.
+
+In addition, I/O operations are observed behavior, which greatly restricts
+optimizations involving them, in turn reducing the need to evaluate
+potential semantic dependencies.
+
+\subsubsection{@@@ Semantic Dependencies Can Be Complex}
+\label{sec:Semantic Dependencies Can Be Complex}
+
+\subsubsection{@@@ Implementers and Users Influence Semantic-Dependencies Definition}
+\label{sec:Implementers and Users Influence Semantic-Dependencies Definition}
+
+\subsubsection{Complex Semantic Dependencies Analysis is a Choice}
+\label{sec:Complex Semantic Dependencies Analysis is a Choice}
+
+All of these tools can incur high overhead, especially manual code
+inspection.
+However, as we will see, the actual C++ implementations themselves need
+not use these tools.
+Their current analysis is guaranteed to be sufficient, courtesy of the
+real-world constraints discussed in the next section, at least assuming
+that their analysis is in fact correct.
+Furthermore, as we will also see, in some special but commonly occurring
+circumstances, there are much simpler methods that incur negligible overhead.
 

### 2024-01-27 ootavacuous.tex
Wordsmith executions and specialization optimizations
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 considering several inline functions that take early exits in that case.
-Tools working with such implementations might therefore choose to focus
-on executions corresponding to a given specialization when generating
-semantic dependencies for code corresponding to that specialization.
+For another example, an implementation might produce separate code for
+cases where an integer is negative, zero, and positive.
+Tools working with such implementations might therefore choose to group
+executions by specialization, and then separately analyze semantic
+dependencies for code corresponding to each specialize execution.
 

### 2024-01-27 ootavacuous.tex
January 27 application of Alan Stern feedback, #1
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \pagebreak
+
+\section{Introduction}
+\label{sec:Introduction}
+
+@@@ Add brief description and roadmap of paper, including the simplifications
+and restrictions.
 
@@
 
-Roughly speaking and all other things being equal, \emph{there is semantic
-dependency from a given load to a given store when a change in the value
-loaded can change the value stored.}
+Roughly speaking, \emph{there is semantic dependency from a given load
+to a given store when, all other things being equal a change in the
+value loaded can change the value stored.}
 
@@
 in using the term \emph{full C++} to denote the standard including the
-above prohibition, and the term \emph{strict C++} to denote a hypothetical
-standard that excludes this prohibition, but is otherwise identical
-to full C++.
+above prohibition, and the term \emph{strict C++}\footnote{
+	@@@ Future bikeshedding of name.}
+% @@@ Alternatives thus far: OOTA-prone C++, OOTA-enabled C++,
+% @@@ relaxed C++, OOTA C++, OOTA-agnostic C++.
+to denote a hypothetical standard that excludes this prohibition, but
+is otherwise identical to full C++.
 Unqualified \emph{C++} means full C++.
@@
 
-This paper will use the \co{herd7} tool to analyze fragments of C++ code.
+This paper will use the \co{herd7}\footnote{
+	Available at \path{https://github.com/herd/herdtools7}.}
+tool to analyze fragments of C++ code.
 This tool carries out the moral equivalent of full state-space searches
-of concurrent code fragments, and further will in some cases identify
-OOTA cycles.
+of concurrent code fragments.
+In some cases its output will include executions with OOTA cycles,
+on occasion reporting undefined values for the variables involved in
+the cycle.
 
@@
 
-\subsection{Semantic Dependencies and Source Code}
+\subsection{Properties of Semantic Dependencies}
+\label{sec:Properties of Semantic Dependencies}
+
+\subsubsection{Semantic Dependencies and Source Code}
 \label{sec:Semantic Dependencies and Source Code}
@@
 
-\subsection{Semantic Dependencies Can Be Many To One}
+\subsubsection{Semantic Dependencies Can Be Many To One}
 \label{sec:Semantic Dependencies Can Be Many To One}
@@
 
-\subsection{Semantic Dependencies Affected by Optimizations}
+\subsubsection{Semantic Dependencies Affected by Optimizations}
 \label{sec:Semantic Dependencies Affected by Optimizations}
@@
 updated pairwise atomically and are subject to change at any time by
-other threads.
-
-\subsection{Semantic Dependencies Affected by \co{if} Statements}
+other threads (a point we will return to).
+
+\subsubsection{Semantic Dependencies Affected by \co{if} Statements}
 \label{sec:Semantic Dependencies Affected by if Statements}
@@
 Although software projects relying on ordering from load-to-store control
-dependencies and using a variety of compilers should take the single-store
-viewpoint~\cite{Howells2009membartxt},
-the fact remains that these two viewpoints lead to differing conclusions
-about whether this code represents one execution or two.
-It is the compiler's choice.
+dependencies and using a variety of compilers should should err on the
+side of caution by assuming there is no semantic
+dependency~\cite{Howells2009membartxt},
+the fact remains that reasonable C++ implementations might disagree as
+to whether this code represents one execution or two.
+It is the implementations's choice.
 

### 2024-01-28 ootavacuous.tex
January 27 application of Alan Stern feedback, #2
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+@@@ Roadmap
+
 \subsubsection{Semantic Dependencies and Source Code}
@@
 \end{quote}
+
+This definition makes OOTA more complicated than a simple cycle but
+we will continue to refer to it as an OOTA cycle out of habit.
+Note that this new definition includes and generalizes our previous
+\co{acyclic(sdep ; rfe)} definition.
 
@@
 traditional compilers.
-Not only that, we will take the notion of semantic dependence (and thus
-OOTA) to be per-compiler rather than absolute.
-
-For the special case of non-volatile atomic operations, we will further
-restrict ourselves to compilers that do single-threaded analysis, with
-the possible addition of specific constraints deduced from multi-threaded
-analysis.
-For volatile atomic operations, compilers are prohibited from inventing
-loads, and we will restrict ourselves to compilers that extend that
-prohibition to non-volatile atomic loads.
+It will then turn out to be necessary to take the notion of semantic
+dependence, and thus OOTA, to be compiler-relative rather than absolute.
+
+We will show that programs built using strict C++ compilers that treat all
+atomic operations as volatile can never exhibit OOTA, and the same is true
+for compilers obeying the weaker constraint that atomic accesses may not
+invented, provided the compiler does not perform multi-thread analysis.
 

### 2024-01-28 ootavacuous.tex
Apply hardware-architect feedback to load/store ordering appendix
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 value returned by that load!
-Which explains why some older hardware systems really do reorder prior
-loads and later stores.
+Which explains why some hardware systems really do reorder prior loads
+and later stores.
 
@@
 
-One possible explanation is ECC errors.
+One explanation is ECC errors.
 
@@
 ECC errors would be directly visible to software.
-If the only possible reaction to an uncorrectable ECC error was to terminate
-the program suffering that ECC error, there would be no problem.
-
-However, some code has ways of handling even uncorrectable errors.
+If the only possible reaction to an uncorrectable ECC error was to
+terminate the program suffering that ECC error, it would be safe to
+allow subsequent stores to commit as soon as all prior loads' address
+translations had completed successfully.
+
+However, some kernels and applications have ways of handling even
+uncorrectable errors.
 Operating-system kernels encountering an ECC error might note that the
-affected data was being used by only one thread, and might react by
-killing that thread.
+affected data was being used by only one user process, and might react by
+killing that user process.
 Some user applications might be note that the affected data affected only
 a particular computation, and might react by restarting that computation.
-Both the kernel and the user application would need to continue execution,
-and might thus desire an exact exception.
+In both cases, it is only reasonable to react to an uncorrectable ECC
+error in a load from a read-only mapped file by re-reading that data
+from that file, then restarting that load instruction.
+In all cases, the kernel and the user application would need to continue
+execution, and would thus require an exact exception.
 
@@
 also require exact exceptions.
-After all, the software is going to need to be able to correct the
-error and then fix up state as if the load had returned the fixed-up
-value.
+After all, the software is going to need to be able to correct the error
+and then fix up the state to make it appear that the load had returned
+the fixed-up value.
 

### 2024-01-29 c2latex.sh
Move c2latex into the repo
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -1,0 +1,23 @@
+#!/bin/sh
+#
+# Numbers a code fragment and converts tabs to spaces.
+#
+# This program is free software; you can redistribute it and/or modify
+# it under the terms of the GNU General Public License as published by
+# the Free Software Foundation; either version 2 of the License, or
+# (at your option) any later version.
+#
+# This program is distributed in the hope that it will be useful,
+# but WITHOUT ANY WARRANTY; without even the implied warranty of
+# MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+# GNU General Public License for more details.
+#
+# You should have received a copy of the GNU General Public License
+# along with this program; if not, you can access it online at
+# http://www.gnu.org/licenses/gpl-2.0.html.
+#
+# Copyright (C) IBM Corporation, 2007
+#
+# Authors: Paul E. McKenney <paulmck@linux.vnet.ibm.com>
+
+awk '{printf "%2d %s\n", NR, $0}' | sed -e 's/	/  /g' | sed -e 's/ *$//'

### 2024-01-29 ootavacuous.sh

@@ -30,7 +30,7 @@ BEGIN {
 	print bs "scriptsize";
 	print bs "begin{verbatim}";
 	print "___EOF___";
-	print "grep -v " sq "^[(/ ]" bs "*" sq " < " path " | c2latex.sh";
+	print "grep -v " sq "^[(/ ]" bs "*" sq " < " path " | ./c2latex.sh";
 	print "cat << " sq "___EOF___" sq;
 	print bs "end{verbatim}";
 	print "}";
@@ -43,14 +43,14 @@ BEGIN {
 	print bs "scriptsize";
 	print bs "begin{verbatim}";
 	print "___EOF___";
-	print "grep -v " sq "^[(/ ]" bs "*" sq " < " path " | c2latex.sh";
+	print "grep -v " sq "^[(/ ]" bs "*" sq " < " path " | ./c2latex.sh";
 	print "cat << " sq "___EOF___" sq;
 	print bs "end{verbatim}";
 	print "";
 	print "Analysis by " dq bs "co{herd7 -c11 " path dq "}:";
 	print bs "begin{verbatim}";
 	print "___EOF___";
-	print "herd7 -c11 " path " | grep -v " sq "^$" sq " | c2latex.sh ";
+	print "herd7 -c11 " path " | grep -v " sq "^$" sq " | ./c2latex.sh ";
 	print "cat << " sq "___EOF___" sq;
 	print bs "end{verbatim}";
 	print "}";
@@ -63,7 +63,7 @@ BEGIN {
 	print bs "scriptsize";
 	print bs "begin{verbatim}";
 	print "___EOF___";
-	print "herd7 -c11 " path " | grep -v " sq "^$" sq " | c2latex.sh ";
+	print "herd7 -c11 " path " | grep -v " sq "^$" sq " | ./c2latex.sh ";
 	print "cat << " sq "___EOF___" sq;
 	print bs "end{verbatim}";
 	print "}";

### 2024-01-29 ootavacuous.tex
Add Stephen Dolan example to the undefined-behavior appendix
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 of the proposed \co{std::observable()}.
+
+\subsection{Lifetime-Based Undefined Behavior}
+\label{sec:Lifetime-Based Undefined Behavior}
+
+Not all undefined behavior can be backstopped using
+\co{std::unreachable()}.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 int *foo(int *r)
+ 2 {
+ 3   b.store(r, memory_order_relaxed);
+ 4   return &global;
+ 5 }
+ 6
+ 7 int go(int *p)
+ 8 {
+ 9   int r;
+10   int *q = foo(&r);
+11
+12   r = 2;
+13   *p = 1;
+14   *q = r;
+15   return *q - r;
+16 }
+17
+18 void t1(void)
+19 {
+20   int r1 = go(a.load(memory_order_relaxed));
+21   printf("go() = %d\n", r1);
+22 }
+23
+24 void t2(void)
+25 {
+26   int *r2 = b.load(memory_order_relaxed);
+27   a.store(r2, memory_order_relaxed);
+28 }
+\end{verbatim}
+\caption{Lifetime-Based Undefined Behavior}
+\label{lst:Lifetime-Based Undefined Behavior}
+\end{listing}
+
+Listing~\ref{lst:Lifetime-Based Undefined Behavior}
+shows an example loosely based on one presented by Stephen Dolan\footnote{
+	\url{https://stedolan.net/talks/2024/loadstore}}
+in which function \co{go()} invokes \co{foo()} with a pointer to \co{go()}'s
+local variable \co{r}.
+However, \co{foo()} uses a relaxed store to place this pointer in
+an atomic variable \co{b}.
+Because this store is relaxed, it is not ordered against other code
+in \co{go()} and \co{t1()}, which means that \co{t2()} might well
+read this pointer from \co{b} either before the beginning or after
+the end of \co{r}'s lifetime, neither of which is well-defined.
+
+Of course, any attempt to dereference the pointer read from either
+\co{a} or \co{b} could result in undefined behavior.
+
+The lesson is clear:
+Properly synchronize reads and writes of pointers with the lifetime
+of the pointed-to object.
+
+This example could make lines~3 and~27 be \co{memory_order_release} and
+lines~20 and~26 be \co{memory_order_acquire} in order to avoid carrying
+a pointer to \co{r} before the lifetime of \co{r} has started.
+Note that \co{std::unreachable()} does not suffice because the CPU
+can also reorder relaxed stores.
+Note also that load-store ordering is not sufficient to prevent a
+pointer to \co{r} from being stored to \co{b} prior to the start of
+\co{r}'s lifetime because there is no atomic load between the start
+of that lifetime and the relaxed store to \co{b}.
+
+There are a number of ways to prevent carrying a pointer to \co{r} after
+the lifetime of \co{r} has ended, and these are left as an exercise for
+the reader.
 
@@
 affected data was being used by only one user process, and might react by
-killing that user process.
+killing that user process, taking care to account for memory shared with
+other processes.
 Some user applications might be note that the affected data affected only
@@
 ECC errors are detected near the end of their corresponding load
-instructions' execution, so that any need for exact exceptions rules
-out the commiting of any subsequent stores before the load has fetched
-its value.
+instructions' execution, so that any need for exact exceptions rules out
+the commiting of any subsequent stores until after the load has fetched
+(and possibly corrected) its value.
 
@@
 \item	Some hardware might choose to forego ECC, for example, in order to
-	reduce cost or to improve energy efficiency.
+	reduce cost for low-end systems or to improve energy efficiency
+	for battery-powered systems.
 \item	ECC error correction might still be done in hardware, avoiding
@@
 	exceptions.
-\item	As late as early 2024, some GPUs have been observed reordering
+\item	As late as early 2024, some GPGPUs have been observed reordering
 	earlier loads against later stores.

### 2024-01-29 ootavacuous.tex
FYI, possible progress on language-level OOTA problem
On Sun, Jan 28, 2024 at 04:37:33PM -0800, Paul E. McKenney wrote:
> > Do I need the c2latex.sh script to build this?  Whatever it is, I don't
> > have it.
>
> You do!  One-line awk script, but attached.
>
> I also moved it into the git archive and pushed it out.
>
> > Also, the \includegraphics LaTeX command doesn't seem to be able to work
> > with .svg files.  I get "File not found" errors.
>
> Ah, apologies, you need to build with "sh ootavacuous.sh".  Very crude,
> I admit.  I created and pushed a README file saying this.

All working no

@@
 
-OOTA occurs when a group of threads load from each others' stores to
-each of a group of objects, but where each thread's store depends on
-the value returned by that thread's load.
-A given out-of-thin-air value passes around the resulting \emph{OOTA cycle}.
+In broad terms,
+OOTA occurs when a group of threads load from each others' stores
+and each thread's store depends on the value returned by that thread's load.
+The collection of loads and stores forms an \emph{OOTA cycle}.
+In the most extreme cases a nonsensical value can pop up ``out of thin air'',
+but this is not necessary.
 
@@
 
-However, it is important to distinguish OOTA cycles from the OOTA-like
-behavior that can be caused by simple reordering.
+It is important to distinguish true OOTA cycles from OOTA-like
+behavior caused by simple reordering.
 An example of simple reordering is shown in
@@
 
-Current C++ implementations can and do exhibit this simple-reordering
-behavior.
+\noindent
+Current C++ implementations can and do exhibit this reordering behavior.
 
@@
 shows a simple example where an OOTA cycle might result in all of \co{x},
-\co{y}, \co{r1}, and \co{r2} have final values of 42, despite the fact
-that there is nothing in the initial values or the executable code that
-would support such an outcome:
+\co{y}, \co{r1}, and \co{r2} having final values of 42, despite the fact
+that there is nothing in the initial values or the executable code to
+support such an outcome:
 
 \begin{enumerate}
-\item   Line~6 loads from \co{x} into \co{r1}, somehow obtaining the
-	value 42, never mind the \co{x}'s initial value of zero.
+\item   Line~6 loads from \co{x} into \co{r1}, claiming to read
+	the value of line~13's store rather than \co{x}'s initial value
+	and somehow obtaining 42.
 \item   Line~7 stores \co{r1}, and thus 42, to \co{y}.
 \item   Line~12 loads 42 from \co{y} to \co{r2}.
-\item   Line~13 stores 42 to \co{x}, justifying the value loaded by
-	line~6 after the fact.
+\item   Line~13 stores 42 to \co{x}, justifying the value loaded by line~6.
 \end{enumerate}
 
-Fortunately, the C++ standard prohibits such OOTA cycles in 33.5.4p8
+Because nothing else in the C++ memory model rules out such OOTA cycles,
+the C++ standard explicitly prohibits them in 33.5.4p8
 (\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
@@
 
+\noindent
 The standard's prohibition of OOTA is of course important, but those of us
@@
 ordering in strongly ordered systems and by data dependency ordering in
-weakly ordered systems.
-Real-world implementations cannot store something until after they load
-it, so there is a \emph{semantic dependency} from line~6 to line~7 and
-another from line~12 to line~13 in
-Listing~\ref{lst:Simple OOTA}.
-These semantic dependencies cause compilers to emit instructions such
-that the hardware enforcement applies.\footnote{
+weakly ordered systems.\footnote{
 	The need to prohibit simple OOTA is one reason why compiler-based
@@
 
-Roughly speaking, \emph{there is semantic dependency from a given load
-to a given store when, all other things being equal a change in the
-value loaded can change the value stored.}
+In Listing~\ref{lst:Simple OOTA}
+there is a \emph{semantic dependency} from line~6 to line~7 and
+another from line~12 to line~13.
+(Roughly speaking, there is a semantic dependency from a given load
+to a given store when \emph{all other things being equal, a change in the
+value loaded can change the value stored or prevent the store from
+occurring at all.}
+Here the dependencies are trivial, because the values stored simply
+\emph{are} the values that were loaded.)
+Since real-world CPUs cannot store something
+until they have determined its value,\footnote{
+	Another way of saying this is that real-world CPUs do not
+	speculate stores.}
+the stores in lines~7 and~13 cannot take place until the CPU
+knows what values are loaded by lines~6 and~12, respectively.
+Thus the hardware orders these stores after their corresponding loads,
+and this ordering prevents the OOTA result.
 
@@
 in using the term \emph{full C++} to denote the standard including the
-above prohibition, and the term \emph{strict C++}\footnote{
+prohibition above, and the term \emph{strict C++}\footnote{
 	@@@ Future bikeshedding of name.}
@@
 
-The next section looks at how prior work has refined this prohibition.
+The next section looks at how prior work has refined these issues.
 
@@
 concerns.
-The remainder of this document will focus on examples lacking UB.
-
-All this work focused on either identifying OOTA or on how C++
+The remainder of this document will concentrate on examples lacking UB.
+
+All this work focused on either identifying OOTA or seeing how C++
 implementations could avoid it.
-None of this work applied real-world hardware ordering to the problem
-of avoiding OOTA cycles.
-Which might explain why no known real-world C++ implementation results
-in OOTA executions.
-
-This paper therefore digs more deeply into OOTA cycles in the light
+None applied real-world hardware ordering constraints to the problem
+of avoiding OOTA cycles,
+yet doing so might help explain why no known real-world C++ implementation
+results in OOTA behavior.
+We therefore dig more deeply into OOTA cycles in the light
 of these real-world constraints.
@@
 on occasion reporting undefined values for the variables involved in
-the cycle.
+the cycle.\footnote{
+	This happens only some of the time because of idiosyncrasies
+	in the algorithm used by \co{herd7}'s self-consistency solver.}
 
@@
 Listing~\ref{lst:OOTA Cycle}
-shows a code fragment that in strict C++ would have an OOTA cycle (but
-that is of course prohibited in full C++).
-This section describes this fragment, thus giving an overview of
+shows a code fragment that under strict C++ has an OOTA cycle (although
+the cycle is of course prohibited in full C++).
+This section describes the fragment, thereby giving an overview of
 the \co{herd7} tool.
@@
 a name.
-Lines~2-5 initialize variables, in this case setting the initial
+Lines~2--5 initialize variables, in this case setting the initial
 values of the global shared variables \co{x} and \co{y} to zero.
-Lines~7-13 define the first thread, \co{P0()}, and lines~15-21
+Lines~7--13 define the first thread, \co{P0()}, and lines~15--21
 define the second thread, \co{P1()}.\footnote{
@@
 and \co{y}.
-The body of each thread specifies a limited subset of C++ code,
-which also happens to be a somewhat less limited subset of C code.
-
-Line~23 gives an example of a \co{locations} clause, which causes
-\co{herd7} to dump out the final value of \co{x} and \co{y}.
-
+The body of each thread contains a limited subset of C++ code
+(written in a slightly stilted manner to keep the load operations separate
+from the rest).
+
+Line~23 has a \co{locations} clause, which causes
+\co{herd7} to dump out the final values of \co{x} and \co{y}.
 Finally, line~24 specifies an \co{exists} clause, which gives a
-condition to check for the final value of the specified local
-variables.
-The \co{0:} prefix specifies a variable local to \co{P0()} and
-the \co{1:} prefix specifies a variable local to \co{P1()}.
-The \co{/\\} specifies a boolean AND, and the \co{=} specifies
-an equality comparison.
-If a variable appears in an \co{exists} clause, then the final value of
+condition to check for the final values of the specified variables.
+The \co{0:} prefix denotes a variable local to \co{P0()} and
+the \co{1:} prefix denotes a variable local to \co{P1()}.
+The \co{/\\} is a boolean AND, and the \co{=} signs are equality comparisons.
+If a variable appears in the \co{exists} clause then the final value of
 that variable constitutes observable behavior.
@@
 counterintuitive outcome where both threads load the value 42.
-Normally these states would include only those variables mentioned
+Normally these lines would include only those variables mentioned
 in the \co{exists} clause, but because of the \co{locations} clause
-on line~23, the values of \co{x} and \co{y} is also included, which
-can be helpful when debugging.
+the values of \co{x} and \co{y} are also listed, which
+can be helpful for debugging.
 % @@@ As we will see, this outcome is not possible in real-world
@@
 and~\ref{lst:Simple OOTA}
-can be recast as similar \co{herd7} litmus tests, which areas shown in
+can be recast as similar \co{herd7} litmus tests, as shown in
 Listing~\ref{lst:Simple Reordering as Litmus Test}
@@
 
-\section{Semantic Dependencies, Executions, and OOTA}
-\label{sec:Semantic Dependencies; Executions; and OOTA}
+\section{OOTA, Semantic Dependencies, and Executions}
+\label{sec:OOTA; Semantic Dependencies; and Executions}
 
@@
 
-However, semantic dependencies are only one type of link in an OOTA cycle.
-The other link extends from a given store to the load that returns the
+Semantic dependencies are only one type of link in an OOTA cycle.
+The other type extends from a given store to the load that returns the
 value stored.
-It is tempting to make the intuitive argument that this store must precede
-this load in global time, and then to combine this with the notion that
+It is tempting to make the intuitive argument that the store must precede
+the load in global time, and then to combine this with the notion that
 any real C++ implementation must consume global time when computing a
 semantic dependency.
-This combination suggests that such implementations cannot exhibit
-OOTA cycles.
-This combination has been formalized by defining an OOTA cycle as a cycle
-in sdep $\cup$ rf, where sdep is the set of semantic dependencies within
-each thread and rf is the set of store-to-load links, whether intra-thread
+This combination suggests that OOTA cycles cannot occur.
+The idea has been formalized by defining an OOTA cycle as a cycle
+in sdep $\cup$ rf [@@@ Reference needed],
+where sdep is the set of semantic dependencies within
+each thread and rf is the set of store-to-load ``reads-from'' links,
+whether intra-thread
 (rfi) or between threads (rfe).\footnote{
@@
 	and properies of rf, rfe, and rfi.}
-
-This is a fine definition, and is consistent with the words in the C++
-standard, but has an intra-thread problem exemplified by the following
+This is a fine definition and is consistent with the words in the C++
+standard, but it has an intra-thread problem exemplified by the following
 code:
@@
 
+\noindent
 This is an elaboration of \co{thread2()} from
 Listing~\ref{lst:Simple OOTA}
-that adds \co{z} along with lines~2 and~3.
-The problem is that the C++ implementation is within its rights to
+that adds \co{z} along with lines~3 and~4.
+The problem is that a C++ implementation is within its rights to
 note that line~3 might well execute immediately after line~2, giving
-threads no chance to modify \co{y}.
+other threads no chance to modify \co{y} in between.
 Such an implementation might therefore behave as if the source code
@@
 
-Here line~3 has been optimized away in favor of line~4 storing the value
-in \co{r2} that was loaded from \co{y} by line~2.
-And given that the load from \co{z} no longer exists, it cannot possibly
+\noindent
+Here line~3 has been optimized away in favor of line~4 storing the same value
+in \co{r2} that was stored to \co{y} by line~2.
+And given that the load from \co{y} no longer exists, it cannot possibly
 act as a temporal constraint.
 
-In order to avoid these rfi links, we substitute rfe
-for rf, thus defining an OOTA cycle as a cycle in sdep $\cup$ rfe (or
-\co{acyclic(sdep ; rfe)} in \co{herd7} notation), where rfe is the set
-of store-to-load links, but excluding intra-thread rfi links, which are
-instead considered to be part of sdep.
+In order to avoid these rfi links we will substitute rfe
+for rf, defining an OOTA cycle---for now---as a cycle in sdep $\cup$ rfe,
+where rfe is the set of store-to-load links excluding intra-thread rfi links.
+Any rfi links in a cycle can instead be interpreted as part of sdep.
 Although this does shunt additional complexity onto the term
-``semantic dependency'', it also enables us to cleanly separate
+``semantic dependency'', it also enables us to separate cleanly
 the inter-thread and intra-thread portions of any given OOTA cycle.
@@
 
-Some discussions of semantic dependencies assume that semantic
-dependencies are a strict function of the source code.
-Although there are some ways of making this work, many uses of
-semantic dependencies require them to be a function of particular
-executions.
+Some discussions of semantic dependencies assume that they
+are strictly functions of the source code.
+Although there are ways of making this work, many instances of
+semantic dependency must be considered functions of particular executions.
 Consider for example:
@@
 
+\noindent
 As long as \co{z} is zero, and all else being equal, changes in the
-value of \co{y} will not cause changes to the value of \co{x}.
-In that case, the implementation is within its rights to discard this
-statement and act as if later accesses to \co{x} were instead the
-constant zero, at least up to the next assignment to \co{x}.
+value of \co{y} will not cause a change in the value stored to \co{x}.
+As a result, the semantic dependency from \co{y} to \co{x} exists only
+in executions where \co{z} is nonzero,
+which shows it is a property of the execution, not just of the source code.
 
@@
 
-Suppose that both \co{y} and \co{z} are zero in the previous example.
-Then changes to either but not both of \co{y} or \co{z} (again, all
-else being equal) will not cause changes to the value of \co{x}.
-In other words, there is no semantic dependency from either \co{y}
-or \co{z} to \co{x} unless both \co{y} and \co{z} change.
-Alternatively, when both \co{y} and \co{z} are zero, there is no semantic
-dependency from either \co{y} or \co{z} to \co{x}, but rather a semantic
-dependency from the combination of \co{y} \emph{and} \co{z} to \co{x}.
-This means that precise definitions of sdep cannot always rely on
-single-variable perturbations.
+Suppose that in some execution of the previous example,
+both \co{y} and \co{z} are zero.
+Then changes to either \co{y} or \co{z} (again, all
+else being equal) will not cause a change in the value stored to \co{x}.
+In other words, in this execution there is no semantic dependency
+from either \co{y} or \co{z} to \co{x}.
+But there \emph{is} a semantic dependency from the \emph{pair}
+\{\co{y},~\co{z}\} to \co{x},
+because changes to both \co{y} and \co{z}
+can cause the value stored in \co{x} to change.
+This means that accurate definitions of sdep cannot always rely on
+single-variable perturbations;
+they must consider changes to multiple variables.
 See Appendix~\ref{app:Non-Trivial Semantic Dependencies}
@@
 
-Roughly speaking and all other things being equal, \emph{there is semantic
-dependency from a given set of loads to a given store when a change in
-the values loaded can change the value stored.}
-
-This definition allows the definition of an OOTA cycle to be expanded
-from \co{acyclic(sdep ; rfe)} to:
+Since we no longer regard sdep as always relating a single load to a store,
+the notion of a cycle involving sdep appears problematic.
+We are forced to change our definition of an OOTA cycle again;
+we will say that an execution is an instance of OOTA if in that execution:
 
@@
 	There are stores
-	$W_i (i = 0, ..., m)$,
+	$W_i$~($i = 0$, \ldots,~$m$),
 	where $W_i$ semantically depends on loads
-	$R_{i,j} (j = 0, ..., n_i)$,
+	$\{R_{i,j}\}$~($j = 0$, \ldots,~$n_i$),
 	such that each $R_{i,j}$ reads from one of the $W_k$
@@
 
-This definition makes OOTA more complicated than a simple cycle but
+\noindent
+This makes OOTA more complicated than a simple cycle but
 we will continue to refer to it as an OOTA cycle out of habit.
-Note that this new definition includes and generalizes our previous
-\co{acyclic(sdep ; rfe)} definition.
-
-\subsubsection{Semantic Dependencies Affected by Optimizations}
-\label{sec:Semantic Dependencies Affected by Optimizations}
+Note that this new definition includes and generalizes the earlier
+``cycle in sdep $\cup$ rfe'' definition.
+
+\subsubsection{Semantic Dependencies Affected by Cross-Thread Optimizations}
+\label{sec:Semantic Dependencies Affected by Cross-Thread Optimizations}
 
@@
 \begin{verbatim}
-x = y + z;
+x = y - z;
 \end{verbatim}
@@
 
-There are semantic dependencies from \co{y} to \co{x} and from \co{z}
+\noindent
+There appear to be semantic dependencies from \co{y} to \co{x} and from \co{z}
 to \co{x}.
-However, if the implementation knows that the value of \co{y} is
-always the negative of \co{z} there is no semantic dependency and the
-implementation can again act as if later accesses to \co{x} were instead
-the constant zero, again, at least until the next assignment to \co{x}.
-Please note that this leaves aside the question of how the compiler
-would know this if \co{y} and \co{z} are atomic objects that cannot be
-updated pairwise atomically and are subject to change at any time by
+However, if the compiler somehow knows that \co{y} is
+always equal to \co{z} at this point then there is no semantic dependency;
+the compiler can act as if the statement were simply ``\co{x = 0;}''.
+We leave aside the question of how the compiler would know this,
+given that \co{y} and \co{z} cannot be updated simultaneously
+and are subject to change at any time by
 other threads (a point we will return to).
@@
 
-Consider the following simple \co{if} statement:
+Consider the following \co{if} statement:
 
@@
 
+\noindent
 Here there is a semantic dependency from \co{x}, but in some executions
 it extends to \co{y} and in others to \co{z}.
-This is an example of the load affecting not the value or address of
+This is an example of the load affecting not the value of
 a given store, but rather whether or not that store is executed at all.
 
-Which raises the question:
-What exactly is an execution?
+We have seen several examples showing that semantic dependencies may vary
+according to the execution.
+This raises the question: What exactly is an execution?
 

### 2024-01-31 ootavacuous.tex
Update IPC plots to x86 server
Also commit a few PDFs to avoid relying on inkscape converting from .svg
to .pdf, which has odd compatibility issues in recent releases.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 a pointer to \co{r} before the lifetime of \co{r} has started.
+Alternatively, threads could be spawned only after \co{r}'s lifetime
+starts and joined before it ends.
+Placing \co{r} in static global storage would avoid lifetime considerations
+totally.
+And there are of course any number of other synchronization schemes
+that might be used to ensure that pointers are in use only while the
+pointed-to objects are alive.
+
 Note that \co{std::unreachable()} does not suffice because the CPU
@@
 are not confined to weakly ordered systems.
-This system is a laptop with Intel(R) Core(TM) i9-10885H CPU @ 2.40GHz,
-having a single socket with 8~cores, each having a pair of hardware
-threads.
+This system is a dual-socket system with
+Intel(R) Xeon(R) Gold 6138 CPU @ 2.00GHz,
+each socket with 20~cores, each having a pair of hardware
+threads, for a grand total of 80 hardware threads.
 The code generating this data may be found in the \path{CodeSamples/cpu}
@@
 \begin{center}
-\resizebox{3in}{!}{\includegraphics{coe-sh-out}}
-\caption{On x86, coe Links Are Atemporal}
-\label{fig:On x86, coe Links Are Atemporal}
+\resizebox{4in}{!}{\includegraphics{coe-nvals}}
+\caption{On x86, Reasonable CPUs Can Disagree}
+\label{fig:On x86; Reasonable CPUs Can Disagree}
 \end{center}
@@
 
-Figure~\ref{fig:On x86, coe Links Are Atemporal}
-shows the results of 15 of the laptop's hardware threads simultaneously
-storing their identifying integer (ranging from 0 to 14) to a shared
+Figure~\ref{fig:On x86; Reasonable CPUs Can Disagree}
+shows the results of 79 of the systems's hardware threads simultaneously
+storing their identifying integer (ranging from 0 to 78) to a shared
 variable, then repeatedly doing timestamped reads from that variable.
-The x-axis displays time in TSC periods, each of which is about 0.417~ns
+The x-axis displays time in TSC periods, each of which is about 0.5~ns
 in duration.
@@
 \begin{center}
+\resizebox{4in}{!}{\includegraphics{coe-sh-out}}
+\caption{On x86, coe Links Are Atemporal}
+\label{fig:On x86; coe Links Are Atemporal}
+\end{center}
+\end{figure}
+
+Not only is this plausible, but as you can see in
+Figure~\ref{fig:On x86; coe Links Are Atemporal},
+it actually happens, and frequently.
+The store-to-store latency is measured from the end of the ``winning''
+store to the beginning of the non-winning store that started latest.
+As you can see, most of the measurements show that the winning store
+was not the last store, even on x86.
+
+\begin{figure}[tb]
+\begin{center}
 \resizebox{5in}{!}{\includegraphics{coe-out-1}}
@@
 
-However, Figure~\ref{fig:On x86, coe Links Are Partially Ordered}
+@@@ However, Figure~\ref{fig:On x86, coe Links Are Partially Ordered}
 shows that the order of reads of each value at each thread are
@@
 Figure~\ref{fig:On x86, fre Links Are Atemporal}
-plots a histogram of the elapsed time from the last load to return the
-old value to the store that provided the new value.
-Note well that most of the data falls into negative time, that is to say,
-the last load of the old value executes about 50 timestamp periods (about
-20~nanoseconds) \emph{after} the store that overwrote that same old value.
+plots a histogram of the elapsed time from the beginning of load to
+return the old value to the end of the store that provided the new value.
+Note well that most of the data falls into negative time, that is to
+say, most frequently, the last load of the old value executes about 60
+timestamp periods (about 30~nanoseconds) \emph{after} the store that
+overwrote that same old value.
 That is, fre links can and do go backwards in time.

### 2024-01-31 ootavacuous.tex
IPC appendix: Wordsmithing
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 store to the beginning of the non-winning store that started latest.
-As you can see, most of the measurements show that the winning store
-was not the last store, even on x86.
+Every data point on the negative x-axis represents a group of runs
+in which the winning store was not the last store.
+Thus, the winning store not being the last store is quite common, even
+on x86.
 
@@
 
-@@@ However, Figure~\ref{fig:On x86, coe Links Are Partially Ordered}
+However, Figure~\ref{fig:On x86, coe Links Are Partially Ordered}
 shows that the order of reads of each value at each thread are
-consistent with a number of global orders, as required.
+consistent with a number of global orders, as required.\footnote{
+	Taken on a 16-CPU x86 laptop to avoid an unreadable
+	diagram containing 79 bubbles.}
 Figure~\ref{fig:On Power5, coe Links Are Also Partially Ordered}
@@
 Note well that most of the data falls into negative time, that is to
-say, most frequently, the last load of the old value executes about 60
-timestamp periods (about 30~nanoseconds) \emph{after} the store that
-overwrote that same old value.
+say, in the most commonly occurring case, the last load of the old value
+executes about 60 timestamp periods (about 30~nanoseconds) \emph{after}
+the store that overwrote that same old value.
 That is, fre links can and do go backwards in time.

### 2024-01-31 ootavacuous.tex
FYI, possible progress on language-level OOTA problem
On Mon, Jan 29, 2024 at 04:38:00PM -0800, Paul E. McKenney wrote:
> I have the pen for the appendixes, and you have the pen for everything
> else.  There will be inevitable overlap, the bibliography if nothing else,
> but that is what "git merge" is for.
>
> So for right now, I am just taking your patches.  At some point, I might
> complain about things like "Another way of saying this is that real-world
> CPUs do not speculate stores", but let's get the overall shape in place
> before such quibbles.  For one thing, such a quibble might be

@@
 	Another way of saying this is that real-world CPUs do not
-	speculate stores.}
+	make their stores visible to other CPUs until those stores
+	are no longer speculative.}
 the stores in lines~7 and~13 cannot take place until the CPU
@@
 The idea has been formalized by defining an OOTA cycle as a cycle
-in sdep $\cup$ rf [@@@ Reference needed],
+in sdep $\cup$ rf~\cite{PaulEMcKenney2014OOTA},
+% @@@ This citation credits the formulation to Ali Sezgin, third author
+% @@@ of N4323 ("Out-of-Thin-Air Execution is Vacuous").
 where sdep is the set of semantic dependencies within
@@
 \noindent
+(Here and below, we have written shared-variable accesses without
+annotations, for brevity.  Please imagine they are all relaxed atomic.)
+
 As long as \co{z} is zero, and all else being equal, changes in the
@@
 This makes OOTA more complicated than a simple cycle but
-we will continue to refer to it as an OOTA cycle out of habit.
+we will continue to refer to OOTA cycles out of habit.
 Note that this new definition includes and generalizes the earlier

### 2024-02-01 ootavacuous.tex
Add cbmc example for C++ non-determinism
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-01-31 ootavacuous.tex
@@
 
+\subsection{Complex Semantic Dependencies}
+\label{app:Complex Semantic Dependencies}
+
+Listing~\ref{lst:Example Code for Evaluation of sdep}
+shows a \co{herd7} thread \co{P0()}.
+Is there a semantic dependency between the load from \co{x} on
+line~3 to the store to \co{y} on line~9?
+Answering this question requires solving a mathematical equation,
+not all of which currently have known solutions.
+However, this section will use a relatively simple equation combined
+with restrictions on domain and range that permit \co{cbmc} to do
+the job.
+
 \begin{listing}[tbp]
@@
 \end{listing}
-
-Listing~\ref{lst:Example Code for Evaluation of sdep}
-shows a \co{herd7} thread \co{P0()}.
-Is there a semantic dependency between the load from \co{x} on
-line~3 to the store to \co{y} on line~9?
 
@@
 	On Debian-based distributions, via \co{sudo apt install cbmc}.}
+
+\subsection{C++ Non-Determinism}
+\label{app:C++ Non-Determinism}
+
+The C++ standard allows non-determinism, for example, the order of
+invocation of side-effects in some circumstances.
+Listing~\ref{lst:Code for cbmc-Based Evaluation of Non-Deterministic sdep}
+shows an example leveraging side-effects in the parameter list shown on
+line~37.
+
+\begin{listing}[tbp]
+@@ DisplayLitmus cbmc/sdep-nondet.c @@
+\caption{Code for cbmc-Based Evaluation of Non-Deterministic sdep}
+\label{lst:Code for cbmc-Based Evaluation of Non-Deterministic sdep}
+\end{listing}
+
+Running \co{cbmc} on this as follows:
+
+\begin{quote}
+\scriptsize
+\begin{verbatim}
+cbmc -DCBMC cbmc/sdep-nondet.c
+\end{verbatim}
+\end{quote}
+
+Gives the result \co{VERIFICATION SUCCESSFUL}, showing that \co{cbmc}
+does not model this non-determinism.
+Which is not surprising, given that many of the safety-critical uses
+to which \co{cbmc} is put would prohibit such non-determinism in their
+code bases.
 
@@
 % @@@ TODO:
-% Create nice bell graph for coe, or at least try!
 % Construct interleaved OOTA examples (see sheets of paper).
-% Check cmbc on nondeterministic i++.
 % Use LKMM as check for dependencies?

### 2024-02-01 cbmc/sdep-nondet.c
Add the source code for the non-determinism cbmc example
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -1,0 +1,54 @@
+/*
+ * Demonstrate cbmc-based computation of sdep given C++ nondeterminism.
+ *
+ * Usage: cbmc -DCBMC sdep-nondet.c
+ *
+ * This can also be run as a stand-alone C program:
+ *	./sdep-nondet val
+ *
+ * Where the command-line argument gives the value of val for each
+ * iteration of the loop.  The same additional preprocessor variables
+ * can be used when compiling.  You can also specify -DCBMC to exercise
+ * CBMC-specific code paths, sort of, anyway.
+ */
+
+#include <stdlib.h>
+#include <stdio.h>
+#include <assert.h>
+
+#ifdef CBMC
+int get_arg(char *arg)
+{
+	return (int)(long)arg;
+}
+#else
+int get_arg(char *arg)
+{
+	return strtol(arg, NULL, 0) & 0x7;
+}
+#endif
+
+int div(int x, int y)
+{
+	return x / y;
+}
+
+int main(int argc, char *argv[])
+{
+	int i;
+	int val = 0;
+	int z[2] = { 0, 0 };
+
+	if (argc > 1)
+		val = get_arg(argv[1]);
+	if (val < 0)
+		val = -val;  // Avoid divide-by-zero UB
+	printf("argc = %d, val = %d\n", argc, val);
+	for (i = 0; i < 2; i++) {
+		int r1 = val;
+
+		z[i] = div(++r1, ++r1);
+		printf("i = %d, r1 = %d, z[i] = %d\n", i, r1, z[i]);
+	}
+	assert(z[0] == z[1]);
+}

### 2024-02-01 ootavacuous.tex
Add citation to Hans Boehm 2020 UB working paper
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
FYI, possible progress on language-level OOTA problem
This fixes up the indenting issues surrounding quoted material.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 and \co{r2} having the value 42 as follows:
-
 \begin{enumerate}
@@
 \end{enumerate}
-
-\noindent
 Current C++ implementations can and do exhibit this reordering behavior.
@@
 (\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
-
 \begin{quote}
@@
 \end{quote}
-
-\noindent
 The standard's prohibition of OOTA is of course important, but those of us
@@
 code:
-
 \begin{quote}
@@
 \end{quote}
-
-\noindent
 This is an elaboration of \co{thread2()} from
@@
 had instead been as follows:
-
 \begin{quote}
@@
 \end{quote}
-
-\noindent
 Here line~3 has been optimized away in favor of line~4 storing the same value
@@
 Consider for example:
-
 \begin{quote}
@@
 \end{quote}
-
-\noindent
 (Here and below, we have written shared-variable accesses without
@@
 we will say that an execution is an instance of OOTA if in that execution:
-
 \begin{quote}
@@
 \end{quote}
-
-\noindent
 This makes OOTA more complicated than a simple cycle but
@@
 Consider the following:
-
 \begin{quote}
@@
 \end{quote}
-
-\noindent
 There appear to be semantic dependencies from \co{y} to \co{x} and from \co{z}
@@
 Consider the following \co{if} statement:
-
 \begin{quote}
@@
 \end{quote}
-
-\noindent
 Here there is a semantic dependency from \co{x}, but in some executions
@@
 The combination of OOTA cycles and UB has proven especially
-challenging~\cite{DavidGoldblatt2019NoElegantOOTAfix}.
+challenging~\cite{DavidGoldblatt2019NoElegantOOTAfix,HansBoehm2020UBalternatives}.
 

### 2024-02-02 ootavacuous.tex
Add discussion of Hans Boehm's promote-to-acquire UB proposal
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 of the proposed \co{std::observable()}.
+
+\subsection{Atomic Stores From Undefined Behavior}
+\label{sec:Atomic Stores From Undefined Behavior}
+
+Another proposal to the C++ standards
+committee~\cite{HansBoehm2020UBalternatives}
+proposes (among other alternatives) that any atomic store resulting
+from undefined behavior have at least release semantics.
+This would prevent such a store from propagating undefined behavior
+backwards within that store's thread.
+In some cases, this could also prevent this undefined behavior from
+propagating back through other threads.
+
+However, it might well force C++ implementations to promote a surprisingly
+large fraction of a program's relaxed stores to release.
 

### 2024-02-03 ootavacuous.tex
Add discussion about load hoisting to invented-loads appendix
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \end{enumerate}
+
+There are situations where inventing loads from non-atomic objects
+is customary, for example, when hoisting a load out of a loop.
+In this case, implementations might hoist the load from x out of
+the loop:
+
+\begin{quote}
+\begin{verbatim}
+do_something_with(x);
+for (i = 0; i < limit i++)
+	y[i] += x;
+\end{verbatim}
+\end{quote}
+
+A C++ implementation might act as if the program had instead been this:
+
+\begin{quote}
+\begin{verbatim}
+r1 = x;
+do_something_with(r1);
+for (i = 0; i < limit i++)
+	y[i] += r1;
+\end{verbatim}
+\end{quote}
+
+Note that passing \co{x} to the call to \co{do_something_with()} is important
+because this allows the C++ implementation to assume that accesses to \co{x}
+will not result in a data race.
+Without this assumption, the above transformation might introduce a data
+race when \co{limit} is less than or equal to zero.
+
+But if \co{x} is a non-volatile atomic, then there are no worries about
+data races.
+In theory, the implementation could make the above transformation without
+worry, especially given that the value loaded from \co{x} is discarded in
+the case where \co{limit} is less than or equal to zero.
+Except that the data-race worry is replaced by a cache-miss worry.
+After all, optimizations are supposed to speed things up.
+This means that a C++ implementation might be wise to adhere to the
+same restriction that avoids data races for loads from non-atomic
+variables in order to avoid expensive cache misses.
 

### 2024-02-04 ootavacuous.tex
FYI, possible progress on language-level OOTA problem
More additions: the section on executions.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\section{OOTA, Semantic Dependencies, and Executions}
-\label{sec:OOTA; Semantic Dependencies; and Executions}
+\section{OOTA and Semantic Dependencies}
+\label{sec:OOTA and Semantic Dependencies}
 
@@
 
-\subsection{What Exactly is an Execution?}
-\label{sec:What Exactly is an Execution?}
+\section{What is an Execution?}
+\label{sec:What is an Execution?}
 
 @@@ Roadmap
+
+\subsection{Abstract Executions}
+\label{sec:Abstract Executions}
+
+The C++ standard describes the execution of a program in terms of
+``a parameterized nondeterministic abstract machine'' in 4.1.2p1
+(\co{intro.abstract}).
+This description specifies how the abstract machine carries out the
+operations of a source program in great, but not complete, detail:
+\begin{itemize}
+\item	Some of the abstract machine's characteristics are
+	implementation defined, including things like the number of
+	bits in the various integer types
+	or whether the \co{char} type is signed.
+\item	Some aspects of an execution are unspecified or nondeterministic,
+	including things like the order of evaluation of the operands
+	of most binary operators or of the arguments in a function call.
+	Implementations may choose from a set of allowed behaviors.
+\item	Some actions are deemed to have undefined results; the standard
+	says essentially nothing about programs that can give rise to
+	undefined behavior.
+\item	Asynchronous actions (i.e., signal handlers) are largely ignored.
+\item	Input and output are not described in any detail.
+\end{itemize}
+In addition to these points, the standard does not specify which store
+an atomic load must read from, beyond requiring that the overall
+pattern of loads and stores be consistent with the C++ memory model.
+We assume that programs will not indulge in any other computations that
+could be expected to vary spontaneously from one execution to another,
+such as basing a dependency on the time of day or a process ID.
+
+The implementation-defined aspects can affect whether or not an
+abstract execution contains a semantic dependency.
+As an example consider the following, where the type of \co{c} is \co{char}:
+\begin{quote}
+\begin{verbatim}
+y = x * (c >= 0);
+\end{verbatim}
+\end{quote}
+Here \co{y} is semantically dependent on \co{c} in executions
+for which \co{char} is a signed type,
+but not those for which it is unsigned.
+
+The same is theoretically true for the nondeterministic aspects of an
+execution.
+Consider this example, with \co{i} initially zero:
+\begin{quote}
+\begin{verbatim}
+r1 = ++i / ++i;
+x = r1 * z;
+\end{verbatim}
+\end{quote}
+Because the order of evaluation of the division's arguments is undetermined,
+the value calculated for \co{r1} might be zero ($1/2$ truncated) or two
+($2/1$).
+In the former case there is no semantic dependency from \co{z} to \co{x},
+but in the latter case there is.\footnote{
+	Thanks to Peter Sewell for pointing out this possibility.}
+(According to the current version of the standard, conflicting side effects
+in unsequenced expressions constitute undefined behavior.
+However there are proposals to make them defined in both
+C++ and C,\footnote{
+	\url{https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2016/p0145r3.pdf}
+	\url{https://www.open-std.org/jtc1/sc22/wg14/www/docs/n3203.htm}}
+which would render this theoretical possibility quite real.)
+
+The abstract executions we use will be fully specified.
+This means that all the missing information must be supplied:
+the implementation-defined characteristics, the selections for the
+nondeterministic pathways, and most notably, for each load, the store
+from which it reads and the value of the load.
+We ignore issues of signal handlers and I/O;
+in any case our litmus-test programs don't use them.
+The totality of this information---along with the program's source
+code, of course---determines within each thread a unique, linearly
+ordered series of steps to be carried out by the abstract machine.
+(However, there is no ordering relation between steps carried out
+in different threads.  Even if a load in one thread reads from a
+store in another thread, we don't require the store to come before
+the load.)
+
+With the compiler-based implementations we are considering,
+the choices for the nondeterministic pathways are ``frozen'' into the
+machine-code executable file and thus are completely determined
+at runtime.
+A consequence of this is that if two abstract executions of the same
+thread under the same implementation agree on the values obtained by
+the load operations during their first $N$ steps then they will agree
+in every respect during those steps, although they may diverge later.
+
+\subsection{Hardware Executions}
+\label{sec:Hardware Executions}
+
+When a given computer executes the machine code in a file, the outcome
+has historically been much more well defined than the executions of the
+C++ abstract machine.
+The hardware's behavior is typically specified with great precision by
+the designer or manufacturer, and there are formal, executable memory models
+describing exactly what patterns of loads and stores can occur.
+Thus, leaving aside questions of asynchronous interrupts and system
+calls, the behavior of a CPU executing a particular thread within a
+program is entirely determined by the values obtained by the various
+load-from-memory instructions.
+
+For this reason, the hardware executions we use will comprise (along
+with machine code being run) the computer architecture and for each
+load instruction, the store instruction from which it reads and the
+value obtained.
+At this level, the fact that the original program was in C++ is
+irrelevant; the same concepts apply to the execution of a program in
+any compiled language.
+
+A computer may execute the instructions in a thread out of order.
+The architecture specifies the extent to which this may happen, and it
+also specifies circumstances under which some pairs of instructions
+must be executed in order.
+Nevertheless, we will consider an execution to be determined by the
+values obtained by its loads.
+As with abstract executions, if two hardware executions of the same
+thread on the same type of computer agree on the values obtained by
+the load instructions during their first $N$ steps then they will
+agree in every respect during those steps, although they may diverge
+later.
+
+\subsection{Relation Between Abstract and Hardware Executions}
+\label{sec:Relation Between Abstract and Hardware Executions}
+
+The C++ standard requires that for any valid implementation, when a
+program runs its observable behavior must be the same as that of some
+abstract execution of the source code given the same input (in the
+absence of any abstract executions containing undefined behavior).\footnote{
+	This requirement is the standard's ``as-if'' rule.}
+This means:\begin{itemize}
+\item	The program's output must be the same as that of the abstract
+	execution.
+\item	Volatile accesses ``are evaluated strictly according to the
+	rules of the abstract machine'' [4.1.2p6.1 (\co{intro.abstract})].
+\item	(There is a condition on the interleaving of input and output,
+	which does not matter for our purposes.)
+\end{itemize}
+We will say that the abstract execution is \emph{realized by} the
+hardware execution.
+
+Under any particular implementation,
+a single program can have many different abstract executions,
+varying in their decisions about which store each load reads from
+and thus the value obtained.
+It's worth noting, however, that not all the possible abstract executions
+of a program need be realizable by the machine-code executable file
+produced by that implementation.
+In fact, we will see that \emph{none} of the possible OOTA executions
+allowed by the strict C++ abstract machine will ever be realized
+by the executables produced by many compilers,
+
+Exactly what the standard's restriction on volatile accesses means
+isn't entirely clear.
+The handling of volatiles, as understood by compiler developers, has
+been described as more folklore or a gentlemen's agreement than
+anything else.
+We will take it to imply that a compiler may not
+invent, omit, or merge volatile accesses, in a sense to be discussed below.
 

### 2024-02-05 ootavacuous.tex
FYI, possible progress on language-level OOTA problem
Some more small changes to the initial parts of the paper.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 In broad terms,
-OOTA occurs when a group of threads load from each others' stores
+OOTA occurs theoretically when a group of threads load from each others' stores
 and each thread's store depends on the value returned by that thread's load.
@@
 Semantic dependencies are only one type of link in an OOTA cycle.
-The other type extends from a given store to the load that returns the
+The other type extends from a given store to a load that returns the
 value stored.
@@
 	and properies of rf, rfe, and rfi.}
+
 This is a fine definition and is consistent with the words in the C++
@@
 Here line~3 has been optimized away in favor of line~4 storing the same value
-in \co{r2} that was stored to \co{y} by line~2.
+to \co{z} that was stored to \co{y} by line~2.
 And given that the load from \co{y} no longer exists, it cannot possibly
@@
 
-Since we no longer regard sdep as always relating a single load to a store,
+Since we can no longer regard sdep as always relating a single load to a store,
 the notion of a cycle involving sdep appears problematic.
@@
 it extends to \co{y} and in others to \co{z}.
-This is an example of the load affecting not the value of
+This is an example of a load affecting not the value of
 a given store, but rather whether or not that store is executed at all.
@@
 ``a parameterized nondeterministic abstract machine'' in 4.1.2p1
-(\co{intro.abstract}).
+(\co{[intro.abstract]}).
 This description specifies how the abstract machine carries out the
@@
 
-The same is theoretically true for the nondeterministic aspects of an
-execution.
+The same is true for the nondeterministic aspects of an execution.
 Consider this example, with \co{i} initially zero:
@@
 \begin{verbatim}
-r1 = ++i / ++i;
+int foo(int a, int b)
+{
+   return a / b;
+}
+
+r1 = foo(++i, ++i);
 x = r1 * z;
@@
 \end{quote}
-Because the order of evaluation of the division's arguments is undetermined,
+Because early C~implementers could not come to agreement, the standard
+does not specify the order of evaluation of function arguments, so
 the value calculated for \co{r1} might be zero ($1/2$ truncated) or two
@@
 	Thanks to Peter Sewell for pointing out this possibility.}
+
 (According to the current version of the standard, conflicting side effects
-in unsequenced expressions constitute undefined behavior.
-However there are proposals to make them defined in both
-C++ and C,\footnote{
+in unsequenced subexpressions constitute undefined behavior,
+although there are proposals to make them defined in both C++ and C.\footnote{
 	\url{https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2016/p0145r3.pdf}
 	\url{https://www.open-std.org/jtc1/sc22/wg14/www/docs/n3203.htm}}
-which would render this theoretical possibility quite real.)
+Nevertheless, the example above is allowed because the order of evaluation
+of arguments to a function call is ``indeterminately sequenced''
+(7.6.1.3p7 \co{[expr.call]}) rather than unsequenced, a subtle distinction.)
+
+\medskip
 
@@
 We ignore issues of signal handlers and I/O;
-in any case our litmus-test programs don't use them.
+in any case our litmus-test programs don't use them
+(but see the discussion of volatile loads in
+section~\ref{sec:Relation Between Abstract and Hardware Executions} below).
 The totality of this information---along with the program's source
@@
 ordered series of steps to be carried out by the abstract machine.
-(However, there is no ordering relation between steps carried out
+However, there is no ordering relation between steps carried out
 in different threads.  Even if a load in one thread reads from a
 store in another thread, we don't require the store to come before
-the load.)
+the load.
 
@@
 	This requirement is the standard's ``as-if'' rule.}
-This means:\begin{itemize}
+This means:
+\begin{itemize}
 \item	The program's output must be the same as that of the abstract
@@
 \item	Volatile accesses ``are evaluated strictly according to the
-	rules of the abstract machine'' [4.1.2p6.1 (\co{intro.abstract})].
+	rules of the abstract machine'' (4.1.2p6.1 \co{[intro.abstract]}).
 \item	(There is a condition on the interleaving of input and output,
@@
 anything else.
-We will take it to imply that a compiler may not
-invent, omit, or merge volatile accesses, in a sense to be discussed below.
+To help guide C++ users and implementers, the standard adds these
+suggestive comments (9.2.9.2p5 and~6 \co{[dcl.type.cv]}):
+\begin{quote}
+	The semantics of an access through a volatile glvalue are
+	implementation-defined.
+
+	\co{volatile} is a hint to the implementation to avoid aggressive
+	optimization involving the object because the value of the object
+	might be changed by means undetectable by an implementation.
+\end{quote}
+Following the folklore, we interpret these comments to mean that
+volatile accesses should be treated as a form of I/O,
+implying that volatile loads can return unpredictable values.
+They don't necessarily read from stores (in contrast to non-volatile loads,
+which must return the value of the store they read from).
+As a corollary, the rfe relation does not apply to volatile loads
+and hence the accesses in an OOTA cycle must not be volatile.
+We also understand that a compiler may not
+invent, omit, merge, or reorder volatile accesses,
+in a sense to be discussed below.
 

### 2024-02-05 ootavacuous.tex
FYI, possible progress on language-level OOTA problem
Updates reflecting my new understanding of volatile accesses and the
order of evaluation of function arguments.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 In broad terms,
-OOTA occurs theoretically when a group of threads load from each others' stores
+OOTA occurs when a group of threads load from each others' stores
 and each thread's store depends on the value returned by that thread's load.
@@
 Semantic dependencies are only one type of link in an OOTA cycle.
-The other type extends from a given store to a load that returns the
+The other type extends from a given store to the load that returns the
 value stored.
@@
 	and properies of rf, rfe, and rfi.}
-
 This is a fine definition and is consistent with the words in the C++
@@
 Here line~3 has been optimized away in favor of line~4 storing the same value
-to \co{z} that was stored to \co{y} by line~2.
+in \co{r2} that was stored to \co{y} by line~2.
 And given that the load from \co{y} no longer exists, it cannot possibly
@@
 
-Since we can no longer regard sdep as always relating a single load to a store,
+Since we no longer regard sdep as always relating a single load to a store,
 the notion of a cycle involving sdep appears problematic.
@@
 it extends to \co{y} and in others to \co{z}.
-This is an example of a load affecting not the value of
+This is an example of the load affecting not the value of
 a given store, but rather whether or not that store is executed at all.
@@
 ``a parameterized nondeterministic abstract machine'' in 4.1.2p1
-(\co{[intro.abstract]}).
+(\co{intro.abstract}).
 This description specifies how the abstract machine carries out the
@@
 \item	Volatile accesses ``are evaluated strictly according to the
-	rules of the abstract machine'' (4.1.2p6.1 \co{[intro.abstract]}).
+	rules of the abstract machine'' [4.1.2p6.1 (\co{intro.abstract})].
 \item	(There is a condition on the interleaving of input and output,

### 2024-02-07 ootavacuous.tex
FYI, possible progress on language-level OOTA problem
Some more instructive examples of semantic dependencies.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-02-05 ootavacuous.tex
@@
 
+\subsubsection{Semantic Dependencies Not Affected by \co{if} Statements}
+\label{sec:Semantic Dependencies Not Affected by if Statements}
+
+Compare this example to the previous one:
+\begin{quote}
+\begin{verbatim}
+if (x > 0)
+    y = 42;
+else
+    y = 42;
+\end{verbatim}
+\end{quote}
+Because the stores executed on each arm of the \co{if} statement write
+identical values to identical addresses, one could equally well regard
+the two statements as performing two different stores or as performing
+for all intents and purposes a single store, independent of x.
+Reasonable C++ implementations might disagree on this matter and
+therefore on whether or not the example has a semantic dependency.
+It is the implementations's choice.
+
+\subsubsection{Semantic Dependencies and Matching Up Stores}
+\label{sec:Semantic Dependencies and Matching Up Stores}
+
+Suppose we take the view that the previous example involves only one
+store.
+This opens up the door to greater complexity:
+\begin{quote}
+\begin{verbatim}
+if (x > 0) {
+    L1: y = 42;
+} else {
+    y = 53;
+    y = 42;
+}
+\end{verbatim}
+\end{quote}
+Consider an execution in which \co{x} is greater than zero, so the
+statement labeled \co{L1} runs.
+Is it semantically dependent on \co{x}?
+The answer isn't immediately clear.
+If the other arm of the \co{if} is taken then the same store of 42 to
+\co{y} occurs, but 53 is written before it.
+Which of these two stores should be compared with the store in \co{L1}?
+
+One way to cut the Gordian knot is to match up the stores by the order
+they occur:
+Since \co{L1} is the first store to \co{y} in its arm of the \co{if}
+statement, it should be matched up with the first store to \co{y} in
+the other arm.
+Those two stores write different values so there is a semantic
+dependency.
+
+On the other hand, a compiler may decide to drop the \co{y =
+53} store entirely, leaving it out of the machine code,
+on the grounds that it's always possible for the two adjacent stores
+to \co{y} to execute in such quick succession that no other thread
+manages to read the value 53 before it gets overwritten with 42.
+If the compiler does this then the first store to \co{y} in that
+arm of the \co{if} statement \emph{would} agree with the store in
+\co{L1}, and so there would not be a semantic dependency.
+Once again, the decision is up to the implementation.
+
+\medskip
+
 We have seen several examples showing that semantic dependencies may vary
-according to the execution.
-This raises the question: What exactly is an execution?
+according to the execution and even the implementation.
+This raises some questions, of which the first is:
+What exactly is an execution?
 
@@
 \begin{verbatim}
-y = x * (c >= 0);
+y = (c >= 0);
 \end{verbatim}
@@
 	the need for precise software-visible exceptions.
-\item	Some systems might prefer to immediately shut down, forbidding
-	software from continuing after an uncorrectable ECC error,
-	perhaps due to safety considerations.
+\item	Some systems might prefer to immediately shut down in response to
+	an uncorrectable ECC error, perhaps due to safety considerations.
+	This would entirely avoid the need for software-visible
+	ECC-related exceptions.\footnote{
+		One motivation for immediate shutdown is that if there
+		are uncorrectable errors, there might soon be errors
+		that translate one valid bit pattern to another valid
+		bit pattern, which could result in a lack of safety.}
 \item	Someone might come up with a clever way of correcting ECC errors

### 2024-02-08 ootavacuous.tex
FYI, possible progress on language-level OOTA problem
Add a section on C++ compilers.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-02-07 ootavacuous.tex
@@
 
+The inability of rfi links to act as temporal constraints is not the
+only, or even the main, weakness in the intuitive argument against
+OOTA cycles.
+The primary difficulty lies in the fact that the code transformations
+performed by optimizing compilers can destroy dependencies, including
+semantic ones (depending on one's definition).
+That is, even though the potential for a dependency exists in the
+source code for a thread, there may be no dependency in the machine
+code produced by a compiler.
+There would then be no constraint forcing the implementation to execute
+the thread's store later in global time than the load it supposedly depends on,
+and thus no impediment to the occurrence of an OOTA cycle.
+We will see examples of this destruction in
+Sections~\ref{sec:Global Optimization Can Destroy Dependencies}
+and~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies} below.
+
 \subsection{Properties of Semantic Dependencies}
@@
 to \co{x}.
-However, if the compiler somehow knows that \co{y} is
+However, if the implementation somehow knows that \co{y} is
 always equal to \co{z} at this point then there is no semantic dependency;
-the compiler can act as if the statement were simply ``\co{x = 0;}''.
-We leave aside the question of how the compiler would know this,
+the implementation can act as if the statement were simply ``\co{x = 0;}''.
+We leave aside the question of how the implementation would know this,
 given that \co{y} and \co{z} cannot be updated simultaneously
 and are subject to change at any time by
-other threads (a point we will return to).
+other threads (a point we will return to in
+section~\ref{sec:Global Optimization Can Destroy Dependencies}).
 
@@
 invent, omit, merge, or reorder volatile accesses,
-in a sense to be discussed below.
+in a sense to be discussed in
+section~\ref{sec:Volatile and Partially Volatile Accesses} below.
+
+\section{C++ Compilers}
+\label{sec:C++ Compilers}
+
+A complete C++ implementation consists of much more than just a compiler.
+Among other things, for example, it might have a collection of
+\co{.h} header files, a linker, runtime libraries, and a dynamic loader.
+Nevertheless, for our purposes the compiler is the most important
+component because it is what primarily determines the translation from
+a C++ source program to directly executable machine code.
+We will therefore use the terms ``compiler'' and ``implementation''
+somewhat interchangeably.
+
+@@@ Fill in roadmap when sectioning is finalized.
+
+\subsection{Users Influence the Behavior of Compilers}
+\label{sec:Users Influence the Behavior of Compilers}
+
+The exact definition of a computer language is subject to some debate,
+with standards, implementations, and users all having some degree of
+influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
+each of which is subject to change over time.
+In areas that are not well settled or where users might reasonably
+want to resist the dictates of the standard,
+compilers often provide switches to override their default behaviors.
+An example is GCC's \co{-funsigned-char} command-line argument,
+which causes it to treat variables of type \co{char} as unsigned.
+More examples of user control over language semantics are given in
+Appendix~\ref{app:User Influence Over Language Semantics}.
+
+We will consider these user-specified compiler switch settings to fall
+within the implementation-defined parameters of the C++ abstract machine.
+They should be provided, implicitly or explicitly, as part of any
+abstract execution.
+
+\subsection{Global Optimization Can Destroy Dependencies}
+\label{sec:Global Optimization Can Destroy Dependencies}
+
+Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
+page~\pageref{lst:Simple OOTA},
+in which \co{thread1()} loads the value of \co{x} and stores it in
+\co{y} while \co{thread2()} does the reverse.
+A globally optimizing compiler given that program might
+transform it to the following before translating it into machine code,
+if the compiler is sufficiently perverse:
+\begin{quote}
+\begin{verbatim}
+ 1 atomic<int> x(0);
+ 2 atomic<int> y(0);
+ 3
+ 4 void thread1()
+ 5 {
+ 6   int r1 = 42;
+ 7   y.store(r1, memory_order_relaxed);
+ 8 }
+ 9
+10 void thread2()
+11 {
+12   int r2 = 42;
+13   x.store(r2, memory_order_relaxed);
+14 }
+\end{verbatim}
+\end{quote}
+The loads previously on lines~6 and~12 have been replaced by constants.
+Such a transformation complies with the strict C++ standard,
+even though the resulting executable file would produce an unintuitive
+OOTA outcome every time it runs!
+
+The only justification a compiler could have for generating output like
+this is that it knows exactly what accesses will be performed by both
+threads, and therefore it knows that it will not violate the strict C++
+memory model by assuming each thread's load reads from the other's
+store.
+A similar justification can underlie the reasoning in
+section~\ref{sec:Semantic Dependencies Affected by Cross-Thread Optimizations};
+in principle an analysis of the complete program could lead a compiler
+to conclude that \co{y} will always be equal to \co{z} whenever a
+particular \co{y - z} expression is evaluated, allowing the compiler
+to replace the expression with a constant \co{0}.
+
+By contrast, a compiler that analyzes only one thread at a time when
+performing its optimizations and other code transformations will not
+have this kind of global knowledge, and consequently it would not
+perform the OOTA-ful transformation shown here.
+
+Because we seek to find characteristics of compilers that will
+guarantee the absence of OOTA behavior in the machine code they generate,
+we will for now confine our attention to compilers that analyze only
+one thread of source code at a time.
+In more precise terms, we want the compilers under consideration
+always to generate the same machine-code output for threads having
+the same source code, regardless of the rest of the programs
+those threads are contained in.
+Later on we will return to globally optimizing compilers.
+
+\subsection{Inventing Atomic Loads Can Destroy Semantic Dependencies}
+\label{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}
+
+Consider this code, in which the final values of \co{r1} and~\co{r2}
+are observable:
+\begin{quote}
+\begin{verbatim}
+ 1  int r1 = (x != 0);
+ 2  int r2 = (y != 0);
+ 3  z = (r1 == r2);
+\end{verbatim}
+\end{quote}
+It is clear that the store to \co{z} semantically depends on the load
+from \co{y}, because the value of \co{z} will change whenever \co{y}
+changes between zero and nonzero (all else being equal).
+
+However, an especially devious compiler might transform the source into
+the following form before translating it to machine code:
+\begin{quote}
+\begin{verbatim}
+ 1  int r1;
+ 2  int r1a = (x != 0);
+ 3  int r1b = (x != 0);
+ 4  int r2 = (y != 0);
+ 5  if (r1a != r1b) {
+ 6      r1 = r2;
+ 7      z = 1;
+ 8  } else {
+ 9      r1 = r1b;
+10      z = (r1 == r2);
+11  }
+\end{verbatim}
+\end{quote}
+The idea is that \co{r1a}, \co{r1b}, and \co{r2} can each be only zero or one,
+so if \co{r1a} and \co{r1b} differ then one of them must be equal to \co{r2}.
+In executions where this happens---because another thread writes to \co{x}
+between the two loads---the implementation can choose at runtime to use
+for \co{r1} whichever value agrees with \co{r2}, as shown on line~6.
+Then the value stored to \co{z} on line~7 will always be one,
+with no dependence on the value loaded from \co{y}.
+This example is discussed more fully in
+Appendix~\ref{app:Inventing Atomic Loads}.
+
+A noteworthy aspect of this transformation is that it invents an atomic load:
+The original form of the code reads \co{x} only once,
+whereas the transformed code reads it twice.
+Therefore we can rule out transformations like this one by insisting
+the compiler not invent atomic loads.
+In fact, we will require that atomic accesses be treated as
+``partially volatile'', in that the compiler is allowed to omit,
+merge, or reorder them but not invent them.
+
+Just what does this mean?
+
+\subsection{Volatile and ``Partially Volatile'' Accesses}
+\label{sec:Volatile and Partially Volatile Accesses}
+
+Declaring objects to be volatile is a way for the programmer to
+indicate that the hardware should perform all accesses to these
+objects exactly as written in the source code (perhaps because they
+represent memory-mapped device registers or DMA buffers rather than
+normal memory locations).
+In any event, we expect compilers' translations of volatile-object accesses
+into machine code to be as close to verbatim as possible.
+
+To express this idea in more formal terms, we will augment the
+requirements for a hardware execution $H$ to realize an abstract
+execution $A$.
+Each realization must include a map from the set of accesses of
+volatile objects in $A$ to the set of instructions in $H$ that access
+these objects, having the following properties:
+\begin{itemize}
+\item	The map connects accesses of the same type (loads to loads
+	and stores to stores).
+\item	The map connects accesses in a thread of $A$ to accesses in
+	the corresponding thread of $H$.
+\item	The map is value-preserving: The value of an access in $A$ must be
+	the same as the value of the access it maps to in $H$.
+\item	The map is order-preserving: Two accesses in the same thread
+	of $A$ must map to accesses occurring in the same order in $H$.
+	(In other words, the compiler may not reorder accesses
+	to volatile objects.)
+\item	The map is onto: For every access in $H$ to a volatile object
+	there must be an access in $A$ that maps to it.
+	(In other words, the compiler may not invent accesses to
+	volatile objects.)
+\item	The map is one-to-one: Different accesses in $A$ map to
+	different accesses in $H$.
+	(In other words, the compiler may not merge accesses to
+	volatile objects.)
+\item	The map is total: Every access in $A$ to a volatile object
+	maps to some access in $H$.
+	(In other words, the compiler may not omit accesses to
+	volatile objects.)
+\end{itemize}
+Most of these are direct consequences of the fact that volatile
+accesses are considered to be a form of I/O.
+
+For accesses to partially volatile objects, we remove some of the
+requirements and add a new one.
+We leave out the last two items; compilers are allowed to merge or
+omit accesses to partially volatile objects.
+And the requirement for order preservation is weakened; it applies
+only to pairs of accesses to the same object.
+Accesses to different objects may be reordered relative to each other.
+The new requirement is that the map must preserve the rf relation.
+That is, if load $R$ in $A$ reads from store $W$ and $R$ is not
+omitted then the instruction it maps to in $H$ must read from the
+instruction that $W$ maps to,
+and consequently $W$ must not be omitted either.
 

### 2024-02-08 ootavacuous.tex
FYI, possible progress on language-level OOTA problem
Add a section on hardware dependencies and instruction ordering.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 and consequently $W$ must not be omitted either.
+
+\section{Hardware Dependencies, Instruction Ordering, and the
+Fundamental Property}
+\label{sec:Hardware Dependencies, Instruction Ordering, and the
+Fundamental Property}
+
+@@@ Roadmap
+
+\subsection{Dependencies at the Hardware Level}
+\label{sec:Dependencies at the Hardware Level}
+
+Dependencies between machine instructions can be understood in terms
+of the flow of information within a CPU.
+Each instruction has a set of inputs and is the source of a set of outputs,
+some of which flow to the inputs for later instructions.
+A few examples:
+\begin{itemize}
+\item	An \co{add} instruction would typically have two inputs
+	(the register values to be added together) and two outputs
+	(the sum to be stored in a general-purpose register and some
+	flag bits---e.g., Zero, Carry, and Overflow---to be stored in
+	a status register).
+\item	A conditional branch instruction would have only one input:
+	the flag bits it tests in the status register.
+	A computed branch instruction would as input the register
+	containing the address to branch to.
+	For both, the output is the new address to be written to the
+	instruction-pointer register.
+\item	A memory-load instruction's input is the address to load from,
+	and its output is the value obtained by the load, to be stored
+	in a register.
+\item	A memory-store instruction's inputs are the value to store and
+	the address at which to store it; there are no outputs.
+\end{itemize}
+Of course, some of an instruction's inputs may be immediate constants
+taken from the instruction itself rather than outputs from some
+earlier instruction.
+Note: Hardware dependency analysis considers only the information
+flowing \emph{within} a CPU, not the information flowing between the
+CPU and memory, which is handled separately.
+
+Using this scheme, we say that an instruction $J$ is dependent on
+another instruction $I$ when any of $I$'s outputs flow into $J$'s inputs,
+perhaps indirectly via some intermediate instructions.
+Tracing back the flow of information between instructions, you can
+see that any hardware dependency must ultimately emanate from a load
+instruction or a thread's initial state (or possibly some sort of
+input instruction, but we will not consider that here).
+These are the only sources of truly new information in a thread.
+
+The concept is simple, but it is important because of the way
+dependencies affect instruction ordering in weakly ordered architectures.
+
+\subsection{Instruction Ordering}
+\label{sec:Instruction Ordering}
+
+A CPU may start executing an instruction speculatively, but at some
+point it must \emph{commit} to a decision either to definitely execute the
+instruction with some well-defined collection of inputs and outputs
+or else to abandon it.
+Thanks to the law of cause and~effect, a CPU is not able to commit an
+instruction until all of its inputs' sources have committed.
+(Or at least, all that can affect the action and outputs of the
+instruction.
+For instance, although no existing architectures do this, there is nothing
+in principle to prevent a CPU from committing a \co{multiply}
+instruction as soon as either one of its inputs has committed to the
+value zero.)
+The reason is obvious: An input is subject to change at any time until
+its source commits, the outputs aren't determined until the inputs
+are, and the instruction can't commit until its outputs are fully
+determined.
+
+This means that, with a few possible exceptions such as the
+multiply-by-zero example, an instruction can't commit until all the
+instructions it depends on have committed.
+And even taking the exceptions into account, if a change in the
+outputs of instruction $I$ would lead to a change in the inputs to
+instruction $J$ then $I$ must commit before $J$.
+Thus dependencies force instructions to commit in order, even on
+weakly ordered architectures.
+
+However, dependencies aren't the only mechanism capable of forcing
+ordering in hardware executions.
+The simplest of these is a load reading from a store in another
+thread, i.e., the rfe relation.
+As mentioned in section~\ref{sec:Simple OOTA Cycle} above, a CPU does
+not make the value of a store instruction available for other CPUs to
+load until the store commits or later.
+And after this happens, it takes some time for the information about
+the store to travel from that CPU to others, owing to the inescapable
+facts that processors have nonzero size and information cannot travel
+faster than the speed of light.
+Since the load instruction cannot commit until its output (the value it
+reads) is fully determined, it must commit after the store it reads
+from.
+
+A similarly straightforward mechanism for ordering is conditional or
+computed branches.
+An instruction executing after such a branch cannot commit
+until the CPU has committed to whether the branch will be taken and
+if taken, where it will branch to;
+until then the CPU cannot know whether the instruction should
+be executed at all.
+Therefore instructions following a conditional or computed branch
+instruction in an execution must commit after anything the branch
+depends on.
+(Again, there could theoretically be an exception for a branch that
+conditionally jumps to the immediately following instruction.
+We can ignore this issue by treating such branches as no-ops.)
+
+Yet another mechanism for ordering crops up when a load reads from a
+store in the same thread (the rfi relation).
+In this situation the load cannot commit before the sources for the
+store's inputs.
+Until the source for the store's value commits, the value of the load
+isn't fully determined and so the load cannot commit.
+And in general, until the source for store's address commits, it's not
+fully determined whether the store and load are for the same address.
+This argument doesn't hold if the store and load have the \emph{same}
+source for their addresses, but another factor is also at play.
+Until the address is fully determined, there is a possibility that it
+will turn out to be invalid (a NULL pointer, for example), causing the
+store instruction to encounter an addressing exception.
+If this should happen the load might not execute at all, and therefore
+it can't commit until the address's source has committed.
+
+\subsection{The Fundamental Property of Semantic Dependencies}
+\label{sec:The Fundamental Property of Semantic Dependencies}
+
+With the contents of the last few sections under our belts, we can
+formulate the Fundamental Property that we would like all semantic
+dependencies to satisfy.
+Note that this formulation makes sense only for implementations in
+which all atomic objects are treated as partially volatile.
+\begin{quote}
+Let $W$ be a store which, in some abstract execution $A$, semantically
+depends on loads $\{R_0$, \ldots,~$R_n\}$, and suppose that $W$ is not
+omitted in some hardware execution $H$ realizing $A$.
+Then for some $i$, load $R_i$ is not omitted and the hardware
+instruction it maps to is ordered before the instruction $W$ maps to.
+\end{quote}
+It's quite straightforward to show that for any implementation in which
+semantic dependencies satisfy the Fundamental Property, no OOTA
+abstract execution can be realized.
+
+Indeed, suppose that abstract execution $A$ is realized by hardware
+execution $H$ and $A$ has an OOTA cycle.
+This means there are stores $W_i$ semantically depending on loads
+$\{R_{i,j}\}$ where each of the loads reads from one of the $W_i$
+stores in a different thread.
+Let $W_i'$ and $R_{i,j}'$ be the hardware instructions these accesses
+map to, if they aren't omitted.
+Assuming that at least one of the stores is not omitted,
+one of the $W_i'$, call it $W_{i_0}'$, must commit first.
+By the Fundamental Property, one of the loads that $W_{i_0}'$ depends on,
+call it $R_{i_0,j_0}'$, is not omitted and is ordered before $W_{i_0}'$.
+But now we have a contradiction:
+\begin{itemize}
+\item	$W_{i_0}'$ commits after $R_{i_0,j_0}'$;
+\item	$R_{i_0,j_0}'$ commits after the store $W_k'$ it reads from;
+\item	$W_k'$ commits no earlier than $W_{i_0}'$.
+\end{itemize}
+
+If all the stores in the OOTA cycle are omitted then all the reads
+must be omitted as well.
+This amounts to a situation where the entire cycle has been optimized
+away by the compiler.
 

### 2024-02-09 ootavacuous.tex
FYI, possible progress on language-level OOTA problem
Here are some small edits for the section on hardware.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	the flag bits it tests in the status register.
-	A computed branch instruction would as input the register
+	A computed branch instruction would have as input the register
 	containing the address to branch to.
@@
 Note: Hardware dependency analysis considers only the information
-flowing \emph{within} a CPU, not the information flowing between the
+flowing \emph{within} a CPU, not information flowing between the
 CPU and memory, which is handled separately.
@@
 see that any hardware dependency must ultimately emanate from a load
-instruction or a thread's initial state (or possibly some sort of
+instruction or the thread's initial register values (or possibly some sort of
 input instruction, but we will not consider that here).
@@
 point it must \emph{commit} to a decision either to definitely execute the
-instruction with some well-defined collection of inputs and outputs
+instruction's action with some collection of well defined inputs and outputs
 or else to abandon it.
@@
 in principle to prevent a CPU from committing a \co{multiply}
-instruction as soon as either one of its inputs has committed to the
+instruction as soon as either one of its inputs' sources has committed to the
 value zero.)
 The reason is obvious: An input is subject to change at any time until
-its source commits, the outputs aren't determined until the inputs
-are, and the instruction can't commit until its outputs are fully
+its source commits,
+the action and outputs aren't fully determined until the inputs
+are, and the instruction can't commit until its action and outputs are fully
 determined.
@@
 And even taking the exceptions into account, if a change in the
-outputs of instruction $I$ would lead to a change in the inputs to
+outputs of instruction $I$ would lead to a change in the action or outputs of
 instruction $J$ then $I$ must commit before $J$.
@@
 not make the value of a store instruction available for other CPUs to
-load until the store commits or later.
+load until the store commits or some time later.
 And after this happens, it takes some time for the information about
@@
 if taken, where it will branch to;
-until then the CPU cannot know whether the instruction should
+until then the CPU cannot know whether the later instruction should
 be executed at all.
@@
 And in general, until the source for store's address commits, it's not
-fully determined whether the store and load are for the same address.
+fully determined whether the store and load will access the same address.
 This argument doesn't hold if the store and load have the \emph{same}
@@
 will turn out to be invalid (a NULL pointer, for example), causing the
-store instruction to encounter an addressing exception.
+store instruction to generate an addressing exception.
 If this should happen the load might not execute at all, and therefore
@@
 It's quite straightforward to show that for any implementation in which
-semantic dependencies satisfy the Fundamental Property, no OOTA
-abstract execution can be realized.
+semantic dependencies satisfy the Fundamental Property, no OOTA cycle
+has a nontrivial realization.
 
@@
 execution $H$ and $A$ has an OOTA cycle.
-This means there are stores $W_i$ semantically depending on loads
-$\{R_{i,j}\}$ where each of the loads reads from one of the $W_i$
+This means there are stores $W_i$ in $A$, semantically depending on loads
+$\{R_{i,j}\}$ where each of the loads reads from one of the $W_k$
 stores in a different thread.
 Let $W_i'$ and $R_{i,j}'$ be the hardware instructions these accesses
-map to, if they aren't omitted.
+map to in $H$, if they aren't omitted.
 Assuming that at least one of the stores is not omitted,
-one of the $W_i'$, call it $W_{i_0}'$, must commit first.
-By the Fundamental Property, one of the loads that $W_{i_0}'$ depends on,
-call it $R_{i_0,j_0}'$, is not omitted and is ordered before $W_{i_0}'$.
+one of the $W_i'$ instructions, let's say $W_0'$, must commit first.
+By the Fundamental Property, one of the loads that $W_0$ depends on,
+let's say $R_{0,0}$, is not omitted and $R_{0,0}'$ is ordered before $W_0'$.
 But now we have a contradiction:
 \begin{itemize}
-\item	$W_{i_0}'$ commits after $R_{i_0,j_0}'$;
-\item	$R_{i_0,j_0}'$ commits after the store $W_k'$ it reads from;
-\item	$W_k'$ commits no earlier than $W_{i_0}'$.
+\item	$W_0'$ commits after $R_{0,0}'$;
+\item	$R_{0,0}'$ commits after the store instruction $W_k'$ it reads from;
+\item	$W_k'$ commits no earlier than $W_0'$.
 \end{itemize}
@@
 This amounts to a situation where the entire cycle has been optimized
-away by the compiler.
+out of existence by the compiler.
+Although we can are unable to prove it, we conjecture that in this
+situation there must be another abstract execution which has the
+same observable effects as $A$ and is also realized by $H$, but in
+which the OOTA cycle does not occur.
+Thus there would be no way to tell, merely by observing the effects of
+$A$, whether there was an OOTA cycle or not.
+For this reason we declare realizations of OOTA cycles in which
+all the accesses are omitted to be trivial.
 

### 2024-02-12 ootavacuous.tex
Update Load/Store appendix to cover tooling
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\section{Load/Store Ordering: Hardware View for Software Hackers}
+\section{But What About Tooling?}
+\label{sec:But What About Tooling?}
+
+This paper focuses primarily on showing that, under commonly occurring
+constraints, OOTA cycles cannot form in real-world C++ implementations.
+
+But what about tooling?
+
+This appendix expands on
+Section~\ref{sec:Semantic Dependencies and Tooling}
+by looking into the possibility of:
+
+\begin{enumerate}
+\item	Tooling focusing on only part of the language,
+\item	Changing the language to eliminate some aspects that are
+	troublesome for tooling (and potentially accepting performance
+	and energy-efficiency shortfalls), and
+\item	Changing the language to better delineate those portions that
+	tooling can easily accommodate.
+\end{enumerate}
+
+But first, the next section looks at why some weakly ordered
+implementations appear to forbid reordering of prior relaxed loads with
+later relaxed stores,\footnote{
+	Or, in memory-order-speak, forbid all load-buffering litmus tests.}
+despite the architecture permitting such reordering.
+
+\subsection{Load/Store Ordering: Hardware View for Software Hackers}
 \label{sec:Load/Store Ordering: Hardware View for Software Hackers}
@@
 continue to refuse.
+
+\subsection{Status Quo and Focused Tooling}
+\label{sec:Status Quo and Focused Tooling}
+
+The theoretical possibility of OOTA cycles causes some tools to have
+difficulty identifying precisely which outcomes are impossible on
+real-world C++ implementations.
+One approach is for tooling to reject programs containing instances
+of \co{memory_order_relaxed} and \co{memory_order_consume}, so that
+developers desiring their code to be analyzed by such tools would
+avoid using these \co{memory_order} values.
+The is strong precedent for this strategy, with the common prohibition
+against side effects in function arguments being but one example.
+
+However, there is a large body of existing code that uses
+\co{memory_order_relaxed}, and it would be unfortunate if such tools
+could not be applied to this code.
+
+Another approach would be to add flags to C++ implementations
+to cause \co{memory_order_relaxed} to be interpreted as either
+\co{memory_order_acquire} (for loads), \co{memory_order_release} (for
+stores), or \co{memory_order_acq_rel} (for read-modify-write operations).
+On TSO systems and on systems featuring exact ECC exceptions, this
+would allow tooling to be brought to bear without performance or
+energy-efficiency consequences beyond forgone optimizations involving
+memory reference reordering.
+
+However, this would mean that programs compiled in this way would not
+be guaranteed to run correctly on weakly ordered systems that lack
+ECC (or that lack precise exceptions).
+In addition, this change is more strict than necessary, forbidding
+optimizations that tooling could in fact analyze.
+Therefore, the next section looks at standardizaing a less severe
+restriction.
+
+\subsection{Change Relaxed to Forbid Load Buffering}
+\label{sec:Change Relaxed to Forbid Load Buffering}
+
+Tooling does not require relaxed accesses become fully acquire and/or
+release, but rather only that implemetnations be forbidden from reordering
+prior relaxed loads with subsequent relaxed stores, as has been suggested
+many times over the
+years~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352}.
+
+This preserves portability while enabling tooling to handle all members
+of the \co{memory_order} enumeration, but inflicts some performance and
+energy-efficiency penalties~\cite{LukeGeeson2023TightenRelaxed} in
+code not requiring these ordering
+restrictions~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
+
+This suggests the addition of a \co{memory_order_load_store} member to
+the \co{memory_order} enumeration, a topic taken up by the next section.
+
+\subsection{Add Load-Store Memory Order that Forbids Load Buffering}
+\label{sec:Add Load-Store Memory Order that Forbids Load Buffering}
+
+The addition of a \co{memory_order_load_store} member to the
+\co{memory_order} enumeration has been suggested starting many years
+ago~\cite{HansBoehm2019OOTArevisitedAgain,DanielLustig2018PlacedBefore}.
+This does not resolve all shortcomings in the C++ memory
+model~\cite{PaulEMcKenney2023P0124R8-LKMM,DanielLustig2018PlacedBefore},
+but it would provide a portable \co{memory_order} that minimally
+restricted compiler and hardware optimizations while still permitting
+full analyzability by current software tools.
+
+This change would also leave \co{memory_order_relaxed} in place,
+allowing minimal-overhead accesses in fastpaths.
+These fastpaths would not be analyzable by generic tools, but could
+be handled by special-case tools that analyze the binaries to verify
+that required orderings are enforced by machine-language properties
+such as dependencies.
+And a tool that checks control dependencies has in fact been
+prototyped~\cite{PaulHeidekrueger2022N4910}.
+
+This suggests a combined strategy of adding \co{memory_order} members
+as needed to extend the reach of general-purpose tooling, while also
+identifying \co{memory_order_relaxed} idioms that are checked at the
+machine level using special-purpose tooling.\footnote{
+	Kudos to Peter Sewell for clearly articulating this possibility
+	as part of an overall strategy.}
 

### 2024-02-13 ootavacuous.tex
February 13 eview of reworked material
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 prohibition above, and the term \emph{strict C++}\footnote{
-	@@@ Future bikeshedding of name.}
+	There is some discomfort with this name, so this footnote
+	is reserved for future bikeshedding.}
 % @@@ Alternatives thus far: OOTA-prone C++, OOTA-enabled C++,
-% @@@ relaxed C++, OOTA C++, OOTA-agnostic C++.
+% @@@ relaxed C++, OOTA C++, OOTA-agnostic C++, loose C++.
 to denote a hypothetical standard that excludes this prohibition, but
@@
 stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
-but this requires real instructions be executed, consuming real
-time and real electrical power to solve a strictly theoretical
-problem.
-This might have been acceptable in the 1960s of my youth, but it is now
-the year 2024.
+but this can require real instructions be
+executed~\cite[Section 7.1]{Maranget2012TutorialARMPower},
+consuming real time and real electrical power to solve a strictly
+theoretical problem.
+This might have been acceptable in the 1960s of some of the authors'
+youths, but it is now the year 2024.
 
@@
 Goldblatt looked at interactions between OOTA cycles and
-undefined behavior (UB)~\cite{DavidGoldblatt2019NoElegantOOTAfix}, and
-Appendix~\ref{app:Aside on Undefined Behavior}
-analyzes his examples and notes ways to separate UB and OOTA-cycle
-concerns.
+undefined behavior (UB)~\cite{DavidGoldblatt2019NoElegantOOTAfix}.
+% Appendix~\ref{app:Aside on Undefined Behavior}
+% analyzes his examples and notes ways to separate UB and OOTA-cycle
+% concerns.
 The remainder of this document will concentrate on examples lacking UB.
@@
 in sdep $\cup$ rf~\cite{PaulEMcKenney2014OOTA},
-% @@@ This citation credits the formulation to Ali Sezgin, third author
-% @@@ of N4323 ("Out-of-Thin-Air Execution is Vacuous").
+% This citation credits the formulation to Ali Sezgin, third author
+% of N4323 ("Out-of-Thin-Air Execution is Vacuous").
 where sdep is the set of semantic dependencies within
@@
 	See Appendix~\ref{app:Inter-Process Communications} for definitions
-	and properies of rf, rfe, and rfi.}
+	and properties of rf, rfe, and rfi.}
 
@@
 
-@@@ Please see
-Appendix~\ref{app:Informal Definition of Semantic Dependency}
-for more a more detailed (albeit quite informal) definition.
-Maybe eventually have an appendix that lists all the definitions
-of both ``OOTA cycle'' and ``semantic dependency''.
+% We no longer need this.  I will move A.1 to the cbmc appendix.
+% Please see
+% Appendix~\ref{app:Informal Definition of Semantic Dependency}
+% for more a more detailed (albeit quite informal) definition.
+% Maybe eventually have an appendix that lists all the definitions
+% of both ``OOTA cycle'' and ``semantic dependency''.
 
@@
 section~\ref{sec:Global Optimization Can Destroy Dependencies}).
+% The usual way in the Linux kernel is via a union, but it is probably
+% best to ignore that possibility.
 
@@
 therefore on whether or not the example has a semantic dependency.
-It is the implementations's choice.
+It is the implementation's choice.
 
@@
 Here \co{y} is semantically dependent on \co{c} in executions
-for which \co{char} is a signed type,
+running on implementations in
+which \co{char} is a signed type,
 but not those for which it is unsigned.
@@
 in unsequenced subexpressions constitute undefined behavior,
-although there are proposals to make them defined in both C++ and C.\footnote{
-	\url{https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2016/p0145r3.pdf}
-	\url{https://www.open-std.org/jtc1/sc22/wg14/www/docs/n3203.htm}}
+although there are proposals to make them defined in both
+C++~\cite{GabrielDosReis2016P0145r3}
+and C~\cite{AlexCeleste2023N3203}.
 Nevertheless, the example above is allowed because the order of evaluation
@@
 The new requirement is that the map must preserve the rf relation.
+% Is this rf?  Or specifically rfe?
 That is, if load $R$ in $A$ reads from store $W$ and $R$ is not
@@
 and consequently $W$ must not be omitted either.
+
+Section~\ref{sec:Constraints of the Standard}
+presents examples illustrating some of these requirements.
 
@@
 
-Consider this example, which appears to be an \co{if} statement:
+Consider this example, similar to that shown in
+Section~\ref{sec:Semantic Dependencies Not Affected by if Statements},
+which appears to be an \co{if} statement:
 
@@
 
-Consider the following:
+Consider the following example derived from that in
+Section~\ref{sec:Semantic Dependencies and Source Code}:
 
@@
 
-\emph{Any analysis of a given execution must either be fully informed
-of the enclosing context, or much make conservative assumptions about
+\emph{Any analysis of a given execution either must be fully informed
+of the enclosing context, or must make conservative assumptions about
 that context.}
@@
 
-Consider this example, with \co{i} initially zero:
+Consider this example, similar to that in
+Section~\ref{sec:Abstract Executions},
+with \co{i} initially zero:
 
@@
 Note that side effects in expressions lacking sequence points is
-undefined behavior, however, there are proposals to define it in
-the C++ and C languages, allowing implementations to nevertheless reorder
-evaluation when permitted by the as-if rule.\footnote{
-	\url{https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2016/p0145r3.pdf}
-	\url{https://www.open-std.org/jtc1/sc22/wg14/www/docs/n3203.htm}}
-
-Compiler-based C++ implementions make their non-deterministic choices
-at compile time, so that the object code is deterministic, at least
-on a per-execution basis.
+undefined behavior, however, there are proposals to define it in the
+C++~\cite{GabrielDosReis2016P0145r3} and
+C~\cite{AlexCeleste2023N3203}
+languages, allowing implementations to nevertheless reorder
+evaluation when permitted by the as-if rule.
+
+Even though the standard is non-deterministic, \emph{compiler-based C++
+implementions make their non-deterministic choices at compile time, so
+that the object code is deterministic, at least on a per-execution basis.}
 
@@
 
-The key point here is that diferent values might result in almost
+The key point here is that different values might result in almost
 completely disjoint executions.
@@
 for at least some approximate definitions of ``semantic dependency''.
-An example use of \co{cbmc} is illustrated in
-Appendix~\ref{sec:Evaluating sdep Using cbmc}.
+% An example use of \co{cbmc} is illustrated in
+% Appendix~\ref{sec:Evaluating sdep Using cbmc}.
 
@@
 especially the implementation.
-However, consider the following code:
+However, consider the following code, similar to that shown in
+Section~\ref{sec:Abstract Executions}:
 
@@
 
-@@@ LTO?
+% @@@ LTO?  Does not appear to be relevant given new material.
 
@@
 constraints introduced by the next section.
-
-@@@  Need to add material on inventing loads, which is currently
-in an appendix.
 
@@
 architecture and design, both of which have been only partially accounted
-for in the standard.
+for in the standard and in prior work.
 One reason for the incomplete accounting for hardware constraints is
@@
 However, no one has managed to construct a real time-travel machine
-or any other apparatus that might result in effect preceding cause,
-as discussed at length in Appendix~\ref{sec:Debates on Causality} and 
-the URL that it provides.
+or any other apparatus that might result in effect preceding cause.
+% as discussed at length in Appendix~\ref{sec:Debates on Causality} and 
+% the URL that it provides.
 
@@
 Section~\ref{sec:But What About Non-Volatile Atomics?}.
+% @@@ Pull that section in here?  Drop this forward reference?
 

### 2024-02-13 ootavacuous.tex
February 13 review of reworked material
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 The new requirement is that the map must preserve the rf relation.
-% Is this rf?  Or specifically rfe?
 That is, if load $R$ in $A$ reads from store $W$ and $R$ is not
@@
 
-Appendix~\ref{app:Informal Definition of Semantic Dependency}
-outlines the informal informal definition of ``semantic dependency''
-used to evaluate the litmus tests in this section,
 Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}
@@
 otherwise gave us pause.
-
-\subsection{Informal Definition of Semantic Dependency}
-\label{app:Informal Definition of Semantic Dependency}
-
-The manual analysis uses the informal rule that a semantic dependency
-exists between a set of relaxed atomic loads\footnote{
-	Please see
-	Appendix~\ref{app:Non-Trivial Semantic Dependencies}
-	for examples showing why this is a set of loads rather than a
-	single load.}
-and a store on any execution where different values loaded would change
-the address stored to, the value stored, or the number of times that
-store would be executed, with the proviso that stores of the same value
-to the same address having the same ordering properties are fungible.
-
-Producing a precise formal statement of this rule turns out to be quite
-challenging, but as noted earlier, there is considerable work along
-various
-lines~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA}.
-For examples of the challenges, please see
-Section~\ref{sec:On OOTA Definitions}.
-Our informal definition meets these challenges as follows:
-
-\begin{enumerate}
-\item	Section~\ref{sec:Semantic Dependencies and Source Code}
-	is responsible for the phrase ``on any execution'' as well
-	as the proviso regarding fungible loads.
-\item	Section~\ref{sec:Semantic Dependencies Depend on Context}'s
-	challenges are met by looking at the full context within
-	a given thread.
-\item	Section~\ref{sec:Semantic Dependencies Can Be Nondeterministic}'s
-	non-determinism challenges are avoided by avoiding litmus tests
-	exhibiting non-deterministic behavior.
-	Again, note that most production C++ code uses compiler-based C++
-	implementations, which freeze their non-deterministic choices
-	into the binary at compile time.
-\item	Section~\ref{sec:Semantic Dependencies And Optimization}'s
-	optimization-choice challenges are met by using the C++
-	implementation's choice of semantic dependency based on the
-	optimizations that the implementation actually applied.
-\item	Section~\ref{sec:Semantic Dependencies and I/O}'s
-	I/O challenges are met by again noting that I/O is observed
-	behavior.
-	And by avoiding litmus tests involving I/O, given that relatively
-	few of the available tools handle I/O.
-\item	Section~\ref{sec:Semantic Dependencies Can Be Complex}'s
-	complexity challenges are met by looking at what the code
-	actually does.
-	For example, if an as-yet unproven mathematical conjecture
-	holds, the code will act consistently with that conjecture,
-	and vice versa.
-\item	Section~\ref{sec:Implementers and Users Influence Semantic-Dependencies Definition}'s
-	implementer-/user-choice challenges are met by accounting for
-	those choices.
-	For example, if a user has specified \co{-funsigned-char},
-	then the actual executions will be consistent with that choice.
-	For more examples, please see
-	Appendix~\ref{app:User Influence Over Language Semantics}.
-\end{enumerate}
-
-In short, we are treating ``semantic dependency'' in a manner similar
-to our treatment of the laws of physics.
-We let the actual executions provide an empirical basis for the analysis.
-
-We do not expect this approach to please everyone.
-However, those who are displeased might do well to reflect on the fact
-that the C++ standard is not mathematical in nature, which implies that
-any mathematical model of C++ is inherently an approximation.
-We look forward to the day when the standard gains more mathematical
-rigor, and perhaps the C++ memory model is a first step in that
-direction.\footnote{
-	But if so, the OOTA situation should be a caution!}
-Nevertheless, until that day, we must live with the standard in its
-current decidedly non-mathematical form.
 
@@
 
-\clearpage
-
-\section{Aside on Undefined Behavior}
-\label{app:Aside on Undefined Behavior}
-
-The combination of OOTA cycles and UB has proven especially
-challenging~\cite{DavidGoldblatt2019NoElegantOOTAfix,HansBoehm2020UBalternatives}.
-
-\subsection{Alignment-Based Undefined Behavior}
-\label{sec:Alignment-Based Undefined Behavior}
-
-The example shown in
-Listing~\ref{lst:Alignment-Based Undefined Behavior (P1916R0)}
-(taken from P1916R0) is especially instructive.
-
-\begin{listing}[tbp]
+\subsubsection{Semantic Dependencies and \co{volatile}}
+\label{app:Semantic Dependencies and volatile}
+
+The Linux kernel uses volatile accesses to constrain the compiler,
+and it is worth looking at the example shown in
+Listing~\ref{lst:Non-Volatile Accesses and Dependencies}
+to see how this works.
+
+\begin{listing}
 \begin{verbatim}
- 1 std::atomic<int> y;
- 2 template <typename T>
- 3 void t1() {
- 4   long long r1 = x.load(std::memory_order_relaxed);
- 5   if (r1 % 4 == 0)
- 6     y.store(1, std::memory_order_relaxed);
- 7   *(T*)r1 = *(T*)r1;
- 8 }
+ 1 atomic<int> x, y;
+ 2
+ 3 void thread1() {
+ 4   int r1 = x.load(memory_order::relaxed);
+ 5   y.store(r1, memory_order::relaxed);
+ 6 }
 \end{verbatim}
-\caption{Alignment-Based Undefined Behavior (P1916R0)}
-\label{lst:Alignment-Based Undefined Behavior (P1916R0)}
-\end{listing}
-
-Because the compiler often assumes that UB cannot happen, it might
-look at line~7 and conclude that the value of \co{r1} necessarily
-corresponded to an address aligned per the requirements of type \co{T}.
-If that type's alignment was larger than four bytes, the compiler could
-then replace the condition in line~5 with \co{true}, which would mean
-that there was no semantic dependency from the load of \co{x} to the
-store of \co{y}.
-
-Except that the developer might have knowingly carried out a misaligned
-access, for example, to test an alignment trap handler or because the
-system on which this code was to run was known to gracefully handle
-misaligned accesses.
-
-\begin{listing}[tbp]
+\caption{Non-Volatile Accesses and Dependencies}
+\label{lst:Non-Volatile Accesses and Dependencies}
+\end{listing}
+
+Because both \co{x} and \co{y} are non-volatile, the compiler is
+free to assume that the value loaded by line~4 will be some value
+stored to \co{x} or its initial value.
+Because there are no stores to \co{x}, the only remaining possibility
+is the default-initialized value of zero.
+Therefore, the C++ implementation is within its rights to substitute
+the constant zero for \co{r1} on line~5, eliminating the semantic
+dependency that would otherwise extend from line~4 to line~5.
+This in turn enables the implementation to reorder line~5 to
+precede line~4.
+
+\begin{listing}
 \begin{verbatim}
- 1 std::atomic<int> y;
- 2 template <typename T>
- 3 void t1() {
- 4   long long r1 = x.load(std::memory_order_relaxed);
- 5   if (r1 % 4 == 0)
- 6     y.store(1, std::memory_order_relaxed);
- 7   std::observable();
- 8   *(T*)r1 = *(T*)r1;
- 9 }
+ 1 atomic<volatile int> x, y;
+ 2
+ 3 void thread1() {
+ 4   int r1 = x.load(memory_order::relaxed);
+ 5   y.store(r1, memory_order::relaxed);
+ 6 }
 \end{verbatim}
-\caption{Alignment-Based Undefined Behavior Fix}
-\label{lst:Alignment-Based Undefined Behavior Fix}
-\end{listing}
-
-However, in that case, the developer is stepping outside the bounds
-of the standard.
-It would be good to have some way for the developer to get the job
-done without confusing the compiler, thus separating OOTA-cycle and
-UB concerns.
-One approach is to use the \co{std::observable()} function
-that has been proposed to the C++ standards
-committee~\cite{DavisHerring2021P1494R2}.
-This function would blocks backwards-propagating UB.
-Then this example could be rewritten as shown in
-Listing~\ref{lst:Alignment-Based Undefined Behavior Fix},
-with the asm on line~7 preventing the compiler from assuming that line~8
-is guaranteed to be executed, thus preventing the compiler from
-applying any lessons that it might learn from line~8 to line~5.
-
-% Clean up <200b> in vim: ":%s/\%u200b//g"
-
-There are a number of ways to implement \co{std::observable()},
-for example, a new \co{"mightnotreturn"} GCC asm clobber, a call to an empty
-function that is invisible to the compiler, and a GCC goto asm that the
-compiler cannot prove falls through.
-
-\subsection{Unreachability-Based Undefined Behavior}
-\label{sec:Unreachability-Based Undefined Behavior}
-
-This section attempts to apply the lessons
-of Appendix~\ref{sec:Unreachability-Based Undefined Behavior} to
-Goldblatt's unreachability-UB example.
-
-The \co{std::observable()} at line~12 of
-Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
-avoids confusing the compiler, instead forcing it to assume that the
-value of \co{local} might not be zero.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 // UB if this function is ever called.
- 2 [[noreturn]]
- 3 inline void unreachable() {
- 4   return;
- 5 }
- 6 void t1() {
- 7   r1 = x.load(std::memory_order_relaxed);
- 8   int local = (r1 & 1);
- 9   if (local == 0)
-10     y.store(1, std::memory_order_relaxed);
-11   if (local != 0) {
-12     std::observable();
-13     // If we get here, UB.
-14     unreachable();
-15   }
-16 }
-\end{verbatim}
-\caption{Unreachability-Based Undefined Behavior Fix}
-\label{lst:Unreachability-Based Undefined Behavior Fix}
-\end{listing}
-
-By the same reasoning, Goldblatt's \co{misbehave} example that
-also involves unreachability can also be addressed through insertion
-of the proposed \co{std::observable()}.
-
-\subsection{Atomic Stores From Undefined Behavior}
-\label{sec:Atomic Stores From Undefined Behavior}
-
-Another proposal to the C++ standards
-committee~\cite{HansBoehm2020UBalternatives}
-proposes (among other alternatives) that any atomic store resulting
-from undefined behavior have at least release semantics.
-This would prevent such a store from propagating undefined behavior
-backwards within that store's thread.
-In some cases, this could also prevent this undefined behavior from
-propagating back through other threads.
-
-However, it might well force C++ implementations to promote a surprisingly
-large fraction of a program's relaxed stores to release.
-
-\subsection{Lifetime-Based Undefined Behavior}
-\label{sec:Lifetime-Based Undefined Behavior}
-
-Not all undefined behavior can be backstopped using
-\co{std::unreachable()}.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 int *foo(int *r)
- 2 {
- 3   b.store(r, memory_order_relaxed);
- 4   return &global;
- 5 }
- 6
- 7 int go(int *p)
- 8 {
- 9   int r;
-10   int *q = foo(&r);
-11
-12   r = 2;
-13   *p = 1;
-14   *q = r;
-15   return *q - r;
-16 }
-17
-18 void t1(void)
-19 {
-20   int r1 = go(a.load(memory_order_relaxed));
-21   printf("go() = %d\n", r1);
-22 }
-23
-24 void t2(void)
-25 {
-26   int *r2 = b.load(memory_order_relaxed);
-27   a.store(r2, memory_order_relaxed);
-28 }
-\end{verbatim}
-\caption{Lifetime-Based Undefined Behavior}
-\label{lst:Lifetime-Based Undefined Behavior}
-\end{listing}
-
-Listing~\ref{lst:Lifetime-Based Undefined Behavior}
-shows an example loosely based on one presented by Stephen Dolan\footnote{
-	\url{https://stedolan.net/talks/2024/loadstore}}
-in which function \co{go()} invokes \co{foo()} with a pointer to \co{go()}'s
-local variable \co{r}.
-However, \co{foo()} uses a relaxed store to place this pointer in
-an atomic variable \co{b}.
-Because this store is relaxed, it is not ordered against other code
-in \co{go()} and \co{t1()}, which means that \co{t2()} might well
-read this pointer from \co{b} either before the beginning or after
-the end of \co{r}'s lifetime, neither of which is well-defined.
-
-Of course, any attempt to dereference the pointer read from either
-\co{a} or \co{b} could result in undefined behavior.
-
-The lesson is clear:
-Properly synchronize reads and writes of pointers with the lifetime
-of the pointed-to object.
-
-This example could make lines~3 and~27 be \co{memory_order_release} and
-lines~20 and~26 be \co{memory_order_acquire} in order to avoid carrying
-a pointer to \co{r} before the lifetime of \co{r} has started.
-Alternatively, threads could be spawned only after \co{r}'s lifetime
-starts and joined before it ends.
-Placing \co{r} in static global storage would avoid lifetime considerations
-totally.
-And there are of course any number of other synchronization schemes
-that might be used to ensure that pointers are in use only while the
-pointed-to objects are alive.
-
-Note that \co{std::unreachable()} does not suffice because the CPU
-can also reorder relaxed stores.
-Note also that load-store ordering is not sufficient to prevent a
-pointer to \co{r} from being stored to \co{b} prior to the start of
-\co{r}'s lifetime because there is no atomic load between the start
-of that lifetime and the relaxed store to \co{b}.
-
-There are a number of ways to prevent carrying a pointer to \co{r} after
-the lifetime of \co{r} has ended, and these are left as an exercise for
-the reader.
-
-\clearpage
-
-\section{Evaluating sdep Using \co{cbmc}}
-\label{sec:Evaluating sdep Using cbmc}
-
-This section demonstrates the evaluation of sdep using
-the C Bounded Model Checker
-(\co{cbmc})~\cite{EdmundClarke2004CBMC}.
-This tool takes a smallish C program as input and constructs a logic
-expression representing that program.
-The logic expression's variables represent bits in the program's inputs
-and state, and the result of that logic expression is \co{true} if that
-program has a bug, for example, if it is possible for that C program to
-trigger an assertion or execute an out-of-bounds array reference.
-The tool then hands this expression to a SAT solver, and if this solver
-finds a satisfying set of values, the tool uses those values provide
-the execution path to the bug.
-
-\subsection{Complex Semantic Dependencies}
-\label{app:Complex Semantic Dependencies}
-
-Listing~\ref{lst:Example Code for Evaluation of sdep}
-shows a \co{herd7} thread \co{P0()}.
-Is there a semantic dependency between the load from \co{x} on
-line~3 to the store to \co{y} on line~9?
-Answering this question requires solving a mathematical equation,
-not all of which currently have known solutions.
-However, this section will use a relatively simple equation combined
-with restrictions on domain and range that permit \co{cbmc} to do
-the job.
-
-\begin{listing}[tbp]
-\scriptsize
-\begin{verbatim}
- 1 P0(atomic_int *x, atomic_int *y, atomic_int *limit)
- 2 {
- 3   r1 = atomic_load_explicit(x, memory_order_relaxed);
- 4   r2 = atomic_load_explicit(limit, memory_order_relaxed);
- 5   r3 = r1 * r1 + 2 * r1 + 2;
- 6   r4 = r2;
- 7   if (r2 >= r3)
- 8     r4 = r3;
- 9   atomic_store_explicit(y, r4, memory_order_relaxed);
-10 }
-\end{verbatim}
-\caption{Example Code for Evaluation of sdep}
-\label{lst:Example Code for Evaluation of sdep}
-\end{listing}
-
-\begin{listing}[tbp]
-\scriptsize
-\begin{verbatim}
- 1 int main(int argc, char *argv[])
- 2 {
- 3   int i;
- 4   int y[2] = { 0, 0 };
- 5
- 6   for (i = 0; i < 2; i++) {
- 7     int r1;
- 8     int r2;
- 9     int r3;
-10     int *yp = &y[i];
-11
-12     r1 = limited_arg(argv[2 * i + 1]);
-13     r2 = const_arg(limited_arg(argv[2 * i + 2]));
-14     r3 = r1 * r1 + 2 * r1 + 2;
-15     *yp = r2 <= r3 ? r2 : r3;
-16     printf("i = %d, r1 = %d, r2 = %d, r3 = %d\n", i, r1, r2, r3);
-17   }
-18   assert(y[0] == y[1]);
-19 }
-\end{verbatim}
-\caption{Code for cbmc-Based Evaluation of sdep}
-\label{lst:Code for cbmc-Based Evaluation of sdep}
-\end{listing}
-
-One way to answer this question is to create a C program
-(see Listing~\ref{lst:Code for cbmc-Based Evaluation of sdep})
-that runs the \co{P0()} thread's code twice with arbitrary inputs
-(the loop spanning lines~6-17) and asserts that the results of
-each iteration be equal (line~18).
-
-Running this with:
-
-\begin{quote}
-\scriptsize
-\begin{verbatim}
-cbmc -DCBMC cbmc/sdep-quadratic.c
-\end{verbatim}
-\end{quote}
-
-gives the result \co{VERIFICATION FAILED}.
-This indicates that there are different values of \co{x} and \co{limit}
-that yield different values of \co{y}, that is, that there can be
-a semantic dependency.
-
-\begin{listing}[tbp]
-@@ DisplayLitmus cbmc/sdep-quadratic.c @@
-\caption{Full Code for cbmc-Based Evaluation of sdep}
-\label{lst:Full Code for cbmc-Based Evaluation of sdep}
-\end{listing}
-
-However, $x^2 + 2x + 2$ has a minimum value of 1 at $x=-1$, so
-forcing \co{limit} to (say) zero should eliminate this static
-dependency:\footnote{
-	See Listing~\ref{lst:Full Code for cbmc-Based Evaluation of sdep}
-	for the full \path{cbmc/sdep-quadratic.c} program.}
-
-\begin{quote}
-\scriptsize
-\begin{verbatim}
-cbmc -DCBMC -DLOAD_CONSTANT=0 cbmc/sdep-quadratic.c
-\end{verbatim}
-\end{quote}
-
-But this also gives the result \co{VERIFICATION FAILED}.
-One reason for this is that this program does 32-bit computer arithmetic,
-and is thus subject to wraparound.
-Limiting the value of \co{x} to 46,339 prevents wraparound:
-
-\begin{quote}
-\scriptsize
-\begin{verbatim}
-cbmc -DCBMC -DLOAD_CONSTANT=0 -DLOAD_LIMITS=46339 cbmc/sdep-quadratic.c
-\end{verbatim}
-\end{quote}
-
-And this yields \co{VERIFICATION SUCCESSFUL}, indicating that there
-is no semantic dependency in this case.
-This is to be expected because the specified value of zero for \co{limit}
-is smaller than the minimum of the quadratic expression, so that the
-value stored to \co{y} is always zero.
-Specifying \co{-DLOAD_CONSTANT=1} also yields \co{VERIFICATION SUCCESSFUL}
-because this value is equal to that minimum, so that the value stored to
-\co{y} is always 1.
-However, specifying any larger value (for example, \co{-DLOAD_CONSTANT=2})
-yields \co{VERIFICATION FAILED} because the value assigned to \co{y}
-might be either 1 or 2, indicating that a semantic dependency exists.
-
-This situation underscores the point that the existance of a semantic
-dependency can be a property of a particular execution rather than a
-universal property of the code itself.
-The interested reader is encouraged to create more complex examples,
-for example, demonstrating semantic dependencies from multiple loads or
-exploring indeterminacy.
-
-There are numerous other tools that can be used, up to and including
-manual inspection, but \co{cbmc} is readily available and easy to
-use.\footnote{
-	On Debian-based distributions, via \co{sudo apt install cbmc}.}
-
-\subsection{C++ Non-Determinism}
-\label{app:C++ Non-Determinism}
-
-The C++ standard allows non-determinism, for example, the order of
-invocation of side-effects in some circumstances.
-Listing~\ref{lst:Code for cbmc-Based Evaluation of Non-Deterministic sdep}
-shows an example leveraging side-effects in the parameter list shown on
-line~37.
-
-\begin{listing}[tbp]
-@@ DisplayLitmus cbmc/sdep-nondet.c @@
-\caption{Code for cbmc-Based Evaluation of Non-Deterministic sdep}
-\label{lst:Code for cbmc-Based Evaluation of Non-Deterministic sdep}
-\end{listing}
-
-Running \co{cbmc} on this as follows:
-
-\begin{quote}
-\scriptsize
-\begin{verbatim}
-cbmc -DCBMC cbmc/sdep-nondet.c
-\end{verbatim}
-\end{quote}
-
-Gives the result \co{VERIFICATION SUCCESSFUL}, showing that \co{cbmc}
-does not model this non-determinism.
-Which is not surprising, given that many of the safety-critical uses
-to which \co{cbmc} is put would prohibit such non-determinism in their
-code bases.
+\caption{Volatile Accesses and Dependencies}
+\label{lst:Volatile Accesses and Dependencies}
+\end{listing}
+
+One way to preserve this dependency is use of \co{volatile}, as shown in
+Listing~\ref{lst:Volatile Accesses and Dependencies}.
+Now the implementation is forbidden from assuming that the value loaded
+by line~4 has any relation to any stores to or initialization of \co{x},
+even if \co{x} is placed in normal memory.
+
+In this case, the \co{volatile} keyword alerts the implementation to
+the possibility of interfering accesses to \co{x} from outside the
+program, for example, due to \co{x} being:
+
+\begin{itemize}
+\item	Allocated in an MMIO region of the address space.
+\item	Modified by an unknown-to-the-implementation thread
+	or signal handler.
+\item	Subject to I/O-device DMA operations.
+\item	Modified by debugger commands.
+\item	Modified by accesses to facilities such as \path{/dev/mem}.
+\item	Modified by as-yet-unwritten dynamically linked libraries.
+\item	Modified by some other mechanism of the reader's choosing.
+\end{itemize}
+
+These possibilities force the implementation to preserve the semantic
+dependency from line~4 to line~5.
+They can also prevent the formation of OOTA cycles local to the
+implementation because such inteference means that the volatile loads
+might be reading from something that the implementation knows nothing
+about.
+
+The developer is free to arrange for \co{x} to be free of any
+interference, thus obtaining predictable behavior along with the
+preserved semantic dependency.
+However, this example invalidates the rough definition of semantic
+dependency from
+Section~\ref{sec:Simple OOTA Cycle}
+because the value loaded from \co{x} will always be the initial value
+of zero, so that it is at best dubious to talk about any changes in
+the value loaded.
+
+One way forward is to assume that interference could happen when
+determining which dependencies are semantic, and then do further analysis
+using that determination.
 
@@
 
+\section{Aside on Undefined Behavior}
+\label{app:Aside on Undefined Behavior}
+
+The combination of OOTA cycles and UB has proven especially
+challenging~\cite{DavidGoldblatt2019NoElegantOOTAfix,HansBoehm2020UBalternatives}.
+
+\subsection{Alignment-Based Undefined Behavior}
+\label{sec:Alignment-Based Undefined Behavior}
+
+The example shown in
+Listing~\ref{lst:Alignment-Based Undefined Behavior (P1916R0)}
+(taken from P1916R0) is especially instructive.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 std::atomic<int> y;
+ 2 template <typename T>
+ 3 void t1() {
+ 4   long long r1 = x.load(std::memory_order_relaxed);
+ 5   if (r1 % 4 == 0)
+ 6     y.store(1, std::memory_order_relaxed);
+ 7   *(T*)r1 = *(T*)r1;
+ 8 }
+\end{verbatim}
+\caption{Alignment-Based Undefined Behavior (P1916R0)}
+\label{lst:Alignment-Based Undefined Behavior (P1916R0)}
+\end{listing}
+
+Because the compiler often assumes that UB cannot happen, it might
+look at line~7 and conclude that the value of \co{r1} necessarily
+corresponded to an address aligned per the requirements of type \co{T}.
+If that type's alignment was larger than four bytes, the compiler could
+then replace the condition in line~5 with \co{true}, which would mean
+that there was no semantic dependency from the load of \co{x} to the
+store of \co{y}.
+
+Except that the developer might have knowingly carried out a misaligned
+access, for example, to test an alignment trap handler or because the
+system on which this code was to run was known to gracefully handle
+misaligned accesses.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 std::atomic<int> y;
+ 2 template <typename T>
+ 3 void t1() {
+ 4   long long r1 = x.load(std::memory_order_relaxed);
+ 5   if (r1 % 4 == 0)
+ 6     y.store(1, std::memory_order_relaxed);
+ 7   std::observable();
+ 8   *(T*)r1 = *(T*)r1;
+ 9 }
+\end{verbatim}
+\caption{Alignment-Based Undefined Behavior Fix}
+\label{lst:Alignment-Based Undefined Behavior Fix}
+\end{listing}
+
+However, in that case, the developer is stepping outside the bounds
+of the standard.
+It would be good to have some way for the developer to get the job
+done without confusing the compiler, thus separating OOTA-cycle and
+UB concerns.
+One approach is to use the \co{std::observable()} function
+that has been proposed to the C++ standards
+committee~\cite{DavisHerring2021P1494R2}.
+This function would blocks backwards-propagating UB.
+Then this example could be rewritten as shown in
+Listing~\ref{lst:Alignment-Based Undefined Behavior Fix},
+with the asm on line~7 preventing the compiler from assuming that line~8
+is guaranteed to be executed, thus preventing the compiler from
+applying any lessons that it might learn from line~8 to line~5.
+
+% Clean up <200b> in vim: ":%s/\%u200b//g"
+
+There are a number of ways to implement \co{std::observable()},
+for example, a new \co{"mightnotreturn"} GCC asm clobber, a call to an empty
+function that is invisible to the compiler, and a GCC goto asm that the
+compiler cannot prove falls through.
+
+\subsection{Unreachability-Based Undefined Behavior}
+\label{sec:Unreachability-Based Undefined Behavior}
+
+This section attempts to apply the lessons
+of Appendix~\ref{sec:Unreachability-Based Undefined Behavior} to
+Goldblatt's unreachability-UB example.
+
+The \co{std::observable()} at line~12 of
+Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
+avoids confusing the compiler, instead forcing it to assume that the
+value of \co{local} might not be zero.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 // UB if this function is ever called.
+ 2 [[noreturn]]
+ 3 inline void unreachable() {
+ 4   return;
+ 5 }
+ 6 void t1() {
+ 7   r1 = x.load(std::memory_order_relaxed);
+ 8   int local = (r1 & 1);
+ 9   if (local == 0)
+10     y.store(1, std::memory_order_relaxed);
+11   if (local != 0) {
+12     std::observable();
+13     // If we get here, UB.
+14     unreachable();
+15   }
+16 }
+\end{verbatim}
+\caption{Unreachability-Based Undefined Behavior Fix}
+\label{lst:Unreachability-Based Undefined Behavior Fix}
+\end{listing}
+
+By the same reasoning, Goldblatt's \co{misbehave} example that
+also involves unreachability can also be addressed through insertion
+of the proposed \co{std::observable()}.
+
+\subsection{Atomic Stores From Undefined Behavior}
+\label{sec:Atomic Stores From Undefined Behavior}
+
+Another proposal to the C++ standards
+committee~\cite{HansBoehm2020UBalternatives}
+proposes (among other alternatives) that any atomic store resulting
+from undefined behavior have at least release semantics.
+This would prevent such a store from propagating undefined behavior
+backwards within that store's thread.
+In some cases, this could also prevent this undefined behavior from
+propagating back through other threads.
+
+However, it might well force C++ implementations to promote a surprisingly
+large fraction of a program's relaxed stores to release.
+
+\subsection{Lifetime-Based Undefined Behavior}
+\label{sec:Lifetime-Based Undefined Behavior}
+
+Not all undefined behavior can be backstopped using
+\co{std::unreachable()}.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 int *foo(int *r)
+ 2 {
+ 3   b.store(r, memory_order_relaxed);
+ 4   return &global;
+ 5 }
+ 6
+ 7 int go(int *p)
+ 8 {
+ 9   int r;
+10   int *q = foo(&r);
+11
+12   r = 2;
+13   *p = 1;
+14   *q = r;
+15   return *q - r;
+16 }
+17
+18 void t1(void)
+19 {
+20   int r1 = go(a.load(memory_order_relaxed));
+21   printf("go() = %d\n", r1);
+22 }
+23
+24 void t2(void)
+25 {
+26   int *r2 = b.load(memory_order_relaxed);
+27   a.store(r2, memory_order_relaxed);
+28 }
+\end{verbatim}
+\caption{Lifetime-Based Undefined Behavior}
+\label{lst:Lifetime-Based Undefined Behavior}
+\end{listing}
+
+Listing~\ref{lst:Lifetime-Based Undefined Behavior}
+shows an example loosely based on one presented by Stephen Dolan\footnote{
+	\url{https://stedolan.net/talks/2024/loadstore}}
+in which function \co{go()} invokes \co{foo()} with a pointer to \co{go()}'s
+local variable \co{r}.
+However, \co{foo()} uses a relaxed store to place this pointer in
+an atomic variable \co{b}.
+Because this store is relaxed, it is not ordered against other code
+in \co{go()} and \co{t1()}, which means that \co{t2()} might well
+read this pointer from \co{b} either before the beginning or after
+the end of \co{r}'s lifetime, neither of which is well-defined.
+
+Of course, any attempt to dereference the pointer read from either
+\co{a} or \co{b} could result in undefined behavior.
+
+The lesson is clear:
+Properly synchronize reads and writes of pointers with the lifetime
+of the pointed-to object.
+
+This example could make lines~3 and~27 be \co{memory_order_release} and
+lines~20 and~26 be \co{memory_order_acquire} in order to avoid carrying
+a pointer to \co{r} before the lifetime of \co{r} has started.
+Alternatively, threads could be spawned only after \co{r}'s lifetime
+starts and joined before it ends.
+Placing \co{r} in static global storage would avoid lifetime considerations
+totally.
+And there are of course any number of other synchronization schemes
+that might be used to ensure that pointers are in use only while the
+pointed-to objects are alive.
+
+Note that \co{std::unreachable()} does not suffice because the CPU
+can also reorder relaxed stores.
+Note also that load-store ordering is not sufficient to prevent a
+pointer to \co{r} from being stored to \co{b} prior to the start of
+\co{r}'s lifetime because there is no atomic load between the start
+of that lifetime and the relaxed store to \co{b}.
+
+There are a number of ways to prevent carrying a pointer to \co{r} after
+the lifetime of \co{r} has ended, and these are left as an exercise for
+the reader.
+
+\clearpage
+
+\section{Evaluating sdep Using \co{cbmc}}
+\label{sec:Evaluating sdep Using cbmc}
+
+This section demonstrates the evaluation of sdep using
+the C Bounded Model Checker
+(\co{cbmc})~\cite{EdmundClarke2004CBMC}.
+This tool takes a smallish C program as input and constructs a logic
+expression representing that program.
+The logic expression's variables represent bits in the program's inputs
+and state, and the result of that logic expression is \co{true} if that
+program has a bug, for example, if it is possible for that C program to
+trigger an assertion or execute an out-of-bounds array reference.
+The tool then hands this expression to a SAT solver, and if this solver
+finds a satisfying set of values, the tool uses those values provide
+the execution path to the bug.
+
+But first, Appendix~\ref{app:Informal Definition of Semantic Dependency}
+outlines an informal informal definition of ``semantic dependency''
+that can be used in \co{cbmc} test cases.
+
+\subsection{Informal Definition of Semantic Dependency}
+\label{app:Informal Definition of Semantic Dependency}
+
+The manual analysis uses the informal rule that a semantic dependency
+exists between a set of relaxed atomic loads\footnote{
+	Please see
+	Appendix~\ref{app:Non-Trivial Semantic Dependencies}
+	for examples showing why this is a set of loads rather than a
+	single load.}
+and a store on any execution where different values loaded would change
+the address stored to, the value stored, or the number of times that
+store would be executed, with the proviso that stores of the same value
+to the same address having the same ordering properties are fungible.
+
+Producing a precise formal statement of this rule turns out to be quite
+challenging, but as noted earlier, there is considerable work along
+various
+lines~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA}.
+For examples of the challenges, please see
+Section~\ref{sec:On OOTA Definitions}.
+Our informal definition meets these challenges as follows:
+
+\begin{enumerate}
+\item	Section~\ref{sec:Semantic Dependencies and Source Code}
+	is responsible for the phrase ``on any execution'' as well
+	as the proviso regarding fungible loads.
+\item	Section~\ref{sec:Semantic Dependencies Depend on Context}'s
+	challenges are met by looking at the full context within
+	a given thread.
+\item	Section~\ref{sec:Semantic Dependencies Can Be Nondeterministic}'s
+	non-determinism challenges are avoided by avoiding litmus tests
+	exhibiting non-deterministic behavior.
+	Again, note that most production C++ code uses compiler-based C++
+	implementations, which freeze their non-deterministic choices
+	into the binary at compile time.
+\item	Section~\ref{sec:Semantic Dependencies And Optimization}'s
+	optimization-choice challenges are met by using the C++
+	implementation's choice of semantic dependency based on the
+	optimizations that the implementation actually applied.
+\item	Section~\ref{sec:Semantic Dependencies and I/O}'s
+	I/O challenges are met by again noting that I/O is observed
+	behavior.
+	And by avoiding litmus tests involving I/O, given that relatively
+	few of the available tools handle I/O.
+\item	Section~\ref{sec:Semantic Dependencies Can Be Complex}'s
+	complexity challenges are met by looking at what the code
+	actually does.
+	For example, if an as-yet unproven mathematical conjecture
+	holds, the code will act consistently with that conjecture,
+	and vice versa.
+\item	Section~\ref{sec:Implementers and Users Influence Semantic-Dependencies Definition}'s
+	implementer-/user-choice challenges are met by accounting for
+	those choices.
+	For example, if a user has specified \co{-funsigned-char},
+	then the actual executions will be consistent with that choice.
+	For more examples, please see
+	Appendix~\ref{app:User Influence Over Language Semantics}.
+\end{enumerate}
+
+In short, we are treating ``semantic dependency'' in a manner similar
+to our treatment of the laws of physics.
+We let the actual executions provide an empirical basis for the analysis.
+
+We do not expect this approach to please everyone.
+However, those who are displeased might do well to reflect on the fact
+that the C++ standard is not mathematical in nature, which implies that
+any mathematical model of C++ is inherently an approximation.
+We look forward to the day when the standard gains more mathematical
+rigor, and perhaps the C++ memory model is a first step in that
+direction.\footnote{
+	But if so, the OOTA situation should be a caution!}
+Nevertheless, until that day, we must live with the standard in its
+current decidedly non-mathematical form.
+
+\subsection{Complex Semantic Dependencies}
+\label{app:Complex Semantic Dependencies}
+
+Listing~\ref{lst:Example Code for Evaluation of sdep}
+shows a \co{herd7} thread \co{P0()}.
+Is there a semantic dependency between the load from \co{x} on
+line~3 to the store to \co{y} on line~9?
+Answering this question requires solving a mathematical equation,
+not all of which currently have known solutions.
+However, this section will use a relatively simple equation combined
+with restrictions on domain and range that permit \co{cbmc} to do
+the job.
+
+\begin{listing}[tbp]
+\scriptsize
+\begin{verbatim}
+ 1 P0(atomic_int *x, atomic_int *y, atomic_int *limit)
+ 2 {
+ 3   r1 = atomic_load_explicit(x, memory_order_relaxed);
+ 4   r2 = atomic_load_explicit(limit, memory_order_relaxed);
+ 5   r3 = r1 * r1 + 2 * r1 + 2;
+ 6   r4 = r2;
+ 7   if (r2 >= r3)
+ 8     r4 = r3;
+ 9   atomic_store_explicit(y, r4, memory_order_relaxed);
+10 }
+\end{verbatim}
+\caption{Example Code for Evaluation of sdep}
+\label{lst:Example Code for Evaluation of sdep}
+\end{listing}
+
+\begin{listing}[tbp]
+\scriptsize
+\begin{verbatim}
+ 1 int main(int argc, char *argv[])
+ 2 {
+ 3   int i;
+ 4   int y[2] = { 0, 0 };
+ 5
+ 6   for (i = 0; i < 2; i++) {
+ 7     int r1;
+ 8     int r2;
+ 9     int r3;
+10     int *yp = &y[i];
+11
+12     r1 = limited_arg(argv[2 * i + 1]);
+13     r2 = const_arg(limited_arg(argv[2 * i + 2]));
+14     r3 = r1 * r1 + 2 * r1 + 2;
+15     *yp = r2 <= r3 ? r2 : r3;
+16     printf("i = %d, r1 = %d, r2 = %d, r3 = %d\n", i, r1, r2, r3);
+17   }
+18   assert(y[0] == y[1]);
+19 }
+\end{verbatim}
+\caption{Code for cbmc-Based Evaluation of sdep}
+\label{lst:Code for cbmc-Based Evaluation of sdep}
+\end{listing}
+
+One way to answer this question is to create a C program
+(see Listing~\ref{lst:Code for cbmc-Based Evaluation of sdep})
+that runs the \co{P0()} thread's code twice with arbitrary inputs
+(the loop spanning lines~6-17) and asserts that the results of
+each iteration be equal (line~18).
+
+Running this with:
+
+\begin{quote}
+\scriptsize
+\begin{verbatim}
+cbmc -DCBMC cbmc/sdep-quadratic.c
+\end{verbatim}
+\end{quote}
+
+gives the result \co{VERIFICATION FAILED}.
+This indicates that there are different values of \co{x} and \co{limit}
+that yield different values of \co{y}, that is, that there can be
+a semantic dependency.
+
+\begin{listing}[tbp]
+@@ DisplayLitmus cbmc/sdep-quadratic.c @@
+\caption{Full Code for cbmc-Based Evaluation of sdep}
+\label{lst:Full Code for cbmc-Based Evaluation of sdep}
+\end{listing}
+
+However, $x^2 + 2x + 2$ has a minimum value of 1 at $x=-1$, so
+forcing \co{limit} to (say) zero should eliminate this static
+dependency:\footnote{
+	See Listing~\ref{lst:Full Code for cbmc-Based Evaluation of sdep}
+	for the full \path{cbmc/sdep-quadratic.c} program.}
+
+\begin{quote}
+\scriptsize
+\begin{verbatim}
+cbmc -DCBMC -DLOAD_CONSTANT=0 cbmc/sdep-quadratic.c
+\end{verbatim}
+\end{quote}
+
+But this also gives the result \co{VERIFICATION FAILED}.
+One reason for this is that this program does 32-bit computer arithmetic,
+and is thus subject to wraparound.
+Limiting the value of \co{x} to 46,339 prevents wraparound:
+
+\begin{quote}
+\scriptsize
+\begin{verbatim}
+cbmc -DCBMC -DLOAD_CONSTANT=0 -DLOAD_LIMITS=46339 cbmc/sdep-quadratic.c
+\end{verbatim}
+\end{quote}
+
+And this yields \co{VERIFICATION SUCCESSFUL}, indicating that there
+is no semantic dependency in this case.
+This is to be expected because the specified value of zero for \co{limit}
+is smaller than the minimum of the quadratic expression, so that the
+value stored to \co{y} is always zero.
+Specifying \co{-DLOAD_CONSTANT=1} also yields \co{VERIFICATION SUCCESSFUL}
+because this value is equal to that minimum, so that the value stored to
+\co{y} is always 1.
+However, specifying any larger value (for example, \co{-DLOAD_CONSTANT=2})
+yields \co{VERIFICATION FAILED} because the value assigned to \co{y}
+might be either 1 or 2, indicating that a semantic dependency exists.
+
+This situation underscores the point that the existance of a semantic
+dependency can be a property of a particular execution rather than a
+universal property of the code itself.
+The interested reader is encouraged to create more complex examples,
+for example, demonstrating semantic dependencies from multiple loads or
+exploring indeterminacy.
+
+There are numerous other tools that can be used, up to and including
+manual inspection, but \co{cbmc} is readily available and easy to
+use.\footnote{
+	On Debian-based distributions, via \co{sudo apt install cbmc}.}
+
+\subsection{C++ Non-Determinism}
+\label{app:C++ Non-Determinism}
+
+The C++ standard allows non-determinism, for example, the order of
+invocation of side-effects in some circumstances.
+Listing~\ref{lst:Code for cbmc-Based Evaluation of Non-Deterministic sdep}
+shows an example leveraging side-effects in the parameter list shown on
+line~37.
+
+\begin{listing}[tbp]
+@@ DisplayLitmus cbmc/sdep-nondet.c @@
+\caption{Code for cbmc-Based Evaluation of Non-Deterministic sdep}
+\label{lst:Code for cbmc-Based Evaluation of Non-Deterministic sdep}
+\end{listing}
+
+Running \co{cbmc} on this as follows:
+
+\begin{quote}
+\scriptsize
+\begin{verbatim}
+cbmc -DCBMC cbmc/sdep-nondet.c
+\end{verbatim}
+\end{quote}
+
+Gives the result \co{VERIFICATION SUCCESSFUL}, showing that \co{cbmc}
+does not model this non-determinism.
+Which is not surprising, given that many of the safety-critical uses
+to which \co{cbmc} is put would prohibit such non-determinism in their
+code bases.
+
+\clearpage
+
 \section{Debates on Causality}

### 2024-02-13 ootavacuous.tex
Move informal sdep definition to cbmc appendix
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-02-12 ootavacuous.tex
@@
 otherwise gave us pause.
+
+\subsection{Litmus Tests from “Causality Test Cases"}
+\label{app:Litmus Tests from “Causality Test Cases"}
+
+These tests might be two decades old, but they are still quite
+relevant.\footnote{
+	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
+
+\subsubsection{Causality Test Case 1}
+\label{app:Causality Test Case 1}
+
+Listing~\ref{lst:Causality Test Case 1}
+shows causality test case 1, for which the \co{r1 == r2 == 1} result
+is to be allowed.
+The \co{herd7} tool finds this litmus test's cycle, even without the
+inter-thread analysis, because C++ does not respect control dependencies
+to relaxed atomic stores.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-1.litmus @@
+\caption{Causality Test Case 1}
+\label{lst:Causality Test Case 1}
+\end{listing}
+
+Implementations that can prove that the value of \co{x} is always
+non-negative can also prove that there is no semantic dependency between
+\co{P0()}'s load from \co{x} on line~8 to its store to \co{y} on line~10,
+that is, no matter which of the possible values is loaded from \co{x},
+the value 1 will always be stored to \co{y}.
+
+Less-omniscient implementations, including those whose code analysis
+is restricted to a thread, would need to act as if there was in fact a
+semantic dependency.
+
+Even omniscient implementations will observe a semantic dependency from
+\co{P1()}'s load from \co{y} on line~14 to its store to \co{x} on line~15.
+
+Either way, there is no OOTA cycle, with omniscient implementations
+having no cycle at all, and other implementations retaining the
+non-semantic dependency.
+
+\subsubsection{Causality Test Case 2}
+\label{app:Causality Test Case 2}
+
+Listing~\ref{lst:Causality Test Case 2}
+shows causality test case 2, for which the \co{r1 == r2 == r3 == 1} result
+is to be allowed.
+The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-2.litmus @@
+\caption{Causality Test Case 2}
+\label{lst:Causality Test Case 2}
+\end{listing}
+
+Because \co{P0()}'s loads from \co{x} on lines~8 and~9 are unordered,
+an implementation can act as if the source code loaded from \co{x} only
+once and assigned the returned value to both \co{r1} and \co{r2}.\footnote{
+	But not in C, where relaxed atomic operations are volatile,
+	and thus observable behavior.}
+The implementation could then prove that the condition on line~10
+would always evaluate to \co{true}, and act as if the source code had
+unconditionally executed the relaxed store to \co{y} on line~11.
+The implementation could then act as if the resulting relaxed accesses
+to \co{x} and \co{y} appeared in reverse order, at which point
+\co{P0()}'s store to \co{y} could be read by \co{P1()} load so that
+\co{P1()} stored 1 to \co{x}, thus resulting in \co{P1()} loading
+that 1 from \co{x}.
+
+Note that \co{P1()} has a trivial semantic dependency from its load
+from \co{y} on line~15 to its store to \co{x} on line~16.
+
+If \co{P0()}'s loads from \co{x} are fused, there is no cycle,
+otherwise the preserved non-semantic dependency prevents an
+OOTA cycle from forming on correct real-world implementations.
+
+\subsubsection{Causality Test Case 3}
+\label{app:Causality Test Case 3}
+
+Listing~\ref{lst:Causality Test Case 3}
+shows causality test case 3, for which the \co{r1 == r2 == r3 == 1} result
+is to be allowed.
+The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
+
+The analysis from Section~\ref{app:Causality Test Case 2}
+applies here, including the difference in volatility between C and C++
+atomics.
+Although the addition of \co{P2} allows sequentially consistent executions
+where \co{P0()}'s two loads from \co{x} might return different values,
+it does not change the fact that those two loads can be merged.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-3.litmus @@
+\caption{Causality Test Case 3}
+\label{lst:Causality Test Case 3}
+\end{listing}
+
+\subsubsection{Causality Test Case 4}
+\label{app:Causality Test Case 4}
+
+Listing~\ref{lst:Causality Test Case 4}
+shows causality test case 4, for which the \co{r1 == r2 == 1} result
+is to be forbidden.
+The \co{herd7} tool finds this OOTA cycle because C++ does not respect
+data dependencies, and flags it as such using \co{S8}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-4.litmus @@
+\caption{Causality Test Case 4}
+\label{lst:Causality Test Case 4}
+\end{listing}
+
+Both \co{P0()} and \co{P1()} have straightforward semantic dependencies
+from their relaxed atomic loads to their relaxed atomic stores, which
+prevents this OOTA cycle on correct real-world C++ implementations.
+
+\subsubsection{Causality Test Case 5}
+\label{app:Causality Test Case 5}
+
+Listing~\ref{lst:Causality Test Case 5}
+shows causality test case 5, for which the \co{r1 == r2 == 1 && r3 == 0}
+result is to be forbidden.
+The \co{herd7} tool finds this OOTA cycle because C++ does not respect
+data dependencies, and flags it as such using \co{S8}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-5.litmus @@
+\caption{Causality Test Case 5}
+\label{lst:Causality Test Case 5}
+\end{listing}
+
+Each of \co{P0()}, \co{P1()}, and \co{P3()} have trivial semantic
+dependencies from their relaxed atomic loads to their relaxed
+atomic stores, which prevents any OOTA cycle from forming in correct
+real-world C++ implementations.
+\co{P2()} has no loads, so cannot participate in an OOTA cycle in
+any case.
+
+\subsubsection{Causality Test Case 6}
+\label{app:Causality Test Case 6}
+
+Listing~\ref{lst:Causality Test Case 6}
+shows causality test case 6, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The \co{herd7} tool finds this cycle, and does not flag it as an OOTA
+cycle because C++ does not respect control dependencies to relaxed
+atomic stores.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-6.litmus @@
+\caption{Causality Test Case 6}
+\label{lst:Causality Test Case 6}
+\end{listing}
+
+A C++ implementation that can prove that \co{y} is always either
+zero or one can also prove that there is no semantic dependency
+from \co{P1()}'s load on line~14 to its stores on lines~16 and~29.
+Such an implementation could act as if lines~15-20 had been replaced
+by a single store to \co{x}, allowing that store to execute first,
+followed by lines~8, 10, and~14, in that order, satisfying the
+\co{exists} clause on line~23.
+
+Less-omniscient implementations will preserve the non-semantic
+dependency, preventing any OOTA loop from forming in a correct real-world C++
+implementation.
+
+\subsubsection{Causality Test Case 7}
+\label{app:Causality Test Case 7}
+
+Listing~\ref{lst:Causality Test Case 7}
+shows causality test case 7, for which the \co{r1 == r2 == r3 == 1}
+result is to be allowed.
+The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-7.litmus @@
+\caption{Causality Test Case 7}
+\label{lst:Causality Test Case 7}
+\end{listing}
+
+Because all accesses are relaxed atomics, this test case has everything
+to do with reordering and nothing to do with OOTA cycles.
+The outcome happens when lines~17, 10, 11, 15, 16 and~9 execute in that
+order.
+Note that this ordering preserves the semantic dependencies from
+lines~10-11 and~15-16, and can thus happen on correct real-world
+C++ implementations.
+
+\subsubsection{Causality Test Case 8}
+\label{app:Causality Test Case 8}
+
+Listing~\ref{lst:Causality Test Case 8}
+shows causality test case 8, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The \co{herd7} tool does not find this cycle, but when \co{r2} is explicitly
+initialized to the constant \co{1}, it does find the cycle and refrains
+from flagging it as an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-8.litmus @@
+\caption{Causality Test Case 8}
+\label{lst:Causality Test Case 8}
+\end{listing}
+
+Omniscient C++ implementations might note that the computation on line~9
+can only have the value one when the load from \co{x} returns either
+zero or one, and both \co{x} and \co{y} are initialized to zero.
+Such an implementation act as if line~10's \co{r2} had instead been
+the value~1, so that the value returned by \co{P0()}'s load could not
+affect the value stored.
+Then there would be no semantic dependency between \co{P0()}'s load
+and store, and thus no OOTA cycle in a correct real-world C++
+implementation.
+
+A less-capable implementation would leave the non-semantic dependency
+in place, also preventing any OOTA cycle in a correct real-world C++
+implementation.
+
+\subsubsection{Causality Test Case 9}
+\label{app:Causality Test Case 9}
+
+Listing~\ref{lst:Causality Test Case 9}
+shows causality test case 9, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The \co{herd7} tool does not find this cycle because it depends on
+scheduling \co{P2()} after the other two threads, and C++ is not
+constrained to do this.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-9.litmus @@
+\caption{Causality Test Case 9}
+\label{lst:Causality Test Case 9}
+\end{listing}
+
+Both threads have straightforward semantic dependencies between their
+respective relaxed atomic loads and relaxed atomic stores, so an OOTA
+cycle cannot form in a correct real-world C++ implementation.
+
+\subsubsection{Causality Test Case 9a}
+\label{app:Causality Test Case 9a}
+
+Listing~\ref{lst:Causality Test Case 9a}
+shows causality test case 9a, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The \co{herd7} tool does not find this cycle because it depends on
+scheduling \co{P2()} before the other two threads, and C++ is not
+constrained to do this.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-9a.litmus @@
+\caption{Causality Test Case 9a}
+\label{lst:Causality Test Case 9a}
+\end{listing}
+
+Both threads have straightforward semantic dependencies between their
+respective relaxed atomic loads and relaxed atomic stores, so an OOTA
+cycle cannot form in a correct real-world C++ implementation.
+
+\subsubsection{Causality Test Case 10}
+\label{app:Causality Test Case 10}
+
+Listing~\ref{lst:Causality Test Case 10}
+shows causality test case 10, for which the \co{r1 == r2 == 1 && r3 == 0}
+result is to be forbidden.
+Because C++ does not respect control dependencies, the \co{herd7} tool
+nevertheless reports this result and does not flag it as an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-10.litmus @@
+\caption{Causality Test Case 10}
+\label{lst:Causality Test Case 10}
+\end{listing}
+
+However, each of \co{P0()}, \co{P1()}, and \co{P3()} have straightforward
+semantic dependencies between their respective loads and stores, and
+thus cannot form an OOTA cycle when running on a correct real-world C++
+implementation.
+For its part, \co{P2()} has no semantic dependency, so it cannot be part
+of an OOTA cycle when running on a correct real-world C++ implementation.
+
+\subsubsection{Causality Test Case 11}
+\label{app:Causality Test Case 11}
+
+Listing~\ref{lst:Causality Test Case 11}
+shows causality test case 11, for which the \co{r1 == r2 == r3 == r4 == 1}
+result is to be allowed.
+The \co{herd7} tool reports this result and does not flag it as an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-11.litmus @@
+\caption{Causality Test Case 11}
+\label{lst:Causality Test Case 11}
+\end{listing}
+
+This test case has nothing to do with OOTA cycles.
+The result may be obtained by executing the statements in the following
+order: line~20, line~12, line~13, line~18, line~19, line~10, line~11,
+and finally line~17.
+
+\subsubsection{Causality Test Case 12}
+\label{app:Causality Test Case 12}
+
+JMM Causality Test Case 12 in
+Listing~\ref{lst:Causality Test Case 12}
+uses arrays, which are not yet supported by the \co{herd7} tool.
+
+\begin{listing}[tbp]
+@@ DisplayLitmus litmus/oota-causality-12.litmus @@
+\caption{Causality Test Case 12}
+\label{lst:Causality Test Case 12}
+\end{listing}
+
+An omniscient C++ implementation could note that only \co{P0()} accesses
+array \co{a[]} and act as if the code was simpler, but there would
+still be semantic dependencies between \co{P0()}'s and \co{P1()}s
+relaxed atomic loads and stores involving \co{x} and \co{y}.
+Those dependencies prevent an OOTA cycle from forming on a correct
+real-world C++ implementation, omniscient or not.
+
+\subsubsection{Causality Test Case 13}
+\label{app:Causality Test Case 13}
+
+Listing~\ref{lst:Causality Test Case 13}
+shows causality test case 13, for which the \co{r1 == r2 == 1}
+result is to be forbidden.
+But because C++ does not respect control dependencies, \co{herd7} tool
+reports this result and does not flag it as an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-13.litmus @@
+\caption{Causality Test Case 13}
+\label{lst:Causality Test Case 13}
+\end{listing}
+
+There are straightforward semantic dependencies between \co{P0()}'s and
+\co{P1()}s relaxed atomic loads and stores involving \co{x} and \co{y}.
+Those dependencies prevent an OOTA cycle from forming on a correct
+real-world C++ implementation.
+
+\subsubsection{Causality Test Case 14}
+\label{app:Causality Test Case 14}
+
+Listing~\ref{lst:Causality Test Case 14}
+shows causality test case 14, for which the \co{r1 == r3 == 1 && r2 == 0}
+result is to be forbidden.
+The \co{herd7} tool nevertheless reports this result because C++ does not
+respect control dependencies for atomic stores.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-14.litmus @@
+\caption{Causality Test Case 14}
+\label{lst:Causality Test Case 14}
+\end{listing}
+
+In C++, this is not a matter of OOTA but rather reordering.
+Dependencies notwithstanding, according to the C++ memory model,
+the statements could execute in the following order:
+Line~17 (so \co{r2 == 0}), line~21, line~9 (so \co{r1 == 1}),
+line~13, line~18 (so \co{r3 == 1)}.
+The sum of \co{r2} and \co{r3} is non-zero, justifying line~21's
+store to \co{x}.
+
+However, correct real-world C++ implementations would have semantic
+dependencies between \co{P0()}'s line~9 and lines~11 and~13, and also
+between \co{P1()}'s lines~17 and~18 and line~21, preventing any
+OOTA cycle from forming.
+
+\subsubsection{Causality Test Case 15}
+\label{app:Causality Test Case 15}
+
+Listing~\ref{lst:Causality Test Case 15}
+shows causality test case 15, for which the \co{r0 == r1 == r3 == 1 && r2 == 0}
+result is to be forbidden.
+The \co{herd7} tool nevertheless reports this result because C++ does not
+respect control dependencies for atomic stores.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-15.litmus @@
+\caption{Causality Test Case 15}
+\label{lst:Causality Test Case 15}
+\end{listing}
+
+In C++, this is not a matter of OOTA but rather reordering.
+Dependencies notwithstanding, according to the C++ memory model,
+the statements could execute in the following order:
+Line~31, line~27, line~10 (so \co{r0 == 1}), line~13 (so \co{r1 == 1}),
+line~19 (because \co{r1 == 1}), line~23 (so \co{r2 == 0}),
+and finally line~24 (so \co{r3 == 1}).
+The sum of \co{r2} and \co{r3} is non-zero, justifying line~27's
+store to \co{u}.
+
+However, correct real-world C++ implementations would have semantic
+dependencies between \co{P0()}'s lines~10 and~13 and lines~17 and~19, and also
+between \co{P1()}'s lines~23 and~24 and line~27, preventing any
+OOTA cycle from forming.
+Having no load, \co{P2()} cannot participate in an OOTA cycle.
+
+\subsubsection{Causality Test Case 16}
+\label{app:Causality Test Case 16}
+
+Listing~\ref{lst:Causality Test Case 16}
+shows causality test case 16, for which the \co{r1 == 2 && r2 == 1}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be expected
+given that C++ does not permit a thread to reorder accesses to a single
+atomic variable, even when those accesses are relaxed.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-16.litmus @@
+\caption{Causality Test Case 16}
+\label{lst:Causality Test Case 16}
+\end{listing}
+
+\subsubsection{Causality Test Case 17}
+\label{app:Causality Test Case 17}
+
+Listing~\ref{lst:Causality Test Case 17}
+shows causality test case 17, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be expected
+given that C++ does not permit a thread to reorder accesses to a single
+atomic variable, even when those accesses are relaxed.
+But note that the \co{herd7} tool correctly notes the presence of an
+OOTA cycle with \co{S17}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-17.litmus @@
+\caption{Causality Test Case 17}
+\label{lst:Causality Test Case 17}
+\end{listing}
+
+\co{P1()} has a straightforward semantic dependency between its
+relaxed atomic load and relaxed atomic store.
+However, suppose that \co{P0()}'s load from \co{x} on line~8 returns
+the value 42.
+In that case, the second load on line~11 can be fused with the first,
+so that it also returns the value 42.
+On the other hand, if the load on line~8 returns some other value, then
+line~10 will store the value 42 to \co{x}, and then the load on line~11
+can be elided\footnote{
+	In C++, but not in C due to atomics being volatile.}
+in favor of using the value 42 stored by line~10.
+
+Either way, the value of \co{r1} will be the constant 42.
+Therefore, the loads from \co{x} cannot affect the store on line~17,
+so there is no semantic dependency.
+This allows that store to \co{y} to be moved to the very beginning of
+\co{P0()}, thus allowing a correct real-world C++ implementation to
+satisfy the \co{exists} clause.
+
+\subsubsection{Causality Test Case 18}
+\label{app:Causality Test Case 18}
+
+Listing~\ref{lst:Causality Test Case 18}
+shows causality test case 18, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be expected
+given that C++ does not permit a thread to reorder accesses to a single
+atomic variable, even when those accesses are relaxed.
+But note that the \co{herd7} tool correctly notes the presence of an
+OOTA cycle with \co{S17}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-18.litmus @@
+\caption{Causality Test Case 18}
+\label{lst:Causality Test Case 18}
+\end{listing}
+
+Both threads have straightforward semantic dependencies between their
+respective relaxed atomic loads and relaxed atomic stores, so an OOTA
+cycle cannot form in a correct real-world C++ implementation.
+
+\subsubsection{Causality Test Case 19}
+\label{app:Causality Test Case 19}
+
+Listing~\ref{lst:Causality Test Case 19}
+shows causality test case 19, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be expected
+given that C++ does not permit a thread to reorder accesses to a single
+atomic variable, even when those accesses are relaxed.
+But note that the \co{herd7} tool correctly notes the presence of an
+OOTA cycle with \co{S8}.
+
+It is interesting that this \co{S8} value does not propagate to
+\co{r3}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-19.litmus @@
+\caption{Causality Test Case 19}
+\label{lst:Causality Test Case 19}
+\end{listing}
+
+All threads have straightforward semantic dependencies between their
+respective relaxed atomic loads and relaxed atomic stores, so an OOTA
+cycle cannot form in a correct real-world C++ implementation.
+
+\subsubsection{Causality Test Case 20}
+\label{app:Causality Test Case 20}
+
+Listing~\ref{lst:Causality Test Case 20}
+shows causality test case 20, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be expected
+given that C++ does not permit a thread to reorder accesses to a single
+atomic variable, even when those accesses are relaxed.
+But note that the \co{herd7} tool correctly notes the presence of an
+OOTA cycle with \co{S8}.
+
+It is interesting that this \co{S8} value does not propagate to
+\co{r3}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-20.litmus @@
+\caption{Causality Test Case 20}
+\label{lst:Causality Test Case 20}
+\end{listing}
+
+All threads have straightforward semantic dependencies between their
+respective relaxed atomic loads and relaxed atomic stores, so an OOTA
+cycle cannot form in a correct real-world C++ implementation.
+
+\subsection{Additional Litmus Tests}
+\label{app:Additional Litmus Tests}
+
+These litmus tests helped illuminate important aspects of the problem
+of defining sdep and identifying OOTA cycles.
+
+\subsubsection{Simple Reordering as Litmus Test}
+\label{app:Simple Reordering as Litmus Test}
+
+Listing~\ref{lst:Simple Reordering as Litmus Test}
+shows how
+Listing~\ref{lst:Simple Reordering}
+can be recast into a \co{herd7} litmus test.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/simple-reordering.litmus @@
+\caption{Simple Reordering as Litmus Test}
+\label{lst:Simple Reordering as Litmus Test}
+\end{listing}
+
+\subsubsection{Non-LB Litmus Tests}
+\label{app:Non-LB Litmus Tests}
+
+Listing~\ref{lst:Non-LB Litmus Test}
+shows a litmus test that can produce an OOTA cycle, but that is not
+a member of the load-buffering (LB) class of litmus tests.
+Suppose that lines~9 and~10 both somehow return the value 21.
+Then line~12 will store the value 42 to \co{x}, which will cause
+line~17 to store 21 to \co{y} and line~22 to store 21 to \co{z},
+thus justifying the values initially loaded by lines~9 and~10.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-non-lb.litmus @@
+\caption{Non-LB Litmus Test}
+\label{lst:Non-LB Litmus Test}
+\end{listing}
+
+All thread's loads' return values affect the value stored, so all
+threads have semantic dependencies between their respective relaxed
+atomic loads and relaxed atomic stores.
+Therefore, an OOTA cycle cannot form in a correct real-world C++
+implementation.
+
+Note that in a C++ implementation that (erroneously) permitted an OOTA
+cycle to form, that cycle would involve both \co{r0} and \co{r1}.
+However, we conjecture that tools designed to find OOTA cycles could
+satisfy themselves with having found either of the two resulting cycles.
+
+\subsubsection{Multi-Cycle Litmus Tests}
+\label{app:Multi-Cycle Litmus Tests}
+
+Listing~\ref{lst:Multi-Cycle Litmus Test}
+shows an expansion to four threads of Causality Test Case 4, which is shown in
+Listing~\ref{lst:Causality Test Case 4}
+and discussed in
+Appendix~\ref{app:Causality Test Case 4}.
+The \co{herd7} tool shows multiple different possible OOTA cycles,
+as expected.
+However, the value returned by each thread's load directly affects the
+value stored, so there is a semantic dependency in each thread.
+Therefore, none of these OOTA cycles can happen in a correct real-world
+C++ implementation.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-two-source.litmus @@
+\caption{Multi-Cycle Litmus Test}
+\label{lst:Multi-Cycle Litmus Test}
+\end{listing}
+
+\subsubsection{Three-Process Version of JMM Causality Test Case 4}
+\label{app:Three-Process Version of JMM Causality Test Case 4}
+
+Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
+shows a different way of expanding
+Listing~\ref{lst:Causality Test Case 4}
+(discussed in
+Appendix~\ref{app:Causality Test Case 4})
+from two threads to three threads.
+The \co{herd7} tool finds OOTA values, as expected.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-3proc.litmus @@
+\caption{Three-Process Version of JMM Causality Test Case 4}
+\label{lst:Three-Process Version of JMM Causality Test Case 4}
+\end{listing}
+
+All threads have straightforward semantic dependencies from their
+relaxed atomic loads to their relaxed atomic stores, which prevents this
+OOTA cycle on correct real-world C++ implementations.
+
+\subsubsection{Why rfe Instead of Tried-And-True rf?}
+\label{app:Why rfe Instead of Tried-And-True rf?}
+
+Although the examples presented in
+Section~\ref{sec:OOTA: rf vs. rfe}
+(and shown as litmus tests in
+Listings~\ref{lst:OOTA With an Intra-Process Store-Load Link}
+and~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized})
+demonstrate that rf can be atemporal, they demonstrate neither
+OOTA cycles nor OOTA-like behavior in full C++ implementations.
+This appendix provides more involved examples that can be optimized
+so as to demonstrate OOTA-like behavior.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-3-2-proc.litmus @@
+\caption{OOTA With an Intra-Process Store-Load Link}
+\label{lst:OOTA With an Intra-Process Store-Load Link}
+\end{listing}
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-3-2-proc-opt.litmus @@
+\caption{OOTA With an Intra-Process Store-Load Link, Optimized}
+\label{lst:OOTA With an Intra-Process Store-Load Link, Optimized}
+\end{listing}
+
+Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
+revises
+Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
+so that the value stored to \co{y} is forced to be at most the
+value 17, and the value stored to \co{z} is forced to be at least the
+value 17.
+This of course means that the store to \co{z} must always be the value
+17, as can be seen from the \co{herd7} output.
+However, a single-thread-at-a-time analysis could not prove this.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-whyrfe-3.litmus @@
+\caption{Three-Process Version of JMM Causality Test Case 4 With Max and Min}
+\label{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
+\end{listing}
+
+Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
+further revises
+Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
+to flatten \co{PO()} and \co{P1()} into a single thread, so that the
+resulting \co{P0()} has an rfi link instead of an rfe link connecting
+the store to and load from \co{y}.
+This means that a C++ implementation could prove that the value 17 is
+always stored to \co{z} by examining \co{P0()} in isolation, given that
+it is also legal for a C++ implementation to replace the value loaded
+from \co{y} with the value just stored.\footnote{
+	This is not an option in C because atomic operations are volatile,
+	and thus observed behavior.}
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-whyrfe.litmus @@
+\caption{Why rfe Instead of Tried-And-True rf?}
+\label{lst:Why rfe Instead of Tried-And-True rf?}
+\end{listing}
+
+In the confines of the C++ abstract machine (omitting the no-OOTA
+restriction), this allows a value-17 OOTA cycle, courtesy of the fact
+that C++ respects neither control nor data dependencies.
+However, the \co{herd7} tool does not identify this cycle, perhaps due
+to its restrictions on its handling of arithmetic on OOTA values.
+
+In a real-world implementation that proves that the value 17 is always
+stored to \co{z}, there is no semantic dependency to
+\co{P0()}'s store to \co{z}, but there is to all of the other relaxed
+atomic stores.
+On the other hand, if the implementation does not prove the value
+stored to \co{z}, there will be a global-time delay between each of
+\co{P0()}'s loads and its store to \co{z}, preventing a value-17 OOTA
+cycle from forming.
+
+Either way, the value-17 OOTA cycle cannot occur in a correct real-world
+C++ implementation.
+Yes, an implementation that proves that the value 17 is always
+stored to \co{z} can satisfy the \co{exists} clause, but because
+there is no semantic dependency to \co{P0()}'s store to \co{z},
+this is not an OOTA cycle.
+
+But what if we chose to define an OOTA cycle as (sdep $\cup$ rf) instead
+of (sdep $\cup$ rfe)?
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-whyrfe-z17.litmus @@
+\caption{Why rfe Instead of Tried-And-True rf When z=17?}
+\label{lst:Why rfe Instead of Tried-And-True rf When z=17?}
+\end{listing}
+
+In that case, it would be necessary to separately consider the dependency
+from \co{P0()}'s load from \co{x} and \co{y} and that from its load
+from \co{y} and its store to \co{z}.
+The problem with this is shown in
+Listing~\ref{lst:Why rfe Instead of Tried-And-True rf When z=17?},
+which shows an equivalent program given the knowledge that the value~17
+will always be stored to \co{z}.
+Because all atomic operations are relaxed and because C++ does not
+respect control dependencies, the store to \co{z} can be moved up
+to the very beginning of \co{P0()}.
+This transformation causes the rfi link associated with \co{y} to
+go backwards in time, which is perfectly legitimate in this case,
+but which makes defining OOTA cycles in terms of (sdep $\cup$ rf)
+problematic.
+
+Again, in C, the load from \co{y} cannot be removed because it is
+volatile, and thus observable behavior.
+C++ could remove that load, but we left it in to demonstrate the all-17s
+outcome.
+
+This example demonstrates why OOTA cycles must be defined in terms
+of (sdep $\cup$ rfe), or in \co{herd7} nomenclature,
+\co{acyclic(sdep; rfe)}.
+
+\subsubsection{Inventing Atomic Loads}
+\label{app:Inventing Atomic Loads}
+
+This section contains several example litmus tests from the active
+mind of Alan Stern.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-no-invented-load.litmus @@
+\caption{No Invented Atomic Loads}
+\label{lst:No Invented Atomic Loads}
+\end{listing}
+
+First, consider Listing~\ref{lst:No Invented Atomic Loads}.
+At first glance, this might appear straightforward.
+\co{P0()}'s store to \co{z} clearly depends on its load from \co{y},
+and \co{P1()}'s store to \co{y} clearly depends on its load from \co{z}.
+If both dependencies are semantic dependency, the OOTA cycle cannot
+form.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-load-invented.litmus @@
+\caption{Atomic Loads Invented}
+\label{lst:Atomic Loads Invented}
+\end{listing}
+
+But is \co{P0()}'s dependency a semantic dependency?
+For example, is a real-world C++ implementation permitted to
+duplicate \co{P0()}'s load from \co{x} and then make use of both loads'
+values?
+If this is permitted, then the implementation would be within its rights
+to transform
+Listing~\ref{lst:No Invented Atomic Loads}
+into
+Listing~\ref{lst:Atomic Loads Invented}.
+
+Because the values of \co{r1}, \co{r2}, and \co{r3} are all known to be
+either zero or one, if the values returned by the two loads from \co{x}
+differ, then one of those two values must be equal to that from \co{y}.
+If the implementation is further permitted to make a runtime choice
+between the two values loaded from \co{x}, it could always choose the
+one that was equal to the value loaded from \co{y}, as shown by the
+\co{if} statement on line~13.
+This would mean that the store to \co{z} was always the constant 1,
+which would permit the store to be ordered before the load from \co{y},
+as shown in lines~14 and~15.
+Doing so permits the OOTA-like outcome of \co{r3 == r4 == 1}.
+
+The implementation might also drop the load from \co{y}, whose value
+is now unused except by the \co{exists} clause.
+However, the \co{exists} clause is considered to be observable behavior,
+which means that the load from \co{y} must remain.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-unused-load.litmus @@
+\caption{Unused Extra Atomic Load}
+\label{lst:Unused Extra Atomic Load}
+\end{listing}
+
+One possible reaction to this situation is to declare that useful
+real-world C++ implementations should never duplicate non-volatile
+relaxed atomic loads.
+But suppose that the initial code was as shown in
+Listing~\ref{lst:Unused Extra Atomic Load}?\footnote{
+	Such unused loads can easily be generated by complex macros or
+	template metaprograms.}
+In this case, the transformation to
+Listing~\ref{lst:Atomic Loads Invented}
+would not require inventing a load, but instead merely inventing a use of
+the previously discarded value in \co{r0}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-invent-int-load.litmus @@
+\caption{Inventing Non-Boolean Atomic Loads}
+\label{lst:Unused Inventing Non-Boolean Atomic Loads}
+\end{listing}
+
+This is not restricted to booleans, as can be seen in
+Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}.
+This example is more straightforward in that it does not make quite so
+creative use of the multiple values loaded.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/inc.litmus @@
+\caption{Only Atomic Increment}
+\label{lst:Only Atomic Increment}
+\end{listing}
+
+But suppose that we drop the single-thread-analysis constraint for
+the moment, and suppose that the C++ implementation is able to prove
+that the only modification of a given atomic variable is to atomically
+increment it, as is the case for \co{y} in
+Listing~\ref{lst:Only Atomic Increment}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/inc-range.litmus @@
+\caption{Only Atomic Increment, Optimized}
+\label{lst:Only Atomic Increment, Optimized}
+\end{listing}
+
+Is a C++ implementation within its rights to behave as if the code
+shown in
+Listing~\ref{lst:Only Atomic Increment}
+was instead as shown in
+Listing~\ref{lst:Only Atomic Increment, Optimized}?
+One argument in favor is that the value of \co{y} must have
+passed through the value 42 in the case where \co{r0} and \co{r1}
+bracket that value.
+Arguments against might cite the added overhead of the invented load
+in Listing~\ref{lst:Only Atomic Increment, Optimized}, or that the
+possibility that the check for the value 42 is intended for statistical
+sampling.
+
+There are a number of possible reactions to these situations:
+
+\begin{enumerate}
+\item	Listing~\ref{lst:No Invented Atomic Loads}
+	looks like it has an OOTA cycle, but there is in fact no
+	semantic dependency.
+\item	Attempts to use real-world constraints to prevent OOTA cycles
+	are futile.
+	(But see Section~\ref{sec:But What About Non-Volatile Atomics?}.)
+\item	Implementations should not invent uses for values from
+	non-volatile relaxed atomic loads.
+	If it turns out that there are useful optimizations that invent
+	such uses, then such optimizations must be applied only with
+	careful attention to the as-if rule.
+	Some might argue that inventing comparisons between a pair of
+	non-volatile relaxed atomic loads from the same object should
+	be permitted only if the two loads could be reordered to be
+	adjacent to each other.
+\item	Any time an implementation might be tempted to invent uses from
+	values from non-volatile relaxed atomic loads from the same
+	object, the loads should instead be fused into a single load.
+	This reaction might be attractive to those who carefully consider
+	the pointlessness of keeping two loads that return the same value
+	on the one hand or the cache-miss overhead incurred when closely
+	spaced loads return different values on the other.
+\item	Implementations that invent non-volatile atomic loads or invent
+	new uses for non-volatile atomic loads can adversely affect
+	observed behavior, for example, by introducing error into code
+	attempting to do statistical sampling.
+\item	What other interesting situations might come to light?
+\end{enumerate}
+
+There are situations where inventing loads from non-atomic objects
+is customary, for example, when hoisting a load out of a loop.
+In this case, implementations might hoist the load from x out of
+the loop:
+
+\begin{quote}
+\begin{verbatim}
+do_something_with(x);
+for (i = 0; i < limit i++)
+	y[i] += x;
+\end{verbatim}
+\end{quote}
+
+A C++ implementation might act as if the program had instead been this:
+
+\begin{quote}
+\begin{verbatim}
+r1 = x;
+do_something_with(r1);
+for (i = 0; i < limit i++)
+	y[i] += r1;
+\end{verbatim}
+\end{quote}
+
+Note that passing \co{x} to the call to \co{do_something_with()} is important
+because this allows the C++ implementation to assume that accesses to \co{x}
+will not result in a data race.
+Without this assumption, the above transformation might introduce a data
+race when \co{limit} is less than or equal to zero.
+
+But if \co{x} is a non-volatile atomic, then there are no worries about
+data races.
+In theory, the implementation could make the above transformation without
+worry, especially given that the value loaded from \co{x} is discarded in
+the case where \co{limit} is less than or equal to zero.
+Except that the data-race worry is replaced by a cache-miss worry.
+After all, optimizations are supposed to speed things up.
+This means that a C++ implementation might be wise to adhere to the
+same restriction that avoids data races for loads from non-atomic
+variables in order to avoid expensive cache misses.
+
+\subsubsection{Non-Trivial Semantic Dependencies}
+\label{app:Non-Trivial Semantic Dependencies}
+
+Most of the examples in this paper involve simple semantic dependencies
+connecting a single load to a single store.
+One exception appears in
+Appendix~\ref{app:Non-LB Litmus Tests},
+but this section will present a more difficult example.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult-0.litmus @@
+\caption{OOTA Multiplication Example, Initial Value Zero}
+\label{lst:OOTA Multiplication Example, Initial Value Zero}
+\end{listing}
+
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+shows an example that builds on the multiplication-by-zero
+discussion in
+Section~\ref{sec:Semantic Dependencies and Source Code}.
+Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
+\co{x} and \co{y} on lines~9 and~10, respectively.
+Similarly, lines~16 and~17 store to \co{x} and \co{y}, respectively,
+the value that line~15 loads from \co{z}.
+Because all three variables are initialized to zero, and because loads
+can be ordered before any store, there is an execution in which all
+three loads return the value zero, in which case all three stores will
+of course store the value zero.
+In any related execution where the value loaded from \co{y} remains zero,
+non-zero values loaded from \co{x} cannot affect the value stored by
+line~11 to \co{z}.
+Only executions which load different (that is, non-zero) values from
+both \co{x} and \co{y} can cause the value stored to \co{z} to change
+from zero to a non-zero value.
+
+Therefore, as noted in
+Section~\ref{sec:Semantic Dependencies and Source Code},
+in an execution where the loads from both \co{x} and \co{y} return zero,
+there is no semantic dependency from the load from \co{x} to the
+store to \co{z}, nor is there a semantic dependency from the load
+from \co{y} to the store to \co{z}.
+However, there \emph{is} a semantic dependency from the \emph{combination}
+of the loads from \co{x} and \co{y} to the store to \co{z}.
+For example, the (forbidden) execution that satisfies the \co{exists}
+clause, which would be an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult-0-cond.litmus @@
+\caption{OOTA Conditional Multiplication Example, Initial Value Zero}
+\label{lst:OOTA Conditional Multiplication Example, Initial Value Zero}
+\end{listing}
+
+For further evidence that there is no semantic dependency from \co{y}
+to \co{z} in executions where the load from \co{x} is zero, see the
+equivalent\footnote{
+	But equivalent only because the atomics are all non-volatile!}
+program shown in
+Listing~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero}.
+Here, there is not even a load from \co{y} in executions where the
+load from \co{x} returns zero, so there cannot possibly be a semantic
+dependency involving this non-existent load.
+A similar transformation would load \co{x} only in executions where the
+load from \co{y} returned non-zero, which similarly emphasizes the lack
+of any semantic dependency between \co{x} and \co{z}.
+
+In both
+Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+and~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero},
+an OOTA cycle (such as the one indicated by their \co{exists} clauses)
+would need to affect both the \co{x} and the \co{y} components of this
+compound semantic dependency.
+However, we again conjecture that tools designed to find OOTA cycles could
+satisfy themselves with having found either of the two resulting cycles.
+This situation poses a challenge to current work on semantic dependencies,
+which typically restricts such dependencies to a single load and store.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult-1.litmus @@
+\caption{OOTA Multiplication Example, Initial Value 1}
+\label{lst:OOTA Multiplication Example, Initial Value 1}
+\end{listing}
+
+In contrast, consider
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1},
+which differs from
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+only in the initial values (1 instead of 0) and the \co{exists}
+clause (0 instead of 1).
+Note that an execution satisfying the \co{exists} clause is again
+a (forbidden) OOTA cycle.
+
+Starting with the intuitive execution where the values loaded from \co{x}
+and \co{y} on lines~9 and~10 are the value 1 (that is, non-zero), executions
+that differ only in the value loaded from either \co{x} or \co{y} will
+still affect the value stored to \co{z} by line~11.
+There is therefore one semantic dependency from the load from \co{x}
+to the store to \co{z} and another separate semantic dependency from
+the load from \co{y} to the store to \co{z}, which matches typical
+definitions of semantic dependency.
+Yet
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+and
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
+can be thought of as being merely two different executions of the same
+program.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult3-0.litmus @@
+\caption{OOTA Three-Factor Multiplication Example, Initial Value Zero}
+\label{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
+\end{listing}
+
+Because the only two solutions to $x^2 = x$ are the values zero and one,
+Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+and~\ref{lst:OOTA Multiplication Example, Initial Value 1}
+each have but one OOTA cycle, and an execution having non-zero non-one
+initial values could have at most two OOTA cycles, all of which are
+(again) forbidden by the C++ language.
+Listing~\ref{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
+addressses this limitation by multiplying the values obtained from three
+different loads.
+Because $x^3 = x$ has three solutions ($\left\{ -1, 0, 1 \right\}$),
+this execution could have up to three (yet again, forbidden) OOTA cycles.
+
+Much more elaborate examples can be constructed, for example,
+involving sums of products of values loaded.
+Reverting back to short-form code fragments, consider the following:
+
+\begin{quote}
+\begin{verbatim}
+z = w * x + w * y + w;
+\end{verbatim}
+\end{quote}
+
+This has the following semantic dependencies when all of \co{w},
+\co{x}, and \co{y} are zero:
+
+\begin{enumerate}
+\item	From the combination of \co{w} and \co{x} to \co{z}.
+\item	From the combination of \co{w} and \co{y} to \co{z}.
+\item	From \co{w} to \co{z}.
+\end{enumerate}
+
+Despite appearances, these are not redundant.
+The last semantic dependency does not imply the first, because given
+that \co{w} becomes non-zero, a change in \co{x} further changes
+the value stored to \co{z}.
+Similar reasoning shows that the last semantic dependency does not
+imply the second.
+Much more elaborate examples can easily be constructed.
+
+This raises the question of how to determine the set of semantic
+dependencies in a given execution of a given fragment of code.
+The short answer is that when running on a real-world system, 
+any C++ program is a finite-state machine, which, unlike Turing-complete
+systems, can be analyzed.
+% @@@ reference to cbmc appendix.
+
+One might ask why semantic dependencies can be non-trivial when hardware
+address, control, and data dependencies are much more straightforward.
+Answers include:
+(1)~Respecting dependencies even when they reduce to a constant,
+(2)~Less-aggressive optimizations, and
+(3)~Lack of undefined behavior.
+
+This situation further underscores the per-execution nature of semantic
+dependencies as well as the need for an improved definition of ``semantic
+dependency''.
+We propose expanding the traditional definition to include dependencies
+extending from groups of loads to a single store.
+
+\subsubsection{OOTA From Undefined-Behavior Optimizations}
+\label{app:OOTA From Undefined-Behavior Optimizations}
+
+Many C++ implementations carry out optimizations based on the assumption
+that UB cannot happen, and these optimizations can result in OOTA cycles,
+with a simple example shown in
+Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-div-ub.litmus @@
+\caption{OOTA Due To Divide-By-Zero UB}
+\label{lst:OOTA Due To Divide-By-Zero UB}
+\end{listing}
+
+On line~14 of this listing, the value of \co{r2} must be either zero
+or one due to the comparison on line~13.
+If it is the value one, then there is a divide-by-zero error, which
+enables the C++ implementation to assume that the value of \co{r1} is
+always zero.
+This in turn allows the implementation to behave as if the expression
+stored to \co{y} on line~14 was the constant 1, at which point there
+is no semantic dependency from the load from \co{x} on line~13 to the
+store to \co{y} on line~14.
+
+The implementation is then within its rights to act as if lines~13 and~14
+were interchanged.
+At which point, the order of execution could be as follows:
+
+\begin{enumerate}
+\item	Line~14 stores the constant 1 to \co{y}.
+\item	Line~8 loads the constant 1 from \co{y}, setting \co{r1} to 1.
+\item	Line~9 stores the constant 1 to \co{x}.
+\item	Line~8 loads the constant 1 from \co{x}, setting \co{r2} to 1.
+\end{enumerate}
+
+Please note that the OOTA-like cycle called out in the \co{exists}
+clause can happen even though the program never actually executes any
+code that would have resulted in division-by-zero UB.
+It is tempting to consider using the techniques described in
+Appendix~\ref{app:Aside on Undefined Behavior}
+that prevent back-propagation of UB, but these are ineffective in this
+case because the C++ implementation can detect and exploit UB when
+considering line~14 in isolation.
+Once the constant 1 is substituted for line~14's expression, the
+hardware is free to reorder lines~13 and~14 even if a compiler-based
+C++ implementation refrains from doing so.
+Note that this OOTA-like cycle can appear even if all of the atomic
+operations in
+Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}
+were converted to be volatile.
+
+Therefore, this paper focuses on UB-free programs, rare though these
+beasts might be among large-scale C++ programs running in production.
+
+\clearpage
+
+\section{Aside on Undefined Behavior}
+\label{app:Aside on Undefined Behavior}
+
+The combination of OOTA cycles and UB has proven especially
+challenging~\cite{DavidGoldblatt2019NoElegantOOTAfix,HansBoehm2020UBalternatives}.
+
+\subsection{Alignment-Based Undefined Behavior}
+\label{sec:Alignment-Based Undefined Behavior}
+
+The example shown in
+Listing~\ref{lst:Alignment-Based Undefined Behavior (P1916R0)}
+(taken from P1916R0) is especially instructive.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 std::atomic<int> y;
+ 2 template <typename T>
+ 3 void t1() {
+ 4   long long r1 = x.load(std::memory_order_relaxed);
+ 5   if (r1 % 4 == 0)
+ 6     y.store(1, std::memory_order_relaxed);
+ 7   *(T*)r1 = *(T*)r1;
+ 8 }
+\end{verbatim}
+\caption{Alignment-Based Undefined Behavior (P1916R0)}
+\label{lst:Alignment-Based Undefined Behavior (P1916R0)}
+\end{listing}
+
+Because the compiler often assumes that UB cannot happen, it might
+look at line~7 and conclude that the value of \co{r1} necessarily
+corresponded to an address aligned per the requirements of type \co{T}.
+If that type's alignment was larger than four bytes, the compiler could
+then replace the condition in line~5 with \co{true}, which would mean
+that there was no semantic dependency from the load of \co{x} to the
+store of \co{y}.
+
+Except that the developer might have knowingly carried out a misaligned
+access, for example, to test an alignment trap handler or because the
+system on which this code was to run was known to gracefully handle
+misaligned accesses.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 std::atomic<int> y;
+ 2 template <typename T>
+ 3 void t1() {
+ 4   long long r1 = x.load(std::memory_order_relaxed);
+ 5   if (r1 % 4 == 0)
+ 6     y.store(1, std::memory_order_relaxed);
+ 7   std::observable();
+ 8   *(T*)r1 = *(T*)r1;
+ 9 }
+\end{verbatim}
+\caption{Alignment-Based Undefined Behavior Fix}
+\label{lst:Alignment-Based Undefined Behavior Fix}
+\end{listing}
+
+However, in that case, the developer is stepping outside the bounds
+of the standard.
+It would be good to have some way for the developer to get the job
+done without confusing the compiler, thus separating OOTA-cycle and
+UB concerns.
+One approach is to use the \co{std::observable()} function
+that has been proposed to the C++ standards
+committee~\cite{DavisHerring2021P1494R2}.
+This function would blocks backwards-propagating UB.
+Then this example could be rewritten as shown in
+Listing~\ref{lst:Alignment-Based Undefined Behavior Fix},
+with the asm on line~7 preventing the compiler from assuming that line~8
+is guaranteed to be executed, thus preventing the compiler from
+applying any lessons that it might learn from line~8 to line~5.
+
+% Clean up <200b> in vim: ":%s/\%u200b//g"
+
+There are a number of ways to implement \co{std::observable()},
+for example, a new \co{"mightnotreturn"} GCC asm clobber, a call to an empty
+function that is invisible to the compiler, and a GCC goto asm that the
+compiler cannot prove falls through.
+
+\subsection{Unreachability-Based Undefined Behavior}
+\label{sec:Unreachability-Based Undefined Behavior}
+
+This section attempts to apply the lessons
+of Appendix~\ref{sec:Unreachability-Based Undefined Behavior} to
+Goldblatt's unreachability-UB example.
+
+The \co{std::observable()} at line~12 of
+Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
+avoids confusing the compiler, instead forcing it to assume that the
+value of \co{local} might not be zero.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 // UB if this function is ever called.
+ 2 [[noreturn]]
+ 3 inline void unreachable() {
+ 4   return;
+ 5 }
+ 6 void t1() {
+ 7   r1 = x.load(std::memory_order_relaxed);
+ 8   int local = (r1 & 1);
+ 9   if (local == 0)
+10     y.store(1, std::memory_order_relaxed);
+11   if (local != 0) {
+12     std::observable();
+13     // If we get here, UB.
+14     unreachable();
+15   }
+16 }
+\end{verbatim}
+\caption{Unreachability-Based Undefined Behavior Fix}
+\label{lst:Unreachability-Based Undefined Behavior Fix}
+\end{listing}
+
+By the same reasoning, Goldblatt's \co{misbehave} example that
+also involves unreachability can also be addressed through insertion
+of the proposed \co{std::observable()}.
+
+\subsection{Atomic Stores From Undefined Behavior}
+\label{sec:Atomic Stores From Undefined Behavior}
+
+Another proposal to the C++ standards
+committee~\cite{HansBoehm2020UBalternatives}
+proposes (among other alternatives) that any atomic store resulting
+from undefined behavior have at least release semantics.
+This would prevent such a store from propagating undefined behavior
+backwards within that store's thread.
+In some cases, this could also prevent this undefined behavior from
+propagating back through other threads.
+
+However, it might well force C++ implementations to promote a surprisingly
+large fraction of a program's relaxed stores to release.
+
+\subsection{Lifetime-Based Undefined Behavior}
+\label{sec:Lifetime-Based Undefined Behavior}
+
+Not all undefined behavior can be backstopped using
+\co{std::unreachable()}.
+
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 int *foo(int *r)
+ 2 {
+ 3   b.store(r, memory_order_relaxed);
+ 4   return &global;
+ 5 }
+ 6
+ 7 int go(int *p)
+ 8 {
+ 9   int r;
+10   int *q = foo(&r);
+11
+12   r = 2;
+13   *p = 1;
+14   *q = r;
+15   return *q - r;
+16 }
+17
+18 void t1(void)
+19 {
+20   int r1 = go(a.load(memory_order_relaxed));
+21   printf("go() = %d\n", r1);
+22 }
+23
+24 void t2(void)
+25 {
+26   int *r2 = b.load(memory_order_relaxed);
+27   a.store(r2, memory_order_relaxed);
+28 }
+\end{verbatim}
+\caption{Lifetime-Based Undefined Behavior}
+\label{lst:Lifetime-Based Undefined Behavior}
+\end{listing}
+
+Listing~\ref{lst:Lifetime-Based Undefined Behavior}
+shows an example loosely based on one presented by Stephen Dolan\footnote{
+	\url{https://stedolan.net/talks/2024/loadstore}}
+in which function \co{go()} invokes \co{foo()} with a pointer to \co{go()}'s
+local variable \co{r}.
+However, \co{foo()} uses a relaxed store to place this pointer in
+an atomic variable \co{b}.
+Because this store is relaxed, it is not ordered against other code
+in \co{go()} and \co{t1()}, which means that \co{t2()} might well
+read this pointer from \co{b} either before the beginning or after
+the end of \co{r}'s lifetime, neither of which is well-defined.
+
+Of course, any attempt to dereference the pointer read from either
+\co{a} or \co{b} could result in undefined behavior.
+
+The lesson is clear:
+Properly synchronize reads and writes of pointers with the lifetime
+of the pointed-to object.
+
+This example could make lines~3 and~27 be \co{memory_order_release} and
+lines~20 and~26 be \co{memory_order_acquire} in order to avoid carrying
+a pointer to \co{r} before the lifetime of \co{r} has started.
+Alternatively, threads could be spawned only after \co{r}'s lifetime
+starts and joined before it ends.
+Placing \co{r} in static global storage would avoid lifetime considerations
+totally.
+And there are of course any number of other synchronization schemes
+that might be used to ensure that pointers are in use only while the
+pointed-to objects are alive.
+
+Note that \co{std::unreachable()} does not suffice because the CPU
+can also reorder relaxed stores.
+Note also that load-store ordering is not sufficient to prevent a
+pointer to \co{r} from being stored to \co{b} prior to the start of
+\co{r}'s lifetime because there is no atomic load between the start
+of that lifetime and the relaxed store to \co{b}.
+
+There are a number of ways to prevent carrying a pointer to \co{r} after
+the lifetime of \co{r} has ended, and these are left as an exercise for
+the reader.
+
+\clearpage
+
+\section{Evaluating sdep Using \co{cbmc}}
+\label{sec:Evaluating sdep Using cbmc}
+
+This section demonstrates the evaluation of sdep using
+the C Bounded Model Checker
+(\co{cbmc})~\cite{EdmundClarke2004CBMC}.
+This tool takes a smallish C program as input and constructs a logic
+expression representing that program.
+The logic expression's variables represent bits in the program's inputs
+and state, and the result of that logic expression is \co{true} if that
+program has a bug, for example, if it is possible for that C program to
+trigger an assertion or execute an out-of-bounds array reference.
+The tool then hands this expression to a SAT solver, and if this solver
+finds a satisfying set of values, the tool uses those values provide
+the execution path to the bug.
+
+But first, Appendix~\ref{app:Informal Definition of Semantic Dependency}
+outlines an informal informal definition of ``semantic dependency''
+that can be used in \co{cbmc} test cases.
 
@@
 current decidedly non-mathematical form.
-
-\subsection{Litmus Tests from “Causality Test Cases"}
-\label{app:Litmus Tests from “Causality Test Cases"}
-
-These tests might be two decades old, but they are still quite
-relevant.\footnote{
-	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
-
-\subsubsection{Causality Test Case 1}
-\label{app:Causality Test Case 1}
-
-Listing~\ref{lst:Causality Test Case 1}
-shows causality test case 1, for which the \co{r1 == r2 == 1} result
-is to be allowed.
-The \co{herd7} tool finds this litmus test's cycle, even without the
-inter-thread analysis, because C++ does not respect control dependencies
-to relaxed atomic stores.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-1.litmus @@
-\caption{Causality Test Case 1}
-\label{lst:Causality Test Case 1}
-\end{listing}
-
-Implementations that can prove that the value of \co{x} is always
-non-negative can also prove that there is no semantic dependency between
-\co{P0()}'s load from \co{x} on line~8 to its store to \co{y} on line~10,
-that is, no matter which of the possible values is loaded from \co{x},
-the value 1 will always be stored to \co{y}.
-
-Less-omniscient implementations, including those whose code analysis
-is restricted to a thread, would need to act as if there was in fact a
-semantic dependency.
-
-Even omniscient implementations will observe a semantic dependency from
-\co{P1()}'s load from \co{y} on line~14 to its store to \co{x} on line~15.
-
-Either way, there is no OOTA cycle, with omniscient implementations
-having no cycle at all, and other implementations retaining the
-non-semantic dependency.
-
-\subsubsection{Causality Test Case 2}
-\label{app:Causality Test Case 2}
-
-Listing~\ref{lst:Causality Test Case 2}
-shows causality test case 2, for which the \co{r1 == r2 == r3 == 1} result
-is to be allowed.
-The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-2.litmus @@
-\caption{Causality Test Case 2}
-\label{lst:Causality Test Case 2}
-\end{listing}
-
-Because \co{P0()}'s loads from \co{x} on lines~8 and~9 are unordered,
-an implementation can act as if the source code loaded from \co{x} only
-once and assigned the returned value to both \co{r1} and \co{r2}.\footnote{
-	But not in C, where relaxed atomic operations are volatile,
-	and thus observable behavior.}
-The implementation could then prove that the condition on line~10
-would always evaluate to \co{true}, and act as if the source code had
-unconditionally executed the relaxed store to \co{y} on line~11.
-The implementation could then act as if the resulting relaxed accesses
-to \co{x} and \co{y} appeared in reverse order, at which point
-\co{P0()}'s store to \co{y} could be read by \co{P1()} load so that
-\co{P1()} stored 1 to \co{x}, thus resulting in \co{P1()} loading
-that 1 from \co{x}.
-
-Note that \co{P1()} has a trivial semantic dependency from its load
-from \co{y} on line~15 to its store to \co{x} on line~16.
-
-If \co{P0()}'s loads from \co{x} are fused, there is no cycle,
-otherwise the preserved non-semantic dependency prevents an
-OOTA cycle from forming on correct real-world implementations.
-
-\subsubsection{Causality Test Case 3}
-\label{app:Causality Test Case 3}
-
-Listing~\ref{lst:Causality Test Case 3}
-shows causality test case 3, for which the \co{r1 == r2 == r3 == 1} result
-is to be allowed.
-The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
-
-The analysis from Section~\ref{app:Causality Test Case 2}
-applies here, including the difference in volatility between C and C++
-atomics.
-Although the addition of \co{P2} allows sequentially consistent executions
-where \co{P0()}'s two loads from \co{x} might return different values,
-it does not change the fact that those two loads can be merged.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-3.litmus @@
-\caption{Causality Test Case 3}
-\label{lst:Causality Test Case 3}
-\end{listing}
-
-\subsubsection{Causality Test Case 4}
-\label{app:Causality Test Case 4}
-
-Listing~\ref{lst:Causality Test Case 4}
-shows causality test case 4, for which the \co{r1 == r2 == 1} result
-is to be forbidden.
-The \co{herd7} tool finds this OOTA cycle because C++ does not respect
-data dependencies, and flags it as such using \co{S8}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-4.litmus @@
-\caption{Causality Test Case 4}
-\label{lst:Causality Test Case 4}
-\end{listing}
-
-Both \co{P0()} and \co{P1()} have straightforward semantic dependencies
-from their relaxed atomic loads to their relaxed atomic stores, which
-prevents this OOTA cycle on correct real-world C++ implementations.
-
-\subsubsection{Causality Test Case 5}
-\label{app:Causality Test Case 5}
-
-Listing~\ref{lst:Causality Test Case 5}
-shows causality test case 5, for which the \co{r1 == r2 == 1 && r3 == 0}
-result is to be forbidden.
-The \co{herd7} tool finds this OOTA cycle because C++ does not respect
-data dependencies, and flags it as such using \co{S8}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-5.litmus @@
-\caption{Causality Test Case 5}
-\label{lst:Causality Test Case 5}
-\end{listing}
-
-Each of \co{P0()}, \co{P1()}, and \co{P3()} have trivial semantic
-dependencies from their relaxed atomic loads to their relaxed
-atomic stores, which prevents any OOTA cycle from forming in correct
-real-world C++ implementations.
-\co{P2()} has no loads, so cannot participate in an OOTA cycle in
-any case.
-
-\subsubsection{Causality Test Case 6}
-\label{app:Causality Test Case 6}
-
-Listing~\ref{lst:Causality Test Case 6}
-shows causality test case 6, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool finds this cycle, and does not flag it as an OOTA
-cycle because C++ does not respect control dependencies to relaxed
-atomic stores.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-6.litmus @@
-\caption{Causality Test Case 6}
-\label{lst:Causality Test Case 6}
-\end{listing}
-
-A C++ implementation that can prove that \co{y} is always either
-zero or one can also prove that there is no semantic dependency
-from \co{P1()}'s load on line~14 to its stores on lines~16 and~29.
-Such an implementation could act as if lines~15-20 had been replaced
-by a single store to \co{x}, allowing that store to execute first,
-followed by lines~8, 10, and~14, in that order, satisfying the
-\co{exists} clause on line~23.
-
-Less-omniscient implementations will preserve the non-semantic
-dependency, preventing any OOTA loop from forming in a correct real-world C++
-implementation.
-
-\subsubsection{Causality Test Case 7}
-\label{app:Causality Test Case 7}
-
-Listing~\ref{lst:Causality Test Case 7}
-shows causality test case 7, for which the \co{r1 == r2 == r3 == 1}
-result is to be allowed.
-The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-7.litmus @@
-\caption{Causality Test Case 7}
-\label{lst:Causality Test Case 7}
-\end{listing}
-
-Because all accesses are relaxed atomics, this test case has everything
-to do with reordering and nothing to do with OOTA cycles.
-The outcome happens when lines~17, 10, 11, 15, 16 and~9 execute in that
-order.
-Note that this ordering preserves the semantic dependencies from
-lines~10-11 and~15-16, and can thus happen on correct real-world
-C++ implementations.
-
-\subsubsection{Causality Test Case 8}
-\label{app:Causality Test Case 8}
-
-Listing~\ref{lst:Causality Test Case 8}
-shows causality test case 8, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not find this cycle, but when \co{r2} is explicitly
-initialized to the constant \co{1}, it does find the cycle and refrains
-from flagging it as an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-8.litmus @@
-\caption{Causality Test Case 8}
-\label{lst:Causality Test Case 8}
-\end{listing}
-
-Omniscient C++ implementations might note that the computation on line~9
-can only have the value one when the load from \co{x} returns either
-zero or one, and both \co{x} and \co{y} are initialized to zero.
-Such an implementation act as if line~10's \co{r2} had instead been
-the value~1, so that the value returned by \co{P0()}'s load could not
-affect the value stored.
-Then there would be no semantic dependency between \co{P0()}'s load
-and store, and thus no OOTA cycle in a correct real-world C++
-implementation.
-
-A less-capable implementation would leave the non-semantic dependency
-in place, also preventing any OOTA cycle in a correct real-world C++
-implementation.
-
-\subsubsection{Causality Test Case 9}
-\label{app:Causality Test Case 9}
-
-Listing~\ref{lst:Causality Test Case 9}
-shows causality test case 9, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not find this cycle because it depends on
-scheduling \co{P2()} after the other two threads, and C++ is not
-constrained to do this.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-9.litmus @@
-\caption{Causality Test Case 9}
-\label{lst:Causality Test Case 9}
-\end{listing}
-
-Both threads have straightforward semantic dependencies between their
-respective relaxed atomic loads and relaxed atomic stores, so an OOTA
-cycle cannot form in a correct real-world C++ implementation.
-
-\subsubsection{Causality Test Case 9a}
-\label{app:Causality Test Case 9a}
-
-Listing~\ref{lst:Causality Test Case 9a}
-shows causality test case 9a, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not find this cycle because it depends on
-scheduling \co{P2()} before the other two threads, and C++ is not
-constrained to do this.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-9a.litmus @@
-\caption{Causality Test Case 9a}
-\label{lst:Causality Test Case 9a}
-\end{listing}
-
-Both threads have straightforward semantic dependencies between their
-respective relaxed atomic loads and relaxed atomic stores, so an OOTA
-cycle cannot form in a correct real-world C++ implementation.
-
-\subsubsection{Causality Test Case 10}
-\label{app:Causality Test Case 10}
-
-Listing~\ref{lst:Causality Test Case 10}
-shows causality test case 10, for which the \co{r1 == r2 == 1 && r3 == 0}
-result is to be forbidden.
-Because C++ does not respect control dependencies, the \co{herd7} tool
-nevertheless reports this result and does not flag it as an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-10.litmus @@
-\caption{Causality Test Case 10}
-\label{lst:Causality Test Case 10}
-\end{listing}
-
-However, each of \co{P0()}, \co{P1()}, and \co{P3()} have straightforward
-semantic dependencies between their respective loads and stores, and
-thus cannot form an OOTA cycle when running on a correct real-world C++
-implementation.
-For its part, \co{P2()} has no semantic dependency, so it cannot be part
-of an OOTA cycle when running on a correct real-world C++ implementation.
-
-\subsubsection{Causality Test Case 11}
-\label{app:Causality Test Case 11}
-
-Listing~\ref{lst:Causality Test Case 11}
-shows causality test case 11, for which the \co{r1 == r2 == r3 == r4 == 1}
-result is to be allowed.
-The \co{herd7} tool reports this result and does not flag it as an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-11.litmus @@
-\caption{Causality Test Case 11}
-\label{lst:Causality Test Case 11}
-\end{listing}
-
-This test case has nothing to do with OOTA cycles.
-The result may be obtained by executing the statements in the following
-order: line~20, line~12, line~13, line~18, line~19, line~10, line~11,
-and finally line~17.
-
-\subsubsection{Causality Test Case 12}
-\label{app:Causality Test Case 12}
-
-JMM Causality Test Case 12 in
-Listing~\ref{lst:Causality Test Case 12}
-uses arrays, which are not yet supported by the \co{herd7} tool.
-
-\begin{listing}[tbp]
-@@ DisplayLitmus litmus/oota-causality-12.litmus @@
-\caption{Causality Test Case 12}
-\label{lst:Causality Test Case 12}
-\end{listing}
-
-An omniscient C++ implementation could note that only \co{P0()} accesses
-array \co{a[]} and act as if the code was simpler, but there would
-still be semantic dependencies between \co{P0()}'s and \co{P1()}s
-relaxed atomic loads and stores involving \co{x} and \co{y}.
-Those dependencies prevent an OOTA cycle from forming on a correct
-real-world C++ implementation, omniscient or not.
-
-\subsubsection{Causality Test Case 13}
-\label{app:Causality Test Case 13}
-
-Listing~\ref{lst:Causality Test Case 13}
-shows causality test case 13, for which the \co{r1 == r2 == 1}
-result is to be forbidden.
-But because C++ does not respect control dependencies, \co{herd7} tool
-reports this result and does not flag it as an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-13.litmus @@
-\caption{Causality Test Case 13}
-\label{lst:Causality Test Case 13}
-\end{listing}
-
-There are straightforward semantic dependencies between \co{P0()}'s and
-\co{P1()}s relaxed atomic loads and stores involving \co{x} and \co{y}.
-Those dependencies prevent an OOTA cycle from forming on a correct
-real-world C++ implementation.
-
-\subsubsection{Causality Test Case 14}
-\label{app:Causality Test Case 14}
-
-Listing~\ref{lst:Causality Test Case 14}
-shows causality test case 14, for which the \co{r1 == r3 == 1 && r2 == 0}
-result is to be forbidden.
-The \co{herd7} tool nevertheless reports this result because C++ does not
-respect control dependencies for atomic stores.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-14.litmus @@
-\caption{Causality Test Case 14}
-\label{lst:Causality Test Case 14}
-\end{listing}
-
-In C++, this is not a matter of OOTA but rather reordering.
-Dependencies notwithstanding, according to the C++ memory model,
-the statements could execute in the following order:
-Line~17 (so \co{r2 == 0}), line~21, line~9 (so \co{r1 == 1}),
-line~13, line~18 (so \co{r3 == 1)}.
-The sum of \co{r2} and \co{r3} is non-zero, justifying line~21's
-store to \co{x}.
-
-However, correct real-world C++ implementations would have semantic
-dependencies between \co{P0()}'s line~9 and lines~11 and~13, and also
-between \co{P1()}'s lines~17 and~18 and line~21, preventing any
-OOTA cycle from forming.
-
-\subsubsection{Causality Test Case 15}
-\label{app:Causality Test Case 15}
-
-Listing~\ref{lst:Causality Test Case 15}
-shows causality test case 15, for which the \co{r0 == r1 == r3 == 1 && r2 == 0}
-result is to be forbidden.
-The \co{herd7} tool nevertheless reports this result because C++ does not
-respect control dependencies for atomic stores.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-15.litmus @@
-\caption{Causality Test Case 15}
-\label{lst:Causality Test Case 15}
-\end{listing}
-
-In C++, this is not a matter of OOTA but rather reordering.
-Dependencies notwithstanding, according to the C++ memory model,
-the statements could execute in the following order:
-Line~31, line~27, line~10 (so \co{r0 == 1}), line~13 (so \co{r1 == 1}),
-line~19 (because \co{r1 == 1}), line~23 (so \co{r2 == 0}),
-and finally line~24 (so \co{r3 == 1}).
-The sum of \co{r2} and \co{r3} is non-zero, justifying line~27's
-store to \co{u}.
-
-However, correct real-world C++ implementations would have semantic
-dependencies between \co{P0()}'s lines~10 and~13 and lines~17 and~19, and also
-between \co{P1()}'s lines~23 and~24 and line~27, preventing any
-OOTA cycle from forming.
-Having no load, \co{P2()} cannot participate in an OOTA cycle.
-
-\subsubsection{Causality Test Case 16}
-\label{app:Causality Test Case 16}
-
-Listing~\ref{lst:Causality Test Case 16}
-shows causality test case 16, for which the \co{r1 == 2 && r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be expected
-given that C++ does not permit a thread to reorder accesses to a single
-atomic variable, even when those accesses are relaxed.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-16.litmus @@
-\caption{Causality Test Case 16}
-\label{lst:Causality Test Case 16}
-\end{listing}
-
-\subsubsection{Causality Test Case 17}
-\label{app:Causality Test Case 17}
-
-Listing~\ref{lst:Causality Test Case 17}
-shows causality test case 17, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be expected
-given that C++ does not permit a thread to reorder accesses to a single
-atomic variable, even when those accesses are relaxed.
-But note that the \co{herd7} tool correctly notes the presence of an
-OOTA cycle with \co{S17}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-17.litmus @@
-\caption{Causality Test Case 17}
-\label{lst:Causality Test Case 17}
-\end{listing}
-
-\co{P1()} has a straightforward semantic dependency between its
-relaxed atomic load and relaxed atomic store.
-However, suppose that \co{P0()}'s load from \co{x} on line~8 returns
-the value 42.
-In that case, the second load on line~11 can be fused with the first,
-so that it also returns the value 42.
-On the other hand, if the load on line~8 returns some other value, then
-line~10 will store the value 42 to \co{x}, and then the load on line~11
-can be elided\footnote{
-	In C++, but not in C due to atomics being volatile.}
-in favor of using the value 42 stored by line~10.
-
-Either way, the value of \co{r1} will be the constant 42.
-Therefore, the loads from \co{x} cannot affect the store on line~17,
-so there is no semantic dependency.
-This allows that store to \co{y} to be moved to the very beginning of
-\co{P0()}, thus allowing a correct real-world C++ implementation to
-satisfy the \co{exists} clause.
-
-\subsubsection{Causality Test Case 18}
-\label{app:Causality Test Case 18}
-
-Listing~\ref{lst:Causality Test Case 18}
-shows causality test case 18, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be expected
-given that C++ does not permit a thread to reorder accesses to a single
-atomic variable, even when those accesses are relaxed.
-But note that the \co{herd7} tool correctly notes the presence of an
-OOTA cycle with \co{S17}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-18.litmus @@
-\caption{Causality Test Case 18}
-\label{lst:Causality Test Case 18}
-\end{listing}
-
-Both threads have straightforward semantic dependencies between their
-respective relaxed atomic loads and relaxed atomic stores, so an OOTA
-cycle cannot form in a correct real-world C++ implementation.
-
-\subsubsection{Causality Test Case 19}
-\label{app:Causality Test Case 19}
-
-Listing~\ref{lst:Causality Test Case 19}
-shows causality test case 19, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be expected
-given that C++ does not permit a thread to reorder accesses to a single
-atomic variable, even when those accesses are relaxed.
-But note that the \co{herd7} tool correctly notes the presence of an
-OOTA cycle with \co{S8}.
-
-It is interesting that this \co{S8} value does not propagate to
-\co{r3}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-19.litmus @@
-\caption{Causality Test Case 19}
-\label{lst:Causality Test Case 19}
-\end{listing}
-
-All threads have straightforward semantic dependencies between their
-respective relaxed atomic loads and relaxed atomic stores, so an OOTA
-cycle cannot form in a correct real-world C++ implementation.
-
-\subsubsection{Causality Test Case 20}
-\label{app:Causality Test Case 20}
-
-Listing~\ref{lst:Causality Test Case 20}
-shows causality test case 20, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be expected
-given that C++ does not permit a thread to reorder accesses to a single
-atomic variable, even when those accesses are relaxed.
-But note that the \co{herd7} tool correctly notes the presence of an
-OOTA cycle with \co{S8}.
-
-It is interesting that this \co{S8} value does not propagate to
-\co{r3}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-20.litmus @@
-\caption{Causality Test Case 20}
-\label{lst:Causality Test Case 20}
-\end{listing}
-
-All threads have straightforward semantic dependencies between their
-respective relaxed atomic loads and relaxed atomic stores, so an OOTA
-cycle cannot form in a correct real-world C++ implementation.
-
-\subsection{Additional Litmus Tests}
-\label{app:Additional Litmus Tests}
-
-These litmus tests helped illuminate important aspects of the problem
-of defining sdep and identifying OOTA cycles.
-
-\subsubsection{Simple Reordering as Litmus Test}
-\label{app:Simple Reordering as Litmus Test}
-
-Listing~\ref{lst:Simple Reordering as Litmus Test}
-shows how
-Listing~\ref{lst:Simple Reordering}
-can be recast into a \co{herd7} litmus test.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/simple-reordering.litmus @@
-\caption{Simple Reordering as Litmus Test}
-\label{lst:Simple Reordering as Litmus Test}
-\end{listing}
-
-\subsubsection{Non-LB Litmus Tests}
-\label{app:Non-LB Litmus Tests}
-
-Listing~\ref{lst:Non-LB Litmus Test}
-shows a litmus test that can produce an OOTA cycle, but that is not
-a member of the load-buffering (LB) class of litmus tests.
-Suppose that lines~9 and~10 both somehow return the value 21.
-Then line~12 will store the value 42 to \co{x}, which will cause
-line~17 to store 21 to \co{y} and line~22 to store 21 to \co{z},
-thus justifying the values initially loaded by lines~9 and~10.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-non-lb.litmus @@
-\caption{Non-LB Litmus Test}
-\label{lst:Non-LB Litmus Test}
-\end{listing}
-
-All thread's loads' return values affect the value stored, so all
-threads have semantic dependencies between their respective relaxed
-atomic loads and relaxed atomic stores.
-Therefore, an OOTA cycle cannot form in a correct real-world C++
-implementation.
-
-Note that in a C++ implementation that (erroneously) permitted an OOTA
-cycle to form, that cycle would involve both \co{r0} and \co{r1}.
-However, we conjecture that tools designed to find OOTA cycles could
-satisfy themselves with having found either of the two resulting cycles.
-
-\subsubsection{Multi-Cycle Litmus Tests}
-\label{app:Multi-Cycle Litmus Tests}
-
-Listing~\ref{lst:Multi-Cycle Litmus Test}
-shows an expansion to four threads of Causality Test Case 4, which is shown in
-Listing~\ref{lst:Causality Test Case 4}
-and discussed in
-Appendix~\ref{app:Causality Test Case 4}.
-The \co{herd7} tool shows multiple different possible OOTA cycles,
-as expected.
-However, the value returned by each thread's load directly affects the
-value stored, so there is a semantic dependency in each thread.
-Therefore, none of these OOTA cycles can happen in a correct real-world
-C++ implementation.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-two-source.litmus @@
-\caption{Multi-Cycle Litmus Test}
-\label{lst:Multi-Cycle Litmus Test}
-\end{listing}
-
-\subsubsection{Three-Process Version of JMM Causality Test Case 4}
-\label{app:Three-Process Version of JMM Causality Test Case 4}
-
-Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
-shows a different way of expanding
-Listing~\ref{lst:Causality Test Case 4}
-(discussed in
-Appendix~\ref{app:Causality Test Case 4})
-from two threads to three threads.
-The \co{herd7} tool finds OOTA values, as expected.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-3proc.litmus @@
-\caption{Three-Process Version of JMM Causality Test Case 4}
-\label{lst:Three-Process Version of JMM Causality Test Case 4}
-\end{listing}
-
-All threads have straightforward semantic dependencies from their
-relaxed atomic loads to their relaxed atomic stores, which prevents this
-OOTA cycle on correct real-world C++ implementations.
-
-\subsubsection{Why rfe Instead of Tried-And-True rf?}
-\label{app:Why rfe Instead of Tried-And-True rf?}
-
-Although the examples presented in
-Section~\ref{sec:OOTA: rf vs. rfe}
-(and shown as litmus tests in
-Listings~\ref{lst:OOTA With an Intra-Process Store-Load Link}
-and~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized})
-demonstrate that rf can be atemporal, they demonstrate neither
-OOTA cycles nor OOTA-like behavior in full C++ implementations.
-This appendix provides more involved examples that can be optimized
-so as to demonstrate OOTA-like behavior.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-3-2-proc.litmus @@
-\caption{OOTA With an Intra-Process Store-Load Link}
-\label{lst:OOTA With an Intra-Process Store-Load Link}
-\end{listing}
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-3-2-proc-opt.litmus @@
-\caption{OOTA With an Intra-Process Store-Load Link, Optimized}
-\label{lst:OOTA With an Intra-Process Store-Load Link, Optimized}
-\end{listing}
-
-Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
-revises
-Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
-so that the value stored to \co{y} is forced to be at most the
-value 17, and the value stored to \co{z} is forced to be at least the
-value 17.
-This of course means that the store to \co{z} must always be the value
-17, as can be seen from the \co{herd7} output.
-However, a single-thread-at-a-time analysis could not prove this.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-whyrfe-3.litmus @@
-\caption{Three-Process Version of JMM Causality Test Case 4 With Max and Min}
-\label{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
-\end{listing}
-
-Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
-further revises
-Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
-to flatten \co{PO()} and \co{P1()} into a single thread, so that the
-resulting \co{P0()} has an rfi link instead of an rfe link connecting
-the store to and load from \co{y}.
-This means that a C++ implementation could prove that the value 17 is
-always stored to \co{z} by examining \co{P0()} in isolation, given that
-it is also legal for a C++ implementation to replace the value loaded
-from \co{y} with the value just stored.\footnote{
-	This is not an option in C because atomic operations are volatile,
-	and thus observed behavior.}
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-whyrfe.litmus @@
-\caption{Why rfe Instead of Tried-And-True rf?}
-\label{lst:Why rfe Instead of Tried-And-True rf?}
-\end{listing}
-
-In the confines of the C++ abstract machine (omitting the no-OOTA
-restriction), this allows a value-17 OOTA cycle, courtesy of the fact
-that C++ respects neither control nor data dependencies.
-However, the \co{herd7} tool does not identify this cycle, perhaps due
-to its restrictions on its handling of arithmetic on OOTA values.
-
-In a real-world implementation that proves that the value 17 is always
-stored to \co{z}, there is no semantic dependency to
-\co{P0()}'s store to \co{z}, but there is to all of the other relaxed
-atomic stores.
-On the other hand, if the implementation does not prove the value
-stored to \co{z}, there will be a global-time delay between each of
-\co{P0()}'s loads and its store to \co{z}, preventing a value-17 OOTA
-cycle from forming.
-
-Either way, the value-17 OOTA cycle cannot occur in a correct real-world
-C++ implementation.
-Yes, an implementation that proves that the value 17 is always
-stored to \co{z} can satisfy the \co{exists} clause, but because
-there is no semantic dependency to \co{P0()}'s store to \co{z},
-this is not an OOTA cycle.
-
-But what if we chose to define an OOTA cycle as (sdep $\cup$ rf) instead
-of (sdep $\cup$ rfe)?
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-whyrfe-z17.litmus @@
-\caption{Why rfe Instead of Tried-And-True rf When z=17?}
-\label{lst:Why rfe Instead of Tried-And-True rf When z=17?}
-\end{listing}
-
-In that case, it would be necessary to separately consider the dependency
-from \co{P0()}'s load from \co{x} and \co{y} and that from its load
-from \co{y} and its store to \co{z}.
-The problem with this is shown in
-Listing~\ref{lst:Why rfe Instead of Tried-And-True rf When z=17?},
-which shows an equivalent program given the knowledge that the value~17
-will always be stored to \co{z}.
-Because all atomic operations are relaxed and because C++ does not
-respect control dependencies, the store to \co{z} can be moved up
-to the very beginning of \co{P0()}.
-This transformation causes the rfi link associated with \co{y} to
-go backwards in time, which is perfectly legitimate in this case,
-but which makes defining OOTA cycles in terms of (sdep $\cup$ rf)
-problematic.
-
-Again, in C, the load from \co{y} cannot be removed because it is
-volatile, and thus observable behavior.
-C++ could remove that load, but we left it in to demonstrate the all-17s
-outcome.
-
-This example demonstrates why OOTA cycles must be defined in terms
-of (sdep $\cup$ rfe), or in \co{herd7} nomenclature,
-\co{acyclic(sdep; rfe)}.
-
-\subsubsection{Inventing Atomic Loads}
-\label{app:Inventing Atomic Loads}
-
-This section contains several example litmus tests from the active
-mind of Alan Stern.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-no-invented-load.litmus @@
-\caption{No Invented Atomic Loads}
-\label{lst:No Invented Atomic Loads}
-\end{listing}
-
-First, consider Listing~\ref{lst:No Invented Atomic Loads}.
-At first glance, this might appear straightforward.
-\co{P0()}'s store to \co{z} clearly depends on its load from \co{y},
-and \co{P1()}'s store to \co{y} clearly depends on its load from \co{z}.
-If both dependencies are semantic dependency, the OOTA cycle cannot
-form.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-load-invented.litmus @@
-\caption{Atomic Loads Invented}
-\label{lst:Atomic Loads Invented}
-\end{listing}
-
-But is \co{P0()}'s dependency a semantic dependency?
-For example, is a real-world C++ implementation permitted to
-duplicate \co{P0()}'s load from \co{x} and then make use of both loads'
-values?
-If this is permitted, then the implementation would be within its rights
-to transform
-Listing~\ref{lst:No Invented Atomic Loads}
-into
-Listing~\ref{lst:Atomic Loads Invented}.
-
-Because the values of \co{r1}, \co{r2}, and \co{r3} are all known to be
-either zero or one, if the values returned by the two loads from \co{x}
-differ, then one of those two values must be equal to that from \co{y}.
-If the implementation is further permitted to make a runtime choice
-between the two values loaded from \co{x}, it could always choose the
-one that was equal to the value loaded from \co{y}, as shown by the
-\co{if} statement on line~13.
-This would mean that the store to \co{z} was always the constant 1,
-which would permit the store to be ordered before the load from \co{y},
-as shown in lines~14 and~15.
-Doing so permits the OOTA-like outcome of \co{r3 == r4 == 1}.
-
-The implementation might also drop the load from \co{y}, whose value
-is now unused except by the \co{exists} clause.
-However, the \co{exists} clause is considered to be observable behavior,
-which means that the load from \co{y} must remain.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-unused-load.litmus @@
-\caption{Unused Extra Atomic Load}
-\label{lst:Unused Extra Atomic Load}
-\end{listing}
-
-One possible reaction to this situation is to declare that useful
-real-world C++ implementations should never duplicate non-volatile
-relaxed atomic loads.
-But suppose that the initial code was as shown in
-Listing~\ref{lst:Unused Extra Atomic Load}?\footnote{
-	Such unused loads can easily be generated by complex macros or
-	template metaprograms.}
-In this case, the transformation to
-Listing~\ref{lst:Atomic Loads Invented}
-would not require inventing a load, but instead merely inventing a use of
-the previously discarded value in \co{r0}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-invent-int-load.litmus @@
-\caption{Inventing Non-Boolean Atomic Loads}
-\label{lst:Unused Inventing Non-Boolean Atomic Loads}
-\end{listing}
-
-This is not restricted to booleans, as can be seen in
-Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}.
-This example is more straightforward in that it does not make quite so
-creative use of the multiple values loaded.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/inc.litmus @@
-\caption{Only Atomic Increment}
-\label{lst:Only Atomic Increment}
-\end{listing}
-
-But suppose that we drop the single-thread-analysis constraint for
-the moment, and suppose that the C++ implementation is able to prove
-that the only modification of a given atomic variable is to atomically
-increment it, as is the case for \co{y} in
-Listing~\ref{lst:Only Atomic Increment}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/inc-range.litmus @@
-\caption{Only Atomic Increment, Optimized}
-\label{lst:Only Atomic Increment, Optimized}
-\end{listing}
-
-Is a C++ implementation within its rights to behave as if the code
-shown in
-Listing~\ref{lst:Only Atomic Increment}
-was instead as shown in
-Listing~\ref{lst:Only Atomic Increment, Optimized}?
-One argument in favor is that the value of \co{y} must have
-passed through the value 42 in the case where \co{r0} and \co{r1}
-bracket that value.
-Arguments against might cite the added overhead of the invented load
-in Listing~\ref{lst:Only Atomic Increment, Optimized}, or that the
-possibility that the check for the value 42 is intended for statistical
-sampling.
-
-There are a number of possible reactions to these situations:
-
-\begin{enumerate}
-\item	Listing~\ref{lst:No Invented Atomic Loads}
-	looks like it has an OOTA cycle, but there is in fact no
-	semantic dependency.
-\item	Attempts to use real-world constraints to prevent OOTA cycles
-	are futile.
-	(But see Section~\ref{sec:But What About Non-Volatile Atomics?}.)
-\item	Implementations should not invent uses for values from
-	non-volatile relaxed atomic loads.
-	If it turns out that there are useful optimizations that invent
-	such uses, then such optimizations must be applied only with
-	careful attention to the as-if rule.
-	Some might argue that inventing comparisons between a pair of
-	non-volatile relaxed atomic loads from the same object should
-	be permitted only if the two loads could be reordered to be
-	adjacent to each other.
-\item	Any time an implementation might be tempted to invent uses from
-	values from non-volatile relaxed atomic loads from the same
-	object, the loads should instead be fused into a single load.
-	This reaction might be attractive to those who carefully consider
-	the pointlessness of keeping two loads that return the same value
-	on the one hand or the cache-miss overhead incurred when closely
-	spaced loads return different values on the other.
-\item	Implementations that invent non-volatile atomic loads or invent
-	new uses for non-volatile atomic loads can adversely affect
-	observed behavior, for example, by introducing error into code
-	attempting to do statistical sampling.
-\item	What other interesting situations might come to light?
-\end{enumerate}
-
-There are situations where inventing loads from non-atomic objects
-is customary, for example, when hoisting a load out of a loop.
-In this case, implementations might hoist the load from x out of
-the loop:
-
-\begin{quote}
-\begin{verbatim}
-do_something_with(x);
-for (i = 0; i < limit i++)
-	y[i] += x;
-\end{verbatim}
-\end{quote}
-
-A C++ implementation might act as if the program had instead been this:
-
-\begin{quote}
-\begin{verbatim}
-r1 = x;
-do_something_with(r1);
-for (i = 0; i < limit i++)
-	y[i] += r1;
-\end{verbatim}
-\end{quote}
-
-Note that passing \co{x} to the call to \co{do_something_with()} is important
-because this allows the C++ implementation to assume that accesses to \co{x}
-will not result in a data race.
-Without this assumption, the above transformation might introduce a data
-race when \co{limit} is less than or equal to zero.
-
-But if \co{x} is a non-volatile atomic, then there are no worries about
-data races.
-In theory, the implementation could make the above transformation without
-worry, especially given that the value loaded from \co{x} is discarded in
-the case where \co{limit} is less than or equal to zero.
-Except that the data-race worry is replaced by a cache-miss worry.
-After all, optimizations are supposed to speed things up.
-This means that a C++ implementation might be wise to adhere to the
-same restriction that avoids data races for loads from non-atomic
-variables in order to avoid expensive cache misses.
-
-\subsubsection{Non-Trivial Semantic Dependencies}
-\label{app:Non-Trivial Semantic Dependencies}
-
-Most of the examples in this paper involve simple semantic dependencies
-connecting a single load to a single store.
-One exception appears in
-Appendix~\ref{app:Non-LB Litmus Tests},
-but this section will present a more difficult example.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult-0.litmus @@
-\caption{OOTA Multiplication Example, Initial Value Zero}
-\label{lst:OOTA Multiplication Example, Initial Value Zero}
-\end{listing}
-
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-shows an example that builds on the multiplication-by-zero
-discussion in
-Section~\ref{sec:Semantic Dependencies and Source Code}.
-Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
-\co{x} and \co{y} on lines~9 and~10, respectively.
-Similarly, lines~16 and~17 store to \co{x} and \co{y}, respectively,
-the value that line~15 loads from \co{z}.
-Because all three variables are initialized to zero, and because loads
-can be ordered before any store, there is an execution in which all
-three loads return the value zero, in which case all three stores will
-of course store the value zero.
-In any related execution where the value loaded from \co{y} remains zero,
-non-zero values loaded from \co{x} cannot affect the value stored by
-line~11 to \co{z}.
-Only executions which load different (that is, non-zero) values from
-both \co{x} and \co{y} can cause the value stored to \co{z} to change
-from zero to a non-zero value.
-
-Therefore, as noted in
-Section~\ref{sec:Semantic Dependencies and Source Code},
-in an execution where the loads from both \co{x} and \co{y} return zero,
-there is no semantic dependency from the load from \co{x} to the
-store to \co{z}, nor is there a semantic dependency from the load
-from \co{y} to the store to \co{z}.
-However, there \emph{is} a semantic dependency from the \emph{combination}
-of the loads from \co{x} and \co{y} to the store to \co{z}.
-For example, the (forbidden) execution that satisfies the \co{exists}
-clause, which would be an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult-0-cond.litmus @@
-\caption{OOTA Conditional Multiplication Example, Initial Value Zero}
-\label{lst:OOTA Conditional Multiplication Example, Initial Value Zero}
-\end{listing}
-
-For further evidence that there is no semantic dependency from \co{y}
-to \co{z} in executions where the load from \co{x} is zero, see the
-equivalent\footnote{
-	But equivalent only because the atomics are all non-volatile!}
-program shown in
-Listing~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero}.
-Here, there is not even a load from \co{y} in executions where the
-load from \co{x} returns zero, so there cannot possibly be a semantic
-dependency involving this non-existent load.
-A similar transformation would load \co{x} only in executions where the
-load from \co{y} returned non-zero, which similarly emphasizes the lack
-of any semantic dependency between \co{x} and \co{z}.
-
-In both
-Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-and~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero},
-an OOTA cycle (such as the one indicated by their \co{exists} clauses)
-would need to affect both the \co{x} and the \co{y} components of this
-compound semantic dependency.
-However, we again conjecture that tools designed to find OOTA cycles could
-satisfy themselves with having found either of the two resulting cycles.
-This situation poses a challenge to current work on semantic dependencies,
-which typically restricts such dependencies to a single load and store.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult-1.litmus @@
-\caption{OOTA Multiplication Example, Initial Value 1}
-\label{lst:OOTA Multiplication Example, Initial Value 1}
-\end{listing}
-
-In contrast, consider
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1},
-which differs from
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-only in the initial values (1 instead of 0) and the \co{exists}
-clause (0 instead of 1).
-Note that an execution satisfying the \co{exists} clause is again
-a (forbidden) OOTA cycle.
-
-Starting with the intuitive execution where the values loaded from \co{x}
-and \co{y} on lines~9 and~10 are the value 1 (that is, non-zero), executions
-that differ only in the value loaded from either \co{x} or \co{y} will
-still affect the value stored to \co{z} by line~11.
-There is therefore one semantic dependency from the load from \co{x}
-to the store to \co{z} and another separate semantic dependency from
-the load from \co{y} to the store to \co{z}, which matches typical
-definitions of semantic dependency.
-Yet
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-and
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
-can be thought of as being merely two different executions of the same
-program.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult3-0.litmus @@
-\caption{OOTA Three-Factor Multiplication Example, Initial Value Zero}
-\label{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
-\end{listing}
-
-Because the only two solutions to $x^2 = x$ are the values zero and one,
-Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-and~\ref{lst:OOTA Multiplication Example, Initial Value 1}
-each have but one OOTA cycle, and an execution having non-zero non-one
-initial values could have at most two OOTA cycles, all of which are
-(again) forbidden by the C++ language.
-Listing~\ref{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
-addressses this limitation by multiplying the values obtained from three
-different loads.
-Because $x^3 = x$ has three solutions ($\left\{ -1, 0, 1 \right\}$),
-this execution could have up to three (yet again, forbidden) OOTA cycles.
-
-Much more elaborate examples can be constructed, for example,
-involving sums of products of values loaded.
-Reverting back to short-form code fragments, consider the following:
-
-\begin{quote}
-\begin{verbatim}
-z = w * x + w * y + w;
-\end{verbatim}
-\end{quote}
-
-This has the following semantic dependencies when all of \co{w},
-\co{x}, and \co{y} are zero:
-
-\begin{enumerate}
-\item	From the combination of \co{w} and \co{x} to \co{z}.
-\item	From the combination of \co{w} and \co{y} to \co{z}.
-\item	From \co{w} to \co{z}.
-\end{enumerate}
-
-Despite appearances, these are not redundant.
-The last semantic dependency does not imply the first, because given
-that \co{w} becomes non-zero, a change in \co{x} further changes
-the value stored to \co{z}.
-Similar reasoning shows that the last semantic dependency does not
-imply the second.
-Much more elaborate examples can easily be constructed.
-
-This raises the question of how to determine the set of semantic
-dependencies in a given execution of a given fragment of code.
-The short answer is that when running on a real-world system, 
-any C++ program is a finite-state machine, which, unlike Turing-complete
-systems, can be analyzed.
-% @@@ reference to cbmc appendix.
-
-One might ask why semantic dependencies can be non-trivial when hardware
-address, control, and data dependencies are much more straightforward.
-Answers include:
-(1)~Respecting dependencies even when they reduce to a constant,
-(2)~Less-aggressive optimizations, and
-(3)~Lack of undefined behavior.
-
-This situation further underscores the per-execution nature of semantic
-dependencies as well as the need for an improved definition of ``semantic
-dependency''.
-We propose expanding the traditional definition to include dependencies
-extending from groups of loads to a single store.
-
-\subsubsection{OOTA From Undefined-Behavior Optimizations}
-\label{app:OOTA From Undefined-Behavior Optimizations}
-
-Many C++ implementations carry out optimizations based on the assumption
-that UB cannot happen, and these optimizations can result in OOTA cycles,
-with a simple example shown in
-Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-div-ub.litmus @@
-\caption{OOTA Due To Divide-By-Zero UB}
-\label{lst:OOTA Due To Divide-By-Zero UB}
-\end{listing}
-
-On line~14 of this listing, the value of \co{r2} must be either zero
-or one due to the comparison on line~13.
-If it is the value one, then there is a divide-by-zero error, which
-enables the C++ implementation to assume that the value of \co{r1} is
-always zero.
-This in turn allows the implementation to behave as if the expression
-stored to \co{y} on line~14 was the constant 1, at which point there
-is no semantic dependency from the load from \co{x} on line~13 to the
-store to \co{y} on line~14.
-
-The implementation is then within its rights to act as if lines~13 and~14
-were interchanged.
-At which point, the order of execution could be as follows:
-
-\begin{enumerate}
-\item	Line~14 stores the constant 1 to \co{y}.
-\item	Line~8 loads the constant 1 from \co{y}, setting \co{r1} to 1.
-\item	Line~9 stores the constant 1 to \co{x}.
-\item	Line~8 loads the constant 1 from \co{x}, setting \co{r2} to 1.
-\end{enumerate}
-
-Please note that the OOTA-like cycle called out in the \co{exists}
-clause can happen even though the program never actually executes any
-code that would have resulted in division-by-zero UB.
-It is tempting to consider using the techniques described in
-Appendix~\ref{app:Aside on Undefined Behavior}
-that prevent back-propagation of UB, but these are ineffective in this
-case because the C++ implementation can detect and exploit UB when
-considering line~14 in isolation.
-Once the constant 1 is substituted for line~14's expression, the
-hardware is free to reorder lines~13 and~14 even if a compiler-based
-C++ implementation refrains from doing so.
-Note that this OOTA-like cycle can appear even if all of the atomic
-operations in
-Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}
-were converted to be volatile.
-
-Therefore, this paper focuses on UB-free programs, rare though these
-beasts might be among large-scale C++ programs running in production.
-
-\clearpage
-
-\section{Aside on Undefined Behavior}
-\label{app:Aside on Undefined Behavior}
-
-The combination of OOTA cycles and UB has proven especially
-challenging~\cite{DavidGoldblatt2019NoElegantOOTAfix,HansBoehm2020UBalternatives}.
-
-\subsection{Alignment-Based Undefined Behavior}
-\label{sec:Alignment-Based Undefined Behavior}
-
-The example shown in
-Listing~\ref{lst:Alignment-Based Undefined Behavior (P1916R0)}
-(taken from P1916R0) is especially instructive.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 std::atomic<int> y;
- 2 template <typename T>
- 3 void t1() {
- 4   long long r1 = x.load(std::memory_order_relaxed);
- 5   if (r1 % 4 == 0)
- 6     y.store(1, std::memory_order_relaxed);
- 7   *(T*)r1 = *(T*)r1;
- 8 }
-\end{verbatim}
-\caption{Alignment-Based Undefined Behavior (P1916R0)}
-\label{lst:Alignment-Based Undefined Behavior (P1916R0)}
-\end{listing}
-
-Because the compiler often assumes that UB cannot happen, it might
-look at line~7 and conclude that the value of \co{r1} necessarily
-corresponded to an address aligned per the requirements of type \co{T}.
-If that type's alignment was larger than four bytes, the compiler could
-then replace the condition in line~5 with \co{true}, which would mean
-that there was no semantic dependency from the load of \co{x} to the
-store of \co{y}.
-
-Except that the developer might have knowingly carried out a misaligned
-access, for example, to test an alignment trap handler or because the
-system on which this code was to run was known to gracefully handle
-misaligned accesses.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 std::atomic<int> y;
- 2 template <typename T>
- 3 void t1() {
- 4   long long r1 = x.load(std::memory_order_relaxed);
- 5   if (r1 % 4 == 0)
- 6     y.store(1, std::memory_order_relaxed);
- 7   std::observable();
- 8   *(T*)r1 = *(T*)r1;
- 9 }
-\end{verbatim}
-\caption{Alignment-Based Undefined Behavior Fix}
-\label{lst:Alignment-Based Undefined Behavior Fix}
-\end{listing}
-
-However, in that case, the developer is stepping outside the bounds
-of the standard.
-It would be good to have some way for the developer to get the job
-done without confusing the compiler, thus separating OOTA-cycle and
-UB concerns.
-One approach is to use the \co{std::observable()} function
-that has been proposed to the C++ standards
-committee~\cite{DavisHerring2021P1494R2}.
-This function would blocks backwards-propagating UB.
-Then this example could be rewritten as shown in
-Listing~\ref{lst:Alignment-Based Undefined Behavior Fix},
-with the asm on line~7 preventing the compiler from assuming that line~8
-is guaranteed to be executed, thus preventing the compiler from
-applying any lessons that it might learn from line~8 to line~5.
-
-% Clean up <200b> in vim: ":%s/\%u200b//g"
-
-There are a number of ways to implement \co{std::observable()},
-for example, a new \co{"mightnotreturn"} GCC asm clobber, a call to an empty
-function that is invisible to the compiler, and a GCC goto asm that the
-compiler cannot prove falls through.
-
-\subsection{Unreachability-Based Undefined Behavior}
-\label{sec:Unreachability-Based Undefined Behavior}
-
-This section attempts to apply the lessons
-of Appendix~\ref{sec:Unreachability-Based Undefined Behavior} to
-Goldblatt's unreachability-UB example.
-
-The \co{std::observable()} at line~12 of
-Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
-avoids confusing the compiler, instead forcing it to assume that the
-value of \co{local} might not be zero.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 // UB if this function is ever called.
- 2 [[noreturn]]
- 3 inline void unreachable() {
- 4   return;
- 5 }
- 6 void t1() {
- 7   r1 = x.load(std::memory_order_relaxed);
- 8   int local = (r1 & 1);
- 9   if (local == 0)
-10     y.store(1, std::memory_order_relaxed);
-11   if (local != 0) {
-12     std::observable();
-13     // If we get here, UB.
-14     unreachable();
-15   }
-16 }
-\end{verbatim}
-\caption{Unreachability-Based Undefined Behavior Fix}
-\label{lst:Unreachability-Based Undefined Behavior Fix}
-\end{listing}
-
-By the same reasoning, Goldblatt's \co{misbehave} example that
-also involves unreachability can also be addressed through insertion
-of the proposed \co{std::observable()}.
-
-\subsection{Atomic Stores From Undefined Behavior}
-\label{sec:Atomic Stores From Undefined Behavior}
-
-Another proposal to the C++ standards
-committee~\cite{HansBoehm2020UBalternatives}
-proposes (among other alternatives) that any atomic store resulting
-from undefined behavior have at least release semantics.
-This would prevent such a store from propagating undefined behavior
-backwards within that store's thread.
-In some cases, this could also prevent this undefined behavior from
-propagating back through other threads.
-
-However, it might well force C++ implementations to promote a surprisingly
-large fraction of a program's relaxed stores to release.
-
-\subsection{Lifetime-Based Undefined Behavior}
-\label{sec:Lifetime-Based Undefined Behavior}
-
-Not all undefined behavior can be backstopped using
-\co{std::unreachable()}.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 int *foo(int *r)
- 2 {
- 3   b.store(r, memory_order_relaxed);
- 4   return &global;
- 5 }
- 6
- 7 int go(int *p)
- 8 {
- 9   int r;
-10   int *q = foo(&r);
-11
-12   r = 2;
-13   *p = 1;
-14   *q = r;
-15   return *q - r;
-16 }
-17
-18 void t1(void)
-19 {
-20   int r1 = go(a.load(memory_order_relaxed));
-21   printf("go() = %d\n", r1);
-22 }
-23
-24 void t2(void)
-25 {
-26   int *r2 = b.load(memory_order_relaxed);
-27   a.store(r2, memory_order_relaxed);
-28 }
-\end{verbatim}
-\caption{Lifetime-Based Undefined Behavior}
-\label{lst:Lifetime-Based Undefined Behavior}
-\end{listing}
-
-Listing~\ref{lst:Lifetime-Based Undefined Behavior}
-shows an example loosely based on one presented by Stephen Dolan\footnote{
-	\url{https://stedolan.net/talks/2024/loadstore}}
-in which function \co{go()} invokes \co{foo()} with a pointer to \co{go()}'s
-local variable \co{r}.
-However, \co{foo()} uses a relaxed store to place this pointer in
-an atomic variable \co{b}.
-Because this store is relaxed, it is not ordered against other code
-in \co{go()} and \co{t1()}, which means that \co{t2()} might well
-read this pointer from \co{b} either before the beginning or after
-the end of \co{r}'s lifetime, neither of which is well-defined.
-
-Of course, any attempt to dereference the pointer read from either
-\co{a} or \co{b} could result in undefined behavior.
-
-The lesson is clear:
-Properly synchronize reads and writes of pointers with the lifetime
-of the pointed-to object.
-
-This example could make lines~3 and~27 be \co{memory_order_release} and
-lines~20 and~26 be \co{memory_order_acquire} in order to avoid carrying
-a pointer to \co{r} before the lifetime of \co{r} has started.
-Alternatively, threads could be spawned only after \co{r}'s lifetime
-starts and joined before it ends.
-Placing \co{r} in static global storage would avoid lifetime considerations
-totally.
-And there are of course any number of other synchronization schemes
-that might be used to ensure that pointers are in use only while the
-pointed-to objects are alive.
-
-Note that \co{std::unreachable()} does not suffice because the CPU
-can also reorder relaxed stores.
-Note also that load-store ordering is not sufficient to prevent a
-pointer to \co{r} from being stored to \co{b} prior to the start of
-\co{r}'s lifetime because there is no atomic load between the start
-of that lifetime and the relaxed store to \co{b}.
-
-There are a number of ways to prevent carrying a pointer to \co{r} after
-the lifetime of \co{r} has ended, and these are left as an exercise for
-the reader.
-
-\clearpage
-
-\section{Evaluating sdep Using \co{cbmc}}
-\label{sec:Evaluating sdep Using cbmc}
-
-This section demonstrates the evaluation of sdep using
-the C Bounded Model Checker
-(\co{cbmc})~\cite{EdmundClarke2004CBMC}.
-This tool takes a smallish C program as input and constructs a logic
-expression representing that program.
-The logic expression's variables represent bits in the program's inputs
-and state, and the result of that logic expression is \co{true} if that
-program has a bug, for example, if it is possible for that C program to
-trigger an assertion or execute an out-of-bounds array reference.
-The tool then hands this expression to a SAT solver, and if this solver
-finds a satisfying set of values, the tool uses those values provide
-the execution path to the bug.
 

### 2024-02-13 ootavacuous.tex
Move UB and cbmc appendices fo follow tooling appendix
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Add "litmus test" showing volatile effects
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\clearpage
-
-\section{Aside on Undefined Behavior}
-\label{app:Aside on Undefined Behavior}
-
-The combination of OOTA cycles and UB has proven especially
-challenging~\cite{DavidGoldblatt2019NoElegantOOTAfix,HansBoehm2020UBalternatives}.
-
-\subsection{Alignment-Based Undefined Behavior}
-\label{sec:Alignment-Based Undefined Behavior}
-
-The example shown in
-Listing~\ref{lst:Alignment-Based Undefined Behavior (P1916R0)}
-(taken from P1916R0) is especially instructive.
-
-\begin{listing}[tbp]
+\subsubsection{Semantic Dependencies and \co{volatile}}
+\label{app:Semantic Dependencies and volatile}
+
+The Linux kernel uses volatile accesses to constrain the compiler,
+and it is worth looking at the example shown in
+Listing~\ref{lst:Non-Volatile Accesses and Dependencies}
+to see how this works.
+
+\begin{listing}
 \begin{verbatim}
- 1 std::atomic<int> y;
- 2 template <typename T>
- 3 void t1() {
- 4   long long r1 = x.load(std::memory_order_relaxed);
- 5   if (r1 % 4 == 0)
- 6     y.store(1, std::memory_order_relaxed);
- 7   *(T*)r1 = *(T*)r1;
- 8 }
+ 1 atomic<int> x, y;
+ 2
+ 3 void thread1() {
+ 4   int r1 = x.load(memory_order::relaxed);
+ 5   y.store(r1, memory_order::relaxed);
+ 6 }
 \end{verbatim}
-\caption{Alignment-Based Undefined Behavior (P1916R0)}
-\label{lst:Alignment-Based Undefined Behavior (P1916R0)}
-\end{listing}
-
-Because the compiler often assumes that UB cannot happen, it might
-look at line~7 and conclude that the value of \co{r1} necessarily
-corresponded to an address aligned per the requirements of type \co{T}.
-If that type's alignment was larger than four bytes, the compiler could
-then replace the condition in line~5 with \co{true}, which would mean
-that there was no semantic dependency from the load of \co{x} to the
-store of \co{y}.
-
-Except that the developer might have knowingly carried out a misaligned
-access, for example, to test an alignment trap handler or because the
-system on which this code was to run was known to gracefully handle
-misaligned accesses.
-
-\begin{listing}[tbp]
+\caption{Non-Volatile Accesses and Dependencies}
+\label{lst:Non-Volatile Accesses and Dependencies}
+\end{listing}
+
+Because both \co{x} and \co{y} are non-volatile, the compiler is
+free to assume that the value loaded by line~4 will be some value
+stored to \co{x} or its initial value.
+Because there are no stores to \co{x}, the only remaining possibility
+is the default-initialized value of zero.
+Therefore, the C++ implementation is within its rights to substitute
+the constant zero for \co{r1} on line~5, eliminating the semantic
+dependency that would otherwise extend from line~4 to line~5.
+This in turn enables the implementation to reorder line~5 to
+precede line~4.
+
+\begin{listing}
 \begin{verbatim}
- 1 std::atomic<int> y;
- 2 template <typename T>
- 3 void t1() {
- 4   long long r1 = x.load(std::memory_order_relaxed);
- 5   if (r1 % 4 == 0)
- 6     y.store(1, std::memory_order_relaxed);
- 7   std::observable();
- 8   *(T*)r1 = *(T*)r1;
- 9 }
+ 1 atomic<volatile int> x, y;
+ 2
+ 3 void thread1() {
+ 4   int r1 = x.load(memory_order::relaxed);
+ 5   y.store(r1, memory_order::relaxed);
+ 6 }
 \end{verbatim}
-\caption{Alignment-Based Undefined Behavior Fix}
-\label{lst:Alignment-Based Undefined Behavior Fix}
-\end{listing}
-
-However, in that case, the developer is stepping outside the bounds
-of the standard.
-It would be good to have some way for the developer to get the job
-done without confusing the compiler, thus separating OOTA-cycle and
-UB concerns.
-One approach is to use the \co{std::observable()} function
-that has been proposed to the C++ standards
-committee~\cite{DavisHerring2021P1494R2}.
-This function would blocks backwards-propagating UB.
-Then this example could be rewritten as shown in
-Listing~\ref{lst:Alignment-Based Undefined Behavior Fix},
-with the asm on line~7 preventing the compiler from assuming that line~8
-is guaranteed to be executed, thus preventing the compiler from
-applying any lessons that it might learn from line~8 to line~5.
-
-% Clean up <200b> in vim: ":%s/\%u200b//g"
-
-There are a number of ways to implement \co{std::observable()},
-for example, a new \co{"mightnotreturn"} GCC asm clobber, a call to an empty
-function that is invisible to the compiler, and a GCC goto asm that the
-compiler cannot prove falls through.
-
-\subsection{Unreachability-Based Undefined Behavior}
-\label{sec:Unreachability-Based Undefined Behavior}
-
-This section attempts to apply the lessons
-of Appendix~\ref{sec:Unreachability-Based Undefined Behavior} to
-Goldblatt's unreachability-UB example.
-
-The \co{std::observable()} at line~12 of
-Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
-avoids confusing the compiler, instead forcing it to assume that the
-value of \co{local} might not be zero.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 // UB if this function is ever called.
- 2 [[noreturn]]
- 3 inline void unreachable() {
- 4   return;
- 5 }
- 6 void t1() {
- 7   r1 = x.load(std::memory_order_relaxed);
- 8   int local = (r1 & 1);
- 9   if (local == 0)
-10     y.store(1, std::memory_order_relaxed);
-11   if (local != 0) {
-12     std::observable();
-13     // If we get here, UB.
-14     unreachable();
-15   }
-16 }
-\end{verbatim}
-\caption{Unreachability-Based Undefined Behavior Fix}
-\label{lst:Unreachability-Based Undefined Behavior Fix}
-\end{listing}
-
-By the same reasoning, Goldblatt's \co{misbehave} example that
-also involves unreachability can also be addressed through insertion
-of the proposed \co{std::observable()}.
-
-\subsection{Atomic Stores From Undefined Behavior}
-\label{sec:Atomic Stores From Undefined Behavior}
-
-Another proposal to the C++ standards
-committee~\cite{HansBoehm2020UBalternatives}
-proposes (among other alternatives) that any atomic store resulting
-from undefined behavior have at least release semantics.
-This would prevent such a store from propagating undefined behavior
-backwards within that store's thread.
-In some cases, this could also prevent this undefined behavior from
-propagating back through other threads.
-
-However, it might well force C++ implementations to promote a surprisingly
-large fraction of a program's relaxed stores to release.
-
-\subsection{Lifetime-Based Undefined Behavior}
-\label{sec:Lifetime-Based Undefined Behavior}
-
-Not all undefined behavior can be backstopped using
-\co{std::unreachable()}.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 int *foo(int *r)
- 2 {
- 3   b.store(r, memory_order_relaxed);
- 4   return &global;
- 5 }
- 6
- 7 int go(int *p)
- 8 {
- 9   int r;
-10   int *q = foo(&r);
-11
-12   r = 2;
-13   *p = 1;
-14   *q = r;
-15   return *q - r;
-16 }
-17
-18 void t1(void)
-19 {
-20   int r1 = go(a.load(memory_order_relaxed));
-21   printf("go() = %d\n", r1);
-22 }
-23
-24 void t2(void)
-25 {
-26   int *r2 = b.load(memory_order_relaxed);
-27   a.store(r2, memory_order_relaxed);
-28 }
-\end{verbatim}
-\caption{Lifetime-Based Undefined Behavior}
-\label{lst:Lifetime-Based Undefined Behavior}
-\end{listing}
-
-Listing~\ref{lst:Lifetime-Based Undefined Behavior}
-shows an example loosely based on one presented by Stephen Dolan\footnote{
-	\url{https://stedolan.net/talks/2024/loadstore}}
-in which function \co{go()} invokes \co{foo()} with a pointer to \co{go()}'s
-local variable \co{r}.
-However, \co{foo()} uses a relaxed store to place this pointer in
-an atomic variable \co{b}.
-Because this store is relaxed, it is not ordered against other code
-in \co{go()} and \co{t1()}, which means that \co{t2()} might well
-read this pointer from \co{b} either before the beginning or after
-the end of \co{r}'s lifetime, neither of which is well-defined.
-
-Of course, any attempt to dereference the pointer read from either
-\co{a} or \co{b} could result in undefined behavior.
-
-The lesson is clear:
-Properly synchronize reads and writes of pointers with the lifetime
-of the pointed-to object.
-
-This example could make lines~3 and~27 be \co{memory_order_release} and
-lines~20 and~26 be \co{memory_order_acquire} in order to avoid carrying
-a pointer to \co{r} before the lifetime of \co{r} has started.
-Alternatively, threads could be spawned only after \co{r}'s lifetime
-starts and joined before it ends.
-Placing \co{r} in static global storage would avoid lifetime considerations
-totally.
-And there are of course any number of other synchronization schemes
-that might be used to ensure that pointers are in use only while the
-pointed-to objects are alive.
-
-Note that \co{std::unreachable()} does not suffice because the CPU
-can also reorder relaxed stores.
-Note also that load-store ordering is not sufficient to prevent a
-pointer to \co{r} from being stored to \co{b} prior to the start of
-\co{r}'s lifetime because there is no atomic load between the start
-of that lifetime and the relaxed store to \co{b}.
-
-There are a number of ways to prevent carrying a pointer to \co{r} after
-the lifetime of \co{r} has ended, and these are left as an exercise for
-the reader.
-
-\clearpage
-
-\section{Evaluating sdep Using \co{cbmc}}
-\label{sec:Evaluating sdep Using cbmc}
-
-This section demonstrates the evaluation of sdep using
-the C Bounded Model Checker
-(\co{cbmc})~\cite{EdmundClarke2004CBMC}.
-This tool takes a smallish C program as input and constructs a logic
-expression representing that program.
-The logic expression's variables represent bits in the program's inputs
-and state, and the result of that logic expression is \co{true} if that
-program has a bug, for example, if it is possible for that C program to
-trigger an assertion or execute an out-of-bounds array reference.
-The tool then hands this expression to a SAT solver, and if this solver
-finds a satisfying set of values, the tool uses those values provide
-the execution path to the bug.
-
-But first, Appendix~\ref{app:Informal Definition of Semantic Dependency}
-outlines an informal informal definition of ``semantic dependency''
-that can be used in \co{cbmc} test cases.
-
-\subsection{Informal Definition of Semantic Dependency}
-\label{app:Informal Definition of Semantic Dependency}
-
-The manual analysis uses the informal rule that a semantic dependency
-exists between a set of relaxed atomic loads\footnote{
-	Please see
-	Appendix~\ref{app:Non-Trivial Semantic Dependencies}
-	for examples showing why this is a set of loads rather than a
-	single load.}
-and a store on any execution where different values loaded would change
-the address stored to, the value stored, or the number of times that
-store would be executed, with the proviso that stores of the same value
-to the same address having the same ordering properties are fungible.
-
-Producing a precise formal statement of this rule turns out to be quite
-challenging, but as noted earlier, there is considerable work along
-various
-lines~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA}.
-For examples of the challenges, please see
-Section~\ref{sec:On OOTA Definitions}.
-Our informal definition meets these challenges as follows:
-
-\begin{enumerate}
-\item	Section~\ref{sec:Semantic Dependencies and Source Code}
-	is responsible for the phrase ``on any execution'' as well
-	as the proviso regarding fungible loads.
-\item	Section~\ref{sec:Semantic Dependencies Depend on Context}'s
-	challenges are met by looking at the full context within
-	a given thread.
-\item	Section~\ref{sec:Semantic Dependencies Can Be Nondeterministic}'s
-	non-determinism challenges are avoided by avoiding litmus tests
-	exhibiting non-deterministic behavior.
-	Again, note that most production C++ code uses compiler-based C++
-	implementations, which freeze their non-deterministic choices
-	into the binary at compile time.
-\item	Section~\ref{sec:Semantic Dependencies And Optimization}'s
-	optimization-choice challenges are met by using the C++
-	implementation's choice of semantic dependency based on the
-	optimizations that the implementation actually applied.
-\item	Section~\ref{sec:Semantic Dependencies and I/O}'s
-	I/O challenges are met by again noting that I/O is observed
-	behavior.
-	And by avoiding litmus tests involving I/O, given that relatively
-	few of the available tools handle I/O.
-\item	Section~\ref{sec:Semantic Dependencies Can Be Complex}'s
-	complexity challenges are met by looking at what the code
-	actually does.
-	For example, if an as-yet unproven mathematical conjecture
-	holds, the code will act consistently with that conjecture,
-	and vice versa.
-\item	Section~\ref{sec:Implementers and Users Influence Semantic-Dependencies Definition}'s
-	implementer-/user-choice challenges are met by accounting for
-	those choices.
-	For example, if a user has specified \co{-funsigned-char},
-	then the actual executions will be consistent with that choice.
-	For more examples, please see
-	Appendix~\ref{app:User Influence Over Language Semantics}.
-\end{enumerate}
-
-In short, we are treating ``semantic dependency'' in a manner similar
-to our treatment of the laws of physics.
-We let the actual executions provide an empirical basis for the analysis.
-
-We do not expect this approach to please everyone.
-However, those who are displeased might do well to reflect on the fact
-that the C++ standard is not mathematical in nature, which implies that
-any mathematical model of C++ is inherently an approximation.
-We look forward to the day when the standard gains more mathematical
-rigor, and perhaps the C++ memory model is a first step in that
-direction.\footnote{
-	But if so, the OOTA situation should be a caution!}
-Nevertheless, until that day, we must live with the standard in its
-current decidedly non-mathematical form.
-
-\subsection{Complex Semantic Dependencies}
-\label{app:Complex Semantic Dependencies}
-
-Listing~\ref{lst:Example Code for Evaluation of sdep}
-shows a \co{herd7} thread \co{P0()}.
-Is there a semantic dependency between the load from \co{x} on
-line~3 to the store to \co{y} on line~9?
-Answering this question requires solving a mathematical equation,
-not all of which currently have known solutions.
-However, this section will use a relatively simple equation combined
-with restrictions on domain and range that permit \co{cbmc} to do
-the job.
-
-\begin{listing}[tbp]
-\scriptsize
-\begin{verbatim}
- 1 P0(atomic_int *x, atomic_int *y, atomic_int *limit)
- 2 {
- 3   r1 = atomic_load_explicit(x, memory_order_relaxed);
- 4   r2 = atomic_load_explicit(limit, memory_order_relaxed);
- 5   r3 = r1 * r1 + 2 * r1 + 2;
- 6   r4 = r2;
- 7   if (r2 >= r3)
- 8     r4 = r3;
- 9   atomic_store_explicit(y, r4, memory_order_relaxed);
-10 }
-\end{verbatim}
-\caption{Example Code for Evaluation of sdep}
-\label{lst:Example Code for Evaluation of sdep}
-\end{listing}
-
-\begin{listing}[tbp]
-\scriptsize
-\begin{verbatim}
- 1 int main(int argc, char *argv[])
- 2 {
- 3   int i;
- 4   int y[2] = { 0, 0 };
- 5
- 6   for (i = 0; i < 2; i++) {
- 7     int r1;
- 8     int r2;
- 9     int r3;
-10     int *yp = &y[i];
-11
-12     r1 = limited_arg(argv[2 * i + 1]);
-13     r2 = const_arg(limited_arg(argv[2 * i + 2]));
-14     r3 = r1 * r1 + 2 * r1 + 2;
-15     *yp = r2 <= r3 ? r2 : r3;
-16     printf("i = %d, r1 = %d, r2 = %d, r3 = %d\n", i, r1, r2, r3);
-17   }
-18   assert(y[0] == y[1]);
-19 }
-\end{verbatim}
-\caption{Code for cbmc-Based Evaluation of sdep}
-\label{lst:Code for cbmc-Based Evaluation of sdep}
-\end{listing}
-
-One way to answer this question is to create a C program
-(see Listing~\ref{lst:Code for cbmc-Based Evaluation of sdep})
-that runs the \co{P0()} thread's code twice with arbitrary inputs
-(the loop spanning lines~6-17) and asserts that the results of
-each iteration be equal (line~18).
-
-Running this with:
-
-\begin{quote}
-\scriptsize
-\begin{verbatim}
-cbmc -DCBMC cbmc/sdep-quadratic.c
-\end{verbatim}
-\end{quote}
-
-gives the result \co{VERIFICATION FAILED}.
-This indicates that there are different values of \co{x} and \co{limit}
-that yield different values of \co{y}, that is, that there can be
-a semantic dependency.
-
-\begin{listing}[tbp]
-@@ DisplayLitmus cbmc/sdep-quadratic.c @@
-\caption{Full Code for cbmc-Based Evaluation of sdep}
-\label{lst:Full Code for cbmc-Based Evaluation of sdep}
-\end{listing}
-
-However, $x^2 + 2x + 2$ has a minimum value of 1 at $x=-1$, so
-forcing \co{limit} to (say) zero should eliminate this static
-dependency:\footnote{
-	See Listing~\ref{lst:Full Code for cbmc-Based Evaluation of sdep}
-	for the full \path{cbmc/sdep-quadratic.c} program.}
-
-\begin{quote}
-\scriptsize
-\begin{verbatim}
-cbmc -DCBMC -DLOAD_CONSTANT=0 cbmc/sdep-quadratic.c
-\end{verbatim}
-\end{quote}
-
-But this also gives the result \co{VERIFICATION FAILED}.
-One reason for this is that this program does 32-bit computer arithmetic,
-and is thus subject to wraparound.
-Limiting the value of \co{x} to 46,339 prevents wraparound:
-
-\begin{quote}
-\scriptsize
-\begin{verbatim}
-cbmc -DCBMC -DLOAD_CONSTANT=0 -DLOAD_LIMITS=46339 cbmc/sdep-quadratic.c
-\end{verbatim}
-\end{quote}
-
-And this yields \co{VERIFICATION SUCCESSFUL}, indicating that there
-is no semantic dependency in this case.
-This is to be expected because the specified value of zero for \co{limit}
-is smaller than the minimum of the quadratic expression, so that the
-value stored to \co{y} is always zero.
-Specifying \co{-DLOAD_CONSTANT=1} also yields \co{VERIFICATION SUCCESSFUL}
-because this value is equal to that minimum, so that the value stored to
-\co{y} is always 1.
-However, specifying any larger value (for example, \co{-DLOAD_CONSTANT=2})
-yields \co{VERIFICATION FAILED} because the value assigned to \co{y}
-might be either 1 or 2, indicating that a semantic dependency exists.
-
-This situation underscores the point that the existance of a semantic
-dependency can be a property of a particular execution rather than a
-universal property of the code itself.
-The interested reader is encouraged to create more complex examples,
-for example, demonstrating semantic dependencies from multiple loads or
-exploring indeterminacy.
-
-There are numerous other tools that can be used, up to and including
-manual inspection, but \co{cbmc} is readily available and easy to
-use.\footnote{
-	On Debian-based distributions, via \co{sudo apt install cbmc}.}
-
-\subsection{C++ Non-Determinism}
-\label{app:C++ Non-Determinism}
-
-The C++ standard allows non-determinism, for example, the order of
-invocation of side-effects in some circumstances.
-Listing~\ref{lst:Code for cbmc-Based Evaluation of Non-Deterministic sdep}
-shows an example leveraging side-effects in the parameter list shown on
-line~37.
-
-\begin{listing}[tbp]
-@@ DisplayLitmus cbmc/sdep-nondet.c @@
-\caption{Code for cbmc-Based Evaluation of Non-Deterministic sdep}
-\label{lst:Code for cbmc-Based Evaluation of Non-Deterministic sdep}
-\end{listing}
-
-Running \co{cbmc} on this as follows:
-
-\begin{quote}
-\scriptsize
-\begin{verbatim}
-cbmc -DCBMC cbmc/sdep-nondet.c
-\end{verbatim}
-\end{quote}
-
-Gives the result \co{VERIFICATION SUCCESSFUL}, showing that \co{cbmc}
-does not model this non-determinism.
-Which is not surprising, given that many of the safety-critical uses
-to which \co{cbmc} is put would prohibit such non-determinism in their
-code bases.
+\caption{Volatile Accesses and Dependencies}
+\label{lst:Volatile Accesses and Dependencies}
+\end{listing}
+
+One way to preserve this dependency is use of \co{volatile}, as shown in
+Listing~\ref{lst:Volatile Accesses and Dependencies}.
+Now the implementation is forbidden from assuming that the value loaded
+by line~4 has any relation to any stores to or initialization of \co{x},
+even if \co{x} is placed in normal memory.
+
+In this case, the \co{volatile} keyword alerts the implementation to
+the possibility of interfering accesses to \co{x} from outside the
+program, for example, due to \co{x} being:
+
+\begin{itemize}
+\item	Allocated in an MMIO region of the address space.
+\item	Modified by an unknown-to-the-implementation thread
+	or signal handler.
+\item	Subject to I/O-device DMA operations.
+\item	Modified by debugger commands.
+\item	Modified by accesses to facilities such as \path{/dev/mem}.
+\item	Modified by as-yet-unwritten dynamically linked libraries.
+\item	Modified by some other mechanism of the reader's choosing.
+\end{itemize}
+
+These possibilities force the implementation to preserve the semantic
+dependency from line~4 to line~5.
+They can also prevent the formation of OOTA cycles local to the
+implementation because such inteference means that the volatile loads
+might be reading from something that the implementation knows nothing
+about.
+
+The developer is free to arrange for \co{x} to be free of any
+interference, thus obtaining predictable behavior along with the
+preserved semantic dependency.
+However, this example invalidates the rough definition of semantic
+dependency from
+Section~\ref{sec:Simple OOTA Cycle}
+because the value loaded from \co{x} will always be the initial value
+of zero, so that it is at best dubious to talk about any changes in
+the value loaded.
+
+One way forward is to assume that interference could happen when
+determining which dependencies are semantic, and then do further analysis
+using that determination.
 

### 2024-02-13 ootavacuous.tex
Remove Appendices F, G, and H
F Evaluating sdep Using cbmc
G Debates on Causality
H Cutting-Room Floor

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-02-13 ootavacuous.tex
@@
 
-\clearpage
-
-\section{Evaluating sdep Using \co{cbmc}}
-\label{sec:Evaluating sdep Using cbmc}
-
-This section demonstrates the evaluation of sdep using
-the C Bounded Model Checker
-(\co{cbmc})~\cite{EdmundClarke2004CBMC}.
-This tool takes a smallish C program as input and constructs a logic
-expression representing that program.
-The logic expression's variables represent bits in the program's inputs
-and state, and the result of that logic expression is \co{true} if that
-program has a bug, for example, if it is possible for that C program to
-trigger an assertion or execute an out-of-bounds array reference.
-The tool then hands this expression to a SAT solver, and if this solver
-finds a satisfying set of values, the tool uses those values provide
-the execution path to the bug.
-
-But first, Appendix~\ref{app:Informal Definition of Semantic Dependency}
-outlines an informal informal definition of ``semantic dependency''
-that can be used in \co{cbmc} test cases.
-
-\subsection{Informal Definition of Semantic Dependency}
-\label{app:Informal Definition of Semantic Dependency}
-
-The manual analysis uses the informal rule that a semantic dependency
-exists between a set of relaxed atomic loads\footnote{
-	Please see
-	Appendix~\ref{app:Non-Trivial Semantic Dependencies}
-	for examples showing why this is a set of loads rather than a
-	single load.}
-and a store on any execution where different values loaded would change
-the address stored to, the value stored, or the number of times that
-store would be executed, with the proviso that stores of the same value
-to the same address having the same ordering properties are fungible.
-
-Producing a precise formal statement of this rule turns out to be quite
-challenging, but as noted earlier, there is considerable work along
-various
-lines~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA}.
-For examples of the challenges, please see
-Section~\ref{sec:On OOTA Definitions}.
-Our informal definition meets these challenges as follows:
-
-\begin{enumerate}
-\item	Section~\ref{sec:Semantic Dependencies and Source Code}
-	is responsible for the phrase ``on any execution'' as well
-	as the proviso regarding fungible loads.
-\item	Section~\ref{sec:Semantic Dependencies Depend on Context}'s
-	challenges are met by looking at the full context within
-	a given thread.
-\item	Section~\ref{sec:Semantic Dependencies Can Be Nondeterministic}'s
-	non-determinism challenges are avoided by avoiding litmus tests
-	exhibiting non-deterministic behavior.
-	Again, note that most production C++ code uses compiler-based C++
-	implementations, which freeze their non-deterministic choices
-	into the binary at compile time.
-\item	Section~\ref{sec:Semantic Dependencies And Optimization}'s
-	optimization-choice challenges are met by using the C++
-	implementation's choice of semantic dependency based on the
-	optimizations that the implementation actually applied.
-\item	Section~\ref{sec:Semantic Dependencies and I/O}'s
-	I/O challenges are met by again noting that I/O is observed
-	behavior.
-	And by avoiding litmus tests involving I/O, given that relatively
-	few of the available tools handle I/O.
-\item	Section~\ref{sec:Semantic Dependencies Can Be Complex}'s
-	complexity challenges are met by looking at what the code
-	actually does.
-	For example, if an as-yet unproven mathematical conjecture
-	holds, the code will act consistently with that conjecture,
-	and vice versa.
-\item	Section~\ref{sec:Implementers and Users Influence Semantic-Dependencies Definition}'s
-	implementer-/user-choice challenges are met by accounting for
-	those choices.
-	For example, if a user has specified \co{-funsigned-char},
-	then the actual executions will be consistent with that choice.
-	For more examples, please see
-	Appendix~\ref{app:User Influence Over Language Semantics}.
-\end{enumerate}
-
-In short, we are treating ``semantic dependency'' in a manner similar
-to our treatment of the laws of physics.
-We let the actual executions provide an empirical basis for the analysis.
-
-We do not expect this approach to please everyone.
-However, those who are displeased might do well to reflect on the fact
-that the C++ standard is not mathematical in nature, which implies that
-any mathematical model of C++ is inherently an approximation.
-We look forward to the day when the standard gains more mathematical
-rigor, and perhaps the C++ memory model is a first step in that
-direction.\footnote{
-	But if so, the OOTA situation should be a caution!}
-Nevertheless, until that day, we must live with the standard in its
-current decidedly non-mathematical form.
-
-\subsection{Complex Semantic Dependencies}
-\label{app:Complex Semantic Dependencies}
-
-Listing~\ref{lst:Example Code for Evaluation of sdep}
-shows a \co{herd7} thread \co{P0()}.
-Is there a semantic dependency between the load from \co{x} on
-line~3 to the store to \co{y} on line~9?
-Answering this question requires solving a mathematical equation,
-not all of which currently have known solutions.
-However, this section will use a relatively simple equation combined
-with restrictions on domain and range that permit \co{cbmc} to do
-the job.
-
-\begin{listing}[tbp]
-\scriptsize
-\begin{verbatim}
- 1 P0(atomic_int *x, atomic_int *y, atomic_int *limit)
- 2 {
- 3   r1 = atomic_load_explicit(x, memory_order_relaxed);
- 4   r2 = atomic_load_explicit(limit, memory_order_relaxed);
- 5   r3 = r1 * r1 + 2 * r1 + 2;
- 6   r4 = r2;
- 7   if (r2 >= r3)
- 8     r4 = r3;
- 9   atomic_store_explicit(y, r4, memory_order_relaxed);
-10 }
-\end{verbatim}
-\caption{Example Code for Evaluation of sdep}
-\label{lst:Example Code for Evaluation of sdep}
-\end{listing}
-
-\begin{listing}[tbp]
-\scriptsize
-\begin{verbatim}
- 1 int main(int argc, char *argv[])
- 2 {
- 3   int i;
- 4   int y[2] = { 0, 0 };
- 5
- 6   for (i = 0; i < 2; i++) {
- 7     int r1;
- 8     int r2;
- 9     int r3;
-10     int *yp = &y[i];
-11
-12     r1 = limited_arg(argv[2 * i + 1]);
-13     r2 = const_arg(limited_arg(argv[2 * i + 2]));
-14     r3 = r1 * r1 + 2 * r1 + 2;
-15     *yp = r2 <= r3 ? r2 : r3;
-16     printf("i = %d, r1 = %d, r2 = %d, r3 = %d\n", i, r1, r2, r3);
-17   }
-18   assert(y[0] == y[1]);
-19 }
-\end{verbatim}
-\caption{Code for cbmc-Based Evaluation of sdep}
-\label{lst:Code for cbmc-Based Evaluation of sdep}
-\end{listing}
-
-One way to answer this question is to create a C program
-(see Listing~\ref{lst:Code for cbmc-Based Evaluation of sdep})
-that runs the \co{P0()} thread's code twice with arbitrary inputs
-(the loop spanning lines~6-17) and asserts that the results of
-each iteration be equal (line~18).
-
-Running this with:
-
-\begin{quote}
-\scriptsize
-\begin{verbatim}
-cbmc -DCBMC cbmc/sdep-quadratic.c
-\end{verbatim}
-\end{quote}
-
-gives the result \co{VERIFICATION FAILED}.
-This indicates that there are different values of \co{x} and \co{limit}
-that yield different values of \co{y}, that is, that there can be
-a semantic dependency.
-
-\begin{listing}[tbp]
-@@ DisplayLitmus cbmc/sdep-quadratic.c @@
-\caption{Full Code for cbmc-Based Evaluation of sdep}
-\label{lst:Full Code for cbmc-Based Evaluation of sdep}
-\end{listing}
-
-However, $x^2 + 2x + 2$ has a minimum value of 1 at $x=-1$, so
-forcing \co{limit} to (say) zero should eliminate this static
-dependency:\footnote{
-	See Listing~\ref{lst:Full Code for cbmc-Based Evaluation of sdep}
-	for the full \path{cbmc/sdep-quadratic.c} program.}
-
-\begin{quote}
-\scriptsize
-\begin{verbatim}
-cbmc -DCBMC -DLOAD_CONSTANT=0 cbmc/sdep-quadratic.c
-\end{verbatim}
-\end{quote}
-
-But this also gives the result \co{VERIFICATION FAILED}.
-One reason for this is that this program does 32-bit computer arithmetic,
-and is thus subject to wraparound.
-Limiting the value of \co{x} to 46,339 prevents wraparound:
-
-\begin{quote}
-\scriptsize
-\begin{verbatim}
-cbmc -DCBMC -DLOAD_CONSTANT=0 -DLOAD_LIMITS=46339 cbmc/sdep-quadratic.c
-\end{verbatim}
-\end{quote}
-
-And this yields \co{VERIFICATION SUCCESSFUL}, indicating that there
-is no semantic dependency in this case.
-This is to be expected because the specified value of zero for \co{limit}
-is smaller than the minimum of the quadratic expression, so that the
-value stored to \co{y} is always zero.
-Specifying \co{-DLOAD_CONSTANT=1} also yields \co{VERIFICATION SUCCESSFUL}
-because this value is equal to that minimum, so that the value stored to
-\co{y} is always 1.
-However, specifying any larger value (for example, \co{-DLOAD_CONSTANT=2})
-yields \co{VERIFICATION FAILED} because the value assigned to \co{y}
-might be either 1 or 2, indicating that a semantic dependency exists.
-
-This situation underscores the point that the existance of a semantic
-dependency can be a property of a particular execution rather than a
-universal property of the code itself.
-The interested reader is encouraged to create more complex examples,
-for example, demonstrating semantic dependencies from multiple loads or
-exploring indeterminacy.
-
-There are numerous other tools that can be used, up to and including
-manual inspection, but \co{cbmc} is readily available and easy to
-use.\footnote{
-	On Debian-based distributions, via \co{sudo apt install cbmc}.}
-
-\subsection{C++ Non-Determinism}
-\label{app:C++ Non-Determinism}
-
-The C++ standard allows non-determinism, for example, the order of
-invocation of side-effects in some circumstances.
-Listing~\ref{lst:Code for cbmc-Based Evaluation of Non-Deterministic sdep}
-shows an example leveraging side-effects in the parameter list shown on
-line~37.
-
-\begin{listing}[tbp]
-@@ DisplayLitmus cbmc/sdep-nondet.c @@
-\caption{Code for cbmc-Based Evaluation of Non-Deterministic sdep}
-\label{lst:Code for cbmc-Based Evaluation of Non-Deterministic sdep}
-\end{listing}
-
-Running \co{cbmc} on this as follows:
-
-\begin{quote}
-\scriptsize
-\begin{verbatim}
-cbmc -DCBMC cbmc/sdep-nondet.c
-\end{verbatim}
-\end{quote}
-
-Gives the result \co{VERIFICATION SUCCESSFUL}, showing that \co{cbmc}
-does not model this non-determinism.
-Which is not surprising, given that many of the safety-critical uses
-to which \co{cbmc} is put would prohibit such non-determinism in their
-code bases.
-
-\clearpage
-
-\section{Debates on Causality}
-\label{sec:Debates on Causality}
-
-This Appendix is provided as a service to readers who want to know more
-about the ongoing debates on causality, provided by the combined efforts
-of Bard, Chat-GPT, and Bing Co-Pilot.
-References may be found here:
-
-{ \scriptsize
-\url{https://docs.google.com/document/d/1MJ6pQ4TcWchUQvoI-qhLjdhPVYF14AytbTW3gr4-xag/edit?usp=sharing}
-}
-
-The concept of causality, the idea that events are caused by preceding
-events in a predictable and consistent manner, is fundamental to many
-scientific theories and our understanding of the world.
-However, when it comes to the universe as a whole, particularly at the
-most fundamental levels such as in quantum physics or the origins of
-the universe, the idea of strict causality can become more complex.
-
-In classical physics, causality is often described in terms of
-deterministic laws, where events are determined by prior causes.
-But at the quantum level, the behavior of particles can be probabilistic
-and uncertain.
-Quantum mechanics introduces a level of randomness and uncertainty that
-challenges our classical notions of causality.
-
-Additionally, when considering the origin of the universe (the Big Bang),
-the laws of physics as we understand them break down at extremely high
-densities and temperatures.
-The conditions at the very beginning of the universe might not conform
-to our classical understanding of causality, and some theories suggest
-that time itself might have begun with the Big Bang, making it difficult
-to define causality in a traditional sense within such a framework.
-
-On the one hand, there is a strong case to be made that the universe
-is causal.
-Our understanding of physics is based on the principle of causality,
-which states that every event has a cause.
-This principle has been incredibly successful in explaining and predicting
-the behavior of the universe at large scales.
-Additionally, our own personal experience seems to confirm causality.
-We constantly observe events being caused by other events, and we have
-no reason to believe that this is an illusion.
-
-However, there are also some challenges to the idea of a perfectly
-causal universe.
-One challenge comes from quantum mechanics, which suggests that some
-events may be fundamentally non-causal.
-For example, the collapse of a wave function in quantum mechanics seems
-to be a random event, with no underlying cause.
-Additionally, our understanding of the universe's origins is still
-very limited.
-We don't know whether the Big Bang itself had a cause, or whether it
-was simply the beginning of everything.
-
-In summary, while causality is a fundamental concept in our understanding
-of the universe, its application and implications can become more complex,
-especially at the quantum level and regarding the origin of the universe.
-There are ongoing debates and explorations within physics and philosophy
-about the nature of causality in these contexts.
-
-\section{Cutting-Room Floor}
-\label{sec:Cutting-Room Floor}
-
-This appendix includes material that was removed but which might later
-prove useful.
-
-@@@ Because there is a cycle, on a real system, at least one of the links
-must be atemporal, that is, either a load returns a value before
-that value is stored, or a store's address or value is computed before
-a prior load returns a value used in that computation.
-
-@@@ Note that C++ implementations are permitted to evaluate to a more
-general notion of OOTA values in cases involving unspecified or undefined
-behavior (UB), such as use of uninitialized objects.
-However, these situations do not involve OOTA cycles, but rather user
-errors and other issues that can lead to unfriendly optimizations that
-in turn result in unpredictable output.
-This paper focuses instead on OOTA cycles.
-
 % \section{History}

### 2024-02-13 ootavacuous.tex
Remove stray reference to Appendix E ("Aside on Undefined Behavior")
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 code that would have resulted in division-by-zero UB.
-It is tempting to consider using the techniques described in
-Appendix~\ref{app:Aside on Undefined Behavior}
+It is tempting to consider using techniques
+% Appendix~\ref{app:Aside on Undefined Behavior}
 that prevent back-propagation of UB, but these are ineffective in this

### 2024-02-13 ootavacuous.tex
Remove Appendix E ("Aside on Undefined Behavior")
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\clearpage
-
-\section{Aside on Undefined Behavior}
-\label{app:Aside on Undefined Behavior}
-
-The combination of OOTA cycles and UB has proven especially
-challenging~\cite{DavidGoldblatt2019NoElegantOOTAfix,HansBoehm2020UBalternatives}.
-
-\subsection{Alignment-Based Undefined Behavior}
-\label{sec:Alignment-Based Undefined Behavior}
-
-The example shown in
-Listing~\ref{lst:Alignment-Based Undefined Behavior (P1916R0)}
-(taken from P1916R0) is especially instructive.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 std::atomic<int> y;
- 2 template <typename T>
- 3 void t1() {
- 4   long long r1 = x.load(std::memory_order_relaxed);
- 5   if (r1 % 4 == 0)
- 6     y.store(1, std::memory_order_relaxed);
- 7   *(T*)r1 = *(T*)r1;
- 8 }
-\end{verbatim}
-\caption{Alignment-Based Undefined Behavior (P1916R0)}
-\label{lst:Alignment-Based Undefined Behavior (P1916R0)}
-\end{listing}
-
-Because the compiler often assumes that UB cannot happen, it might
-look at line~7 and conclude that the value of \co{r1} necessarily
-corresponded to an address aligned per the requirements of type \co{T}.
-If that type's alignment was larger than four bytes, the compiler could
-then replace the condition in line~5 with \co{true}, which would mean
-that there was no semantic dependency from the load of \co{x} to the
-store of \co{y}.
-
-Except that the developer might have knowingly carried out a misaligned
-access, for example, to test an alignment trap handler or because the
-system on which this code was to run was known to gracefully handle
-misaligned accesses.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 std::atomic<int> y;
- 2 template <typename T>
- 3 void t1() {
- 4   long long r1 = x.load(std::memory_order_relaxed);
- 5   if (r1 % 4 == 0)
- 6     y.store(1, std::memory_order_relaxed);
- 7   std::observable();
- 8   *(T*)r1 = *(T*)r1;
- 9 }
-\end{verbatim}
-\caption{Alignment-Based Undefined Behavior Fix}
-\label{lst:Alignment-Based Undefined Behavior Fix}
-\end{listing}
-
-However, in that case, the developer is stepping outside the bounds
-of the standard.
-It would be good to have some way for the developer to get the job
-done without confusing the compiler, thus separating OOTA-cycle and
-UB concerns.
-One approach is to use the \co{std::observable()} function
-that has been proposed to the C++ standards
-committee~\cite{DavisHerring2021P1494R2}.
-This function would blocks backwards-propagating UB.
-Then this example could be rewritten as shown in
-Listing~\ref{lst:Alignment-Based Undefined Behavior Fix},
-with the asm on line~7 preventing the compiler from assuming that line~8
-is guaranteed to be executed, thus preventing the compiler from
-applying any lessons that it might learn from line~8 to line~5.
-
-% Clean up <200b> in vim: ":%s/\%u200b//g"
-
-There are a number of ways to implement \co{std::observable()},
-for example, a new \co{"mightnotreturn"} GCC asm clobber, a call to an empty
-function that is invisible to the compiler, and a GCC goto asm that the
-compiler cannot prove falls through.
-
-\subsection{Unreachability-Based Undefined Behavior}
-\label{sec:Unreachability-Based Undefined Behavior}
-
-This section attempts to apply the lessons
-of Appendix~\ref{sec:Unreachability-Based Undefined Behavior} to
-Goldblatt's unreachability-UB example.
-
-The \co{std::observable()} at line~12 of
-Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
-avoids confusing the compiler, instead forcing it to assume that the
-value of \co{local} might not be zero.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 // UB if this function is ever called.
- 2 [[noreturn]]
- 3 inline void unreachable() {
- 4   return;
- 5 }
- 6 void t1() {
- 7   r1 = x.load(std::memory_order_relaxed);
- 8   int local = (r1 & 1);
- 9   if (local == 0)
-10     y.store(1, std::memory_order_relaxed);
-11   if (local != 0) {
-12     std::observable();
-13     // If we get here, UB.
-14     unreachable();
-15   }
-16 }
-\end{verbatim}
-\caption{Unreachability-Based Undefined Behavior Fix}
-\label{lst:Unreachability-Based Undefined Behavior Fix}
-\end{listing}
-
-By the same reasoning, Goldblatt's \co{misbehave} example that
-also involves unreachability can also be addressed through insertion
-of the proposed \co{std::observable()}.
-
-\subsection{Atomic Stores From Undefined Behavior}
-\label{sec:Atomic Stores From Undefined Behavior}
-
-Another proposal to the C++ standards
-committee~\cite{HansBoehm2020UBalternatives}
-proposes (among other alternatives) that any atomic store resulting
-from undefined behavior have at least release semantics.
-This would prevent such a store from propagating undefined behavior
-backwards within that store's thread.
-In some cases, this could also prevent this undefined behavior from
-propagating back through other threads.
-
-However, it might well force C++ implementations to promote a surprisingly
-large fraction of a program's relaxed stores to release.
-
-\subsection{Lifetime-Based Undefined Behavior}
-\label{sec:Lifetime-Based Undefined Behavior}
-
-Not all undefined behavior can be backstopped using
-\co{std::unreachable()}.
-
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 int *foo(int *r)
- 2 {
- 3   b.store(r, memory_order_relaxed);
- 4   return &global;
- 5 }
- 6
- 7 int go(int *p)
- 8 {
- 9   int r;
-10   int *q = foo(&r);
-11
-12   r = 2;
-13   *p = 1;
-14   *q = r;
-15   return *q - r;
-16 }
-17
-18 void t1(void)
-19 {
-20   int r1 = go(a.load(memory_order_relaxed));
-21   printf("go() = %d\n", r1);
-22 }
-23
-24 void t2(void)
-25 {
-26   int *r2 = b.load(memory_order_relaxed);
-27   a.store(r2, memory_order_relaxed);
-28 }
-\end{verbatim}
-\caption{Lifetime-Based Undefined Behavior}
-\label{lst:Lifetime-Based Undefined Behavior}
-\end{listing}
-
-Listing~\ref{lst:Lifetime-Based Undefined Behavior}
-shows an example loosely based on one presented by Stephen Dolan\footnote{
-	\url{https://stedolan.net/talks/2024/loadstore}}
-in which function \co{go()} invokes \co{foo()} with a pointer to \co{go()}'s
-local variable \co{r}.
-However, \co{foo()} uses a relaxed store to place this pointer in
-an atomic variable \co{b}.
-Because this store is relaxed, it is not ordered against other code
-in \co{go()} and \co{t1()}, which means that \co{t2()} might well
-read this pointer from \co{b} either before the beginning or after
-the end of \co{r}'s lifetime, neither of which is well-defined.
-
-Of course, any attempt to dereference the pointer read from either
-\co{a} or \co{b} could result in undefined behavior.
-
-The lesson is clear:
-Properly synchronize reads and writes of pointers with the lifetime
-of the pointed-to object.
-
-This example could make lines~3 and~27 be \co{memory_order_release} and
-lines~20 and~26 be \co{memory_order_acquire} in order to avoid carrying
-a pointer to \co{r} before the lifetime of \co{r} has started.
-Alternatively, threads could be spawned only after \co{r}'s lifetime
-starts and joined before it ends.
-Placing \co{r} in static global storage would avoid lifetime considerations
-totally.
-And there are of course any number of other synchronization schemes
-that might be used to ensure that pointers are in use only while the
-pointed-to objects are alive.
-
-Note that \co{std::unreachable()} does not suffice because the CPU
-can also reorder relaxed stores.
-Note also that load-store ordering is not sufficient to prevent a
-pointer to \co{r} from being stored to \co{b} prior to the start of
-\co{r}'s lifetime because there is no atomic load between the start
-of that lifetime and the relaxed store to \co{b}.
-
-There are a number of ways to prevent carrying a pointer to \co{r} after
-the lifetime of \co{r} has ended, and these are left as an exercise for
-the reader.
-
 % \section{History}

### 2024-02-13 ootavacuous.tex
Remove Section 9 ("How to Form an OOTA Cycle?")
This commit also removes a couple of stray references.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-For more discussion of potential optimizations, please see
-Section~\ref{sec:But What About Non-Volatile Atomics?}.
-% @@@ Pull that section in here?  Drop this forward reference?
-
 \subsection{C++ Implementations}
@@
 
-\section{How to Form an OOTA Cycle?}
-\label{sec:How to Form an OOTA Cycle?}
-
-Rather than look at how to prevent an implementation from forming an
-OOTA cycle, this section instead looks at how to form one.
-This will lead to the conclusion that a correct real-world C
-implementation cannot form OOTA cycles,\footnote{
-	Keep in mind that C does not have non-volatile atomic
-	operations.}
-and provide guidance on how to avoid OOTA cycles for non-volatile relaxed
-atomics in C++ implementations.
-
-Again, an OOTA cycle consists of loads from other threads' stores where
-a given store's address or value depends on the value returned by
-the corresponding thread's prior load.
-All of the links from one thread to the next are reads-from links, which
-are temporal in nature, that is, a given store must execute earlier in
-global time than any load from that stored-to object that returns the
-value stored.
-
-But it is also the case that an OOTA cycle must have at least one link
-that goes backwards in time.
-Given that the inter-thread links are temporal with a give store preceding
-any load returning the value stored, the only possible candidate atemporal
-links are those confined to a given thread.
-This in turn requires that a store that depends on an earlier load
-be executed earlier in time than that load.
-The two ways that this can happen are when the implementation:
-
-\begin{enumerate}
-\item	Is able to prove the location and value of the store.
-\item	Guesses the location and value of the store and has some means
-	to take corrective action should any guess prove incorrect.
-\end{enumerate}
-
-Each of these possibilities is discussed in the next sections.
-followed by a section on the possibility of flattening a multithreaded
-program into a single thread.
-
-\subsection{Implementation Proves Value}
-\label{sec:Implementation Proves Value}
-
-Implementation can sometimes prove the location and value of a given
-store.
-For example, if the value loaded is multiplied by zero, the result
-is known to be zero, so that the store can proceed prior to the load.
-But in this case, there is no way that the value returned from
-the load that can affect the value stored.
-Therefore, that store cannot possibly take part in an OOTA cycle.
-
-\subsection{Implementation Guesses at Value}
-\label{sec:Implementation Guesses at Value}
-
-Hardware speculative execution is commonplace, and permits the
-hardware to guess at values, squashing the speculation if any
-of the guesses prove incorrect.
-But in this case, a given store is not committed until all guesses that
-this store depends on have been confirmed, and thus no uncommitted store
-is visible to any other thread.
-Therefore, any hardware speculation that results in OOTA cycles is doomed
-to be squashed an discarded, and by design, even on weakly ordered
-systems~\cite{ARMv7A:2010,ARMv8A:2017,PowerISA2.07-2013}.
-This same constraint applies to software, which might use compare-and-swap
-loops or hardware transactional memory to confirm guesses.
-
-Either way, the need to wait for guesses to be confirmed before committing
-stores prevents those stores from becoming visible prior to the
-completion of any loads that those stores depend on.
-
-\subsection{But What About Flattening?}
-\label{sec:But What About Flattening?}
-
-The concept of flattening multiple threads into a single thread is
-tantalizing, especially given the large cache-miss penalties inherent
-to large multicore systems, which can have many sockets containing
-thousands of CPUs.
-One might object to the whole concept of flattening as an
-optimization given the difficulty inherent in avoiding deadlocks
-and other hazards.
-Nevertheless, it is only reasonable to ask whether flattening can
-result in OOTA cycles.
-As we will see, the answer is ``no''.
-
-The reason for this answer is that any fully flattened program will be
-single threaded, which implies that each of its atomic load operations
-will return the last value stored to that same object in sequenced-before
-order (or the initial value if there is no such store).
-Thus, any sequence of loads and stores must have a first operation and
-a last operation, and the last operation cannot affect the first
-operation.
-Therefore, cycles cannot possibly form in a flattened program,
-which in turn implies that OOTA cycles cannot form.
-
-\subsection{But What About Non-Volatile Atomics?}
-\label{sec:But What About Non-Volatile Atomics?}
-
-Volatile atomic accesses have some very attractive properties from
-the viewpoint of analyzing executions for OOTA behavior because C++
-compilers are prohibited from reordering, omitting, fusing, or inventing
-volatile accesses.
-Thus, if the compiler emits code that faithfully reflects the behaviors
-specified by the source code, the hardware memory model will prohibit
-OOTA cycles.
-
-However, there is some question as to what exactly a C++ implementation
-is permitted to do with non-volatile atomic accesses~\cite{JFBastien2015N4455}.
-For example,
-Appendix~\ref{app:Inventing Atomic Loads}
-shows a clever litmus test that results in OOTA-like behavior
-if a C++ implementation duplicates an atomic load, compares the two values
-loaded, and then, if the two loads return different values, chooses the
-one that eliminates a semantic dependency that would otherwise remain
-in place.
-Perhaps in the fullness of time, such examples will be considered to be
-simple optimizations rather than full-on OOTA cycles, but in the meantime
-C++ implementations might wish to prohibit such behavior.
-
-Keeping in mind that some early C++ implementations treated non-volatile
-atomic operations as if they were volatile, one way to accomplish this
-while still permitting optimization of code containing non-volatile
-atomics is to model non-volatile behavior on that of volatile atomics,
-but with some specific weakenings, for example:
-
-\begin{itemize}
-\item	Loads that do not affect observable behavior can be omitted,
-	along with any computations based on their return values.
-\item	Given a pair of loads from the same object that can be
-	reordered to be adjacent to each other, if one of the pair is a
-	relaxed load, it can be omitted, with uses of the omitted load
-	instead using the value returned by the surviving load.
-	This is termed \emph{load fusing}.
-\item	Given a pair of stores to the same object that can be reordered
-	to be adjacent to each other, if the first of the pair is a
-	relaxed store, it can be omitted.
-	This is termed \emph{store fusing}.
-\item	Suppose the source code contains a pair of non-volatile atomic
-	loads from objects that are adjacent in memory.
-	Suppose further that these loads can be reordered to be adjacent
-	to each other, and that the last of the two is a relaxed load.
-	Then this pair of loads can be replaced with a single load
-	from both objects using the first load's memory order.
-	This is termed \emph{load consolidation}.
-\item	Suppose the source code contains a pair of non-volatile atomic
-	stores to objects that are adjacent in memory.
-	Suppose further that these stores can be reordered to be adjacent
-	to each other, and that the first of the two is a relaxed load.
-	Then this pair of loads can be replaced with a single store
-	to both objects using the last store's memory order.
-	This is termed \emph{store consolidation}.
-\item	If an expression can be proven to always result in a constant,
-	that constant can be substituted for that expression.
-	Note well that combining this with specialization optimizations
-	can convert address and data dependencies to weaker control
-	dependencies.
-	Address and data dependencies order later loads and stores,
-	and control dependencies order only later stores.\footnote{
-		Yes, yes, the C++ standard does not mention address or
-		data dependencies, and mandates control dependencies
-		only indirectly (via the prohibition against introducing
-		data races).
-		Such dependencies nevertheless still exist.}
-\end{itemize}
-
-Note that load and store fusing can only be applied a finite number of
-times.
-In practice, this means that fusing can be used to unroll loops, but
-not to hoist atomic loads or stores out of loops than cannot be proven
-to be always finite.
-
-Note also that many users would consider it to be an extremely unfriendly
-act for a C++ implementation to combine loads or stores to adjacent
-objects where each object \co{is_lock_free()}, but the combination is not.
-
-These optimizations are examples, not necessarily a full set.
-Additional optimizations might be added over time, but great care will
-of course be required.
-
-\subsection{But What About Multi-Threaded Optimizations?}
-\label{sec:But What About Multi-Threaded Optimizations?}
-
-Although this paper mostly ignores multi-threaded implementations,
-JMM Causality Test Cases 1, 8, 9, and 9a discuss inter-thread
-analysis or constraints on thread scheduling.
-This topic is therefore worth a short discussion.
-
-In order to set context, it is important to note that an oracular C++
-implementation could in theory read a program and its input, and then
-perform a sequence of observable behaviors consistent with that program
-running on that input without actually executing that program's code.
-This admittedly extreme example shows that it is not unreasonable to
-impose some real-world constraints on a real C++ implementation, and
-limiting analysis to a single thread is one of the constraints that we
-have chosen.
-
-At the same time, it is worth calling out some optimizations based on
-multi-threaded analysis that can be introduced without risk of
-introducing OOTA cycles.
-
-First, given a variable that the implementation can prove is always equal
-to some constant, the implementation may act as if the program's uses of
-that variable's value instead used that constant.
-
-Second, given a pair of variables that the implementation can prove always
-have a given algebraic relation to each other, the implementation can
-act as if that program's uses of one of the variable's values instead
-used the corresponding function of the other variable's value.
-However, if the values of those variables are subject to change, there
-must be some form of synchronization in effect to prevent accesses
-to those variables' values during times when a given modification has
-changed the value of one of the variables but not yet that of the other.
-If there is no such synchronization, then this optimization cannot
-be applied.
-
-Third, given that the implementation can prove that one thread always
-terminates before another thread starts, and if one or another of the
-threads does nothing related to thread identity, the implementation
-may act as if these two threads had been fused into a single thread.
-
-Fourth and finally, if the implementation can prove that a given thread
-never executes, the implementation may act as if code specific to that
-thread had been omitted from the program.
-
-Although these last two examples might appear to be a bit utopian,
-the first two provide ample room for multi-threaded analysis.
-
-\subsection{Correct Implementations Cannot Form OOTA Cycles}
-\label{sec:Correct Implementations Cannot Form OOTA Cycles}
-
-This section has shown that an implementation that correctly handles
-the memory model, observable behavior, and the as-if rule cannot form
-OOTA cycles from volatile atomics when running UB-free programs on real
-systems in this universe.
-It has also presented a substantial list of optimizations that can be
-applied to non-volatile atomics without the risk of creating OOTA cycles.
-
 \section{Conclusion}
@@
 	are futile.
-	(But see Section~\ref{sec:But What About Non-Volatile Atomics?}.)
 \item	Implementations should not invent uses for values from

### 2024-02-13 ootavacuous.tex
Remove Section 8 ("@@@ This marks the end of the reworked material")
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 potentially all the way down to zero.
-
-\section{@@@ This marks the end of the reworked material}
-
-\subsection{But what if you want source-code dependencies?}
-
-But what if you really need a definition of semantic dependency
-for a given code fragment that covers all executions?
-One approach is to take the union of semantic dependencies across
-all executions, so that if any execution has a semantic dependency,
-the source code fragment has a semantic dependency.
-This approach might be useful when carrying out static analysis.
-
-On the other hand, many C++ implementations carry out specialization
-optimizations, which consider groups of executions.
-For example, an implementation might special-case a \co{NULL} pointer when
-considering several inline functions that take early exits in that case.
-For another example, an implementation might produce separate code for
-cases where an integer is negative, zero, and positive.
-Tools working with such implementations might therefore choose to group
-executions by specialization, and then separately analyze semantic
-dependencies for code corresponding to each specialize execution.
-
-Other tools might need to take the intersection of semantic dependencies
-across all executions, so that the source code fragment has a semantic
-dependency only if all executions have a semantic dependency.
-This approach might be useful when analyzing ordering properties
-of that code fragment.
-
-Nevertheless, in general, sdep must be defined on a per-execution basis.
-
-\subsection{Where Do Loaded Values Come From?}
-\label{sec:Where Do Loaded Values Come From?}
-
-Relying on any rf link, whether the intra-thread rfi links or the
-inter-thread rfe links, raises the question of what restrictions
-might apply to C++ implementations' choice of value loaded.
-For non-volatile atomic loads, this is question is answered by
-the C++ standard in 6.9.2p14
-\co{[intro.races]}~\cite{ThomasKoeppe2023N4950}:
-
-\begin{quote}
-	The value of an atomic object M, as determined by evaluation B,
-	shall be the value stored by some side effect A that modifies M,
-	where B does not happen before A.
-\end{quote}
-
-This wording presumably includes the initial value of M as one of the
-possible side effects.
-This means that an implementation might determine that a given atomic
-object always has a given value, and such an implementation might
-skip the load in favor of that value.\footnote{
-	There is some work starting to forbid the implementation
-	from using this assumption, for example, by requiring that
-	the implementation assume that there is at least one thread
-	that it does not know about.
-	\url{https://hackmd.io/@gonzalob/Bk_ufPZ6o/edit}}
-
-Please note that this does not apply to volatile atomic objects,
-for which the implementation is prohibited from assuming that the
-value loaded has any relationship whatsoever to any initialization
-or any values stored.
-This prohibition may come as a surprise to those carefully reading the
-C++ standard~\cite{ThomasKoeppe2023N4950}
-and finding only this wording constraining volatile accessess in
-4.1.2p6
-\co{[intro.abstract]}:
-
-\begin{quote}
-	Accesses through volatile glvalues are evaluated strictly
-	according to the rules of the abstract machine.
-\end{quote}
-
-Unfortunately, volatile semantics are mostly folklore passed by word
-of mouth among some compiler writers and some device-driver developers.
-The key point is that volatile accesses must interact with MMIO
-registers in such a way as to permit device drivers to be written.
-And until such time as the semantics of all MMIO registers of all
-devices have been formalized, published, and incorporated into
-C++ implementations,\footnote{
-	Given that some CPU vendors have started publishing formal
-	memory models, this happy day might actually arrive.
-	But we strongly recommend against holding your breath while
-	waiting for it.}
-these implementations cannot asssume that they can predict the value
-returned by a volatile load.
-
-\subsection{On OOTA Definitions}
-\label{sec:On OOTA Definitions}
-
-The earliest work defines ``OOTA cycle'' by example, without a precise
-definition.
-Sometimes ``causal cycle'' is used as if it was a definition, but
-without a clear decision thread for what does and does not
-constitute a causal cycle.
-P0442R0 defines an OOTA cycle as a fixed-point computation that is
-destroyed by perturbations, which makes perfect sense to that paper's
-authors, but has left others unsatisfied.
-Still others argue (perhaps correctly) that it is impossible to provide
-a general definition of ``OOTA cycle'', but this simply raises
-the question of whether a \emph{useful} definition can be formulated.
-
-@@@ The C++ standard says this in 33.5.4p8
-(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
-
-\begin{quote}
-	Implementations should ensure that no “out-of-thin-air” values
-	are computed that circularly depend on their own computation.
-\end{quote}
-
-The following sections discuss properties of and constraints on
-useful definitions of ``semantic dependency''.
-
-\subsubsection{@@@ Semantic Dependencies Are Per Execution}
-\label{sec:Semantic Dependencies Are Per Execution}
-
-\subsubsection{@@@ Semantic Dependencies Depend on Context}
-\label{sec:Semantic Dependencies Depend on Context}
-
-\subsubsection{@@@ Semantic Dependencies Can Be Nondeterministic}
-\label{sec:Semantic Dependencies Can Be Nondeterministic}
-
-\subsubsection{Semantic Dependencies And Optimization}
-\label{sec:Semantic Dependencies And Optimization}
-
-@@@ What do we do with this one???  Not sure we need it.
-
-Consider again the following:
-
-\begin{quote}
-\begin{verbatim}
-x = y + z;
-\end{verbatim}
-\end{quote}
-
-Suppose that the implementation can prove that at any time that
-the above statement might execute, \co{y} is equal to \co{z}.
-That would allow the implementation to optimize, that is, to act as if
-the source code was instead as follows:
-
-\begin{quote}
-\begin{verbatim}
-x = 2 * y;
-\end{verbatim}
-\end{quote}
-
-Or equally valid:
-
-\begin{quote}
-\begin{verbatim}
-x = 2 * z;
-\end{verbatim}
-\end{quote}
-
-In this case, is the semantic dependency from \co{y} to \co{x},
-from \co{z} to \co{x}, or both \co{y} and \co{z} to \co{x}?
-The answer is that this is a free choice on the part of the
-implementation.
-Those wishing to change the value of \co{y} or \co{z} externally
-(for example, using a debugger) would do well to avail themselves
-of the \co{volatile} keyword.
-
-\emph{Semantic dependencies can be a free choice on the part of the
-C++ implementation.}
-
-\subsubsection{Semantic Dependencies and I/O}
-\label{sec:Semantic Dependencies and I/O}
-
-@@@ Not sure we need this one.
-
-A computation might include I/O:
-
-\begin{quote}
-\begin{verbatim}
-x = y * input_int();
-\end{verbatim}
-\end{quote}
-
-Then there is a semantic dependency from \co{y} to \co{x} only if
-the number input is non-zero.
-Similar issues arise when reading timestamps.
-
-In addition, I/O operations are observed behavior, which greatly restricts
-optimizations involving them, in turn reducing the need to evaluate
-potential semantic dependencies.
-
-\subsubsection{@@@ Semantic Dependencies Can Be Complex}
-\label{sec:Semantic Dependencies Can Be Complex}
-
-\subsubsection{@@@ Implementers and Users Influence Semantic-Dependencies Definition}
-\label{sec:Implementers and Users Influence Semantic-Dependencies Definition}
-
-\subsubsection{Complex Semantic Dependencies Analysis is a Choice}
-\label{sec:Complex Semantic Dependencies Analysis is a Choice}
-
-All of these tools can incur high overhead, especially manual code
-inspection.
-However, as we will see, the actual C++ implementations themselves need
-not use these tools.
-Their current analysis is guaranteed to be sufficient, courtesy of the
-real-world constraints discussed in the next section, at least assuming
-that their analysis is in fact correct.
-Furthermore, as we will also see, in some special but commonly occurring
-circumstances, there are much simpler methods that incur negligible overhead.
 

### 2024-02-13 cuttingroom.tex
Add a file to hold material cut from paper
8 @@@ This marks the end of the reworked material
9 How to Form an OOTA Cycle?
E Aside on Undefined Behavior
F Evaluating sdep Using cbmc
G Debates on Causality
H Cutting-Room Floor

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\section{@@@ This marks the end of the reworked material}

\subsection{But what if you want source-code dependencies?}

But what if you really need a definition of semantic dependency
for a given code fragment that covers all executions?
One approach is to take the union of semantic dependencies across
all executions, so that if any execution has a semantic dependency,
the source code fragment has a semantic dependency.
This approach might be useful when carrying out static analysis.

On the other hand, many C++ implementations carry out specialization
optimizations, which consider groups of executions.
For example, an implementation might special-case a \co{NULL} pointer when
considering several inline functions that take early exits in that case.
For another example, an implementation might produce separate code for
cases where an integer is negative, zero, and positive.
Tools working with such implementations might therefore choose to group
executions by specialization, and then separately analyze semantic
dependencies for code corresponding to each specialize execution.

Other tools might need to take the intersection of semantic dependencies
across all executions, so that the source code fragment has a semantic
dependency only if all executions have a semantic dependency.
This approach might be useful when analyzing ordering properties
of that code fragment.

Nevertheless, in general, sdep must be defined on a per-execution basis.

\subsection{Where Do Loaded Values Come From?}
\label{sec:Where Do Loaded Values Come From?}

Relying on any rf link, whether the intra-thread rfi links or the
inter-thread rfe links, raises the question of what restrictions
might apply to C++ implementations' choice of value loaded.
For non-volatile atomic loads, this is question is answered by
the C++ standard in 6.9.2p14
\co{[intro.races]}~\cite{ThomasKoeppe2023N4950}:

\begin{quote}
	The value of an atomic object M, as determined by evaluation B,
	shall be the value stored by some side effect A that modifies M,
	where B does not happen before A.
\end{quote}

This wording presumably includes the initial value of M as one of the
possible side effects.
This means that an implementation might determine that a given atomic
object always has a given value, and such an implementation might
skip the load in favor of that value.\footnote{
	There is some work starting to forbid the implementation
	from using this assumption, for example, by requiring that
	the implementation assume that there is at least one thread
	that it does not know about.
	\url{https://hackmd.io/@gonzalob/Bk_ufPZ6o/edit}}

Please note that this does not apply to volatile atomic objects,
for which the implementation is prohibited from assuming that the
value loaded has any relationship whatsoever to any initialization
or any values stored.
This prohibition may come as a surprise to those carefully reading the
C++ standard~\cite{ThomasKoeppe2023N4950}
and finding only this wording constraining volatile accessess in
4.1.2p6
\co{[intro.abstract]}:

\begin{quote}
	Accesses through volatile glvalues are evaluated strictly
	according to the rules of the abstract machine.
\end{quote}

Unfortunately, volatile semantics are mostly folklore passed by word
of mouth among some compiler writers and some device-driver developers.
The key point is that volatile accesses must interact with MMIO
registers in such a way as to permit device drivers to be written.
And until such time as the semantics of all MMIO registers of all
devices have been formalized, published, and incorporated into
C++ implementations,\footnote{
	Given that some CPU vendors have started publishing formal
	memory models, this happy day might actually arrive.
	But we strongly recommend against holding your breath while
	waiting for it.}
these implementations cannot asssume that they can predict the value
returned by a volatile load.

\subsection{On OOTA Definitions}
\label{sec:On OOTA Definitions}

The earliest work defines ``OOTA cycle'' by example, without a precise
definition.
Sometimes ``causal cycle'' is used as if it was a definition, but
without a clear decision thread for what does and does not
constitute a causal cycle.
P0442R0 defines an OOTA cycle as a fixed-point computation that is
destroyed by perturbations, which makes perfect sense to that paper's
authors, but has left others unsatisfied.
Still others argue (perhaps correctly) that it is impossible to provide
a general definition of ``OOTA cycle'', but this simply raises
the question of whether a \emph{useful} definition can be formulated.

@@@ The C++ standard says this in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:

\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}

The following sections discuss properties of and constraints on
useful definitions of ``semantic dependency''.

\subsubsection{@@@ Semantic Dependencies Are Per Execution}
\label{sec:Semantic Dependencies Are Per Execution}

\subsubsection{@@@ Semantic Dependencies Depend on Context}
\label{sec:Semantic Dependencies Depend on Context}

\subsubsection{@@@ Semantic Dependencies Can Be Nondeterministic}
\label{sec:Semantic Dependencies Can Be Nondeterministic}

\subsubsection{Semantic Dependencies And Optimization}
\label{sec:Semantic Dependencies And Optimization}

@@@ What do we do with this one???  Not sure we need it.

Consider again the following:

\begin{quote}
\begin{verbatim}
x = y + z;
\end{verbatim}
\end{quote}

Suppose that the implementation can prove that at any time that
the above statement might execute, \co{y} is equal to \co{z}.
That would allow the implementation to optimize, that is, to act as if
the source code was instead as follows:

\begin{quote}
\begin{verbatim}
x = 2 * y;
\end{verbatim}
\end{quote}

Or equally valid:

\begin{quote}
\begin{verbatim}
x = 2 * z;
\end{verbatim}
\end{quote}

In this case, is the semantic dependency from \co{y} to \co{x},
from \co{z} to \co{x}, or both \co{y} and \co{z} to \co{x}?
The answer is that this is a free choice on the part of the
implementation.
Those wishing to change the value of \co{y} or \co{z} externally
(for example, using a debugger) would do well to avail themselves
of the \co{volatile} keyword.

\emph{Semantic dependencies can be a free choice on the part of the
C++ implementation.}

\subsubsection{Semantic Dependencies and I/O}
\label{sec:Semantic Dependencies and I/O}

@@@ Not sure we need this one.

A computation might include I/O:

\begin{quote}
\begin{verbatim}
x = y * input_int();
\end{verbatim}
\end{quote}

Then there is a semantic dependency from \co{y} to \co{x} only if
the number input is non-zero.
Similar issues arise when reading timestamps.

In addition, I/O operations are observed behavior, which greatly restricts
optimizations involving them, in turn reducing the need to evaluate
potential semantic dependencies.

\subsubsection{@@@ Semantic Dependencies Can Be Complex}
\label{sec:Semantic Dependencies Can Be Complex}

\subsubsection{@@@ Implementers and Users Influence Semantic-Dependencies Definition}
\label{sec:Implementers and Users Influence Semantic-Dependencies Definition}

\subsubsection{Complex Semantic Dependencies Analysis is a Choice}
\label{sec:Complex Semantic Dependencies Analysis is a Choice}

All of these tools can incur high overhead, especially manual code
inspection.
However, as we will see, the actual C++ implementations themselves need
not use these tools.
Their current analysis is guaranteed to be sufficient, courtesy of the
real-world constraints discussed in the next section, at least assuming
that their analysis is in fact correct.
Furthermore, as we will also see, in some special but commonly occurring
circumstances, there are much simpler methods that incur negligible overhead.

\section{How to Form an OOTA Cycle?}
\label{sec:How to Form an OOTA Cycle?}

Rather than look at how to prevent an implementation from forming an
OOTA cycle, this section instead looks at how to form one.
This will lead to the conclusion that a correct real-world C
implementation cannot form OOTA cycles,\footnote{
	Keep in mind that C does not have non-volatile atomic
	operations.}
and provide guidance on how to avoid OOTA cycles for non-volatile relaxed
atomics in C++ implementations.

Again, an OOTA cycle consists of loads from other threads' stores where
a given store's address or value depends on the value returned by
the corresponding thread's prior load.
All of the links from one thread to the next are reads-from links, which
are temporal in nature, that is, a given store must execute earlier in
global time than any load from that stored-to object that returns the
value stored.

But it is also the case that an OOTA cycle must have at least one link
that goes backwards in time.
Given that the inter-thread links are temporal with a give store preceding
any load returning the value stored, the only possible candidate atemporal
links are those confined to a given thread.
This in turn requires that a store that depends on an earlier load
be executed earlier in time than that load.
The two ways that this can happen are when the implementation:

\begin{enumerate}
\item	Is able to prove the location and value of the store.
\item	Guesses the location and value of the store and has some means
	to take corrective action should any guess prove incorrect.
\end{enumerate}

Each of these possibilities is discussed in the next sections.
followed by a section on the possibility of flattening a multithreaded
program into a single thread.

\subsection{Implementation Proves Value}
\label{sec:Implementation Proves Value}

Implementation can sometimes prove the location and value of a given
store.
For example, if the value loaded is multiplied by zero, the result
is known to be zero, so that the store can proceed prior to the load.
But in this case, there is no way that the value returned from
the load that can affect the value stored.
Therefore, that store cannot possibly take part in an OOTA cycle.

\subsection{Implementation Guesses at Value}
\label{sec:Implementation Guesses at Value}

Hardware speculative execution is commonplace, and permits the
hardware to guess at values, squashing the speculation if any
of the guesses prove incorrect.
But in this case, a given store is not committed until all guesses that
this store depends on have been confirmed, and thus no uncommitted store
is visible to any other thread.
Therefore, any hardware speculation that results in OOTA cycles is doomed
to be squashed an discarded, and by design, even on weakly ordered
systems~\cite{ARMv7A:2010,ARMv8A:2017,PowerISA2.07-2013}.
This same constraint applies to software, which might use compare-and-swap
loops or hardware transactional memory to confirm guesses.

Either way, the need to wait for guesses to be confirmed before committing
stores prevents those stores from becoming visible prior to the
completion of any loads that those stores depend on.

\subsection{But What About Flattening?}
\label{sec:But What About Flattening?}

The concept of flattening multiple threads into a single thread is
tantalizing, especially given the large cache-miss penalties inherent
to large multicore systems, which can have many sockets containing
thousands of CPUs.
One might object to the whole concept of flattening as an
optimization given the difficulty inherent in avoiding deadlocks
and other hazards.
Nevertheless, it is only reasonable to ask whether flattening can
result in OOTA cycles.
As we will see, the answer is ``no''.

The reason for this answer is that any fully flattened program will be
single threaded, which implies that each of its atomic load operations
will return the last value stored to that same object in sequenced-before
order (or the initial value if there is no such store).
Thus, any sequence of loads and stores must have a first operation and
a last operation, and the last operation cannot affect the first
operation.
Therefore, cycles cannot possibly form in a flattened program,
which in turn implies that OOTA cycles cannot form.

\subsection{But What About Non-Volatile Atomics?}
\label{sec:But What About Non-Volatile Atomics?}

Volatile atomic accesses have some very attractive properties from
the viewpoint of analyzing executions for OOTA behavior because C++
compilers are prohibited from reordering, omitting, fusing, or inventing
volatile accesses.
Thus, if the compiler emits code that faithfully reflects the behaviors
specified by the source code, the hardware memory model will prohibit
OOTA cycles.

However, there is some question as to what exactly a C++ implementation
is permitted to do with non-volatile atomic accesses~\cite{JFBastien2015N4455}.
For example,
Appendix~\ref{app:Inventing Atomic Loads}
shows a clever litmus test that results in OOTA-like behavior
if a C++ implementation duplicates an atomic load, compares the two values
loaded, and then, if the two loads return different values, chooses the
one that eliminates a semantic dependency that would otherwise remain
in place.
Perhaps in the fullness of time, such examples will be considered to be
simple optimizations rather than full-on OOTA cycles, but in the meantime
C++ implementations might wish to prohibit such behavior.

Keeping in mind that some early C++ implementations treated non-volatile
atomic operations as if they were volatile, one way to accomplish this
while still permitting optimization of code containing non-volatile
atomics is to model non-volatile behavior on that of volatile atomics,
but with some specific weakenings, for example:

\begin{itemize}
\item	Loads that do not affect observable behavior can be omitted,
	along with any computations based on their return values.
\item	Given a pair of loads from the same object that can be
	reordered to be adjacent to each other, if one of the pair is a
	relaxed load, it can be omitted, with uses of the omitted load
	instead using the value returned by the surviving load.
	This is termed \emph{load fusing}.
\item	Given a pair of stores to the same object that can be reordered
	to be adjacent to each other, if the first of the pair is a
	relaxed store, it can be omitted.
	This is termed \emph{store fusing}.
\item	Suppose the source code contains a pair of non-volatile atomic
	loads from objects that are adjacent in memory.
	Suppose further that these loads can be reordered to be adjacent
	to each other, and that the last of the two is a relaxed load.
	Then this pair of loads can be replaced with a single load
	from both objects using the first load's memory order.
	This is termed \emph{load consolidation}.
\item	Suppose the source code contains a pair of non-volatile atomic
	stores to objects that are adjacent in memory.
	Suppose further that these stores can be reordered to be adjacent
	to each other, and that the first of the two is a relaxed load.
	Then this pair of loads can be replaced with a single store
	to both objects using the last store's memory order.
	This is termed \emph{store consolidation}.
\item	If an expression can be proven to always result in a constant,
	that constant can be substituted for that expression.
	Note well that combining this with specialization optimizations
	can convert address and data dependencies to weaker control
	dependencies.
	Address and data dependencies order later loads and stores,
	and control dependencies order only later stores.\footnote{
		Yes, yes, the C++ standard does not mention address or
		data dependencies, and mandates control dependencies
		only indirectly (via the prohibition against introducing
		data races).
		Such dependencies nevertheless still exist.}
\end{itemize}

Note that load and store fusing can only be applied a finite number of
times.
In practice, this means that fusing can be used to unroll loops, but
not to hoist atomic loads or stores out of loops than cannot be proven
to be always finite.

Note also that many users would consider it to be an extremely unfriendly
act for a C++ implementation to combine loads or stores to adjacent
objects where each object \co{is_lock_free()}, but the combination is not.

These optimizations are examples, not necessarily a full set.
Additional optimizations might be added over time, but great care will
of course be required.

\subsection{But What About Multi-Threaded Optimizations?}
\label{sec:But What About Multi-Threaded Optimizations?}

Although this paper mostly ignores multi-threaded implementations,
JMM Causality Test Cases 1, 8, 9, and 9a discuss inter-thread
analysis or constraints on thread scheduling.
This topic is therefore worth a short discussion.

In order to set context, it is important to note that an oracular C++
implementation could in theory read a program and its input, and then
perform a sequence of observable behaviors consistent with that program
running on that input without actually executing that program's code.
This admittedly extreme example shows that it is not unreasonable to
impose some real-world constraints on a real C++ implementation, and
limiting analysis to a single thread is one of the constraints that we
have chosen.

At the same time, it is worth calling out some optimizations based on
multi-threaded analysis that can be introduced without risk of
introducing OOTA cycles.

First, given a variable that the implementation can prove is always equal
to some constant, the implementation may act as if the program's uses of
that variable's value instead used that constant.

Second, given a pair of variables that the implementation can prove always
have a given algebraic relation to each other, the implementation can
act as if that program's uses of one of the variable's values instead
used the corresponding function of the other variable's value.
However, if the values of those variables are subject to change, there
must be some form of synchronization in effect to prevent accesses
to those variables' values during times when a given modification has
changed the value of one of the variables but not yet that of the other.
If there is no such synchronization, then this optimization cannot
be applied.

Third, given that the implementation can prove that one thread always
terminates before another thread starts, and if one or another of the
threads does nothing related to thread identity, the implementation
may act as if these two threads had been fused into a single thread.

Fourth and finally, if the implementation can prove that a given thread
never executes, the implementation may act as if code specific to that
thread had been omitted from the program.

Although these last two examples might appear to be a bit utopian,
the first two provide ample room for multi-threaded analysis.

\subsection{Correct Implementations Cannot Form OOTA Cycles}
\label{sec:Correct Implementations Cannot Form OOTA Cycles}

This section has shown that an implementation that correctly handles
the memory model, observable behavior, and the as-if rule cannot form
OOTA cycles from volatile atomics when running UB-free programs on real
systems in this universe.
It has also presented a substantial list of optimizations that can be
applied to non-volatile atomics without the risk of creating OOTA cycles.

\clearpage

\section{Aside on Undefined Behavior}
\label{app:Aside on Undefined Behavior}

The combination of OOTA cycles and UB has proven especially
challenging~\cite{DavidGoldblatt2019NoElegantOOTAfix,HansBoehm2020UBalternatives}.

\subsection{Alignment-Based Undefined Behavior}
\label{sec:Alignment-Based Undefined Behavior}

The example shown in
Listing~\ref{lst:Alignment-Based Undefined Behavior (P1916R0)}
(taken from P1916R0) is especially instructive.

\begin{listing}[tbp]
\begin{verbatim}
 1 std::atomic<int> y;
 2 template <typename T>
 3 void t1() {
 4   long long r1 = x.load(std::memory_order_relaxed);
 5   if (r1 % 4 == 0)
 6     y.store(1, std::memory_order_relaxed);
 7   *(T*)r1 = *(T*)r1;
 8 }
\end{verbatim}
\caption{Alignment-Based Undefined Behavior (P1916R0)}
\label{lst:Alignment-Based Undefined Behavior (P1916R0)}
\end{listing}

Because the compiler often assumes that UB cannot happen, it might
look at line~7 and conclude that the value of \co{r1} necessarily
corresponded to an address aligned per the requirements of type \co{T}.
If that type's alignment was larger than four bytes, the compiler could
then replace the condition in line~5 with \co{true}, which would mean
that there was no semantic dependency from the load of \co{x} to the
store of \co{y}.

Except that the developer might have knowingly carried out a misaligned
access, for example, to test an alignment trap handler or because the
system on which this code was to run was known to gracefully handle
misaligned accesses.

\begin{listing}[tbp]
\begin{verbatim}
 1 std::atomic<int> y;
 2 template <typename T>
 3 void t1() {
 4   long long r1 = x.load(std::memory_order_relaxed);
 5   if (r1 % 4 == 0)
 6     y.store(1, std::memory_order_relaxed);
 7   std::observable();
 8   *(T*)r1 = *(T*)r1;
 9 }
\end{verbatim}
\caption{Alignment-Based Undefined Behavior Fix}
\label{lst:Alignment-Based Undefined Behavior Fix}
\end{listing}

However, in that case, the developer is stepping outside the bounds
of the standard.
It would be good to have some way for the developer to get the job
done without confusing the compiler, thus separating OOTA-cycle and
UB concerns.
One approach is to use the \co{std::observable()} function
that has been proposed to the C++ standards
committee~\cite{DavisHerring2021P1494R2}.
This function would blocks backwards-propagating UB.
Then this example could be rewritten as shown in
Listing~\ref{lst:Alignment-Based Undefined Behavior Fix},
with the asm on line~7 preventing the compiler from assuming that line~8
is guaranteed to be executed, thus preventing the compiler from
applying any lessons that it might learn from line~8 to line~5.

% Clean up <200b> in vim: ":%s/\%u200b//g"

There are a number of ways to implement \co{std::observable()},
for example, a new \co{"mightnotreturn"} GCC asm clobber, a call to an empty
function that is invisible to the compiler, and a GCC goto asm that the
compiler cannot prove falls through.

\subsection{Unreachability-Based Undefined Behavior}
\label{sec:Unreachability-Based Undefined Behavior}

This section attempts to apply the lessons
of Appendix~\ref{sec:Unreachability-Based Undefined Behavior} to
Goldblatt's unreachability-UB example.

The \co{std::observable()} at line~12 of
Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
avoids confusing the compiler, instead forcing it to assume that the
value of \co{local} might not be zero.

\begin{listing}[tbp]
\begin{verbatim}
 1 // UB if this function is ever called.
 2 [[noreturn]]
 3 inline void unreachable() {
 4   return;
 5 }
 6 void t1() {
 7   r1 = x.load(std::memory_order_relaxed);
 8   int local = (r1 & 1);
 9   if (local == 0)
10     y.store(1, std::memory_order_relaxed);
11   if (local != 0) {
12     std::observable();
13     // If we get here, UB.
14     unreachable();
15   }
16 }
\end{verbatim}
\caption{Unreachability-Based Undefined Behavior Fix}
\label{lst:Unreachability-Based Undefined Behavior Fix}
\end{listing}

By the same reasoning, Goldblatt's \co{misbehave} example that
also involves unreachability can also be addressed through insertion
of the proposed \co{std::observable()}.

\subsection{Atomic Stores From Undefined Behavior}
\label{sec:Atomic Stores From Undefined Behavior}

Another proposal to the C++ standards
committee~\cite{HansBoehm2020UBalternatives}
proposes (among other alternatives) that any atomic store resulting
from undefined behavior have at least release semantics.
This would prevent such a store from propagating undefined behavior
backwards within that store's thread.
In some cases, this could also prevent this undefined behavior from
propagating back through other threads.

However, it might well force C++ implementations to promote a surprisingly
large fraction of a program's relaxed stores to release.

\subsection{Lifetime-Based Undefined Behavior}
\label{sec:Lifetime-Based Undefined Behavior}

Not all undefined behavior can be backstopped using
\co{std::unreachable()}.

\begin{listing}[tbp]
\begin{verbatim}
 1 int *foo(int *r)
 2 {
 3   b.store(r, memory_order_relaxed);
 4   return &global;
 5 }
 6
 7 int go(int *p)
 8 {
 9   int r;
10   int *q = foo(&r);
11
12   r = 2;
13   *p = 1;
14   *q = r;
15   return *q - r;
16 }
17
18 void t1(void)
19 {
20   int r1 = go(a.load(memory_order_relaxed));
21   printf("go() = %d\n", r1);
22 }
23
24 void t2(void)
25 {
26   int *r2 = b.load(memory_order_relaxed);
27   a.store(r2, memory_order_relaxed);
28 }
\end{verbatim}
\caption{Lifetime-Based Undefined Behavior}
\label{lst:Lifetime-Based Undefined Behavior}
\end{listing}

Listing~\ref{lst:Lifetime-Based Undefined Behavior}
shows an example loosely based on one presented by Stephen Dolan\footnote{
	\url{https://stedolan.net/talks/2024/loadstore}}
in which function \co{go()} invokes \co{foo()} with a pointer to \co{go()}'s
local variable \co{r}.
However, \co{foo()} uses a relaxed store to place this pointer in
an atomic variable \co{b}.
Because this store is relaxed, it is not ordered against other code
in \co{go()} and \co{t1()}, which means that \co{t2()} might well
read this pointer from \co{b} either before the beginning or after
the end of \co{r}'s lifetime, neither of which is well-defined.

Of course, any attempt to dereference the pointer read from either
\co{a} or \co{b} could result in undefined behavior.

The lesson is clear:
Properly synchronize reads and writes of pointers with the lifetime
of the pointed-to object.

This example could make lines~3 and~27 be \co{memory_order_release} and
lines~20 and~26 be \co{memory_order_acquire} in order to avoid carrying
a pointer to \co{r} before the lifetime of \co{r} has started.
Alternatively, threads could be spawned only after \co{r}'s lifetime
starts and joined before it ends.
Placing \co{r} in static global storage would avoid lifetime considerations
totally.
And there are of course any number of other synchronization schemes
that might be used to ensure that pointers are in use only while the
pointed-to objects are alive.

Note that \co{std::unreachable()} does not suffice because the CPU
can also reorder relaxed stores.
Note also that load-store ordering is not sufficient to prevent a
pointer to \co{r} from being stored to \co{b} prior to the start of
\co{r}'s lifetime because there is no atomic load between the start
of that lifetime and the relaxed store to \co{b}.

There are a number of ways to prevent carrying a pointer to \co{r} after
the lifetime of \co{r} has ended, and these are left as an exercise for
the reader.

\clearpage

\section{Evaluating sdep Using \co{cbmc}}
\label{sec:Evaluating sdep Using cbmc}

This section demonstrates the evaluation of sdep using
the C Bounded Model Checker
(\co{cbmc})~\cite{EdmundClarke2004CBMC}.
This tool takes a smallish C program as input and constructs a logic
expression representing that program.
The logic expression's variables represent bits in the program's inputs
and state, and the result of that logic expression is \co{true} if that
program has a bug, for example, if it is possible for that C program to
trigger an assertion or execute an out-of-bounds array reference.
The tool then hands this expression to a SAT solver, and if this solver
finds a satisfying set of values, the tool uses those values provide
the execution path to the bug.

But first, Appendix~\ref{app:Informal Definition of Semantic Dependency}
outlines an informal informal definition of ``semantic dependency''
that can be used in \co{cbmc} test cases.

\subsection{Informal Definition of Semantic Dependency}
\label{app:Informal Definition of Semantic Dependency}

The manual analysis uses the informal rule that a semantic dependency
exists between a set of relaxed atomic loads\footnote{
	Please see
	Appendix~\ref{app:Non-Trivial Semantic Dependencies}
	for examples showing why this is a set of loads rather than a
	single load.}
and a store on any execution where different values loaded would change
the address stored to, the value stored, or the number of times that
store would be executed, with the proviso that stores of the same value
to the same address having the same ordering properties are fungible.

Producing a precise formal statement of this rule turns out to be quite
challenging, but as noted earlier, there is considerable work along
various
lines~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA}.
For examples of the challenges, please see
Section~\ref{sec:On OOTA Definitions}.
Our informal definition meets these challenges as follows:

\begin{enumerate}
\item	Section~\ref{sec:Semantic Dependencies and Source Code}
	is responsible for the phrase ``on any execution'' as well
	as the proviso regarding fungible loads.
\item	Section~\ref{sec:Semantic Dependencies Depend on Context}'s
	challenges are met by looking at the full context within
	a given thread.
\item	Section~\ref{sec:Semantic Dependencies Can Be Nondeterministic}'s
	non-determinism challenges are avoided by avoiding litmus tests
	exhibiting non-deterministic behavior.
	Again, note that most production C++ code uses compiler-based C++
	implementations, which freeze their non-deterministic choices
	into the binary at compile time.
\item	Section~\ref{sec:Semantic Dependencies And Optimization}'s
	optimization-choice challenges are met by using the C++
	implementation's choice of semantic dependency based on the
	optimizations that the implementation actually applied.
\item	Section~\ref{sec:Semantic Dependencies and I/O}'s
	I/O challenges are met by again noting that I/O is observed
	behavior.
	And by avoiding litmus tests involving I/O, given that relatively
	few of the available tools handle I/O.
\item	Section~\ref{sec:Semantic Dependencies Can Be Complex}'s
	complexity challenges are met by looking at what the code
	actually does.
	For example, if an as-yet unproven mathematical conjecture
	holds, the code will act consistently with that conjecture,
	and vice versa.
\item	Section~\ref{sec:Implementers and Users Influence Semantic-Dependencies Definition}'s
	implementer-/user-choice challenges are met by accounting for
	those choices.
	For example, if a user has specified \co{-funsigned-char},
	then the actual executions will be consistent with that choice.
	For more examples, please see
	Appendix~\ref{app:User Influence Over Language Semantics}.
\end{enumerate}

In short, we are treating ``semantic dependency'' in a manner similar
to our treatment of the laws of physics.
We let the actual executions provide an empirical basis for the analysis.

We do not expect this approach to please everyone.
However, those who are displeased might do well to reflect on the fact
that the C++ standard is not mathematical in nature, which implies that
any mathematical model of C++ is inherently an approximation.
We look forward to the day when the standard gains more mathematical
rigor, and perhaps the C++ memory model is a first step in that
direction.\footnote{
	But if so, the OOTA situation should be a caution!}
Nevertheless, until that day, we must live with the standard in its
current decidedly non-mathematical form.

\subsection{Complex Semantic Dependencies}
\label{app:Complex Semantic Dependencies}

Listing~\ref{lst:Example Code for Evaluation of sdep}
shows a \co{herd7} thread \co{P0()}.
Is there a semantic dependency between the load from \co{x} on
line~3 to the store to \co{y} on line~9?
Answering this question requires solving a mathematical equation,
not all of which currently have known solutions.
However, this section will use a relatively simple equation combined
with restrictions on domain and range that permit \co{cbmc} to do
the job.

\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 P0(atomic_int *x, atomic_int *y, atomic_int *limit)
 2 {
 3   r1 = atomic_load_explicit(x, memory_order_relaxed);
 4   r2 = atomic_load_explicit(limit, memory_order_relaxed);
 5   r3 = r1 * r1 + 2 * r1 + 2;
 6   r4 = r2;
 7   if (r2 >= r3)
 8     r4 = r3;
 9   atomic_store_explicit(y, r4, memory_order_relaxed);
10 }
\end{verbatim}
\caption{Example Code for Evaluation of sdep}
\label{lst:Example Code for Evaluation of sdep}
\end{listing}

\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 int main(int argc, char *argv[])
 2 {
 3   int i;
 4   int y[2] = { 0, 0 };
 5
 6   for (i = 0; i < 2; i++) {
 7     int r1;
 8     int r2;
 9     int r3;
10     int *yp = &y[i];
11
12     r1 = limited_arg(argv[2 * i + 1]);
13     r2 = const_arg(limited_arg(argv[2 * i + 2]));
14     r3 = r1 * r1 + 2 * r1 + 2;
15     *yp = r2 <= r3 ? r2 : r3;
16     printf("i = %d, r1 = %d, r2 = %d, r3 = %d\n", i, r1, r2, r3);
17   }
18   assert(y[0] == y[1]);
19 }
\end{verbatim}
\caption{Code for cbmc-Based Evaluation of sdep}
\label{lst:Code for cbmc-Based Evaluation of sdep}
\end{listing}

One way to answer this question is to create a C program
(see Listing~\ref{lst:Code for cbmc-Based Evaluation of sdep})
that runs the \co{P0()} thread's code twice with arbitrary inputs
(the loop spanning lines~6-17) and asserts that the results of
each iteration be equal (line~18).

Running this with:

\begin{quote}
\scriptsize
\begin{verbatim}
cbmc -DCBMC cbmc/sdep-quadratic.c
\end{verbatim}
\end{quote}

gives the result \co{VERIFICATION FAILED}.
This indicates that there are different values of \co{x} and \co{limit}
that yield different values of \co{y}, that is, that there can be
a semantic dependency.

\begin{listing}[tbp]
@@ DisplayLitmus cbmc/sdep-quadratic.c @@
\caption{Full Code for cbmc-Based Evaluation of sdep}
\label{lst:Full Code for cbmc-Based Evaluation of sdep}
\end{listing}

However, $x^2 + 2x + 2$ has a minimum value of 1 at $x=-1$, so
forcing \co{limit} to (say) zero should eliminate this static
dependency:\footnote{
	See Listing~\ref{lst:Full Code for cbmc-Based Evaluation of sdep}
	for the full \path{cbmc/sdep-quadratic.c} program.}

\begin{quote}
\scriptsize
\begin{verbatim}
cbmc -DCBMC -DLOAD_CONSTANT=0 cbmc/sdep-quadratic.c
\end{verbatim}
\end{quote}

But this also gives the result \co{VERIFICATION FAILED}.
One reason for this is that this program does 32-bit computer arithmetic,
and is thus subject to wraparound.
Limiting the value of \co{x} to 46,339 prevents wraparound:

\begin{quote}
\scriptsize
\begin{verbatim}
cbmc -DCBMC -DLOAD_CONSTANT=0 -DLOAD_LIMITS=46339 cbmc/sdep-quadratic.c
\end{verbatim}
\end{quote}

And this yields \co{VERIFICATION SUCCESSFUL}, indicating that there
is no semantic dependency in this case.
This is to be expected because the specified value of zero for \co{limit}
is smaller than the minimum of the quadratic expression, so that the
value stored to \co{y} is always zero.
Specifying \co{-DLOAD_CONSTANT=1} also yields \co{VERIFICATION SUCCESSFUL}
because this value is equal to that minimum, so that the value stored to
\co{y} is always 1.
However, specifying any larger value (for example, \co{-DLOAD_CONSTANT=2})
yields \co{VERIFICATION FAILED} because the value assigned to \co{y}
might be either 1 or 2, indicating that a semantic dependency exists.

This situation underscores the point that the existance of a semantic
dependency can be a property of a particular execution rather than a
universal property of the code itself.
The interested reader is encouraged to create more complex examples,
for example, demonstrating semantic dependencies from multiple loads or
exploring indeterminacy.

There are numerous other tools that can be used, up to and including
manual inspection, but \co{cbmc} is readily available and easy to
use.\footnote{
	On Debian-based distributions, via \co{sudo apt install cbmc}.}

\subsection{C++ Non-Determinism}
\label{app:C++ Non-Determinism}

The C++ standard allows non-determinism, for example, the order of
invocation of side-effects in some circumstances.
Listing~\ref{lst:Code for cbmc-Based Evaluation of Non-Deterministic sdep}
shows an example leveraging side-effects in the parameter list shown on
line~37.

\begin{listing}[tbp]
@@ DisplayLitmus cbmc/sdep-nondet.c @@
\caption{Code for cbmc-Based Evaluation of Non-Deterministic sdep}
\label{lst:Code for cbmc-Based Evaluation of Non-Deterministic sdep}
\end{listing}

Running \co{cbmc} on this as follows:

\begin{quote}
\scriptsize
\begin{verbatim}
cbmc -DCBMC cbmc/sdep-nondet.c
\end{verbatim}
\end{quote}

Gives the result \co{VERIFICATION SUCCESSFUL}, showing that \co{cbmc}
does not model this non-determinism.
Which is not surprising, given that many of the safety-critical uses
to which \co{cbmc} is put would prohibit such non-determinism in their
code bases.

\clearpage

\section{Debates on Causality}
\label{sec:Debates on Causality}

This Appendix is provided as a service to readers who want to know more
about the ongoing debates on causality, provided by the combined efforts
of Bard, Chat-GPT, and Bing Co-Pilot.
References may be found here:

{ \scriptsize
\url{https://docs.google.com/document/d/1MJ6pQ4TcWchUQvoI-qhLjdhPVYF14AytbTW3gr4-xag/edit?usp=sharing}
}

The concept of causality, the idea that events are caused by preceding
events in a predictable and consistent manner, is fundamental to many
scientific theories and our understanding of the world.
However, when it comes to the universe as a whole, particularly at the
most fundamental levels such as in quantum physics or the origins of
the universe, the idea of strict causality can become more complex.

In classical physics, causality is often described in terms of
deterministic laws, where events are determined by prior causes.
But at the quantum level, the behavior of particles can be probabilistic
and uncertain.
Quantum mechanics introduces a level of randomness and uncertainty that
challenges our classical notions of causality.

Additionally, when considering the origin of the universe (the Big Bang),
the laws of physics as we understand them break down at extremely high
densities and temperatures.
The conditions at the very beginning of the universe might not conform
to our classical understanding of causality, and some theories suggest
that time itself might have begun with the Big Bang, making it difficult
to define causality in a traditional sense within such a framework.

On the one hand, there is a strong case to be made that the universe
is causal.
Our understanding of physics is based on the principle of causality,
which states that every event has a cause.
This principle has been incredibly successful in explaining and predicting
the behavior of the universe at large scales.
Additionally, our own personal experience seems to confirm causality.
We constantly observe events being caused by other events, and we have
no reason to believe that this is an illusion.

However, there are also some challenges to the idea of a perfectly
causal universe.
One challenge comes from quantum mechanics, which suggests that some
events may be fundamentally non-causal.
For example, the collapse of a wave function in quantum mechanics seems
to be a random event, with no underlying cause.
Additionally, our understanding of the universe's origins is still
very limited.
We don't know whether the Big Bang itself had a cause, or whether it
was simply the beginning of everything.

In summary, while causality is a fundamental concept in our understanding
of the universe, its application and implications can become more complex,
especially at the quantum level and regarding the origin of the universe.
There are ongoing debates and explorations within physics and philosophy
about the nature of causality in these contexts.

\section{Cutting-Room Floor}
\label{sec:Cutting-Room Floor}

This appendix includes material that was removed but which might later
prove useful.

@@@ Because there is a cycle, on a real system, at least one of the links
must be atemporal, that is, either a load returns a value before
that value is stored, or a store's address or value is computed before
a prior load returns a value used in that computation.

@@@ Note that C++ implementations are permitted to evaluate to a more
general notion of OOTA values in cases involving unspecified or undefined
behavior (UB), such as use of uninitialized objects.
However, these situations do not involve OOTA cycles, but rather user
errors and other issues that can lead to unfriendly optimizations that
in turn result in unpredictable output.
This paper focuses instead on OOTA cycles.

### 2024-02-14 ootavacuous.tex
Change "partially volatile" to "quasi volatile"
At the risk of seeming over-eager, I'm going to change the name we use
from "partially volatile" to "quasi volatile" now.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 in a sense to be discussed in
-section~\ref{sec:Volatile and Partially Volatile Accesses} below.
+section~\ref{sec:Volatile and Quasi Volatile Accesses} below.
 
@@
 In fact, we will require that atomic accesses be treated as
-``partially volatile'', in that the compiler is allowed to omit,
+``quasi volatile'', in that the compiler is allowed to omit,
 merge, or reorder them but not invent them.
@@
 
-\subsection{Volatile and ``Partially Volatile'' Accesses}
-\label{sec:Volatile and Partially Volatile Accesses}
+\subsection{Volatile and ``Quasi Volatile'' Accesses}
+\label{sec:Volatile and Quasi Volatile Accesses}
 
@@
 
-For accesses to partially volatile objects, we remove some of the
+For accesses to quasi volatile objects, we remove some of the
 requirements and add a new one.
 We leave out the last two items; compilers are allowed to merge or
-omit accesses to partially volatile objects.
+omit accesses to quasi volatile objects.
 And the requirement for order preservation is weakened; it applies
@@
 Note that this formulation makes sense only for implementations in
-which all atomic objects are treated as partially volatile.
+which all atomic objects are treated as quasi volatile.
 \begin{quote}

### 2024-02-14 ootavacuous.tex
Change "strict C++" to "loose C++"
Since we seem to be drifting in the direction of using "loose C++" for
what Hans Boehm called "strict C++", let's put it in the paper.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 in using the term \emph{full C++} to denote the standard including the
-prohibition above, and the term \emph{strict C++}\footnote{
+prohibition above.
+{\it Contra\/} that article, we will use the term \emph{loose C++}\footnote{
 	There is some discomfort with this name, so this footnote
@@
 % @@@ Alternatives thus far: OOTA-prone C++, OOTA-enabled C++,
-% @@@ relaxed C++, OOTA C++, OOTA-agnostic C++, loose C++.
+% @@@ relaxed C++, OOTA C++, OOTA-agnostic C++.
+(rather than \emph{strict C++}, which seems too similar to
+\emph{full C++} for comfort)
 to denote a hypothetical standard that excludes this prohibition, but
@@
 Listing~\ref{lst:OOTA Cycle}
-shows a code fragment that under strict C++ has an OOTA cycle (although
+shows a code fragment that under loose C++ has an OOTA cycle (although
 the cycle is of course prohibited in full C++).
@@
 In fact, we will see that \emph{none} of the possible OOTA executions
-allowed by the strict C++ abstract machine will ever be realized
+allowed by the loose C++ abstract machine will ever be realized
 by the executables produced by many compilers,
@@
 The loads previously on lines~6 and~12 have been replaced by constants.
-Such a transformation complies with the strict C++ standard,
+Such a transformation complies with the loose C++ standard,
 even though the resulting executable file would produce an unintuitive
@@
 this is that it knows exactly what accesses will be performed by both
-threads, and therefore it knows that it will not violate the strict C++
+threads, and therefore it knows that it will not violate the loose C++
 memory model by assuming each thread's load reads from the other's
@@
 
-We will show that programs built using strict C++ compilers that treat all
+We will show that programs built using loose C++ compilers that treat all
 atomic operations as volatile can never exhibit OOTA, and the same is true

### 2024-02-14 ootavacuous.tex
Updates to the discussion of hardware dependencies
Change the discussion of hardware dependencies to avoid making the
assumption that an instruction must commit its action and all its
outputs at the same time.

Also remove the paragraph on ordering connected with rfi.  I don't
think it will be needed later.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 some of which flow to the inputs for later instructions.
+The inputs determine what an instruction will do.
 A few examples:
@@
 	(the sum to be stored in a general-purpose register and some
-	flag bits---e.g., Zero, Carry, and Overflow---to be stored in
-	a status register).
-\item	A conditional branch instruction would have only one input:
-	the flag bits it tests in the status register.
-	A computed branch instruction would have as input the register
-	containing the address to branch to.
-	For both, the output is the new address to be written to the
+	condition-code bits---e.g., Zero, Carry, and Overflow---to be
+	stored in a status register).
+\item	A conditional move instruction would have as inputs the
+	condition-code bits to test, the source register value, and the
+	target register value; the output would be the value to be stored
+	in the target register.
+\item	A conditional or computed branch instruction would have as inputs
+	the condition-code bits to test, the address of the following
+	instruction (to be used if the condition is false) and the
+	destination address (to be used if the condition is true).
+	The output is the new address to be written to the
 	instruction-pointer register.
@@
 \end{itemize}
-Of course, some of an instruction's inputs may be immediate constants
-taken from the instruction itself rather than outputs from some
-earlier instruction.
 Note: Hardware dependency analysis considers only the information
 flowing \emph{within} a CPU, not information flowing between the
-CPU and memory, which is handled separately.
+CPU and memory, which is handled separately as part of the action
+taken by an instruction.
 
@@
 instruction or the thread's initial register values (or possibly some sort of
-input instruction, but we will not consider that here).
+input instruction, but we will not consider that complication here).
 These are the only sources of truly new information in a thread.
@@
 Thanks to the law of cause and~effect, a CPU is not able to commit an
-instruction until all of its inputs' sources have committed.
-(Or at least, all that can affect the action and outputs of the
-instruction.
-For instance, although no existing architectures do this, there is nothing
+instruction or its outputs until the relevant inputs' sources have committed.
+The reason is obvious: An input is subject to change at any time until
+its source commits to its value,
+and the instruction and its outputs can't commit until the CPU has fully
+determined what it will do and what they will be.
+(On the other hand, the CPU need not wait for inputs that won't affect
+the instruction's results.
+For instance, a conditional move instruction wouldn't need to wait for
+the source register input to be determined once it knew that the
+condition was definitely false.
+And, although no existing architectures do this, there is nothing
 in principle to prevent a CPU from committing a \co{multiply}
@@
 value zero.)
-The reason is obvious: An input is subject to change at any time until
-its source commits,
-the action and outputs aren't fully determined until the inputs
-are, and the instruction can't commit until its action and outputs are fully
-determined.
-
-This means that, with a few possible exceptions such as the
-multiply-by-zero example, an instruction can't commit until all the
-instructions it depends on have committed.
-And even taking the exceptions into account, if a change in the
-outputs of instruction $I$ would lead to a change in the action or outputs of
-instruction $J$ then $I$ must commit before $J$.
+
+Of course, an instruction's inputs don't all have to be outputs from
+earlier instructions; some of them may be immediate constants present
+in the instruction itself.
+In this case there is no need to wait for those inputs because
+their values are fully determined from the start.
+
+The overall effect of these hardware dependencies is that if a change to an
+output of instruction $I$ would lead to a change in the action or outputs of
+a later instruction $J$, then the CPU must commit $I$'s output before
+committing $J$'s action and outputs.
 Thus dependencies force instructions to commit in order, even on
@@
 ordering in hardware executions.
-The simplest of these is a load reading from a store in another
+The simplest alternative is a load reading from a store in another
 thread, i.e., the rfe relation.
@@
 until the CPU has committed to whether the branch will be taken and
-if taken, where it will branch to;
+if taken, where it will branch to (that is, the output value it will
+send to the instruction-pointer register);
 until then the CPU cannot know whether the later instruction should
@@
 Therefore instructions following a conditional or computed branch
-instruction in an execution must commit after anything the branch
-depends on.
-(Again, there could theoretically be an exception for a branch that
+instruction in an execution must commit after the branch, and hence
+after the source for the branch's condition or destination input.
+(There could theoretically be an exception for a branch that
 conditionally jumps to the immediately following instruction.
-We can ignore this issue by treating such branches as no-ops.)
-
-Yet another mechanism for ordering crops up when a load reads from a
-store in the same thread (the rfi relation).
-In this situation the load cannot commit before the sources for the
-store's inputs.
-Until the source for the store's value commits, the value of the load
-isn't fully determined and so the load cannot commit.
-And in general, until the source for store's address commits, it's not
-fully determined whether the store and load will access the same address.
-This argument doesn't hold if the store and load have the \emph{same}
-source for their addresses, but another factor is also at play.
-Until the address is fully determined, there is a possibility that it
-will turn out to be invalid (a NULL pointer, for example), causing the
-store instruction to generate an addressing exception.
-If this should happen the load might not execute at all, and therefore
-it can't commit until the address's source has committed.
+We can ignore this possibility by treating such branches as no-ops.)
 
@@
 execution $H$ and $A$ has an OOTA cycle.
-This means there are stores $W_i$ in $A$, semantically depending on loads
+This means there are atomic stores $W_i$ in $A$,
+semantically depending on atomic loads
 $\{R_{i,j}\}$ where each of the loads reads from one of the $W_k$
@@
 map to in $H$, if they aren't omitted.
-Assuming that at least one of the stores is not omitted,
+Assuming that the stores are not all omitted,
 one of the $W_i'$ instructions, let's say $W_0'$, must commit first.
@@
 must be omitted as well.
-This amounts to a situation where the entire cycle has been optimized
+In effect, the entire cycle has been optimized
 out of existence by the compiler.
-Although we can are unable to prove it, we conjecture that in this
+Although we are unable to prove it, we conjecture that in this
 situation there must be another abstract execution which has the

### 2024-02-14 ootavacuous.tex
Update discussion of volatile
Change the discussion of volatile-object accesses to talk about the
difference between execution in a benign and a non-benign environment.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \end{quote}
-Following the folklore, we interpret these comments to mean that
-volatile accesses should be treated as a form of I/O,
-implying that volatile loads can return unpredictable values.
+
+Taking our cue from the folklore, we propose to recognize formally
+that programs with volatile objects can execute in two different kinds
+of environment: a benign one in which accesses to these objects work
+the same as non-volatile memory accesses, and a non-benign one in which
+accesses to volatile objects are subject to outside interference and
+act more like I/O.
+In particular, when running in a non-benign environment a program's
+volatile loads can return unpredictable values.
 They don't necessarily read from stores (in contrast to non-volatile loads,
-which must return the value of the store they read from).
-As a corollary, the rfe relation does not apply to volatile loads
-and hence the accesses in an OOTA cycle must not be volatile.
-We also understand that a compiler may not
-invent, omit, merge, or reorder volatile accesses,
-in a sense to be discussed in
+which always must return the value of the store they read from).
+This implies that volatile load-acquire's do not synchronize with
+volatile store-releases in the sense of the C++ memory model,
+so they do not contribute in the same way to the happens-before relation.
+Also, in these environments the rfe relation does not apply
+to volatile loads and hence the accesses in an OOTA cycle must be
+non-volatile.
+
+Of course, the machine-code file produced by a compiler must work
+properly in either kind of environment.
+Therefore the compiler must generally treat accesses to volatile objects
+as a form of I/O, and it may not
+invent, omit, merge, or reorder these accesses, as we will discuss in
 section~\ref{sec:Volatile and Quasi Volatile Accesses} below.
@@
 
-To express this idea in more formal terms, we will augment the
+To express this idea in more formal terms,
+and to explain what we mean by ``quasi-volatile'' object accesses,
+we augment the
 requirements for a hardware execution $H$ to realize an abstract
@@
 \item	The map connects accesses of the same type (loads to loads
-	and stores to stores).
+	and stores to stores) and to the same object.
 \item	The map connects accesses in a thread of $A$ to accesses in
@@
 	the same as the value of the access it maps to in $H$.
+\item	In benign environments the map must preserve the rf relation.
+	That is, if volatile load $R$ in $A$ reads from store $W$ then
+	then the instruction it maps to in $H$ must read from the
+	instruction that $W$ maps to.
 \item	The map is order-preserving: Two accesses in the same thread
@@
 \end{itemize}
-Most of these are direct consequences of the fact that volatile
-accesses are considered to be a form of I/O.
-
-For accesses to quasi volatile objects, we remove some of the
-requirements and add a new one.
-We leave out the last two items; compilers are allowed to merge or
-omit accesses to quasi volatile objects.
-And the requirement for order preservation is weakened; it applies
-only to pairs of accesses to the same object.
+Most of these are direct consequences of the fact that volatile-object
+accesses are considered to be a form of I/O when the program runs
+in a non-benign environment.
+But to be clear, these requirements apply to all environments.
+
+By contrast, accesses to quasi-volatile objects are normal memory
+accesses, not subject to unpredictable interference in non-benign
+environments (otherwise the program's behavior would be undefined).
+However, we do impose most of the requirements above on quasi-volatile
+object accesses.
+The last two items are left out: Compilers are allowed to merge or
+omit accesses to these objects.
+Because of this, the item about preserving the rf relation
+applies only when $R$ is not omitted, in which case $W$ must not
+be omitted either, but now it applies in all environments.
+Lastly, the requirement for order preservation is weakened; it applies
+only to pairs of accesses to the same quasi-volatile object.
 Accesses to different objects may be reordered relative to each other.
-The new requirement is that the map must preserve the rf relation.
-That is, if load $R$ in $A$ reads from store $W$ and $R$ is not
-omitted then the instruction it maps to in $H$ must read from the
-instruction that $W$ maps to,
-and consequently $W$ must not be omitted either.
 

### 2024-02-14 ootavacuous.tex
abstract: Pull in quasi-volatile objects
Note that we only conjecture that reasonable implementations should
adhere to quasi-volatile behavior for relaxed non-volatile atomics.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
introduction: Add roadmap
And a LaTeX comment for a future "Future Directions" section.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	restricted to single-threaded analysis in our current universe.
-	We further conjecture that, given current optimizations, these
-	constraints also prevent OOTA cycles involving non-volatile
-	atomics.
+	We extend this work to non-volatile atomics by defining
+	quasi-volatile behavior that we conjecture that usable real-world
+	C++ implementations must adhere to.
 \end{abstract}
@@
 
-@@@ Add brief description and roadmap of paper, including the simplifications
-and restrictions.
+This paper shows that C++ implementations subject to reasonable
+real-world constraints cannot exhibit OOTA cycles.
+Section~\ref{sec:Background}
+provides background on the problem of OOTA cycles, including a
+brief summary of prior work.
+Section~\ref{sec:OOTA and Semantic Dependencies}
+motivates a shift in definition of OOTA cycle and reviews properties of
+semantic dependencies, showing that these properties are a function of
+executions rather than simply of the source code.
+This leads into
+Section~\ref{sec:What is an Execution?},
+which describes properties of abstract and hardware executions, along
+with the relationship between the two.
+
+This in turn leads to
+Section~\ref{sec:C++ Compilers}, which describes properties of the tool
+commonly used to create hardware executions with the proper relationship
+to a given set of abstract executions, namely  the compiler.
+Section~\ref{sec:Hardware Dependencies, Instruction Ordering, and the
+Fundamental Property}
+then combines hardware dependencies and instruction ordering to arrive
+at the fundamental property of semantic dependencies.
+
+This sets the stage for
+Section~\ref{sec:Real-World Constraints}
+to bring real-world constraints to bear, starting with the laws of
+physics, informally demonstrating that these constraints prevent
+OOTA cycles containing only volatile and quasi-volatile atomics from
+occurring in undefined-behavior-free C++ programs running on correctly
+constructed C++ compiler-based implemen- tations that are restricted to
+single-threaded analysis in our current universe.
+
+Finally,
+Section~\ref{sec:Conclusion}
+presents concluding remarks.
+% @@@ Future directions at some point.
 

### 2024-02-14 ootavacuous.tex
OOTA: Add roadmap
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-@@@ Fill in roadmap when sectioning is finalized.
+Section~\ref{sec:OOTA: rf vs. rfe}
+demonstrates advantages of formulating an OOTA cycle as a cycle in
+sdep $\cup$ rfe instead of the traditional sdep $\cup$ rf.
+Section~\ref{sec:Properties of Semantic Dependencies}
+then discusses properties of semantic dependences, showing that they
+are functions of executions rather than strictly of source code.
 

### 2024-02-14 ootavacuous.tex
sdep: Add roadmap
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-@@@ Roadmap
+This section uncovers some semantic-dependency complexities.
+Section~\ref{sec:Semantic Dependencies and Source Code}
+shows that semantic dependencies are functions of executions,
+rather than being strict functions of the source code.
+Section~\ref{sec:Semantic Dependencies Can Be Many To One}
+shows that semantic dependencies do not necessarily extend
+from a single load to a single store, but can instead involve
+multiple loads.
+Section~\ref{sec:Semantic Dependencies Affected by Cross-Thread Optimizations}
+shows that semantic dependencies can be affected by cross-thread
+optimizations.
+Sections~\ref{sec:Semantic Dependencies Affected by if Statements}
+and~\ref{sec:Semantic Dependencies Not Affected by if Statements}
+shows that \co{if} statements can have surprising effects on semantic
+dependencies, up to and including eliminating them completely.
+Finally,
+Section~\ref{sec:Semantic Dependencies and Matching Up Stores}
+demonstrates some challenges in determining which of a group of
+stores is involved in a given semantic dependency.
 

### 2024-02-14 ootavacuous.tex
execution: Add roadmap
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-@@@ Roadmap
+This section looks more carefully at executions, in some cases revisiting
+example code from
+Section~\ref{sec:Properties of Semantic Dependencies}.
+Section~\ref{sec:Abstract Executions}
+looks at executions from the viewpoint of the abstract machine, and
+Section~\ref{sec:Hardware Executions}
+looks at them from the viewpoint of the computer hardware.
+Finally,
+Section~\ref{sec:Relation Between Abstract and Hardware Executions}
+describes how to relate these two viewpoints.
 

### 2024-02-14 ootavacuous.tex
compilers: Add roadmap
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-@@@ Fill in roadmap when sectioning is finalized.
+Section~\ref{sec:Users Influence the Behavior of Compilers}
+shows that C++ compilers can be influenced by their users as well
+as by the standard.
+Section~\ref{sec:Global Optimization Can Destroy Dependencies}
+presents an example showing that global optimizations can destroy
+semantic dependencies, and then
+Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}
+analyzes examples showing that inventing atomic loads can destroy
+semantic dependencies.
+Section~\ref{sec:Volatile and Quasi Volatile Accesses}
+then presents required properties of volatile atomic operations, and
+also defines quasi-volatile properties for non-volatile atomic operations.
 

### 2024-02-14 ootavacuous.sh
ootavacuous.sh: Take early exit after pdflatex/bibtex failure
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -89,8 +89,8 @@ do
 	fi
 done
 
-pdflatex ootavacuous-flat.tex
-bibtex ootavacuous-flat.aux
-pdflatex ootavacuous-flat.tex
-pdflatex ootavacuous-flat.tex
+pdflatex ootavacuous-flat.tex || exit
+bibtex ootavacuous-flat.aux || exit
+pdflatex ootavacuous-flat.tex || exit
+pdflatex ootavacuous-flat.tex || exit
 cp ootavacuous-flat.pdf ootavacuous.pdf

### 2024-02-14 ootavacuous.tex
fundamental: Add roadmap
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-@@@ Roadmap
+Section~\ref{sec:Dependencies at the Hardware Level}
+discusses hardware-level dependencies involving machine instructions.
+Section~\ref{sec:Instruction Ordering}
+then examines the relationships between various instruction-ordering
+mechanisms and semantic dependencies.
+This material feeds into
+Section~\ref{sec:The Fundamental Property of Semantic Dependencies},
+which presents the fundamental property of semantic dependencies
+and related complications.
 

### 2024-02-14 ootavacuous.tex
executions: Add roadmap
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 all the accesses are omitted to be trivial.
+
+The following sections look at semantic-dependency complications that
+can arise in the context of executions.
+Section~\ref{sec:Executions Can Be Merged}
+shows that executions can be merged,
+Section~\ref{sec:Executions Depend on Context}
+shows that executions depend on context,
+Section~\ref{sec:Executions Can Be Nondeterministic}
+looks at nondeterministic executions,
+Section~\ref{sec:Executions Can Be Complex}
+looks at complex executions,
+Section~\ref{sec:Implementers and Users Influence the Definition of Execution}
+describes how implementers and users (not just the standard!) can
+influence executions,
+and finally
+Section~\ref{sec:So What Exactly is an Execution, Anyway?}
+summarizes this material in preparation for bringing the real world
+to bear on these issues.
 

### 2024-02-14 ootavacuous.tex
Point to OOTA example with non-nonsensical values
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Peter Sewell: Hang a lantern on second thanks footnote
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 In the most extreme cases a nonsensical value can pop up ``out of thin air'',
-but this is not necessary.
+however, as shown by
+Listing~\ref{lst:OOTA Cycle},
+OOTA cycles need not involve nonsensical values.
 
@@
 by the C++ implementation.\footnote{
-	Thanks to Peter Sewell for pointing out this possibility.}
+	Again, thanks to Peter Sewell for pointing out this possibility.}
 

### 2024-02-14 ootavacuous.tex
Hardware Executions: s/more well/better/
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 When a given computer executes the machine code in a file, the outcome
-has historically been much more well defined than the executions of the
+has historically been much better defined than the executions of the
 C++ abstract machine.

### 2024-02-14 ootavacuous.tex
Provide volatile non-synchronization example
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 which always must return the value of the store they read from).
-This implies that volatile load-acquire's do not synchronize with
-volatile store-releases in the sense of the C++ memory model,
+This implies that volatile load-acquire's do not necessarily synchronize with
+volatile store-releases in the sense of the C++ memory model,\footnote{
+	For example, roughly speaking, they might instead synchronize
+	with load-acquires and store-releases in device firmware.}
 so they do not contribute in the same way to the happens-before relation.
-Also, in these environments the rfe relation does not apply
+Also, in these environments the rfe relation does not necessarily apply
 to volatile loads and hence the accesses in an OOTA cycle must be

### 2024-02-14 ootavacuous.tex
Add transition paragraph to "C++ Compilers" section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Back-reference atomic-load invention
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 section~\ref{sec:Volatile and Quasi Volatile Accesses} below.
+
+Given this relation between abstract and hardware executions, it is time
+to turn our attention to the tool that selects hardware execution so as
+to preserve that relation, namely, compilers.
 
@@
 This section contains several example litmus tests from the active
-mind of Alan Stern.
+mind of Alan Stern, expanding on the material presented in
+Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.
 

### 2024-02-14 ootavacuous.tex
Demote quasi-volatile parentheses to comma clause
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 indicate that the hardware should perform all accesses to these
-objects exactly as written in the source code (perhaps because they
+objects exactly as written in the source code, perhaps because they
 represent memory-mapped device registers or DMA buffers rather than
-normal memory locations).
+normal memory locations.
 In any event, we expect compilers' translations of volatile-object accesses

### 2024-02-14 ootavacuous.tex
tooling: Add other-problems disclaimer
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-02-14 ootavacuous.tex
@@
 	restricted to single-threaded analysis in our current universe.
-	We extend this work to non-volatile atomics by defining
-	quasi-volatile behavior that we conjecture that usable real-world
-	C++ implementations must adhere to.
+	We further conjecture that, given current optimizations, these
+	constraints also prevent OOTA cycles involving non-volatile
+	atomics.
 \end{abstract}
@@
 
-This paper shows that C++ implementations subject to reasonable
-real-world constraints cannot exhibit OOTA cycles.
-Section~\ref{sec:Background}
-provides background on the problem of OOTA cycles, including a
-brief summary of prior work.
-Section~\ref{sec:OOTA and Semantic Dependencies}
-motivates a shift in definition of OOTA cycle and reviews properties of
-semantic dependencies, showing that these properties are a function of
-executions rather than simply of the source code.
-This leads into
-Section~\ref{sec:What is an Execution?},
-which describes properties of abstract and hardware executions, along
-with the relationship between the two.
-
-This in turn leads to
-Section~\ref{sec:C++ Compilers}, which describes properties of the tool
-commonly used to create hardware executions with the proper relationship
-to a given set of abstract executions, namely  the compiler.
-Section~\ref{sec:Hardware Dependencies, Instruction Ordering, and the
-Fundamental Property}
-then combines hardware dependencies and instruction ordering to arrive
-at the fundamental property of semantic dependencies.
-
-This sets the stage for
-Section~\ref{sec:Real-World Constraints}
-to bring real-world constraints to bear, starting with the laws of
-physics, informally demonstrating that these constraints prevent
-OOTA cycles containing only volatile and quasi-volatile atomics from
-occurring in undefined-behavior-free C++ programs running on correctly
-constructed C++ compiler-based implemen- tations that are restricted to
-single-threaded analysis in our current universe.
-
-Finally,
-Section~\ref{sec:Conclusion}
-presents concluding remarks.
-% @@@ Future directions at some point.
+@@@ Add brief description and roadmap of paper, including the simplifications
+and restrictions.
 
@@
 In the most extreme cases a nonsensical value can pop up ``out of thin air'',
-however, as shown by
-Listing~\ref{lst:OOTA Cycle},
-OOTA cycles need not involve nonsensical values.
+but this is not necessary.
 
@@
 
-Section~\ref{sec:OOTA: rf vs. rfe}
-demonstrates advantages of formulating an OOTA cycle as a cycle in
-sdep $\cup$ rfe instead of the traditional sdep $\cup$ rf.
-Section~\ref{sec:Properties of Semantic Dependencies}
-then discusses properties of semantic dependences, showing that they
-are functions of executions rather than strictly of source code.
+@@@ Fill in roadmap when sectioning is finalized.
 
@@
 
-This section uncovers some semantic-dependency complexities.
-Section~\ref{sec:Semantic Dependencies and Source Code}
-shows that semantic dependencies are functions of executions,
-rather than being strict functions of the source code.
-Section~\ref{sec:Semantic Dependencies Can Be Many To One}
-shows that semantic dependencies do not necessarily extend
-from a single load to a single store, but can instead involve
-multiple loads.
-Section~\ref{sec:Semantic Dependencies Affected by Cross-Thread Optimizations}
-shows that semantic dependencies can be affected by cross-thread
-optimizations.
-Sections~\ref{sec:Semantic Dependencies Affected by if Statements}
-and~\ref{sec:Semantic Dependencies Not Affected by if Statements}
-shows that \co{if} statements can have surprising effects on semantic
-dependencies, up to and including eliminating them completely.
-Finally,
-Section~\ref{sec:Semantic Dependencies and Matching Up Stores}
-demonstrates some challenges in determining which of a group of
-stores is involved in a given semantic dependency.
+@@@ Roadmap
 
@@
 
-This section looks more carefully at executions, in some cases revisiting
-example code from
-Section~\ref{sec:Properties of Semantic Dependencies}.
-Section~\ref{sec:Abstract Executions}
-looks at executions from the viewpoint of the abstract machine, and
-Section~\ref{sec:Hardware Executions}
-looks at them from the viewpoint of the computer hardware.
-Finally,
-Section~\ref{sec:Relation Between Abstract and Hardware Executions}
-describes how to relate these two viewpoints.
+@@@ Roadmap
 
@@
 When a given computer executes the machine code in a file, the outcome
-has historically been much better defined than the executions of the
+has historically been much more well defined than the executions of the
 C++ abstract machine.
@@
 which always must return the value of the store they read from).
-This implies that volatile load-acquire's do not necessarily synchronize with
-volatile store-releases in the sense of the C++ memory model,\footnote{
-	For example, roughly speaking, they might instead synchronize
-	with load-acquires and store-releases in device firmware.}
+This implies that volatile load-acquire's do not synchronize with
+volatile store-releases in the sense of the C++ memory model,
 so they do not contribute in the same way to the happens-before relation.
-Also, in these environments the rfe relation does not necessarily apply
+Also, in these environments the rfe relation does not apply
 to volatile loads and hence the accesses in an OOTA cycle must be
@@
 
-Section~\ref{sec:Users Influence the Behavior of Compilers}
-shows that C++ compilers can be influenced by their users as well
-as by the standard.
-Section~\ref{sec:Global Optimization Can Destroy Dependencies}
-presents an example showing that global optimizations can destroy
-semantic dependencies, and then
-Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}
-analyzes examples showing that inventing atomic loads can destroy
-semantic dependencies.
-Section~\ref{sec:Volatile and Quasi Volatile Accesses}
-then presents required properties of volatile atomic operations, and
-also defines quasi-volatile properties for non-volatile atomic operations.
+@@@ Fill in roadmap when sectioning is finalized.
 
@@
 
-Section~\ref{sec:Dependencies at the Hardware Level}
-discusses hardware-level dependencies involving machine instructions.
-Section~\ref{sec:Instruction Ordering}
-then examines the relationships between various instruction-ordering
-mechanisms and semantic dependencies.
-This material feeds into
-Section~\ref{sec:The Fundamental Property of Semantic Dependencies},
-which presents the fundamental property of semantic dependencies
-and related complications.
+@@@ Roadmap
 
@@
 
-The following sections look at semantic-dependency complications that
-can arise in the context of executions.
-Section~\ref{sec:Executions Can Be Merged}
-shows that executions can be merged,
-Section~\ref{sec:Executions Depend on Context}
-shows that executions depend on context,
-Section~\ref{sec:Executions Can Be Nondeterministic}
-looks at nondeterministic executions,
-Section~\ref{sec:Executions Can Be Complex}
-looks at complex executions,
-Section~\ref{sec:Implementers and Users Influence the Definition of Execution}
-describes how implementers and users (not just the standard!) can
-influence executions,
-and finally
-Section~\ref{sec:So What Exactly is an Execution, Anyway?}
-summarizes this material in preparation for bringing the real world
-to bear on these issues.
-
 \subsubsection{Executions Can Be Merged}
@@
 by the C++ implementation.\footnote{
-	Again, thanks to Peter Sewell for pointing out this possibility.}
+	Thanks to Peter Sewell for pointing out this possibility.}
 
@@
 
-This appendix expands on
+One entirely reasonable reaction is that, given the issues raised in
+Sections~\ref{sec:Properties of Semantic Dependencies}
+and~\ref{sec:The Fundamental Property of Semantic Dependencies},
+load/store reordering is the least of the problems faced by tooling.
+Nevertheless, this appendix expands on
 Section~\ref{sec:Semantic Dependencies and Tooling}

### 2024-02-14 ootavacuous.tex
Apply Alan Stern feedback to non-benign example
6fb6358577e8 ("Provide volatile non-synchronization example")

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-02-14 ootavacuous.tex
@@
 which always must return the value of the store they read from).
-This implies that volatile load-acquire's do not necessarily synchronize with
+This implies that volatile load-acquire's do not synchronize with
 volatile store-releases in the sense of the C++ memory model,\footnote{
-	For example, roughly speaking, they might instead synchronize
-	with load-acquires and store-releases in device firmware.}
+	However, they might instead synchronize with store-releases in
+	device firmware (or vice versa), roughly speaking.}
 so they do not contribute in the same way to the happens-before relation.
-Also, in these environments the rfe relation does not necessarily apply
-to volatile loads and hence the accesses in an OOTA cycle must be
-non-volatile.
+Also, in these environments the rfe relation does not apply to volatile
+loads and hence the accesses in an OOTA cycle must be non-volatile.
 

### 2024-02-15 ootavacuous.tex
Add a "Future Directions" section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 potentially all the way down to zero.
+
+\section{Future Directions}
+\label{sec:Future Directions}
+
+This paper focuses solely on compilers, but it should be possible to
+extend these results to some classes of interpreters and JITs on the
+one hand and to link-time optimizations (LTO) on the other.
+
+This paper focuses further on compilers that do only local per-thread
+analysis.
+Further explorations could consider some classes of global analysis,
+for example, analyses that demonstrate that a given expression
+will always evaluate to a particular constant.
+JMM Causality Test Case 1 discussed in
+Appendix~\ref{app:Causality Test Case 1}
+provides an example of such a global analysis.
+
+This paper assumes that the user defines threads, and that the
+C++ implementation executes those threads unchanged.
+Future analyses might examine the possibility of ``flattening''
+optimizations that combine multiple threads into one.
 

### 2024-02-16 ootavacuous.tex
Define semantic dependency
Add a chapter defining semantic dependency, proving that the
Fundamental Property holds, and raising some questions.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-02-14 ootavacuous.tex
@@
 all the accesses are omitted to be trivial.
+
+\section{A Definition of Semantic Dependency}
+\label{sec:A Definition of Semantic Dependency}
+
+Semantic dependency is a notoriously difficult concept to define
+rigorously and precisely.
+A large part of the reason is because it was never a completely clear
+concept to begin with, especially when there are multiple accesses to
+the variables involved.
+In this section we will stick our necks out by offering just such a
+definition.
+No doubt many people will object to it for various reasons, but we
+hope it will provide a starting point for discussions that move the
+entire field forward.
+
+The definition given below is applicable only to C++ implementations
+that treat all atomic objects as though they are volatile or
+quasi-volatile.
+For compilers that perform only single-thread analysis, not global
+analysis, quasi-volatile is sufficient.
+In this setting we can relate abstract and hardware executions by
+means of the map of accesses described in
+section~\ref{sec:Volatile and Quasi Volatile Accesses}.
+The key insight is that this allows us to consider semantic
+dependencies at the level of the machine code, where they are much
+more tractable.
+
+@@@ Roadmap
+
+\subsection{For Compilers Using Single-Thread Analysis}
+\label{sec:For Compilers Using Single-Thread Analysis}
+
+In this section we consider implementations whose compilers perform
+only single-thread analysis and treat all atomic objects as
+quasi-volatile.
+This implies that if two different programs contain the same thread
+(i.e., the same source code for the functions in the thread), the
+machine code generated by the compiler for the thread will be the
+same in the two programs.
+
+We begin by recognizing that semantic dependencies are relative to a
+particular execution and a particular implementation,
+as described in section~\ref{sec:Properties of Semantic Dependencies}.
+The same source code may or may not contain a semantic dependency,
+according to the details of the execution in question and the
+machine code produced by the compiler.
+For this reason we will characterize semantic dependencies in a given
+abstract execution realized by a given hardware execution.
+(While it possible to argue about semantic dependencies in abstract
+executions that have no hardware realizations, doing so seems like a
+pointless activity.)
+
+Let $A$ be an abstract execution of some program $P$ containing a
+thread $T$, and let $H$ be a hardware execution realizing $A$.
+Let $W$ be an store in $T$ to an atomic object and let $\{R_i \mid i =
+0, \ldots, n\}$ be a set of loads in $T$ from atomic objects on which
+$W$ might or might not depend.
+We can dispose of one case immediately: If $W$ is omitted in $H$ then
+the issue of semantic dependency is moot.
+You can give either answer since the question is basically meaningless.
+Therefore we'll assume that $W$ is not omitted.
+Then:
+\begin{quote}
+There is a semantic dependency from $\{R_i\}$ to $W$ in $A$ and $H$,
+relative to the compiler used to produce $H$, if there is another
+abstract execution $B$ realized by hardware execution $G$ under the
+same compiler that together \emph{witness the semantic dependency}.
+\end{quote}
+
+To be a proper witness, $B$ must be an execution of some program $Q$,
+not necessarily the same as $P$ but which contains the same thread $T$.
+The thread should start out with the same initial state in $A$ and
+$B$, and all loads in $A$ coming before any of the $R_i$ should obtain
+the same value as they do in $B$ (this is part of our interpretation
+of ``all other things being equal'').
+It follows that the two abstract executions of $T$ will be identical
+up to the first of the $R_i$ loads.
+
+Let $W'$ and $\{R'_i\}$ be the accesses in $H$ that $W$ and the
+non-omitted $\{R_i\}$ loads map to.
+We then require that the hardware executions of $T$ in $H$ and $G$ be
+identical for an initial period lasting up to the first of the $R'_i$.
+Following this initial period there will be a common period, during
+which $H$ and $G$ execute the same machine instructions but do not
+necessarily compute the same values.
+This common period ends when one of the hardware executions
+takes a conditional branch that the other doesn't, or when a computed
+branch leads to different addresses in the two executions, or when $T$
+ends, whichever comes first.
+Past this point $H$ and $G$ diverge and are no longer directly
+comparable, as they execute different instructions.
+Our third requirement for being a witness is that each load in the
+common portion either must obtain the same value in $H$ and $G$, or
+must itself be one of the $R'_i$ loads, or must be ordered in $H$
+after one of the $R'_i$ loads (this is the remaining part of our
+interpretation of ``all other things being equal''.)
+
+Finally, we need to determine an instruction $X'$ in $G$ that
+corresponds to $W'$.
+If $W'$ is part of the initial or common portion of $H$ this is no
+problem; we can take $X'$ to be $W'$ itself.
+But if $W'$ is in the divergent portion of $H$ then things aren't so
+simple.
+The choice is somewhat arbitrary, and so we will fall back on the
+earlier proposal of matching up stores by the order they occur.
+Let \co{y} be the atomic object that $W'$ stores to, and suppose $W'$
+is the $N$th store to \co{y} within the divergent portion of $H$.
+Then $X'$ will be the $N$th store to \co{y} in the divergent portion
+of $G$, if such a store exists.
+Our last requirement for being a witness to a semantic dependency
+is that $X'$ act differently from $W'$: it doesn't exist, it stores
+a different value, or it stores to an object other than \co{y}.
+
+\subsection{For Compilers Using Global Analysis}
+\label{sec:For Compilers Using Global Analysis}
+
+As promised earlier, we now consider implementations whose compilers
+may use global analysis.
+In order to obtain the desired results we have to require that these
+compilers treat all atomic objects as volatile.
+Equivalently, the machine code generated by such a compiler must be
+the same for a given program as for a ``volatilized'' form of the
+program in which all atomic objects are defined to be volatile.
+
+In this context our definition of semantic dependency is essentially
+the same as before.
+Since we can no longer expect the machine code for a thread to be the
+same regardless of the program it belongs to, the program $Q$ in the
+earlier definition (of which $B$ and $G$ are executions) must now be
+$P$ or its volatilized form.
+However, we do now allow the possibility that the executions $B$ and
+$G$ take place in a non-benign environment.
+Aside from these minor adjustments, the definition remains unchanged.
+
+\subsection{Verifying the Fundamental Property}
+\label{sec:Verifying the Fundamental Property}
+
+Of course we want to check that our definition of semantic dependency
+satisfies the Fundamental Property of
+section~\ref{sec:The Fundamental Property of Semantic Dependencies}.
+Given the information we have already presented, the demonstration is
+easy.
+
+Suppose we have $W$, $\{R_i\}$, $A$, and $H$ as in the definition.
+The Fundamental Property assumes that $W$ is not omitted in $H$, so
+there is an abtract execution $B$ with hardware realization $G$
+witnessing the semantic dependency of the store $W$ on the loads
+$\{R_i\}$ in $A$ and $H$.
+We must show that some $R_i$ is not omitted and $R'_i$ is ordered
+before $W'$ in $H$.
+The proof splits into three cases.
+
+First case: $W'$ lies in the initial portion of $H$.
+During the initial portion of the hardware executions, $H$ and $G$
+behave identically and therefore $W'$ performs the same write in
+both.
+This contradicts the fact that $B$ and $G$ witness the semantic
+dependency.
+
+Second case: $W'$ lies in the common portion of $H$.
+Since the action of $W'$ in $H$ is different from its action in $G$,
+at least one of its inputs must differ between the two hardware
+executions.
+Therefore the source instruction for that input must behave
+differently, and so must one of its sources, going back until we reach
+a load instruction that obtains differing values in $H$ and $G$.
+Then $W'$ depends on this load and so is ordered after it.
+And since the load must lie in the common portion of $H$, by the
+definition of semantic dependency it must either be one of the
+$R'_i$ or be ordered after one of them.
+Therefore $W'$ is ordered after one of the $R'_i$ in $H$,
+which certainly means that $R_i$ is not omitted.
+
+Third case: $W'$ lies in the divergent portion of $H$.
+This happens when $W'$ comes after the conditional or computed branch
+which marks the end of the common period by going different ways in
+$H$ and $G$.
+Just as in the previous case, since the branch behaves differently in
+the two executions it must be ordered after one of the $R'_i$ loads.
+And then so must $W'$, because any instruction following a conditional
+or computed branch instruction is ordered after the branch.
+QED.
+
+\medskip
+
+A corollary of this result is that if an implementation's compiler
+either
+\begin{itemize}
+\item	uses single-thread analysis and treats atomic objects as
+	quasi volatile, or
+\item	uses global analysis and treats atomic objects as volatile,
+\end{itemize}
+then programs produced by that implementation will never exhibit OOTA.
+Thus the implementation will automatically comply with full C++, even
+though it may be been designed only to comply with loose C++.
+
+\subsection{Outstanding Issues}
+\label{sec:Outstanding Issues}
+
+Here we consider some general questions related to our definition of
+semantic dependency.
+
+\subsubsection{Relative versus Absolute Dependency}
+\label{sec:Relative versus Absolute Dependency}
+
+A drawback of the definition is that it is only relative to a specific
+compiler or implementation.
+This may strike some people as too compromising, in that a given piece
+of code should either contain or not contain a semantic dependency
+independent of the implementation used to run it.
+
+We can address this drawback by defining an \emph{absolute semantic
+dependency} as one that is present relative to any valid C++
+implementation, past, present, or future.
+Of course this notion has its own problems, including that it is
+extremely non-constructive and impossible to apply in practice.
+However it may be the best we can do with our current understanding
+of computing systems.
+
+There is one thing we can definitely state:
+Programs produced by an implementation of the right sort will never
+exhibit absolute OOTA (that is, an OOTA cycle in which all the
+semantic dependencies are absolute).
+This is simply because a semantic dependency that is absolute is {\it
+a fortiori\/} a semantic dependency relative to the compiler in use.
+
+But in fact we have shown more than this:
+Programs will never exhibit an OOTA cycle relative to the compiler
+used to build them, even when that cycle is not absolute.
+In this sense we have gone beyond the requirement of full C++.
+
+\subsubsection{Global Analysis and Volatile versus Quasi Volatile}
+\label{sec:Global Analysis and Volatile versus Quasi Volatile}
+
+In principle we don't need to require global-analysis compilers to
+treat atomic objects as volatile; our results would still hold if they
+merely treated them as quasi volatile.
+We chose not to do this because it would violate our intuitions about
+semantic dependencies.
+
+For example, consider the Simple OOTA program, repeated here in
+simplified form:
+\begin{quote}
+\begin{verbatim}
+void thread1() {
+   int r1 = x;
+   y = r1;
+}
+
+void thread2() {
+   int r2 = y;
+   x = r2;
+}
+\end{verbatim}
+\end{quote}
+A compiler using global analysis and treating \co{x} and \co{y} as
+quasi-volatile objects could omit the two loads, replacing them in the
+machine code with simple assignments ``\co{r1 = 42}'' and
+``\co{r2 = 42}''.
+This would be a valid transformation, and the resulting behavior would
+not count as OOTA because according to our definition, the
+dependencies in \co{thread1} and \co{thread2} would not be semantic.
+
+To see why, recall that a semantic dependency must have a witness,
+another execution in which the store acts differently.
+But this transformed program has no other executions; every time it
+runs it will write 42 to both \co{x} and \co{y}.
+(Keep in mind also that since the atomic objects are not treated as
+volatile, they are not subject to unspecified interference when the
+program runs in a non-benign environment.)
+
+This unintuitive behavior could not occur if the two loads were not
+omitted.
+In fact, the definition of semantic dependency might remain
+perfectly acceptable if the requirement for global-analysis compilers
+were weakened, if the compiler treated atomic objects as quasi
+volatile and in addition was not allowed to omit accesses to them.
+This is a possible topic for future research.
+
+\subsubsection{Effect of Memory Layout}
+\label{sec:Effect of Memory Layout}
+
+Part of our demonstration of the Fundamental Property of semantic
+dependencies relies on the fact, stated in
+section~\ref{sec:Abstract Executions},
+that an abstract execution of a thread is entirely determined by the
+values obtained for its loads.
+But when we compare abstract executions of the same thread in two
+different programs, this may no longer be entirely true owing to the
+effect of differing memory layouts.
+
+Consider this simple example:
+\begin{quote}
+\begin{verbatim}
+x = (int) &x;
+\end{verbatim}
+\end{quote}
+Even though the example contains no loads at all, it may store
+different values when running in different programs because the
+variable \co{x} may be allocated at differing addresses in those
+programs.
+According to our definition, this could count as a degenerate OOTA
+cycle of length one, in which the store is semantically dependent on
+an empty set of loads!
+
+To rule out such pathological counterexamples we should require that
+in a witness to a semantic dependency, the addresses of all the
+objects and functions referred to in the thread $T$ are the same as
+in the original execution.
+This is a very technical restriction but there are occasions when
+the issue might realistically arise, such as when computing a hash
+value based on an object's address.
+
+\subsubsection{Merging Quasi-Volatile Loads}
+\label{sec:Merging Quasi-Volatile Loads}
+
+The compiler is permitted to merge quasi-volatile loads.
+This can lead to surprising results because a particular load
+may be merged with an earlier load in one execution and with a
+later load in another.
+This is demonstrated in the following, which is a variant of
+the example in
+section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}:
+\begin{quote}
+\begin{verbatim}
+int r1 = (x != 0);
+int r2 = (x != 0);
+int r3 = (x != 0);
+int r4 = y;
+z = (r2 == r4);
+\end{verbatim}
+\end{quote}
+Consider an abstract execution in which \co{r1}, \co{r2}, and \co{r4}
+are zero and \co{r3} is one (because another thread changed the value
+of \co{x} between two of the loads).
+We would expect that the store to \co{z} would be semantically
+dependent on the load of \co{y} in this execution.
+
+However, a single-thread analysis compiler can translate this into the
+machine code equivalent of:
+\begin{quote}
+\begin{verbatim}
+int r1 = (x != 0);
+int r2;
+int r3 = (x != 0);
+int r4 = y;
+if (r1 != r3) {
+   r2 = r4;
+   z = 1;
+} else {
+   r2 = r1;
+   z = (r2 == r4);
+}
+\end{verbatim}
+\end{quote}
+In effect, the \co{r2} load is merged with the \co{r1} load in
+executions where \co{r1} is equal to \co{r3} or to \co{r4}, and it is
+merged with the \co{r3} load in other executions.
+
+To demonstrate the semantic dependency, a suitable witness would have
+to include a hardware realization of the abstract execution in which
+\co{r1} and \co{r2} are zero and \co{r3} and \co{r4} are one.
+But there are no hardware realizations of this execution with the
+machine code indicated above!
+Since \co{r1} is different from both \co{r3} and \co{r4}, the \co{r2}
+load will be merged with the \co{r3} load and so \co{r2} will
+necessarily be one, not zero.
+Thus, relative to this compiler the example does not contain a
+semantic dependency.
+
+Although it is unexpected, we cannot say that this conclusion is
+definitely wrong, because our intuitive notions of semantic
+dependency are not clear in cases where multiple loads of the same
+variable are present.
+
+\section{To Be Moved Or Removed...}
+
+@@@ Review
 

### 2024-02-16 ootavacuous.tex

@@
 This implies that volatile load-acquire's do not synchronize with
-volatile store-releases in the sense of the C++ memory model,
+volatile store-releases in the sense of the C++ memory model,\footnote{
+	However, they might instead synchronize with store-releases in
+	device firmware (or vice versa), roughly speaking.}
 so they do not contribute in the same way to the happens-before relation.
-Also, in these environments the rfe relation does not apply
-to volatile loads and hence the accesses in an OOTA cycle must be
-non-volatile.
+Also, in these environments the rfe relation does not apply to volatile
+loads and hence the accesses in an OOTA cycle must be non-volatile.
 
@@
 
+\section{Future Directions}
+\label{sec:Future Directions}
+
+This paper focuses solely on compilers, but it should be possible to
+extend these results to some classes of interpreters and JITs on the
+one hand and to link-time optimizations (LTO) on the other.
+
+This paper focuses further on compilers that do only local per-thread
+analysis.
+Further explorations could consider some classes of global analysis,
+for example, analyses that demonstrate that a given expression
+will always evaluate to a particular constant.
+JMM Causality Test Case 1 discussed in
+Appendix~\ref{app:Causality Test Case 1}
+provides an example of such a global analysis.
+
+This paper assumes that the user defines threads, and that the
+C++ implementation executes those threads unchanged.
+Future analyses might examine the possibility of ``flattening''
+optimizations that combine multiple threads into one.
+
 \section{Conclusion}

### 2024-02-16 cuttingroom.tex
Move "complex sdep" to cbmc portion of cutting-room.tex
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 outlines an informal informal definition of ``semantic dependency''
-that can be used in \co{cbmc} test cases.
+that can be used in \co{cbmc} test cases, and then
+Appendix~\ref{sec:Executions Can Be Complex} gives some examples
+of mathematically complex semantic dependencies.
 
@@
 
+\subsection{Executions Can Be Complex}
+\label{sec:Executions Can Be Complex}
+
+Consider this example:
+
+\begin{quote}
+\begin{verbatim}
+x = horribly_complex(y, z);
+\end{verbatim}
+\end{quote}
+
+Here, the presence or absence of a semantic dependency from \co{y} to
+\co{x} depends not only on the value of \co{z}, but also and execution
+that passes through the code in a function that, although single threaded,
+is horribly complex.
+
+A special case is as follows:
+
+\begin{quote}
+\begin{verbatim}
+x = y & 0x1 ? do_odd(y) : do_even(y);
+\end{verbatim}
+\end{quote}
+
+The key point here is that different values might result in almost
+completely disjoint executions.
+
+Fortunately, there are now tools that can help.
+For example, the SAT-solver-based \co{cbmc} tool
+has been used to mechanically verify signicant portions of Linux-kernel
+RCU from the C-language source
+code~\cite{LihaoLiang2016VerifyTreeRCU,LanceRoy2017CBMC-SRCU}.
+Another tool, Nidhugg, which is based on partial-order
+reduction, has been used to carry out a similar
+verification~\cite{MichalisKokologiannakis2017NidhuggRCU,SMC-TreeRCU,MichalisKokologiannakis2019RCUstatelessModelCheck}.
+Both tools can easily check whether or not a given dependency is semantic
+for at least some approximate definitions of ``semantic dependency''.
+% An example use of \co{cbmc} is illustrated in
+% Appendix~\ref{sec:Evaluating sdep Using cbmc}.
+
+There are also many other tools, but failing that, it is always possible
+to fall back to older software-verification tools such as manual
+code inspection.
+
+But consider the following code fragment:
+
+\begin{quote}
+\begin{verbatim}
+x = collatz_cycle_min(y);
+\end{verbatim}
+\end{quote}
+
+Here, \co{collatz_cycle_min()}\footnote{
+	\url{https://en.wikipedia.org/wiki/Collatz_conjecture}}
+takes its arbitrary-precision-integer argument and repeatedly either
+divides it by two if it is even, or multiplies it by three and adds one
+if it is odd.
+Should the resulting sequence of arbitrary-precision integers reach
+a cycle, the function returns the minimum element of that cycle.
+Leaving aside the practical complication of a value of \co{y} that
+consumes more than half of the available memory, there is reason to
+believe that the above code fragment will always assign the value 1 to
+\co{x}.
+Except that this reason for belief is based on the Collatz conjecture,
+which is currently unproven and might never be proven (or disproven).
+
+A similar example could be constructed from Fermat's Last Theorem
+(FLT),\footnote{
+\url{https://en.wikipedia.org/wiki/Fermat\%27s_Last_Theorem}}
+which was first stated in the year 1637 and finally proved more than
+300~years later, in 1994.
+Therefore, it was only after 1994 that certain types of executions
+relying on FLT were known not to have a semantic dependency.
+
+Many analysis tools would respond to these examples by failing, for example,
+by consuming all available CPU and memory.
+
+One could quite reasonably argue that these two examples are irrelevant
+to any real-world C++ implementation.
+After all, evidence to date suggests that code relying on either the
+Collatz Conjecture or FLT is quite rare, and thus not worth optimizing.
+And the value of any given optimization is another real-world constraint
+on the problem of locating OOTA cycles, but one that will not be considered
+further in this document.
+
+\emph{Because of unsolved mathematical problems, precise definitions of
+``execution'' have limitations.
+Therefore, C++ implementations must assume that a dependency passing
+through a given execution is a semantic dependency unless and until it
+can prove otherwise.}
+
 \subsection{Complex Semantic Dependencies}

### 2024-02-16 ootavacuous.tex
Move "complex sdep" to cbmc portion of cutting-room.tex
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 that the object code is deterministic, at least on a per-execution basis.}
-
-\subsubsection{Executions Can Be Complex}
-\label{sec:Executions Can Be Complex}
-
-Consider this example:
-
-\begin{quote}
-\begin{verbatim}
-x = horribly_complex(y, z);
-\end{verbatim}
-\end{quote}
-
-Here, the presence or absence of a semantic dependency from \co{y} to
-\co{x} depends not only on the value of \co{z}, but also and execution
-that passes through the code in a function that, although single threaded,
-is horribly complex.
-
-A special case is as follows:
-
-\begin{quote}
-\begin{verbatim}
-x = y & 0x1 ? do_odd(y) : do_even(y);
-\end{verbatim}
-\end{quote}
-
-The key point here is that different values might result in almost
-completely disjoint executions.
-
-Fortunately, there are now tools that can help.
-For example, the SAT-solver-based \co{cbmc} tool
-has been used to mechanically verify signicant portions of Linux-kernel
-RCU from the C-language source
-code~\cite{LihaoLiang2016VerifyTreeRCU,LanceRoy2017CBMC-SRCU}.
-Another tool, Nidhugg, which is based on partial-order
-reduction, has been used to carry out a similar
-verification~\cite{MichalisKokologiannakis2017NidhuggRCU,SMC-TreeRCU,MichalisKokologiannakis2019RCUstatelessModelCheck}.
-Both tools can easily check whether or not a given dependency is semantic
-for at least some approximate definitions of ``semantic dependency''.
-% An example use of \co{cbmc} is illustrated in
-% Appendix~\ref{sec:Evaluating sdep Using cbmc}.
-
-There are also many other tools, but failing that, it is always possible
-to fall back to older software-verification tools such as manual
-code inspection.
-
-But consider the following code fragment:
-
-\begin{quote}
-\begin{verbatim}
-x = collatz_cycle_min(y);
-\end{verbatim}
-\end{quote}
-
-Here, \co{collatz_cycle_min()}\footnote{
-	\url{https://en.wikipedia.org/wiki/Collatz_conjecture}}
-takes its arbitrary-precision-integer argument and repeatedly either
-divides it by two if it is even, or multiplies it by three and adds one
-if it is odd.
-Should the resulting sequence of arbitrary-precision integers reach
-a cycle, the function returns the minimum element of that cycle.
-Leaving aside the practical complication of a value of \co{y} that
-consumes more than half of the available memory, there is reason to
-believe that the above code fragment will always assign the value 1 to
-\co{x}.
-Except that this reason for belief is based on the Collatz conjecture,
-which is currently unproven and might never be proven (or disproven).
-
-A similar example could be constructed from Fermat's Last Theorem
-(FLT),\footnote{
-\url{https://en.wikipedia.org/wiki/Fermat\%27s_Last_Theorem}}
-which was first stated in the year 1637 and finally proved more than
-300~years later, in 1994.
-Therefore, it was only after 1994 that certain types of executions
-relying on FLT were known not to have a semantic dependency.
-
-Many analysis tools would respond to these examples by failing, for example,
-by consuming all available CPU and memory.
-
-One could quite reasonably argue that these two examples are irrelevant
-to any real-world C++ implementation.
-After all, evidence to date suggests that code relying on either the
-Collatz Conjecture or FLT is quite rare, and thus not worth optimizing.
-And the value of any given optimization is another real-world constraint
-on the problem of locating OOTA cycles, but one that will not be considered
-further in this document.
-
-\emph{Because of unsolved mathematical problems, precise definitions of
-``execution'' have limitations.
-Therefore, C++ implementations must assume that a dependency passing
-through a given execution is a semantic dependency unless and until it
-can prove otherwise.}
 

### 2024-02-16 cuttingroom.tex
Move additional material to cuttingroom.tex
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 program into a single thread.
+
+\subsubsection{Executions Can Be Merged}
+\label{sec:Executions Can Be Merged}
+
+Consider this example, similar to that shown in
+Section~\ref{sec:Semantic Dependencies Not Affected by if Statements},
+which appears to be an \co{if} statement:
+
+\begin{quote}
+\begin{verbatim}
+r1 = x;
+if (r1 > 0)
+    y = 42;
+else
+    y = 42;
+\end{verbatim}
+\end{quote}
+
+Because the stores executed on each branch of that \co{if} statement store
+identical values to identical addresses, one could equally well regard
+the two statements as performing two different stores or as performing
+(for all intents and purposes) a single store.
+Although software projects relying on ordering from load-to-store control
+dependencies and using a variety of compilers should should err on the
+side of caution by assuming there is no semantic
+dependency~\cite{Howells2009membartxt},
+the fact remains that reasonable C++ implementations might disagree as
+to whether this code represents one execution or two.
+It is the implementations's choice.
+
+A similar caution applies to loads:
+
+\begin{quote}
+\begin{verbatim}
+r1 = x;
+if (r1 > 42)
+    r2 = y;
+else
+    r2 = y;
+z = r2;
+\end{verbatim}
+\end{quote}
+
+The two loads from \co{y} are fungible, and so a C++ implementation
+could behave as if this \co{if} statement had instead been a single load
+from \co{y}.
+
+These examples demonstrate a key point: 
+\emph{In general, executions do not necessarily map one-to-one onto
+paths through the source code}.% \footnote{
+% 	Thanks to Alan Stern for the examples in this section and
+% 	in the preceding one.}
+
+\subsubsection{Executions Depend on Context}
+\label{sec:Executions Depend on Context}
+
+Consider the following example derived from that in
+Section~\ref{sec:Semantic Dependencies and Source Code}:
+
+\begin{quote}
+\begin{verbatim}
+r1 = x;
+// r1 = 0;
+z = y * r1;
+// z = y / r1;
+\end{verbatim}
+\end{quote}
+
+As written, there is a semantic dependency from \co{x} and \co{y}
+to \co{z}.
+But if the second line is uncommented, there is no semantic
+dependency.
+If the second line is left commented, but the last line is uncommented,
+then the compiler is free to backwards-propagate the potential
+divide-by-zero undefined behavior to precede the first assignment
+to \co{z}.
+This could prevent both assignments to \co{z} from executing, for example,
+due to a divide-by-zero exception.
+
+More subtle changes in context can also have an effect.
+For example:
+
+\begin{quote}
+\begin{verbatim}
+r1 = x;
+z = y * r1;
+z = 42;
+\end{verbatim}
+\end{quote}
+
+Given that there is no ordering, a C++ implementation could choose to
+act as if the first store to \co{z} was omitted, even if that store was
+a relaxed atomic store.
+But if \co{z} was marked volatile, both stores would need to be executed
+in strict accordance with the rules of the abstract machine.
+
+\emph{Any analysis of a given execution either must be fully informed
+of the enclosing context, or must make conservative assumptions about
+that context.}
+
+\subsubsection{Executions Can Be Nondeterministic}
+\label{sec:Executions Can Be Nondeterministic}
+
+Consider this example, similar to that in
+Section~\ref{sec:Abstract Executions},
+with \co{i} initially zero:
+
+\begin{quote}
+\begin{verbatim}
+int foo(int a, int b)
+{
+   return a / b;
+}
+
+y = foo(++i, ++i);
+x = y * z;
+\end{verbatim}
+\end{quote}
+
+Because early C~implementers could not come to agreement, standard
+does not specify the order of evaluation of function arguments, so the
+resulting value of \co{y} might be zero ($\frac{1}{2}$ truncated) or two
+($\frac{2}{1}$).
+In the former case, there is no semantic dependency from \co{z} to
+\co{x}, but in the latter case there is.
+Thus, semantic dependencies are not just a function of a particular
+execution through the source code, but also of arbitrary choices made
+by the C++ implementation.\footnote{
+	Again, thanks to Peter Sewell for pointing out this possibility.}
+
+Note well that C++ implementations can and do act as if portions of
+the source code was duplicated in order to apply specialization
+optimizations, in which case any non-determinism might also be
+a function of a given execution.
+
+Many projects' coding guidelines prohibit side effects in expressions,
+their goal being to obtain portable code whose behavior does not depend
+on arbitrary choices on the part of the implementation.
+However, the above code nevertheless conforms to the standard.
+Note that side effects in expressions lacking sequence points is
+undefined behavior, however, there are proposals to define it in the
+C++~\cite{GabrielDosReis2016P0145r3} and
+C~\cite{AlexCeleste2023N3203}
+languages, allowing implementations to nevertheless reorder
+evaluation when permitted by the as-if rule.
+
+Even though the standard is non-deterministic, \emph{compiler-based C++
+implementions make their non-deterministic choices at compile time, so
+that the object code is deterministic, at least on a per-execution basis.}
+
+\subsubsection{Implementers and Users Influence the Definition of Execution}
+\label{sec:Implementers and Users Influence the Definition of Execution}
+
+The exact definition of a computer language is subject to some debate,
+with standards, implementations, and users all having some degree of
+influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
+and each of which is subject to change over time.
+Users  and implementers are of course wise to assume that their opinions
+might be overridden by those of the standard and, for the users,
+especially the implementation.
+However, consider the following code, similar to that shown in
+Section~\ref{sec:Abstract Executions}:
+
+\begin{quote}
+\begin{verbatim}
+void foo(int x, char c)
+{
+    return x * (c >= 0);
+}
+\end{verbatim}
+\end{quote}
+
+The standard does not specify whether or not type \co{char} is signed,
+leaving that choice to the implementations, which follow the lead
+of the architects of a given CPU family.
+Therefore, the code above has a semantic dependency in executions running
+on systems choosing signed \co{char}, but not those running on systems
+choosing unsigned \co{char}.
+So score one for the implementers, albeit by conscious choice on the
+part of the standards committee.
+
+Except that GCC provides the \co{-funsigned-char} command-line
+argument that causes this implementation to treat variables of
+type \co{char} as unsigned, regardless of the architects' wishes.
+Given this command-line argument, there is always a semantic dependency
+in the execution running from the \co{foo()} function's argument to its
+return value.
+So score one for at least some implementations' users.
+
+Please see Appendix~\ref{app:User Influence Over Language Semantics}
+for more examples of user control over semantics, including semantic
+dependencies.
+
+\emph{A definition of ``execution'' drawn strictly from the standard will
+be at best an approximation to a definition that is useful in practice.}
+
+\subsubsection{So What Exactly is an Execution, Anyway?}
+\label{sec:So What Exactly is an Execution, Anyway?}
+
+Given all of these complications in C++ executions, and given that
+semantic dependencies must be defined in terms of executions, it makes
+sense to lower our sights from full generality.
+We therefore restrict ourselves to a specific context:
+traditional compilers.
+It will then turn out to be necessary to take the notion of semantic
+dependence, and thus OOTA, to be compiler-relative rather than absolute.
+
+We will show that programs built using loose C++ compilers that treat all
+atomic operations as volatile can never exhibit OOTA, and the same is true
+for compilers obeying the weaker constraint that atomic accesses may not
+invented, provided the compiler does not perform multi-thread analysis.
+
+% @@@ LTO?  Does not appear to be relevant given new material.
+
+These restrictions permit us to bring to bear the real-world
+constraints introduced by the next section.
 

### 2024-02-16 ootavacuous.tex
Move additional material to cuttingroom.tex
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\section{To Be Moved Or Removed...}
-
-@@@ Review
-
-\subsubsection{Executions Can Be Merged}
-\label{sec:Executions Can Be Merged}
-
-Consider this example, similar to that shown in
-Section~\ref{sec:Semantic Dependencies Not Affected by if Statements},
-which appears to be an \co{if} statement:
-
-\begin{quote}
-\begin{verbatim}
-r1 = x;
-if (r1 > 0)
-    y = 42;
-else
-    y = 42;
-\end{verbatim}
-\end{quote}
-
-Because the stores executed on each branch of that \co{if} statement store
-identical values to identical addresses, one could equally well regard
-the two statements as performing two different stores or as performing
-(for all intents and purposes) a single store.
-Although software projects relying on ordering from load-to-store control
-dependencies and using a variety of compilers should should err on the
-side of caution by assuming there is no semantic
-dependency~\cite{Howells2009membartxt},
-the fact remains that reasonable C++ implementations might disagree as
-to whether this code represents one execution or two.
-It is the implementations's choice.
-
-A similar caution applies to loads:
-
-\begin{quote}
-\begin{verbatim}
-r1 = x;
-if (r1 > 42)
-    r2 = y;
-else
-    r2 = y;
-z = r2;
-\end{verbatim}
-\end{quote}
-
-The two loads from \co{y} are fungible, and so a C++ implementation
-could behave as if this \co{if} statement had instead been a single load
-from \co{y}.
-
-These examples demonstrate a key point: 
-\emph{In general, executions do not necessarily map one-to-one onto
-paths through the source code}.% \footnote{
-% 	Thanks to Alan Stern for the examples in this section and
-% 	in the preceding one.}
-
-\subsubsection{Executions Depend on Context}
-\label{sec:Executions Depend on Context}
-
-Consider the following example derived from that in
-Section~\ref{sec:Semantic Dependencies and Source Code}:
-
-\begin{quote}
-\begin{verbatim}
-r1 = x;
-// r1 = 0;
-z = y * r1;
-// z = y / r1;
-\end{verbatim}
-\end{quote}
-
-As written, there is a semantic dependency from \co{x} and \co{y}
-to \co{z}.
-But if the second line is uncommented, there is no semantic
-dependency.
-If the second line is left commented, but the last line is uncommented,
-then the compiler is free to backwards-propagate the potential
-divide-by-zero undefined behavior to precede the first assignment
-to \co{z}.
-This could prevent both assignments to \co{z} from executing, for example,
-due to a divide-by-zero exception.
-
-More subtle changes in context can also have an effect.
-For example:
-
-\begin{quote}
-\begin{verbatim}
-r1 = x;
-z = y * r1;
-z = 42;
-\end{verbatim}
-\end{quote}
-
-Given that there is no ordering, a C++ implementation could choose to
-act as if the first store to \co{z} was omitted, even if that store was
-a relaxed atomic store.
-But if \co{z} was marked volatile, both stores would need to be executed
-in strict accordance with the rules of the abstract machine.
-
-\emph{Any analysis of a given execution either must be fully informed
-of the enclosing context, or must make conservative assumptions about
-that context.}
-
-\subsubsection{Executions Can Be Nondeterministic}
-\label{sec:Executions Can Be Nondeterministic}
-
-Consider this example, similar to that in
-Section~\ref{sec:Abstract Executions},
-with \co{i} initially zero:
-
-\begin{quote}
-\begin{verbatim}
-int foo(int a, int b)
-{
-   return a / b;
-}
-
-y = foo(++i, ++i);
-x = y * z;
-\end{verbatim}
-\end{quote}
-
-Because early C~implementers could not come to agreement, standard
-does not specify the order of evaluation of function arguments, so the
-resulting value of \co{y} might be zero ($\frac{1}{2}$ truncated) or two
-($\frac{2}{1}$).
-In the former case, there is no semantic dependency from \co{z} to
-\co{x}, but in the latter case there is.
-Thus, semantic dependencies are not just a function of a particular
-execution through the source code, but also of arbitrary choices made
-by the C++ implementation.\footnote{
-	Again, thanks to Peter Sewell for pointing out this possibility.}
-
-Note well that C++ implementations can and do act as if portions of
-the source code was duplicated in order to apply specialization
-optimizations, in which case any non-determinism might also be
-a function of a given execution.
-
-Many projects' coding guidelines prohibit side effects in expressions,
-their goal being to obtain portable code whose behavior does not depend
-on arbitrary choices on the part of the implementation.
-However, the above code nevertheless conforms to the standard.
-Note that side effects in expressions lacking sequence points is
-undefined behavior, however, there are proposals to define it in the
-C++~\cite{GabrielDosReis2016P0145r3} and
-C~\cite{AlexCeleste2023N3203}
-languages, allowing implementations to nevertheless reorder
-evaluation when permitted by the as-if rule.
-
-Even though the standard is non-deterministic, \emph{compiler-based C++
-implementions make their non-deterministic choices at compile time, so
-that the object code is deterministic, at least on a per-execution basis.}
-
-\subsubsection{Implementers and Users Influence the Definition of Execution}
-\label{sec:Implementers and Users Influence the Definition of Execution}
-
-The exact definition of a computer language is subject to some debate,
-with standards, implementations, and users all having some degree of
-influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
-and each of which is subject to change over time.
-Users  and implementers are of course wise to assume that their opinions
-might be overridden by those of the standard and, for the users,
-especially the implementation.
-However, consider the following code, similar to that shown in
-Section~\ref{sec:Abstract Executions}:
-
-\begin{quote}
-\begin{verbatim}
-void foo(int x, char c)
-{
-    return x * (c >= 0);
-}
-\end{verbatim}
-\end{quote}
-
-The standard does not specify whether or not type \co{char} is signed,
-leaving that choice to the implementations, which follow the lead
-of the architects of a given CPU family.
-Therefore, the code above has a semantic dependency in executions running
-on systems choosing signed \co{char}, but not those running on systems
-choosing unsigned \co{char}.
-So score one for the implementers, albeit by conscious choice on the
-part of the standards committee.
-
-Except that GCC provides the \co{-funsigned-char} command-line
-argument that causes this implementation to treat variables of
-type \co{char} as unsigned, regardless of the architects' wishes.
-Given this command-line argument, there is always a semantic dependency
-in the execution running from the \co{foo()} function's argument to its
-return value.
-So score one for at least some implementations' users.
-
-Please see Appendix~\ref{app:User Influence Over Language Semantics}
-for more examples of user control over semantics, including semantic
-dependencies.
-
-\emph{A definition of ``execution'' drawn strictly from the standard will
-be at best an approximation to a definition that is useful in practice.}
-
-\subsubsection{So What Exactly is an Execution, Anyway?}
-\label{sec:So What Exactly is an Execution, Anyway?}
-
-Given all of these complications in C++ executions, and given that
-semantic dependencies must be defined in terms of executions, it makes
-sense to lower our sights from full generality.
-We therefore restrict ourselves to a specific context:
-traditional compilers.
-It will then turn out to be necessary to take the notion of semantic
-dependence, and thus OOTA, to be compiler-relative rather than absolute.
-
-We will show that programs built using loose C++ compilers that treat all
-atomic operations as volatile can never exhibit OOTA, and the same is true
-for compilers obeying the weaker constraint that atomic accesses may not
-invented, provided the compiler does not perform multi-thread analysis.
-
-% @@@ LTO?  Does not appear to be relevant given new material.
-
-These restrictions permit us to bring to bear the real-world
-constraints introduced by the next section.
+Despite these outstanding issues, our definition of semantic dependency
+allows us to bring to bear the real-world constraints outlined in the
+next section.
 

### 2024-02-16 ootavacuous.tex
Adjust roadmap in Introduction
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 then combines hardware dependencies and instruction ordering to arrive
-at the fundamental property of semantic dependencies.
+at the fundamental property of semantic dependencies, which
+Section~\ref{sec:A Definition of Semantic Dependency} uses to
+arrive at a definition of the term ``semantic dependency''.
 
@@
 Finally,
+Section~\ref{sec:Future Directions}
+lists a few possible future directions for this work and
 Section~\ref{sec:Conclusion}
 presents concluding remarks.
-% @@@ Future directions at some point.
 

### 2024-02-16 ootavacuous.tex
Add roadmap for "A Definition of Semantic Dependency"
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-@@@ Roadmap
+Section~\ref{sec:For Compilers Using Single-Thread Analysis}
+focuses on compilers that restrict their analysis to a single
+thread,
+Section~\ref{sec:For Compilers Using Global Analysis}
+slightly relaxes this single-thread restriction,
+Section~\ref{sec:Verifying the Fundamental Property}
+verifies the fundamantal property of semantic dependencies,
+and
+Section~\ref{sec:Outstanding Issues}
+discusses a general questions regarding our definitions.
 

### 2024-02-16 ootavacuous.tex
Update "Future Directions" section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-This paper focuses further on compilers that do only local per-thread
+This paper focuses primarily on compilers that do only local per-thread
 analysis.
-Further explorations could consider some classes of global analysis,
-for example, analyses that demonstrate that a given expression
-will always evaluate to a particular constant.
+Further explorations could consider additional classes of global analysis,
+for example, analyses that demonstrate that a given expression will
+always evaluate to a particular constant.
 JMM Causality Test Case 1 discussed in

### 2024-02-16 ootavacuous.tex
Reword hope for definition of semantic dependency
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
"Relative versus Absolute Dependency" s/compromising/restrictive/
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 No doubt many people will object to it for various reasons, but we
-hope it will provide a starting point for discussions that move the
-entire field forward.
+nevertheless hope it will help move the discussion forward.
 
@@
 compiler or implementation.
-This may strike some people as too compromising, in that a given piece
+This may strike some people as too restrictive, in that a given piece
 of code should either contain or not contain a semantic dependency

### 2024-02-17 ootavacuous.tex
Alan Stern February 16 feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{document}
-\title{D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem}
+\title{D3064R0: How to Avoid OOTA Without Really Trying}
 
@@
 compiler or implementation.
-This may strike some people as too restrictive, in that a given piece
-of code should either contain or not contain a semantic dependency
+This may strike some people as avoiding the real problem, in that a given
+piece of code should either contain or not contain a semantic dependency
 independent of the implementation used to run it.
@@
 This would be a valid transformation, and the resulting behavior would
-not count as OOTA because according to our definition, the
+not count as OOTA according to our definition because the
 dependencies in \co{thread1} and \co{thread2} would not be semantic.

### 2024-02-17 ootavacuous.tex
More Alan Stern February 16 feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 compiler or implementation.
-This may strike some people as avoiding the real problem, in that a given
-piece of code should either contain or not contain a semantic dependency
-independent of the implementation used to run it.
+This may strike some people as wishy-washy and as avoiding the real
+problem, in that a given piece of code should either contain or not
+contain a semantic dependency independent of the implementation used to
+run it.
 

### 2024-02-17 ootavacuous.tex
Update authors and date
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \author{
-Paul E.~McKenney\\\email{paulmck@kernel.org} \and
-Michael Wong\\\email{fraggamuffin@gmail.com} \and
-Maged Michael\\\email{maged.michael@gmail.com} \and
-Alan Stern\\\email{stern@rowland.harvard.edu} \and
+Paul E.~McKenney\\{\small \email{paulmck@kernel.org}} \and
+Alan Stern\\{\small \email{stern@rowland.harvard.edu}} \and
+Michael Wong\\{\small \email{fraggamuffin@gmail.com}} \and
+Maged Michael\\{\small \email{maged.michael@gmail.com}} \and
 The Indefatigible TBD
 }
-\date{January 7, 2024 (Pre-Tokyo)}
+\date{Feburary 16, 2024 (Post-Pre-Tokyo)}
 \maketitle{}

### 2024-02-17 cuttingroom.tex
Add footnote on Rust asms
Reported-by: Ralf Jung <post@ralfj.de>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 function that is invisible to the compiler, and a GCC goto asm that the
-compiler cannot prove falls through.
+compiler cannot prove falls through.\footnote{
+	Rust defines non-pure asms as potentially looping forever.
+	\url{https://github.com/rust-lang/reference/pull/1442}}
 

### 2024-02-18 ootavacuous.tex
Remove hyphen in "non-" prefixes
According to my style book, a hyphen should be used after a "non-"
prefix only when the following word is a proper name or when leaving
out the hyphen would promote an incorrect pronunciation.  So let's
remove lots of extraneous hyphens.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	restricted to single-threaded analysis in our current universe.
-	We extend this work to non-volatile atomics by defining
+	We extend this work to nonvolatile atomics by defining
 	quasi-volatile behavior that we conjecture that usable real-world
@@
 	Others cleverly avoid this issue by forbidding atomic
-	stores of non-constant values~\cite{MarkBatty2011cppmem}.}
+	stores of nonconstant values~\cite{MarkBatty2011cppmem}.}
 % @@@ Better herd7 C11 citation?
@@
 of environment: a benign one in which accesses to these objects work
-the same as non-volatile memory accesses, and a non-benign one in which
+the same as nonvolatile memory accesses, and a nonbenign one in which
 accesses to volatile objects are subject to outside interference and
 act more like I/O.
-In particular, when running in a non-benign environment a program's
+In particular, when running in a nonbenign environment a program's
 volatile loads can return unpredictable values.
-They don't necessarily read from stores (in contrast to non-volatile loads,
+They don't necessarily read from stores (in contrast to nonvolatile loads,
 which always must return the value of the store they read from).
@@
 Also, in these environments the rfe relation does not apply to volatile
-loads and hence the accesses in an OOTA cycle must be non-volatile.
+loads and hence the accesses in an OOTA cycle must be nonvolatile.
 
@@
 then presents required properties of volatile atomic operations, and
-also defines quasi-volatile properties for non-volatile atomic operations.
+also defines quasi-volatile properties for nonvolatile atomic operations.
 
@@
 accesses are considered to be a form of I/O when the program runs
-in a non-benign environment.
+in a nonbenign environment.
 But to be clear, these requirements apply to all environments.
@@
 By contrast, accesses to quasi-volatile objects are normal memory
-accesses, not subject to unpredictable interference in non-benign
+accesses, not subject to unpredictable interference in nonbenign
 environments (otherwise the program's behavior would be undefined).
@@
 However, we do now allow the possibility that the executions $B$ and
-$G$ take place in a non-benign environment.
+$G$ take place in a nonbenign environment.
 Aside from these minor adjustments, the definition remains unchanged.
@@
 Of course this notion has its own problems, including that it is
-extremely non-constructive and impossible to apply in practice.
+extremely nonconstructive and impossible to apply in practice.
 However it may be the best we can do with our current understanding
@@
 volatile, they are not subject to unspecified interference when the
-program runs in a non-benign environment.)
+program runs in a nonbenign environment.)
 
@@
 A final set of constraints is imposed by the C++ implementations
-themselves, some of which have resulted in non-determinisms within
+themselves, some of which have resulted in nondeterminisms within
 the standard.\footnote{
@@
 the speed of light is finite~\cite{OleRoemer1671SpeedOfLight}, and
-atoms are of non-zero size~\cite{JeanBaptistePerrin1923AtomSize}.
+atoms are of nonzero size~\cite{JeanBaptistePerrin1923AtomSize}.
 These fundamental laws of physics are empirical in nature, but
@@
 However, there has to date been no successful attempt to construct a
-zero-sized computing device capable of running non-trivial C++ programs.
+zero-sized computing device capable of running nontrivial C++ programs.
 
 This paper will therefore assume that any computing device capable of
-running C++ will be of non-zero size, which rules out formation of
+running C++ will be of nonzero size, which rules out formation of
 causal loops based on zero-size computing systems.
@@
 
-Therefore, non-speculative loads are prohibited from loading values from
+Therefore, nonspeculative loads are prohibited from loading values from
 speculative stores.
@@
 As a special case of this prohibition, a value stored speculatively must
-not be exposed to non-speculative loads from other threads.
+not be exposed to nonspeculative loads from other threads.
 In the further special case of multiple hardware threads sharing a core,
@@
 Applying this to Listing~\ref{lst:OOTA Cycle}, because values from
-speculative stores cannot be visible to non-speculative loads, lines~10
+speculative stores cannot be visible to nonspeculative loads, lines~10
 and~12 cannot be committed until after the load in line~8 completes
-and the conditional on line~9 is evaluated non-speculatively.\footnote{
+and the conditional on line~9 is evaluated nonspeculatively.\footnote{
 	Alternatively, and as noted earlier, one could imagine that both
@@
 Similarly, lines~18 and~20 cannot be committed until after lines~16
-and~17 complete non-speculatively.
+and~17 complete nonspeculatively.
 
@@
 all loads in that set must have been confirmed by corresponding
-non-speculative stores.
+nonspeculative stores.
 
@@
 
-Although C++ non-volatile accesses to atomic objects are not observable
+Although C++ nonvolatile accesses to atomic objects are not observable
 behavior, any implementation that restricts its code analysis to a
@@
 the extent that a volatile relaxed atomic store would, but it does add
-significant constrants over those related to non-volatile non-atomic
+significant constrants over those related to nonvolatile non-atomic
 stores.
 
-For example, non-volatile relaxed atomic accesses are subject to the
+For example, nonvolatile relaxed atomic accesses are subject to the
 memory model.
@@
 when redundant atomic stores may safely be omitted and on fusing of
-non-volatile atomic loads from adjacent objects.
+nonvolatile atomic loads from adjacent objects.
 
@@
 \paragraph{Omitted Redundant Atomic Stores}
-In contrast, a pair of back-to-back non-volatile atomic stores to \co{x}
+In contrast, a pair of back-to-back nonvolatile atomic stores to \co{x}
 might be executed such that no other thread accesses \co{x}
@@
 code cannot create an OOTA cycle because any valid execution (OOTA or not)
-of the program with the first of a back-to-back pair of non-volatile
+of the program with the first of a back-to-back pair of nonvolatile
 relaxed atomic stores omitted is essentially a valid execution of the
@@
 
-Suppose there are a pair of adjacent non-volatile atomic objects,
+Suppose there are a pair of adjacent nonvolatile atomic objects,
 and that these objects taken together are machine-word aligned
@@
 the code and the time spent actually executing it.
-One type of optimization is the identification of non-semantic dependencies
+One type of optimization is the identification of nonsemantic dependencies
 and the replacement of them with simpler computations, up to and including
@@
 The second row depicts the case where the implementation misclassifies
-the dependency as non-semantic, which will result in an incorrect
+the dependency as nonsemantic, which will result in an incorrect
 execution.
@@
 The fourth and final row depicts the case where the implementation
-correctly classifies the dependency as non-semantic, in which case
+correctly classifies the dependency as nonsemantic, in which case
 the execution is both correct and atemporal.
@@
 C++ implementations undertake a wide variety of optimizations, in
-some cases taking advantage of non-determinism and in others applying
+some cases taking advantage of nondeterminism and in others applying
 the results of code analysis to produce faster code.
@@
 
-In C++ implementations using compilers, non-determinism of a given
+In C++ implementations using compilers, nondeterminism of a given
 (post-optimization) code path is ``frozen'' into the resulting binary,
@@
 Other C++ implementations (for example, those using interpreters or JITs)
-might need to directly model any non-determinism.
+might need to directly model any nondeterminism.
 This latter situation might motivate additional C++ projects to ban
-non-determinism in favor of portable code.
+nondeterminism in favor of portable code.
 
@@
 not.
-In the former case on a real-world implementation, a non-negative interval
+In the former case on a real-world implementation, a nonnegative interval
 of time will elapse between the load and the store, and in the latter
@@
 might show that one of the dependencies that the implementation treated
-as a semantic dependency was in fact non-semantic.
+as a semantic dependency was in fact nonsemantic.
 And this is a belt-and-suspenders situation:
 The corresponding load-store pair cannot be a member of an OOTA cycle,
-and a non-negative interval of time will elapse between the load and
+and a nonnegative interval of time will elapse between the load and
 the store.
@@
 Such tools can simply use that implementation's classification of
-dependencies as semantic on the one hand or non-semantic on the
+dependencies as semantic on the one hand or nonsemantic on the
 other, whether by working with that implementation's intermediate
@@
 to a specific load in the abstract machine are incapable of producing
-OOTA cycles in UB-free programs using non-volatile as well
+OOTA cycles in UB-free programs using nonvolatile as well
 as volatile atomics.
@@
 Implementations that can prove that the value of \co{x} is always
-non-negative can also prove that there is no semantic dependency between
+nonnegative can also prove that there is no semantic dependency between
 \co{P0()}'s load from \co{x} on line~8 to its store to \co{y} on line~10,
@@
 having no cycle at all, and other implementations retaining the
-non-semantic dependency.
+nonsemantic dependency.
 
@@
 If \co{P0()}'s loads from \co{x} are fused, there is no cycle,
-otherwise the preserved non-semantic dependency prevents an
+otherwise the preserved nonsemantic dependency prevents an
 OOTA cycle from forming on correct real-world implementations.
@@
 
-Less-omniscient implementations will preserve the non-semantic
+Less-omniscient implementations will preserve the nonsemantic
 dependency, preventing any OOTA loop from forming in a correct real-world C++
@@
 
-A less-capable implementation would leave the non-semantic dependency
+A less-capable implementation would leave the nonsemantic dependency
 in place, also preventing any OOTA cycle in a correct real-world C++
@@
 line~13, line~18 (so \co{r3 == 1)}.
-The sum of \co{r2} and \co{r3} is non-zero, justifying line~21's
+The sum of \co{r2} and \co{r3} is nonzero, justifying line~21's
 store to \co{x}.
@@
 and finally line~24 (so \co{r3 == 1}).
-The sum of \co{r2} and \co{r3} is non-zero, justifying line~27's
+The sum of \co{r2} and \co{r3} is nonzero, justifying line~27's
 store to \co{u}.
@@
 One possible reaction to this situation is to declare that useful
-real-world C++ implementations should never duplicate non-volatile
+real-world C++ implementations should never duplicate nonvolatile
 relaxed atomic loads.
@@
 \item	Implementations should not invent uses for values from
-	non-volatile relaxed atomic loads.
+	nonvolatile relaxed atomic loads.
 	If it turns out that there are useful optimizations that invent
@@
 	Some might argue that inventing comparisons between a pair of
-	non-volatile relaxed atomic loads from the same object should
+	nonvolatile relaxed atomic loads from the same object should
 	be permitted only if the two loads could be reordered to be
@@
 \item	Any time an implementation might be tempted to invent uses from
-	values from non-volatile relaxed atomic loads from the same
+	values from nonvolatile relaxed atomic loads from the same
 	object, the loads should instead be fused into a single load.
@@
 	spaced loads return different values on the other.
-\item	Implementations that invent non-volatile atomic loads or invent
-	new uses for non-volatile atomic loads can adversely affect
+\item	Implementations that invent nonvolatile atomic loads or invent
+	new uses for nonvolatile atomic loads can adversely affect
 	observed behavior, for example, by introducing error into code
@@
 
-But if \co{x} is a non-volatile atomic, then there are no worries about
+But if \co{x} is a nonvolatile atomic, then there are no worries about
 data races.
@@
 In any related execution where the value loaded from \co{y} remains zero,
-non-zero values loaded from \co{x} cannot affect the value stored by
+nonzero values loaded from \co{x} cannot affect the value stored by
 line~11 to \co{z}.
-Only executions which load different (that is, non-zero) values from
+Only executions which load different (that is, nonzero) values from
 both \co{x} and \co{y} can cause the value stored to \co{z} to change
-from zero to a non-zero value.
+from zero to a nonzero value.
 
@@
 equivalent\footnote{
-	But equivalent only because the atomics are all non-volatile!}
+	But equivalent only because the atomics are all nonvolatile!}
 program shown in
@@
 A similar transformation would load \co{x} only in executions where the
-load from \co{y} returned non-zero, which similarly emphasizes the lack
+load from \co{y} returned nonzero, which similarly emphasizes the lack
 of any semantic dependency between \co{x} and \co{z}.
@@
 Starting with the intuitive execution where the values loaded from \co{x}
-and \co{y} on lines~9 and~10 are the value 1 (that is, non-zero), executions
+and \co{y} on lines~9 and~10 are the value 1 (that is, nonzero), executions
 that differ only in the value loaded from either \co{x} or \co{y} will
@@
 and~\ref{lst:OOTA Multiplication Example, Initial Value 1}
-each have but one OOTA cycle, and an execution having non-zero non-one
+each have but one OOTA cycle, and an execution having nonzero non-one
 initial values could have at most two OOTA cycles, all of which are
@@
 The last semantic dependency does not imply the first, because given
-that \co{w} becomes non-zero, a change in \co{x} further changes
+that \co{w} becomes nonzero, a change in \co{x} further changes
 the value stored to \co{z}.
@@
 
-One might ask why semantic dependencies can be non-trivial when hardware
+One might ask why semantic dependencies can be nontrivial when hardware
 address, control, and data dependencies are much more straightforward.
@@
 
-Because both \co{x} and \co{y} are non-volatile, the compiler is
+Because both \co{x} and \co{y} are nonvolatile, the compiler is
 free to assume that the value loaded by line~4 will be some value
@@
 The store-to-store latency is measured from the end of the ``winning''
-store to the beginning of the non-winning store that started latest.
+store to the beginning of the nonwinning store that started latest.
 Every data point on the negative x-axis represents a group of runs
@@
 
-In other words, user preference can exert a non-trivial influence over
+In other words, user preference can exert a nontrivial influence over
 language semantics, and in particular can affect some aspects of the
@@
 Martin Uecker contributed valuable insights on backwards-propagating UB.
-Gonzalo Brito Gadeschi shared an alternative way of handling non-volatile
+Gonzalo Brito Gadeschi shared an alternative way of handling nonvolatile
 atomic operations.

### 2024-02-18 ootavacuous.tex
More minor edits
Here's a number of small changes to the text.  They are intended to
improve grammar, style, and consistency.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Some workers recommend avoiding OOTA by ordering prior relaxed
-loads before subsequent relaxed
+Some workers recommend avoiding OOTA by forcing prior relaxed
+loads to be ordered before subsequent relaxed
 stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
@@
 
-As long as \co{z} is zero, and all else being equal, changes in the
+As long as \co{z} is zero, changes in the
 value of \co{y} will not cause a change in the value stored to \co{x}.
@@
 both \co{y} and \co{z} are zero.
-Then changes to either \co{y} or \co{z} (again, all
-else being equal) will not cause a change in the value stored to \co{x}.
+Then changes to either \co{y} or \co{z}
+will not cause a change in the value stored to \co{x}.
 In other words, in this execution there is no semantic dependency
@@
 \begin{quote}
-	There are stores
-	$W_i$~($i = 0$, \ldots,~$m$),
-	where $W_i$ semantically depends on loads
-	$\{R_{i,j}\}$~($j = 0$, \ldots,~$n_i$),
+	There are stores $W_0,$ \ldots,~$W_m$,
+	where each $W_i$ semantically depends on a set of loads
+	$\{R_{i,0},$ \ldots,~$R_{i,n_i}\}$,
 	such that each $R_{i,j}$ reads from one of the $W_k$
@@
 always equal to \co{z} at this point then there is no semantic dependency;
-the implementation can act as if the statement were simply ``\co{x = 0;}''.
+the implementation can act as if the statement were simply ``\co{x = 0}''.
 We leave aside the question of how the implementation would know this,
@@
 The answer isn't immediately clear.
-If the other arm of the \co{if} is taken then the same store of 42 to
+If the other arm of the \co{if} is taken then a store of the same value 42 to
 \co{y} occurs, but 53 is written before it.
@@
 according to the execution and even the implementation.
-This raises some questions, of which the first is:
+This raises several questions, of which the first is:
 What exactly is an execution?
@@
 
-When a given computer executes the machine code in a file, the outcome
+The outcome when a given computer executes the machine code in a file
 has historically been much better defined than the executions of the
@@
 program is entirely determined by the values obtained by the various
-load-from-memory instructions.
+memory-load instructions.
+(We count read-modify-write instructions as both memory loads and
+memory stores.)
 
@@
 \begin{verbatim}
- 1  int r1 = (x != 0);
- 2  int r2 = (y != 0);
- 3  z = (r1 == r2);
+   int r1 = (x != 0);
+   int r2 = (y != 0);
+   z = (r1 == r2);
 \end{verbatim}
@@
 in a nonbenign environment.
-But to be clear, these requirements apply to all environments.
+But to be clear, these requirements apply in all environments.
 
@@
 	stored in a status register).
-\item	A conditional move instruction would have as inputs the
+\item	A conditional-move instruction would have as inputs the
 	condition-code bits to test, the source register value, and the
@@
 	in the target register.
-\item	A conditional or computed branch instruction would have as inputs
+\item	A conditional- or computed-branch instruction would have as inputs
 	the condition-code bits to test, the address of the following
@@
 the instruction's results.
-For instance, a conditional move instruction wouldn't need to wait for
+For instance, a conditional-move instruction wouldn't need to wait for
 the source register input to be determined once it knew that the
@@
 be executed at all.
-Therefore instructions following a conditional or computed branch
+Therefore instructions following a conditional- or computed-branch
 instruction in an execution must commit after the branch, and hence
@@
 Note that this formulation makes sense only for implementations in
-which all atomic objects are treated as quasi volatile.
+which all atomic objects are treated as volatile or quasi volatile.
 \begin{quote}
-Let $W$ be a store which, in some abstract execution $A$, semantically
-depends on loads $\{R_0$, \ldots,~$R_n\}$, and suppose that $W$ is not
+Let $W$ be a store which semantically depends on loads
+$\{R_0$, \ldots,~$R_n\}$ in some abstract execution $A$,
+and suppose that $W$ is not
 omitted in some hardware execution $H$ realizing $A$.
@@
 thread $T$, and let $H$ be a hardware execution realizing $A$.
-Let $W$ be an store in $T$ to an atomic object and let $\{R_i \mid i =
-0, \ldots, n\}$ be a set of loads in $T$ from atomic objects on which
+Let $W$ in $T$ be a store to an atomic object, and let $\{R_0,$
+\ldots,~$R_n\}$ in $T$ be a set of loads from atomic objects on which
 $W$ might or might not depend.
@@
 the same value as they do in $B$ (this is part of our interpretation
-of ``all other things being equal'').
+of ``all else being equal'').
 It follows that the two abstract executions of $T$ will be identical
@@
 after one of the $R'_i$ loads (this is the remaining part of our
-interpretation of ``all other things being equal''.)
+interpretation of ``all else being equal''.)
 
@@
 earlier proposal of matching up stores by the order they occur.
-Let \co{y} be the atomic object that $W'$ stores to, and suppose $W'$
-is the $N$th store to \co{y} within the divergent portion of $H$.
-Then $X'$ will be the $N$th store to \co{y} in the divergent portion
+Let $y$ be the atomic object that $W'$ stores to, and suppose $W'$
+is the $N$th store to $y$ within the divergent portion of $H$.
+Then $X'$ will be the $N$th store to $y$ in the divergent portion
 of $G$, if such a store exists.
@@
 is that $X'$ act differently from $W'$: it doesn't exist, it stores
-a different value, or it stores to an object other than \co{y}.
+a different value, or it stores to an object other than $y$.
 
@@
 the same for a given program as for a ``volatilized'' form of the
-program in which all atomic objects are defined to be volatile.
+program in which all the atomic objects are defined to be volatile.
 
@@
 same regardless of the program it belongs to, the program $Q$ in the
-earlier definition (of which $B$ and $G$ are executions) must now be
+earlier definition (of which $B$ and $G$ are executions) must be
 $P$ or its volatilized form.
@@
 the two executions it must be ordered after one of the $R'_i$ loads.
-And then so must $W'$, because any instruction following a conditional
-or computed branch instruction is ordered after the branch.
+And then so must $W'$, because any instruction following a conditional-
+or computed-branch instruction is ordered after that branch.
 QED.
@@
 problem, in that a given piece of code should either contain or not
-contain a semantic dependency independent of the implementation used to
+contain a semantic dependency, independent of the implementation used to
 run it.
@@
 We can address this drawback by defining an \emph{absolute semantic
-dependency} as one that is present relative to any valid C++
-implementation, past, present, or future.
+dependency} as one that is present relative to any valid loose C++
+implementation, past, present, or future, real or imagined.
 Of course this notion has its own problems, including that it is
@@
 semantic dependencies are absolute).
-This is simply because a semantic dependency that is absolute is {\it
+This is simply because an absolute semantic dependency is {\it
 a fortiori\/} a semantic dependency relative to the compiler in use.
@@
 But this transformed program has no other executions; every time it
-runs it will write 42 to both \co{x} and \co{y}.
+runs it will store 42 to both \co{x} and \co{y}.
 (Keep in mind also that since the atomic objects are not treated as
@@
 different values when running in different programs because the
-variable \co{x} may be allocated at differing addresses in those
+object \co{x} may be allocated at differing addresses in those
 programs.
@@
 However, a single-thread analysis compiler can translate this into the
-machine code equivalent of:
+machine-code equivalent of:
 \begin{quote}
@@
  3 void thread1() {
- 4   int r1 = x.load(memory_order::relaxed);
- 5   y.store(r1, memory_order::relaxed);
+ 4   int r1 = x.load(memory_order_relaxed);
+ 5   y.store(r1, memory_order_relaxed);
  6 }

### 2024-02-20 cuttingroom.tex
Apply Feb 18 Alan Stern feedback, part 2
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 This paper focuses instead on OOTA cycles.
+
+In happy contrast to the situation at the start of the C++11 memory-model
+effort, heavily used CPU families now have accurate and executable formal
+memory models, all of which prohibit the causal cycles required to form
+OOTA cycles.
+Even more important, there is clarity on exactly what hardware can and
+can not do in the course of speculative execution.
+Finally, computation is based on instructions, and instructions require
+finite time to execute, even when executed speculatively.

### 2024-02-20 ootavacuous.tex
Apply Feb 18 Alan Stern feedback, part 1
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Apply Feb 18 Alan Stern feedback, part 2
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-This paper shows that C++ implementations subject to reasonable
-real-world constraints cannot exhibit OOTA cycles.
+This paper shows that compiler-based C++ implementations subject to
+reasonable real-world constraints cannot exhibit OOTA cycles.
 Section~\ref{sec:Background}
@@
 
-In happy contrast to the situation at the start of the C++11 memory-model
-effort, heavily used CPU families now have accurate and executable formal
-memory models, all of which prohibit the causal cycles required to form
-OOTA cycles.
-Even more important, there is clarity on exactly what hardware can and
-can not do in the course of speculative execution.
-Finally, computation is based on instructions, and instructions require
-finite time to execute, even when executed speculatively.
-
-\begin{listing}[tbp]
-@@ DisplayLitmus litmus/speculative-store.litmus @@
-\caption{Speculated Store and Non-Speculated Load}
-\label{lst:Speculated Store and Non-Speculated Load}
-\end{listing}
-
+On real systems, instructions require time to execute, even
+when executed speculatively.
 For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
@@
 
+\begin{listing}[tbp]
+@@ DisplayLitmus litmus/speculative-store.litmus @@
+\caption{Speculated Store and Non-Speculated Load}
+\label{lst:Speculated Store and Non-Speculated Load}
+\end{listing}
+
 Therefore, nonspeculative loads are prohibited from loading values from
@@
 
-Note that if both legs of the \co{if} statement spanning lines~9-12 were
-to store the same value, the implementation would be within its rights
-to hoist the store out of that \co{if} statement.
-However, in this case, the two stores on lines~10 and~12 are storing
-distinct values, preventing the hoisting.
-
-Note also that some architectures provide conditional-move instructions,
-which allows line~9 to avoid emitting a conditional branch.
-However, such architectures would still treat a store of the
-target of the conditional-move instructions as speculative until
-such time as the conditional moves (and any loads that they depend
-on) complete.
+One could imagine a conditional-store instruction, similar to existing
+conditional-move instructions, but with condition controlling a store
+to memory rather than a move to a register.
+The condition would control whether or not the store was executed,
+and prior conditional-move instructions could control the value
+stored and the address stored to.
+With or without prior conditional-move instructions, but assuming a
+compiler doing single-thread analysis, a conditional-store instruction
+permits generation of code corresponding to lines~10 and~11 of
+Listing~\ref{lst:Causality Test Case 2}
+on page~\pageref{lst:Causality Test Case 2}
+without use of a conditional branch.
+This means that the control dependency would extend from the loads on
+lines~8 and~9 to the store on line~11 and nowhere else, potentially
+enabling more hardware optimizations than would be permitted for code
+relying on conditional branches.
+Nevertheless, the conditional store could not be committed while any
+the instructions it depends on were still being speculatively executed.
 
@@
 The key point is that a store having a semantic dependency on a given
-set of loads must execute after the completion of the last of those
+set of loads must execute after the completion of at least some of those
 loads in global time.

### 2024-02-21 cuttingroom.tex
Drop "Laws of Physics" section
But it lives on in cuttingroom.tex.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 finite time to execute, even when executed speculatively.
+
+\subsection{Laws of Physics}
+\label{sec:Laws of Physics}
+
+Evidence to date suggests that the universe is
+causal~\cite{Plato360BC-causality},
+the speed of light is finite~\cite{OleRoemer1671SpeedOfLight}, and
+atoms are of nonzero size~\cite{JeanBaptistePerrin1923AtomSize}.
+These fundamental laws of physics are empirical in nature, but
+are backed by a great many observations extending back to a time
+preceding electronic computers, let alone computer languages that
+support concurrency.
+
+The relationship between these three laws of physics and OOTA
+cycles is described in the following sections.
+
+\subsubsection{Causality}
+\label{sec:Causality}
+
+Plato's ``Timaeus'' notwithstanding, there has been much recent debate
+as to whether the universe is in fact causal with cause necessarily
+preceding effect.
+However, no one has managed to construct a real time-travel machine
+or any other apparatus that might result in effect preceding cause.
+% as discussed at length in Appendix~\ref{sec:Debates on Causality} and 
+% the URL that it provides.
+
+This paper will therefore assume that effect cannot precede cause in
+any real-world computing system.
+
+\subsubsection{Finite Speed of Light}
+\label{sec:Finite Speed of Light}
+
+If the speed of light was infinite, then information could be transmitted
+at infinite speed, arriving at any destination, no matter how remote,
+in zero time.
+This situation would permit a load to return the value stored by some
+other thread at the exact instant that the store executed, which
+would in turn remove propagation delay as a factor preventing
+causal loops such as OOTA cycles from forming.
+
+However, all macroscopic measurements to date have shown the speed of
+light to be finite.
+This paper will therefore assume that the speed of light is finite, and
+that the maximum speed at which information can propagate is also finite.
+
+As long as empirical evidence continues to validate this assumption, it
+is not possible to form causal loops based on infinite-speed propagation
+of information.
+
+\subsubsection{Non-Zero Sized Atoms}
+\label{sec:Non-Zero Sized Atoms}
+
+If atoms were of zero size, a zero-sized computer might be constructed.
+In such a computer, even given finite speed of light, information could
+propagate among computing elements in zero time.
+This situation would again permit a load to return the value stored by
+some other thread at the exact instant that the store executed, which
+would again in turn remove propagation delay as a factor preventing
+causal loops such as OOTA cycles from forming.
+
+However, there has to date been no successful attempt to construct a
+zero-sized computing device capable of running nontrivial C++ programs.
+
+This paper will therefore assume that any computing device capable of
+running C++ will be of nonzero size, which rules out formation of
+causal loops based on zero-size computing systems.
+
+\subsubsection{Laws of Physics: Consequences}
+\label{sec:Laws of Physics: Consequences}
+
+These three laws of physics mean that if one thread loads the value
+stored by some other thread, that load must have occurred later
+in global time than did the store.
+In other words, the reads-from relation is temporal in nature.
+
+In contrast, the C++11 modification-order relation is atemporal,
+because computing systems can and observably do determine the
+modification order of a set of concurrent stores long after the fact
+(see Appendix~\ref{app:Inter-Process Communications}).
+This means that the store that executed earlier in global time
+might appear later in the stored-to object's modification
+order~\cite{McKenney20xxParallelProgramming}.
+The atemporal nature of C++11 modification order is due to the hardware
+store-buffer optimizations used by modern multicore systems.
+
+This atemporality is also why cycles are not considered to have
+OOTA cycles when at least one link from one thread to the next
+involves both threads storing to the same object.
+Similarly, cycles having at least link where one thread loads from an
+object and the next thread stores to that object are also not considered
+to be OOTA cycles.
+
+Current programming-language abstract machines do not capture these
+temporal constraints, which makes it more difficult for them to
+efficiently rule out OOTA behavior.

### 2024-02-21 ootavacuous.tex
Drop "Laws of Physics" section
But it lives on in cuttingroom.tex.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 any tooling that exists separately for these C++ implementations.
-
-\subsection{Laws of Physics}
-\label{sec:Laws of Physics}
-
-Evidence to date suggests that the universe is
-causal~\cite{Plato360BC-causality},
-the speed of light is finite~\cite{OleRoemer1671SpeedOfLight}, and
-atoms are of nonzero size~\cite{JeanBaptistePerrin1923AtomSize}.
-These fundamental laws of physics are empirical in nature, but
-are backed by a great many observations extending back to a time
-preceding electronic computers, let alone computer languages that
-support concurrency.
-
-The relationship between these three laws of physics and OOTA
-cycles is described in the following sections.
-
-\subsubsection{Causality}
-\label{sec:Causality}
-
-Plato's ``Timaeus'' notwithstanding, there has been much recent debate
-as to whether the universe is in fact causal with cause necessarily
-preceding effect.
-However, no one has managed to construct a real time-travel machine
-or any other apparatus that might result in effect preceding cause.
-% as discussed at length in Appendix~\ref{sec:Debates on Causality} and 
-% the URL that it provides.
-
-This paper will therefore assume that effect cannot precede cause in
-any real-world computing system.
-
-\subsubsection{Finite Speed of Light}
-\label{sec:Finite Speed of Light}
-
-If the speed of light was infinite, then information could be transmitted
-at infinite speed, arriving at any destination, no matter how remote,
-in zero time.
-This situation would permit a load to return the value stored by some
-other thread at the exact instant that the store executed, which
-would in turn remove propagation delay as a factor preventing
-causal loops such as OOTA cycles from forming.
-
-However, all macroscopic measurements to date have shown the speed of
-light to be finite.
-This paper will therefore assume that the speed of light is finite, and
-that the maximum speed at which information can propagate is also finite.
-
-As long as empirical evidence continues to validate this assumption, it
-is not possible to form causal loops based on infinite-speed propagation
-of information.
-
-\subsubsection{Non-Zero Sized Atoms}
-\label{sec:Non-Zero Sized Atoms}
-
-If atoms were of zero size, a zero-sized computer might be constructed.
-In such a computer, even given finite speed of light, information could
-propagate among computing elements in zero time.
-This situation would again permit a load to return the value stored by
-some other thread at the exact instant that the store executed, which
-would again in turn remove propagation delay as a factor preventing
-causal loops such as OOTA cycles from forming.
-
-However, there has to date been no successful attempt to construct a
-zero-sized computing device capable of running nontrivial C++ programs.
-
-This paper will therefore assume that any computing device capable of
-running C++ will be of nonzero size, which rules out formation of
-causal loops based on zero-size computing systems.
-
-\subsubsection{Laws of Physics: Consequences}
-\label{sec:Laws of Physics: Consequences}
-
-These three laws of physics mean that if one thread loads the value
-stored by some other thread, that load must have occurred later
-in global time than did the store.
-In other words, the reads-from relation is temporal in nature.
-
-In contrast, the C++11 modification-order relation is atemporal,
-because computing systems can and observably do determine the
-modification order of a set of concurrent stores long after the fact
-(see Appendix~\ref{app:Inter-Process Communications}).
-This means that the store that executed earlier in global time
-might appear later in the stored-to object's modification
-order~\cite{McKenney20xxParallelProgramming}.
-The atemporal nature of C++11 modification order is due to the hardware
-store-buffer optimizations used by modern multicore systems.
-
-This atemporality is also why cycles are not considered to have
-OOTA cycles when at least one link from one thread to the next
-involves both threads storing to the same object.
-Similarly, cycles having at least link where one thread loads from an
-object and the next thread stores to that object are also not considered
-to be OOTA cycles.
-
-Current programming-language abstract machines do not capture these
-temporal constraints, which makes it more difficult for them to
-efficiently rule out OOTA behavior.
 

### 2024-02-21 ootavacuous.tex
Tie Section 8.2 discussion to quasi-volatile properties
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-The following sections discuss these constraints, starting with the
-laws of physics, continuing with hardware constraints, continuing
+The following sections discuss these constraints, starting with
+with hardware constraints, continuing
 further with constraints imposed by the standard, and culminating
@@
 memory model.
-In particular, implementations must not behave as if the source code has
+The ``order-preserving'' bullet from
+Section~\ref{sec:Volatile and Quasi Volatile Accesses}
+prohibits reordering only of quasi volatile accesses to the same object,
+but such reordering is in any case directly forbidden by the standard.
+Referring also to the ``onto'' bullet from
+Section~\ref{sec:Volatile and Quasi Volatile Accesses},
+implementations must not behave as if the source code has
 additional:
@@
 \item	Duplicated atomic stores.
+\item	Invented atomic loads.
 \end{enumerate}
@@
 nonvolatile atomic loads from adjacent objects.
+These topics correspond to the ``one-to-one'' and ``total'' bullets of
+Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
 
@@
 
-Duplicating atomic stores can also introduce new and undesirable
-behaviors.
+Duplicating atomic stores is a special case of inventing them, but it is
+worth illustrating the fact that duplicating stores can also introduce
+new and undesirable behaviors.
 To see this, consider Listing~\ref{lst:Example Duplicated Store}, a
@@
 are no other accesses to \co{x} at the time of the invented store.
+
+\paragraph{Invented Atomic Loads}
+Although one can argue that it is functionally correct to invent atomic
+loads as long as the loaded value is not used, doing so can result in
+excessive cache-miss overheads.
+
+\begin{listing}[tbp]
+\scriptsize
+\begin{verbatim}
+ 1 struct foo {
+ 2   int a;
+ 3   int b;
+ 4 };
+ 5 struct foo *globalfoop;
+ 6
+ 7 void bar()
+ 8 {
+ 9   int a;
+10   int b;
+11   struct foo *fp;
+12
+13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
+14   a = fp->a;
+15   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
+16   b = fp->b;
+17   do_something_with(a, b);
+18 }
+\end{verbatim}
+\caption{Do Not Duplicate Atomic Loads}
+\label{lst:Do Not Duplicate Atomic Loads}
+\end{listing}
+
+However, even duplicating an existing atomic load can lead to incorrect
+results.
+To see this, please refer to
+Listing{lst:Do Not Duplicate Atomic Loads}.
+Suppose that the load on line~13 were duplicated, for example, by
+uncommenting line~15.
+Then the values of \co{a} and \co{b} passed to line~17's call to
+\co{do_something_with()} might be from different instances of the
+\co{foo} structure, something that \co{do_something_with()} might
+not be prepared to deal with.
+
+Therefore, C++ implementations should not invent loads, and should
+not even duplicate them.
 

### 2024-02-21 cuttingroom.tex
Drop "C++ Implementation" section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 efficiently rule out OOTA behavior.
+
+\subsection{C++ Implementations}
+\label{sec:C++ Implementations}
+
+As noted earlier, this initial work focuses on C++ implementations based
+on compilers and computing hardware (including CPUs and GPGPUs).
+
+C++ implementations execute C++ programs, and use a range of techniques.
+For example, many implementations take steps to optimize the code to
+be executed.
+This optimization presents a trade-off between the time spent optimizing
+the code and the time spent actually executing it.
+One type of optimization is the identification of nonsemantic dependencies
+and the replacement of them with simpler computations, up to and including
+replacement of arbitrarily complex passages of code with constants.\footnote{
+	There are a great many other optimizations, but this paper
+	necessarily focuses on semantic dependencies.}
+
+This optimization thread may, roughly speaking, be divided into two
+activities, analysis of the code and classification of dependencies.
+Two sections on these topics are followed by sections on optimizations
+and on consequences.
+
+\subsubsection{Code Analysis}
+\label{sec:Code Analysis}
+
+As would be expected given the various optimization/execution tradeoffs,
+different C++ implementations carry out different degrees of analysis.
+
+First, at one extreme, ``simple implementations'' do absolutely no
+analysis beyond that absolutely required to correctly execute the program.
+
+Second, other implementations do local analysis.
+This analysis might be on a per-code-path basis, within a single function,
+within a single function after inlining, and so forth.
+
+Third, one could imagine implementations whose analysis is restricted to
+a thread.
+In this case, the analysis stops at loads to objects that were last
+stored to by other threads (``rfe''), stores to objects where the next
+and/or previous stores in that object's modification order were executed
+by other threads (``coe''), and loads from objects for which the value
+returned by a given load was overwritten by some other thread
+(``fre'').\footnote{
+	See Appendix~\ref{app:Inter-Process Communications} for more
+	information on these memory-reference inter-process communications
+	links.}
+
+Fourth, one could also imagine implementations whose analysis is
+completely unrestricted, up to and including the limiting case where
+execution is oracular in nature.
+Of course, if we are going to imagine omniscient analysis and oracles,
+we can also rely on them to locate any OOTA cycles that might appear
+in environments not constrained by the laws of physics, hardware
+limitations, and C++ implementation concerns.
+
+However, for more realistic implementations in the third category,
+analysis relevant to OOTA cycles begins with loads from objects that were
+last stored to by some other thread and ends with stores to objects that
+might be loaded by other threads.
+
+If additional inter-thread information is available, the corresponding
+constraints may be applied.
+For example, if an implementation proves that the values of a pair of
+objects are always equal at any time that a given thread might access
+them, that implementation could behave as if the that thread's code had
+substituted a load from one for a load from the other.
+
+Note that if all of the inter-thread relaxed accesses are volatile,
+as they are in the C language, then all of the inter-thread traffic is
+observed behavior.
+In this case, even omniscient oracular implementations are required
+to confine their analyses to within a single thread.
+
+The next section focuses on the key analysis step for OOTA cycles,
+namely dependency classification.
+
+\subsubsection{Dependency Classification}
+\label{sec:Dependency Classification}
+
+Carrying out this type of optimization requires that the implementation
+classify dependencies as maybe semantic or definitely not.
+Given the limited resources available, almost all implementations will
+choose to approximate sdep, which entails some misclassifications, with
+the consequences summarized in
+Table~\ref{tab:Semantic-Dependency Classification}.
+The first row of the table depicts the case where the implementation
+correctly identifies a semantic dependency, and thus exhibits temporal
+behavior for the execution at hand of the corresponding code.
+The second row depicts the case where the implementation misclassifies
+the dependency as nonsemantic, which will result in an incorrect
+execution.
+In this case, the proper course of action is to fix that bug, which is
+most likely a bug even in the absence of OOTA cycles.
+The third row depicts the case where the implementation misclassifies
+the dependency as semantic, thus missing a chance to optimize.
+The resulting execution is nevertheless correct, and is also temporal.
+The fourth and final row depicts the case where the implementation
+correctly classifies the dependency as nonsemantic, in which case
+the execution is both correct and atemporal.
+
+\begin{table}
+\centering
+\begin{tabular}{c|c|l}
+Actual Execution	& Implementation	& Result \\
+\hline
+sdep			& sdep			& Temporal \\
+\cline{2-3}
+			& $\neg$sdep		& Fix C++ bug! \\
+\hline
+$\neg$sde		& sdep			& Temporal \\
+\cline{2-3}
+			& $\neg$sdep		& Atemporal \\
+\end{tabular}
+\caption{Semantic-Dependency Classification}
+\label{tab:Semantic-Dependency Classification}
+\end{table}
+
+Note well that in the absence of bugs in the implementation (which
+again should be fixed), it is the implementation's classification of
+the dependency that dictates whether or not the corresponding execution
+is temporal.
+
+\subsubsection{C++ Implementations: Optimization}
+\label{sec:C++ Implementations: Optimization}
+
+C++ implementations undertake a wide variety of optimizations, in
+some cases taking advantage of nondeterminism and in others applying
+the results of code analysis to produce faster code.
+The changes can be profound, for example, with large computations being
+replaced by simple constants.
+Either way, the implementation is behaving as if the original program
+had been written more optimally.
+
+In such cases, one option is to carry out analysis of semantic dependencies
+on the optimized code.
+The seL4 project took this approach to the extreme of analyzing
+the binary output~\cite{ThomasSewell2013L4binaryVerification}.
+
+In C++ implementations using compilers, nondeterminism of a given
+(post-optimization) code path is ``frozen'' into the resulting binary,
+which greatly simplifies analysis of semantic dependencies.
+Other C++ implementations (for example, those using interpreters or JITs)
+might need to directly model any nondeterminism.
+This latter situation might motivate additional C++ projects to ban
+nondeterminism in favor of portable code.
+
+Another question is how to prove that a given optimization is valid.
+This has been a hard question in the past, and is likely to remain so
+both for sequential and concurrent programs.
+The consequences of using an invalid optimization can be quite severe,
+and the evidence thus far indicates that OOTA cycles are not a bit issue.
+Nevertheless, a proof that the same value is always stored implies a
+proof that there is no semantic dependency to that relaxed atomic store,
+and thus no chance of that store participating in an OOTA cycle.
+
+\subsubsection{C++ Implementations: Consequences}
+\label{sec:C++ Implementations: Consequences}
+
+Given a correct C++ implementation, a given store will either be
+executed as if is was semantically dependent on a given prior load or
+not.
+In the former case on a real-world implementation, a nonnegative interval
+of time will elapse between the load and the store, and in the latter
+case, that load-store pair cannot possibly be a member of an OOTA cycle.
+
+Of course, dependencies can be separately analyzed, and such analysis
+might show that one of the dependencies that the implementation treated
+as a semantic dependency was in fact nonsemantic.
+And this is a belt-and-suspenders situation:
+The corresponding load-store pair cannot be a member of an OOTA cycle,
+and a nonnegative interval of time will elapse between the load and
+the store.

### 2024-02-21 ootavacuous.tex
Drop "C++ Implementation" section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 pair of objects, with the memory-ordering semantics of the first load.
-
-\subsection{C++ Implementations}
-\label{sec:C++ Implementations}
-
-As noted earlier, this initial work focuses on C++ implementations based
-on compilers and computing hardware (including CPUs and GPGPUs).
-
-C++ implementations execute C++ programs, and use a range of techniques.
-For example, many implementations take steps to optimize the code to
-be executed.
-This optimization presents a trade-off between the time spent optimizing
-the code and the time spent actually executing it.
-One type of optimization is the identification of nonsemantic dependencies
-and the replacement of them with simpler computations, up to and including
-replacement of arbitrarily complex passages of code with constants.\footnote{
-	There are a great many other optimizations, but this paper
-	necessarily focuses on semantic dependencies.}
-
-This optimization thread may, roughly speaking, be divided into two
-activities, analysis of the code and classification of dependencies.
-Two sections on these topics are followed by sections on optimizations
-and on consequences.
-
-\subsubsection{Code Analysis}
-\label{sec:Code Analysis}
-
-As would be expected given the various optimization/execution tradeoffs,
-different C++ implementations carry out different degrees of analysis.
-
-First, at one extreme, ``simple implementations'' do absolutely no
-analysis beyond that absolutely required to correctly execute the program.
-
-Second, other implementations do local analysis.
-This analysis might be on a per-code-path basis, within a single function,
-within a single function after inlining, and so forth.
-
-Third, one could imagine implementations whose analysis is restricted to
-a thread.
-In this case, the analysis stops at loads to objects that were last
-stored to by other threads (``rfe''), stores to objects where the next
-and/or previous stores in that object's modification order were executed
-by other threads (``coe''), and loads from objects for which the value
-returned by a given load was overwritten by some other thread
-(``fre'').\footnote{
-	See Appendix~\ref{app:Inter-Process Communications} for more
-	information on these memory-reference inter-process communications
-	links.}
-
-Fourth, one could also imagine implementations whose analysis is
-completely unrestricted, up to and including the limiting case where
-execution is oracular in nature.
-Of course, if we are going to imagine omniscient analysis and oracles,
-we can also rely on them to locate any OOTA cycles that might appear
-in environments not constrained by the laws of physics, hardware
-limitations, and C++ implementation concerns.
-
-However, for more realistic implementations in the third category,
-analysis relevant to OOTA cycles begins with loads from objects that were
-last stored to by some other thread and ends with stores to objects that
-might be loaded by other threads.
-
-If additional inter-thread information is available, the corresponding
-constraints may be applied.
-For example, if an implementation proves that the values of a pair of
-objects are always equal at any time that a given thread might access
-them, that implementation could behave as if the that thread's code had
-substituted a load from one for a load from the other.
-
-Note that if all of the inter-thread relaxed accesses are volatile,
-as they are in the C language, then all of the inter-thread traffic is
-observed behavior.
-In this case, even omniscient oracular implementations are required
-to confine their analyses to within a single thread.
-
-The next section focuses on the key analysis step for OOTA cycles,
-namely dependency classification.
-
-\subsubsection{Dependency Classification}
-\label{sec:Dependency Classification}
-
-Carrying out this type of optimization requires that the implementation
-classify dependencies as maybe semantic or definitely not.
-Given the limited resources available, almost all implementations will
-choose to approximate sdep, which entails some misclassifications, with
-the consequences summarized in
-Table~\ref{tab:Semantic-Dependency Classification}.
-The first row of the table depicts the case where the implementation
-correctly identifies a semantic dependency, and thus exhibits temporal
-behavior for the execution at hand of the corresponding code.
-The second row depicts the case where the implementation misclassifies
-the dependency as nonsemantic, which will result in an incorrect
-execution.
-In this case, the proper course of action is to fix that bug, which is
-most likely a bug even in the absence of OOTA cycles.
-The third row depicts the case where the implementation misclassifies
-the dependency as semantic, thus missing a chance to optimize.
-The resulting execution is nevertheless correct, and is also temporal.
-The fourth and final row depicts the case where the implementation
-correctly classifies the dependency as nonsemantic, in which case
-the execution is both correct and atemporal.
-
-\begin{table}
-\centering
-\begin{tabular}{c|c|l}
-Actual Execution	& Implementation	& Result \\
-\hline
-sdep			& sdep			& Temporal \\
-\cline{2-3}
-			& $\neg$sdep		& Fix C++ bug! \\
-\hline
-$\neg$sde		& sdep			& Temporal \\
-\cline{2-3}
-			& $\neg$sdep		& Atemporal \\
-\end{tabular}
-\caption{Semantic-Dependency Classification}
-\label{tab:Semantic-Dependency Classification}
-\end{table}
-
-Note well that in the absence of bugs in the implementation (which
-again should be fixed), it is the implementation's classification of
-the dependency that dictates whether or not the corresponding execution
-is temporal.
-
-\subsubsection{C++ Implementations: Optimization}
-\label{sec:C++ Implementations: Optimization}
-
-C++ implementations undertake a wide variety of optimizations, in
-some cases taking advantage of nondeterminism and in others applying
-the results of code analysis to produce faster code.
-The changes can be profound, for example, with large computations being
-replaced by simple constants.
-Either way, the implementation is behaving as if the original program
-had been written more optimally.
-
-In such cases, one option is to carry out analysis of semantic dependencies
-on the optimized code.
-The seL4 project took this approach to the extreme of analyzing
-the binary output~\cite{ThomasSewell2013L4binaryVerification}.
-
-In C++ implementations using compilers, nondeterminism of a given
-(post-optimization) code path is ``frozen'' into the resulting binary,
-which greatly simplifies analysis of semantic dependencies.
-Other C++ implementations (for example, those using interpreters or JITs)
-might need to directly model any nondeterminism.
-This latter situation might motivate additional C++ projects to ban
-nondeterminism in favor of portable code.
-
-Another question is how to prove that a given optimization is valid.
-This has been a hard question in the past, and is likely to remain so
-both for sequential and concurrent programs.
-The consequences of using an invalid optimization can be quite severe,
-and the evidence thus far indicates that OOTA cycles are not a bit issue.
-Nevertheless, a proof that the same value is always stored implies a
-proof that there is no semantic dependency to that relaxed atomic store,
-and thus no chance of that store participating in an OOTA cycle.
-
-\subsubsection{C++ Implementations: Consequences}
-\label{sec:C++ Implementations: Consequences}
-
-Given a correct C++ implementation, a given store will either be
-executed as if is was semantically dependent on a given prior load or
-not.
-In the former case on a real-world implementation, a nonnegative interval
-of time will elapse between the load and the store, and in the latter
-case, that load-store pair cannot possibly be a member of an OOTA cycle.
-
-Of course, dependencies can be separately analyzed, and such analysis
-might show that one of the dependencies that the implementation treated
-as a semantic dependency was in fact nonsemantic.
-And this is a belt-and-suspenders situation:
-The corresponding load-store pair cannot be a member of an OOTA cycle,
-and a nonnegative interval of time will elapse between the load and
-the store.
 

### 2024-02-21 ootavacuous.tex
Remove backwards references from Section 8.3
("Semantic Dependencies and Tooling")

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 of possible executions of that same program.
-This difference in function results in different sensitivities to
-sdep misclassifications, as shown in
+While implementations are never supposed to break a semantic dependency,
+tools are in some cases permitted to misclassify them, as shown in
 Table~\ref{tab:Semantic-Dependency Classification For Tools}.
@@
 executing the program.
-While taking this approach sacrifices significant generality, it has the
-significant benefit of greatly reducing the cost of dependency analysis,
-potentially all the way down to zero.
+Either way, this approach reduces the problem of analysis from the
+level of the C++ abstract machine to a much simpler lower level of
+abstraction.
+While taking this approach sacrifices significant generality, it also
+has the significant benefit of greatly reducing the cost of dependency
+analysis, potentially all the way down to zero.
 

### 2024-02-21 ootavacuous.tex
Alan Stern February 21 feedback, Section 8
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Real-world constraints are imposed by the standard, which have been
-considered in prior work.
+The C++ standard imposes many constraints, and these have been considered
+in prior work.
 However, in correct real-world C++ implementations, the standard's
@@
 This mapping has not been considered by prior work, and brings to
-bear additional constraints from the laws of physics and from hardware
-architecture and design, both of which have been only partially accounted
-for in the standard and in prior work.
+bear additional constraints from from hardware architecture and
+design,\footnote{
+	Which is in turn constrained by the laws of physics.}
+which has been only partially accounted for in the standard and in
+prior work.
 One reason for the incomplete accounting for hardware constraints is
@@
 models did not appear until after the standard was released.
-A final set of constraints is imposed by the C++ implementations
-themselves, some of which have resulted in nondeterminisms within
-the standard.\footnote{
-	For example, when pre-existing implementations could not agree
-	on any specific deterministic handling of side-effects within
-	expressions.
-	On the other hand, implementer freedom does have significant
-	value.}
-
-The following sections discuss these constraints, starting with
-with hardware constraints, continuing
-further with constraints imposed by the standard, and culminating
-with constraints imposed by the C++ implementations themselves.
-An additional section discusses the OOTA-cycle implications for
-any tooling that exists separately for these C++ implementations.
+
+The following sections discuss these constraints, starting with with
+hardware constraints and continuing with additional constraints imposed
+by the standard.
+An additional section discusses the OOTA-cycle implications for C++
+tooling.
 

### 2024-02-21 ootavacuous.tex
Alan Stern February 21 feedback, Section 8.1
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-On real systems, instructions require time to execute, even
-when executed speculatively.
+Computer systems are expected to continue increasing their use of
+speculative execution.
+Nevertheless, the main points of this paper will remain unaffected.
+To see this, keep in mind that on real systems, observable effects
+must eventually be committed, but that nothing can be committed
+until all of the computations that determine those effects remain
+speculative.
+Furthermore, instructions take time to execute, even when executed
+speculatively.
 For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
@@
 and~12 cannot be committed until after the load in line~8 completes
-and the conditional on line~9 is evaluated nonspeculatively.\footnote{
-	Alternatively, and as noted earlier, one could imagine that both
-	the speculation and the squashing of failed speculation might
-	span multiple threads.
-	This might make sense for multiple hardware threads sharing
-	a core.}
+and the conditional on line~9 is evaluated nonspeculatively.
 Similarly, lines~18 and~20 cannot be committed until after lines~16

### 2024-02-21 ootavacuous.tex
Alan Stern February 21 feedback, Section 8.2
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 machine.
-This means that any volatile atomic operation involving a store
-must execute as if all loads whose return values are used to compute
-that store's address or value have already returned.
+This means that any volatile atomic operation involving a store may
+execute only after the committing of all loads whose return values are
+used to compute that store's address or value.
 OOTA afficienados will recognize this as a special case of ordering
@@
 Section~\ref{sec:Volatile and Quasi Volatile Accesses},
-implementations must not behave as if the source code has
+implementations should not behave as if the source code has
 additional:
@@
  4 };
- 5 struct foo *globalfoop;
+ 5 _Atomic struct foo *globalfoop;
  6
@@
 To see this, please refer to
-Listing{lst:Do Not Duplicate Atomic Loads}.
+Listing~\ref{lst:Do Not Duplicate Atomic Loads}.
 Suppose that the load on line~13 were duplicated, for example, by
@@
 
-Therefore, C++ implementations should not invent loads, and should
-not even duplicate them.
+Therefore, C++ implementations should avoid inventing atomic loads,
+and in fact should avoid even duplicating them.
 
@@
 pair of objects, with the memory-ordering semantics of the first load.
+
+This provides an example of the ``one-to-one'' bullet of
+Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
 

### 2024-02-21 ootavacuous.tex
Alan Stern February 21 feedback, Section 8.3
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 of possible executions of that same program.
-While implementations are never supposed to break a semantic dependency,
+While implementations should avoid breaking semantic dependencies,
 tools are in some cases permitted to misclassify them, as shown in

### 2024-02-23 ootavacuous.tex
Apply Alan Stern feedback on store-load commit order
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 must eventually be committed, but that nothing can be committed
-until all of the computations that determine those effects remain
-speculative.
+until all of the computations that determine those effects have
+committed.
 Furthermore, instructions take time to execute, even when executed
@@
 
-Therefore, nonspeculative loads are prohibited from loading values from
-speculative stores.
+Therefore, a given load is prohibited from committing until after the
+store that it reads from has committed.
 And yes, this does mean that hardware can produce OOTA cycles during
@@
 speculative OOTA cycles from ever committing.
-
-As a special case of this prohibition, a value stored speculatively must
-not be exposed to nonspeculative loads from other threads.
-In the further special case of multiple hardware threads sharing a core,
-one might imagine speculation involving multiple threads.
-However, in this case, any squashing of speculation must squash the
-full extent of that speculation, across all of the hardware threads
-that were involved.
-
-Applying this to Listing~\ref{lst:OOTA Cycle}, because values from
-speculative stores cannot be visible to nonspeculative loads, lines~10
-and~12 cannot be committed until after the load in line~8 completes
-and the conditional on line~9 is evaluated nonspeculatively.
+As a special case of this prohibition, if a speculated load obtains
+a value from a store that has not yet committed, then if that store
+is squashed, the load must also be squashed, even if that load and
+that store are executed by different CPUs.
+
+Applying this to Listing~\ref{lst:OOTA Cycle}, lines~10 and~12 cannot be
+committed until after the load in line~8 and the conditional on line~9
+have both committed.
 Similarly, lines~18 and~20 cannot be committed until after lines~16
-and~17 complete nonspeculatively.
+and~17 have committed.
 
@@
 relying on conditional branches.
-Nevertheless, the conditional store could not be committed while any
-the instructions it depends on were still being speculatively executed.
+Nevertheless, the conditional store could not be committed until all of
+the instructions it depends on have committed.
 
@@
 This means that for any set of speculative accesses to be committed,
-all loads in that set must have been confirmed by corresponding
-nonspeculative stores.
+all loads in that set must have committed, which in turn requires that
+all store providing the loads' values must also have committed.
 

### 2024-02-23 ootavacuous.tex
Update the commentary on the causality test cases
The commentary on the causality test cases is out of date as compared
with the rest of the paper.  Update it, while removing some unneeded
parts.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 is to be allowed.
-The \co{herd7} tool finds this litmus test's cycle, even without the
-inter-thread analysis, because C++ does not respect control dependencies
-to relaxed atomic stores.
+And indeed, this result does show up in the output from the \co{herd7} tool.
 
@@
 
-Implementations that can prove that the value of \co{x} is always
-nonnegative can also prove that there is no semantic dependency between
-\co{P0()}'s load from \co{x} on line~8 to its store to \co{y} on line~10,
+Compilers that can prove the value of \co{x} is always
+nonnegative can also determine that there is no semantic dependency between
+\co{P0()}'s load from \co{x} on line~8 and its store to \co{y} on line~10,
 that is, no matter which of the possible values is loaded from \co{x},
 the value 1 will always be stored to \co{y}.
-
-Less-omniscient implementations, including those whose code analysis
-is restricted to a thread, would need to act as if there was in fact a
-semantic dependency.
-
-Even omniscient implementations will observe a semantic dependency from
+Relative to less-omniscient compilers, however,
+including those using single-thread analysis or treating atomic
+objects as volatile, there \emph{is} a semantic dependency.
+
+Even omniscient compilers will observe a semantic dependency from
 \co{P1()}'s load from \co{y} on line~14 to its store to \co{x} on line~15.
 
-Either way, there is no OOTA cycle, with omniscient implementations
-having no cycle at all, and other implementations retaining the
-nonsemantic dependency.
+Either way no OOTA cycle will occur, with omniscient compilers
+having no cycle at all because of the lack of an sdep link in \co{P0()},
+and other implementations prevented by hardware
+dependency ordering from realizing the cycle.
 
@@
 is to be allowed.
-The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
+This result appears in the \co{herd7} output.
 
@@
 Because \co{P0()}'s loads from \co{x} on lines~8 and~9 are unordered,
-an implementation can act as if the source code loaded from \co{x} only
-once and assigned the returned value to both \co{r1} and \co{r2}.\footnote{
+a C++ compiler that treats atomics as quasi volatile
+can act as if the source code loaded from \co{x} only
+once and assigned the returned value to both \co{r1} and \co{r2},
+merging the loads.\footnote{
 	But not in C, where relaxed atomic operations are volatile,
-	and thus observable behavior.}
-The implementation could then prove that the condition on line~10
-would always evaluate to \co{true}, and act as if the source code had
-unconditionally executed the relaxed store to \co{y} on line~11.
-The implementation could then act as if the resulting relaxed accesses
-to \co{x} and \co{y} appeared in reverse order, at which point
-\co{P0()}'s store to \co{y} could be read by \co{P1()} load so that
-\co{P1()} stored 1 to \co{x}, thus resulting in \co{P1()} loading
-that 1 from \co{x}.
-
-Note that \co{P1()} has a trivial semantic dependency from its load
-from \co{y} on line~15 to its store to \co{x} on line~16.
-
-If \co{P0()}'s loads from \co{x} are fused, there is no cycle,
-otherwise the preserved nonsemantic dependency prevents an
-OOTA cycle from forming on correct real-world implementations.
+	and thus are observable behavior not subject to merging.}
+The compiler could then prove that the condition on line~10
+would always true, and act as if the source code had
+unconditionally executed the relaxed store to \co{y} on line~11
+independent of the value loaded from \co{x}.
+Without this dependency there would then be no OOTA cycle.
+
+For compilers that do not merge the two loads, the hardware
+dependency ordering arising from the conditional test on line~10
+will prevent the realization of an OOTA cycle.
 
@@
 is to be allowed.
-The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
-
-The analysis from Section~\ref{app:Causality Test Case 2}
-applies here, including the difference in volatility between C and C++
-atomics.
-Although the addition of \co{P2} allows sequentially consistent executions
-where \co{P0()}'s two loads from \co{x} might return different values,
+The \co{herd7} tool reports this result.
+
+The analysis from the previous section applies here.
+Although the addition of \co{P2()} allows sequentially consistent executions
+in which \co{P0()}'s two loads from \co{x} return different values,
 it does not change the fact that those two loads can be merged.
@@
 is to be forbidden.
-The \co{herd7} tool finds this OOTA cycle because C++ does not respect
-data dependencies, and flags it as such using \co{S8}.
+The \co{herd7} tool finds this OOTA result;
+the \co{S8} values given on line~3 of its output indicate
+that in this execution the variables \co{r1} and~\co{r2} were never
+assigned any specific value, so they could just as well be equal to
+one as to anything else.
 
@@
 
-Both \co{P0()} and \co{P1()} have straightforward semantic dependencies
-from their relaxed atomic loads to their relaxed atomic stores, which
-prevents this OOTA cycle on correct real-world C++ implementations.
+This test case is virtually the same as the Simple OOTA cycle
+discussed in section~\ref{sec:Simple OOTA Cycle}, the only difference
+being the final values of the variables.
 
@@
 Listing~\ref{lst:Causality Test Case 5}
-shows causality test case 5, for which the \co{r1 == r2 == 1 && r3 == 0}
+shows causality test case 5, for which the \co{r1 == r2 == 1} and \co{r3 == 0}
 result is to be forbidden.
-The \co{herd7} tool finds this OOTA cycle because C++ does not respect
-data dependencies, and flags it as such using \co{S8}.
+The \co{herd7} tool reports this OOTA result in line~3 of its output.
 
@@
 
-Each of \co{P0()}, \co{P1()}, and \co{P3()} have trivial semantic
-dependencies from their relaxed atomic loads to their relaxed
-atomic stores, which prevents any OOTA cycle from forming in correct
-real-world C++ implementations.
-\co{P2()} has no loads, so cannot participate in an OOTA cycle in
+This is the same as causality test case~4 with \co{P2()} and \co{P3()}
+added, but those routines play no role in the OOTA cycle.
+\co{P2()} has no loads and so cannot participate in an OOTA cycle in
 any case.
@@
 result is to be allowed.
-The \co{herd7} tool finds this cycle, and does not flag it as an OOTA
-cycle because C++ does not respect control dependencies to relaxed
-atomic stores.
+The \co{herd7} tool reports this result.
 
@@
 
-A C++ implementation that can prove that \co{y} is always either
+A compiler that can prove \co{y} is always either
 zero or one can also prove that there is no semantic dependency
-from \co{P1()}'s load on line~14 to its stores on lines~16 and~29.
-Such an implementation could act as if lines~15-20 had been replaced
-by a single store to \co{x}, allowing that store to execute first,
-followed by lines~8, 10, and~14, in that order, satisfying the
-\co{exists} clause on line~23.
-
-Less-omniscient implementations will preserve the nonsemantic
-dependency, preventing any OOTA loop from forming in a correct real-world C++
-implementation.
+from \co{P1()}'s load on line~14 to its stores on lines~16 and~19.
+Such a compiler could act as if lines~15--20 had been replaced
+by a single store to \co{x}, showing that the dependency isn't
+semantic.
+
+Less-omniscient compilers will preserve the dependency, and then
+hardware dependency ordering will prevent the OOTA cycle from
+being realized.
 
@@
 result is to be allowed.
-The \co{herd7} tool finds this cycle, and does not flag it as an OOTA cycle.
+The \co{herd7} tool finds this result.
 
@@
 
-Because all accesses are relaxed atomics, this test case has everything
+This test case has everything
 to do with reordering and nothing to do with OOTA cycles.
-The outcome happens when lines~17, 10, 11, 15, 16 and~9 execute in that
-order.
-Note that this ordering preserves the semantic dependencies from
-lines~10-11 and~15-16, and can thus happen on correct real-world
-C++ implementations.
+The \co{r1 == r2 == r3 == 1} outcome happens when lines~17, 10, 11,
+15, 16 and~9 execute in that order.
+Note that this execution order respects the semantic dependencies in
+lines~10--11 and~15--16, and thus is not ruled out by hardware
+dependency ordering.
+It could have been ruled out if some of the accesses were acquire or
+release, but here they are all relaxed.
 
@@
 result is to be allowed.
-The \co{herd7} tool does not find this cycle, but when \co{r2} is explicitly
-initialized to the constant \co{1}, it does find the cycle and refrains
-from flagging it as an OOTA cycle.
+The \co{herd7} tool does not report this result, but it does if
+\co{r2} is initialized to one before being set to
+\co{1 + r1 * r1 - r1}.
 
@@
 
-Omniscient C++ implementations might note that the computation on line~9
-can only have the value one when the load from \co{x} returns either
-zero or one, and both \co{x} and \co{y} are initialized to zero.
-Such an implementation act as if line~10's \co{r2} had instead been
-the value~1, so that the value returned by \co{P0()}'s load could not
-affect the value stored.
+Omniscient compilers might note that the computation on line~9
+will have the value one when the load from \co{x} returns either
+zero or one, which are the only values \co{x} and \co{y} can take.
+Such a compiler could then realize that line~9 always sets \co{r2} to one,
+independent of the value loaded in line~8.
 Then there would be no semantic dependency between \co{P0()}'s load
-and store, and thus no OOTA cycle in a correct real-world C++
-implementation.
-
-A less-capable implementation would leave the nonsemantic dependency
-in place, also preventing any OOTA cycle in a correct real-world C++
-implementation.
+and store, and thus no OOTA cycle relative to this compiler.
+
+A less-capable compiler would leave the dependency in place,
+and the resulting hardware dependency ordering would prevent
+the OOTA cycle from being realized.
 
@@
 result is to be allowed.
-The \co{herd7} tool does not find this cycle because it depends on
-scheduling \co{P2()} after the other two threads, and C++ is not
-constrained to do this.
+The \co{herd7} tool does not report this result.
 
@@
 
-Both threads have straightforward semantic dependencies between their
-respective relaxed atomic loads and relaxed atomic stores, so an OOTA
-cycle cannot form in a correct real-world C++ implementation.
+This is the same as causality test case~8 but with \co{P2()} added;
+the store to \co{x} in \co{P2} can interfere with the formation of the
+OOTA cycle.
+
+Unlike in causality test case~8, the computation on line~9 must be
+carried out because now the load on line~8 might return two.
+Consequently \co{P0} and~\co{P1} both have semantic dependencies,
+and hardware dependency ordering will prevent the OOTA cycle from
+being realized.
 
@@
 result is to be allowed.
-The \co{herd7} tool does not find this cycle because it depends on
-scheduling \co{P2()} before the other two threads, and C++ is not
-constrained to do this.
+The \co{herd7} tool does not report this result.
 
@@
 
-Both threads have straightforward semantic dependencies between their
-respective relaxed atomic loads and relaxed atomic stores, so an OOTA
-cycle cannot form in a correct real-world C++ implementation.
+This is the same as causality test case~9 except that here \co{P2()}
+sets \co{x} to zero instead of two.
+As as result the computation on line~9 is once again independent of
+the value loaded on line~8, which makes the analysis nearly the same
+as that of causality test case~8.
 
@@
 Listing~\ref{lst:Causality Test Case 10}
-shows causality test case 10, for which the \co{r1 == r2 == 1 && r3 == 0}
+shows causality test case 10, for which the \co{r1 == r2 == 1} and \co{r3 == 0}
 result is to be forbidden.
-Because C++ does not respect control dependencies, the \co{herd7} tool
-nevertheless reports this result and does not flag it as an OOTA cycle.
+The \co{herd7} tool reports this result.
 
@@
 
-However, each of \co{P0()}, \co{P1()}, and \co{P3()} have straightforward
+\co{P0()} and \co{P1()} have straightforward
 semantic dependencies between their respective loads and stores, and
-thus cannot form an OOTA cycle when running on a correct real-world C++
-implementation.
-For its part, \co{P2()} has no semantic dependency, so it cannot be part
-of an OOTA cycle when running on a correct real-world C++ implementation.
+so hardware dependency ordering will prevent an OOTA cycle from being
+realized.
+\co{P2()} has no semantic dependency, and \co{P3()}'s store does not
+occur in this execution because \co{r3} is zero.
+Thus neither of them contributes to an OOTA cycle.
 
@@
 result is to be allowed.
-The \co{herd7} tool reports this result and does not flag it as an OOTA cycle.
+The \co{herd7} tool reports this result.
 
@@
 
-This test case has nothing to do with OOTA cycles.
-The result may be obtained by executing the statements in the following
-order: line~20, line~12, line~13, line~18, line~19, line~10, line~11,
-and finally line~17.
+Like causality test case~7, this test case has nothing to do with OOTA cycles.
+The result may be obtained by executing lines~20, 12, 13, 18, 19, 10, 11,
+and~17 in that order.
 
@@
 
-JMM Causality Test Case 12 in
+Causality Test Case 12 in
 Listing~\ref{lst:Causality Test Case 12}
@@
 
-An omniscient C++ implementation could note that only \co{P0()} accesses
+An omniscient compiler could note that only \co{P0()} accesses
 array \co{a[]} and act as if the code was simpler, but there would
@@
 relaxed atomic loads and stores involving \co{x} and \co{y}.
-Those dependencies prevent an OOTA cycle from forming on a correct
-real-world C++ implementation, omniscient or not.
+The resulting hardware dependency ordering will prevent an OOTA cycle
+from being realized, whether the compiler is omniscient or not.
 
@@
 result is to be forbidden.
-But because C++ does not respect control dependencies, \co{herd7} tool
-reports this result and does not flag it as an OOTA cycle.
+The \co{herd7} tool reports this result.
 
@@
 
-There are straightforward semantic dependencies between \co{P0()}'s and
-\co{P1()}s relaxed atomic loads and stores involving \co{x} and \co{y}.
-Those dependencies prevent an OOTA cycle from forming on a correct
-real-world C++ implementation.
+This is the same as causality test case~10 without the confounding
+influence of \co{P2()} and~\co{P3()}.
 
@@
 Listing~\ref{lst:Causality Test Case 14}
-shows causality test case 14, for which the \co{r1 == r3 == 1 && r2 == 0}
+shows causality test case 14, for which the \co{r1 == r3 == 1} and \co{r2 == 0}
 result is to be forbidden.
-The \co{herd7} tool nevertheless reports this result because C++ does not
-respect control dependencies for atomic stores.
+The \co{herd7} tool reports this result.
 
@@
 
-In C++, this is not a matter of OOTA but rather reordering.
-Dependencies notwithstanding, according to the C++ memory model,
-the statements could execute in the following order:
-Line~17 (so \co{r2 == 0}), line~21, line~9 (so \co{r1 == 1}),
-line~13, line~18 (so \co{r3 == 1)}.
-The sum of \co{r2} and \co{r3} is nonzero, justifying line~21's
-store to \co{x}.
-
-However, correct real-world C++ implementations would have semantic
-dependencies between \co{P0()}'s line~9 and lines~11 and~13, and also
-between \co{P1()}'s lines~17 and~18 and line~21, preventing any
-OOTA cycle from forming.
+In this execution, \co{P0()} has a semantic dependency from line~9 to
+line~13 and \co{P1()} has a semantic dependency from line~18 to
+line~21.
+(There's also a semantic dependency from line~17 to line~21 but it
+doesn't enter into this OOTA cycle, which involves only \co{x} and~\co{y}.)
+The \co{memory_order_seq_cst} specification on line~17 has no important effect;
+hardware dependency ordering will prevent the cycle from being realized.
 
@@
 Listing~\ref{lst:Causality Test Case 15}
-shows causality test case 15, for which the \co{r0 == r1 == r3 == 1 && r2 == 0}
+shows causality test case 15, for which the \co{r0 == r1 == r3 == 1}
+and \co{r2 == 0}
 result is to be forbidden.
-The \co{herd7} tool nevertheless reports this result because C++ does not
-respect control dependencies for atomic stores.
+The \co{herd7} tool reports this result.
 
@@
 
-In C++, this is not a matter of OOTA but rather reordering.
-Dependencies notwithstanding, according to the C++ memory model,
-the statements could execute in the following order:
-Line~31, line~27, line~10 (so \co{r0 == 1}), line~13 (so \co{r1 == 1}),
-line~19 (because \co{r1 == 1}), line~23 (so \co{r2 == 0}),
-and finally line~24 (so \co{r3 == 1}).
-The sum of \co{r2} and \co{r3} is nonzero, justifying line~27's
-store to \co{u}.
-
-However, correct real-world C++ implementations would have semantic
-dependencies between \co{P0()}'s lines~10 and~13 and lines~17 and~19, and also
-between \co{P1()}'s lines~23 and~24 and line~27, preventing any
-OOTA cycle from forming.
+This execution has an OOTA cycle involving \co{u} and~\co{v}.
+The dependency in \co{P0()} from line~13 to line~19 and the dependency
+in \co{P1()} from line~24 to line~27 are both semantic.
+The resulting hardware dependency ordering prevents the cycle from
+being realized.
+
 Having no load, \co{P2()} cannot participate in an OOTA cycle.
@@
 Listing~\ref{lst:Causality Test Case 16}
-shows causality test case 16, for which the \co{r1 == 2 && r2 == 1}
+shows causality test case 16, for which the \co{r1 == 2} and \co{r2 == 1}
 result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be expected
-given that C++ does not permit a thread to reorder accesses to a single
-atomic variable, even when those accesses are relaxed.
+The \co{herd7} tool does not report this result, which is to be
+expected because it would violate the C++ memory model's read-write
+coherence rule regarding the modification order of an atomic object
+(6.9.2.2p17 \co{[intro.races]}).
+Because no abstract execution can obtain \co{r1 == 2} and \co{r2 == 1}, no
+valid realization will either.
 
@@
 result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be expected
-given that C++ does not permit a thread to reorder accesses to a single
-atomic variable, even when those accesses are relaxed.
-But note that the \co{herd7} tool correctly notes the presence of an
-OOTA cycle with \co{S17}.
+The \co{herd7} tool does not report this result, although it does
+report a different OOTA cycle involving lines 11, 12, 16, and~17
+(the \co{S17} value on line~3 of the \co{herd7} output).
 
@@
 
-\co{P1()} has a straightforward semantic dependency between its
-relaxed atomic load and relaxed atomic store.
-However, suppose that \co{P0()}'s load from \co{x} on line~8 returns
-the value 42.
-In that case, the second load on line~11 can be fused with the first,
+A compiler that treats atomics as quasi volatile could reason as follows:
+Suppose that \co{P0()}'s load from \co{x} on line~8 returns the value 42.
+In that case, the second load on line~11 can be merged with the first,
 so that it also returns the value 42.
-On the other hand, if the load on line~8 returns some other value, then
+On the other hand, if the load on line~8 returns some other value then
 line~10 will store the value 42 to \co{x}, and then the load on line~11
-can be elided\footnote{
-	In C++, but not in C due to atomics being volatile.}
-in favor of using the value 42 stored by line~10.
-
-Either way, the value of \co{r1} will be the constant 42.
-Therefore, the loads from \co{x} cannot affect the store on line~17,
+can be omitted in favor of using the value 42 stored by line~10.
+
+Either way, the value of \co{r1} will be 42.
+Therefore the loads from \co{x} cannot affect the store on line~17,
 so there is no semantic dependency.
-This allows that store to \co{y} to be moved to the very beginning of
-\co{P0()}, thus allowing a correct real-world C++ implementation to
-satisfy the \co{exists} clause.
+This allows the cycle to be realized,
+but because the dependency in \co{P0()} is not semantic,
+the cycle is not OOTA.
 
@@
 result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be expected
-given that C++ does not permit a thread to reorder accesses to a single
-atomic variable, even when those accesses are relaxed.
-But note that the \co{herd7} tool correctly notes the presence of an
-OOTA cycle with \co{S17}.
+The \co{herd7} tool does not report this result.
 
@@
 
-Both threads have straightforward semantic dependencies between their
-respective relaxed atomic loads and relaxed atomic stores, so an OOTA
-cycle cannot form in a correct real-world C++ implementation.
+This is the same as causality test case~17, even to the alternate OOTA
+cycle reported by the \co{herd7} tool, except that line~9 tests
+\co{r3} for equality to zero rather than non-equality to 42.
+
+Perhaps surprisingly, this makes a difference.
 
@@
 result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be expected
-given that C++ does not permit a thread to reorder accesses to a single
-atomic variable, even when those accesses are relaxed.
-But note that the \co{herd7} tool correctly notes the presence of an
-OOTA cycle with \co{S8}.
-
-It is interesting that this \co{S8} value does not propagate to
-\co{r3}.
+The \co{herd7} tool does not report this result but it does report the
+OOTA cycle in line~3 of its output.
 
@@
 
-All threads have straightforward semantic dependencies between their
-respective relaxed atomic loads and relaxed atomic stores, so an OOTA
-cycle cannot form in a correct real-world C++ implementation.
+This is nearly the same as the Simple OOTA cycle and causality test
+case~4, with the addition of \co{P2()}.
+Just as in those examples, the OOTA cycle cannot be realized by
+compilers that treat atomic objects as volatile or that do
+single-thread analysis and treat atomic objects as quasi volatile.
 
@@
 result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be expected
-given that C++ does not permit a thread to reorder accesses to a single
-atomic variable, even when those accesses are relaxed.
-But note that the \co{herd7} tool correctly notes the presence of an
-OOTA cycle with \co{S8}.
-
-It is interesting that this \co{S8} value does not propagate to
-\co{r3}.
+The \co{herd7} tool does not report this result.
 
@@
 
-All threads have straightforward semantic dependencies between their
-respective relaxed atomic loads and relaxed atomic stores, so an OOTA
-cycle cannot form in a correct real-world C++ implementation.
+This is the same as causality test case~19, even to the OOTA cycle
+reported by the \co{herd7} tool, except that line~19 tests
+\co{r3} for equality to zero rather than non-equality to 42.
+The same analysis applies.
 

### 2024-02-23 ootavacuous.tex
Finish changes to causality test case 18
In an earlier patch, I forgot to complete the changes to the comments on
causality test case 18.  Here is the missing material.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 This is the same as causality test case~17, even to the alternate OOTA
-cycle reported by the \co{herd7} tool, except that line~9 tests
+cycle reported by \co{herd7}, except that line~9 tests
 \co{r3} for equality to zero rather than non-equality to 42.
-
 Perhaps surprisingly, this makes a difference.
+
+A compiler that treats atomic objects as volatile or uses
+single-thread analysis cannot assume that \co{x} will necessarily be
+equal to 42 following lines~9 and~10; it might have some other value
+such as 1.
+Therefore the compiler cannot conclude that \co{r1} will always be 42
+and must leave the semantic dependency between lines~11 and~12 intact.
+Of course, the resulting OOTA cycle will not be realized, because of
+hardware dependency ordering.
 
@@
 This is the same as causality test case~19, even to the OOTA cycle
-reported by the \co{herd7} tool, except that line~19 tests
+reported by \co{herd7}, except that line~19 tests
 \co{r3} for equality to zero rather than non-equality to 42.

### 2024-02-24 ootavacuous.tex
Apply Alan Stern February 22 feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 This mapping has not been considered by prior work, and brings to
-bear additional constraints from from hardware architecture and
+bear additional constraints from hardware architecture and
 design,\footnote{
@@
 To see this, keep in mind that on real systems, observable effects
-must eventually be committed, but that nothing can be committed
-until all of the computations that determine those effects have
-committed.
+must eventually be committed, but that no effect can be committed
+while any of the computations that determine the effect remain
+speculative.
 Furthermore, instructions take time to execute, even when executed
@@
 This means that any volatile atomic operation involving a store may
-execute only after the committing of all loads whose return values are
-used to compute that store's address or value.
+execute only after after all loads whose values are used to compute the
+store's address or value have committed.
 OOTA afficienados will recognize this as a special case of ordering
@@
 excessive cache-miss overheads.
+Even duplicating an atomic load can lead to incorrect results if done
+carelessly.
+To see this, please refer to
+Listing~\ref{lst:Do Not Duplicate Atomic Loads}.
+Suppose that the load on line~13 were duplicated, for example, by
+uncommenting line~15.
+Then the values of \co{a} and \co{b} passed to line~17's call to
+\co{do_something_with()} might be from different instances of the
+\co{foo} structure, something that \co{do_something_with()} might
+not be prepared to deal with.
 
@@
 
-However, even duplicating an existing atomic load can lead to incorrect
-results.
-To see this, please refer to
-Listing~\ref{lst:Do Not Duplicate Atomic Loads}.
-Suppose that the load on line~13 were duplicated, for example, by
-uncommenting line~15.
-Then the values of \co{a} and \co{b} passed to line~17's call to
-\co{do_something_with()} might be from different instances of the
-\co{foo} structure, something that \co{do_something_with()} might
-not be prepared to deal with.
-
-Therefore, C++ implementations should avoid inventing atomic loads,
-and in fact should avoid even duplicating them.
+For these reasons, C++ implementations should avoid inventing atomic
+loads, and in fact should avoid even duplicating them.
 

### 2024-02-24 cuttingroom.tex
February 23 feedback from Alan Stern
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 the store.
+
+@@@  disagreements.
+
+\begin{figure}[tb]
+\begin{center}
+\resizebox{4in}{!}{\includegraphics{coe-nvals}}
+\caption{On x86, Reasonable CPUs Can Disagree}
+\label{fig:On x86; Reasonable CPUs Can Disagree}
+\end{center}
+\end{figure}
+
+Figure~\ref{fig:On x86; Reasonable CPUs Can Disagree}
+shows the results of 79 of the systems's hardware threads simultaneously
+storing their identifying integer (ranging from 0 to 78) to a shared
+variable, then repeatedly doing timestamped reads from that variable.
+The x-axis displays time in TSC periods, each of which is about 0.5~ns
+in duration.
+The y-axis shows the number of distinct opinions that the hardware
+threads have as a function of time, a number that is frequently rather
+larger than one.
+The backwards-in-time coe link from Thread~1 to Thread~0 in
+Figure~\ref{fig:IPC Diagram for coe, fre, and rfe}
+is therefore entirely plausible.
+
+\begin{figure}[tb]
+\begin{center}
+\resizebox{3in}{!}{\rotatebox{90}{\includegraphics{store15tred}}}
+\caption{On Power5, coe Links Are Also Partially Ordered}
+\label{fig:On Power5, coe Links Are Also Partially Ordered}
+\end{center}
+\end{figure}

### 2024-02-24 ootavacuous.tex
February 23 feedback from Alan Stern
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 The code generating this data may be found in the \path{CodeSamples/cpu}
-directory of the
-perfbook~\cite{McKenney2018ParallelProgramming-2018-12-08a}.\footnote{
-	\co{git clone git://git.kernel.org/pub/scm/linux/kernel/git/paulmck/perfbook.git}}
-
-\begin{figure}[tb]
-\begin{center}
-\resizebox{4in}{!}{\includegraphics{coe-nvals}}
-\caption{On x86, Reasonable CPUs Can Disagree}
-\label{fig:On x86; Reasonable CPUs Can Disagree}
-\end{center}
-\end{figure}
-
-Figure~\ref{fig:On x86; Reasonable CPUs Can Disagree}
-shows the results of 79 of the systems's hardware threads simultaneously
-storing their identifying integer (ranging from 0 to 78) to a shared
-variable, then repeatedly doing timestamped reads from that variable.
-The x-axis displays time in TSC periods, each of which is about 0.5~ns
-in duration.
-The y-axis shows the number of distinct opinions that the hardware
-threads have as a function of time, a number that is frequently rather
-larger than one.
-The backwards-in-time coe link from Thread~1 to Thread~0 in
-Figure~\ref{fig:IPC Diagram for coe, fre, and rfe}
-is therefore entirely plausible.
+directory of the git archive of the book
+``Is Parallel Programming Hard, And, If So, What Can You Do About
+It?''~\cite{McKenney2018ParallelProgramming-2018-12-08a}.\footnote{
+	\co{git://git.kernel.org/pub/scm/linux/kernel/git/paulmck/perfbook.git}}
 
@@
 
-Not only is this plausible, but as you can see in
-Figure~\ref{fig:On x86; coe Links Are Atemporal},
-it actually happens, and frequently.
+Figure~\ref{fig:On x86; coe Links Are Atemporal}
+shows that coe links really are atemporal, that is, they can go backwards
+in time, and quite frequently at that, even on x86.
 The store-to-store latency is measured from the end of the ``winning''
@@
 \resizebox{5in}{!}{\includegraphics{coe-out-1}}
-\caption{On x86, coe Links Are Partially Ordered}
-\label{fig:On x86, coe Links Are Partially Ordered}
+\caption{On x86, coe Links Form a Coherent Partial Order}
+\label{fig:On x86, coe Links Form a Coherent Partial Order}
 \end{center}
@@
 
-\begin{figure}[tb]
-\begin{center}
-\resizebox{3in}{!}{\rotatebox{90}{\includegraphics{store15tred}}}
-\caption{On Power5, coe Links Are Also Partially Ordered}
-\label{fig:On Power5, coe Links Are Also Partially Ordered}
-\end{center}
-\end{figure}
-
-However, Figure~\ref{fig:On x86, coe Links Are Partially Ordered}
+However, Figure~\ref{fig:On x86, coe Links Form a Coherent Partial Order}
 shows that the order of reads of each value at each thread are
 consistent with a number of global orders, as required.\footnote{
-	Taken on a 16-CPU x86 laptop to avoid an unreadable
-	diagram containing 79 bubbles.}
-Figure~\ref{fig:On Power5, coe Links Are Also Partially Ordered}
-shows a rather more elaborate partial order from a Power5
-system~\cite{McKenney2018ParallelProgramming-2018-12-08a}.
+	Note that this data was not measured on the 80-CPU system, but
+	rather on a 16-CPU x86 laptop to avoid an unreadable diagram
+	containing 79 bubbles.}
+Working from the far left, one CPU saw the values 0, 11, and 9,
+a second CPU saw 3, 11, and 9, and so on.
+Although different CPUs saw different sequences of values, there is no
+disagreement on the order of the values that they could see.
+For example, all CPUs agree that the value 9 came last.
+This is an example of single-variable sequential consistency, and
+meets the coherence requirements of the C++ standard.
 
@@
 by propagation delay.
-In the case of coe, the two stores take time to propagate across the
-system, and their ordering can be and often is decided after the fact.
-In the case of fre, the store again takes time to propagate across
-the system, and so a load might execute later in global time than
-did the store, but not late enough to return that store's value.
-In the case of rfe, the load cannot return a given store's value until
-the value has propagated to that part of the system executing that load.
+The upper portion of the figure shows an atemporal coe, in which
+CPU~0 stores 1 to \co{x} after CPU~3 stores 2 to \co{x}, but
+the ordering of the stores is decided after the fact.
+In this case, the value of 2 from CPU~3's store overwrites the value
+of 1 from CPU~0's store, despite the fact that CPU~0's store
+happened later in time.
+The middle portion of the figure shows an atemporal fre, in which
+CPU~3's load obtains the old value of zero from \co{x} despite
+having executed long after CPU~0's store, courtesy of the fact
+that the new value of \co{x} had not yet propagated to CPU~3.
+The lower portion of the figure shows a temporal rfe.
+Because CPU~3's load cannot obtain the value stored by CPU~0 until
+after that value has propagated to CPU~3, the fact that CPU~3 obtains
+the new value implies that the load was executed after the store.
 

### 2024-02-24 ootavacuous.tex
Self-review of Figure 6 text
Better explain atemporal coe.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Figure~\ref{fig:Propagation Delay and Temporal Properties of coe, fre, and rfe}
-shows how the temporal properties of coe, fre, and rfe are explained
-by propagation delay.
+shows how propagation delay explains the temporal properties of coe,
+fre, and rfe.
 The upper portion of the figure shows an atemporal coe, in which
@@
 happened later in time.
+One way that this could happen is that the cache line containing
+\co{x} happened to arrive at CPU~3 before arriving at CPU~0.
+A more fanciful explanation is that both values arrived at the
+interconnect at the saem time, and the interconnect made an
+arbitrary choice between the two.
+Either way, the end result is that an earlier store can overwrite
+a later store.
+
 The middle portion of the figure shows an atemporal fre, in which
@@
 that the new value of \co{x} had not yet propagated to CPU~3.
+
 The lower portion of the figure shows a temporal rfe.

### 2024-02-24 ootavacuous.tex
Split litmus-test appendix into two
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Other OOTA-related litmus tests may be found in
-Appendix~\ref{app:OOTA-Related Litmus Tests}.
+Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}
+and
+Appendix~\ref{app:Additional Litmus Tests}.
 
@@
 
-\section{OOTA-Related Litmus Tests}
-\label{app:OOTA-Related Litmus Tests}
-
-This appendix evaluates a number of OOTA-related litmus tests.
-The evaluations are carried out using the \co{herd7} tool and also
-using manual analysis.
-
-Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}
-evaluates the obligatory litmus tests from the JMM Causality Test
-Cases, and
-Appendix~\ref{app:Additional Litmus Tests}
-evaluates some additional litmus tests that affected our thinking or
-otherwise gave us pause.
-
-\subsection{Litmus Tests from “Causality Test Cases"}
+\section{Litmus Tests from “Causality Test Cases"}
 \label{app:Litmus Tests from “Causality Test Cases"}
@@
 	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
-
-\subsubsection{Causality Test Case 1}
+This appendix evaluates them using the \co{herd7} tool and also using
+manual analysis.
+
+\subsection{Causality Test Case 1}
 \label{app:Causality Test Case 1}
@@
 
-\subsubsection{Causality Test Case 2}
+\subsection{Causality Test Case 2}
 \label{app:Causality Test Case 2}
@@
 
-\subsubsection{Causality Test Case 3}
+\subsection{Causality Test Case 3}
 \label{app:Causality Test Case 3}
@@
 
-\subsubsection{Causality Test Case 4}
+\subsection{Causality Test Case 4}
 \label{app:Causality Test Case 4}
@@
 
-\subsubsection{Causality Test Case 5}
+\subsection{Causality Test Case 5}
 \label{app:Causality Test Case 5}
@@
 
-\subsubsection{Causality Test Case 6}
+\subsection{Causality Test Case 6}
 \label{app:Causality Test Case 6}
@@
 
-\subsubsection{Causality Test Case 7}
+\subsection{Causality Test Case 7}
 \label{app:Causality Test Case 7}
@@
 
-\subsubsection{Causality Test Case 8}
+\subsection{Causality Test Case 8}
 \label{app:Causality Test Case 8}
@@
 
-\subsubsection{Causality Test Case 9}
+\subsection{Causality Test Case 9}
 \label{app:Causality Test Case 9}
@@
 
-\subsubsection{Causality Test Case 9a}
+\subsection{Causality Test Case 9a}
 \label{app:Causality Test Case 9a}
@@
 
-\subsubsection{Causality Test Case 10}
+\subsection{Causality Test Case 10}
 \label{app:Causality Test Case 10}
@@
 
-\subsubsection{Causality Test Case 11}
+\subsection{Causality Test Case 11}
 \label{app:Causality Test Case 11}
@@
 
-\subsubsection{Causality Test Case 12}
+\subsection{Causality Test Case 12}
 \label{app:Causality Test Case 12}
@@
 
-\subsubsection{Causality Test Case 13}
+\subsection{Causality Test Case 13}
 \label{app:Causality Test Case 13}
@@
 
-\subsubsection{Causality Test Case 14}
+\subsection{Causality Test Case 14}
 \label{app:Causality Test Case 14}
@@
 
-\subsubsection{Causality Test Case 15}
+\subsection{Causality Test Case 15}
 \label{app:Causality Test Case 15}
@@
 
-\subsubsection{Causality Test Case 16}
+\subsection{Causality Test Case 16}
 \label{app:Causality Test Case 16}
@@
 
-\subsubsection{Causality Test Case 17}
+\subsection{Causality Test Case 17}
 \label{app:Causality Test Case 17}
@@
 
-\subsubsection{Causality Test Case 18}
+\subsection{Causality Test Case 18}
 \label{app:Causality Test Case 18}
@@
 
-\subsubsection{Causality Test Case 19}
+\subsection{Causality Test Case 19}
 \label{app:Causality Test Case 19}
@@
 
-\subsubsection{Causality Test Case 20}
+\subsection{Causality Test Case 20}
 \label{app:Causality Test Case 20}
@@
 
-\subsection{Additional Litmus Tests}
+\section{Additional Litmus Tests}
 \label{app:Additional Litmus Tests}
@@
 of defining sdep and identifying OOTA cycles.
-
-\subsubsection{Simple Reordering as Litmus Test}
+This appendix evaluates them using the \co{herd7} tool and also using
+manual analysis.
+
+\subsection{Simple Reordering as Litmus Test}
 \label{app:Simple Reordering as Litmus Test}
@@
 
-\subsubsection{Non-LB Litmus Tests}
+\subsection{Non-LB Litmus Tests}
 \label{app:Non-LB Litmus Tests}
@@
 
-\subsubsection{Multi-Cycle Litmus Tests}
+\subsection{Multi-Cycle Litmus Tests}
 \label{app:Multi-Cycle Litmus Tests}
@@
 
-\subsubsection{Three-Process Version of JMM Causality Test Case 4}
+\subsection{Three-Process Version of JMM Causality Test Case 4}
 \label{app:Three-Process Version of JMM Causality Test Case 4}
@@
 
-\subsubsection{Why rfe Instead of Tried-And-True rf?}
+\subsection{Why rfe Instead of Tried-And-True rf?}
 \label{app:Why rfe Instead of Tried-And-True rf?}
@@
 
-\subsubsection{Inventing Atomic Loads}
+\subsection{Inventing Atomic Loads}
 \label{app:Inventing Atomic Loads}
@@
 
-\subsubsection{Non-Trivial Semantic Dependencies}
+\subsection{Non-Trivial Semantic Dependencies}
 \label{app:Non-Trivial Semantic Dependencies}
@@
 
-\subsubsection{OOTA From Undefined-Behavior Optimizations}
+\subsection{OOTA From Undefined-Behavior Optimizations}
 \label{app:OOTA From Undefined-Behavior Optimizations}
@@
 
-\subsubsection{Semantic Dependencies and \co{volatile}}
+\subsection{Semantic Dependencies and \co{volatile}}
 \label{app:Semantic Dependencies and volatile}

### 2024-02-24 ootavacuous.tex
Fix typo s/r1/r2/ in first paragraph of 8.1
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	nor cares, nor should it.}
-Then line~11 can cause the value 42 to be loaded into \co{r1}, when
+Then line~11 can cause the value 42 to be loaded into \co{r2},
+thus satisfying the \co{exists} clause on line~20, when
 by the rules of the C++ abstract machine the value must instead be

### 2024-02-24 ootavacuous.tex
Translation from Java to C++ and different verdicts
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
-This appendix evaluates them using the \co{herd7} tool and also using
-manual analysis.
+We have translated them from Java to C++, and we evaluate them using
+the \co{herd7} tool and also using manual analysis.
+Please note that differences in the semantics of these two languages
+result in changes in verdict for some litmus tests.
 

### 2024-02-25 ootavacuous.tex
Add pointer about speculative execution
Footnote 2 in section 2.1.2 talks about speculative execution.  Add a
pointer to the additional discussion in section 8.1.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	make their stores visible to other CPUs until those stores
-	are no longer speculative.}
+	are no longer speculative.
+	See Section~\ref{sec:Hardware Architecture and Design}
+	for more discussion about hardware speculation.}
 the stores in lines~7 and~13 cannot take place until the CPU

### 2024-02-25 ootavacuous.tex
Make "this execution" more explicit in appendix A.1
The discussion of several of the causality test cases mentions "this
execution", usually meaning the execution described by the "exists"
clause.  Make this explicit in the text.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 the \co{S8} values given on line~3 of its output indicate
-that in this execution the variables \co{r1} and~\co{r2} were never
+that in the execution it describes, the variables \co{r1} and~\co{r2} were never
 assigned any specific value, so they could just as well be equal to
@@
 \co{P2()} has no semantic dependency, and \co{P3()}'s store does not
-occur in this execution because \co{r3} is zero.
+occur in this execution (i.e., the one described by the \co{exists} clause)
+because \co{r3} is zero.
 Thus neither of them contributes to an OOTA cycle.
@@
 
-In this execution, \co{P0()} has a semantic dependency from line~9 to
+In the execution described by the \co{exists} clause,
+\co{P0()} has a semantic dependency from line~9 to
 line~13 and \co{P1()} has a semantic dependency from line~18 to
@@
 
-This execution has an OOTA cycle involving \co{u} and~\co{v}.
+The execution described by the \co{exists} clause
+has an OOTA cycle involving \co{u} and~\co{v}.
 The dependency in \co{P0()} from line~13 to line~19 and the dependency

### 2024-02-25 ootavacuous.tex
Improve mention of fused loads for non-volatiles
Section 8.2 has a section about fusing loads of non-volatile atomic
objects.  Make it clear that this applies even to quasi-volatile
objects, as mentioned in section 5.4.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-This provides an example of the ``one-to-one'' bullet of
-Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
+This holds even when the atomic objects are quasi volatile,
+providing an example of how the ``one-to-one'' bullet of
+Section~\ref{sec:Volatile and Quasi Volatile Accesses} does not apply
+to them.
 

### 2024-02-25 ootavacuous.tex
Rename "Additional Litmus Tests" to "Illustrative Litmus Tests"
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 and
-Appendix~\ref{app:Additional Litmus Tests}.
+Appendix~\ref{app:Illustrative Litmus Tests}.
 
@@
 
-\section{Additional Litmus Tests}
-\label{app:Additional Litmus Tests}
+\clearpage
+
+\section{Illustrative Litmus Tests}
+\label{app:Illustrative Litmus Tests}
 

### 2024-02-25 ootavacuous.tex
Move "Illustrative Litmus Tests" to end
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Move "Litmus Tests from “Causality Test Cases"" to end
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Other OOTA-related litmus tests may be found in
-Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}
+Appendix~\ref{app:Illustrative Litmus Tests}
 and
-Appendix~\ref{app:Illustrative Litmus Tests}.
+Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}.
 
@@
 \appendix
-
-\section{Litmus Tests from “Causality Test Cases"}
-\label{app:Litmus Tests from “Causality Test Cases"}
-
-These tests might be two decades old, but they are still quite
-relevant.\footnote{
-	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
-We have translated them from Java to C++, and we evaluate them using
-the \co{herd7} tool and also using manual analysis.
-Please note that differences in the semantics of these two languages
-result in changes in verdict for some litmus tests.
-
-\subsection{Causality Test Case 1}
-\label{app:Causality Test Case 1}
-
-Listing~\ref{lst:Causality Test Case 1}
-shows causality test case 1, for which the \co{r1 == r2 == 1} result
-is to be allowed.
-And indeed, this result does show up in the output from the \co{herd7} tool.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-1.litmus @@
-\caption{Causality Test Case 1}
-\label{lst:Causality Test Case 1}
-\end{listing}
-
-Compilers that can prove the value of \co{x} is always
-nonnegative can also determine that there is no semantic dependency between
-\co{P0()}'s load from \co{x} on line~8 and its store to \co{y} on line~10,
-that is, no matter which of the possible values is loaded from \co{x},
-the value 1 will always be stored to \co{y}.
-Relative to less-omniscient compilers, however,
-including those using single-thread analysis or treating atomic
-objects as volatile, there \emph{is} a semantic dependency.
-
-Even omniscient compilers will observe a semantic dependency from
-\co{P1()}'s load from \co{y} on line~14 to its store to \co{x} on line~15.
-
-Either way no OOTA cycle will occur, with omniscient compilers
-having no cycle at all because of the lack of an sdep link in \co{P0()},
-and other implementations prevented by hardware
-dependency ordering from realizing the cycle.
-
-\subsection{Causality Test Case 2}
-\label{app:Causality Test Case 2}
-
-Listing~\ref{lst:Causality Test Case 2}
-shows causality test case 2, for which the \co{r1 == r2 == r3 == 1} result
-is to be allowed.
-This result appears in the \co{herd7} output.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-2.litmus @@
-\caption{Causality Test Case 2}
-\label{lst:Causality Test Case 2}
-\end{listing}
-
-Because \co{P0()}'s loads from \co{x} on lines~8 and~9 are unordered,
-a C++ compiler that treats atomics as quasi volatile
-can act as if the source code loaded from \co{x} only
-once and assigned the returned value to both \co{r1} and \co{r2},
-merging the loads.\footnote{
-	But not in C, where relaxed atomic operations are volatile,
-	and thus are observable behavior not subject to merging.}
-The compiler could then prove that the condition on line~10
-would always true, and act as if the source code had
-unconditionally executed the relaxed store to \co{y} on line~11
-independent of the value loaded from \co{x}.
-Without this dependency there would then be no OOTA cycle.
-
-For compilers that do not merge the two loads, the hardware
-dependency ordering arising from the conditional test on line~10
-will prevent the realization of an OOTA cycle.
-
-\subsection{Causality Test Case 3}
-\label{app:Causality Test Case 3}
-
-Listing~\ref{lst:Causality Test Case 3}
-shows causality test case 3, for which the \co{r1 == r2 == r3 == 1} result
-is to be allowed.
-The \co{herd7} tool reports this result.
-
-The analysis from the previous section applies here.
-Although the addition of \co{P2()} allows sequentially consistent executions
-in which \co{P0()}'s two loads from \co{x} return different values,
-it does not change the fact that those two loads can be merged.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-3.litmus @@
-\caption{Causality Test Case 3}
-\label{lst:Causality Test Case 3}
-\end{listing}
-
-\subsection{Causality Test Case 4}
-\label{app:Causality Test Case 4}
-
-Listing~\ref{lst:Causality Test Case 4}
-shows causality test case 4, for which the \co{r1 == r2 == 1} result
-is to be forbidden.
-The \co{herd7} tool finds this OOTA result;
-the \co{S8} values given on line~3 of its output indicate
-that in the execution it describes, the variables \co{r1} and~\co{r2} were never
-assigned any specific value, so they could just as well be equal to
-one as to anything else.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-4.litmus @@
-\caption{Causality Test Case 4}
-\label{lst:Causality Test Case 4}
-\end{listing}
-
-This test case is virtually the same as the Simple OOTA cycle
-discussed in section~\ref{sec:Simple OOTA Cycle}, the only difference
-being the final values of the variables.
-
-\subsection{Causality Test Case 5}
-\label{app:Causality Test Case 5}
-
-Listing~\ref{lst:Causality Test Case 5}
-shows causality test case 5, for which the \co{r1 == r2 == 1} and \co{r3 == 0}
-result is to be forbidden.
-The \co{herd7} tool reports this OOTA result in line~3 of its output.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-5.litmus @@
-\caption{Causality Test Case 5}
-\label{lst:Causality Test Case 5}
-\end{listing}
-
-This is the same as causality test case~4 with \co{P2()} and \co{P3()}
-added, but those routines play no role in the OOTA cycle.
-\co{P2()} has no loads and so cannot participate in an OOTA cycle in
-any case.
-
-\subsection{Causality Test Case 6}
-\label{app:Causality Test Case 6}
-
-Listing~\ref{lst:Causality Test Case 6}
-shows causality test case 6, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool reports this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-6.litmus @@
-\caption{Causality Test Case 6}
-\label{lst:Causality Test Case 6}
-\end{listing}
-
-A compiler that can prove \co{y} is always either
-zero or one can also prove that there is no semantic dependency
-from \co{P1()}'s load on line~14 to its stores on lines~16 and~19.
-Such a compiler could act as if lines~15--20 had been replaced
-by a single store to \co{x}, showing that the dependency isn't
-semantic.
-
-Less-omniscient compilers will preserve the dependency, and then
-hardware dependency ordering will prevent the OOTA cycle from
-being realized.
-
-\subsection{Causality Test Case 7}
-\label{app:Causality Test Case 7}
-
-Listing~\ref{lst:Causality Test Case 7}
-shows causality test case 7, for which the \co{r1 == r2 == r3 == 1}
-result is to be allowed.
-The \co{herd7} tool finds this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-7.litmus @@
-\caption{Causality Test Case 7}
-\label{lst:Causality Test Case 7}
-\end{listing}
-
-This test case has everything
-to do with reordering and nothing to do with OOTA cycles.
-The \co{r1 == r2 == r3 == 1} outcome happens when lines~17, 10, 11,
-15, 16 and~9 execute in that order.
-Note that this execution order respects the semantic dependencies in
-lines~10--11 and~15--16, and thus is not ruled out by hardware
-dependency ordering.
-It could have been ruled out if some of the accesses were acquire or
-release, but here they are all relaxed.
-
-\subsection{Causality Test Case 8}
-\label{app:Causality Test Case 8}
-
-Listing~\ref{lst:Causality Test Case 8}
-shows causality test case 8, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not report this result, but it does if
-\co{r2} is initialized to one before being set to
-\co{1 + r1 * r1 - r1}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-8.litmus @@
-\caption{Causality Test Case 8}
-\label{lst:Causality Test Case 8}
-\end{listing}
-
-Omniscient compilers might note that the computation on line~9
-will have the value one when the load from \co{x} returns either
-zero or one, which are the only values \co{x} and \co{y} can take.
-Such a compiler could then realize that line~9 always sets \co{r2} to one,
-independent of the value loaded in line~8.
-Then there would be no semantic dependency between \co{P0()}'s load
-and store, and thus no OOTA cycle relative to this compiler.
-
-A less-capable compiler would leave the dependency in place,
-and the resulting hardware dependency ordering would prevent
-the OOTA cycle from being realized.
-
-\subsection{Causality Test Case 9}
-\label{app:Causality Test Case 9}
-
-Listing~\ref{lst:Causality Test Case 9}
-shows causality test case 9, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not report this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-9.litmus @@
-\caption{Causality Test Case 9}
-\label{lst:Causality Test Case 9}
-\end{listing}
-
-This is the same as causality test case~8 but with \co{P2()} added;
-the store to \co{x} in \co{P2} can interfere with the formation of the
-OOTA cycle.
-
-Unlike in causality test case~8, the computation on line~9 must be
-carried out because now the load on line~8 might return two.
-Consequently \co{P0} and~\co{P1} both have semantic dependencies,
-and hardware dependency ordering will prevent the OOTA cycle from
-being realized.
-
-\subsection{Causality Test Case 9a}
-\label{app:Causality Test Case 9a}
-
-Listing~\ref{lst:Causality Test Case 9a}
-shows causality test case 9a, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not report this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-9a.litmus @@
-\caption{Causality Test Case 9a}
-\label{lst:Causality Test Case 9a}
-\end{listing}
-
-This is the same as causality test case~9 except that here \co{P2()}
-sets \co{x} to zero instead of two.
-As as result the computation on line~9 is once again independent of
-the value loaded on line~8, which makes the analysis nearly the same
-as that of causality test case~8.
-
-\subsection{Causality Test Case 10}
-\label{app:Causality Test Case 10}
-
-Listing~\ref{lst:Causality Test Case 10}
-shows causality test case 10, for which the \co{r1 == r2 == 1} and \co{r3 == 0}
-result is to be forbidden.
-The \co{herd7} tool reports this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-10.litmus @@
-\caption{Causality Test Case 10}
-\label{lst:Causality Test Case 10}
-\end{listing}
-
-\co{P0()} and \co{P1()} have straightforward
-semantic dependencies between their respective loads and stores, and
-so hardware dependency ordering will prevent an OOTA cycle from being
-realized.
-\co{P2()} has no semantic dependency, and \co{P3()}'s store does not
-occur in this execution (i.e., the one described by the \co{exists} clause)
-because \co{r3} is zero.
-Thus neither of them contributes to an OOTA cycle.
-
-\subsection{Causality Test Case 11}
-\label{app:Causality Test Case 11}
-
-Listing~\ref{lst:Causality Test Case 11}
-shows causality test case 11, for which the \co{r1 == r2 == r3 == r4 == 1}
-result is to be allowed.
-The \co{herd7} tool reports this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-11.litmus @@
-\caption{Causality Test Case 11}
-\label{lst:Causality Test Case 11}
-\end{listing}
-
-Like causality test case~7, this test case has nothing to do with OOTA cycles.
-The result may be obtained by executing lines~20, 12, 13, 18, 19, 10, 11,
-and~17 in that order.
-
-\subsection{Causality Test Case 12}
-\label{app:Causality Test Case 12}
-
-Causality Test Case 12 in
-Listing~\ref{lst:Causality Test Case 12}
-uses arrays, which are not yet supported by the \co{herd7} tool.
-
-\begin{listing}[tbp]
-@@ DisplayLitmus litmus/oota-causality-12.litmus @@
-\caption{Causality Test Case 12}
-\label{lst:Causality Test Case 12}
-\end{listing}
-
-An omniscient compiler could note that only \co{P0()} accesses
-array \co{a[]} and act as if the code was simpler, but there would
-still be semantic dependencies between \co{P0()}'s and \co{P1()}s
-relaxed atomic loads and stores involving \co{x} and \co{y}.
-The resulting hardware dependency ordering will prevent an OOTA cycle
-from being realized, whether the compiler is omniscient or not.
-
-\subsection{Causality Test Case 13}
-\label{app:Causality Test Case 13}
-
-Listing~\ref{lst:Causality Test Case 13}
-shows causality test case 13, for which the \co{r1 == r2 == 1}
-result is to be forbidden.
-The \co{herd7} tool reports this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-13.litmus @@
-\caption{Causality Test Case 13}
-\label{lst:Causality Test Case 13}
-\end{listing}
-
-This is the same as causality test case~10 without the confounding
-influence of \co{P2()} and~\co{P3()}.
-
-\subsection{Causality Test Case 14}
-\label{app:Causality Test Case 14}
-
-Listing~\ref{lst:Causality Test Case 14}
-shows causality test case 14, for which the \co{r1 == r3 == 1} and \co{r2 == 0}
-result is to be forbidden.
-The \co{herd7} tool reports this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-14.litmus @@
-\caption{Causality Test Case 14}
-\label{lst:Causality Test Case 14}
-\end{listing}
-
-In the execution described by the \co{exists} clause,
-\co{P0()} has a semantic dependency from line~9 to
-line~13 and \co{P1()} has a semantic dependency from line~18 to
-line~21.
-(There's also a semantic dependency from line~17 to line~21 but it
-doesn't enter into this OOTA cycle, which involves only \co{x} and~\co{y}.)
-The \co{memory_order_seq_cst} specification on line~17 has no important effect;
-hardware dependency ordering will prevent the cycle from being realized.
-
-\subsection{Causality Test Case 15}
-\label{app:Causality Test Case 15}
-
-Listing~\ref{lst:Causality Test Case 15}
-shows causality test case 15, for which the \co{r0 == r1 == r3 == 1}
-and \co{r2 == 0}
-result is to be forbidden.
-The \co{herd7} tool reports this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-15.litmus @@
-\caption{Causality Test Case 15}
-\label{lst:Causality Test Case 15}
-\end{listing}
-
-The execution described by the \co{exists} clause
-has an OOTA cycle involving \co{u} and~\co{v}.
-The dependency in \co{P0()} from line~13 to line~19 and the dependency
-in \co{P1()} from line~24 to line~27 are both semantic.
-The resulting hardware dependency ordering prevents the cycle from
-being realized.
-
-Having no load, \co{P2()} cannot participate in an OOTA cycle.
-
-\subsection{Causality Test Case 16}
-\label{app:Causality Test Case 16}
-
-Listing~\ref{lst:Causality Test Case 16}
-shows causality test case 16, for which the \co{r1 == 2} and \co{r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be
-expected because it would violate the C++ memory model's read-write
-coherence rule regarding the modification order of an atomic object
-(6.9.2.2p17 \co{[intro.races]}).
-Because no abstract execution can obtain \co{r1 == 2} and \co{r2 == 1}, no
-valid realization will either.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-16.litmus @@
-\caption{Causality Test Case 16}
-\label{lst:Causality Test Case 16}
-\end{listing}
-
-\subsection{Causality Test Case 17}
-\label{app:Causality Test Case 17}
-
-Listing~\ref{lst:Causality Test Case 17}
-shows causality test case 17, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result, although it does
-report a different OOTA cycle involving lines 11, 12, 16, and~17
-(the \co{S17} value on line~3 of the \co{herd7} output).
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-17.litmus @@
-\caption{Causality Test Case 17}
-\label{lst:Causality Test Case 17}
-\end{listing}
-
-A compiler that treats atomics as quasi volatile could reason as follows:
-Suppose that \co{P0()}'s load from \co{x} on line~8 returns the value 42.
-In that case, the second load on line~11 can be merged with the first,
-so that it also returns the value 42.
-On the other hand, if the load on line~8 returns some other value then
-line~10 will store the value 42 to \co{x}, and then the load on line~11
-can be omitted in favor of using the value 42 stored by line~10.
-
-Either way, the value of \co{r1} will be 42.
-Therefore the loads from \co{x} cannot affect the store on line~17,
-so there is no semantic dependency.
-This allows the cycle to be realized,
-but because the dependency in \co{P0()} is not semantic,
-the cycle is not OOTA.
-
-\subsection{Causality Test Case 18}
-\label{app:Causality Test Case 18}
-
-Listing~\ref{lst:Causality Test Case 18}
-shows causality test case 18, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-18.litmus @@
-\caption{Causality Test Case 18}
-\label{lst:Causality Test Case 18}
-\end{listing}
-
-This is the same as causality test case~17, even to the alternate OOTA
-cycle reported by \co{herd7}, except that line~9 tests
-\co{r3} for equality to zero rather than non-equality to 42.
-Perhaps surprisingly, this makes a difference.
-
-A compiler that treats atomic objects as volatile or uses
-single-thread analysis cannot assume that \co{x} will necessarily be
-equal to 42 following lines~9 and~10; it might have some other value
-such as 1.
-Therefore the compiler cannot conclude that \co{r1} will always be 42
-and must leave the semantic dependency between lines~11 and~12 intact.
-Of course, the resulting OOTA cycle will not be realized, because of
-hardware dependency ordering.
-
-\subsection{Causality Test Case 19}
-\label{app:Causality Test Case 19}
-
-Listing~\ref{lst:Causality Test Case 19}
-shows causality test case 19, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result but it does report the
-OOTA cycle in line~3 of its output.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-19.litmus @@
-\caption{Causality Test Case 19}
-\label{lst:Causality Test Case 19}
-\end{listing}
-
-This is nearly the same as the Simple OOTA cycle and causality test
-case~4, with the addition of \co{P2()}.
-Just as in those examples, the OOTA cycle cannot be realized by
-compilers that treat atomic objects as volatile or that do
-single-thread analysis and treat atomic objects as quasi volatile.
-
-\subsection{Causality Test Case 20}
-\label{app:Causality Test Case 20}
-
-Listing~\ref{lst:Causality Test Case 20}
-shows causality test case 20, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-20.litmus @@
-\caption{Causality Test Case 20}
-\label{lst:Causality Test Case 20}
-\end{listing}
-
-This is the same as causality test case~19, even to the OOTA cycle
-reported by \co{herd7}, except that line~19 tests
-\co{r3} for equality to zero rather than non-equality to 42.
-The same analysis applies.
-
-\clearpage
-
-\section{Illustrative Litmus Tests}
-\label{app:Illustrative Litmus Tests}
-
-These litmus tests helped illuminate important aspects of the problem
-of defining sdep and identifying OOTA cycles.
-This appendix evaluates them using the \co{herd7} tool and also using
-manual analysis.
-
-\subsection{Simple Reordering as Litmus Test}
-\label{app:Simple Reordering as Litmus Test}
-
-Listing~\ref{lst:Simple Reordering as Litmus Test}
-shows how
-Listing~\ref{lst:Simple Reordering}
-can be recast into a \co{herd7} litmus test.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/simple-reordering.litmus @@
-\caption{Simple Reordering as Litmus Test}
-\label{lst:Simple Reordering as Litmus Test}
-\end{listing}
-
-\subsection{Non-LB Litmus Tests}
-\label{app:Non-LB Litmus Tests}
-
-Listing~\ref{lst:Non-LB Litmus Test}
-shows a litmus test that can produce an OOTA cycle, but that is not
-a member of the load-buffering (LB) class of litmus tests.
-Suppose that lines~9 and~10 both somehow return the value 21.
-Then line~12 will store the value 42 to \co{x}, which will cause
-line~17 to store 21 to \co{y} and line~22 to store 21 to \co{z},
-thus justifying the values initially loaded by lines~9 and~10.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-non-lb.litmus @@
-\caption{Non-LB Litmus Test}
-\label{lst:Non-LB Litmus Test}
-\end{listing}
-
-All thread's loads' return values affect the value stored, so all
-threads have semantic dependencies between their respective relaxed
-atomic loads and relaxed atomic stores.
-Therefore, an OOTA cycle cannot form in a correct real-world C++
-implementation.
-
-Note that in a C++ implementation that (erroneously) permitted an OOTA
-cycle to form, that cycle would involve both \co{r0} and \co{r1}.
-However, we conjecture that tools designed to find OOTA cycles could
-satisfy themselves with having found either of the two resulting cycles.
-
-\subsection{Multi-Cycle Litmus Tests}
-\label{app:Multi-Cycle Litmus Tests}
-
-Listing~\ref{lst:Multi-Cycle Litmus Test}
-shows an expansion to four threads of Causality Test Case 4, which is shown in
-Listing~\ref{lst:Causality Test Case 4}
-and discussed in
-Appendix~\ref{app:Causality Test Case 4}.
-The \co{herd7} tool shows multiple different possible OOTA cycles,
-as expected.
-However, the value returned by each thread's load directly affects the
-value stored, so there is a semantic dependency in each thread.
-Therefore, none of these OOTA cycles can happen in a correct real-world
-C++ implementation.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-two-source.litmus @@
-\caption{Multi-Cycle Litmus Test}
-\label{lst:Multi-Cycle Litmus Test}
-\end{listing}
-
-\subsection{Three-Process Version of JMM Causality Test Case 4}
-\label{app:Three-Process Version of JMM Causality Test Case 4}
-
-Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
-shows a different way of expanding
-Listing~\ref{lst:Causality Test Case 4}
-(discussed in
-Appendix~\ref{app:Causality Test Case 4})
-from two threads to three threads.
-The \co{herd7} tool finds OOTA values, as expected.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-3proc.litmus @@
-\caption{Three-Process Version of JMM Causality Test Case 4}
-\label{lst:Three-Process Version of JMM Causality Test Case 4}
-\end{listing}
-
-All threads have straightforward semantic dependencies from their
-relaxed atomic loads to their relaxed atomic stores, which prevents this
-OOTA cycle on correct real-world C++ implementations.
-
-\subsection{Why rfe Instead of Tried-And-True rf?}
-\label{app:Why rfe Instead of Tried-And-True rf?}
-
-Although the examples presented in
-Section~\ref{sec:OOTA: rf vs. rfe}
-(and shown as litmus tests in
-Listings~\ref{lst:OOTA With an Intra-Process Store-Load Link}
-and~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized})
-demonstrate that rf can be atemporal, they demonstrate neither
-OOTA cycles nor OOTA-like behavior in full C++ implementations.
-This appendix provides more involved examples that can be optimized
-so as to demonstrate OOTA-like behavior.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-3-2-proc.litmus @@
-\caption{OOTA With an Intra-Process Store-Load Link}
-\label{lst:OOTA With an Intra-Process Store-Load Link}
-\end{listing}
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-3-2-proc-opt.litmus @@
-\caption{OOTA With an Intra-Process Store-Load Link, Optimized}
-\label{lst:OOTA With an Intra-Process Store-Load Link, Optimized}
-\end{listing}
-
-Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
-revises
-Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
-so that the value stored to \co{y} is forced to be at most the
-value 17, and the value stored to \co{z} is forced to be at least the
-value 17.
-This of course means that the store to \co{z} must always be the value
-17, as can be seen from the \co{herd7} output.
-However, a single-thread-at-a-time analysis could not prove this.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-whyrfe-3.litmus @@
-\caption{Three-Process Version of JMM Causality Test Case 4 With Max and Min}
-\label{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
-\end{listing}
-
-Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
-further revises
-Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
-to flatten \co{PO()} and \co{P1()} into a single thread, so that the
-resulting \co{P0()} has an rfi link instead of an rfe link connecting
-the store to and load from \co{y}.
-This means that a C++ implementation could prove that the value 17 is
-always stored to \co{z} by examining \co{P0()} in isolation, given that
-it is also legal for a C++ implementation to replace the value loaded
-from \co{y} with the value just stored.\footnote{
-	This is not an option in C because atomic operations are volatile,
-	and thus observed behavior.}
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-whyrfe.litmus @@
-\caption{Why rfe Instead of Tried-And-True rf?}
-\label{lst:Why rfe Instead of Tried-And-True rf?}
-\end{listing}
-
-In the confines of the C++ abstract machine (omitting the no-OOTA
-restriction), this allows a value-17 OOTA cycle, courtesy of the fact
-that C++ respects neither control nor data dependencies.
-However, the \co{herd7} tool does not identify this cycle, perhaps due
-to its restrictions on its handling of arithmetic on OOTA values.
-
-In a real-world implementation that proves that the value 17 is always
-stored to \co{z}, there is no semantic dependency to
-\co{P0()}'s store to \co{z}, but there is to all of the other relaxed
-atomic stores.
-On the other hand, if the implementation does not prove the value
-stored to \co{z}, there will be a global-time delay between each of
-\co{P0()}'s loads and its store to \co{z}, preventing a value-17 OOTA
-cycle from forming.
-
-Either way, the value-17 OOTA cycle cannot occur in a correct real-world
-C++ implementation.
-Yes, an implementation that proves that the value 17 is always
-stored to \co{z} can satisfy the \co{exists} clause, but because
-there is no semantic dependency to \co{P0()}'s store to \co{z},
-this is not an OOTA cycle.
-
-But what if we chose to define an OOTA cycle as (sdep $\cup$ rf) instead
-of (sdep $\cup$ rfe)?
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-whyrfe-z17.litmus @@
-\caption{Why rfe Instead of Tried-And-True rf When z=17?}
-\label{lst:Why rfe Instead of Tried-And-True rf When z=17?}
-\end{listing}
-
-In that case, it would be necessary to separately consider the dependency
-from \co{P0()}'s load from \co{x} and \co{y} and that from its load
-from \co{y} and its store to \co{z}.
-The problem with this is shown in
-Listing~\ref{lst:Why rfe Instead of Tried-And-True rf When z=17?},
-which shows an equivalent program given the knowledge that the value~17
-will always be stored to \co{z}.
-Because all atomic operations are relaxed and because C++ does not
-respect control dependencies, the store to \co{z} can be moved up
-to the very beginning of \co{P0()}.
-This transformation causes the rfi link associated with \co{y} to
-go backwards in time, which is perfectly legitimate in this case,
-but which makes defining OOTA cycles in terms of (sdep $\cup$ rf)
-problematic.
-
-Again, in C, the load from \co{y} cannot be removed because it is
-volatile, and thus observable behavior.
-C++ could remove that load, but we left it in to demonstrate the all-17s
-outcome.
-
-This example demonstrates why OOTA cycles must be defined in terms
-of (sdep $\cup$ rfe), or in \co{herd7} nomenclature,
-\co{acyclic(sdep; rfe)}.
-
-\subsection{Inventing Atomic Loads}
-\label{app:Inventing Atomic Loads}
-
-This section contains several example litmus tests from the active
-mind of Alan Stern, expanding on the material presented in
-Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-no-invented-load.litmus @@
-\caption{No Invented Atomic Loads}
-\label{lst:No Invented Atomic Loads}
-\end{listing}
-
-First, consider Listing~\ref{lst:No Invented Atomic Loads}.
-At first glance, this might appear straightforward.
-\co{P0()}'s store to \co{z} clearly depends on its load from \co{y},
-and \co{P1()}'s store to \co{y} clearly depends on its load from \co{z}.
-If both dependencies are semantic dependency, the OOTA cycle cannot
-form.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-load-invented.litmus @@
-\caption{Atomic Loads Invented}
-\label{lst:Atomic Loads Invented}
-\end{listing}
-
-But is \co{P0()}'s dependency a semantic dependency?
-For example, is a real-world C++ implementation permitted to
-duplicate \co{P0()}'s load from \co{x} and then make use of both loads'
-values?
-If this is permitted, then the implementation would be within its rights
-to transform
-Listing~\ref{lst:No Invented Atomic Loads}
-into
-Listing~\ref{lst:Atomic Loads Invented}.
-
-Because the values of \co{r1}, \co{r2}, and \co{r3} are all known to be
-either zero or one, if the values returned by the two loads from \co{x}
-differ, then one of those two values must be equal to that from \co{y}.
-If the implementation is further permitted to make a runtime choice
-between the two values loaded from \co{x}, it could always choose the
-one that was equal to the value loaded from \co{y}, as shown by the
-\co{if} statement on line~13.
-This would mean that the store to \co{z} was always the constant 1,
-which would permit the store to be ordered before the load from \co{y},
-as shown in lines~14 and~15.
-Doing so permits the OOTA-like outcome of \co{r3 == r4 == 1}.
-
-The implementation might also drop the load from \co{y}, whose value
-is now unused except by the \co{exists} clause.
-However, the \co{exists} clause is considered to be observable behavior,
-which means that the load from \co{y} must remain.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-unused-load.litmus @@
-\caption{Unused Extra Atomic Load}
-\label{lst:Unused Extra Atomic Load}
-\end{listing}
-
-One possible reaction to this situation is to declare that useful
-real-world C++ implementations should never duplicate nonvolatile
-relaxed atomic loads.
-But suppose that the initial code was as shown in
-Listing~\ref{lst:Unused Extra Atomic Load}?\footnote{
-	Such unused loads can easily be generated by complex macros or
-	template metaprograms.}
-In this case, the transformation to
-Listing~\ref{lst:Atomic Loads Invented}
-would not require inventing a load, but instead merely inventing a use of
-the previously discarded value in \co{r0}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-invent-int-load.litmus @@
-\caption{Inventing Non-Boolean Atomic Loads}
-\label{lst:Unused Inventing Non-Boolean Atomic Loads}
-\end{listing}
-
-This is not restricted to booleans, as can be seen in
-Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}.
-This example is more straightforward in that it does not make quite so
-creative use of the multiple values loaded.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/inc.litmus @@
-\caption{Only Atomic Increment}
-\label{lst:Only Atomic Increment}
-\end{listing}
-
-But suppose that we drop the single-thread-analysis constraint for
-the moment, and suppose that the C++ implementation is able to prove
-that the only modification of a given atomic variable is to atomically
-increment it, as is the case for \co{y} in
-Listing~\ref{lst:Only Atomic Increment}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/inc-range.litmus @@
-\caption{Only Atomic Increment, Optimized}
-\label{lst:Only Atomic Increment, Optimized}
-\end{listing}
-
-Is a C++ implementation within its rights to behave as if the code
-shown in
-Listing~\ref{lst:Only Atomic Increment}
-was instead as shown in
-Listing~\ref{lst:Only Atomic Increment, Optimized}?
-One argument in favor is that the value of \co{y} must have
-passed through the value 42 in the case where \co{r0} and \co{r1}
-bracket that value.
-Arguments against might cite the added overhead of the invented load
-in Listing~\ref{lst:Only Atomic Increment, Optimized}, or that the
-possibility that the check for the value 42 is intended for statistical
-sampling.
-
-There are a number of possible reactions to these situations:
-
-\begin{enumerate}
-\item	Listing~\ref{lst:No Invented Atomic Loads}
-	looks like it has an OOTA cycle, but there is in fact no
-	semantic dependency.
-\item	Attempts to use real-world constraints to prevent OOTA cycles
-	are futile.
-\item	Implementations should not invent uses for values from
-	nonvolatile relaxed atomic loads.
-	If it turns out that there are useful optimizations that invent
-	such uses, then such optimizations must be applied only with
-	careful attention to the as-if rule.
-	Some might argue that inventing comparisons between a pair of
-	nonvolatile relaxed atomic loads from the same object should
-	be permitted only if the two loads could be reordered to be
-	adjacent to each other.
-\item	Any time an implementation might be tempted to invent uses from
-	values from nonvolatile relaxed atomic loads from the same
-	object, the loads should instead be fused into a single load.
-	This reaction might be attractive to those who carefully consider
-	the pointlessness of keeping two loads that return the same value
-	on the one hand or the cache-miss overhead incurred when closely
-	spaced loads return different values on the other.
-\item	Implementations that invent nonvolatile atomic loads or invent
-	new uses for nonvolatile atomic loads can adversely affect
-	observed behavior, for example, by introducing error into code
-	attempting to do statistical sampling.
-\item	What other interesting situations might come to light?
-\end{enumerate}
-
-There are situations where inventing loads from non-atomic objects
-is customary, for example, when hoisting a load out of a loop.
-In this case, implementations might hoist the load from x out of
-the loop:
-
-\begin{quote}
-\begin{verbatim}
-do_something_with(x);
-for (i = 0; i < limit i++)
-    y[i] += x;
-\end{verbatim}
-\end{quote}
-
-A C++ implementation might act as if the program had instead been this:
-
-\begin{quote}
-\begin{verbatim}
-r1 = x;
-do_something_with(r1);
-for (i = 0; i < limit i++)
-    y[i] += r1;
-\end{verbatim}
-\end{quote}
-
-Note that passing \co{x} to the call to \co{do_something_with()} is important
-because this allows the C++ implementation to assume that accesses to \co{x}
-will not result in a data race.
-Without this assumption, the above transformation might introduce a data
-race when \co{limit} is less than or equal to zero.
-
-But if \co{x} is a nonvolatile atomic, then there are no worries about
-data races.
-In theory, the implementation could make the above transformation without
-worry, especially given that the value loaded from \co{x} is discarded in
-the case where \co{limit} is less than or equal to zero.
-Except that the data-race worry is replaced by a cache-miss worry.
-After all, optimizations are supposed to speed things up.
-This means that a C++ implementation might be wise to adhere to the
-same restriction that avoids data races for loads from non-atomic
-variables in order to avoid expensive cache misses.
-
-\subsection{Non-Trivial Semantic Dependencies}
-\label{app:Non-Trivial Semantic Dependencies}
-
-Most of the examples in this paper involve simple semantic dependencies
-connecting a single load to a single store.
-One exception appears in
-Appendix~\ref{app:Non-LB Litmus Tests},
-but this section will present a more difficult example.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult-0.litmus @@
-\caption{OOTA Multiplication Example, Initial Value Zero}
-\label{lst:OOTA Multiplication Example, Initial Value Zero}
-\end{listing}
-
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-shows an example that builds on the multiplication-by-zero
-discussion in
-Section~\ref{sec:Semantic Dependencies and Source Code}.
-Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
-\co{x} and \co{y} on lines~9 and~10, respectively.
-Similarly, lines~16 and~17 store to \co{x} and \co{y}, respectively,
-the value that line~15 loads from \co{z}.
-Because all three variables are initialized to zero, and because loads
-can be ordered before any store, there is an execution in which all
-three loads return the value zero, in which case all three stores will
-of course store the value zero.
-In any related execution where the value loaded from \co{y} remains zero,
-nonzero values loaded from \co{x} cannot affect the value stored by
-line~11 to \co{z}.
-Only executions which load different (that is, nonzero) values from
-both \co{x} and \co{y} can cause the value stored to \co{z} to change
-from zero to a nonzero value.
-
-Therefore, as noted in
-Section~\ref{sec:Semantic Dependencies and Source Code},
-in an execution where the loads from both \co{x} and \co{y} return zero,
-there is no semantic dependency from the load from \co{x} to the
-store to \co{z}, nor is there a semantic dependency from the load
-from \co{y} to the store to \co{z}.
-However, there \emph{is} a semantic dependency from the \emph{combination}
-of the loads from \co{x} and \co{y} to the store to \co{z}.
-For example, the (forbidden) execution that satisfies the \co{exists}
-clause, which would be an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult-0-cond.litmus @@
-\caption{OOTA Conditional Multiplication Example, Initial Value Zero}
-\label{lst:OOTA Conditional Multiplication Example, Initial Value Zero}
-\end{listing}
-
-For further evidence that there is no semantic dependency from \co{y}
-to \co{z} in executions where the load from \co{x} is zero, see the
-equivalent\footnote{
-	But equivalent only because the atomics are all nonvolatile!}
-program shown in
-Listing~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero}.
-Here, there is not even a load from \co{y} in executions where the
-load from \co{x} returns zero, so there cannot possibly be a semantic
-dependency involving this non-existent load.
-A similar transformation would load \co{x} only in executions where the
-load from \co{y} returned nonzero, which similarly emphasizes the lack
-of any semantic dependency between \co{x} and \co{z}.
-
-In both
-Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-and~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero},
-an OOTA cycle (such as the one indicated by their \co{exists} clauses)
-would need to affect both the \co{x} and the \co{y} components of this
-compound semantic dependency.
-However, we again conjecture that tools designed to find OOTA cycles could
-satisfy themselves with having found either of the two resulting cycles.
-This situation poses a challenge to current work on semantic dependencies,
-which typically restricts such dependencies to a single load and store.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult-1.litmus @@
-\caption{OOTA Multiplication Example, Initial Value 1}
-\label{lst:OOTA Multiplication Example, Initial Value 1}
-\end{listing}
-
-In contrast, consider
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1},
-which differs from
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-only in the initial values (1 instead of 0) and the \co{exists}
-clause (0 instead of 1).
-Note that an execution satisfying the \co{exists} clause is again
-a (forbidden) OOTA cycle.
-
-Starting with the intuitive execution where the values loaded from \co{x}
-and \co{y} on lines~9 and~10 are the value 1 (that is, nonzero), executions
-that differ only in the value loaded from either \co{x} or \co{y} will
-still affect the value stored to \co{z} by line~11.
-There is therefore one semantic dependency from the load from \co{x}
-to the store to \co{z} and another separate semantic dependency from
-the load from \co{y} to the store to \co{z}, which matches typical
-definitions of semantic dependency.
-Yet
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-and
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
-can be thought of as being merely two different executions of the same
-program.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult3-0.litmus @@
-\caption{OOTA Three-Factor Multiplication Example, Initial Value Zero}
-\label{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
-\end{listing}
-
-Because the only two solutions to $x^2 = x$ are the values zero and one,
-Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-and~\ref{lst:OOTA Multiplication Example, Initial Value 1}
-each have but one OOTA cycle, and an execution having nonzero non-one
-initial values could have at most two OOTA cycles, all of which are
-(again) forbidden by the C++ language.
-Listing~\ref{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
-addressses this limitation by multiplying the values obtained from three
-different loads.
-Because $x^3 = x$ has three solutions ($\left\{ -1, 0, 1 \right\}$),
-this execution could have up to three (yet again, forbidden) OOTA cycles.
-
-Much more elaborate examples can be constructed, for example,
-involving sums of products of values loaded.
-Reverting back to short-form code fragments, consider the following:
-
-\begin{quote}
-\begin{verbatim}
-z = w * x + w * y + w;
-\end{verbatim}
-\end{quote}
-
-This has the following semantic dependencies when all of \co{w},
-\co{x}, and \co{y} are zero:
-
-\begin{enumerate}
-\item	From the combination of \co{w} and \co{x} to \co{z}.
-\item	From the combination of \co{w} and \co{y} to \co{z}.
-\item	From \co{w} to \co{z}.
-\end{enumerate}
-
-Despite appearances, these are not redundant.
-The last semantic dependency does not imply the first, because given
-that \co{w} becomes nonzero, a change in \co{x} further changes
-the value stored to \co{z}.
-Similar reasoning shows that the last semantic dependency does not
-imply the second.
-Much more elaborate examples can easily be constructed.
-
-This raises the question of how to determine the set of semantic
-dependencies in a given execution of a given fragment of code.
-The short answer is that when running on a real-world system, 
-any C++ program is a finite-state machine, which, unlike Turing-complete
-systems, can be analyzed.
-% @@@ reference to cbmc appendix.
-
-One might ask why semantic dependencies can be nontrivial when hardware
-address, control, and data dependencies are much more straightforward.
-Answers include:
-(1)~Respecting dependencies even when they reduce to a constant,
-(2)~Less-aggressive optimizations, and
-(3)~Lack of undefined behavior.
-
-This situation further underscores the per-execution nature of semantic
-dependencies as well as the need for an improved definition of ``semantic
-dependency''.
-We propose expanding the traditional definition to include dependencies
-extending from groups of loads to a single store.
-
-\subsection{OOTA From Undefined-Behavior Optimizations}
-\label{app:OOTA From Undefined-Behavior Optimizations}
-
-Many C++ implementations carry out optimizations based on the assumption
-that UB cannot happen, and these optimizations can result in OOTA cycles,
-with a simple example shown in
-Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-div-ub.litmus @@
-\caption{OOTA Due To Divide-By-Zero UB}
-\label{lst:OOTA Due To Divide-By-Zero UB}
-\end{listing}
-
-On line~14 of this listing, the value of \co{r2} must be either zero
-or one due to the comparison on line~13.
-If it is the value one, then there is a divide-by-zero error, which
-enables the C++ implementation to assume that the value of \co{r1} is
-always zero.
-This in turn allows the implementation to behave as if the expression
-stored to \co{y} on line~14 was the constant 1, at which point there
-is no semantic dependency from the load from \co{x} on line~13 to the
-store to \co{y} on line~14.
-
-The implementation is then within its rights to act as if lines~13 and~14
-were interchanged.
-At which point, the order of execution could be as follows:
-
-\begin{enumerate}
-\item	Line~14 stores the constant 1 to \co{y}.
-\item	Line~8 loads the constant 1 from \co{y}, setting \co{r1} to 1.
-\item	Line~9 stores the constant 1 to \co{x}.
-\item	Line~8 loads the constant 1 from \co{x}, setting \co{r2} to 1.
-\end{enumerate}
-
-Please note that the OOTA-like cycle called out in the \co{exists}
-clause can happen even though the program never actually executes any
-code that would have resulted in division-by-zero UB.
-It is tempting to consider using techniques
-% Appendix~\ref{app:Aside on Undefined Behavior}
-that prevent back-propagation of UB, but these are ineffective in this
-case because the C++ implementation can detect and exploit UB when
-considering line~14 in isolation.
-Once the constant 1 is substituted for line~14's expression, the
-hardware is free to reorder lines~13 and~14 even if a compiler-based
-C++ implementation refrains from doing so.
-Note that this OOTA-like cycle can appear even if all of the atomic
-operations in
-Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}
-were converted to be volatile.
-
-Therefore, this paper focuses on UB-free programs, rare though these
-beasts might be among large-scale C++ programs running in production.
-
-\subsection{Semantic Dependencies and \co{volatile}}
-\label{app:Semantic Dependencies and volatile}
-
-The Linux kernel uses volatile accesses to constrain the compiler,
-and it is worth looking at the example shown in
-Listing~\ref{lst:Non-Volatile Accesses and Dependencies}
-to see how this works.
-
-\begin{listing}
-\begin{verbatim}
- 1 atomic<int> x, y;
- 2
- 3 void thread1() {
- 4   int r1 = x.load(memory_order_relaxed);
- 5   y.store(r1, memory_order_relaxed);
- 6 }
-\end{verbatim}
-\caption{Non-Volatile Accesses and Dependencies}
-\label{lst:Non-Volatile Accesses and Dependencies}
-\end{listing}
-
-Because both \co{x} and \co{y} are nonvolatile, the compiler is
-free to assume that the value loaded by line~4 will be some value
-stored to \co{x} or its initial value.
-Because there are no stores to \co{x}, the only remaining possibility
-is the default-initialized value of zero.
-Therefore, the C++ implementation is within its rights to substitute
-the constant zero for \co{r1} on line~5, eliminating the semantic
-dependency that would otherwise extend from line~4 to line~5.
-This in turn enables the implementation to reorder line~5 to
-precede line~4.
-
-\begin{listing}
-\begin{verbatim}
- 1 atomic<volatile int> x, y;
- 2
- 3 void thread1() {
- 4   int r1 = x.load(memory_order_relaxed);
- 5   y.store(r1, memory_order_relaxed);
- 6 }
-\end{verbatim}
-\caption{Volatile Accesses and Dependencies}
-\label{lst:Volatile Accesses and Dependencies}
-\end{listing}
-
-One way to preserve this dependency is use of \co{volatile}, as shown in
-Listing~\ref{lst:Volatile Accesses and Dependencies}.
-Now the implementation is forbidden from assuming that the value loaded
-by line~4 has any relation to any stores to or initialization of \co{x},
-even if \co{x} is placed in normal memory.
-
-In this case, the \co{volatile} keyword alerts the implementation to
-the possibility of interfering accesses to \co{x} from outside the
-program, for example, due to \co{x} being:
-
-\begin{itemize}
-\item	Allocated in an MMIO region of the address space.
-\item	Modified by an unknown-to-the-implementation thread
-	or signal handler.
-\item	Subject to I/O-device DMA operations.
-\item	Modified by debugger commands.
-\item	Modified by accesses to facilities such as \path{/dev/mem}.
-\item	Modified by as-yet-unwritten dynamically linked libraries.
-\item	Modified by some other mechanism of the reader's choosing.
-\end{itemize}
-
-These possibilities force the implementation to preserve the semantic
-dependency from line~4 to line~5.
-They can also prevent the formation of OOTA cycles local to the
-implementation because such inteference means that the volatile loads
-might be reading from something that the implementation knows nothing
-about.
-
-The developer is free to arrange for \co{x} to be free of any
-interference, thus obtaining predictable behavior along with the
-preserved semantic dependency.
-However, this example invalidates the rough definition of semantic
-dependency from
-Section~\ref{sec:Simple OOTA Cycle}
-because the value loaded from \co{x} will always be the initial value
-of zero, so that it is at best dubious to talk about any changes in
-the value loaded.
-
-One way forward is to assume that interference could happen when
-determining which dependencies are semantic, and then do further analysis
-using that determination.
-
-\clearpage
 
@@
 
+\clearpage
+
+\section{Illustrative Litmus Tests}
+\label{app:Illustrative Litmus Tests}
+
+These litmus tests helped illuminate important aspects of the problem
+of defining sdep and identifying OOTA cycles.
+This appendix evaluates them using the \co{herd7} tool and also using
+manual analysis.
+
+\subsection{Simple Reordering as Litmus Test}
+\label{app:Simple Reordering as Litmus Test}
+
+Listing~\ref{lst:Simple Reordering as Litmus Test}
+shows how
+Listing~\ref{lst:Simple Reordering}
+can be recast into a \co{herd7} litmus test.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/simple-reordering.litmus @@
+\caption{Simple Reordering as Litmus Test}
+\label{lst:Simple Reordering as Litmus Test}
+\end{listing}
+
+\subsection{Non-LB Litmus Tests}
+\label{app:Non-LB Litmus Tests}
+
+Listing~\ref{lst:Non-LB Litmus Test}
+shows a litmus test that can produce an OOTA cycle, but that is not
+a member of the load-buffering (LB) class of litmus tests.
+Suppose that lines~9 and~10 both somehow return the value 21.
+Then line~12 will store the value 42 to \co{x}, which will cause
+line~17 to store 21 to \co{y} and line~22 to store 21 to \co{z},
+thus justifying the values initially loaded by lines~9 and~10.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-non-lb.litmus @@
+\caption{Non-LB Litmus Test}
+\label{lst:Non-LB Litmus Test}
+\end{listing}
+
+All thread's loads' return values affect the value stored, so all
+threads have semantic dependencies between their respective relaxed
+atomic loads and relaxed atomic stores.
+Therefore, an OOTA cycle cannot form in a correct real-world C++
+implementation.
+
+Note that in a C++ implementation that (erroneously) permitted an OOTA
+cycle to form, that cycle would involve both \co{r0} and \co{r1}.
+However, we conjecture that tools designed to find OOTA cycles could
+satisfy themselves with having found either of the two resulting cycles.
+
+\subsection{Multi-Cycle Litmus Tests}
+\label{app:Multi-Cycle Litmus Tests}
+
+Listing~\ref{lst:Multi-Cycle Litmus Test}
+shows an expansion to four threads of Causality Test Case 4, which is shown in
+Listing~\ref{lst:Causality Test Case 4}
+and discussed in
+Appendix~\ref{app:Causality Test Case 4}.
+The \co{herd7} tool shows multiple different possible OOTA cycles,
+as expected.
+However, the value returned by each thread's load directly affects the
+value stored, so there is a semantic dependency in each thread.
+Therefore, none of these OOTA cycles can happen in a correct real-world
+C++ implementation.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-two-source.litmus @@
+\caption{Multi-Cycle Litmus Test}
+\label{lst:Multi-Cycle Litmus Test}
+\end{listing}
+
+\subsection{Three-Process Version of JMM Causality Test Case 4}
+\label{app:Three-Process Version of JMM Causality Test Case 4}
+
+Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
+shows a different way of expanding
+Listing~\ref{lst:Causality Test Case 4}
+(discussed in
+Appendix~\ref{app:Causality Test Case 4})
+from two threads to three threads.
+The \co{herd7} tool finds OOTA values, as expected.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-3proc.litmus @@
+\caption{Three-Process Version of JMM Causality Test Case 4}
+\label{lst:Three-Process Version of JMM Causality Test Case 4}
+\end{listing}
+
+All threads have straightforward semantic dependencies from their
+relaxed atomic loads to their relaxed atomic stores, which prevents this
+OOTA cycle on correct real-world C++ implementations.
+
+\subsection{Why rfe Instead of Tried-And-True rf?}
+\label{app:Why rfe Instead of Tried-And-True rf?}
+
+Although the examples presented in
+Section~\ref{sec:OOTA: rf vs. rfe}
+(and shown as litmus tests in
+Listings~\ref{lst:OOTA With an Intra-Process Store-Load Link}
+and~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized})
+demonstrate that rf can be atemporal, they demonstrate neither
+OOTA cycles nor OOTA-like behavior in full C++ implementations.
+This appendix provides more involved examples that can be optimized
+so as to demonstrate OOTA-like behavior.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-3-2-proc.litmus @@
+\caption{OOTA With an Intra-Process Store-Load Link}
+\label{lst:OOTA With an Intra-Process Store-Load Link}
+\end{listing}
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-3-2-proc-opt.litmus @@
+\caption{OOTA With an Intra-Process Store-Load Link, Optimized}
+\label{lst:OOTA With an Intra-Process Store-Load Link, Optimized}
+\end{listing}
+
+Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
+revises
+Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
+so that the value stored to \co{y} is forced to be at most the
+value 17, and the value stored to \co{z} is forced to be at least the
+value 17.
+This of course means that the store to \co{z} must always be the value
+17, as can be seen from the \co{herd7} output.
+However, a single-thread-at-a-time analysis could not prove this.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-whyrfe-3.litmus @@
+\caption{Three-Process Version of JMM Causality Test Case 4 With Max and Min}
+\label{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
+\end{listing}
+
+Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
+further revises
+Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
+to flatten \co{PO()} and \co{P1()} into a single thread, so that the
+resulting \co{P0()} has an rfi link instead of an rfe link connecting
+the store to and load from \co{y}.
+This means that a C++ implementation could prove that the value 17 is
+always stored to \co{z} by examining \co{P0()} in isolation, given that
+it is also legal for a C++ implementation to replace the value loaded
+from \co{y} with the value just stored.\footnote{
+	This is not an option in C because atomic operations are volatile,
+	and thus observed behavior.}
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-whyrfe.litmus @@
+\caption{Why rfe Instead of Tried-And-True rf?}
+\label{lst:Why rfe Instead of Tried-And-True rf?}
+\end{listing}
+
+In the confines of the C++ abstract machine (omitting the no-OOTA
+restriction), this allows a value-17 OOTA cycle, courtesy of the fact
+that C++ respects neither control nor data dependencies.
+However, the \co{herd7} tool does not identify this cycle, perhaps due
+to its restrictions on its handling of arithmetic on OOTA values.
+
+In a real-world implementation that proves that the value 17 is always
+stored to \co{z}, there is no semantic dependency to
+\co{P0()}'s store to \co{z}, but there is to all of the other relaxed
+atomic stores.
+On the other hand, if the implementation does not prove the value
+stored to \co{z}, there will be a global-time delay between each of
+\co{P0()}'s loads and its store to \co{z}, preventing a value-17 OOTA
+cycle from forming.
+
+Either way, the value-17 OOTA cycle cannot occur in a correct real-world
+C++ implementation.
+Yes, an implementation that proves that the value 17 is always
+stored to \co{z} can satisfy the \co{exists} clause, but because
+there is no semantic dependency to \co{P0()}'s store to \co{z},
+this is not an OOTA cycle.
+
+But what if we chose to define an OOTA cycle as (sdep $\cup$ rf) instead
+of (sdep $\cup$ rfe)?
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-whyrfe-z17.litmus @@
+\caption{Why rfe Instead of Tried-And-True rf When z=17?}
+\label{lst:Why rfe Instead of Tried-And-True rf When z=17?}
+\end{listing}
+
+In that case, it would be necessary to separately consider the dependency
+from \co{P0()}'s load from \co{x} and \co{y} and that from its load
+from \co{y} and its store to \co{z}.
+The problem with this is shown in
+Listing~\ref{lst:Why rfe Instead of Tried-And-True rf When z=17?},
+which shows an equivalent program given the knowledge that the value~17
+will always be stored to \co{z}.
+Because all atomic operations are relaxed and because C++ does not
+respect control dependencies, the store to \co{z} can be moved up
+to the very beginning of \co{P0()}.
+This transformation causes the rfi link associated with \co{y} to
+go backwards in time, which is perfectly legitimate in this case,
+but which makes defining OOTA cycles in terms of (sdep $\cup$ rf)
+problematic.
+
+Again, in C, the load from \co{y} cannot be removed because it is
+volatile, and thus observable behavior.
+C++ could remove that load, but we left it in to demonstrate the all-17s
+outcome.
+
+This example demonstrates why OOTA cycles must be defined in terms
+of (sdep $\cup$ rfe), or in \co{herd7} nomenclature,
+\co{acyclic(sdep; rfe)}.
+
+\subsection{Inventing Atomic Loads}
+\label{app:Inventing Atomic Loads}
+
+This section contains several example litmus tests from the active
+mind of Alan Stern, expanding on the material presented in
+Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-no-invented-load.litmus @@
+\caption{No Invented Atomic Loads}
+\label{lst:No Invented Atomic Loads}
+\end{listing}
+
+First, consider Listing~\ref{lst:No Invented Atomic Loads}.
+At first glance, this might appear straightforward.
+\co{P0()}'s store to \co{z} clearly depends on its load from \co{y},
+and \co{P1()}'s store to \co{y} clearly depends on its load from \co{z}.
+If both dependencies are semantic dependency, the OOTA cycle cannot
+form.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-load-invented.litmus @@
+\caption{Atomic Loads Invented}
+\label{lst:Atomic Loads Invented}
+\end{listing}
+
+But is \co{P0()}'s dependency a semantic dependency?
+For example, is a real-world C++ implementation permitted to
+duplicate \co{P0()}'s load from \co{x} and then make use of both loads'
+values?
+If this is permitted, then the implementation would be within its rights
+to transform
+Listing~\ref{lst:No Invented Atomic Loads}
+into
+Listing~\ref{lst:Atomic Loads Invented}.
+
+Because the values of \co{r1}, \co{r2}, and \co{r3} are all known to be
+either zero or one, if the values returned by the two loads from \co{x}
+differ, then one of those two values must be equal to that from \co{y}.
+If the implementation is further permitted to make a runtime choice
+between the two values loaded from \co{x}, it could always choose the
+one that was equal to the value loaded from \co{y}, as shown by the
+\co{if} statement on line~13.
+This would mean that the store to \co{z} was always the constant 1,
+which would permit the store to be ordered before the load from \co{y},
+as shown in lines~14 and~15.
+Doing so permits the OOTA-like outcome of \co{r3 == r4 == 1}.
+
+The implementation might also drop the load from \co{y}, whose value
+is now unused except by the \co{exists} clause.
+However, the \co{exists} clause is considered to be observable behavior,
+which means that the load from \co{y} must remain.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-unused-load.litmus @@
+\caption{Unused Extra Atomic Load}
+\label{lst:Unused Extra Atomic Load}
+\end{listing}
+
+One possible reaction to this situation is to declare that useful
+real-world C++ implementations should never duplicate nonvolatile
+relaxed atomic loads.
+But suppose that the initial code was as shown in
+Listing~\ref{lst:Unused Extra Atomic Load}?\footnote{
+	Such unused loads can easily be generated by complex macros or
+	template metaprograms.}
+In this case, the transformation to
+Listing~\ref{lst:Atomic Loads Invented}
+would not require inventing a load, but instead merely inventing a use of
+the previously discarded value in \co{r0}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-invent-int-load.litmus @@
+\caption{Inventing Non-Boolean Atomic Loads}
+\label{lst:Unused Inventing Non-Boolean Atomic Loads}
+\end{listing}
+
+This is not restricted to booleans, as can be seen in
+Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}.
+This example is more straightforward in that it does not make quite so
+creative use of the multiple values loaded.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/inc.litmus @@
+\caption{Only Atomic Increment}
+\label{lst:Only Atomic Increment}
+\end{listing}
+
+But suppose that we drop the single-thread-analysis constraint for
+the moment, and suppose that the C++ implementation is able to prove
+that the only modification of a given atomic variable is to atomically
+increment it, as is the case for \co{y} in
+Listing~\ref{lst:Only Atomic Increment}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/inc-range.litmus @@
+\caption{Only Atomic Increment, Optimized}
+\label{lst:Only Atomic Increment, Optimized}
+\end{listing}
+
+Is a C++ implementation within its rights to behave as if the code
+shown in
+Listing~\ref{lst:Only Atomic Increment}
+was instead as shown in
+Listing~\ref{lst:Only Atomic Increment, Optimized}?
+One argument in favor is that the value of \co{y} must have
+passed through the value 42 in the case where \co{r0} and \co{r1}
+bracket that value.
+Arguments against might cite the added overhead of the invented load
+in Listing~\ref{lst:Only Atomic Increment, Optimized}, or that the
+possibility that the check for the value 42 is intended for statistical
+sampling.
+
+There are a number of possible reactions to these situations:
+
+\begin{enumerate}
+\item	Listing~\ref{lst:No Invented Atomic Loads}
+	looks like it has an OOTA cycle, but there is in fact no
+	semantic dependency.
+\item	Attempts to use real-world constraints to prevent OOTA cycles
+	are futile.
+\item	Implementations should not invent uses for values from
+	nonvolatile relaxed atomic loads.
+	If it turns out that there are useful optimizations that invent
+	such uses, then such optimizations must be applied only with
+	careful attention to the as-if rule.
+	Some might argue that inventing comparisons between a pair of
+	nonvolatile relaxed atomic loads from the same object should
+	be permitted only if the two loads could be reordered to be
+	adjacent to each other.
+\item	Any time an implementation might be tempted to invent uses from
+	values from nonvolatile relaxed atomic loads from the same
+	object, the loads should instead be fused into a single load.
+	This reaction might be attractive to those who carefully consider
+	the pointlessness of keeping two loads that return the same value
+	on the one hand or the cache-miss overhead incurred when closely
+	spaced loads return different values on the other.
+\item	Implementations that invent nonvolatile atomic loads or invent
+	new uses for nonvolatile atomic loads can adversely affect
+	observed behavior, for example, by introducing error into code
+	attempting to do statistical sampling.
+\item	What other interesting situations might come to light?
+\end{enumerate}
+
+There are situations where inventing loads from non-atomic objects
+is customary, for example, when hoisting a load out of a loop.
+In this case, implementations might hoist the load from x out of
+the loop:
+
+\begin{quote}
+\begin{verbatim}
+do_something_with(x);
+for (i = 0; i < limit i++)
+    y[i] += x;
+\end{verbatim}
+\end{quote}
+
+A C++ implementation might act as if the program had instead been this:
+
+\begin{quote}
+\begin{verbatim}
+r1 = x;
+do_something_with(r1);
+for (i = 0; i < limit i++)
+    y[i] += r1;
+\end{verbatim}
+\end{quote}
+
+Note that passing \co{x} to the call to \co{do_something_with()} is important
+because this allows the C++ implementation to assume that accesses to \co{x}
+will not result in a data race.
+Without this assumption, the above transformation might introduce a data
+race when \co{limit} is less than or equal to zero.
+
+But if \co{x} is a nonvolatile atomic, then there are no worries about
+data races.
+In theory, the implementation could make the above transformation without
+worry, especially given that the value loaded from \co{x} is discarded in
+the case where \co{limit} is less than or equal to zero.
+Except that the data-race worry is replaced by a cache-miss worry.
+After all, optimizations are supposed to speed things up.
+This means that a C++ implementation might be wise to adhere to the
+same restriction that avoids data races for loads from non-atomic
+variables in order to avoid expensive cache misses.
+
+\subsection{Non-Trivial Semantic Dependencies}
+\label{app:Non-Trivial Semantic Dependencies}
+
+Most of the examples in this paper involve simple semantic dependencies
+connecting a single load to a single store.
+One exception appears in
+Appendix~\ref{app:Non-LB Litmus Tests},
+but this section will present a more difficult example.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult-0.litmus @@
+\caption{OOTA Multiplication Example, Initial Value Zero}
+\label{lst:OOTA Multiplication Example, Initial Value Zero}
+\end{listing}
+
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+shows an example that builds on the multiplication-by-zero
+discussion in
+Section~\ref{sec:Semantic Dependencies and Source Code}.
+Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
+\co{x} and \co{y} on lines~9 and~10, respectively.
+Similarly, lines~16 and~17 store to \co{x} and \co{y}, respectively,
+the value that line~15 loads from \co{z}.
+Because all three variables are initialized to zero, and because loads
+can be ordered before any store, there is an execution in which all
+three loads return the value zero, in which case all three stores will
+of course store the value zero.
+In any related execution where the value loaded from \co{y} remains zero,
+nonzero values loaded from \co{x} cannot affect the value stored by
+line~11 to \co{z}.
+Only executions which load different (that is, nonzero) values from
+both \co{x} and \co{y} can cause the value stored to \co{z} to change
+from zero to a nonzero value.
+
+Therefore, as noted in
+Section~\ref{sec:Semantic Dependencies and Source Code},
+in an execution where the loads from both \co{x} and \co{y} return zero,
+there is no semantic dependency from the load from \co{x} to the
+store to \co{z}, nor is there a semantic dependency from the load
+from \co{y} to the store to \co{z}.
+However, there \emph{is} a semantic dependency from the \emph{combination}
+of the loads from \co{x} and \co{y} to the store to \co{z}.
+For example, the (forbidden) execution that satisfies the \co{exists}
+clause, which would be an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult-0-cond.litmus @@
+\caption{OOTA Conditional Multiplication Example, Initial Value Zero}
+\label{lst:OOTA Conditional Multiplication Example, Initial Value Zero}
+\end{listing}
+
+For further evidence that there is no semantic dependency from \co{y}
+to \co{z} in executions where the load from \co{x} is zero, see the
+equivalent\footnote{
+	But equivalent only because the atomics are all nonvolatile!}
+program shown in
+Listing~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero}.
+Here, there is not even a load from \co{y} in executions where the
+load from \co{x} returns zero, so there cannot possibly be a semantic
+dependency involving this non-existent load.
+A similar transformation would load \co{x} only in executions where the
+load from \co{y} returned nonzero, which similarly emphasizes the lack
+of any semantic dependency between \co{x} and \co{z}.
+
+In both
+Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+and~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero},
+an OOTA cycle (such as the one indicated by their \co{exists} clauses)
+would need to affect both the \co{x} and the \co{y} components of this
+compound semantic dependency.
+However, we again conjecture that tools designed to find OOTA cycles could
+satisfy themselves with having found either of the two resulting cycles.
+This situation poses a challenge to current work on semantic dependencies,
+which typically restricts such dependencies to a single load and store.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult-1.litmus @@
+\caption{OOTA Multiplication Example, Initial Value 1}
+\label{lst:OOTA Multiplication Example, Initial Value 1}
+\end{listing}
+
+In contrast, consider
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1},
+which differs from
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+only in the initial values (1 instead of 0) and the \co{exists}
+clause (0 instead of 1).
+Note that an execution satisfying the \co{exists} clause is again
+a (forbidden) OOTA cycle.
+
+Starting with the intuitive execution where the values loaded from \co{x}
+and \co{y} on lines~9 and~10 are the value 1 (that is, nonzero), executions
+that differ only in the value loaded from either \co{x} or \co{y} will
+still affect the value stored to \co{z} by line~11.
+There is therefore one semantic dependency from the load from \co{x}
+to the store to \co{z} and another separate semantic dependency from
+the load from \co{y} to the store to \co{z}, which matches typical
+definitions of semantic dependency.
+Yet
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+and
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
+can be thought of as being merely two different executions of the same
+program.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult3-0.litmus @@
+\caption{OOTA Three-Factor Multiplication Example, Initial Value Zero}
+\label{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
+\end{listing}
+
+Because the only two solutions to $x^2 = x$ are the values zero and one,
+Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+and~\ref{lst:OOTA Multiplication Example, Initial Value 1}
+each have but one OOTA cycle, and an execution having nonzero non-one
+initial values could have at most two OOTA cycles, all of which are
+(again) forbidden by the C++ language.
+Listing~\ref{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
+addressses this limitation by multiplying the values obtained from three
+different loads.
+Because $x^3 = x$ has three solutions ($\left\{ -1, 0, 1 \right\}$),
+this execution could have up to three (yet again, forbidden) OOTA cycles.
+
+Much more elaborate examples can be constructed, for example,
+involving sums of products of values loaded.
+Reverting back to short-form code fragments, consider the following:
+
+\begin{quote}
+\begin{verbatim}
+z = w * x + w * y + w;
+\end{verbatim}
+\end{quote}
+
+This has the following semantic dependencies when all of \co{w},
+\co{x}, and \co{y} are zero:
+
+\begin{enumerate}
+\item	From the combination of \co{w} and \co{x} to \co{z}.
+\item	From the combination of \co{w} and \co{y} to \co{z}.
+\item	From \co{w} to \co{z}.
+\end{enumerate}
+
+Despite appearances, these are not redundant.
+The last semantic dependency does not imply the first, because given
+that \co{w} becomes nonzero, a change in \co{x} further changes
+the value stored to \co{z}.
+Similar reasoning shows that the last semantic dependency does not
+imply the second.
+Much more elaborate examples can easily be constructed.
+
+This raises the question of how to determine the set of semantic
+dependencies in a given execution of a given fragment of code.
+The short answer is that when running on a real-world system, 
+any C++ program is a finite-state machine, which, unlike Turing-complete
+systems, can be analyzed.
+% @@@ reference to cbmc appendix.
+
+One might ask why semantic dependencies can be nontrivial when hardware
+address, control, and data dependencies are much more straightforward.
+Answers include:
+(1)~Respecting dependencies even when they reduce to a constant,
+(2)~Less-aggressive optimizations, and
+(3)~Lack of undefined behavior.
+
+This situation further underscores the per-execution nature of semantic
+dependencies as well as the need for an improved definition of ``semantic
+dependency''.
+We propose expanding the traditional definition to include dependencies
+extending from groups of loads to a single store.
+
+\subsection{OOTA From Undefined-Behavior Optimizations}
+\label{app:OOTA From Undefined-Behavior Optimizations}
+
+Many C++ implementations carry out optimizations based on the assumption
+that UB cannot happen, and these optimizations can result in OOTA cycles,
+with a simple example shown in
+Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-div-ub.litmus @@
+\caption{OOTA Due To Divide-By-Zero UB}
+\label{lst:OOTA Due To Divide-By-Zero UB}
+\end{listing}
+
+On line~14 of this listing, the value of \co{r2} must be either zero
+or one due to the comparison on line~13.
+If it is the value one, then there is a divide-by-zero error, which
+enables the C++ implementation to assume that the value of \co{r1} is
+always zero.
+This in turn allows the implementation to behave as if the expression
+stored to \co{y} on line~14 was the constant 1, at which point there
+is no semantic dependency from the load from \co{x} on line~13 to the
+store to \co{y} on line~14.
+
+The implementation is then within its rights to act as if lines~13 and~14
+were interchanged.
+At which point, the order of execution could be as follows:
+
+\begin{enumerate}
+\item	Line~14 stores the constant 1 to \co{y}.
+\item	Line~8 loads the constant 1 from \co{y}, setting \co{r1} to 1.
+\item	Line~9 stores the constant 1 to \co{x}.
+\item	Line~8 loads the constant 1 from \co{x}, setting \co{r2} to 1.
+\end{enumerate}
+
+Please note that the OOTA-like cycle called out in the \co{exists}
+clause can happen even though the program never actually executes any
+code that would have resulted in division-by-zero UB.
+It is tempting to consider using techniques
+% Appendix~\ref{app:Aside on Undefined Behavior}
+that prevent back-propagation of UB, but these are ineffective in this
+case because the C++ implementation can detect and exploit UB when
+considering line~14 in isolation.
+Once the constant 1 is substituted for line~14's expression, the
+hardware is free to reorder lines~13 and~14 even if a compiler-based
+C++ implementation refrains from doing so.
+Note that this OOTA-like cycle can appear even if all of the atomic
+operations in
+Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}
+were converted to be volatile.
+
+Therefore, this paper focuses on UB-free programs, rare though these
+beasts might be among large-scale C++ programs running in production.
+
+\subsection{Semantic Dependencies and \co{volatile}}
+\label{app:Semantic Dependencies and volatile}
+
+The Linux kernel uses volatile accesses to constrain the compiler,
+and it is worth looking at the example shown in
+Listing~\ref{lst:Non-Volatile Accesses and Dependencies}
+to see how this works.
+
+\begin{listing}
+\begin{verbatim}
+ 1 atomic<int> x, y;
+ 2
+ 3 void thread1() {
+ 4   int r1 = x.load(memory_order_relaxed);
+ 5   y.store(r1, memory_order_relaxed);
+ 6 }
+\end{verbatim}
+\caption{Non-Volatile Accesses and Dependencies}
+\label{lst:Non-Volatile Accesses and Dependencies}
+\end{listing}
+
+Because both \co{x} and \co{y} are nonvolatile, the compiler is
+free to assume that the value loaded by line~4 will be some value
+stored to \co{x} or its initial value.
+Because there are no stores to \co{x}, the only remaining possibility
+is the default-initialized value of zero.
+Therefore, the C++ implementation is within its rights to substitute
+the constant zero for \co{r1} on line~5, eliminating the semantic
+dependency that would otherwise extend from line~4 to line~5.
+This in turn enables the implementation to reorder line~5 to
+precede line~4.
+
+\begin{listing}
+\begin{verbatim}
+ 1 atomic<volatile int> x, y;
+ 2
+ 3 void thread1() {
+ 4   int r1 = x.load(memory_order_relaxed);
+ 5   y.store(r1, memory_order_relaxed);
+ 6 }
+\end{verbatim}
+\caption{Volatile Accesses and Dependencies}
+\label{lst:Volatile Accesses and Dependencies}
+\end{listing}
+
+One way to preserve this dependency is use of \co{volatile}, as shown in
+Listing~\ref{lst:Volatile Accesses and Dependencies}.
+Now the implementation is forbidden from assuming that the value loaded
+by line~4 has any relation to any stores to or initialization of \co{x},
+even if \co{x} is placed in normal memory.
+
+In this case, the \co{volatile} keyword alerts the implementation to
+the possibility of interfering accesses to \co{x} from outside the
+program, for example, due to \co{x} being:
+
+\begin{itemize}
+\item	Allocated in an MMIO region of the address space.
+\item	Modified by an unknown-to-the-implementation thread
+	or signal handler.
+\item	Subject to I/O-device DMA operations.
+\item	Modified by debugger commands.
+\item	Modified by accesses to facilities such as \path{/dev/mem}.
+\item	Modified by as-yet-unwritten dynamically linked libraries.
+\item	Modified by some other mechanism of the reader's choosing.
+\end{itemize}
+
+These possibilities force the implementation to preserve the semantic
+dependency from line~4 to line~5.
+They can also prevent the formation of OOTA cycles local to the
+implementation because such inteference means that the volatile loads
+might be reading from something that the implementation knows nothing
+about.
+
+The developer is free to arrange for \co{x} to be free of any
+interference, thus obtaining predictable behavior along with the
+preserved semantic dependency.
+However, this example invalidates the rough definition of semantic
+dependency from
+Section~\ref{sec:Simple OOTA Cycle}
+because the value loaded from \co{x} will always be the initial value
+of zero, so that it is at best dubious to talk about any changes in
+the value loaded.
+
+One way forward is to assume that interference could happen when
+determining which dependencies are semantic, and then do further analysis
+using that determination.
+
 % \section{History}
 % \label{sec:History}
+
+\clearpage
+
+\section{Litmus Tests from “Causality Test Cases"}
+\label{app:Litmus Tests from “Causality Test Cases"}
+
+These tests might be two decades old, but they are still quite
+relevant.\footnote{
+	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
+We have translated them from Java to C++, and we evaluate them using
+the \co{herd7} tool and also using manual analysis.
+Please note that differences in the semantics of these two languages
+result in changes in verdict for some litmus tests.
+
+\subsection{Causality Test Case 1}
+\label{app:Causality Test Case 1}
+
+Listing~\ref{lst:Causality Test Case 1}
+shows causality test case 1, for which the \co{r1 == r2 == 1} result
+is to be allowed.
+And indeed, this result does show up in the output from the \co{herd7} tool.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-1.litmus @@
+\caption{Causality Test Case 1}
+\label{lst:Causality Test Case 1}
+\end{listing}
+
+Compilers that can prove the value of \co{x} is always
+nonnegative can also determine that there is no semantic dependency between
+\co{P0()}'s load from \co{x} on line~8 and its store to \co{y} on line~10,
+that is, no matter which of the possible values is loaded from \co{x},
+the value 1 will always be stored to \co{y}.
+Relative to less-omniscient compilers, however,
+including those using single-thread analysis or treating atomic
+objects as volatile, there \emph{is} a semantic dependency.
+
+Even omniscient compilers will observe a semantic dependency from
+\co{P1()}'s load from \co{y} on line~14 to its store to \co{x} on line~15.
+
+Either way no OOTA cycle will occur, with omniscient compilers
+having no cycle at all because of the lack of an sdep link in \co{P0()},
+and other implementations prevented by hardware
+dependency ordering from realizing the cycle.
+
+\subsection{Causality Test Case 2}
+\label{app:Causality Test Case 2}
+
+Listing~\ref{lst:Causality Test Case 2}
+shows causality test case 2, for which the \co{r1 == r2 == r3 == 1} result
+is to be allowed.
+This result appears in the \co{herd7} output.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-2.litmus @@
+\caption{Causality Test Case 2}
+\label{lst:Causality Test Case 2}
+\end{listing}
+
+Because \co{P0()}'s loads from \co{x} on lines~8 and~9 are unordered,
+a C++ compiler that treats atomics as quasi volatile
+can act as if the source code loaded from \co{x} only
+once and assigned the returned value to both \co{r1} and \co{r2},
+merging the loads.\footnote{
+	But not in C, where relaxed atomic operations are volatile,
+	and thus are observable behavior not subject to merging.}
+The compiler could then prove that the condition on line~10
+would always true, and act as if the source code had
+unconditionally executed the relaxed store to \co{y} on line~11
+independent of the value loaded from \co{x}.
+Without this dependency there would then be no OOTA cycle.
+
+For compilers that do not merge the two loads, the hardware
+dependency ordering arising from the conditional test on line~10
+will prevent the realization of an OOTA cycle.
+
+\subsection{Causality Test Case 3}
+\label{app:Causality Test Case 3}
+
+Listing~\ref{lst:Causality Test Case 3}
+shows causality test case 3, for which the \co{r1 == r2 == r3 == 1} result
+is to be allowed.
+The \co{herd7} tool reports this result.
+
+The analysis from the previous section applies here.
+Although the addition of \co{P2()} allows sequentially consistent executions
+in which \co{P0()}'s two loads from \co{x} return different values,
+it does not change the fact that those two loads can be merged.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-3.litmus @@
+\caption{Causality Test Case 3}
+\label{lst:Causality Test Case 3}
+\end{listing}
+
+\subsection{Causality Test Case 4}
+\label{app:Causality Test Case 4}
+
+Listing~\ref{lst:Causality Test Case 4}
+shows causality test case 4, for which the \co{r1 == r2 == 1} result
+is to be forbidden.
+The \co{herd7} tool finds this OOTA result;
+the \co{S8} values given on line~3 of its output indicate
+that in the execution it describes, the variables \co{r1} and~\co{r2} were never
+assigned any specific value, so they could just as well be equal to
+one as to anything else.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-4.litmus @@
+\caption{Causality Test Case 4}
+\label{lst:Causality Test Case 4}
+\end{listing}
+
+This test case is virtually the same as the Simple OOTA cycle
+discussed in section~\ref{sec:Simple OOTA Cycle}, the only difference
+being the final values of the variables.
+
+\subsection{Causality Test Case 5}
+\label{app:Causality Test Case 5}
+
+Listing~\ref{lst:Causality Test Case 5}
+shows causality test case 5, for which the \co{r1 == r2 == 1} and \co{r3 == 0}
+result is to be forbidden.
+The \co{herd7} tool reports this OOTA result in line~3 of its output.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-5.litmus @@
+\caption{Causality Test Case 5}
+\label{lst:Causality Test Case 5}
+\end{listing}
+
+This is the same as causality test case~4 with \co{P2()} and \co{P3()}
+added, but those routines play no role in the OOTA cycle.
+\co{P2()} has no loads and so cannot participate in an OOTA cycle in
+any case.
+
+\subsection{Causality Test Case 6}
+\label{app:Causality Test Case 6}
+
+Listing~\ref{lst:Causality Test Case 6}
+shows causality test case 6, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The \co{herd7} tool reports this result.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-6.litmus @@
+\caption{Causality Test Case 6}
+\label{lst:Causality Test Case 6}
+\end{listing}
+
+A compiler that can prove \co{y} is always either
+zero or one can also prove that there is no semantic dependency
+from \co{P1()}'s load on line~14 to its stores on lines~16 and~19.
+Such a compiler could act as if lines~15--20 had been replaced
+by a single store to \co{x}, showing that the dependency isn't
+semantic.
+
+Less-omniscient compilers will preserve the dependency, and then
+hardware dependency ordering will prevent the OOTA cycle from
+being realized.
+
+\subsection{Causality Test Case 7}
+\label{app:Causality Test Case 7}
+
+Listing~\ref{lst:Causality Test Case 7}
+shows causality test case 7, for which the \co{r1 == r2 == r3 == 1}
+result is to be allowed.
+The \co{herd7} tool finds this result.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-7.litmus @@
+\caption{Causality Test Case 7}
+\label{lst:Causality Test Case 7}
+\end{listing}
+
+This test case has everything
+to do with reordering and nothing to do with OOTA cycles.
+The \co{r1 == r2 == r3 == 1} outcome happens when lines~17, 10, 11,
+15, 16 and~9 execute in that order.
+Note that this execution order respects the semantic dependencies in
+lines~10--11 and~15--16, and thus is not ruled out by hardware
+dependency ordering.
+It could have been ruled out if some of the accesses were acquire or
+release, but here they are all relaxed.
+
+\subsection{Causality Test Case 8}
+\label{app:Causality Test Case 8}
+
+Listing~\ref{lst:Causality Test Case 8}
+shows causality test case 8, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The \co{herd7} tool does not report this result, but it does if
+\co{r2} is initialized to one before being set to
+\co{1 + r1 * r1 - r1}.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-8.litmus @@
+\caption{Causality Test Case 8}
+\label{lst:Causality Test Case 8}
+\end{listing}
+
+Omniscient compilers might note that the computation on line~9
+will have the value one when the load from \co{x} returns either
+zero or one, which are the only values \co{x} and \co{y} can take.
+Such a compiler could then realize that line~9 always sets \co{r2} to one,
+independent of the value loaded in line~8.
+Then there would be no semantic dependency between \co{P0()}'s load
+and store, and thus no OOTA cycle relative to this compiler.
+
+A less-capable compiler would leave the dependency in place,
+and the resulting hardware dependency ordering would prevent
+the OOTA cycle from being realized.
+
+\subsection{Causality Test Case 9}
+\label{app:Causality Test Case 9}
+
+Listing~\ref{lst:Causality Test Case 9}
+shows causality test case 9, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The \co{herd7} tool does not report this result.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-9.litmus @@
+\caption{Causality Test Case 9}
+\label{lst:Causality Test Case 9}
+\end{listing}
+
+This is the same as causality test case~8 but with \co{P2()} added;
+the store to \co{x} in \co{P2} can interfere with the formation of the
+OOTA cycle.
+
+Unlike in causality test case~8, the computation on line~9 must be
+carried out because now the load on line~8 might return two.
+Consequently \co{P0} and~\co{P1} both have semantic dependencies,
+and hardware dependency ordering will prevent the OOTA cycle from
+being realized.
+
+\subsection{Causality Test Case 9a}
+\label{app:Causality Test Case 9a}
+
+Listing~\ref{lst:Causality Test Case 9a}
+shows causality test case 9a, for which the \co{r1 == r2 == 1}
+result is to be allowed.
+The \co{herd7} tool does not report this result.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-9a.litmus @@
+\caption{Causality Test Case 9a}
+\label{lst:Causality Test Case 9a}
+\end{listing}
+
+This is the same as causality test case~9 except that here \co{P2()}
+sets \co{x} to zero instead of two.
+As as result the computation on line~9 is once again independent of
+the value loaded on line~8, which makes the analysis nearly the same
+as that of causality test case~8.
+
+\subsection{Causality Test Case 10}
+\label{app:Causality Test Case 10}
+
+Listing~\ref{lst:Causality Test Case 10}
+shows causality test case 10, for which the \co{r1 == r2 == 1} and \co{r3 == 0}
+result is to be forbidden.
+The \co{herd7} tool reports this result.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-10.litmus @@
+\caption{Causality Test Case 10}
+\label{lst:Causality Test Case 10}
+\end{listing}
+
+\co{P0()} and \co{P1()} have straightforward
+semantic dependencies between their respective loads and stores, and
+so hardware dependency ordering will prevent an OOTA cycle from being
+realized.
+\co{P2()} has no semantic dependency, and \co{P3()}'s store does not
+occur in this execution (i.e., the one described by the \co{exists} clause)
+because \co{r3} is zero.
+Thus neither of them contributes to an OOTA cycle.
+
+\subsection{Causality Test Case 11}
+\label{app:Causality Test Case 11}
+
+Listing~\ref{lst:Causality Test Case 11}
+shows causality test case 11, for which the \co{r1 == r2 == r3 == r4 == 1}
+result is to be allowed.
+The \co{herd7} tool reports this result.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-11.litmus @@
+\caption{Causality Test Case 11}
+\label{lst:Causality Test Case 11}
+\end{listing}
+
+Like causality test case~7, this test case has nothing to do with OOTA cycles.
+The result may be obtained by executing lines~20, 12, 13, 18, 19, 10, 11,
+and~17 in that order.
+
+\subsection{Causality Test Case 12}
+\label{app:Causality Test Case 12}
+
+Causality Test Case 12 in
+Listing~\ref{lst:Causality Test Case 12}
+uses arrays, which are not yet supported by the \co{herd7} tool.
+
+\begin{listing}[tbp]
+@@ DisplayLitmus litmus/oota-causality-12.litmus @@
+\caption{Causality Test Case 12}
+\label{lst:Causality Test Case 12}
+\end{listing}
+
+An omniscient compiler could note that only \co{P0()} accesses
+array \co{a[]} and act as if the code was simpler, but there would
+still be semantic dependencies between \co{P0()}'s and \co{P1()}s
+relaxed atomic loads and stores involving \co{x} and \co{y}.
+The resulting hardware dependency ordering will prevent an OOTA cycle
+from being realized, whether the compiler is omniscient or not.
+
+\subsection{Causality Test Case 13}
+\label{app:Causality Test Case 13}
+
+Listing~\ref{lst:Causality Test Case 13}
+shows causality test case 13, for which the \co{r1 == r2 == 1}
+result is to be forbidden.
+The \co{herd7} tool reports this result.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-13.litmus @@
+\caption{Causality Test Case 13}
+\label{lst:Causality Test Case 13}
+\end{listing}
+
+This is the same as causality test case~10 without the confounding
+influence of \co{P2()} and~\co{P3()}.
+
+\subsection{Causality Test Case 14}
+\label{app:Causality Test Case 14}
+
+Listing~\ref{lst:Causality Test Case 14}
+shows causality test case 14, for which the \co{r1 == r3 == 1} and \co{r2 == 0}
+result is to be forbidden.
+The \co{herd7} tool reports this result.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-14.litmus @@
+\caption{Causality Test Case 14}
+\label{lst:Causality Test Case 14}
+\end{listing}
+
+In the execution described by the \co{exists} clause,
+\co{P0()} has a semantic dependency from line~9 to
+line~13 and \co{P1()} has a semantic dependency from line~18 to
+line~21.
+(There's also a semantic dependency from line~17 to line~21 but it
+doesn't enter into this OOTA cycle, which involves only \co{x} and~\co{y}.)
+The \co{memory_order_seq_cst} specification on line~17 has no important effect;
+hardware dependency ordering will prevent the cycle from being realized.
+
+\subsection{Causality Test Case 15}
+\label{app:Causality Test Case 15}
+
+Listing~\ref{lst:Causality Test Case 15}
+shows causality test case 15, for which the \co{r0 == r1 == r3 == 1}
+and \co{r2 == 0}
+result is to be forbidden.
+The \co{herd7} tool reports this result.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-15.litmus @@
+\caption{Causality Test Case 15}
+\label{lst:Causality Test Case 15}
+\end{listing}
+
+The execution described by the \co{exists} clause
+has an OOTA cycle involving \co{u} and~\co{v}.
+The dependency in \co{P0()} from line~13 to line~19 and the dependency
+in \co{P1()} from line~24 to line~27 are both semantic.
+The resulting hardware dependency ordering prevents the cycle from
+being realized.
+
+Having no load, \co{P2()} cannot participate in an OOTA cycle.
+
+\subsection{Causality Test Case 16}
+\label{app:Causality Test Case 16}
+
+Listing~\ref{lst:Causality Test Case 16}
+shows causality test case 16, for which the \co{r1 == 2} and \co{r2 == 1}
+result is to be allowed.
+The \co{herd7} tool does not report this result, which is to be
+expected because it would violate the C++ memory model's read-write
+coherence rule regarding the modification order of an atomic object
+(6.9.2.2p17 \co{[intro.races]}).
+Because no abstract execution can obtain \co{r1 == 2} and \co{r2 == 1}, no
+valid realization will either.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-16.litmus @@
+\caption{Causality Test Case 16}
+\label{lst:Causality Test Case 16}
+\end{listing}
+
+\subsection{Causality Test Case 17}
+\label{app:Causality Test Case 17}
+
+Listing~\ref{lst:Causality Test Case 17}
+shows causality test case 17, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result, although it does
+report a different OOTA cycle involving lines 11, 12, 16, and~17
+(the \co{S17} value on line~3 of the \co{herd7} output).
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-17.litmus @@
+\caption{Causality Test Case 17}
+\label{lst:Causality Test Case 17}
+\end{listing}
+
+A compiler that treats atomics as quasi volatile could reason as follows:
+Suppose that \co{P0()}'s load from \co{x} on line~8 returns the value 42.
+In that case, the second load on line~11 can be merged with the first,
+so that it also returns the value 42.
+On the other hand, if the load on line~8 returns some other value then
+line~10 will store the value 42 to \co{x}, and then the load on line~11
+can be omitted in favor of using the value 42 stored by line~10.
+
+Either way, the value of \co{r1} will be 42.
+Therefore the loads from \co{x} cannot affect the store on line~17,
+so there is no semantic dependency.
+This allows the cycle to be realized,
+but because the dependency in \co{P0()} is not semantic,
+the cycle is not OOTA.
+
+\subsection{Causality Test Case 18}
+\label{app:Causality Test Case 18}
+
+Listing~\ref{lst:Causality Test Case 18}
+shows causality test case 18, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-18.litmus @@
+\caption{Causality Test Case 18}
+\label{lst:Causality Test Case 18}
+\end{listing}
+
+This is the same as causality test case~17, even to the alternate OOTA
+cycle reported by \co{herd7}, except that line~9 tests
+\co{r3} for equality to zero rather than non-equality to 42.
+Perhaps surprisingly, this makes a difference.
+
+A compiler that treats atomic objects as volatile or uses
+single-thread analysis cannot assume that \co{x} will necessarily be
+equal to 42 following lines~9 and~10; it might have some other value
+such as 1.
+Therefore the compiler cannot conclude that \co{r1} will always be 42
+and must leave the semantic dependency between lines~11 and~12 intact.
+Of course, the resulting OOTA cycle will not be realized, because of
+hardware dependency ordering.
+
+\subsection{Causality Test Case 19}
+\label{app:Causality Test Case 19}
+
+Listing~\ref{lst:Causality Test Case 19}
+shows causality test case 19, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result but it does report the
+OOTA cycle in line~3 of its output.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-19.litmus @@
+\caption{Causality Test Case 19}
+\label{lst:Causality Test Case 19}
+\end{listing}
+
+This is nearly the same as the Simple OOTA cycle and causality test
+case~4, with the addition of \co{P2()}.
+Just as in those examples, the OOTA cycle cannot be realized by
+compilers that treat atomic objects as volatile or that do
+single-thread analysis and treat atomic objects as quasi volatile.
+
+\subsection{Causality Test Case 20}
+\label{app:Causality Test Case 20}
+
+Listing~\ref{lst:Causality Test Case 20}
+shows causality test case 20, for which the \co{r1 == r2 == r3 == 42}
+result is to be allowed.
+The \co{herd7} tool does not report this result.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-20.litmus @@
+\caption{Causality Test Case 20}
+\label{lst:Causality Test Case 20}
+\end{listing}
+
+This is the same as causality test case~19, even to the OOTA cycle
+reported by \co{herd7}, except that line~19 tests
+\co{r3} for equality to zero rather than non-equality to 42.
+The same analysis applies.
 

### 2024-02-26 ootavacuous.tex
Changes for chapter 8
Here is a bunch of edits for chapter 8.  Nothing really major but they
are quite extensive.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 More examples of user control over language semantics are given in
-Appendix~\ref{app:User Influence Over Language Semantics}.
+Appendix~\ref{app:User Influence Over Language Semantics}
+and the discussion
+in~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.
 
@@
 in prior work.
-However, in correct real-world C++ implementations, the standard's
-abstract machine must be mapped onto real matter and energy.
-This mapping has not been considered by prior work, and brings to
-bear additional constraints from hardware architecture and
-design,\footnote{
-	Which is in turn constrained by the laws of physics.}
-which has been only partially accounted for in the standard and in
-prior work.
-One reason for the incomplete accounting for hardware constraints is
+In real-world C++ implementations, the standard's
+abstract machine must be mapped onto real hardware, however, and
+this mapping has not been considered to any significant degree.
+It imposes additional constraints stemming from hardware architecture and
+design,
+which have been only partially accounted for in the standard and other work.
+Part of the reason for this is
 that accurate and executable formal descriptions of hardware memory
 models did not appear until after the standard was released.
+But they do exist now.
 
@@
 Nevertheless, the main points of this paper will remain unaffected.
-To see this, keep in mind that on real systems, observable effects
-must eventually be committed, but that no effect can be committed
+To see why, keep in mind that on real systems observable effects
+must eventually be committed, but no effect can be committed
 while any of the computations that determine the effect remain
 speculative.
-Furthermore, instructions take time to execute, even when executed
-speculatively.
-For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
-suppose that the storage for \co{x} is located immediately before that
-for \co{y[]}.
-Suppose further that hardware speculation incorrectly guesses the value
-of \co{i} to be -1, so that the speculated store of 42 on line~9 uses
+Combining this with the obvious fact that a machine instruction cannot
+commit until it is no longer speculative, we can draw two conclusions:
+\begin{itemize}
+\item	A load cannot commit until the store it reads from has
+	committed.\footnote{
+	An early ARMv8 memory model did allow loads that read from an
+	earlier store in the same thread to commit before the store.
+	This apparent exception merely reflects a difference in
+	nomenclature; in the memory model a store was said to commit at
+	the time when it makes its new value available to other
+	threads, whereas we say that a store commits at the time when
+	the CPU has irrevocably decided that it will take place with a
+	particular fully determined value and address.}
+\item	A store cannot commit until all the loads on which it has a
+	hardware dependency have committed.
+\end{itemize}
+
+As an example showing what can go wrong when these principles are
+violated, consider Listing~\ref{lst:Speculated Store and Non-Speculated Load}.
+Suppose that the storage for \co{x} is located immediately before that
+for \co{y[]}, and
+suppose further that hardware speculation incorrectly guesses the value
+of \co{r1} will be $-1$, so that the speculated store of 42 on line~11 uses
 the address of \co{x}.\footnote{
@@
 	nor cares, nor should it.}
-Then line~11 can cause the value 42 to be loaded into \co{r2},
-thus satisfying the \co{exists} clause on line~20, when
-by the rules of the C++ abstract machine the value must instead be
-either zero or one.
+Then if line~13 reads from the speculative store and commits, it will load
+the value 42 into \co{r2}---an observable effect since \co{r2}
+is mentioned in the \co{exists} clause on line~16---although
+by the rules of the C++ abstract machine the value must instead be zero.
 
@@
 
-Therefore, a given load is prohibited from committing until after the
-store that it reads from has committed.
-And yes, this does mean that hardware can produce OOTA cycles during
-speculative execution, but the hardware is required to prevent all such
-speculative OOTA cycles from ever committing.
-As a special case of this prohibition, if a speculated load obtains
-a value from a store that has not yet committed, then if that store
-is squashed, the load must also be squashed, even if that load and
-that store are executed by different CPUs.
-
-Applying this to Listing~\ref{lst:OOTA Cycle}, lines~10 and~12 cannot be
-committed until after the load in line~8 and the conditional on line~9
-have both committed.
-Similarly, lines~18 and~20 cannot be committed until after lines~16
-and~17 have committed.
-
-One could imagine a conditional-store instruction, similar to existing
-conditional-move instructions, but with condition controlling a store
-to memory rather than a move to a register.
-The condition would control whether or not the store was executed,
-and prior conditional-move instructions could control the value
-stored and the address stored to.
-With or without prior conditional-move instructions, but assuming a
-compiler doing single-thread analysis, a conditional-store instruction
-permits generation of code corresponding to lines~10 and~11 of
+Applied to Listing~\ref{lst:OOTA Cycle}, for another example,
+the principles dictate that if the load on line~8 reads from the store
+on line~18 and the load on line~16 reads from the store on line~10
+(as they do in the execution described by the listing's \co{exists} clause),
+then each of the stores must commit before its corresponding load.
+A further application says that
+neither line~10 nor line~12 can commit until after the load
+in line~8 (and also the conditional on line~9) has committed.
+Similarly, neither line~18 nor line~20 can commit until after line~16
+(and~17) has committed.
+Taken together, these constraints imply that the OOTA cycle in the
+listing cannot be realized, because to do so would require that
+none of the machine instructions corresponding to lines~8, 10, 16,
+and~18 could commit before the others.
+Furthermore, it's not possible to circumvent this reasoning by suggesting
+that some of those lines could commit at the same time,
+because instructions take time to execute even when executing
+speculatively.
+
+Yes, this does mean that hardware can produce OOTA cycles during
+speculative execution, but the hardware is required to prevent such
+speculative cycles from committing.
+As a special case of this requirement, if a load obtains
+a value from a speculated store that has not yet committed,
+and the store gets squelched,
+then the load must be restarted before it commits even if the load and
+the store are executed by different CPUs.
+
+The requirement applies within a single
+multi-threaded core as well as between cores.
+The fact that intra-core communication is faster than inter-core
+communication does not magically give loads the ability to commit
+before the stores they read from, no matter what cores they execute on.
+
+\medskip
+
+Another feature we can expect of upcoming computer systems is the
+addition of new hardware instructions.
+For example, one could imagine a conditional-store instruction,
+similar to existing conditional-move instructions but carrying out
+a store to memory rather than a move to a register.
+The condition would control whether or not the store takes place.
+A compiler doing single-thread analysis could use a conditional-store
+instruction to generate code corresponding to lines~10 and~11 of
 Listing~\ref{lst:Causality Test Case 2}
 on page~\pageref{lst:Causality Test Case 2}
-without use of a conditional branch.
-This means that the control dependency would extend from the loads on
-lines~8 and~9 to the store on line~11 and nowhere else, potentially
-enabling more hardware optimizations than would be permitted for code
-relying on conditional branches.
-Nevertheless, the conditional store could not be committed until all of
-the instructions it depends on have committed.
-
-Finally, the prohibition of hardware OOTA applies within a single
-multi-threaded core as well as between cores.
-This means that for any set of speculative accesses to be committed,
-all loads in that set must have committed, which in turn requires that
-all store providing the loads' values must also have committed.
-
-The key point is that a store having a semantic dependency on a given
-set of loads must execute after the completion of at least some of those
-loads in global time.
+without use of a conditional branch, potentially
+enabling more hardware optimizations than would otherwise be permitted.
+Nevertheless, the conditional store could not be committed until
+the instructions it depends on (lines~8 and~9) have committed.
+Any new instruction will be subject to this constraint,
+just as the instructions in current systems are.
+
+A key conclusion to draw from this discussion is that
+regardless of how advanced a computer system may be,
+it cannot commit a store having a hardware dependency on some
+set of loads until after (in global time) those loads have committed.
 
@@
 
-Because volatile atomics constitute observable
-behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2023N4950},
-they must be executed in strict accordance with the rules of the abstract
+Volatile atomic accesses constitute observable
+behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2023N4950} and
+must be executed in strict accordance with the rules of the C++ abstract
 machine.
-This means that any volatile atomic operation involving a store may
-execute only after after all loads whose values are used to compute the
-store's address or value have committed.
-OOTA afficienados will recognize this as a special case of ordering
+This means that a volatile atomic store in the source code
+corresponds directly to a machine-code store instruction, which can
+commit only after all loads whose values are used to compute the
+store's address and value have committed.
+OOTA aficionados will recognize this as a special case of ordering
 relaxed loads before relaxed stores, albeit one not requiring
@@
 Although C++ nonvolatile accesses to atomic objects are not observable
-behavior, any implementation that restricts its code analysis to a
+behavior, any compiler that restricts its code analysis to a
 single thread must assume (unless it can prove otherwise) that a given
-relaxed atomic store might affect observable behavior following any
-load that returns the value stored.
-This assumption does not constrain the implementation to anywhere near
+relaxed atomic store can affect observable behavior in a different
+thread that loads the value stored.
+This assumption does not constrain the compiler to anywhere near
 the extent that a volatile relaxed atomic store would, but it does add
-significant constrants over those related to nonvolatile non-atomic
+significant constraints over those related to nonvolatile non-atomic
 stores.
@@
 For example, nonvolatile relaxed atomic accesses are subject to the
-memory model.
+C++ memory model.
 The ``order-preserving'' bullet from
 Section~\ref{sec:Volatile and Quasi Volatile Accesses}
-prohibits reordering only of quasi volatile accesses to the same object,
-but such reordering is in any case directly forbidden by the standard.
-Referring also to the ``onto'' bullet from
-Section~\ref{sec:Volatile and Quasi Volatile Accesses},
-implementations should not behave as if the source code has
-additional:
-
-\begin{enumerate}
-\item	Invented atomic stores.
-\item	Duplicated atomic stores.
-\item	Invented atomic loads.
-\end{enumerate}
-
-The discussion of these two constraints is followed by a discussion of
-when redundant atomic stores may safely be omitted and on fusing of
-nonvolatile atomic loads from adjacent objects.
-These topics correspond to the ``one-to-one'' and ``total'' bullets of
+prohibits reordering of quasi volatile accesses when they are
+to the same object,
+but such reordering is in any case directly forbidden by the memory model
+(the four coherence rules in 6.9.2.2p14--19).
+The ``onto'' bullet also largely reiterates restrictions that follow
+from the standard, in particular,
+that compilers should not invent atomic stores,
+duplicate atomic stores, or invent atomic loads.
+On the other hand, it is permissible for compilers to omit redundant
+atomic stores or fuse nonvolatile atomic accesses of adjacent objects
+under appropriate circumstances;
+these topics correspond to the ``total'' and ``one-to-one'' bullets of
 Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
 
-It is also important to note that implementations and users have a
-surprising degree of influence over language
-semantics~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.
-Please see
-Appendix~\ref{app:User Influence Over Language Semantics}
-for examples.
-
 \paragraph{Invented Atomic Stores}
-
 \begin{listing}[tbp]
@@
 \end{listing}
-
-The reason that atomic stores cannot be invented is that doing so can
+The reason that atomic stores should not be invented is that doing so can
 introduce new (and almost certainly undesirable) behaviors that are
@@
 With line~8, the additional value~42 can also be loaded into \co{r1}.
-The implementation is therefore forbidden from inventing that store of 42
+The compiler is therefore forbidden from inventing the store of 42
 unless it can prove that doing so does not negatively affect the
-program's observable behaviors, independently of OOTA cycles.
-For example, the implementation might be able to prove that there
-are no other accesses to \co{x} at the time of the invented store.
+program's observable behaviors.
+For example, the compiler might be able to prove that there
+are no other accesses to \co{x} at the time of the invented
+store---but it's very hard to imagine how a compiler that analyzes
+only a single thread at a time could prove such a thing.
 
 \paragraph{Duplicated Atomic Stores}
-
 \begin{listing}[tbp]
@@
 \end{listing}
-
 Duplicating atomic stores is a special case of inventing them, but it is
@@
 litmus test that demonstrates such a behavior.
-Without line~9, if the value loaded into \co{r1} is 1, then the final value
-of \co{x} must be 2.
-With line~9, the final value of \co{x} can be 1 even when the
-value loaded into \co{r1} is 1.
-The implementation is therefore forbidden from duplicating that store of 1
+Without line~9, if the value loaded into \co{r1} is one then the final value
+of \co{x} must be two.
+With line~9, the final value of \co{x} can be one even when the
+value loaded into \co{r1} is one.
+The compiler is therefore forbidden from duplicating the store of one
 unless it can prove that doing so does not negatively affect the
-program's observable behaviors, independently of OOTA cycles.
-Again, the implementation might be able to prove that there
-are no other accesses to \co{x} at the time of the invented store.
+program's observable behaviors.
+Again, the compiler might be able to prove that there
+are no other accesses to \co{x} at the time of the duplicated store.
 
 \paragraph{Invented Atomic Loads}
-Although one can argue that it is functionally correct to invent atomic
-loads as long as the loaded value is not used, doing so can result in
-excessive cache-miss overheads.
-Even duplicating an atomic load can lead to incorrect results if done
-carelessly.
-To see this, please refer to
-Listing~\ref{lst:Do Not Duplicate Atomic Loads}.
-Suppose that the load on line~13 were duplicated, for example, by
-uncommenting line~15.
-Then the values of \co{a} and \co{b} passed to line~17's call to
-\co{do_something_with()} might be from different instances of the
-\co{foo} structure, something that \co{do_something_with()} might
-not be prepared to deal with.
-
 \begin{listing}[tbp]
@@
 \end{verbatim}
-\caption{Do Not Duplicate Atomic Loads}
-\label{lst:Do Not Duplicate Atomic Loads}
-\end{listing}
-
-For these reasons, C++ implementations should avoid inventing atomic
+\caption{Example Duplicated Load}
+\label{lst:Example Duplicated Load}
+\end{listing}
+Although one can argue that it is functionally correct to invent atomic
+loads as long as the loaded value is not used, doing so can result in
+extra cache misses for no good purpose.
+And even duplicating an atomic load can lead to incorrect results if done
+carelessly.
+To see this, consider
+Listing~\ref{lst:Example Duplicated Load}.
+Suppose that the load on line~13 were duplicated, for example, by
+uncommenting line~15.
+Then the values of \co{a} and \co{b} passed to line~17's call to
+\co{do_something_with()} might be from different instances of the
+\co{foo} structure, something that \co{do_something_with()} probably
+isn't prepared to deal with.
+
+For these reasons, C++ compilers should avoid inventing atomic
 loads, and in fact should avoid even duplicating them.
@@
 \paragraph{Omitted Redundant Atomic Stores}
-In contrast, a pair of back-to-back nonvolatile atomic stores to \co{x}
-might be executed such that no other thread accesses \co{x}
+In contrast, a pair of back-to-back nonvolatile atomic stores to the
+same object
+always might be executed such that no other thread accesses the object
 during the time between those two stores.
-This means that, if the implementation acted as if the first store was omitted
-from the source, the user would have no way to prove that fact short of
+This means that if the compiler omitted the first store,
+the user would have no way to prove this fact short of
 inspecting the assembly code.
@@
 	clobber.}
-to prevent the implementation from omitting a store.
-However, acting as if such a redundant store was omitted from the source
-code cannot create an OOTA cycle because any valid execution (OOTA or not)
+to prevent them.
+
+Acting as if a redundant store was omitted from the source
+code cannot create an OOTA cycle, because any valid execution (OOTA or not)
 of the program with the first of a back-to-back pair of nonvolatile
@@
 
-\paragraph{Loads From Adjacent Non-Volatile Atomic Objects}
-
-Suppose there are a pair of adjacent nonvolatile atomic objects,
-and that these objects taken together are machine-word aligned
-and machine-word sized.
+\paragraph{Fused Accesses of Adjacent Nonvolatile Atomic Objects}
+Suppose a pair of adjacent nonvolatile atomic objects, when combined,
+form a single machine-word aligned and machine-word sized unit.
 Suppose further that the source code contains a pair of atomic
@@
 being a relaxed load.
-Then the C++ implementation could choose to act as if the source code
-had instead contained a single atomic load from the combined
-pair of objects, with the memory-ordering semantics of the first load.
-
-This holds even when the atomic objects are quasi volatile,
+Then the compiler could generate instead
+a single atomic load of the combined pair,
+with the memory-ordering semantics of the first load.
+A similar fact applies to pairs of stores.
+
+This holds even when the atomic objects are treated as quasi volatile,
 providing an example of how the ``one-to-one'' bullet of
@@
 tools are in some cases permitted to misclassify them, as shown in
-Table~\ref{tab:Semantic-Dependency Classification For Tools}.
+Table~\ref{tab:Dependency Classification For Tools}.
 
@@
 \hline
-sdep			& sdep			& Temporal \\
+sdep			& sdep			& Ordered \\
 \cline{2-3}
@@
 \cline{2-3}
-			& $\neg$sdep		& Atemporal \\
+			& $\neg$sdep		& Unordered \\
 \end{tabular}
-\caption{Semantic-Dependency Classification For Tools}
-\label{tab:Semantic-Dependency Classification For Tools}
+\caption{Dependency Classification For Tools}
+\label{tab:Dependency Classification For Tools}
 \end{table}
@@
 that the code is too complex for it to analyze.
-Some projects would choose to restructure the code in such cases on the
+Some projects would choose to restructure the code in such cases, on the
 grounds that code that is difficult for an automated tool to understand
-is likely also being misunderstood by its all-too-human developers.
+is also likely to be misunderstood by its all-too-human developers.
 
@@
 However, there is an important special case for tools that are closely
-associated with a C++ implementation.
-Such tools can simply use that implementation's classification of
+associated with a C++ compiler.
+Such tools can simply use that compiler's classification of
 dependencies as semantic on the one hand or nonsemantic on the
-other, whether by working with that implementation's intermediate
-representations, binaries produced by that implementation, by tracing
-the program's execution, or, in the case of dynamic tools, by actually
+other, whether by working with the compiler's intermediate
+representations, examining binaries produced by the compiler, tracing
+the program's execution, or, in the case of dynamic tools, actually
 executing the program.

### 2024-02-26 ootavacuous.tex
Further tweaks for Chapter 8
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Appendix~\ref{app:User Influence Over Language Semantics}
-and the discussion
+and by the discussion
 in~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.
@@
 but such reordering is in any case directly forbidden by the memory model
-(the four coherence rules in 6.9.2.2p14--19).
+(the four coherence rules in 6.9.2.2p14--19 \co{[intro.races]}).
 The ``onto'' bullet also largely reiterates restrictions that follow

### 2024-02-27 ootavacuous.tex
Add executable formal hardware memory-model citations
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 that accurate and executable formal descriptions of hardware memory
-models did not appear until after the standard was released.
+models did not appear until after the standard was
+released~\cite{JadeAlglave2011ppcmem,Maranget2012TutorialARMPower}.
 But they do exist now.

### 2024-02-27 ootavacuous.tex
Edits for the IPC appendix
Here is a number of fairly small changes to the appendix on IPC.
Probably the largest change is in the name, replacing "process" with
"thread".

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 (rfi) or between threads (rfe).\footnote{
-	See Appendix~\ref{app:Inter-Process Communications} for definitions
+	See Appendix~\ref{app:Interthread Communications} for definitions
 	and properties of rf, rfe, and rfi.}
@@
 
-\section{Inter-Process Communications}
-\label{app:Inter-Process Communications}
-
-This section more clearly defines the ``coe'', ``fre'', and ``rfe''
-types of interprocess communication (also collectively called ``links'')
-and gives evidence on the temporal qualities of these different types.
-
-First, context.
-The final ``e'' in all three acronyms stands for ``external'' as opposed
-to within-thread communication (``coi'', ``fri'', and ``rfi'') and as
-also opposed to all communication, whether inter-thread or intra-thread
-(``co'', ``fr'', and ``rf'').
+\section{Interthread Communications}
+\label{app:Interthread Communications}
+
+This section more precisely defines the ``coe'', ``fre'', and ``rfe''
+types of interthread communication (also collectively called ``links'')
+and presents information about their temporal qualities.
+
+First, naming.
+The final ``e'' in all three acronyms stands for ``external'' (that
+is, between threads) as opposed to ``internal'' communication (within
+a thread: ``coi'', ``fri'', and ``rfi'') and also as opposed to all
+communication, whether interthread or intrathread (``co'', ``fr'', and
+``rf'').
 
 Next, definitions:
-
 \begin{description}
 \item[coe]
-	Coherency (external), which connects a store to any other store
-	to that same object that overwrites either the first store's
-	value or that overwrites some later value.
+	Coherence-order (external), which connects a store to any other store
+	for the same object that comes later in the object's
+	modification order, i.e., that overwrites either the first store's
+	value or some later value.
 \item[fre]
-	From-read (external), which connects a load to any other store
-	object that overwrites either the value that the load returned
-	or that overwrites some later value.
+	From-read (external), which connects a load to any store for
+	the same object that overwrites either the value that the load
+	returned or some later value.
 \item[rfe]
 	Reads-from (external), which connects a store to any load from
-	that same object that returns the value stored.
+	the same object that returns the value stored.
+	(Note: This means that the load retrieves the information
+	written by the store, as opposed to retrieving the information
+	from a different store that happened to write the same value.)
 \end{description}
@@
 \resizebox{4in}{!}{\includegraphics{ipc}}
-\caption{IPC Diagram for coe, fre, and rfe}
-\label{fig:IPC Diagram for coe, fre, and rfe}
+\caption{ITC Diagram for coe, fre, and rfe}
+\label{fig:ITC Diagram for coe, fre, and rfe}
 \end{center}
@@
 
-Thirdly, Figure~\ref{fig:IPC Diagram for coe, fre, and rfe} illustrates
+Thirdly, Figure~\ref{fig:ITC Diagram for coe, fre, and rfe} illustrates
 the coe, fre, fri, and rfe links for four threads with time advancing
-from the top to the bottom of the figure.  Starting with coe, even though
-Thread~1's store was later in global time than that of Thread~0, that of
-Thread~1 came first in coe order, followed by those of Threads~2 and ~3.
+from the top to the bottom of the figure.  Start with coe: Even though
+Thread~1's store is later in global time than that of Thread~0, the
+Thread~1 store comes first in \co{x}'s modification order, then Thread-0's
+store, then those of Threads~2 and ~3.
 There is thus a coe link from Thread~1's store to every other store,
@@
 
-Thread~3's load reads the value stored by Thread~0's store, so there is
-an rfe link from that store to that load.  There is also an fre link from
-that load to Thread~2's store and an fri link to Thread~3's later store.
-Note that this fre link goes backwards in time.
-
-Finally, there might be a desire for hard evidence that coe and fre
+Thread~3's load reads the value written by Thread~0's store, so there is
+an rfe link from that store to that load.  In addition there is an fre link from
+the load to Thread~2's store and an fri link to Thread~3's later store.
+Note that the fre link goes backwards in time.
+
+Finally, the reader might desire hard evidence that coe and fre
 links really can go backwards in time.
 We provide this evidence on x86 to demonstrate that these effects
-are not confined to weakly ordered systems.
-This system is a dual-socket system with
-Intel(R) Xeon(R) Gold 6138 CPU @ 2.00GHz,
-each socket with 20~cores, each having a pair of hardware
+are not confined to weakly ordered architectures.
+The machine is a dual-socket system with
+Intel(R) Xeon(R) Gold 6138 CPUs @ 2.00~GHz,
+each socket with 20~cores and each core having a pair of hardware
 threads, for a grand total of 80 hardware threads.
@@
 in time, and quite frequently at that, even on x86.
-The store-to-store latency is measured from the end of the ``winning''
-store to the beginning of the nonwinning store that started latest.
-Every data point on the negative x-axis represents a group of runs
-in which the winning store was not the last store.
-Thus, the winning store not being the last store is quite common, even
-on x86.
+During a run of the test, each of the threads stores a distinct value to
+the same atomic integer variable (all at about the same time), and each
+one records timestamps just before and just after its store.
+The ``winning'' store is the one whose value is retained in the
+variable at the end of the run, overwriting all the others.
+The store-to-store latency is measured from the beginning of the
+nonwinning store that started latest to the end of the winning store.
+Every data point on the negative x-axis thus represents a group of
+runs in which a nonwinning store started after the winning store
+finished.
+This demonstrates that the winning store quite commonly is not the
+last one, even on x86.
 
@@
 
-However, Figure~\ref{fig:On x86, coe Links Form a Coherent Partial Order}
-shows that the order of reads of each value at each thread are
-consistent with a number of global orders, as required.\footnote{
-	Note that this data was not measured on the 80-CPU system, but
-	rather on a 16-CPU x86 laptop to avoid an unreadable diagram
+Figure~\ref{fig:On x86, coe Links Form a Coherent Partial Order} shows
+that the sequence of values read from the atomic variable by each
+thread is consistent with a number of global orders, as
+required.\footnote{
+	Note that this data was not measured on the 80-thread system but
+	rather on a 16-thread x86 laptop, to avoid an unreadable diagram
 	containing 79 bubbles.}
-Working from the far left, one CPU saw the values 0, 11, and 9,
-a second CPU saw 3, 11, and 9, and so on.
-Although different CPUs saw different sequences of values, there is no
+Working from the far left, one thread saw the values 0, 11, and 9,
+a second thread saw 3, 11, and 9, and so on.
+Although different threads saw different sequences of values, there is no
 disagreement on the order of the values that they could see.
-For example, all CPUs agree that the value 9 came last.
-This is an example of single-variable sequential consistency, and
+For example, all threads agree that the value 9 came last.
+This is an example of single-variable sequential consistency, and it
 meets the coherence requirements of the C++ standard.
@@
 Figure~\ref{fig:On x86, fre Links Are Atemporal}
-plots a histogram of the elapsed time from the beginning of load to
-return the old value to the end of the store that provided the new value.
-Note well that most of the data falls into negative time, that is to
-say, in the most commonly occurring case, the last load of the old value
+plots a histogram of the elapsed time from the beginning of a load that
+returned an old value to the end of the store that provided the new value.
+Most of the data falls into negative time.
+In fact, in the most commonly occurring case the last load of an old value
 executes about 60 timestamp periods (about 30~nanoseconds) \emph{after}
-the store that overwrote that same old value.
-That is, fre links can and do go backwards in time.
-This corresponds to the fre link between Thread~3 and Thread~2 depicted in
-Figure~\ref{fig:IPC Diagram for coe, fre, and rfe}.
+the store which overwrote that old value.
+This shows that fre links can and do go backwards in global time,
+corresponding to the fre link between Thread~3 and Thread~2 in
+Figure~\ref{fig:ITC Diagram for coe, fre, and rfe}.
 
@@
 Figure~\ref{fig:On x86, rfe Links Are Temporal}
-plots a histogram of the elapsed time from a store of a new value
+plots a histogram of the elapsed time from the store of a new value
 to the first load of that new value.
-Note well that all of the data falls into positive time, indicating
-that rfe links always go forward in time, as those familiar with
+Here all of the data falls into positive time, indicating
+that rfe links always go forward in global time, as those familiar with
 computer hardware and the laws of physics would expect.
@@
 The upper portion of the figure shows an atemporal coe, in which
-CPU~0 stores 1 to \co{x} after CPU~3 stores 2 to \co{x}, but
-the ordering of the stores is decided after the fact.
+CPU~0 stores 1 to \co{x} after CPU~3 stores 2 but
+the modification order of \co{x} is decided after the fact.
 In this case, the value of 2 from CPU~3's store overwrites the value
 of 1 from CPU~0's store, despite the fact that CPU~0's store
-happened later in time.
-One way that this could happen is that the cache line containing
-\co{x} happened to arrive at CPU~3 before arriving at CPU~0.
-A more fanciful explanation is that both values arrived at the
-interconnect at the saem time, and the interconnect made an
+happened later in global time.
+One way this could happen is if the cache line containing
+\co{x} arrived at CPU~3 before arriving at CPU~0.
+A more fanciful explanation is that both values arrived at an
+interconnect at the same time, and the interconnect made an
 arbitrary choice between the two.
@@
 The lower portion of the figure shows a temporal rfe.
-Because CPU~3's load cannot obtain the value stored by CPU~0 until
+Because a load on CPU~3 cannot obtain the value stored by CPU~0 until
 after that value has propagated to CPU~3, the fact that CPU~3 obtains
@@
 And this is exactly why the C++ memory model guarantees ordering from
-rfe links but not the coe and fre links for relaxed accesses in the
+rfe links but not from coe and fre links for relaxed accesses (in the
 absence of other ordering from stronger atomic memory accesses or
-\co{atomic_thread_fence(memory_order_seq_cst}.
+\co{atomic_thread_fence(memory_order_seq_cst)}).
 

### 2024-02-29 ootavacuous.tex
Edits for Appendices B and C
Here is a bunch of small changes to Appendices B and C.  The largest set
of changes involves altering the line breaks in the GCC command-line
example to prevent overfull lines.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 some degree of
-influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
-and each of which is subject to change over time.
+influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.
 It is natural to dismiss user influence when compared to the text of
@@
 change and do change over time.
-The implementation is especially subject to change, for example,
-consider the following command line used to build the Linux kernel's
+An especially easy way for users to influence the implementation is by
+means of command-line flags and switch settings.
+As an example,
+consider the following command line used to compile the Linux kernel's
 \co{kernel/rcu/tree.c} C-language source file:
 
-~\\
+\begin{quote}
 {
 	\scriptsize
-	\texttt{
+	\texttt{\noindent
 	gcc -Wp,-MMD,kernel/rcu/.tree.o.d -nostdinc
@@
 	-I./arch/x86/include/generated/uapi
-	-I./include/uapi -I./include/generated/uapi \\
-	-include ./include/linux/compiler-version.h
+	-I./include/uapi \\
+	-I./include/generated/uapi \\
+	-include ./include/linux/compiler-version.h \\
 	-include ./include/linux/kconfig.h \\
-	-include ./include/linux/compiler\_types.h -D\_\_KERNEL\_\_
-	-fmacro-prefix-map=./= \\
-	-Werror -std=gnu11 -fshort-wchar
-	-funsigned-char -fno-common -fno-PIE \\
-	-fno-strict-aliasing -mno-sse
+	-include ./include/linux/compiler\_types.h \\
+	-D\_\_KERNEL\_\_
+	-fmacro-prefix-map=./=
+	-Werror -std=gnu11 -fshort-wchar \\
+	-funsigned-char -fno-common -fno-PIE
+	-fno-strict-aliasing \\
+	-mno-sse
 	-mno-mmx -mno-sse2 -mno-3dnow -mno-avx \\
 	-fcf-protection=branch
-	-fno-jump-tables -m64 -falign-jumps=1 -falign-loops=1 \\
-	-mno-80387 -mno-fp-ret-in-387 -mpreferred-stack-boundary=3
-	-mskip-rax-setup \\
+	-fno-jump-tables -m64 -falign-jumps=1 \\
+	-falign-loops=1 -mno-80387 -mno-fp-ret-in-387 \\
+	-mpreferred-stack-boundary=3 -mskip-rax-setup \\
 	-mtune=generic -mno-red-zone -mcmodel=kernel
@@
 	-mindirect-branch-register
-	-mindirect-branch-cs-prefix -mfunction-return=thunk-extern
-	-fno-jump-tables -fpatchable-function-entry=16,16
-	-fno-delete-null-pointer-checks -O2 \\
-	-fno-allow-store-data-races
-	-fstack-protector-strong -fomit-frame-pointer \\
-	-fno-stack-clash-protection -falign-functions=16
-	-fno-strict-overflow -fno-stack-check \\
-	-fconserve-stack
-	-Wall -Wundef -Werror=implicit-function-declaration \\
-	-Werror=implicit-int -Werror=return-type -Werror=strict-prototypes \\
-	-Wno-format-security -Wno-trigraphs -Wno-frame-address
-	-Wno-address-of-packed-member \\
-	-Wframe-larger-than=2048 -Wno-main
+	-mindirect-branch-cs-prefix \\
+	-mfunction-return=thunk-extern -fno-jump-tables \\
+	-fpatchable-function-entry=16,16
+	-fno-delete-null-pointer-checks \\
+	-O2 -fno-allow-store-data-races
+	-fstack-protector-strong \\
+	-fomit-frame-pointer
+	-fno-stack-clash-protection -falign-functions=16 \\
+	-fno-strict-overflow -fno-stack-check -fconserve-stack
+	-Wall -Wundef \\
+	-Werror=implicit-function-declaration
+	-Werror=implicit-int \\
+	-Werror=return-type
+	-Werror=strict-prototypes -Wno-format-security \\
+	-Wno-trigraphs -Wno-frame-address \\
+	-Wno-address-of-packed-member
+	-Wframe-larger-than=2048 -Wno-main \\
+	-Wno-unused-but-set-variable -Wno-unused-const-variable -Wvla \\
+	-Wno-pointer-sign -Wcast-function-type -Wno-array-bounds \\
+	-Wno-alloc-size-larger-than -Wimplicit-fallthrough=5 \\
+	-Werror=date-time -Werror=incompatible-pointer-types \\
+	-Werror=designated-init -Wenum-conversion
 	-Wno-unused-but-set-variable \\
-	-Wno-unused-const-variable -Wvla
-	-Wno-pointer-sign -Wcast-function-type \\
-	-Wno-array-bounds
-	-Wno-alloc-size-larger-than -Wimplicit-fallthrough=5 \\
-	-Werror=date-time -Werror=incompatible-pointer-types
-	-Werror=designated-init \\
-	-Wenum-conversion
-	-Wno-unused-but-set-variable -Wno-unused-const-variable \\
-	-Wno-restrict -Wno-packed-not-aligned -Wno-format-overflow
-	-Wno-format-truncation \\
-	-Wno-stringop-overflow
-	-Wno-stringop-truncation -Wno-missing-field-initializers \\
-	-Wno-type-limits -Wno-shift-negative-value
-	-Wno-maybe-uninitialized \\
+	-Wno-unused-const-variable -Wno-restrict -Wno-packed-not-aligned \\
+	-Wno-format-overflow 
+	-Wno-format-truncation -Wno-stringop-overflow \\
+	-Wno-stringop-truncation
+	-Wno-missing-field-initializers \\
+	-Wno-type-limits
+	-Wno-shift-negative-value -Wno-maybe-uninitialized \\
 	-Wno-sign-compare 
-	-DKBUILD\_MODFILE='"kernel/rcu/tree"'
-	-DKBUILD\_BASENAME='"tree"' \\
-	-DKBUILD\_MODNAME='"tree"' -D\_\_KBUILD\_MODNAME=kmod\_tree \\
+	-DKBUILD\_MODFILE='"kernel/rcu/tree"' \\
+	-DKBUILD\_BASENAME='"tree"' -DKBUILD\_MODNAME='"tree"' \\
+	-D\_\_KBUILD\_MODNAME=kmod\_tree \\
 	-c -o kernel/rcu/tree.o kernel/rcu/tree.c
@@
 }
+\end{quote}
 
@@
 
-The \co{-funsigned-char} causes type \co{char} to be unsigned, which
+The \co{-funsigned-char} causes the \co{char} type to be unsigned, which
 overrides per-architecture defaults, some of which treat \co{char}
@@
 the arithmetic proporties of this type.
-In theory, the standard could specify the signedness of \co{char}, but the
-variety of existing practice in the 1980s prevented this.
+In theory, the standard could have specified the signedness of \co{char}
+but the variety of existing practice in the 1980s prevented this.
 Plus machine instructions were much more expensive back then, so the
 signedness of \co{char} was much more of a performance concern than
-it is toda than
-it is todayy.
-
-The \co{-mno-sse} prevents GCC from making use of the SSE hardware.
+it is today.
+
+The \co{-mno-sse} prevents GCC from making use of the processor's SSE hardware.
 This is done for performance reasons, as it avoids the overhead of
@@
 monitoring.
-This is again clearly outside of the scope of the standard.
+This is again clearly outside the scope of the standard.
 
 The \co{-fstack-protector-strong} causes GCC to emit code that provides
-some protection against some clases of attacks based on buffer overflows.
+some protection against some classes of attacks based on buffer overflows.
 One could rightly argue code should simply avoid ever overflowing buffers,
@@
 The \co{-fno-strict-overflow} causes GCC to act as if signed integer
-overflow is defined behavior.
-Note well that this command-line option also affects the definition of
-``semantic dependency''.
+overflow is defined behavior,
+which also affects the definition of ``semantic dependency''.
 This might be a controversial choice, and another option would be to add
@@
 This helps avoid certain classes of bugs.
-Warnings are by design outside of the scope of the standard.
-
-In other words, user preference can exert a nontrivial influence over
+Warnings are by design outside the scope of the standard.
+
+To sum up, user preference can exert a nontrivial influence over
 language semantics, and in particular can affect some aspects of the
@@
 by looking into the possibility of:
-
-\begin{enumerate}
+\begin{itemize}
 \item	Tooling focusing on only part of the language,
@@
 	tooling can easily accommodate.
-\end{enumerate}
+\end{itemize}
 
@@
 
-Both the ARM and PowerPC memory models permit reordering prior loads
+Both the ARM and PowerPC architectural memory models permit reordering prior loads
 against subsequent stores, but actual tests on recent hardware fail to
@@
 
-However, if that load suffers a cache miss, address translation will have
+However, if the load suffers a cache miss, address translation will have
 completed long before the load returns its value, meaning that a later
 store might well commit before the load completes.
-In fact, that later store might commit before the store that supplies the
-value returned by that load!
+In fact, that later store might commit before the store which supplies the
+value returned by the load!
 Which explains why some hardware systems really do reorder prior loads
@@
 other processes.
-Some user applications might be note that the affected data affected only
+Some user applications might note that the corrupted data affected only
 a particular computation, and might react by restarting that computation.
@@
 
-If you would like an authoratative answer to this question, you should of
+If you would like an authoritative answer to this question, you should of
 course ask your friendly local hardware architect.
@@
 
-\begin{enumerate}
+\begin{itemize}
 \item	Some hardware might choose to forego ECC, for example, in order to
@@
 	this option open for their own systems.
-\end{enumerate}
+\end{itemize}
 
@@
 optimizations that tooling could in fact analyze.
-Therefore, the next section looks at standardizaing a less severe
+Therefore, the next section looks at standardizing a less severe
 restriction.

### 2024-03-01 ootavacuous.tex
Litmus name to filename and default initialization
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 The first line identifies it as a C-language litmus test and gives it
-a name.
+a name, and this name identifies the litmus test's source file within
+the \path{litmus} directory in the \url{https://github.com/paulmckrcu/oota}
+repository.
 Lines~2--5 initialize variables, in this case setting the initial
 values of the global shared variables \co{x} and \co{y} to zero.
+Variables that are not explicitly initialized are initialized by
+default to zero.
 Lines~7--13 define the first thread, \co{P0()}, and lines~15--21

### 2024-03-01 ootavacuous.tex
Call out older single-variable perturbation citation
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 can cause the value stored in \co{x} to change.
-This means that accurate definitions of sdep cannot always rely on
+This means that, prior work~\cite{PaulEMcKenney2016OOTA}
+notwithstanding, accurate definitions of sdep cannot always rely on
 single-variable perturbations;

### 2024-03-01 ootavacuous.tex
abstract: Add "shared memory" to avoid I/O
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Abstract Executions: Standard does not require store before load
In contrast, we do require stores to happen before loads yielding
the stored value.  In this, we defer to the objective universe.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	long-standing laws of physics and from hardware systems.
-	In this initial work, we focus on C++ implementations based
-	on compilers and computing hardware (including CPUs and GPGPUs).
+	In this initial work, we focus on shared-memory programs running
+	on C++ implementations based on compilers and computing hardware
+	(including CPUs and GPGPUs).
 	We present informal arguments that these constraints and
@@
 in different threads.  Even if a load in one thread reads from a
-store in another thread, we don't require the store to come before
-the load.
+store in another thread, the standard does not require the store to come
+before the load.
 

### 2024-03-01 ootavacuous.tex
Relation Between...: Remove extraneous apostrophe
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 which always must return the value of the store they read from).
-This implies that volatile load-acquire's do not synchronize with
+This implies that volatile load-acquires do not synchronize with
 volatile store-releases in the sense of the C++ memory model,\footnote{

### 2024-03-01 ootavacuous.tex
Global Optimization...: Loose C++ and avoid OOTA
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \co{y} while \co{thread2()} does the reverse.
-A globally optimizing compiler given that program might
+A globally optimizing loose C++ compiler given that program might
 transform it to the following before translating it into machine code,
@@
 memory model by assuming each thread's load reads from the other's
-store.
+store.\footnote{
+	Such a compiler could presumably instead choose to avoid this
+	OOTA cycle, for example, by retaining the loads from \co{x}
+	and \co{y}.}
 A similar justification can underlie the reasoning in

### 2024-03-01 ootavacuous.tex
Definition sdep: s/fundamantal/fundamental/
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Section~\ref{sec:Verifying the Fundamental Property}
-verifies the fundamantal property of semantic dependencies,
+verifies the fundamental property of semantic dependencies,
 and

### 2024-03-01 ootavacuous.tex
Verifying Fundamental...: s/abtract/abstract/
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 The Fundamental Property assumes that $W$ is not omitted in $H$, so
-there is an abtract execution $B$ with hardware realization $G$
+there is an abstract execution $B$ with hardware realization $G$
 witnessing the semantic dependency of the store $W$ on the loads

### 2024-03-01 ootavacuous.tex
Verifying Fundamental...: Add "the execution of"
The memory model allows reordering of loads after conditional
branches, so explicitly state that the ordering refers to the
execution (as opposed to the memory effects).

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Global Analysis...: Qualify with "loose C++"
In contrast, a strict C++ compiler would need to keep the loads or
otherwise avoid the OOTA cycle.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 the two executions it must be ordered after one of the $R'_i$ loads.
-And then so must $W'$, because any instruction following a conditional-
-or computed-branch instruction is ordered after that branch.
+And then so must $W'$, because the execution of any instruction following
+a conditional- or computed-branch instruction is ordered after that
+branch.
 QED.
@@
 \end{quote}
-A compiler using global analysis and treating \co{x} and \co{y} as
+A loose C++ compiler using global analysis and treating \co{x} and \co{y} as
 quasi-volatile objects could omit the two loads, replacing them in the

### 2024-03-01 ootavacuous.tex
Add to Future Directions section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 optimizations that combine multiple threads into one.
+
+This paper considers only fragments of programs whose threads
+communicate using shared memory.
+Future work might include the effects of input and output, as well
+as code that reads out the current time, that uses thread IDs,
+or that operates on pointer values.
+
+Finally, future work might more precisely delineate the limits of
+quasi-volatile behavior.
 

### 2024-03-01 ootavacuous.tex
Conclusion: Rework based on changes
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \co{acyclic(sdep; rf)} to \co{acyclic(sdep; rfe)}, which enables the
-application the laws of physics (and thus of temporal reasoning) to OOTA
+application of the laws of physics (and thus of temporal reasoning) to OOTA
 analysis of the C++ memory model.
-This reasoning shows that correct real-world C++ implementations relying
-on single-threaded analyses are inherently incapable of producing OOTA
-cycles in UB-free programs using volatile atomics.
-
-C++ implementations that additionally refrain from inventing atomic loads
-and that also ensure that each atomic load that is executed corresponds
-to a specific load in the abstract machine are incapable of producing
-OOTA cycles in UB-free programs using nonvolatile as well
-as volatile atomics.
-
-This paper also uncovers some shortcomings in typical definitions of
-``semantic dependency'', pointing out examples where a single semantic
-dependency extends from multiple loads to a store, but not from any one
-of those loads to that same store.
-We therefore propose that the definition of semantic dependency be
-extended accordingly.
+This refinement, combined with a survey of code containing non-trivial
+semantic dependencies, uncovered some shortcomings in typical definitions of
+``semantic dependency'', pointing out examples where semantic dependencies
+are a function of an execution rather than of the source code, and also
+examples where a single semantic dependency extends from multiple loads
+to a store, but not from any one of those loads to that same store.
+These complex dependencies motivated a generalization of the definition
+of ``OOTA cycle'', which is shown in
+Section~\ref{sec:Semantic Dependencies Can Be Many To One}.
+
+This generalization, combined with a focus on compiler-based C++
+implementations using single-threaded analyses, led to the fundamental
+property of semantic dependency shown in
+Section~\ref{sec:The Fundamental Property of Semantic Dependencies}.
+This property in turn led to the precise definition of ``semantic
+dependency'' shown in
+Section~\ref{sec:For Compilers Using Single-Thread Analysis}.
+These definitions are then used in
+Section~\ref{sec:Verifying the Fundamental Property}
+to demonstrate that correct real-world compiler-based C++
+implementations relying on single-threaded analyses are inherently
+incapable of producing OOTA cycles in UB-free programs using volatile
+atomics.
+
+C++ implementations that additionally confine optimizations of code
+containing nonvolatile atomic operations to the quasi-volatile limitations
+outlined in
+Section~\ref{sec:Volatile and Quasi Volatile Accesses}
+are incapable of producing OOTA cycles in UB-free programs using
+nonvolatile as well as volatile atomics.
 

### 2024-03-01 ootavacuous.tex
Future Directions: Add material from Section 7.4
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+This paper shows that semantic dependencies are relative to specific
+executions of a program rather than strictly of the source code of that
+program.
+Future work might expand on
+Section~\ref{sec:Relative versus Absolute Dependency},
+further exploring absolute semantic dependencies that are special cases
+where semantic dependencies can be a function strictly of the source code.
+
 This paper considers only fragments of programs whose threads
 communicate using shared memory.
-Future work might include the effects of input and output, as well
-as code that reads out the current time, that uses thread IDs,
-or that operates on pointer values.
-
-Finally, future work might more precisely delineate the limits of
-quasi-volatile behavior.
+Future work might include the effects of input and output, as well as
+code that reads out the current time, that uses thread IDs, or that
+operates on pointer values
+(see Section~\ref{sec:Effect of Memory Layout}).
+
+Finally, future work might expand on
+Sections~\ref{sec:Global Analysis and Volatile versus Quasi Volatile},
+and~\ref{sec:Merging Quasi-Volatile Loads}
+so as to more precisely delineate the limits of permissible quasi-volatile
+behavior.
 

### 2024-03-01 ootavacuous.tex
March 1 Alan Stern feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 the two executions it must be ordered after one of the $R'_i$ loads.
-And then so must $W'$, because the execution of any instruction following
-a conditional- or computed-branch instruction is ordered after that
-branch.
+And then so must $W'$, because any instruction following a conditional-
+or computed-branch instruction must commit after that branch
+commits.\footnote{
+	But note that a load might obtain its value from memory before
+	previous instructions have committed.}
 QED.

### 2024-03-02 ootavacuous.tex
Capitalize "Section"
The word "section" should be capitalized when it refers to a section of
the paper.  Make it so.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 other threads (a point we will return to in
-section~\ref{sec:Global Optimization Can Destroy Dependencies}).
+Section~\ref{sec:Global Optimization Can Destroy Dependencies}).
 % The usual way in the Linux kernel is via a union, but it is probably
@@
 (but see the discussion of volatile loads in
-section~\ref{sec:Relation Between Abstract and Hardware Executions} below).
+Section~\ref{sec:Relation Between Abstract and Hardware Executions} below).
 The totality of this information---along with the program's source
@@
 invent, omit, merge, or reorder these accesses, as we will discuss in
-section~\ref{sec:Volatile and Quasi Volatile Accesses} below.
+Section~\ref{sec:Volatile and Quasi Volatile Accesses} below.
 
@@
 A similar justification can underlie the reasoning in
-section~\ref{sec:Semantic Dependencies Affected by Cross-Thread Optimizations};
+Section~\ref{sec:Semantic Dependencies Affected by Cross-Thread Optimizations};
 in principle an analysis of the complete program could lead a compiler
@@
 thread, i.e., the rfe relation.
-As mentioned in section~\ref{sec:Simple OOTA Cycle} above, a CPU does
+As mentioned in Section~\ref{sec:Simple OOTA Cycle} above, a CPU does
 not make the value of a store instruction available for other CPUs to
@@
 means of the map of accesses described in
-section~\ref{sec:Volatile and Quasi Volatile Accesses}.
+Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
 The key insight is that this allows us to consider semantic
@@
 particular execution and a particular implementation,
-as described in section~\ref{sec:Properties of Semantic Dependencies}.
+as described in Section~\ref{sec:Properties of Semantic Dependencies}.
 The same source code may or may not contain a semantic dependency,
@@
 satisfies the Fundamental Property of
-section~\ref{sec:The Fundamental Property of Semantic Dependencies}.
+Section~\ref{sec:The Fundamental Property of Semantic Dependencies}.
 Given the information we have already presented, the demonstration is
@@
 dependencies relies on the fact, stated in
-section~\ref{sec:Abstract Executions},
+Section~\ref{sec:Abstract Executions},
 that an abstract execution of a thread is entirely determined by the
@@
 the example in
-section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}:
+Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}:
 \begin{quote}
@@
 This test case is virtually the same as the Simple OOTA cycle
-discussed in section~\ref{sec:Simple OOTA Cycle}, the only difference
+discussed in Section~\ref{sec:Simple OOTA Cycle}, the only difference
 being the final values of the variables.

### 2024-03-02 ootavacuous.tex
Hyphenate "quasi-volatile" correctly
The term "quasi-volatile" should be hyphenated when and only when it
modifies an immediately following noun.  Make it so.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 then presents required properties of volatile atomic operations, and
-also defines quasi-volatile properties for nonvolatile atomic operations.
+also defines properties of quasi-volatile atomic operations.
 
@@
 that treat all atomic objects as though they are volatile or
-quasi-volatile.
+quasi volatile.
 For compilers that perform only single-thread analysis, not global
-analysis, quasi-volatile is sufficient.
+analysis, quasi volatile is sufficient.
 In this setting we can relate abstract and hardware executions by
@@
 only single-thread analysis and treat all atomic objects as
-quasi-volatile.
+quasi volatile.
 This implies that if two different programs contain the same thread
@@
 Section~\ref{sec:Volatile and Quasi Volatile Accesses}
-prohibits reordering of quasi volatile accesses when they are
+prohibits reordering of quasi-volatile accesses when they are
 to the same object,
@@
 and~\ref{sec:Merging Quasi-Volatile Loads}
-so as to more precisely delineate the limits of permissible quasi-volatile
-behavior.
+so as to more precisely delineate the limits of permissible
+behavior for quasi-volatile object accesses.
 

### 2024-03-03 ootavacuous.tex
Add explanation of ordering vis à vis commit order and loads
Rework the recently added disclaimer about loads and commit ordering:
Move it up to the place where the concept is first introduced and
explain more explicitly what "ordered after" means.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-However, dependencies aren't the only mechanism capable of forcing
+We will say that one instruction is ordered after another to mean that
+the CPU forces it to commit after the other one commits.
+Hence instructions are ordered after the instructions they depend on.
+(Note that this implies nothing about when a load instruction retrieves
+its value from memory; it may do so long before it or a prior
+instruction commits.)
+
+Dependencies aren't the only mechanism that can lead to
 ordering in hardware executions.
@@
 And then so must $W'$, because any instruction following a conditional-
-or computed-branch instruction must commit after that branch
-commits.\footnote{
-	But note that a load might obtain its value from memory before
-	previous instructions have committed.}
+or computed-branch instruction must commit after the branch commits.
 QED.

### 2024-03-03 ootavacuous.tex
Edits on top of the recent batch of changes
These edits refine the changes committed in the last day or two.  They
also remove the waffling on the term "loose C++", nobody seems to mind
it any more.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 prohibition above.
-{\it Contra\/} that article, we will use the term \emph{loose C++}\footnote{
-	There is some discomfort with this name, so this footnote
-	is reserved for future bikeshedding.}
-% @@@ Alternatives thus far: OOTA-prone C++, OOTA-enabled C++,
-% @@@ relaxed C++, OOTA C++, OOTA-agnostic C++.
+{\it Contra\/} that article, we will use the term \emph{loose C++}
 (rather than \emph{strict C++}, which seems too similar to
@@
 values of the global shared variables \co{x} and \co{y} to zero.
-Variables that are not explicitly initialized are initialized by
-default to zero.
+Variables that are not explicitly initialized are initialized
+to zero by default.
 Lines~7--13 define the first thread, \co{P0()}, and lines~15--21
@@
 store.\footnote{
-	Such a compiler could presumably instead choose to avoid this
-	OOTA cycle, for example, by retaining the loads from \co{x}
-	and \co{y}.}
+	A less perverse compiler could choose to avoid the
+	OOTA cycle simply by not making this transformation.}
 A similar justification can underlie the reasoning in
@@
 This paper shows that semantic dependencies are relative to specific
-executions of a program rather than strictly of the source code of that
-program.
+executions of a program and the specific compiler in use rather than
+being determined solely by the source code.
 Future work might expand on
 Section~\ref{sec:Relative versus Absolute Dependency},
-further exploring absolute semantic dependencies that are special cases
-where semantic dependencies can be a function strictly of the source code.
-
-This paper considers only fragments of programs whose threads
+exploring special cases in which absolute semantic dependencies are
+functions just of the source code.
+
+This paper considers only programs whose threads
 communicate using shared memory.
-Future work might include the effects of input and output, as well as
-code that reads out the current time, that uses thread IDs, or that
-operates on pointer values
+Future work might include the effects of input and output or
+other operations which can vary from one execution to another,
+such as those based on the current time, process IDs, or pointer values
 (see Section~\ref{sec:Effect of Memory Layout}).
@@
 
-This paper has refined the definition of ``OOTA cycle'' from
-\co{acyclic(sdep; rf)} to \co{acyclic(sdep; rfe)}, which enables the
-application of the laws of physics (and thus of temporal reasoning) to OOTA
-analysis of the C++ memory model.
-This refinement, combined with a survey of code containing non-trivial
-semantic dependencies, uncovered some shortcomings in typical definitions of
-``semantic dependency'', pointing out examples where semantic dependencies
-are a function of an execution rather than of the source code, and also
-examples where a single semantic dependency extends from multiple loads
-to a store, but not from any one of those loads to that same store.
+This paper has presented samples of code containing non-trivial
+semantic dependencies, uncovering some shortcomings in typical definitions of
+``semantic dependency''.
+It pointed out examples where semantic dependencies
+are a function of an execution rather than of the source code, and examples
+where a single semantic dependency extends from multiple loads
+to a store but not from any one of those loads.
 These complex dependencies motivated a generalization of the definition
@@
 
-This generalization, combined with a focus on compiler-based C++
-implementations using single-threaded analyses, led to the fundamental
-property of semantic dependency shown in
+This generalization, combined with a focus on compiler-based loose C++
+implementations using single-thread analysis, and a consideration of
+the relation between source code and machine code, led to the formulation
+of the Fundamental Property of semantic dependency shown in
 Section~\ref{sec:The Fundamental Property of Semantic Dependencies}.
-This property in turn led to the precise definition of ``semantic
-dependency'' shown in
+This in turn led to a precise definition of ``semantic dependency'' given in
 Section~\ref{sec:For Compilers Using Single-Thread Analysis}.
-These definitions are then used in
+This definition was used in
 Section~\ref{sec:Verifying the Fundamental Property}
-to demonstrate that correct real-world compiler-based C++
-implementations relying on single-threaded analyses are inherently
-incapable of producing OOTA cycles in UB-free programs using volatile
-atomics.
-
-C++ implementations that additionally confine optimizations of code
-containing nonvolatile atomic operations to the quasi-volatile limitations
-outlined in
-Section~\ref{sec:Volatile and Quasi Volatile Accesses}
-are incapable of producing OOTA cycles in UB-free programs using
-nonvolatile as well as volatile atomics.
+to demonstrate that these implementations are
+incapable of producing OOTA cycles in UB-free programs,
+provided they treat all nonvolatile atomic objects as
+quasi volatile, obeying the limitations outlined in
+Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
+
+A variant definition of ``semantic dependency'' given in
+Section~\ref{sec:For Compilers Using Global Analysis}, appropriate for
+compiler-based loose C++ implementations that don't restrict
+their code transformations to those based on single-thread analysis,
+was used to show that such implementations
+are incapable of producing OOTA cycles in UB-free programs
+provided they treat all atomic objects as volatile.
 

### 2024-03-03 ootavacuous.tex
"Real-World Constraints": Wordsmith
Consolidate duplicate lack of consideration/accounting.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 in prior work.
-In real-world C++ implementations, the standard's
-abstract machine must be mapped onto real hardware, however, and
-this mapping has not been considered to any significant degree.
-It imposes additional constraints stemming from hardware architecture and
-design,
-which have been only partially accounted for in the standard and other work.
+In real-world C++ implementations, the standard's abstract machine
+must be mapped onto real hardware, however, and this imposes additional
+constraints stemming from hardware architecture and design.
+This mapping and these constraints have been only partially accounted
+for in the standard and other work.
 Part of the reason for this is

### 2024-03-03 ootavacuous.tex
"Constraints of the Standard": Reword
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 value loaded into \co{r1} is one.
-The compiler is therefore forbidden from duplicating the store of one
+The compiler is therefore forbidden from duplicating atomic stores
 unless it can prove that doing so does not negatively affect the
@@
 extra cache misses for no good purpose.
-And even duplicating an atomic load can lead to incorrect results if done
+And duplicating an atomic load can even lead to incorrect results if done
 carelessly.
@@
 of the program with the first of a back-to-back pair of nonvolatile
-relaxed atomic stores omitted is essentially a valid execution of the
+relaxed atomic stores omitted is a valid execution of the
 program with the store present.

### 2024-03-03 ootavacuous.tex
"Interthread Communication": Reword
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	Coherence-order (external), which connects a store to any other store
-	for the same object that comes later in the object's
+	to the same object that comes later in the object's
 	modification order, i.e., that overwrites either the first store's
@@
 \item[fre]
-	From-read (external), which connects a load to any store for
+	From-read (external), which connects a load to any store to
 	the same object that overwrites either the value that the load
@@
 Thread~1's store is later in global time than that of Thread~0, the
-Thread~1 store comes first in \co{x}'s modification order, then Thread-0's
-store, then those of Threads~2 and ~3.
+Thread~1 store comes first in \co{x}'s modification order, then Thread~0's
+store, then those of Threads~2 and~3.
 There is thus a coe link from Thread~1's store to every other store,
@@
 an rfe link from that store to that load.  In addition there is an fre link from
-the load to Thread~2's store and an fri link to Thread~3's later store.
+the load to Thread~2's earlier store and an fri link to Thread~3's later store.
 Note that the fre link goes backwards in time.

### 2024-03-03 ootavacuous.tex
"But What About Tooling": Reword
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 One entirely reasonable reaction is that, given the issues raised in
-Sections~\ref{sec:Properties of Semantic Dependencies}
-and~\ref{sec:The Fundamental Property of Semantic Dependencies},
+Section~\ref{sec:Properties of Semantic Dependencies},
+Section~\ref{sec:C++ Compilers},
+and
+Section~\ref{sec:Outstanding Issues},
 load/store reordering is the least of the problems faced by tooling.
@@
 But first, the next section looks at why some weakly ordered
-implementations appear to forbid reordering of prior relaxed loads with
+computer hardware appears to forbid reordering of prior relaxed loads with
 later relaxed stores,\footnote{
@@
 Tooling does not require relaxed accesses become fully acquire and/or
-release, but rather only that implemetnations be forbidden from reordering
+release, but rather only that implementations be forbidden from reordering
 prior relaxed loads with subsequent relaxed stores, as has been suggested

### 2024-03-04 ootavacuous.tex
Updates for the Abstract
The abstract is rather old and in need of updating.  These changes will
put it more in tune with the content of the paper.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{abstract}
-	The out-of-thin-air (OOTA) properties of the specification
-	of \co{memory_order_relaxed} have resulted in considerable
-	consternation over the years.
+	The out-of-thin-air (OOTA) properties
+	of \co{memory_order_relaxed} in the C++ memory model
+	have caused considerable consternation over the years.
 	Attempts to create memory models that rule out OOTA behaviors
-	have either been non-executable, complex, or unloved by C++
-	implementers, and this paper summarizes some of the challenges.
+	have been either non-executable, complex, or unloved by C++
+	implementers
+	(and this paper describes some of the challenges of coming to
+	grips with the key notion of semantic dependency).
 	But at the same time, we know of no instances of OOTA behavior
@@
 
-	This paper refactors the commonly used condition for classifying
-	OOTA behavior, which permits bringing to bear constraints from
-	long-standing laws of physics and from hardware systems.
-	In this initial work, we focus on shared-memory programs running
-	on C++ implementations based on compilers and computing hardware
-	(including CPUs and GPGPUs).
-	We present informal arguments that these constraints and
-	limitations prevent OOTA cycles containing only volatile atomics
-	from occurring in undefined-behavior-free C++ programs running on
-	correctly constructed C++ compiler-based implementations that are
-	restricted to single-threaded analysis in our current universe.
-	We extend this work to nonvolatile atomics by defining
-	quasi-volatile behavior that we conjecture that usable real-world
-	C++ implementations must adhere to.
+	We focus on shared-memory programs and
+	C++ implementations based on traditional compilers and
+	computing hardware, including CPUs and GPGPUs.
+	This context permits us to consider the detailed relations
+	between source code and machine code required by the restricted
+	nature of volatile atomic accesses.
+	The OOTA problem is much more tractable at the hardware level
+	than at the source level, thanks to existing formal hardware
+	models, and
+	we show that the constraints in these models prevent OOTA cycles
+	involving only volatile atomics from occurring in
+	undefined-behavior-free C++ programs (under these implementations).
+	We also extend this work to nonvolatile atomics by defining
+	``quasi-volatile'' behavior that we expect C++ implementations
+	will adhere to if they perform single-thread analysis.
 \end{abstract}

### 2024-03-04 ootavacuous.tex
Reorder the examples in section 2.1
Section 2.1 is odd in that it presents an example of non-OOTA behavior
before an example of OOTA.  Swap them so that it makes more sense.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 The collection of loads and stores forms an \emph{OOTA cycle}.
-In the most extreme cases a nonsensical value can pop up ``out of thin air'',
+In the most extreme cases a nonsensical value can pop up ``out of thin air'';
 however, as shown by
@@
 
-\subsubsection{Simple Reordering}
-\label{sec:Simple Reordering}
-
-It is important to distinguish true OOTA cycles from OOTA-like
-behavior caused by simple reordering.
-An example of simple reordering is shown in
-Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
-Both the C++ compiler and the CPU are within their rights to reorder
-lines~12 and~13, which can result in all of \co{x}, \co{y}, \co{r1},
-and \co{r2} having the value 42 as follows:
-\begin{enumerate}
-\item   Line~13 stores 42 to \co{x}.
-\item   Line~6 loads 42 from \co{x} into \co{r1}.
-\item   Line~7 stores \co{r1}, and thus 42, to \co{y}.
-\item   Line~12 loads 42 from \co{y} to \co{r2}.
-\end{enumerate}
-Current C++ implementations can and do exhibit this reordering behavior.
-
 \subsubsection{Simple OOTA Cycle}
@@
 
-In contrast,
+\begin{listing}[tbp]
+\begin{verbatim}
+ 1 atomic<int> x(0);
+ 2 atomic<int> y(0);
+ 3
+ 4 void thread1()
+ 5 {
+ 6   int r1 = x.load(memory_order_relaxed);
+ 7   y.store(r1, memory_order_relaxed);
+ 8 }
+ 9
+10 void thread2()
+11 {
+12   int r2 = y.load(memory_order_relaxed);
+13   x.store(r2, memory_order_relaxed);
+14 }
+\end{verbatim}
+\caption{Simple OOTA}
+\label{lst:Simple OOTA}
+\end{listing}
+
 Listing~\ref{lst:Simple OOTA}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
@@
 
-This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
-in using the term \emph{full C++} to denote the standard including the
-prohibition above.
-{\it Contra\/} that article, we will use the term \emph{loose C++}
-(rather than \emph{strict C++}, which seems too similar to
-\emph{full C++} for comfort)
-to denote a hypothetical standard that excludes this prohibition, but
-is otherwise identical to full C++.
-Unqualified \emph{C++} means full C++.
-
-The next section looks at how prior work has refined these issues.
+\subsubsection{Simple Reordering}
+\label{sec:Simple Reordering}
 
@@
 
-\begin{listing}[tbp]
-\begin{verbatim}
- 1 atomic<int> x(0);
- 2 atomic<int> y(0);
- 3
- 4 void thread1()
- 5 {
- 6   int r1 = x.load(memory_order_relaxed);
- 7   y.store(r1, memory_order_relaxed);
- 8 }
- 9
-10 void thread2()
-11 {
-12   int r2 = y.load(memory_order_relaxed);
-13   x.store(r2, memory_order_relaxed);
-14 }
-\end{verbatim}
-\caption{Simple OOTA}
-\label{lst:Simple OOTA}
-\end{listing}
+It is important to distinguish true OOTA cycles from OOTA-like
+behavior caused by simple reordering.
+An example of simple reordering is shown in
+Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
+Both the C++ compiler and the CPU are within their rights to reorder
+lines~12 and~13, which can result in all of \co{x}, \co{y}, \co{r1},
+and \co{r2} having the value 42 as follows:
+\begin{enumerate}
+\item   Line~13 stores 42 to \co{x}.
+\item   Line~6 loads 42 from \co{x} into \co{r1}.
+\item   Line~7 stores \co{r1}, and thus 42, to \co{y}.
+\item   Line~12 loads 42 from \co{y} to \co{r2}.
+\end{enumerate}
+Current C++ implementations can and do exhibit this reordering behavior.
+
+\medskip
+
+This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
+in using the term \emph{full C++} to denote the standard including the
+prohibition of OOTA mentioned above.
+{\it Contra\/} that article, we will use the term \emph{loose C++}
+(rather than \emph{strict C++}, which seems too similar to
+\emph{full C++} for comfort)
+to denote a hypothetical standard that excludes this prohibition, but
+is otherwise identical to full C++.
+Unqualified \emph{C++} means full C++.
+
+The next section looks at how prior work has refined these issues.
 

### 2024-03-04 ootavacuous.tex
"Illustrative Litmus Tests": s/correct real-world/compiler-based/
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 occurring in undefined-behavior-free C++ programs running on correctly
-constructed C++ compiler-based implemen- tations that are restricted to
+constructed C++ compiler-based implementations that are restricted to
 single-threaded analysis in our current universe.
@@
 atomic loads and relaxed atomic stores.
-Therefore, an OOTA cycle cannot form in a correct real-world C++
+Therefore, an OOTA cycle cannot form in a compiler-based C++
 implementation.
@@
 value stored, so there is a semantic dependency in each thread.
-Therefore, none of these OOTA cycles can happen in a correct real-world
+Therefore, none of these OOTA cycles can happen in a compiler-based
 C++ implementation.
@@
 relaxed atomic loads to their relaxed atomic stores, which prevents this
-OOTA cycle on correct real-world C++ implementations.
+OOTA cycle on compiler-based C++ implementations.
 
@@
 
-Either way, the value-17 OOTA cycle cannot occur in a correct real-world
+Either way, the value-17 OOTA cycle cannot occur in a compiler-based
 C++ implementation.

### 2024-03-04 ootavacuous.tex
"Illustrative Litmus Tests": Remove bogus conjectures
There really are cases where locating an OOTA cycle would require
looking at more than one branch of a complex semantic dependency.
Therefore, remove counterfactual conjectures.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
"Illustrative Litmus Tests": Add section references
The work suggested in the "Non-Trivial Semantic Dependencies" appendix
has been done, so add references to the corresponding sections.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 cycle to form, that cycle would involve both \co{r0} and \co{r1}.
-However, we conjecture that tools designed to find OOTA cycles could
-satisfy themselves with having found either of the two resulting cycles.
 
@@
 compound semantic dependency.
-However, we again conjecture that tools designed to find OOTA cycles could
-satisfy themselves with having found either of the two resulting cycles.
 This situation poses a challenge to current work on semantic dependencies,
@@
 any C++ program is a finite-state machine, which, unlike Turing-complete
-systems, can be analyzed.
-% @@@ reference to cbmc appendix.
+systems, can be analyzed, as demonstrated in
+Section~\ref{sec:Hardware Dependencies, Instruction Ordering, and the Fundamental Property}.
 
@@
 dependency''.
-We propose expanding the traditional definition to include dependencies
-extending from groups of loads to a single store.
+In
+Section~\ref{sec:A Definition of Semantic Dependency},
+we expand the traditional definition to include dependencies extending
+from groups of loads to a single store.
 

### 2024-03-05 ootavacuous.tex
"Illustrative Litmus Tests": Reorder in decreasing importance
This commit pulls to the beginning the most important, surprising,
and illuminating sections of this appendix.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
"Illustrative Litmus Tests": Update lead-in paragraph
Document the rationale for the section ordering.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
abstract: Rework sdep challenges
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	have been either non-executable, complex, or unloved by C++
-	implementers
-	(and this paper describes some of the challenges of coming to
-	grips with the key notion of semantic dependency).
+	implementers.
 	But at the same time, we know of no instances of OOTA behavior
@@
 	nature of volatile atomic accesses.
-	The OOTA problem is much more tractable at the hardware level
-	than at the source level, thanks to existing formal hardware
-	models, and
-	we show that the constraints in these models prevent OOTA cycles
-	involving only volatile atomics from occurring in
+	The OOTA problem and the related challenge of coming to grips
+	with semantic dependency are much more tractable at the hardware
+	level than at the source level, thanks to existing formal
+	hardware models.
+	We show that these models' constraints prevent OOTA cycles
+	that involve only volatile atomics from occurring in
 	undefined-behavior-free C++ programs (under these implementations).
@@
 of defining sdep and identifying OOTA cycles.
-This appendix evaluates them using the \co{herd7} tool and also using
-manual analysis.
-
-\subsection{Simple Reordering as Litmus Test}
-\label{app:Simple Reordering as Litmus Test}
-
-Listing~\ref{lst:Simple Reordering as Litmus Test}
-shows how
-Listing~\ref{lst:Simple Reordering}
-can be recast into a \co{herd7} litmus test.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/simple-reordering.litmus @@
-\caption{Simple Reordering as Litmus Test}
-\label{lst:Simple Reordering as Litmus Test}
-\end{listing}
-
-\subsection{Non-LB Litmus Tests}
-\label{app:Non-LB Litmus Tests}
-
-Listing~\ref{lst:Non-LB Litmus Test}
-shows a litmus test that can produce an OOTA cycle, but that is not
-a member of the load-buffering (LB) class of litmus tests.
-Suppose that lines~9 and~10 both somehow return the value 21.
-Then line~12 will store the value 42 to \co{x}, which will cause
-line~17 to store 21 to \co{y} and line~22 to store 21 to \co{z},
-thus justifying the values initially loaded by lines~9 and~10.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-non-lb.litmus @@
-\caption{Non-LB Litmus Test}
-\label{lst:Non-LB Litmus Test}
-\end{listing}
-
-All thread's loads' return values affect the value stored, so all
-threads have semantic dependencies between their respective relaxed
-atomic loads and relaxed atomic stores.
-Therefore, an OOTA cycle cannot form in a compiler-based C++
-implementation.
-
-Note that in a C++ implementation that (erroneously) permitted an OOTA
-cycle to form, that cycle would involve both \co{r0} and \co{r1}.
-
-\subsection{Multi-Cycle Litmus Tests}
-\label{app:Multi-Cycle Litmus Tests}
-
-Listing~\ref{lst:Multi-Cycle Litmus Test}
-shows an expansion to four threads of Causality Test Case 4, which is shown in
-Listing~\ref{lst:Causality Test Case 4}
-and discussed in
-Appendix~\ref{app:Causality Test Case 4}.
-The \co{herd7} tool shows multiple different possible OOTA cycles,
-as expected.
-However, the value returned by each thread's load directly affects the
-value stored, so there is a semantic dependency in each thread.
-Therefore, none of these OOTA cycles can happen in a compiler-based
-C++ implementation.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-two-source.litmus @@
-\caption{Multi-Cycle Litmus Test}
-\label{lst:Multi-Cycle Litmus Test}
-\end{listing}
-
-\subsection{Three-Process Version of JMM Causality Test Case 4}
-\label{app:Three-Process Version of JMM Causality Test Case 4}
-
-Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
-shows a different way of expanding
-Listing~\ref{lst:Causality Test Case 4}
-(discussed in
-Appendix~\ref{app:Causality Test Case 4})
-from two threads to three threads.
-The \co{herd7} tool finds OOTA values, as expected.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-3proc.litmus @@
-\caption{Three-Process Version of JMM Causality Test Case 4}
-\label{lst:Three-Process Version of JMM Causality Test Case 4}
-\end{listing}
-
-All threads have straightforward semantic dependencies from their
-relaxed atomic loads to their relaxed atomic stores, which prevents this
-OOTA cycle on compiler-based C++ implementations.
+This appendix presents these tests in roughly decreasing order of
+importance, surprise, and illumination.
+It evaluates them using the \co{herd7} tool and also using manual
+analysis.
+
+\subsection{Semantic Dependencies and \co{volatile}}
+\label{app:Semantic Dependencies and volatile}
+
+The Linux kernel uses volatile accesses to constrain the compiler,
+and it is worth looking at the example shown in
+Listing~\ref{lst:Non-Volatile Accesses and Dependencies}
+to see how this works.
+
+\begin{listing}
+\begin{verbatim}
+ 1 atomic<int> x, y;
+ 2
+ 3 void thread1() {
+ 4   int r1 = x.load(memory_order_relaxed);
+ 5   y.store(r1, memory_order_relaxed);
+ 6 }
+\end{verbatim}
+\caption{Non-Volatile Accesses and Dependencies}
+\label{lst:Non-Volatile Accesses and Dependencies}
+\end{listing}
+
+Because both \co{x} and \co{y} are nonvolatile, the compiler is
+free to assume that the value loaded by line~4 will be some value
+stored to \co{x} or its initial value.
+Because there are no stores to \co{x}, the only remaining possibility
+is the default-initialized value of zero.
+Therefore, the C++ implementation is within its rights to substitute
+the constant zero for \co{r1} on line~5, eliminating the semantic
+dependency that would otherwise extend from line~4 to line~5.
+This in turn enables the implementation to reorder line~5 to
+precede line~4.
+
+\begin{listing}
+\begin{verbatim}
+ 1 atomic<volatile int> x, y;
+ 2
+ 3 void thread1() {
+ 4   int r1 = x.load(memory_order_relaxed);
+ 5   y.store(r1, memory_order_relaxed);
+ 6 }
+\end{verbatim}
+\caption{Volatile Accesses and Dependencies}
+\label{lst:Volatile Accesses and Dependencies}
+\end{listing}
+
+One way to preserve this dependency is use of \co{volatile}, as shown in
+Listing~\ref{lst:Volatile Accesses and Dependencies}.
+Now the implementation is forbidden from assuming that the value loaded
+by line~4 has any relation to any stores to or initialization of \co{x},
+even if \co{x} is placed in normal memory.
+
+In this case, the \co{volatile} keyword alerts the implementation to
+the possibility of interfering accesses to \co{x} from outside the
+program, for example, due to \co{x} being:
+
+\begin{itemize}
+\item	Allocated in an MMIO region of the address space.
+\item	Modified by an unknown-to-the-implementation thread
+	or signal handler.
+\item	Subject to I/O-device DMA operations.
+\item	Modified by debugger commands.
+\item	Modified by accesses to facilities such as \path{/dev/mem}.
+\item	Modified by as-yet-unwritten dynamically linked libraries.
+\item	Modified by some other mechanism of the reader's choosing.
+\end{itemize}
+
+These possibilities force the implementation to preserve the semantic
+dependency from line~4 to line~5.
+They can also prevent the formation of OOTA cycles local to the
+implementation because such inteference means that the volatile loads
+might be reading from something that the implementation knows nothing
+about.
+
+The developer is free to arrange for \co{x} to be free of any
+interference, thus obtaining predictable behavior along with the
+preserved semantic dependency.
+However, this example invalidates the rough definition of semantic
+dependency from
+Section~\ref{sec:Simple OOTA Cycle}
+because the value loaded from \co{x} will always be the initial value
+of zero, so that it is at best dubious to talk about any changes in
+the value loaded.
+
+One way forward is to assume that interference could happen when
+determining which dependencies are semantic, and then do further analysis
+using that determination.
+
+\subsection{Non-Trivial Semantic Dependencies}
+\label{app:Non-Trivial Semantic Dependencies}
+
+Most of the examples in this paper involve simple semantic dependencies
+connecting a single load to a single store.
+One exception appears in
+Appendix~\ref{app:Non-LB Litmus Tests},
+but this section will present a more difficult example.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult-0.litmus @@
+\caption{OOTA Multiplication Example, Initial Value Zero}
+\label{lst:OOTA Multiplication Example, Initial Value Zero}
+\end{listing}
+
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+shows an example that builds on the multiplication-by-zero
+discussion in
+Section~\ref{sec:Semantic Dependencies and Source Code}.
+Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
+\co{x} and \co{y} on lines~9 and~10, respectively.
+Similarly, lines~16 and~17 store to \co{x} and \co{y}, respectively,
+the value that line~15 loads from \co{z}.
+Because all three variables are initialized to zero, and because loads
+can be ordered before any store, there is an execution in which all
+three loads return the value zero, in which case all three stores will
+of course store the value zero.
+In any related execution where the value loaded from \co{y} remains zero,
+nonzero values loaded from \co{x} cannot affect the value stored by
+line~11 to \co{z}.
+Only executions which load different (that is, nonzero) values from
+both \co{x} and \co{y} can cause the value stored to \co{z} to change
+from zero to a nonzero value.
+
+Therefore, as noted in
+Section~\ref{sec:Semantic Dependencies and Source Code},
+in an execution where the loads from both \co{x} and \co{y} return zero,
+there is no semantic dependency from the load from \co{x} to the
+store to \co{z}, nor is there a semantic dependency from the load
+from \co{y} to the store to \co{z}.
+However, there \emph{is} a semantic dependency from the \emph{combination}
+of the loads from \co{x} and \co{y} to the store to \co{z}.
+For example, the (forbidden) execution that satisfies the \co{exists}
+clause, which would be an OOTA cycle.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult-0-cond.litmus @@
+\caption{OOTA Conditional Multiplication Example, Initial Value Zero}
+\label{lst:OOTA Conditional Multiplication Example, Initial Value Zero}
+\end{listing}
+
+For further evidence that there is no semantic dependency from \co{y}
+to \co{z} in executions where the load from \co{x} is zero, see the
+equivalent\footnote{
+	But equivalent only because the atomics are all nonvolatile!}
+program shown in
+Listing~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero}.
+Here, there is not even a load from \co{y} in executions where the
+load from \co{x} returns zero, so there cannot possibly be a semantic
+dependency involving this non-existent load.
+A similar transformation would load \co{x} only in executions where the
+load from \co{y} returned nonzero, which similarly emphasizes the lack
+of any semantic dependency between \co{x} and \co{z}.
+
+In both
+Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+and~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero},
+an OOTA cycle (such as the one indicated by their \co{exists} clauses)
+would need to affect both the \co{x} and the \co{y} components of this
+compound semantic dependency.
+This situation poses a challenge to current work on semantic dependencies,
+which typically restricts such dependencies to a single load and store.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult-1.litmus @@
+\caption{OOTA Multiplication Example, Initial Value 1}
+\label{lst:OOTA Multiplication Example, Initial Value 1}
+\end{listing}
+
+In contrast, consider
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1},
+which differs from
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+only in the initial values (1 instead of 0) and the \co{exists}
+clause (0 instead of 1).
+Note that an execution satisfying the \co{exists} clause is again
+a (forbidden) OOTA cycle.
+
+Starting with the intuitive execution where the values loaded from \co{x}
+and \co{y} on lines~9 and~10 are the value 1 (that is, nonzero), executions
+that differ only in the value loaded from either \co{x} or \co{y} will
+still affect the value stored to \co{z} by line~11.
+There is therefore one semantic dependency from the load from \co{x}
+to the store to \co{z} and another separate semantic dependency from
+the load from \co{y} to the store to \co{z}, which matches typical
+definitions of semantic dependency.
+Yet
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+and
+Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
+can be thought of as being merely two different executions of the same
+program.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-mult3-0.litmus @@
+\caption{OOTA Three-Factor Multiplication Example, Initial Value Zero}
+\label{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
+\end{listing}
+
+Because the only two solutions to $x^2 = x$ are the values zero and one,
+Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
+and~\ref{lst:OOTA Multiplication Example, Initial Value 1}
+each have but one OOTA cycle, and an execution having nonzero non-one
+initial values could have at most two OOTA cycles, all of which are
+(again) forbidden by the C++ language.
+Listing~\ref{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
+addressses this limitation by multiplying the values obtained from three
+different loads.
+Because $x^3 = x$ has three solutions ($\left\{ -1, 0, 1 \right\}$),
+this execution could have up to three (yet again, forbidden) OOTA cycles.
+
+Much more elaborate examples can be constructed, for example,
+involving sums of products of values loaded.
+Reverting back to short-form code fragments, consider the following:
+
+\begin{quote}
+\begin{verbatim}
+z = w * x + w * y + w;
+\end{verbatim}
+\end{quote}
+
+This has the following semantic dependencies when all of \co{w},
+\co{x}, and \co{y} are zero:
+
+\begin{enumerate}
+\item	From the combination of \co{w} and \co{x} to \co{z}.
+\item	From the combination of \co{w} and \co{y} to \co{z}.
+\item	From \co{w} to \co{z}.
+\end{enumerate}
+
+Despite appearances, these are not redundant.
+The last semantic dependency does not imply the first, because given
+that \co{w} becomes nonzero, a change in \co{x} further changes
+the value stored to \co{z}.
+Similar reasoning shows that the last semantic dependency does not
+imply the second.
+Much more elaborate examples can easily be constructed.
+
+This raises the question of how to determine the set of semantic
+dependencies in a given execution of a given fragment of code.
+The short answer is that when running on a real-world system, 
+any C++ program is a finite-state machine, which, unlike Turing-complete
+systems, can be analyzed, as demonstrated in
+Section~\ref{sec:Hardware Dependencies, Instruction Ordering, and the Fundamental Property}.
+
+One might ask why semantic dependencies can be nontrivial when hardware
+address, control, and data dependencies are much more straightforward.
+Answers include:
+(1)~Respecting dependencies even when they reduce to a constant,
+(2)~Less-aggressive optimizations, and
+(3)~Lack of undefined behavior.
+
+This situation further underscores the per-execution nature of semantic
+dependencies as well as the need for an improved definition of ``semantic
+dependency''.
+In
+Section~\ref{sec:A Definition of Semantic Dependency},
+we expand the traditional definition to include dependencies extending
+from groups of loads to a single store.
 
@@
 
-\subsection{Non-Trivial Semantic Dependencies}
-\label{app:Non-Trivial Semantic Dependencies}
-
-Most of the examples in this paper involve simple semantic dependencies
-connecting a single load to a single store.
-One exception appears in
-Appendix~\ref{app:Non-LB Litmus Tests},
-but this section will present a more difficult example.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult-0.litmus @@
-\caption{OOTA Multiplication Example, Initial Value Zero}
-\label{lst:OOTA Multiplication Example, Initial Value Zero}
-\end{listing}
-
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-shows an example that builds on the multiplication-by-zero
-discussion in
-Section~\ref{sec:Semantic Dependencies and Source Code}.
-Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
-\co{x} and \co{y} on lines~9 and~10, respectively.
-Similarly, lines~16 and~17 store to \co{x} and \co{y}, respectively,
-the value that line~15 loads from \co{z}.
-Because all three variables are initialized to zero, and because loads
-can be ordered before any store, there is an execution in which all
-three loads return the value zero, in which case all three stores will
-of course store the value zero.
-In any related execution where the value loaded from \co{y} remains zero,
-nonzero values loaded from \co{x} cannot affect the value stored by
-line~11 to \co{z}.
-Only executions which load different (that is, nonzero) values from
-both \co{x} and \co{y} can cause the value stored to \co{z} to change
-from zero to a nonzero value.
-
-Therefore, as noted in
-Section~\ref{sec:Semantic Dependencies and Source Code},
-in an execution where the loads from both \co{x} and \co{y} return zero,
-there is no semantic dependency from the load from \co{x} to the
-store to \co{z}, nor is there a semantic dependency from the load
-from \co{y} to the store to \co{z}.
-However, there \emph{is} a semantic dependency from the \emph{combination}
-of the loads from \co{x} and \co{y} to the store to \co{z}.
-For example, the (forbidden) execution that satisfies the \co{exists}
-clause, which would be an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult-0-cond.litmus @@
-\caption{OOTA Conditional Multiplication Example, Initial Value Zero}
-\label{lst:OOTA Conditional Multiplication Example, Initial Value Zero}
-\end{listing}
-
-For further evidence that there is no semantic dependency from \co{y}
-to \co{z} in executions where the load from \co{x} is zero, see the
-equivalent\footnote{
-	But equivalent only because the atomics are all nonvolatile!}
-program shown in
-Listing~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero}.
-Here, there is not even a load from \co{y} in executions where the
-load from \co{x} returns zero, so there cannot possibly be a semantic
-dependency involving this non-existent load.
-A similar transformation would load \co{x} only in executions where the
-load from \co{y} returned nonzero, which similarly emphasizes the lack
-of any semantic dependency between \co{x} and \co{z}.
-
-In both
-Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-and~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero},
-an OOTA cycle (such as the one indicated by their \co{exists} clauses)
-would need to affect both the \co{x} and the \co{y} components of this
-compound semantic dependency.
-This situation poses a challenge to current work on semantic dependencies,
-which typically restricts such dependencies to a single load and store.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult-1.litmus @@
-\caption{OOTA Multiplication Example, Initial Value 1}
-\label{lst:OOTA Multiplication Example, Initial Value 1}
-\end{listing}
-
-In contrast, consider
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1},
-which differs from
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-only in the initial values (1 instead of 0) and the \co{exists}
-clause (0 instead of 1).
-Note that an execution satisfying the \co{exists} clause is again
-a (forbidden) OOTA cycle.
-
-Starting with the intuitive execution where the values loaded from \co{x}
-and \co{y} on lines~9 and~10 are the value 1 (that is, nonzero), executions
-that differ only in the value loaded from either \co{x} or \co{y} will
-still affect the value stored to \co{z} by line~11.
-There is therefore one semantic dependency from the load from \co{x}
-to the store to \co{z} and another separate semantic dependency from
-the load from \co{y} to the store to \co{z}, which matches typical
-definitions of semantic dependency.
-Yet
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-and
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
-can be thought of as being merely two different executions of the same
-program.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult3-0.litmus @@
-\caption{OOTA Three-Factor Multiplication Example, Initial Value Zero}
-\label{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
-\end{listing}
-
-Because the only two solutions to $x^2 = x$ are the values zero and one,
-Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-and~\ref{lst:OOTA Multiplication Example, Initial Value 1}
-each have but one OOTA cycle, and an execution having nonzero non-one
-initial values could have at most two OOTA cycles, all of which are
-(again) forbidden by the C++ language.
-Listing~\ref{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
-addressses this limitation by multiplying the values obtained from three
-different loads.
-Because $x^3 = x$ has three solutions ($\left\{ -1, 0, 1 \right\}$),
-this execution could have up to three (yet again, forbidden) OOTA cycles.
-
-Much more elaborate examples can be constructed, for example,
-involving sums of products of values loaded.
-Reverting back to short-form code fragments, consider the following:
-
-\begin{quote}
-\begin{verbatim}
-z = w * x + w * y + w;
-\end{verbatim}
-\end{quote}
-
-This has the following semantic dependencies when all of \co{w},
-\co{x}, and \co{y} are zero:
-
-\begin{enumerate}
-\item	From the combination of \co{w} and \co{x} to \co{z}.
-\item	From the combination of \co{w} and \co{y} to \co{z}.
-\item	From \co{w} to \co{z}.
-\end{enumerate}
-
-Despite appearances, these are not redundant.
-The last semantic dependency does not imply the first, because given
-that \co{w} becomes nonzero, a change in \co{x} further changes
-the value stored to \co{z}.
-Similar reasoning shows that the last semantic dependency does not
-imply the second.
-Much more elaborate examples can easily be constructed.
-
-This raises the question of how to determine the set of semantic
-dependencies in a given execution of a given fragment of code.
-The short answer is that when running on a real-world system, 
-any C++ program is a finite-state machine, which, unlike Turing-complete
-systems, can be analyzed, as demonstrated in
-Section~\ref{sec:Hardware Dependencies, Instruction Ordering, and the Fundamental Property}.
-
-One might ask why semantic dependencies can be nontrivial when hardware
-address, control, and data dependencies are much more straightforward.
-Answers include:
-(1)~Respecting dependencies even when they reduce to a constant,
-(2)~Less-aggressive optimizations, and
-(3)~Lack of undefined behavior.
-
-This situation further underscores the per-execution nature of semantic
-dependencies as well as the need for an improved definition of ``semantic
-dependency''.
-In
-Section~\ref{sec:A Definition of Semantic Dependency},
-we expand the traditional definition to include dependencies extending
-from groups of loads to a single store.
-
 \subsection{OOTA From Undefined-Behavior Optimizations}
@@
 
-\subsection{Semantic Dependencies and \co{volatile}}
-\label{app:Semantic Dependencies and volatile}
-
-The Linux kernel uses volatile accesses to constrain the compiler,
-and it is worth looking at the example shown in
-Listing~\ref{lst:Non-Volatile Accesses and Dependencies}
-to see how this works.
-
-\begin{listing}
-\begin{verbatim}
- 1 atomic<int> x, y;
- 2
- 3 void thread1() {
- 4   int r1 = x.load(memory_order_relaxed);
- 5   y.store(r1, memory_order_relaxed);
- 6 }
-\end{verbatim}
-\caption{Non-Volatile Accesses and Dependencies}
-\label{lst:Non-Volatile Accesses and Dependencies}
-\end{listing}
-
-Because both \co{x} and \co{y} are nonvolatile, the compiler is
-free to assume that the value loaded by line~4 will be some value
-stored to \co{x} or its initial value.
-Because there are no stores to \co{x}, the only remaining possibility
-is the default-initialized value of zero.
-Therefore, the C++ implementation is within its rights to substitute
-the constant zero for \co{r1} on line~5, eliminating the semantic
-dependency that would otherwise extend from line~4 to line~5.
-This in turn enables the implementation to reorder line~5 to
-precede line~4.
-
-\begin{listing}
-\begin{verbatim}
- 1 atomic<volatile int> x, y;
- 2
- 3 void thread1() {
- 4   int r1 = x.load(memory_order_relaxed);
- 5   y.store(r1, memory_order_relaxed);
- 6 }
-\end{verbatim}
-\caption{Volatile Accesses and Dependencies}
-\label{lst:Volatile Accesses and Dependencies}
-\end{listing}
-
-One way to preserve this dependency is use of \co{volatile}, as shown in
-Listing~\ref{lst:Volatile Accesses and Dependencies}.
-Now the implementation is forbidden from assuming that the value loaded
-by line~4 has any relation to any stores to or initialization of \co{x},
-even if \co{x} is placed in normal memory.
-
-In this case, the \co{volatile} keyword alerts the implementation to
-the possibility of interfering accesses to \co{x} from outside the
-program, for example, due to \co{x} being:
-
-\begin{itemize}
-\item	Allocated in an MMIO region of the address space.
-\item	Modified by an unknown-to-the-implementation thread
-	or signal handler.
-\item	Subject to I/O-device DMA operations.
-\item	Modified by debugger commands.
-\item	Modified by accesses to facilities such as \path{/dev/mem}.
-\item	Modified by as-yet-unwritten dynamically linked libraries.
-\item	Modified by some other mechanism of the reader's choosing.
-\end{itemize}
-
-These possibilities force the implementation to preserve the semantic
-dependency from line~4 to line~5.
-They can also prevent the formation of OOTA cycles local to the
-implementation because such inteference means that the volatile loads
-might be reading from something that the implementation knows nothing
-about.
-
-The developer is free to arrange for \co{x} to be free of any
-interference, thus obtaining predictable behavior along with the
-preserved semantic dependency.
-However, this example invalidates the rough definition of semantic
-dependency from
-Section~\ref{sec:Simple OOTA Cycle}
-because the value loaded from \co{x} will always be the initial value
-of zero, so that it is at best dubious to talk about any changes in
-the value loaded.
-
-One way forward is to assume that interference could happen when
-determining which dependencies are semantic, and then do further analysis
-using that determination.
+\subsection{Three-Process Version of JMM Causality Test Case 4}
+\label{app:Three-Process Version of JMM Causality Test Case 4}
+
+Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
+shows a different way of expanding
+Listing~\ref{lst:Causality Test Case 4}
+(discussed in
+Appendix~\ref{app:Causality Test Case 4})
+from two threads to three threads.
+The \co{herd7} tool finds OOTA values, as expected.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-3proc.litmus @@
+\caption{Three-Process Version of JMM Causality Test Case 4}
+\label{lst:Three-Process Version of JMM Causality Test Case 4}
+\end{listing}
+
+All threads have straightforward semantic dependencies from their
+relaxed atomic loads to their relaxed atomic stores, which prevents this
+OOTA cycle on compiler-based C++ implementations.
+
+\subsection{Non-LB Litmus Tests}
+\label{app:Non-LB Litmus Tests}
+
+Listing~\ref{lst:Non-LB Litmus Test}
+shows a litmus test that can produce an OOTA cycle, but that is not
+a member of the load-buffering (LB) class of litmus tests.
+Suppose that lines~9 and~10 both somehow return the value 21.
+Then line~12 will store the value 42 to \co{x}, which will cause
+line~17 to store 21 to \co{y} and line~22 to store 21 to \co{z},
+thus justifying the values initially loaded by lines~9 and~10.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-non-lb.litmus @@
+\caption{Non-LB Litmus Test}
+\label{lst:Non-LB Litmus Test}
+\end{listing}
+
+All thread's loads' return values affect the value stored, so all
+threads have semantic dependencies between their respective relaxed
+atomic loads and relaxed atomic stores.
+Therefore, an OOTA cycle cannot form in a compiler-based C++
+implementation.
+
+Note that in a C++ implementation that (erroneously) permitted an OOTA
+cycle to form, that cycle would involve both \co{r0} and \co{r1}.
+
+\subsection{Multi-Cycle Litmus Tests}
+\label{app:Multi-Cycle Litmus Tests}
+
+Listing~\ref{lst:Multi-Cycle Litmus Test}
+shows an expansion to four threads of Causality Test Case 4, which is shown in
+Listing~\ref{lst:Causality Test Case 4}
+and discussed in
+Appendix~\ref{app:Causality Test Case 4}.
+The \co{herd7} tool shows multiple different possible OOTA cycles,
+as expected.
+However, the value returned by each thread's load directly affects the
+value stored, so there is a semantic dependency in each thread.
+Therefore, none of these OOTA cycles can happen in a compiler-based
+C++ implementation.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-two-source.litmus @@
+\caption{Multi-Cycle Litmus Test}
+\label{lst:Multi-Cycle Litmus Test}
+\end{listing}
+
+\subsection{Simple Reordering as Litmus Test}
+\label{app:Simple Reordering as Litmus Test}
+
+Listing~\ref{lst:Simple Reordering as Litmus Test}
+shows how
+Listing~\ref{lst:Simple Reordering}
+can be recast into a \co{herd7} litmus test.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/simple-reordering.litmus @@
+\caption{Simple Reordering as Litmus Test}
+\label{lst:Simple Reordering as Litmus Test}
+\end{listing}
 

### 2024-03-05 ootavacuous.tex
Fallout from reversing Listings 1 and 2
Now that Listings 1 and 2 have been swapped, the references to them
should be swapped as well.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 The earlier examples shown in
-Listings~\ref{lst:Simple Reordering}
-and~\ref{lst:Simple OOTA}
+Listings~\ref{lst:Simple OOTA}
+and~\ref{lst:Simple Reordering}
 can be recast as similar \co{herd7} litmus tests, as shown in
-Listing~\ref{lst:Simple Reordering as Litmus Test}
-on page~\pageref{lst:Simple Reordering as Litmus Test}
-and Listing~\ref{lst:Causality Test Case 4}
-on page~\pageref{lst:Causality Test Case 4}, respectively.
+Listing~\ref{lst:Causality Test Case 4}
+on page~\pageref{lst:Causality Test Case 4}
+and Listing~\ref{lst:Simple Reordering as Litmus Test}
+on page~\pageref{lst:Simple Reordering as Litmus Test}, respectively.
 Other OOTA-related litmus tests may be found in

### 2024-03-05 ootavacuous.tex
Add disclaimer about rfe "ordering"
The rfe relation doesn't _cause_ ordering of hardware instructions; it's
more like a _result_ of the order.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Listing 10: s/atomic<volatile int>/volatile atomic<int>
Make it match the similar use cases in the standard.

Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 from.
+(To be fair, this should be considered more as a \emph{result} of
+ordering rather than a \emph{cause} of ordering, in that
+if a load commits before a store on another thread then it
+cannot possibly read from that store.)
 
@@
 \begin{verbatim}
- 1 atomic<volatile int> x, y;
+ 1 volatile atomic<int> x, y;
  2

### 2024-03-05 ootavacuous.tex
volatile: Remove reordering and normal-memory wording
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 dependency that would otherwise extend from line~4 to line~5.
-This in turn enables the implementation to reorder line~5 to
-precede line~4.
 
@@
 Now the implementation is forbidden from assuming that the value loaded
-by line~4 has any relation to any stores to or initialization of \co{x},
-even if \co{x} is placed in normal memory.
+by line~4 has any relation to any stores to or initialization of \co{x}.
 

### 2024-03-05 ootavacuous.tex
volatile: Clarify preserved dependencies and prevented OOTA cycles
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 dependency from line~4 to line~5.
-They can also prevent the formation of OOTA cycles local to the
-implementation because such inteference means that the volatile loads
-might be reading from something that the implementation knows nothing
-about.
+This preserved dependency helps prevent the formation of OOTA cycles
+even in situations where there is no interference.\footnote{
+	When there is interference, the definition of ``OOTA cycle''
+	does not apply.}
 

### 2024-03-05 ootavacuous.tex
volatile: Refer back to Section 7.4.2
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
"Illustrative Litmus Tests": s/implementation/compiler/
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 is the default-initialized value of zero.
-Therefore, the C++ implementation is within its rights to substitute
+Therefore, the C++ compiler is within its rights to substitute
 the constant zero for \co{r1} on line~5, eliminating the semantic
@@
 Listing~\ref{lst:Volatile Accesses and Dependencies}.
-Now the implementation is forbidden from assuming that the value loaded
+Now the compiler is forbidden from assuming that the value loaded
 by line~4 has any relation to any stores to or initialization of \co{x}.
 
-In this case, the \co{volatile} keyword alerts the implementation to
+In this case, the \co{volatile} keyword alerts the compiler to
 the possibility of interfering accesses to \co{x} from outside the
@@
 \item	Allocated in an MMIO region of the address space.
-\item	Modified by an unknown-to-the-implementation thread
+\item	Modified by an unknown-to-the-compiler thread
 	or signal handler.
@@
 
-These possibilities force the implementation to preserve the semantic
+These possibilities force the compiler to preserve the semantic
 dependency from line~4 to line~5.
@@
 using that determination.
+This forms the basis the quasi-volatile approach presented in
+Section~\ref{sec:Global Analysis and Volatile versus Quasi Volatile}.
 
@@
 demonstrate that rf can be atemporal, they demonstrate neither
-OOTA cycles nor OOTA-like behavior in full C++ implementations.
+OOTA cycles nor OOTA-like behavior in full C++ compiler.
 This appendix provides more involved examples that can be optimized
@@
 the store to and load from \co{y}.
-This means that a C++ implementation could prove that the value 17 is
+This means that a C++ compiler could prove that the value 17 is
 always stored to \co{z} by examining \co{P0()} in isolation, given that
-it is also legal for a C++ implementation to replace the value loaded
+it is also legal for a C++ compiler to replace the value loaded
 from \co{y} with the value just stored.\footnote{
@@
 
-In a real-world implementation that proves that the value 17 is always
-stored to \co{z}, there is no semantic dependency to
-\co{P0()}'s store to \co{z}, but there is to all of the other relaxed
+Given a compiler that proves that the value 17 is always
+stored to \co{z}, there might be no semantic dependency to
+\co{P0()}'s store to \co{z}, but must be to all of the other relaxed
 atomic stores.
-On the other hand, if the implementation does not prove the value
+On the other hand, if the compiler does not prove the value
 stored to \co{z}, there will be a global-time delay between each of
@@
 C++ implementation.
-Yes, an implementation that proves that the value 17 is always
+Yes, a compielr that proves that the value 17 is always
 stored to \co{z} can satisfy the \co{exists} clause, but because
@@
 But is \co{P0()}'s dependency a semantic dependency?
-For example, is a real-world C++ implementation permitted to
+For example, is a C++ compiler permitted to
 duplicate \co{P0()}'s load from \co{x} and then make use of both loads'
 values?
-If this is permitted, then the implementation would be within its rights
+If this is permitted, then the compiler would be within its rights
 to transform
@@
 differ, then one of those two values must be equal to that from \co{y}.
-If the implementation is further permitted to make a runtime choice
+If the compiler is further permitted to make a runtime choice
 between the two values loaded from \co{x}, it could always choose the
@@
 
-The implementation might also drop the load from \co{y}, whose value
+The compiler might also drop the load from \co{y}, whose value
 is now unused except by the \co{exists} clause.
@@
 One possible reaction to this situation is to declare that useful
-real-world C++ implementations should never duplicate nonvolatile
+C++ compilers should never duplicate nonvolatile
 relaxed atomic loads.
@@
 But suppose that we drop the single-thread-analysis constraint for
-the moment, and suppose that the C++ implementation is able to prove
+the moment, and suppose that the compiler is able to prove
 that the only modification of a given atomic variable is to atomically
@@
 
-Is a C++ implementation within its rights to behave as if the code
+Is a C++ compiler within its rights to behave as if the code
 shown in
@@
 	adjacent to each other.
-\item	Any time an implementation might be tempted to invent uses from
+\item	Any time a compiler might be tempted to invent uses from
 	values from nonvolatile relaxed atomic loads from the same
@@
 is customary, for example, when hoisting a load out of a loop.
-In this case, implementations might hoist the load from x out of
+In this case, compilers might hoist the load from x out of
 the loop:
@@
 
-A C++ implementation might act as if the program had instead been this:
+A compiler might act as if the program had instead been this:
 
@@
 Note that passing \co{x} to the call to \co{do_something_with()} is important
-because this allows the C++ implementation to assume that accesses to \co{x}
+because this allows the compiler to assume that accesses to \co{x}
 will not result in a data race.
@@
 data races.
-In theory, the implementation could make the above transformation without
+In theory, the compiler could make the above transformation without
 worry, especially given that the value loaded from \co{x} is discarded in
@@
 After all, optimizations are supposed to speed things up.
-This means that a C++ implementation might be wise to adhere to the
+This means that a compiler might be wise to adhere to the
 same restriction that avoids data races for loads from non-atomic
@@
 
-Many C++ implementations carry out optimizations based on the assumption
+Many compilers carry out optimizations based on the assumption
 that UB cannot happen, and these optimizations can result in OOTA cycles,
@@
 If it is the value one, then there is a divide-by-zero error, which
-enables the C++ implementation to assume that the value of \co{r1} is
+enables the compiler to assume that the value of \co{r1} is
 always zero.
-This in turn allows the implementation to behave as if the expression
+This in turn allows the compiler to behave as if the expression
 stored to \co{y} on line~14 was the constant 1, at which point there
@@
 
-The implementation is then within its rights to act as if lines~13 and~14
+The compiler is then within its rights to act as if lines~13 and~14
 were interchanged.
@@
 that prevent back-propagation of UB, but these are ineffective in this
-case because the C++ implementation can detect and exploit UB when
+case because the compiler can detect and exploit UB when
 considering line~14 in isolation.
@@
 having no cycle at all because of the lack of an sdep link in \co{P0()},
-and other implementations prevented by hardware
+and other compilers prevented by hardware
 dependency ordering from realizing the cycle.

### 2024-03-05 ootavacuous.tex
Abstract: Further rewording of second-to-last sentence
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	We show that these models' constraints prevent OOTA cycles
-	that involve only volatile atomics from occurring in
-	undefined-behavior-free C++ programs (under these implementations).
+	from occurring in undefined-behavior-free C++ programs running
+	on compiler-based implementations, provided the cycles involve
+	only volatile atomics.
 	We also extend this work to nonvolatile atomics by defining

### 2024-03-05 ootavacuous.tex
"Non-Trivial Semantic Dependencies": Wordsmith
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 of the loads from \co{x} and \co{y} to the store to \co{z}.
-For example, the (forbidden) execution that satisfies the \co{exists}
-clause, which would be an OOTA cycle.
+An example is given by the (forbidden) execution that satisfies the
+\co{exists} clause on line~20, which would be an OOTA cycle.
 
@@
 A similar transformation would load \co{x} only in executions where the
-load from \co{y} returned nonzero, which similarly emphasizes the lack
-of any semantic dependency between \co{x} and \co{z}.
+load from \co{y} returned nonzero, which similarly illustrates the lack
+of a semantic dependency between \co{x} and \co{z}.
 
@@
 and~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero},
-an OOTA cycle (such as the one indicated by their \co{exists} clauses)
+an OOTA cycle (such as those indicated by their respective \co{exists} clauses)
 would need to affect both the \co{x} and the \co{y} components of this
@@
 This situation poses a challenge to current work on semantic dependencies,
-which typically restricts such dependencies to a single load and store.
+which are typically restricted to a single load and store.
 
@@
 Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
-can be thought of as being merely two different executions of the same
+can be thought of as being two different executions of the exact same
 program.
@@
 and~\ref{lst:OOTA Multiplication Example, Initial Value 1}
-each have but one OOTA cycle, and an execution having nonzero non-one
-initial values could have at most two OOTA cycles, all of which are
+each have only one OOTA cycle, and an execution having initial values
+of (say) two could have at most two OOTA cycles, both of which are
 (again) forbidden by the C++ language.
 Listing~\ref{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
-addressses this limitation by multiplying the values obtained from three
+goes one step further by multiplying the values obtained from three
 different loads.
@@
 address, control, and data dependencies are much more straightforward.
-Answers include:
-(1)~Respecting dependencies even when they reduce to a constant,
-(2)~Less-aggressive optimizations, and
-(3)~Lack of undefined behavior.
+The reason is that, unlike compilers, hardware:
+(1)~Respects dependencies even when they reduce to a constant,
+(2)~Less-aggressively optimizes, and
+(3)~Lacks undefined behavior.
 

### 2024-03-05 ootavacuous.tex
rf-vs-rfe: Reword
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 demonstrate that rf can be atemporal, they demonstrate neither
-OOTA cycles nor OOTA-like behavior in full C++ compiler.
+OOTA cycles nor OOTA-like behavior in full C++ compilers.
 This appendix provides more involved examples that can be optimized
@@
 
-Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
+Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
 revises
-Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
+Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
 so that the value stored to \co{y} is forced to be at most the
@@
 C++ implementation.
-Yes, a compielr that proves that the value 17 is always
+Yes, a compiler that proves that the value 17 is always
 stored to \co{z} can satisfy the \co{exists} clause, but because
@@
 of (sdep $\cup$ rfe), or in \co{herd7} nomenclature,
-\co{acyclic(sdep; rfe)}.
+\co{acyclic(sdep; rfe)}, as discussed in
+Section~\ref{sec:OOTA: rf vs. rfe}.
 

### 2024-03-05 ootavacuous.tex
"Inventing Atomic Loads": Reword
Also stomp out a couple more "implementations" that hid previous
efforts by virtue of being capitalized.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 and \co{P1()}'s store to \co{y} clearly depends on its load from \co{z}.
-If both dependencies are semantic dependency, the OOTA cycle cannot
+If both dependencies are semantic dependencies, the OOTA cycle cannot
 form.
@@
 	are futile.
-\item	Implementations should not invent uses for values from
+\item	Compilers should not invent uses for values from
 	nonvolatile relaxed atomic loads.
@@
 	spaced loads return different values on the other.
-\item	Implementations that invent nonvolatile atomic loads or invent
+\item	Compilers that invent nonvolatile atomic loads or invent
 	new uses for nonvolatile atomic loads can adversely affect
@@
 Note that passing \co{x} to the call to \co{do_something_with()} is important
-because this allows the compiler to assume that accesses to \co{x}
-will not result in a data race.
+because this allows the compiler to assume that inventing additional
+accesses to \co{x} will not result in a data race.
 Without this assumption, the above transformation might introduce a data

### 2024-03-05 ootavacuous.tex
OOTA from UB: Wordsmith
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}
-were converted to be volatile.
+are converted to volatile.
 

### 2024-03-05 ootavacuous.tex
Three-Process OOTA: Wordsmith.
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
-shows a different way of expanding
+expands
 Listing~\ref{lst:Causality Test Case 4}

### 2024-03-05 ootavacuous.tex
Move multi-cycle litmus test after three-process test
The two litmus tests are closely related, so move their sections
together.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
"Multi-Cycle Litmus Test": Reword
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+\subsection{Multi-Cycle Litmus Test}
+\label{app:Multi-Cycle Litmus Test}
+
+Listing~\ref{lst:Multi-Cycle Litmus Test}
+expands
+Listing~\ref{lst:Causality Test Case 4}
+in a different way, moving to four threads but having only the
+original two variables \co{x} and \co{y}.
+The \co{herd7} tool shows multiple different possible OOTA cycles,
+as expected.
+However, the value returned by each thread's load directly affects the
+value stored, so there is a semantic dependency in each thread.
+Therefore, none of these OOTA cycles can arise in a compiler-based
+C++ implementation.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-two-source.litmus @@
+\caption{Multi-Cycle Litmus Test}
+\label{lst:Multi-Cycle Litmus Test}
+\end{listing}
+
 \subsection{Non-LB Litmus Tests}
@@
 cycle to form, that cycle would involve both \co{r0} and \co{r1}.
-
-\subsection{Multi-Cycle Litmus Tests}
-\label{app:Multi-Cycle Litmus Tests}
-
-Listing~\ref{lst:Multi-Cycle Litmus Test}
-shows an expansion to four threads of Causality Test Case 4, which is shown in
-Listing~\ref{lst:Causality Test Case 4}
-and discussed in
-Appendix~\ref{app:Causality Test Case 4}.
-The \co{herd7} tool shows multiple different possible OOTA cycles,
-as expected.
-However, the value returned by each thread's load directly affects the
-value stored, so there is a semantic dependency in each thread.
-Therefore, none of these OOTA cycles can happen in a compiler-based
-C++ implementation.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-two-source.litmus @@
-\caption{Multi-Cycle Litmus Test}
-\label{lst:Multi-Cycle Litmus Test}
-\end{listing}
 

### 2024-03-05 ootavacuous.tex
"Non-LB Litmus Test": Reword
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 One exception appears in
-Appendix~\ref{app:Non-LB Litmus Tests},
+Appendix~\ref{app:Non-LB Litmus Test},
 but this section will present a more difficult example.
@@
 
-\subsection{Non-LB Litmus Tests}
-\label{app:Non-LB Litmus Tests}
+\subsection{Non-LB Litmus Test}
+\label{app:Non-LB Litmus Test}
 
@@
 Note that in a C++ implementation that (erroneously) permitted an OOTA
-cycle to form, that cycle would involve both \co{r0} and \co{r1}.
+cycle to form, that cycle would involve both of \co{P0()}'s \co{r0}
+and \co{r1} local variables.
 

### 2024-03-06 ootavacuous.tex
Combine Introduction and Background
Aside from its first sentence, the Introduction section contains
essentially no information that isn't already present in the table of
contents.  While "Tell them what you're going to tell them, then tell
them, then tell them what you've told them" might be a good strategy for
an oral presentation where listeners can't easily refer back to earlier
portions, it's wasteful and distracting for written work.

Therefore I removed all the redundant material from the Introduction,
and since only one sentence remained, I decided to combine it with the
following Back

@@
 
-\section{Introduction}
+\section{Introduction and Background}
 \label{sec:Introduction}
@@
 This paper shows that compiler-based C++ implementations subject to
-reasonable real-world constraints cannot exhibit OOTA cycles.
-Section~\ref{sec:Background}
-provides background on the problem of OOTA cycles, including a
-brief summary of prior work.
-Section~\ref{sec:OOTA and Semantic Dependencies}
-motivates a shift in definition of OOTA cycle and reviews properties of
-semantic dependencies, showing that these properties are a function of
-executions rather than simply of the source code.
-This leads into
-Section~\ref{sec:What is an Execution?},
-which describes properties of abstract and hardware executions, along
-with the relationship between the two.
-
-This in turn leads to
-Section~\ref{sec:C++ Compilers}, which describes properties of the tool
-commonly used to create hardware executions with the proper relationship
-to a given set of abstract executions, namely  the compiler.
-Section~\ref{sec:Hardware Dependencies, Instruction Ordering, and the
-Fundamental Property}
-then combines hardware dependencies and instruction ordering to arrive
-at the fundamental property of semantic dependencies, which
-Section~\ref{sec:A Definition of Semantic Dependency} uses to
-arrive at a definition of the term ``semantic dependency''.
-
-This sets the stage for
-Section~\ref{sec:Real-World Constraints}
-to bring real-world constraints to bear, starting with the laws of
-physics, informally demonstrating that these constraints prevent
-OOTA cycles containing only volatile and quasi-volatile atomics from
-occurring in undefined-behavior-free C++ programs running on correctly
-constructed C++ compiler-based implementations that are restricted to
-single-threaded analysis in our current universe.
-
-Finally,
-Section~\ref{sec:Future Directions}
-lists a few possible future directions for this work and
-Section~\ref{sec:Conclusion}
-presents concluding remarks.
-
-\section{Background}
-\label{sec:Background}
-
-This section provides a brief overview of the OOTA problem, followed
+reasonable constraints on how they treat accesses to atomic objects
+cannot exhibit out-of-thin-air (OOTA) cycles.
+It follows that these implementations hardly need to take any special
+action to avoid OOTA.
+In fact, for many compilers the constraints boil down to a simple
+``Don't invent or duplicate atomic accesses'', which is probably
+something the compiler wouldn't do anyway.
+
+We begin with a brief overview of the OOTA problem, followed
 by an equally brief summary of prior work in this area, and ending

### 2024-03-06 ootavacuous.tex
"Non-Trivial Semantic Dependencies": Add "for", drop "quasi-"
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Make "Illustrative Litmus Tests" less forbidding
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 using that determination.
-This forms the basis the quasi-volatile approach presented in
+This forms the basis for the volatile approach presented in
 Section~\ref{sec:Global Analysis and Volatile versus Quasi Volatile}.
@@
 of the loads from \co{x} and \co{y} to the store to \co{z}.
-An example is given by the (forbidden) execution that satisfies the
+An example is given by the execution that satisfies the
 \co{exists} clause on line~20, which would be an OOTA cycle.
@@
 Note that an execution satisfying the \co{exists} clause is again
-a (forbidden) OOTA cycle.
+a OOTA cycle.
 
@@
 each have only one OOTA cycle, and an execution having initial values
-of (say) two could have at most two OOTA cycles, both of which are
-(again) forbidden by the C++ language.
+of (say) two could have at most two OOTA cycles.
 Listing~\ref{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
@@
 Because $x^3 = x$ has three solutions ($\left\{ -1, 0, 1 \right\}$),
-this execution could have up to three (yet again, forbidden) OOTA cycles.
+this execution could have up to three OOTA cycles.
 

### 2024-03-06 ootavacuous.tex
"Non-Trivial Semantic Dependencies": Distinguish init from input
Initial values are part of the program, so talk instead about the
alternative case where those values were taken as input by a single
program.

Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 definitions of semantic dependency.
-Yet
+Yet if the initial values were instead obtained as input,
 Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
@@
 Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
-can be thought of as being two different executions of the exact same
+could be thought of as being two different executions of the exact same
 program.

### 2024-03-06 ootavacuous.tex
"Non-Trivial Semantic Dependencies": Add initial-value/OOTA footnote
Examining an execution can identify OOTA cycles whose values happen to
match the initial values, even though software could not distinguish them.

Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 and~\ref{lst:OOTA Multiplication Example, Initial Value 1}
-each have only one OOTA cycle, and an execution having initial values
+each have only one OOTA cycle,\footnote{
+	Note well that the initial values cannot be said to have come
+	out of thin air.
+	Although it is possible to have an OOTA cycle that does not
+	involve the initial values, but happens to have the same value,
+	the software would not be able to tell the difference.}
+and an execution having initial values
 of (say) two could have at most two OOTA cycles.

### 2024-03-06 ootavacuous.tex
"Non-Trivial Semantic Dependencies": Delete ornate example
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Much more elaborate examples can be constructed, for example,
-involving sums of products of values loaded.
-Reverting back to short-form code fragments, consider the following:
-
-\begin{quote}
-\begin{verbatim}
-z = w * x + w * y + w;
-\end{verbatim}
-\end{quote}
-
-This has the following semantic dependencies when all of \co{w},
-\co{x}, and \co{y} are zero:
-
-\begin{enumerate}
-\item	From the combination of \co{w} and \co{x} to \co{z}.
-\item	From the combination of \co{w} and \co{y} to \co{z}.
-\item	From \co{w} to \co{z}.
-\end{enumerate}
-
-Despite appearances, these are not redundant.
-The last semantic dependency does not imply the first, because given
-that \co{w} becomes nonzero, a change in \co{x} further changes
-the value stored to \co{z}.
-Similar reasoning shows that the last semantic dependency does not
-imply the second.
-Much more elaborate examples can easily be constructed.
-
-This raises the question of how to determine the set of semantic
+Much more elaborate examples can be constructed,
+which raises the question of how to determine the set of semantic
 dependencies in a given execution of a given fragment of code.

### 2024-03-06 ootavacuous.tex
"Non-Trivial Semantic Dependencies": Fix HW/compiler comparison
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-One might ask why semantic dependencies can be nontrivial when hardware
+One might ask why semantic dependencies can be complicated when hardware
 address, control, and data dependencies are much more straightforward.
@@
 (1)~Respects dependencies even when they reduce to a constant,
-(2)~Less-aggressively optimizes, and
+(2)~Optimizes much less aggressively, if at all, and
 (3)~Lacks undefined behavior.

### 2024-03-06 ootavacuous.tex
Small changes for D.1 and D.2
Here is a set of small edits to sections D.1 and D.2.  They are mostly
just stylistic improvements.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-03-06 ootavacuous.tex
@@
 
-Because both \co{x} and \co{y} are nonvolatile, the compiler is
-free to assume that the value loaded by line~4 will be some value
+Because both \co{x} and \co{y} are nonvolatile, a C++ compiler is
+free to assume that the value loaded by line~4 will be either some value
 stored to \co{x} or its initial value.
-Because there are no stores to \co{x}, the only remaining possibility
+And because there are no stores to \co{x}, the only remaining possibility
 is the default-initialized value of zero.
-Therefore, the C++ compiler is within its rights to substitute
+Therefore, the compiler is within its rights to substitute
 the constant zero for \co{r1} on line~5, eliminating the semantic
@@
 In this case, the \co{volatile} keyword alerts the compiler to
-the possibility of interfering accesses to \co{x} from outside the
+the possibility of interfering changes to \co{x} from outside the
 program, for example, due to \co{x} being:
-
 \begin{itemize}
@@
 \end{itemize}
-
 These possibilities force the compiler to preserve the semantic
-dependency from line~4 to line~5.
-This preserved dependency helps prevent the formation of OOTA cycles
-even in situations where there is no interference.\footnote{
+dependency from line~4 to line~5,
+and this preserved dependency helps prevent the formation of OOTA cycles
+in environments where there is no interference.\footnote{
 	When there is interference, the definition of ``OOTA cycle''
@@
 that differ only in the value loaded from either \co{x} or \co{y} will
-still affect the value stored to \co{z} by line~11.
+now affect the value stored to \co{z} by line~11.
 There is therefore one semantic dependency from the load from \co{x}
@@
 
-One might ask why semantic dependencies can be nontrivial when hardware
-address, control, and data dependencies are much more straightforward.
+One might ask why semantic dependencies can be so complicated when hardware
+dependencies are much more straightforward.
 The reason is that, unlike compilers, hardware:
-(1)~Respects dependencies even when they reduce to a constant,
-(2)~Less-aggressively optimizes, and
-(3)~Lacks undefined behavior.
-
+\begin{itemize}
+\item	Respects dependencies even when they reduce to a constant,
+\item	Optimizes much less aggressively, if at all, and
+\item	Lacks undefined behavior.
+\end{itemize}
 This situation further underscores the per-execution nature of semantic

### 2024-03-07 ootavacuous.tex
"Non-Trivial Semantic Dependencies": Allow initial-value OOTA
If the initial values happen to match those of an OOTA cycle, then
executions yielding those values might or might not be an OOTA
cycle, depending on whether or not the cycle's loads take values
only from stores within that same cycle.

Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 and~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero},
-an OOTA cycle (such as those indicated by their respective \co{exists} clauses)
-would need to affect both the \co{x} and the \co{y} components of this
-compound semantic dependency.
+an OOTA cycle having non-zero values (such as those indicated by their
+respective \co{exists} clauses) would need to affect both the \co{x}
+and the \co{y} components of this compound semantic dependency.
 This situation poses a challenge to current work on semantic dependencies,
@@
 and~\ref{lst:OOTA Multiplication Example, Initial Value 1}
-each have only one OOTA cycle,\footnote{
-	Note well that the initial values cannot be said to have come
-	out of thin air.
-	Although it is possible to have an OOTA cycle that does not
-	involve the initial values, but happens to have the same value,
-	the software would not be able to tell the difference.}
-and an execution having initial values
-of (say) two could have at most two OOTA cycles.
+each have only two OOTA cycles.
+Note well that although software cannot distinguish the initial
+values from those of an OOTA cycle having those same values, there is
+a very real difference in the corresponding executions.
+To wit, an execution not corresponding to an OOTA cycle will read at
+least one initialization value, while an execution corresponding to an
+OOTA cycle with those same values will read only from that cycle's stores.
+On the other hand, in a similar program having initial values of (say)
+two, both executions containing OOTA cycles would be distinguishable
+from the initial values.
+
 Listing~\ref{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}

### 2024-03-08 ootavacuous.tex
"Why rfe" appendix: It is not the compiler, but the implementation
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 demonstrate that rf can be atemporal, they demonstrate neither
-OOTA cycles nor OOTA-like behavior in full C++ compilers.
+OOTA cycles nor OOTA-like behavior in compiler-based C++ implementations.
 This appendix provides more involved examples that can be optimized
@@
 
-In the confines of the C++ abstract machine (omitting the no-OOTA
+Within the confines of the C++ abstract machine (omitting the no-OOTA
 restriction), this allows a value-17 OOTA cycle, courtesy of the fact

### 2024-03-08 ootavacuous.tex
"Why rfe" appendix: Rework first paragraph and pull in JMM CT 4
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+Listings~\ref{lst:OOTA With an Intra-Process Store-Load Link}
+and~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized}
+show the examples from
+Section~\ref{sec:OOTA: rf vs. rfe}
+in the form of litmus tests.
+A compiler is allowed to transform the program in
+Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link}
+to the form of
+Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized},
+which demonstrates that intra-thread reads-from (rfi) links may be
+eliminated by the compiler and therefore should not be used in a
+definition of OOTA.
+This section provides more involved examples to drive this point home.
+
 Although the examples presented in
-Section~\ref{sec:OOTA: rf vs. rfe}
 (and shown as litmus tests in
@@
 \label{lst:OOTA With an Intra-Process Store-Load Link, Optimized}
+\end{listing}
+
+The first step towards these more involved examples is shown in
+Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4},
+which expands
+Listing~\ref{lst:Causality Test Case 4}
+(discussed in
+Appendix~\ref{app:Causality Test Case 4})
+from two threads to three threads.
+The \co{herd7} tool finds OOTA values, as expected.
+All threads have straightforward semantic dependencies from their
+relaxed atomic loads to their relaxed atomic stores, which prevents this
+OOTA cycle on compiler-based C++ implementations.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-3proc.litmus @@
+\caption{Three-Process Version of JMM Causality Test Case 4}
+\label{lst:Three-Process Version of JMM Causality Test Case 4}
 \end{listing}
@@
 
-\subsection{Three-Process Version of JMM Causality Test Case 4}
-\label{app:Three-Process Version of JMM Causality Test Case 4}
-
-Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
-expands
-Listing~\ref{lst:Causality Test Case 4}
-(discussed in
-Appendix~\ref{app:Causality Test Case 4})
-from two threads to three threads.
-The \co{herd7} tool finds OOTA values, as expected.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-3proc.litmus @@
-\caption{Three-Process Version of JMM Causality Test Case 4}
-\label{lst:Three-Process Version of JMM Causality Test Case 4}
-\end{listing}
-
-All threads have straightforward semantic dependencies from their
-relaxed atomic loads to their relaxed atomic stores, which prevents this
-OOTA cycle on compiler-based C++ implementations.
-
 \subsection{Multi-Cycle Litmus Test}

### 2024-03-08 ootavacuous.tex
"Why rfe" appendix: Remove second paragraph
It is now redundant with the reworked first paragraph.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Although the examples presented in
-(and shown as litmus tests in
-Listings~\ref{lst:OOTA With an Intra-Process Store-Load Link}
-and~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized})
-demonstrate that rf can be atemporal, they demonstrate neither
-OOTA cycles nor OOTA-like behavior in compiler-based C++ implementations.
-This appendix provides more involved examples that can be optimized
-so as to demonstrate OOTA-like behavior.
-
 \begin{listing}[tbp]

### 2024-03-09 ootavacuous.tex
Drop simple-reordering appendix, keep litmus test.
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Further revisions to section D.3
Here are some fairly intensive revisions to section D.3.  Excess
material is removed, the remaining parts made more concise, and the
whole thing is now consistent with the main part of the paper.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 definition of OOTA.
-This section provides more involved examples to drive this point home.
+This section provides a more involved example to drive the point home.
 
@@
 
-The first step towards these more involved examples is shown in
+The first step in this direction is
 Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4},
@@
 Appendix~\ref{app:Causality Test Case 4})
-from two threads to three threads.
-The \co{herd7} tool finds OOTA values, as expected.
-All threads have straightforward semantic dependencies from their
-relaxed atomic loads to their relaxed atomic stores, which prevents this
-OOTA cycle on compiler-based C++ implementations.
+from two threads to three.
+The \co{herd7} tool reports OOTA values, and
+all threads have straightforward semantic dependencies from their
+loads to their stores.
 
@@
 Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
-revises
-Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
-so that the value stored to \co{y} is forced to be at most the
-value 17, and the value stored to \co{z} is forced to be at least the
-value 17.
-This of course means that the store to \co{z} must always be the value
-17, as can be seen from the \co{herd7} output.
-However, a single-thread-at-a-time analysis could not prove this.
+revises Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
+so that the value stored to \co{y} is forced to be at most 17
+and the value stored to \co{z} is forced to be at least 17.
+This of course means that the value stored to \co{z} must always be
+17, as can be seen in the \co{herd7} output.
+(However, a single-thread analysis could not prove this fact.)
 
@@
 Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
-further revises
-Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
-to flatten \co{PO()} and \co{P1()} into a single thread, so that the
-resulting \co{P0()} has an rfi link instead of an rfe link connecting
-the store to and load from \co{y}.
-This means that a C++ compiler could prove that the value 17 is
-always stored to \co{z} by examining \co{P0()} in isolation, given that
-it is also legal for a C++ compiler to replace the value loaded
-from \co{y} with the value just stored.\footnote{
-	This is not an option in C because atomic operations are volatile,
-	and thus observed behavior.}
+further revises the litmus test
+to flatten \co{P0()} and \co{P1()} into a single thread, so that the
+resulting \co{P0()} has an rfi link connecting
+the store to and load from \co{y}, rather than the rfe link in
+Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}.
+As a result, even a single-thread analysis could conclude that the value
+stored to \co{z} will always be 17, by examining \co{P0()} in isolation and
+using the fact that it is valid for a C++ compiler to assume the value
+loaded from \co{y} will always be the value just stored and so to omit
+the load.\footnote{
+	This is not an option in C because in that language,
+	atomic accesses are volatile and hence must be preserved
+	in the machine code.}
 
@@
 
-Within the confines of the C++ abstract machine (omitting the no-OOTA
-restriction), this allows a value-17 OOTA cycle, courtesy of the fact
-that C++ respects neither control nor data dependencies.
-However, the \co{herd7} tool does not identify this cycle, perhaps due
-to its restrictions on its handling of arithmetic on OOTA values.
-
-Given a compiler that proves that the value 17 is always
-stored to \co{z}, there might be no semantic dependency to
-\co{P0()}'s store to \co{z}, but must be to all of the other relaxed
-atomic stores.
-On the other hand, if the compiler does not prove the value
-stored to \co{z}, there will be a global-time delay between each of
-\co{P0()}'s loads and its store to \co{z}, preventing a value-17 OOTA
-cycle from forming.
-
-Either way, the value-17 OOTA cycle cannot occur in a compiler-based
-C++ implementation.
-Yes, a compiler that proves that the value 17 is always
-stored to \co{z} can satisfy the \co{exists} clause, but because
-there is no semantic dependency to \co{P0()}'s store to \co{z},
-this is not an OOTA cycle.
-
-But what if we chose to define an OOTA cycle as (sdep $\cup$ rf) instead
+If a compiler makes this observation and optimizes the program by
+storing a constant 17 to \co{z} rather than going through the
+computations on lines~15--17 of
+Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?},
+it will generate an executable which could produce the all-17's result
+described by the ``exists'' clause.
+(See Listing~\ref{lst:Why rfe Instead of Tried-And-True rf When z=17?},
+where the store to \co{z} has been moved up to line~9.\footnote{
+	The load from \co{y} on line~15 is retained even though it is
+	not used, because \co{r3} appears in the ``exists'' clause and
+	hence its final value is considered observable behavior.})
+This would not be an OOTA cycle, because in either form of the program
+the store to~\co{z} in \co{P0} is not semantically dependent on the
+load from~\co{x}, so there is no cycle in (sdep $\cup$ rfe).
+A different compiler that does not make this optimization will
+generate an executable which cannot produce the all-17's result at all,
+because of hardware dependencies and consequent instruction ordering
+in the machine code.
+
+Either way, no OOTA cycle can occur.
+But what if we chose to define an OOTA cycle with (sdep $\cup$ rf) instead
 of (sdep $\cup$ rfe)?
@@
 
-In that case, it would be necessary to separately consider the dependency
-from \co{P0()}'s load from \co{x} and \co{y} and that from its load
-from \co{y} and its store to \co{z}.
-The problem with this is shown in
-Listing~\ref{lst:Why rfe Instead of Tried-And-True rf When z=17?},
-which shows an equivalent program given the knowledge that the value~17
-will always be stored to \co{z}.
-Because all atomic operations are relaxed and because C++ does not
-respect control dependencies, the store to \co{z} can be moved up
-to the very beginning of \co{P0()}.
-This transformation causes the rfi link associated with \co{y} to
-go backwards in time, which is perfectly legitimate in this case,
-but which makes defining OOTA cycles in terms of (sdep $\cup$ rf)
-problematic.
-
-Again, in C, the load from \co{y} cannot be removed because it is
-volatile, and thus observable behavior.
-C++ could remove that load, but we left it in to demonstrate the all-17s
-outcome.
-
-This example demonstrates why OOTA cycles must be defined in terms
-of (sdep $\cup$ rfe), or in \co{herd7} nomenclature,
-\co{acyclic(sdep; rfe)}, as discussed in
-Section~\ref{sec:OOTA: rf vs. rfe}.
+In that case, it would be necessary to consider separately the dependency
+in Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
+connecting \co{P0()}'s load from \co{x} to its store to \co{y},
+and that connecting its load from \co{y} to its store to \co{z}.
+The first is a genuine semantic dependency; the second might or might
+not be, depending on one's definition.
+A single-thread analysis would conclude that it is, because of the
+possibility that the value loaded from \co{y} might be larger than 17.
+From this point of view, the all-17's result \emph{would} be
+considered an OOTA cycle.
+
+This example demonstrates another reason why defining OOTA cycles
+in terms of (sdep $\cup$ rf) is problematic.
+Doing so can lead to classifying a cycle as OOTA even though there
+is no genuine semantic dependency in one of the participating threads.
 
@@
 
-\subsection{Simple Reordering as Litmus Test}
-\label{app:Simple Reordering as Litmus Test}
-
-Listing~\ref{lst:Simple Reordering as Litmus Test}
-shows how
-Listing~\ref{lst:Simple Reordering}
-can be recast into a \co{herd7} litmus test.
+% \subsection{Simple Reordering as Litmus Test}
+% \label{app:Simple Reordering as Litmus Test}
+
+% Listing~\ref{lst:Simple Reordering as Litmus Test}
+% shows how
+% Listing~\ref{lst:Simple Reordering}
+% can be recast into a \co{herd7} litmus test.
 

### 2024-03-10 ootavacuous.tex
Move simple-reordering litmus test to Section 1.3
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-The earlier examples shown in
-Listings~\ref{lst:Simple OOTA}
-and~\ref{lst:Simple Reordering}
-can be recast as similar \co{herd7} litmus tests, as shown in
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/simple-reordering.litmus @@
+\caption{Simple Reordering as Litmus Test}
+\label{lst:Simple Reordering as Litmus Test}
+\end{listing}
+
+Later examples will usually combine the litmus test and the \co{herd7}
+output into one listing, as shown in
+Listing~\ref{lst:Simple Reordering as Litmus Test},
+which recasts
+Listing~\ref{lst:Simple Reordering}
+into litmus-test form.
+Listing~\ref{lst:Simple OOTA}
+can be also recast as a \co{herd7} litmus test, as shown in
 Listing~\ref{lst:Causality Test Case 4}
-on page~\pageref{lst:Causality Test Case 4}
-and Listing~\ref{lst:Simple Reordering as Litmus Test}
-on page~\pageref{lst:Simple Reordering as Litmus Test}, respectively.
+on page~\pageref{lst:Causality Test Case 4}.
 Other OOTA-related litmus tests may be found in
@@
 
-% \subsection{Simple Reordering as Litmus Test}
-% \label{app:Simple Reordering as Litmus Test}
-
-% Listing~\ref{lst:Simple Reordering as Litmus Test}
-% shows how
-% Listing~\ref{lst:Simple Reordering}
-% can be recast into a \co{herd7} litmus test.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/simple-reordering.litmus @@
-\caption{Simple Reordering as Litmus Test}
-\label{lst:Simple Reordering as Litmus Test}
-\end{listing}
-
 % \section{History}

### 2024-03-18 ootavacuous.tex
Record copies sent on March 18th
Olivier and Gonzalez overheard Jade and I discussing it and wanted
copies.  ;-)

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Alan Stern March 17 feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-This is not restricted to booleans, as can be seen in
+This issue is not restricted to booleans.
+For example, if given the code fragment shown in
+Figure~\ref{lst:OOTA Cycle}
+on
+page~\pageref{lst:OOTA Cycle},
+is the compiler permitted to act as if the code was instead as shown in
+Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}?
+Here, if the value loaded by line~9 is something other than 42,
+the load is retried on line~11, and perhaps a third time on line~14.
+If the value loaded in all three cases is something other than 42,
+then line~16 stores the last value loaded, and otherwise, the constant
+42 is stored by either line~18, 20, or~23.
+Despite the fact that the code in both listings can be argued to be
+equivalent, it is difficult to argue that a reasonable compiler
+would act as if the code in
+Figure~\ref{lst:OOTA Cycle}
+was instead that shown in
 Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}.
-This example is more straightforward in that it does not make quite so
-creative use of the multiple values loaded.
 
@@
 
+This section departs from one of the rules for this paper by considering
+an example having undefined behavior.
+
 Many compilers carry out optimizations based on the assumption
@@
 % Luc Maranget: Memory models.
-% Greg Kroah-Hartman and Linus Torvalds: FYI, trouble being caused.
 % Christoph Hellwig: Why slow on BPF memory model.
@@
 % Hans Boehm: Skeptic, memory models, OOTA.  Suggestions for others?
-% Olivier Giroux: Skeptic, memory models, NVIDIA.
 % Luke Geeson: Whatever...  Review.
@@
 
-% Done as of December 18, 2023:
-% David Goldblatt
-% Boqun Feng, Uladzislau Rezki, Neeraj Upadhyay, Joel Fernandes, and
+% Done as of December 18, 2023 %% as of March 18, 2023
+%% David Goldblatt
+%% Boqun Feng, Uladzislau Rezki, Neeraj Upadhyay, Joel Fernandes, and
 %	Frederic Weisbecker (RCU proteges)
-% Michael Wong
-% Maged Michael
+%% Michael Wong
+%% Maged Michael
 % John Wickerson: Thank you.  OK for Peter Sewell, Mark Batty, etc.
-% Akira Yokosawa: Professional courtesy, perfbook memory models.
-% Dan Kelley, Alexei Starovoitov, Mykola Lysenko: FYI.  OOTA dropped
+%% Akira Yokosawa: Professional courtesy, perfbook memory models.
+%% Dan Kelley, Alexei Starovoitov, Mykola Lysenko: FYI.  OOTA dropped
 % Alan Jeffrey: Fixed-point insight P0422R0.  Reached out via LinkedIn.
-% Jade Alglave: C++ herd model?  Memory models and variety of hardware.
-% Alan Stern: Memory models, mathematical logic, and variety of hardware.
+%% Jade Alglave: C++ herd model?  Memory models and variety of hardware.
+%% Alan Stern: Memory models, mathematical logic, and variety of hardware.
 % Martin Uecker (offered)
@@
 % Will Deacon: ARM hardware, C++, and OOTA.
+%% Greg Kroah-Hartman and Linus Torvalds: FYI, trouble being caused.
+%% Olivier Giroux: Skeptic, memory models, NVIDIA, requested.
+%% Gonzalo Brito <gonzalob@nvidia.com>
 

### 2024-03-18 ootavacuous.tex
Jade Alglave March 18 feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 that rfe links always go forward in global time, as those familiar with
-computer hardware and the laws of physics would expect.
+computer hardware, limits on speculative execution, and the laws of
+physics would expect.\footnote{
+	Our apologies to those who might feel that this is belaboring
+	the obvious.
+	And please understand that these points have proven helpful to
+	some of those whose whose work has never strayed too close to
+	hardware.}
 

### 2024-03-19 ootavacuous.tex
"Interthread Communications": Call out perftemporal.sh
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Use "exists" consistently
Most places put "exists" in the code font, but a couple enclose it in
quotes instead.  Make them consistent with the others.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	\co{git://git.kernel.org/pub/scm/linux/kernel/git/paulmck/perfbook.git}}
+See especially the \path{perftemporal.sh} script.
 
@@
 it will generate an executable which could produce the all-17's result
-described by the ``exists'' clause.
+described by the \co{exists} clause.
 (See Listing~\ref{lst:Why rfe Instead of Tried-And-True rf When z=17?},
@@
 	The load from \co{y} on line~15 is retained even though it is
-	not used, because \co{r3} appears in the ``exists'' clause and
+	not used, because \co{r3} appears in the \co{exists} clause and
 	hence its final value is considered observable behavior.})

### 2024-03-19 ootavacuous.tex
Updates for Appendix D4
This is an extensive set of edits for Appendix D4.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-This section contains several example litmus tests from the active
-mind of Alan Stern, expanding on the material presented in
+This section contains several example litmus tests
+expanding on the material presented in
 Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.
@@
 
-First, consider Listing~\ref{lst:No Invented Atomic Loads}.
-At first glance, this might appear straightforward.
+Consider Listing~\ref{lst:No Invented Atomic Loads}.
+At first glance, it might appear straightforward.
 \co{P0()}'s store to \co{z} clearly depends on its load from \co{y},
 and \co{P1()}'s store to \co{y} clearly depends on its load from \co{z}.
-If both dependencies are semantic dependencies, the OOTA cycle cannot
-form.
+If both dependencies are semantic, the OOTA cycle cannot be realized.
 
@@
 
-But is \co{P0()}'s dependency a semantic dependency?
-For example, is a C++ compiler permitted to
+But is \co{P0()}'s dependency truly semantic?
+Imagine what might happen if a C++ compiler is permitted to
 duplicate \co{P0()}'s load from \co{x} and then make use of both loads'
-values?
-If this is permitted, then the compiler would be within its rights
-to transform
+values.
+This would allow a compiler to transform
 Listing~\ref{lst:No Invented Atomic Loads}
@@
 
-Because the values of \co{r1}, \co{r2}, and \co{r3} are all known to be
-either zero or one, if the values returned by the two loads from \co{x}
-differ, then one of those two values must be equal to that from \co{y}.
-If the compiler is further permitted to make a runtime choice
-between the two values loaded from \co{x}, it could always choose the
-one that was equal to the value loaded from \co{y}, as shown by the
-\co{if} statement on line~13.
-This would mean that the store to \co{z} was always the constant 1,
-which would permit the store to be ordered before the load from \co{y},
-as shown in lines~14 and~15.
-Doing so permits the OOTA-like outcome of \co{r3 == r4 == 1}.
-
-The compiler might also drop the load from \co{y}, whose value
-is now unused except by the \co{exists} clause.
-However, the \co{exists} clause is considered to be observable behavior,
-which means that the load from \co{y} must remain.
+Because of the comparisons against zero on lines~9 and~10 of
+Listing~\ref{lst:No Invented Atomic Loads},
+the values of \co{r1} and \co{r2} are known to be either zero or one.
+Hence if the values of \co{r1a} and \co{r1b} on lines~9 and~10 of
+Listing~\ref{lst:Atomic Loads Invented}
+differ (say because \co{P2()} changes the value of \co{x}
+during the time between the two loads)
+then one of them must be equal to the value that would
+be obtained for \co{r2}.
+If the compiler is further permitted to make a choice at runtime
+between the two values loaded from \co{x}, it could always choose
+to use for \co{r1} the one that would be equal to \co{r2} (see line~16).
+Then the value stored to \co{z} would always be one,
+as shown on line~14, where the store has been moved before
+the load from \co{y} on line~15 to emphasize the fact that
+it does not depend on that load.
+The dependency from~\co{y} to~\co{z} would not be semantic after all,
+and the outcome \co{r1 == r2 == 1} would indeed be possible
+(although it would not be an example of OOTA because of the lack of a
+semantic dependency).
+
+This is a simple example of a more general phenomenon.
+Suppose $z$ is given by some function $f$ of $x$ and $y$.
+Under what conditions can we say that the value of $z$ doesn't depend
+on $y$?
+The standard answer is that this happens when there are values
+$z_0$ and $x_0$ such that for any $y$, we have $z_0 = f(x_0, y)$.
+But if we are allowed to choose from among multiple values of $x$,
+the situation gets more complicated.
+Then the answer would be that there is a value $z_0$ and
+a set $X$ of values for $x$ such that for any $y$, there is some
+$x \in X$ with $z_0 = f(x, y)$.
+In Listing~\ref{lst:Atomic Loads Invented},
+$f$ is the equality function and $X$ is simply the set $\{0, 1\}$;
+however, the reasoning applies in any situation where $x$ ranges
+over a finite collection of possible values.
+And as the example shows, the more general condition can hold in situations
+where the simpler condition does not, indicating that our intuitive
+notions of semantic dependency are not adequate when there can be
+multiple loads of the same variable.
 
@@
 
-One possible reaction to this situation is to declare that useful
-C++ compilers should never duplicate nonvolatile
-relaxed atomic loads.
-But suppose that the initial code was as shown in
+Clearly something has gone wrong if a valid (in loose C++) transformation
+like this one is capable of destroying what should be an obvious semantic
+dependency.
+One possible reaction is to declare that useful C++ compilers should never
+invent or duplicate nonvolatile relaxed atomic loads.
+But what if the initial code was as shown in
 Listing~\ref{lst:Unused Extra Atomic Load}?\footnote{
-	Such unused loads can easily be generated by complex macros or
-	template metaprograms.}
+	Unused loads can easily be generated by complex macros or
+	template metaprograms.
+	Typically compilers then remove them, but they are not obliged to.}
 In this case, the transformation to
 Listing~\ref{lst:Atomic Loads Invented}
-would not require inventing a load, but instead merely inventing a use of
+would not require inventing or duplicating a load,
+but instead merely inventing a use for
 the previously discarded value in \co{r0}.
+
+\medskip
 
@@
 Despite the fact that the code in both listings can be argued to be
-equivalent, it is difficult to argue that a reasonable compiler
+equivalent, it is difficult to imagine that a reasonable compiler
 would act as if the code in
@@
 
-But suppose that we drop the single-thread-analysis constraint for
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/inc-range.litmus @@
+\caption{Only Atomic Increment, Optimized}
+\label{lst:Only Atomic Increment, Optimized}
+\end{listing}
+
+Let us drop the single-thread-analysis constraint for
 the moment, and suppose that the compiler is able to prove
@@
 Listing~\ref{lst:Only Atomic Increment}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/inc-range.litmus @@
-\caption{Only Atomic Increment, Optimized}
-\label{lst:Only Atomic Increment, Optimized}
-\end{listing}
-
-Is a C++ compiler within its rights to behave as if the code
-shown in
-Listing~\ref{lst:Only Atomic Increment}
-was instead as shown in
+Is it okay for a C++ compiler to transform this program into
 Listing~\ref{lst:Only Atomic Increment, Optimized}?
-One argument in favor is that the value of \co{y} must have
-passed through the value 42 in the case where \co{r0} and \co{r1}
+An argument in favor is that the value of \co{y} must have
+passed through the value 42 if \co{r0} and \co{r1}
 bracket that value.
-Arguments against might cite the added overhead of the invented load
-in Listing~\ref{lst:Only Atomic Increment, Optimized}, or that the
-possibility that the check for the value 42 is intended for statistical
+Arguments against might cite the added overhead of the invented load,
+or the possibility that the check for the value 42 is intended for statistical
 sampling.
@@
 	adjacent to each other.
-\item	Any time a compiler might be tempted to invent uses from
-	values from nonvolatile relaxed atomic loads from the same
-	object, the loads should instead be fused into a single load.
+\item	Any time a compiler might be tempted to invent uses for
+	values of nonvolatile relaxed atomic loads from the same
+	object, the loads should instead be merged into a single load.
 	This reaction might be attractive to those who carefully consider
@@
 is customary, for example, when hoisting a load out of a loop.
-In this case, compilers might hoist the load from x out of
+In this case, compilers might hoist the load from \co{x} out of
 the loop:
-
 \begin{quote}
@@
 do_something_with(x);
-for (i = 0; i < limit i++)
+for (i = 0; i < limit; i++)
     y[i] += x;
@@
 \end{quote}
-
 A compiler might act as if the program had instead been this:
-
 \begin{quote}
@@
 \end{quote}
-
-Note that passing \co{x} to the call to \co{do_something_with()} is important
+Note that passing \co{x} to \co{do_something_with()} is important
 because this allows the compiler to assume that inventing additional
@@
 
-But if \co{x} is a nonvolatile atomic, then there are no worries about
+But if \co{x} is a nonvolatile atomic, there are no worries about
 data races.
-In theory, the compiler could make the above transformation without
+In theory, if the \co{do_something_with()} call was not present
+the compiler could still make the above transformation without
 worry, especially given that the value loaded from \co{x} is discarded in
@@
 After all, optimizations are supposed to speed things up.
-This means that a compiler might be wise to adhere to the
+A compiler might be wise to adhere to the
 same restriction that avoids data races for loads from non-atomic
-variables in order to avoid expensive cache misses.
+variables in order to avoid potentially expensive cache misses.
 

### 2024-03-19 ootavacuous.tex
Edits for Appendix D5
The text for Appendix D5 was wrong in several respects.  The program
doesn't have UB, and the OOTA-like result of the optimization proves
that the optimization was invalid.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\subsection{OOTA From Undefined-Behavior Optimizations}
-\label{app:OOTA From Undefined-Behavior Optimizations}
-
-This section departs from one of the rules for this paper by considering
-an example having undefined behavior.
+\subsection{Undefined Behavior and Unwise Optimization}
+\label{app:Undefined Behavior and Unwise Optimization}
 
 Many compilers carry out optimizations based on the assumption
-that UB cannot happen, and these optimizations can result in OOTA cycles,
-with a simple example shown in
-Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}.
+that undefined behavior (UB) cannot happen,
+and if not performed carefully, these optimizations may be invalid
+because of OOTA-like interactions.
+A simple example is shown in
+Listing~\ref{lst:OOTA-Like Behavior Due To Divide-By-Zero UB}.
 
@@
 @@ DisplayRunLitmus litmus/oota-div-ub.litmus @@
-\caption{OOTA Due To Divide-By-Zero UB}
-\label{lst:OOTA Due To Divide-By-Zero UB}
-\end{listing}
-
-On line~14 of this listing, the value of \co{r2} must be either zero
-or one due to the comparison on line~13.
-If it is the value one, then there is a divide-by-zero error, which
-enables the compiler to assume that the value of \co{r1} is
-always zero.
-This in turn allows the compiler to behave as if the expression
-stored to \co{y} on line~14 was the constant 1, at which point there
-is no semantic dependency from the load from \co{x} on line~13 to the
-store to \co{y} on line~14.
-
-The compiler is then within its rights to act as if lines~13 and~14
-were interchanged.
-At which point, the order of execution could be as follows:
-
+\caption{OOTA-Like Behavior Due To Divide-By-Zero UB}
+\label{lst:OOTA-Like Behavior Due To Divide-By-Zero UB}
+\end{listing}
+
+On line~14 of this listing, the computation \co{1 / (r2 <= 0)}
+will yield one if \co{r2} is not positive and will result in UB
+otherwise (a divide-by-zero error).
+It may not be obvious, but the program is in fact UB-free---there are
+no abstract executions in which \co{r2} is greater than zero.
+To see why not, consider that in any such execution
+\co{P1()} would not store anything to \co{y};
+it certainly would not store a positive value.
+This means that the value loaded on line~8 would have to be \co{y}'s
+initial value of zero, and so would the value stored to \co{x} on
+line~9, and hence it would not be possible for the load on line~13
+to get a positive value for \co{r2} after all.
+
+Even for a compiler analyzing \co{P1()} in isolation,
+it is clear that either line~14 will store one to \co{y} or
+else the division by zero will cause UB.
+This allows the compiler to assume that the store will always
+take place, on the grounds that UB cannot happen,
+and so the compiler may optimize the comparison and division
+by replacing the whole expression with a constant \co{1}.
+There will then be no dependency from the load on line~13
+to the store and thus nothing forcing the store to execute
+after the load.
+
+When the resulting program runs on a weakly ordered architecture,
+the execution could go as follows:
 \begin{enumerate}
-\item	Line~14 stores the constant 1 to \co{y}.
-\item	Line~8 loads the constant 1 from \co{y}, setting \co{r1} to 1.
-\item	Line~9 stores the constant 1 to \co{x}.
-\item	Line~8 loads the constant 1 from \co{x}, setting \co{r2} to 1.
+\item	Line~14 executes out of order, storing the constant \co{1} to \co{y}.
+\item	Line~8 loads one from \co{y}, setting \co{r1} to one.
+\item	Line~9 stores one to \co{x}.
+\item	Line~13 loads one from \co{x}, setting \co{r2} to one.
 \end{enumerate}
-
-Please note that the OOTA-like cycle called out in the \co{exists}
-clause can happen even though the program never actually executes any
-code that would have resulted in division-by-zero UB.
+The final result would satisfy the \co{exists} clause,
+exhibiting observable behavior (\co{r2} is one at the end)
+that no abstract execution of the original program could produce.
+(This could happen even if all of the atomic objects in
+Listing~\ref{lst:OOTA-Like Behavior Due To Divide-By-Zero UB}
+were converted to volatile.)
+
+This result shows that the proposed optimization was invalid.
+To prevent the unwanted behavior, the compiler would have to take an
+extra step to force the optimized store on line~14 to be ordered after
+the load on line~13, possibly by changing the load to a load-acquire
+or changing the store to a store-release.
+
+In other words, the compiler would have to preserve the dependency
+ordering that the UB-based optimization would otherwise destroy.
 It is tempting to consider using techniques
@@
 considering line~14 in isolation.
-Once the constant 1 is substituted for line~14's expression, the
-hardware is free to reorder lines~13 and~14 even if a compiler-based
-C++ implementation refrains from doing so.
-Note that this OOTA-like cycle can appear even if all of the atomic
-operations in
-Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}
-are converted to volatile.
-
-Therefore, this paper focuses on UB-free programs, rare though these
-beasts might be among large-scale C++ programs running in production.
 

### 2024-03-21 ootavacuous.tex
Simplify inc.litmus and inc-range.litmus
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Create summary appendix and relegate a couple tests to it
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 @@ DisplayRunLitmus litmus/inc-range.litmus @@
-\caption{Only Atomic Increment, Optimized}
-\label{lst:Only Atomic Increment, Optimized}
+\caption{Only Atomic Increment, Extended}
+\label{lst:Only Atomic Increment, Extended}
 \end{listing}
@@
 Is it okay for a C++ compiler to transform this program into
-Listing~\ref{lst:Only Atomic Increment, Optimized}?
+Listing~\ref{lst:Only Atomic Increment, Extended}?
 An argument in favor is that the value of \co{y} must have
@@
 
-\subsection{Multi-Cycle Litmus Test}
-\label{app:Multi-Cycle Litmus Test}
-
-Listing~\ref{lst:Multi-Cycle Litmus Test}
-expands
-Listing~\ref{lst:Causality Test Case 4}
-in a different way, moving to four threads but having only the
-original two variables \co{x} and \co{y}.
-The \co{herd7} tool shows multiple different possible OOTA cycles,
-as expected.
-However, the value returned by each thread's load directly affects the
-value stored, so there is a semantic dependency in each thread.
-Therefore, none of these OOTA cycles can arise in a compiler-based
-C++ implementation.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-two-source.litmus @@
-\caption{Multi-Cycle Litmus Test}
-\label{lst:Multi-Cycle Litmus Test}
-\end{listing}
-
-\subsection{Non-LB Litmus Test}
-\label{app:Non-LB Litmus Test}
-
-Listing~\ref{lst:Non-LB Litmus Test}
-shows a litmus test that can produce an OOTA cycle, but that is not
-a member of the load-buffering (LB) class of litmus tests.
-Suppose that lines~9 and~10 both somehow return the value 21.
-Then line~12 will store the value 42 to \co{x}, which will cause
-line~17 to store 21 to \co{y} and line~22 to store 21 to \co{z},
-thus justifying the values initially loaded by lines~9 and~10.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-non-lb.litmus @@
-\caption{Non-LB Litmus Test}
-\label{lst:Non-LB Litmus Test}
-\end{listing}
-
-All thread's loads' return values affect the value stored, so all
-threads have semantic dependencies between their respective relaxed
-atomic loads and relaxed atomic stores.
-Therefore, an OOTA cycle cannot form in a compiler-based C++
-implementation.
-
-Note that in a C++ implementation that (erroneously) permitted an OOTA
-cycle to form, that cycle would involve both of \co{P0()}'s \co{r0}
-and \co{r1} local variables.
+\subsection{Additional Litmus Tests}
+\label{app:Additional Litmus Tests}
+
+This section lists a couple of other litmus tests of interest:
+
+\begin{itemize}
+\item	The litmus test at
+	\url{https://github.com/paulmckrcu/oota/blob/master/litmus/oota-two-source.litmus}
+	shows multiple overlapping OOTA cycles.
+\item	The litmus test at
+	\url{https://github.com/paulmckrcu/oota/blob/master/litmus/oota-non-lb.litmus}
+	shows that OOTA cycles are not confined to the load-buffering (LB)
+	pattern.
+\end{itemize}
 

### 2024-03-22 ootavacuous.tex
Many small edits throughout the paper
These changes do fairly minor things, such as fixing grammatical errors
or inconsistencies, punching up the language, clarifying the text, and
so on.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 however, as shown by
-Listing~\ref{lst:OOTA Cycle},
+Listing~\ref{lst:OOTA Cycle} below,
 OOTA cycles need not involve nonsensical values.
@@
 occurring at all.}
-Here the dependencies are trivial, because the values stored simply
+Here the dependencies are trivial because the values stored simply
 \emph{are} the values that were loaded.)
@@
 values of the global shared variables \co{x} and \co{y} to zero.
-Variables that are not explicitly initialized are initialized
-to zero by default.
+(Variables that are not explicitly initialized are initialized
+to zero by default.)
 Lines~7--13 define the first thread, \co{P0()}, and lines~15--21
@@
 (written in a slightly stilted manner to keep the load operations separate
-from the rest).
+from the rest and because \co{herd7}'s knowledge of C++ is limited).
 
@@
 Listing~\ref{lst:Simple OOTA}
-can be also recast as a \co{herd7} litmus test, as shown in
+can also be recast as a \co{herd7} litmus test, as shown in
 Listing~\ref{lst:Causality Test Case 4}
@@
 then discusses properties of semantic dependences, showing that they
-are functions of executions rather than strictly of source code.
-
-\subsection{OOTA: rf vs. rfe}
+are functions of executions rather than strictly of source code,
+among other things.
+
+\subsection{OOTA: rf vs.\ rfe}
 \label{sec:OOTA: rf vs. rfe}
 
-Semantic dependencies are only one type of link in an OOTA cycle.
+Semantic dependencies form only one type of link in an OOTA cycle.
 The other type extends from a given store to a load that returns the
 value stored.
-It is tempting to make the intuitive argument that the store must precede
-the load in global time, and then to combine this with the notion that
+It is tempting to argue that the store must precede
+the load in global time and combine this with the intuitive notion that
 any real C++ implementation must consume global time when computing a
@@
 each thread and rf is the set of store-to-load ``reads-from'' links,
-whether intra-thread
+whether within a thread
 (rfi) or between threads (rfe).\footnote{
@@
 This is a fine definition and is consistent with the words in the C++
-standard, but it has an intra-thread problem exemplified by the following
-code:
+standard, but it has a problem with intrathread rfi links
+exemplified by the following code:
 \begin{quote}
@@
 that adds \co{z} along with lines~3 and~4.
-The problem is that a C++ implementation is within its rights to
-note that line~3 might well execute immediately after line~2, giving
+The problem is that a C++ implementation may
+note that line~3 could well execute immediately after line~2, giving
 other threads no chance to modify \co{y} in between.
@@
 for rf, defining an OOTA cycle---for now---as a cycle in sdep $\cup$ rfe,
-where rfe is the set of store-to-load links excluding intra-thread rfi links.
+where rfe is the set of store-to-load links between different threads.
 Any rfi links in a cycle can instead be interpreted as part of sdep.
@@
 ``semantic dependency'', it also enables us to separate cleanly
-the inter-thread and intra-thread portions of any given OOTA cycle.
+the interthread and intrathread portions of any given OOTA cycle.
 % \footnote{
@@
 semantic ones (depending on one's definition).
-That is, even though the potential for a dependency exists in the
+That is, even when the potential for a dependency exists in the
 source code for a thread, there may be no dependency in the machine
@@
 and~\ref{sec:Semantic Dependencies Not Affected by if Statements}
-shows that \co{if} statements can have surprising effects on semantic
+show that \co{if} statements can have surprising effects on semantic
 dependencies, up to and including eliminating them completely.
@@
 This makes OOTA more complicated than a simple cycle but
-we will continue to refer to OOTA cycles out of habit.
+we will continue to refer to ``OOTA cycles'' out of habit.
 Note that this new definition includes and generalizes the earlier
@@
 the two statements as performing two different stores or as performing
-for all intents and purposes a single store, independent of x.
+for all intents and purposes a single store, independent of \co{x}.
 Reasonable C++ implementations might disagree on this matter and
@@
 pattern of loads and stores be consistent with the C++ memory model.
-We assume that programs will not indulge in any other computations that
-could be expected to vary spontaneously from one execution to another,
+We assume that programs will not indulge in any computations that
+could vary spontaneously from one execution to another,
 such as basing a dependency on the time of day or a process ID.
@@
 in different threads.  Even if a load in one thread reads from a
-store in another thread, the standard does not require the store to come
-before the load.
+store in another thread, the standard generally does not require
+the store to come before the load in any meaningful way.
 
@@
 For this reason, the hardware executions we use will comprise (along
-with machine code being run) the computer architecture and for each
+with the machine code being run) the computer architecture and for each
 load instruction, the store instruction from which it reads and the
@@
 	rules of the abstract machine'' (4.1.2p6.1 \co{[intro.abstract]}).
-\item	(There is a condition on the interleaving of input and output,
+\item	(There is a condition on the timing and interleaving of input and output,
 	which does not matter for our purposes.)
@@
 act more like I/O.
-In particular, when running in a nonbenign environment a program's
+In particular, when it runs in a nonbenign environment, a program's
 volatile loads can return unpredictable values.
@@
 	device firmware (or vice versa), roughly speaking.}
-so they do not contribute in the same way to the happens-before relation.
+so they do not contribute to the happens-before relation.
 Also, in these environments the rfe relation does not apply to volatile
-loads and hence the accesses in an OOTA cycle must be nonvolatile.
+loads and stores, and hence the accesses in an OOTA cycle must be nonvolatile.
 
@@
 Given this relation between abstract and hardware executions, it is time
-to turn our attention to the tool that selects hardware execution so as
-to preserve that relation, namely, compilers.
+to turn our attention to the tools that manage hardware executions so as
+to enforce that relation, namely, compilers.
 
@@
 always to generate the same machine-code output for threads having
-the same source code, regardless of the rest of the programs
-those threads are contained in.
+the same source code, regardless of the rest of the code in the programs
+containing those threads.
 Later on we will return to globally optimizing compilers.
@@
 
-\subsection{Volatile and ``Quasi Volatile'' Accesses}
+\subsection{Volatile and Quasi Volatile Accesses}
 \label{sec:Volatile and Quasi Volatile Accesses}
@@
 	volatile objects.)
-\item	The map is one-to-one: Different accesses in $A$ map to
+\item	The map is one to one: Different accesses in $A$ map to
 	different accesses in $H$.
@@
 (To be fair, this should be considered more as a \emph{result} of
-ordering rather than a \emph{cause} of ordering, in that
+ordering than as a \emph{cause} of ordering, in that
 if a load commits before a store on another thread then it
@@
 Therefore instructions following a conditional- or computed-branch
-instruction in an execution must commit after the branch, and hence
+instruction in a hardware execution must commit after the branch, and hence
 after the source for the branch's condition or destination input.
@@
 omitted in some hardware execution $H$ realizing $A$.
-Then for some $i$, load $R_i$ is not omitted and the hardware
+Then for some $i$, load $R_i$ is not omitted in $H$ and the
 instruction it maps to is ordered before the instruction $W$ maps to.
@@
 Thus there would be no way to tell, merely by observing the effects of
-$A$, whether there was an OOTA cycle or not.
+$H$, whether there was an OOTA cycle or not.
 For this reason we declare realizations of OOTA cycles in which
@@
 quasi volatile.
-For compilers that perform only single-thread analysis, not global
-analysis, quasi volatile is sufficient.
+(For compilers that perform only single-thread analysis, not global
+analysis, quasi volatile is sufficient.)
 In this setting we can relate abstract and hardware executions by
@@
 Section~\ref{sec:Outstanding Issues}
-discusses a general questions regarding our definitions.
+discusses general questions regarding our definitions.
 
@@
 In this section we consider implementations whose compilers perform
-only single-thread analysis and treat all atomic objects as
+only single-thread analysis and treat atomic objects as
 quasi volatile.
@@
 (While it possible to argue about semantic dependencies in abstract
-executions that have no hardware realizations, doing so seems like a
-pointless activity.)
+executions that have no hardware realizations, doing so seems pointless.)
 
@@
 corresponds to $W'$.
-If $W'$ is part of the initial or common portion of $H$ this is no
+If $W'$ is in the initial or common portion of $H$ this is no
 problem; we can take $X'$ to be $W'$ itself.
@@
 compiler or implementation.
-This may strike some people as wishy-washy and as avoiding the real
+This may strike some people as wishy-washy and avoiding the real
 problem, in that a given piece of code should either contain or not
@@
 
-To see why, recall that a semantic dependency must have a witness,
+To see why not, recall that a semantic dependency must have a witness,
 another execution in which the store acts differently.
-But this transformed program has no other executions; every time it
+But this transformed program has no other hardware executions; every time it
 runs it will store 42 to both \co{x} and \co{y}.
@@
 int r3 = (x != 0);
-int r4 = y;
+int r4 = (y != 0);
 z = (r2 == r4);
@@
 int r3 = (x != 0);
-int r4 = y;
+int r4 = (y != 0);
 if (r1 != r3) {
@@
 
-To demonstrate the semantic dependency, a suitable witness would have
+To demonstrate the semantic dependency in the original code,
+a suitable witness would have
 to include a hardware realization of the abstract execution in which
@@
 
-The following sections discuss these constraints, starting with with
+The following sections discuss these constraints, starting with
 hardware constraints and continuing with additional constraints imposed
@@
 
-Applied to Listing~\ref{lst:OOTA Cycle}, for another example,
+Applied to Listing~\ref{lst:OOTA Cycle} on
+page~\pageref{lst:OOTA Cycle}, for another example,
 the principles dictate that if the load on line~8 reads from the store
@@
 The requirement applies within a single
-multi-threaded core as well as between cores.
-The fact that intra-core communication is faster than inter-core
+multithreaded core as well as between cores.
+The fact that intracore communication is faster than intercore
 communication does not magically give loads the ability to commit
@@
 extra cache misses for no good purpose.
-And duplicating an atomic load can even lead to incorrect results if done
+And duplicating an atomic load can also lead to incorrect results if done
 carelessly.
@@
 	And please understand that these points have proven helpful to
-	some of those whose whose work has never strayed too close to
+	some of those whose work has never strayed too close to
 	hardware.}
@@
 It is natural to dismiss user influence when compared to the text of
-standards or the code in implementations, but both are subject to
+standards or the code in implementations, but both of these are subject to
 change and do change over time.
@@
 In all cases, the kernel and the user application would need to continue
-execution, and would thus require an exact exception.
+execution, and would thus require a precise exception.
 
 Systems that offload processing of correctable ECC errors to software
-also require exact exceptions.
+also require precise exceptions.
 After all, the software is going to need to be able to correct the error
@@
 ECC errors are detected near the end of their corresponding load
-instructions' execution, so that any need for exact exceptions rules out
-the commiting of any subsequent stores until after the load has fetched
+instructions' execution, and so any need for precise exceptions rules out
+committing any subsequent stores until after the load has fetched
 (and possibly corrected) its value.
@@
 		are uncorrectable errors, there might soon be errors
-		that translate one valid bit pattern to another valid
+		that change one valid bit pattern to another valid
 		bit pattern, which could result in a lack of safety.}
@@
 stores), or \co{memory_order_acq_rel} (for read-modify-write operations).
-On TSO systems and on systems featuring exact ECC exceptions, this
+On TSO systems and on systems featuring precise ECC exceptions, this
 would allow tooling to be brought to bear without performance or
@@
 
-Listings~\ref{lst:OOTA With an Intra-Process Store-Load Link}
-and~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized}
+Listings~\ref{lst:OOTA With an Intrathread Store-Load Link}
+and~\ref{lst:OOTA With an Intrathread Store-Load Link, Optimized}
 show the examples from
@@
 A compiler is allowed to transform the program in
-Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link}
+Listing~\ref{lst:OOTA With an Intrathread Store-Load Link}
 to the form of
-Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized},
-which demonstrates that intra-thread reads-from (rfi) links may be
+Listing~\ref{lst:OOTA With an Intrathread Store-Load Link, Optimized},
+which demonstrates that intrathread reads-from (rfi) links may be
 eliminated by the compiler and therefore should not be used in a
@@
 @@ DisplayRunLitmus litmus/oota-3-2-proc.litmus @@
-\caption{OOTA With an Intra-Process Store-Load Link}
-\label{lst:OOTA With an Intra-Process Store-Load Link}
+\caption{OOTA With an Intrathread Store-Load Link}
+\label{lst:OOTA With an Intrathread Store-Load Link}
 \end{listing}
@@
 @@ DisplayRunLitmus litmus/oota-3-2-proc-opt.litmus @@
-\caption{OOTA With an Intra-Process Store-Load Link, Optimized}
-\label{lst:OOTA With an Intra-Process Store-Load Link, Optimized}
+\caption{OOTA With an Intrathread Store-Load Link, Optimized}
+\label{lst:OOTA With an Intrathread Store-Load Link, Optimized}
 \end{listing}
@@
 %	that you either move further up the stack or take early
-%	retirement."  Note that that C++ committee's concurrency
+%	retirement."  Note that the C++ committee's concurrency
 %	expertise has grown beyond recognition over the past 15 years.

### 2024-03-22 ootavacuous.tex
Move oota-invent-int-load.litmus to D.6
Also fix up dead cross-reference while in the area.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 One exception appears in
-Appendix~\ref{app:Non-LB Litmus Test},
+Appendix~\ref{app:Additional Litmus Tests},
 but this section will present a more difficult example.
@@
 the previously discarded value in \co{r0}.
-
-\medskip
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-invent-int-load.litmus @@
-\caption{Inventing Non-Boolean Atomic Loads}
-\label{lst:Unused Inventing Non-Boolean Atomic Loads}
-\end{listing}
-
-This issue is not restricted to booleans.
-For example, if given the code fragment shown in
-Figure~\ref{lst:OOTA Cycle}
-on
-page~\pageref{lst:OOTA Cycle},
-is the compiler permitted to act as if the code was instead as shown in
-Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}?
-Here, if the value loaded by line~9 is something other than 42,
-the load is retried on line~11, and perhaps a third time on line~14.
-If the value loaded in all three cases is something other than 42,
-then line~16 stores the last value loaded, and otherwise, the constant
-42 is stored by either line~18, 20, or~23.
-Despite the fact that the code in both listings can be argued to be
-equivalent, it is difficult to imagine that a reasonable compiler
-would act as if the code in
-Figure~\ref{lst:OOTA Cycle}
-was instead that shown in
-Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}.
 
@@
 
-This section lists a couple of other litmus tests of interest:
+This section lists a few other litmus tests of interest:
 
@@
 	pattern.
+\item	The litmus test at
+	\url{https://github.com/paulmckrcu/oota/blob/master/litmus/oota-invent-int-load.litmus}
+	shows another invented-load scenario loosely based on
+	Figure~\ref{lst:OOTA Cycle}
+	on
+	page~\pageref{lst:OOTA Cycle}.
+	See Appendix~\ref{app:Inventing Atomic Loads}
+	for more discussion on this topic.
 \end{itemize}

### 2024-03-24 ootavacuous.tex
Change "vs." to "versus"
To be consistent with the title of section 6.4.2, change the title of
section 2.1 to "OOTA: rf versus rfe".  Besides, "rf" and "rfe" are
already acronyms so the title doesn't really need another abbreviation.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Section~\ref{sec:OOTA: rf vs. rfe}
+Section~\ref{sec:OOTA: rf versus rfe}
 demonstrates advantages of formulating an OOTA cycle as a cycle in
@@
 
-\subsection{OOTA: rf vs.\ rfe}
-\label{sec:OOTA: rf vs. rfe}
+\subsection{OOTA: rf versus rfe}
+\label{sec:OOTA: rf versus rfe}
 
@@
 show the examples from
-Section~\ref{sec:OOTA: rf vs. rfe}
+Section~\ref{sec:OOTA: rf versus rfe}
 in the form of litmus tests.

### 2024-03-25 ootavacuous.tex
Final (?) set of small edits
I hope there won't need to be any more small edits for the main part of
the paper.  The appendices might still need a little attention.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \emph{full C++} for comfort)
-to denote a hypothetical standard that excludes this prohibition, but
+to denote a hypothetical standard that excludes this prohibition but
 is otherwise identical to full C++.
@@
 	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
-which also may be found in
+a version of which may be found in
 Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}.
@@
 standard, but it has a problem with intrathread rfi links
-exemplified by the following code:
+as exemplified by the following code:
 \begin{quote}
@@
 ordered series of steps to be carried out by the abstract machine.
-However, there is no ordering relation between steps carried out
-in different threads.  Even if a load in one thread reads from a
-store in another thread, the standard generally does not require
-the store to come before the load in any meaningful way.
+However, with a few exceptions\footnote{
+	Such as a load-acquire synchronizing with a store-release.}
+there is no ordering relation between steps carried out
+in different threads.
+Even if a relaxed atomic load in one thread reads from a relaxed
+atomic store in another thread, the standard does not require the
+store to come before the load in any meaningful way.
 
@@
 	volatile objects.)
-\item	The map is one to one: Different accesses in $A$ map to
+\item	The map is one-to-one: Different accesses in $A$ map to
 	different accesses in $H$.
@@
 object accesses.
-The last two items are left out: Compilers are allowed to merge or
+The last two bullet points are left out: Compilers are allowed to merge or
 omit accesses to these objects.
-Because of this, the item about preserving the rf relation
+Because of this, the bullet point about preserving the rf relation
 applies only when $R$ is not omitted, in which case $W$ must not
@@
 Section~\ref{sec:For Compilers Using Global Analysis}
-slightly relaxes this single-thread restriction,
+relaxes this single-thread restriction,
 Section~\ref{sec:Verifying the Fundamental Property}
@@
 the issue of semantic dependency is moot.
-You can give either answer since the question is basically meaningless.
+You can give either answer since it will have no effect.
 Therefore we'll assume that $W$ is not omitted.
@@
 
+\medskip
+
 Despite these outstanding issues, our definition of semantic dependency
@@
 in prior work.
-In real-world C++ implementations, the standard's abstract machine
-must be mapped onto real hardware, however, and this imposes additional
+In real-world C++ implementations, however, the standard's abstract machine
+must be mapped onto real hardware, and this imposes additional
 constraints stemming from hardware architecture and design.
@@
 C++ memory model.
-The ``order-preserving'' bullet from
+The ``order-preserving'' bullet point from
 Section~\ref{sec:Volatile and Quasi Volatile Accesses}
@@
 (the four coherence rules in 6.9.2.2p14--19 \co{[intro.races]}).
-The ``onto'' bullet also largely reiterates restrictions that follow
+The ``onto'' bullet point also largely reiterates restrictions that follow
 from the standard, in particular,
@@
 under appropriate circumstances;
-these topics correspond to the ``total'' and ``one-to-one'' bullets of
+these topics correspond to the ``total'' and ``one-to-one'' bullet points of
 Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
@@
 This holds even when the atomic objects are treated as quasi volatile,
-providing an example of how the ``one-to-one'' bullet of
+providing an example of how the ``one-to-one'' bullet point of
 Section~\ref{sec:Volatile and Quasi Volatile Accesses} does not apply

### 2024-03-29 ootavacuous.tex
Make instruction cost commentary consistent
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Final (?) edits for appendices
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	Coherence-order (external), which connects a store to any other store
+	(in a different thread)
 	to the same object that comes later in the object's
@@
 \item[fre]
-	From-read (external), which connects a load to any store to
+	From-read (external), which connects a load to any store
+	(in a different thread) to
 	the same object that overwrites either the value that the load
@@
 \item[rfe]
-	Reads-from (external), which connects a store to any load from
+	Reads-from (external), which connects a store to any load
+	(in a different thread) from
 	the same object that returns the value stored.
@@
 One way this could happen is if the cache line containing
-\co{x} arrived at CPU~3 before arriving at CPU~0.
+\co{x} arrived at CPU~0 before arriving at CPU~3.
 A more fanciful explanation is that both values arrived at an
@@
 but the variety of existing practice in the 1980s prevented this.
-Plus machine instructions were much more expensive back then, so the
-signedness of \co{char} was much more of a performance concern than
-it is today.
+Plus, optimizers were much less capable back then, so the signedness of
+\co{char} was much more of a performance concern than it is today.
 
@@
 optimizations that tooling could in fact analyze.
-Therefore, the next section looks at standardizing a less severe
-restriction.
+Therefore, the next two sections look at standardizing less severe
+restrictions.
 
@@
 restrictions~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
-
-This suggests the addition of a \co{memory_order_load_store} member to
-the \co{memory_order} enumeration, a topic taken up by the next section.
 
@@
 the case where \co{limit} is less than or equal to zero.
-Except that the data-race worry is replaced by a cache-miss worry.
-After all, optimizations are supposed to speed things up.
+Except that the data-race worry is replaced by a cache-miss worry;
+after all, optimizations are supposed to speed things up,
+not slow them down.
 A compiler might be wise to adhere to the
@@
 line~9, and hence it would not be possible for the load on line~13
-to get a positive value for \co{r2} after all.
+to get a positive value for \co{r2} in the first place.
 
@@
 
-In other words, the compiler would have to preserve the dependency
+In other words, the compiler would have to preserve the dependency-induced
 ordering that the UB-based optimization would otherwise destroy.

### 2024-03-29 ootavacuous.tex
Additional minor edits
Here are a few very minor last-minute changes.  The biggest one is
probably my decision that "many-to-one" ought to be hyphenated because
it is not a modifier-noun combination.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 rather than being strict functions of the source code.
-Section~\ref{sec:Semantic Dependencies Can Be Many To One}
+Section~\ref{sec:Semantic Dependencies Can Be Many-To-One}
 shows that semantic dependencies do not necessarily extend
@@
 
-\subsubsection{Semantic Dependencies Can Be Many To One}
-\label{sec:Semantic Dependencies Can Be Many To One}
+\subsubsection{Semantic Dependencies Can Be Many-To-One}
+\label{sec:Semantic Dependencies Can Be Many-To-One}
 
@@
 
-The exact definition of a computer language is subject to some debate,
+The exact definition of a computer language is a subject of some debate,
 with standards, implementations, and users all having some degree of
 influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
-each of which is subject to change over time.
+and each being prone to change over time.
 In areas that are not well settled or where users might reasonably
@@
 Therefore we can rule out transformations like this one by insisting
-the compiler not invent atomic loads.
+the compiler not invent (or duplicate) atomic loads.
 In fact, we will require that atomic accesses be treated as
@@
 
-We will say that one instruction is ordered after another to mean that
+We will say that one instruction is \emph{ordered after} another to mean that
 the CPU forces it to commit after the other one commits.
@@
 \end{quote}
-It's quite straightforward to show that for any implementation in which
+It's quite straightforward to show that under any implementation in which
 semantic dependencies satisfy the Fundamental Property, no OOTA cycle
@@
 For this reason we declare realizations of OOTA cycles in which
-all the accesses are omitted to be trivial.
+all the accesses are omitted to be \emph{trivial}.
 
@@
 	nomenclature; in the memory model a store was said to commit at
-	the time when it makes its new value available to other
+	the time when it made its new value available to other
 	threads, whereas we say that a store commits at the time when
@@
 of ``OOTA cycle'', which is shown in
-Section~\ref{sec:Semantic Dependencies Can Be Many To One}.
+Section~\ref{sec:Semantic Dependencies Can Be Many-To-One}.
 
@@
 \bibliographystyle{plain}
+\addcontentsline{toc}{section}{References}
 \bibliography{bib/RCU,bib/WFS,bib/hw,bib/os,bib/parallelsys,bib/patterns,bib/perfmeas,bib/refs,bib/syncrefs,bib/search,bib/swtools,bib/realtime,bib/TM,bib/standards,bib/memorymodel.bib}

### 2024-04-01 ootavacuous.tex
Add references to table of contents
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Self-review up through the end of Appendix C
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Although this does shunt additional complexity onto the term
-``semantic dependency'', it also enables us to separate cleanly
+``semantic dependency'', it also enables us to cleanly separate
 the interthread and intrathread portions of any given OOTA cycle.
@@
 That is, even when the potential for a dependency exists in the
-source code for a thread, there may be no dependency in the machine
+source code for a thread, there might be no dependency in the machine
 code produced by a compiler.
@@
 We leave aside the question of how the implementation would know this,
-given that \co{y} and \co{z} cannot be updated simultaneously
+given that \co{y} and \co{z} cannot be updated simultaneously\footnote{
+	At least not by any means within the confines of the standard.}
 and are subject to change at any time by
@@
 memory-load instructions.
-(We count read-modify-write instructions as both memory loads and
+(We count read-modify-write instructions as comprising both memory loads and
 memory stores.)
@@
 We will therefore use the terms ``compiler'' and ``implementation''
-somewhat interchangeably.
+interchangeably.
 
@@
 value zero.)
+% Are we sure that no CPUs do this?
 
@@
 variable are present.
+% Can we instead say that transforming the code might require that the
+% witness also be transformed?
 
@@
 but such reordering is in any case directly forbidden by the memory model
-(the four coherence rules in 6.9.2.2p14--19 \co{[intro.races]}).
+(the four coherence rules in 6.9.2.2p14--18 \co{[intro.races]}).
 The ``onto'' bullet point also largely reiterates restrictions that follow
@@
 program with the store present.
+% This is awkward because a reader might reasonably extend the "Acting
+% as if" all the way to the comma.  The best I could do was "It is not
+% possible to create an OOTA cycle by acting as if a redundant store
+% was omitted from the source code", which is admittedly not all that
+% great.
 
@@
 Future analyses might examine the possibility of ``flattening''
-optimizations that combine multiple threads into one.
+optimizations that combine multiple threads into one, but note
+that such optimizations are more difficult than they first might
+appear.
 
@@
 
-% @@@ Get citations
-
 Both the ARM and PowerPC architectural memory models permit reordering prior loads
 against subsequent stores, but actual tests on recent hardware fail to
-produce any evidence of any such reordering.
+produce any evidence of such reordering.
 Nevertheless, thus far neither ARM nor PowerPC hardware architects
 have been willing to strengthen their memory models so as to forbid
-such reordering.
-% @@@ Cite Grisenthwaite
+such reordering, despite a number requests to do
+so~\cite{RichardGrisenthwaite2023TightenRelaxedJustSayNo}.
 
@@
 a particular computation, and might react by restarting that computation.
-In both cases, it is only reasonable to react to an uncorrectable ECC
+On the other hand, it is only reasonable to react to an uncorrectable ECC
 error in a load from a read-only mapped file by re-reading that data
 from that file, then restarting that load instruction.
-In all cases, the kernel and the user application would need to continue
+In such cases, the kernel and the user application would need to continue
 execution, and would thus require a precise exception.
@@
 the fixed-up value.
+% I was tempted to add a sentence saying that although one could imaging
+% ECC on memory but not on caches, today's caches are a bit large for this
+% to be a winning strategy.  One might nevertheless consider omitting ECC on
+% the smallest caches closest to the CPU, but this is unlikely to allow LB.
 

### 2024-04-03 ootavacuous.tex
Apply Alan Stern's April Fool's feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 prohibition of OOTA mentioned above.
-{\it Contra\/} that article, we will use the term \emph{loose C++}
+Unlike that article, we will use the term \emph{loose C++}
 (rather than \emph{strict C++}, which seems too similar to
@@
 but such reordering is in any case directly forbidden by the memory model
-(the four coherence rules in 6.9.2.2p14--18 \co{[intro.races]}).
+(the four coherence rules discussed in 6.9.2.2p14--19 \co{[intro.races]}).
 The ``onto'' bullet point also largely reiterates restrictions that follow
@@
 
-Acting as if a redundant store was omitted from the source
-code cannot create an OOTA cycle, because any valid execution (OOTA or not)
+Omitting a redundant store cannot create an OOTA cycle,
+because any valid execution (OOTA or not)
 of the program with the first of a back-to-back pair of nonvolatile
@@
 Future analyses might examine the possibility of ``flattening''
-optimizations that combine multiple threads into one, but note
-that such optimizations are more difficult than they first might
-appear.
+optimizations that combine multiple threads into one.
+(Note that such optimizations are more difficult than they first might appear).
 
@@
 have been willing to strengthen their memory models so as to forbid
-such reordering, despite a number requests to do
+such reordering, despite a number of requests to do
 so~\cite{RichardGrisenthwaite2023TightenRelaxedJustSayNo}.

### 2024-04-03 ootavacuous.tex
Even more minor edits
These changes remove unnecessary words, improve grammar and consistency,
fix some ambiguities and errors, etc.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 cannot exhibit out-of-thin-air (OOTA) cycles.
-It follows that these implementations hardly need to take any special
-action to avoid OOTA.
+It follows that these implementations need to take almost no special
+actions to avoid OOTA.
 In fact, for many compilers the constraints boil down to a simple
-``Don't invent or duplicate atomic accesses'', which is probably
-something the compiler wouldn't do anyway.
+``Don't invent or duplicate atomic accesses'', which the compiler
+probably wouldn't do anyway.
 
@@
 And in these implementations, this prohibition is in fact enforced by TSO
-ordering in strongly ordered systems and by data dependency ordering in
+ordering in strongly ordered systems and by data-dependency ordering in
 weakly ordered systems.\footnote{
@@
 % concerns.
-The remainder of this document will concentrate on examples lacking UB.
+This document will concentrate on examples lacking UB.
 
@@
 We therefore dig more deeply into OOTA cycles in the light
-of these real-world constraints.
+of these constraints.
 
@@
 and \co{y}.
-The body of each thread contains a limited subset of C++ code
-(written in a slightly stilted manner to keep the load operations separate
-from the rest and because \co{herd7}'s knowledge of C++ is limited).
+The body of each thread contains C++ code,
+written in a slightly stilted manner to keep the load operations separate
+from the rest and because \co{herd7}'s knowledge of C++ is limited.
 
@@
 In order to avoid these rfi links we will substitute rfe
-for rf, defining an OOTA cycle---for now---as a cycle in sdep $\cup$ rfe,
-where rfe is the set of store-to-load links between different threads.
+for rf, defining an OOTA cycle---for now---as a cycle in sdep $\cup$ rfe.
 Any rfi links in a cycle can instead be interpreted as part of sdep.
@@
 \begin{verbatim}
-int foo(int a, int b)
-{
+int foo(int a, int b) {
    return a / b;
@@
 program is entirely determined by the values obtained by the various
-memory-load instructions.
-(We count read-modify-write instructions as comprising both memory loads and
-memory stores.)
+memory-load instructions.\footnote{
+	We regard read-modify-write instructions as consisting of both a
+	memory load and a memory store.}
 
@@
 Each instruction has a set of inputs and is the source of a set of outputs,
-some of which flow to the inputs for later instructions.
+some of which flow to the inputs of later instructions.
 The inputs determine what an instruction will do.
@@
 condition was definitely false.
-And, although no existing architectures do this, there is nothing
+And, although no architectures we are aware of do this, there is nothing
 in principle to prevent a CPU from committing a \co{multiply}
@@
 value zero.)
-% Are we sure that no CPUs do this?
 
@@
 thread, i.e., the rfe relation.
-As mentioned in Section~\ref{sec:Simple OOTA Cycle} above, a CPU does
+As mentioned in Section~\ref{sec:Simple OOTA Cycle}, a CPU does
 not make the value of a store instruction available for other CPUs to
@@
 This implies that if two different programs contain the same thread
-(i.e., the same source code for the functions in the thread), the
+(i.e., the same source code for the functions and objects in the thread), the
 machine code generated by the compiler for the thread will be the
@@
 Our third requirement for being a witness is that each load in the
-common portion either must obtain the same value in $H$ and $G$, or
+common period either must obtain the same value in $H$ and $G$, or
 must itself be one of the $R'_i$ loads, or must be ordered in $H$
@@
 corresponds to $W'$.
-If $W'$ is in the initial or common portion of $H$ this is no
+If $W'$ is in the initial or common period of $H$ this is no
 problem; we can take $X'$ to be $W'$ itself.
-But if $W'$ is in the divergent portion of $H$ then things aren't so
+But if $W'$ is in the divergent part of $H$ then things aren't so
 simple.
@@
 Let $y$ be the atomic object that $W'$ stores to, and suppose $W'$
-is the $N$th store to $y$ within the divergent portion of $H$.
-Then $X'$ will be the $N$th store to $y$ in the divergent portion
+is the $N$th store to $y$ within the divergent part of $H$.
+Then $X'$ will be the $N$th store to $y$ in the divergent part
 of $G$, if such a store exists.
@@
 
-First case: $W'$ lies in the initial portion of $H$.
-During the initial portion of the hardware executions, $H$ and $G$
+First case: $W'$ lies in the initial period of $H$.
+During the initial period of the hardware executions, $H$ and $G$
 behave identically and therefore $W'$ performs the same write in
@@
 
-Second case: $W'$ lies in the common portion of $H$.
+Second case: $W'$ lies in the common period of $H$.
 Since the action of $W'$ in $H$ is different from its action in $G$,
@@
 Then $W'$ depends on this load and so is ordered after it.
-And since the load must lie in the common portion of $H$, by the
+And since the load must lie in the common period of $H$, by the
 definition of semantic dependency it must either be one of the
@@
 
-Third case: $W'$ lies in the divergent portion of $H$.
+Third case: $W'$ lies in the divergent part of $H$.
 This happens when $W'$ comes after the conditional or computed branch
@@
 optimizations that combine multiple threads into one.
-(Note that such optimizations are more difficult than they first might appear).
+(Note that such optimizations are more difficult than they might first appear.)
 
@@
 overrides per-architecture defaults, some of which treat \co{char}
-as signed and others as \co{unsigned}.
+as signed and others as unsigned.
 This choice prevents a class of bugs, and also allows the kernel to make
@@
 The first is a genuine semantic dependency; the second might or might
-not be, depending on one's definition.
-A single-thread analysis would conclude that it is, because of the
+not be, depending on the implementation.
+A compiler that uses single-thread analysis and does not omit the load
+from \co{y} would conclude that it is, because of the
 possibility that the value loaded from \co{y} might be larger than 17.
@@
 do_something_with(r1);
-for (i = 0; i < limit i++)
+for (i = 0; i < limit; i++)
     y[i] += r1;
@@
 Note that passing \co{x} to \co{do_something_with()} is important
+when \co{x} is not atomic,
 because this allows the compiler to assume that inventing additional
@@
 
-But if \co{x} is a nonvolatile atomic, there are no worries about
+But when \co{x} is a nonvolatile atomic, there are no worries about
 data races.
 In theory, if the \co{do_something_with()} call was not present
-the compiler could still make the above transformation without
-worry, especially given that the value loaded from \co{x} is discarded in
+the compiler could still make the transformation above with no
+qualms, especially given that the value loaded from \co{x} is discarded in
 the case where \co{limit} is less than or equal to zero.
-Except that the data-race worry is replaced by a cache-miss worry;
+Except that now the data-race worry is replaced by a cache-miss worry;
 after all, optimizations are supposed to speed things up,
@@
 the \co{herd7} tool and also using manual analysis.
-Please note that differences in the semantics of these two languages
+Please note that differences in the semantics of the two languages
 result in changes in verdict for some litmus tests.

### 2024-04-03 ootavacuous.tex
Appendix D self-review
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-04-03 ootavacuous.tex
@@
 \item	Modified by debugger commands.
-\item	Modified by accesses to facilities such as \path{/dev/mem}.
+\item	Modified using facilities such as \path{/dev/mem}.
 \item	Modified by as-yet-unwritten dynamically linked libraries.
@@
 discussion in
-Section~\ref{sec:Semantic Dependencies and Source Code}.
+Section~\ref{sec:Semantic Dependencies and Source Code}
+on
+page~\pageref{sec:Semantic Dependencies and Source Code}.
 Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
@@
 clause (0 instead of 1).
-Note that an execution satisfying the \co{exists} clause is again
+Note that an execution satisfying the updated \co{exists} clause is again
 a OOTA cycle.
@@
 Section~\ref{sec:OOTA: rf versus rfe}
+on
+page~\pageref{sec:OOTA: rf versus rfe}
 in the form of litmus tests.
@@
 Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.
+on
+page~\pageref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.
 
@@
 or changing the store to a store-release.
+Hans Boehm has proposed this store-release alternative as his
+Interpretation~B'~\cite{HansBoehm2020UBalternatives}.
+% @@@ Do we want to add a disclaimer about completeness or lack thereof
+% @@@ of this solution?
 

### 2024-04-04 ootavacuous.tex
C has bracket on next line, litmus tests on same line
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{verbatim}
-int foo(int a, int b) {
+int foo(int a, int b)
+{
    return a / b;
@@
  2
- 3 void thread1() {
- 4   int r1 = x.load(memory_order_relaxed);
- 5   y.store(r1, memory_order_relaxed);
- 6 }
+ 3 void thread1()
+ 4 {
+ 5   int r1 = x.load(memory_order_relaxed);
+ 6   y.store(r1, memory_order_relaxed);
+ 7 }
 \end{verbatim}
@@
 Because both \co{x} and \co{y} are nonvolatile, a C++ compiler is
-free to assume that the value loaded by line~4 will be either some value
+free to assume that the value loaded by line~5 will be either some value
 stored to \co{x} or its initial value.
@@
 Therefore, the compiler is within its rights to substitute
-the constant zero for \co{r1} on line~5, eliminating the semantic
-dependency that would otherwise extend from line~4 to line~5.
+the constant zero for \co{r1} on line~6, eliminating the semantic
+dependency that would otherwise extend from line~5 to line~6.
 
@@
 Now the compiler is forbidden from assuming that the value loaded
-by line~4 has any relation to any stores to or initialization of \co{x}.
+by line~5 has any relation to any stores to or initialization of \co{x}.
 

### 2024-04-04 ootavacuous.tex
Updates to Appendix E and Acknowledgements
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 This test case is virtually the same as the Simple OOTA cycle
-discussed in Section~\ref{sec:Simple OOTA Cycle}, the only difference
-being the final values of the variables.
+discussed in
+Section~\ref{sec:Simple OOTA Cycle}
+on
+page~\pageref{sec:Simple OOTA Cycle},
+the only difference being the final values of the variables.
 
@@
 that prevent hardware OOTA.
+Mark Batty encouraged our work on new OOTA-related litmus tests.
 

### 2024-04-05 ootavacuous.tex
Fix typo
This typo has been around for a long time and I noticed it on several
occasions, but I haven't gotten around to fixing it until now.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Make myself the first author
This takes up Paul's generous offer to give first authorship to me.  It
also removes TBD (who doesn't seem to have contributed anything yet) and
changes the paper's date from February to April.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \author{
+Alan Stern\\{\small \email{stern@rowland.harvard.edu}} \and
 Paul E.~McKenney\\{\small \email{paulmck@kernel.org}} \and
-Alan Stern\\{\small \email{stern@rowland.harvard.edu}} \and
 Michael Wong\\{\small \email{fraggamuffin@gmail.com}} \and
-Maged Michael\\{\small \email{maged.michael@gmail.com}} \and
-The Indefatigible TBD
+Maged Michael\\{\small \email{maged.michael@gmail.com}}
 }
-\date{Feburary 16, 2024 (Post-Pre-Tokyo)}
+\date{April 5, 2024 (Post-Pre-Tokyo)}
 \maketitle{}
@@
 Note that an execution satisfying the updated \co{exists} clause is again
-a OOTA cycle.
+an OOTA cycle.
 

### 2024-05-04 ootavacuous.tex
Tokyo is done, so post-Tokyo it is!
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Upgrade from "D" to "P" for submission
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Add more compelling examples to "Inventing Atomic Loads"
Also add subsections.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{document}
-\title{D3064R0: How to Avoid OOTA Without Really Trying}
+\title{P3064R0: How to Avoid OOTA Without Really Trying}
 
@@
 }
-\date{April 5, 2024 (Post-Pre-Tokyo)}
+\date{April 5, 2024 (Post-Tokyo)}
 \maketitle{}
@@
 
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-no-invented-load.litmus @@
-\caption{No Invented Atomic Loads}
-\label{lst:No Invented Atomic Loads}
-\end{listing}
+\subsubsection{Illustrative Atomic-Load Invention}
+\label{app:Illustrative Atomic-Load Invention}
 
@@
 If both dependencies are semantic, the OOTA cycle cannot be realized.
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-no-invented-load.litmus @@
+\caption{No Invented Atomic Loads}
+\label{lst:No Invented Atomic Loads}
+\end{listing}
 
@@
 \end{listing}
+
+\subsubsection{Illustrative Atomic-Load Use Invention}
+\label{app:Illustrative Atomic-Load Use Invention}
 
@@
 
+However, before making such a choice, we should look at other situations.
+
+\subsubsection{Customary Non-Atomic-Load ``Invention''}
+\label{app:Customary Non-Atomic-Load Invention}
+
 There are situations where inventing loads from non-atomic objects
@@
 race when \co{limit} is less than or equal to zero.
+Hence the use of quotation marks in the section heading:
+Here the load is not actually invented, but instead fused with an
+earlier load with no synchronization between the two loads.
 
@@
 variables in order to avoid potentially expensive cache misses.
+
+In addition, compiler writers should carefully consider the issues
+raised by the following section:
+
+\subsubsection{Other Cautions for Atomic-Load Invention}
+\label{app:Other Cautions for Atomic-Load Invention}
+
+Consider the code below:
+\begin{quote}
+\begin{verbatim}
+int r0 = x.load(memory_order_relaxed);
+int r1 = r0 * r0 + 2 * r0 + 1;
+\end{verbatim}
+\end{quote}
+A developer might be forgiven for assuming that the value of \co{r1}
+is guaranteed to be a perfect square, at least for sufficiently small
+values of \co{x}.
+
+But if the compiler is permitted to invent atomic loads, there is no such
+guarantee, as can be seen from this transformed version of the above code:
+\begin{quote}
+\begin{verbatim}
+int r0 = x.load(memory_order_relaxed);
+int invented = x.load(memory_order_relaxed);
+int r1 = r0 * r0 + 2 * invented + 1;
+\end{verbatim}
+\end{quote}
+For example, if the load into \co{r0} happened when the value of \co{x}
+was zero and the load into \co{invented} happened when the value of \co{x}
+was ten, then the value of \co{r1} will be twenty, which is a perfectly
+legitimate integer, but not a perfect square.
+
+Similar issues can arise even when the compiler merely repurposes a
+value from an atomic load, as can be seen from this example:
+\begin{quote}
+\begin{verbatim}
+do_something(x.load(memory_order_relaxed));
+int r0 = x.load(memory_order_relaxed);
+int r1 = r0 * r0 + 2 * r0 + 1;
+\end{verbatim}
+\end{quote}
+This again ensures that \co{r1} will be a perfect square given sufficiently
+small values of \co{x}.
+But is the compiler allowed to transform this example into the following?
+\begin{quote}
+\begin{verbatim}
+repurposed = x.load(memory_order_relaxed);
+do_something(repurposed);
+int r0 = x.load(memory_order_relaxed);
+int r1 = r0 * r0 + 2 * repurposed + 1;
+\end{verbatim}
+\end{quote}
+Again, this can result in the value of \co{r1} being something other than
+a perfect square.
+%
+But it would be just fine if the compiler fully fused the two loads from
+\co{x}, for example, as follows:
+\begin{quote}
+\begin{verbatim}
+r0 = x.load(memory_order_relaxed);
+do_something(r0);
+int r0 = x.load(memory_order_relaxed);
+int r1 = r0 * r0 + 2 * r0 + 1;
+\end{verbatim}
+\end{quote}
+
+Similar observations apply to atomic loads from pointers.
+For example, consider a load from an atomic pointer and subsequent uses
+of the pointed-to fields:
+\begin{quote}
+\begin{verbatim}
+do_something(gp.load(memory_order_relaxed));
+p = gp.load(memory_order_relaxed);
+do_something_else(p->a, p->b,p->c);
+\end{verbatim}
+\end{quote}
+Users are within their rights to assume that the three arguments to
+\co{do_something_else()} will be taken from the same instance of the pointed-to
+structure, at least assuming that the lifetime of that structure extends
+across all of the above statements.
+However, further assuming that \co{do_something()} is free of
+synchronization, a compiler would be within its rights to fuse the two
+loads from \co{gp}, for example, as follows:
+\begin{quote}
+\begin{verbatim}
+p = gp.load(memory_order_relaxed);
+do_something(p);
+do_something_else(p->a, p->b,p->c);
+\end{verbatim}
+\end{quote}
+
+It is imperative that users be able to rely on arithmetic and pointer accesses.
+Therefore, compilers must refrain from inventing atomic loads and from repurposing
+their values, but compilers are permitted to fuse adjacent atomic loads from the
+same variable.
 

### 2024-05-05 ootavacuous.tex
Update paper number and date
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Appendix D.4 ("Inventing Atomic Loads"): Apply Alan Stern feedback
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-04-05 ootavacuous.tex
@@
 \begin{document}
-\title{D3064R0: How to Avoid OOTA Without Really Trying}
+\title{D3064R1: How to Avoid OOTA Without Really Trying}
 
@@
 }
-\date{April 5, 2024 (Post-Pre-Tokyo)}
+\date{May 4, 2024 (Pre-St. Louis)}
 \maketitle{}
@@
 race when \co{limit} is less than or equal to zero.
+Hence the use of quotation marks in the section heading:
+Here the load is not actually invented, but instead merged with an
+earlier load with no synchronization between the two loads.
 
@@
 variables in order to avoid potentially expensive cache misses.
+
+In addition, compiler writers should carefully consider the issues
+raised by the following section:
+
+\subsubsection{Other Cautions for Atomic-Load Invention}
+\label{app:Other Cautions for Atomic-Load Invention}
+
+Consider the code below:
+\begin{quote}
+\begin{verbatim}
+int r0 = x;
+int r1 = r0 * r0 + 2 * r0 + 1;
+\end{verbatim}
+\end{quote}
+A developer might be forgiven for assuming that the value of \co{r1}
+is guaranteed to be a perfect square, at least for sufficiently small
+values of \co{x}.
+
+But if the compiler is permitted to invent atomic loads, there is no such
+guarantee, as can be seen from this transformed version of the above code:
+\begin{quote}
+\begin{verbatim}
+int r0 = x;
+int invented = x;
+int r1 = r0 * r0 + 2 * invented + 1;
+\end{verbatim}
+\end{quote}
+For example, if the load into \co{r0} happened when the value of \co{x}
+was zero and the load into \co{invented} happened when the value of
+\co{x} was ten, then the value of \co{r1} will be twenty, which is not
+a perfect square.
+
+Similar issues can arise even when the compiler merely repurposes a
+value from an atomic load, as can be seen from this example:
+\begin{quote}
+\begin{verbatim}
+do_something(x);
+int r0 = x;
+int r1 = r0 * r0 + 2 * r0 + 1;
+\end{verbatim}
+\end{quote}
+Again, a developer might be forgiven for assuming that the value of
+\co{r1} is guaranteed to be a perfect square, at least for sufficiently
+small values of \co{x}.
+But is the compiler allowed to transform this example into the following?
+\begin{quote}
+\begin{verbatim}
+repurposed = x;
+do_something(repurposed);
+int r0 = x;
+int r1 = r0 * r0 + 2 * repurposed + 1;
+\end{verbatim}
+\end{quote}
+Again, this can result in the value of \co{r1} being something other than
+a perfect square.
+%
+But it would be just fine if the compiler fully merged the two loads from
+\co{x}, for example, as follows:
+\begin{quote}
+\begin{verbatim}
+int r0 = x;
+do_something(r0);
+int r1 = r0 * r0 + 2 * r0 + 1;
+\end{verbatim}
+\end{quote}
+
+Similar observations apply to atomic loads from pointers.
+For example, consider
+Listing~\ref{lst:Example Duplicated Load} on
+page~\pageref{lst:Example Duplicated Load},
+in which the developer expected line~19 to pass two fields from the
+same structure to \co{do_something_with()}, which is an entirely
+reasonable expectation.
+This expectation will not always be met if the compiler is permitted
+to invent atomic loads as shown by the commented-out line~15.
+\begin{listing}[tbp]
+\scriptsize
+\begin{verbatim}
+ 1 struct foo {
+ 2   int a;
+ 3   int b;
+ 4 };
+ 5 _Atomic struct foo *globalfoop;
+ 6
+ 7 void bar()
+ 8 {
+ 9   int a;
+10   int b;
+11   struct foo *fp;
+12
+13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
+14   do_something(fp);
+15   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
+16   a = fp->a;
+17   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
+18   b = fp->b;
+19   do_something_with(a, b);
+20 }
+\end{verbatim}
+\caption{Example Merging of Duplicated Loads}
+\label{lst:Example Merging of Duplicated Loads}
+\end{listing}
+But suppose that there was an earlier load from \co{globalfoop}, as
+shown on lines~13 and~14 in
+Listing~\ref{lst:Example Merging of Duplicated Loads}.
+Then, assuming that \co{do_something()} is synchronization-free,
+the second load on line~15 could be removed without violating user
+expectations.
+
+It is imperative that users be able to rely on arithmetic and pointer accesses.
+Therefore, compilers must refrain from inventing atomic loads and from repurposing
+their values, but compilers are permitted to merge adjacent atomic loads from the
+same variable.
+These restrictions correspond to our notion of quasi volatile defined in
+Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
 

### 2024-05-12 ootavacuous.tex
Invented loads: Justify restrictions based on abstract machine
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 This example is discussed more fully in
-Appendix~\ref{app:Inventing Atomic Loads}.
+Appendix~\ref{app:Inventing Atomic Loads},
+which also discusses other examples in which invented atomic loads lead
+to incorrect behavior, that is, behavior that is not exhibited by the
+abstract machine.\footnote{
+	See also the ``Invented Atomic Loads'' paragraph of
+	Section~\ref{sec:Constraints of the Standard}.}
 
@@
 \co{foo} structure, something that \co{do_something_with()} probably
-isn't prepared to deal with.
+isn't prepared to deal with and that represents incorrect behavior,
+that is, behavior that the abstract machine does not exhibit.
+See
+Appendix~\ref{app:Inventing Atomic Loads}
+for more details on this example and other examples of invented
+atomic loads.
 
@@
 Here the load is not actually invented, but instead merged with an
-earlier load with no synchronization between the two loads.
+earlier load such that there is no synchronization between the two loads.
 
@@
 In addition, compiler writers should carefully consider the issues
-raised by the following section:
+raised by the following section.
 
@@
 \end{quote}
-A developer might be forgiven for assuming that the value of \co{r1}
-is guaranteed to be a perfect square, at least for sufficiently small
-values of \co{x}.
-
-But if the compiler is permitted to invent atomic loads, there is no such
-guarantee, as can be seen from this transformed version of the above code:
+For sufficiently small values of \co{x}, the abstract machine is
+guaranteed to produce a perfect square in \co{r1}.
+
+But if the compiler is permitted to invent atomic loads, then this
+guarantee might not hold, as can be seen from this transformed version of
+the above code:
 \begin{quote}
@@
 was zero and the load into \co{invented} happened when the value of
-\co{x} was ten, then the value of \co{r1} will be twenty, which is not
+\co{x} was 10, then the value of \co{r1} will be 21, which is not
 a perfect square.
@@
 \end{quote}
-Again, a developer might be forgiven for assuming that the value of
-\co{r1} is guaranteed to be a perfect square, at least for sufficiently
-small values of \co{x}.
-But is the compiler allowed to transform this example into the following?
+Again, for sufficiently small values of \co{x}, the abstract machine is
+guaranteed to produce a perfect square in \co{r1}.
+So is the compiler allowed to transform this example into the following?
 \begin{quote}
@@
 \end{quote}
-Again, this can result in the value of \co{r1} being something other than
-a perfect square.
+This can result in the value of \co{r1} being something other than
+a perfect square, which is incorrect behavior.
 %
@@
 page~\pageref{lst:Example Duplicated Load},
-in which the developer expected line~19 to pass two fields from the
-same structure to \co{do_something_with()}, which is an entirely
-reasonable expectation.
-This expectation will not always be met if the compiler is permitted
-to invent atomic loads as shown by the commented-out line~15.
+in which the abstract machine guarantees that line~17 passes two
+fields from the same structure to \co{do_something_with()}.
+This guarantee will not always be met if the compiler is permitted to
+invent atomic loads such as that contained in the commented-out line~15.
 \begin{listing}[tbp]
@@
 Listing~\ref{lst:Example Merging of Duplicated Loads}.
-Then, assuming that \co{do_something()} is synchronization-free,
-the second load on line~15 could be removed without violating user
-expectations.
-
-It is imperative that users be able to rely on arithmetic and pointer accesses.
+Then, assuming that \co{do_something()} is synchronization-free, the
+second load on line~15 could be removed without resulting in incorrect
+behavior.
+
+Please note that these examples are not just accidents of the current state
+of the C++ abstract machine.
+After all, it is imperative that users be able to rely on the laws of
+arithmetic and on atomic accesses to pointers.
 Therefore, compilers must refrain from inventing atomic loads and from repurposing
-their values, but compilers are permitted to merge adjacent atomic loads from the
-same variable.
+their values.
+However, compilers are permitted to merge adjacent atomic loads from
+the same variable.
 These restrictions correspond to our notion of quasi volatile defined in

### 2024-05-13 ootavacuous.tex
Upgrade to P3064R1 for submission
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Section 4.3: Move forward references to a footnote
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{document}
-\title{D3064R1: How to Avoid OOTA Without Really Trying}
+\title{P3064R1: How to Avoid OOTA Without Really Trying}
 
@@
 with no dependence on the value loaded from \co{y}.
-This example is discussed more fully in
-Appendix~\ref{app:Inventing Atomic Loads},
-which also discusses other examples in which invented atomic loads lead
-to incorrect behavior, that is, behavior that is not exhibited by the
-abstract machine.\footnote{
-	See also the ``Invented Atomic Loads'' paragraph of
-	Section~\ref{sec:Constraints of the Standard}.}
 
@@
 Therefore we can rule out transformations like this one by insisting
-the compiler not invent (or duplicate) atomic loads.
+the compiler not invent (or duplicate) atomic loads.\footnote{
+	This issue is discussed more fully in
+	Appendix~\ref{app:Inventing Atomic Loads},
+	which also discusses other examples in which invented atomic
+	loads lead to incorrect behavior, that is, behavior that is not
+	exhibited by the abstract machine.
+	See also the ``Invented Atomic Loads'' paragraph of
+        Section~\ref{sec:Constraints of the Standard}.}
 In fact, we will require that atomic accesses be treated as

### 2024-05-13 ootavacuous.tex
May 12 feedback from Alan Stern
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Listing~\ref{lst:Example Duplicated Load}.
-Suppose that the load on line~13 were duplicated, for example, by
+Suppose that the load on line~13 were duplicated, for example, as if the
+compiler had uncommented line~15.
 uncommenting line~15.
@@
 
-But if the compiler is permitted to invent atomic loads, then this
-guarantee might not hold, as can be seen from this transformed version of
-the above code:
+But if the compiler is permitted to invent atomic loads, then compiled
+code might not respect this guarantee, thus (incorrectly) deviating
+from the compiled code.
+This issue is evident in this transformed version of the above code:
 \begin{quote}
@@
 Then, assuming that \co{do_something()} is synchronization-free, the
-second load on line~15 could be removed without resulting in incorrect
-behavior.
+second load on line~15 if that listing could be removed without resulting
+in incorrect behavior.
 

### 2024-05-13 ootavacuous.tex
Fixes to commits a186bdfbf6fe and 9a8e75ed86a7
Commits a186bdfbf6fe ("Section 4.3: Move forward references to a
footnote") and 9a8e75ed86a7 ("May 12 feedback from Alan Stern") could
stand a little improvement.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	Appendix~\ref{app:Inventing Atomic Loads},
-	which also discusses other examples in which invented atomic
-	loads lead to incorrect behavior, that is, behavior that is not
+	which includes other examples of invented atomic
+	loads leading to incorrect behavior, that is, behavior not
 	exhibited by the abstract machine.
@@
 compiler had uncommented line~15.
-uncommenting line~15.
 Then the values of \co{a} and \co{b} passed to line~17's call to
@@
 
-But if the compiler is permitted to invent atomic loads, then compiled
-code might not respect this guarantee, thus (incorrectly) deviating
-from the compiled code.
-This issue is evident in this transformed version of the above code:
+But if the compiler is permitted to invent atomic loads then the compiled
+code might not respect this guarantee, thus incorrectly deviating
+from the abstract machine.
+This issue is evident in a transformed version of the above code:
 \begin{quote}
@@
 Then, assuming that \co{do_something()} is synchronization-free, the
-second load on line~15 if that listing could be removed without resulting
+second load on line~15 of that listing could be removed without resulting
 in incorrect behavior.

### 2024-07-12 ootavacuous.tex
Add advice per Michael Wong feedback from St. Louis
Explicitly and prominently state that nothing need change for the
common-case implementations in the scope of this paper.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-We begin with a brief overview of the OOTA problem, followed
-by an equally brief summary of prior work in this area, and ending
-with a quick overview of the \co{herd7} tool that will be used to
-evaluate litmus tests.
+We begin with a summary of what needs to change and a brief overview of
+the OOTA problem, followed by an equally brief summary of prior work in
+this area, and ending with a quick overview of the \co{herd7} tool that
+will be used to evaluate litmus tests.
+
+\subsection{What Needs to Change?}
+\label{sec:What Needs to Change?}
+
+What changes are needed to prevent OOTA from C++ implementations using
+traditional compilers and traditional hardware, including CPUs and
+GPGPUs?
+
+\begin{itemize}
+\item	To the standard:
+	No changes.
+	The indefinite prohibition of out-of-thin-air values needs
+	to remain.
+\item	To implementations:
+	No changes.
+	Future implementations must avoid optimizations that invent,
+	duplicate, or repurpose atomic loads.
+	However, such optimizations must be avoided for other reasons,
+	including preserving basic laws of arithmetic
+	(see Appendix~\ref{app:Inventing Atomic Loads} for more detail).
+\item	To user code:
+	No changes.
+	Users must avoid undefined behavior in portable code, just as
+	is the case now.
+\end{itemize}
+
+In short, no changes are needed because current implementations already
+avoid OOTA cycles.
+The point of this paper is to give a semi-formal demonstration of
+why this is the case.
 
@@
 \end{quote}
-The standard's prohibition of OOTA is of course important, but those of us
-writing code in the real world must rely on actual C++ implementations.
+The standard's prohibition of OOTA is of course important, for example, to
+prevent misapplication of the as-if rule in oracular C++ implementations,
+but those of us writing code in the real world must rely on actual C++
+implementations.
 And in these implementations, this prohibition is in fact enforced by TSO
@@
 provided they treat all atomic objects as volatile.
+
+This work shows that avoiding OOTA cycles requires no changes to the
+standard, to current implementions, or to user practices for portable
+code.
+Future implementations must avoid optimizations that invent, duplicate,
+or repurpose atomic loads.
+More work is required to delineate any further restrictions on future
+implementations that might carry out optimizations based on global
+inter-thread analysis.
 

### 2024-07-12 ootavacuous.tex
Update paper number and add a history appendix
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{document}
-\title{P3064R1: How to Avoid OOTA Without Really Trying}
+\title{P3064R2: How to Avoid OOTA Without Really Trying}
 
@@
 }
-\date{May 4, 2024 (Pre-St. Louis)}
+\date{July 12, 2024 (Post-St. Louis)}
 \maketitle{}
@@
 
-% \section{History}
-% \label{sec:History}
-
 \clearpage
@@
 
-% Done as of December 18, 2023 %% as of March 18, 2023
+% Done as of December 18, 2023 %% as of March 18, 2024
 %% David Goldblatt
@@
 
+\section{History}
+\label{sec:History}
+
+P3064R2 incorporates feedback from the St.~Louis meeting, primarily
+incorporating clear statements that nothing need change to avoid OOTA.
+In other words, the title really means what it says.
+
+P3064R1 adds non-OOTA examples that show that inventing, duplicating,
+or repurposing atomic loads can violate basic laws of arithmetic.
+
 \bibliographystyle{plain}

### 2024-08-19 paper/litmus2table.sh
paper: Add litmus2table.sh prototype script
The idea is to save space displaying litmus tests.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -1,0 +1,73 @@
+#!/bin/bash
+#
+# litmus2table.sh whatever.litmus
+#
+# Outputs an abbreviated LaTeX table variant.
+
+awk -v amp='&' -v bs='\\' -v dq='"' '
+BEGIN {
+	procnum = -1;
+	stmtno = 0;
+	memorder["memory_order_relaxed"] = "rlx";
+	memorder["memory_order_acquire"] = "acq";
+	memorder["memory_order_release"] = "rel";
+}
+
+inproc != 0 && $0 != "}" {
+	stmt = $0;
+	moa = $0;
+	gsub(/	/, "~~", stmt);
+	if (moa ~ /memory_order_/) {
+		gsub(/^.*memory_order_/, "memory_order_", moa);
+		gsub(/\);/, "", moa);
+		moa = memorder[moa];
+	}
+	if (stmt ~ /= atomic_load_explicit\(/) {
+		gsub(/= atomic_load_explicit\(/, "=" bs "textsubscript{" moa "} ", stmt);
+		gsub(/, memory_order_[a-z_]*\);/, "", stmt);
+	} else if (stmt != /atomic_store_explicit\(/) {
+		gsub(/atomic_store_explicit\(/, "", stmt);
+		gsub(/, memory_order_[a-z_]*\);/, "", stmt);
+		gsub(/,/, " =" bs "textsubscript{" moa "}", stmt);
+	}
+	stmtlist[++stmtno][procnum] = stmt;
+	# print procnum ":" stmtno, stmt
+}
+
+/^P[0-9]*/ {
+	inproc = 1;
+	stmtno = 0;
+	procnum++;
+}
+
+/^}$/ {
+	inproc = 0;
+}
+
+END {
+	cols = "";
+	head = "";
+	for (i = 0; i <= procnum; i++) {
+		if (cols != "") {
+			cols = cols "|";
+			head = head " & ";
+		}
+		cols = cols "l";
+		head = head "P" i;
+	}
+	print bs "begin{tabular}{" cols "}"
+	print head " " bs bs;
+	print bs "hline"
+	for (i in stmtlist) {
+		curstmt = "";
+		tabs = "";
+		for (j in stmtlist[i]) {
+			if (tabs != "")
+				printf " &\n";
+			printf "%s", tabs stmtlist[i][j];
+			tabs = tabs "	";
+		}
+		print " " bs bs;
+	}
+	print bs "end{tabular}"
+}'

### 2024-08-19 paper/litmus2table.sh
paper/litmus2table.sh: Add line numbers
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -45,26 +45,21 @@ inproc != 0 && $0 != "}" {
 }
 
 END {
-	cols = "";
+	cols = "r";
 	head = "";
 	for (i = 0; i <= procnum; i++) {
-		if (cols != "") {
-			cols = cols "|";
-			head = head " & ";
-		}
-		cols = cols "l";
-		head = head "P" i;
+		cols = cols "|l";
+		head = head "& P" i " ";
 	}
 	print bs "begin{tabular}{" cols "}"
-	print head " " bs bs;
+	print head bs bs;
 	print bs "hline"
+	stmtno = 0;
 	for (i in stmtlist) {
-		curstmt = "";
 		tabs = "";
-		for (j in stmtlist[i]) {
-			if (tabs != "")
-				printf " &\n";
-			printf "%s", tabs stmtlist[i][j];
+		printf "%d.", ++stmtno;
+		for (j = 0; j <= procnum; j++) {
+			printf " &\n%s", tabs stmtlist[i][j];
 			tabs = tabs "	";
 		}
 		print " " bs bs;

### 2024-08-19 paper/c2latex.sh
paper: Make ootavacuous.tex build as-is
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -1,0 +1,1 @@
+../c2latex.sh

### 2024-08-19 paper/ootavacuous.sh

@@ -1,0 +1,96 @@
+#!/bin/bash
+#
+# Usage: sh ootavacuous.sh
+#
+# Commands:
+#
+# @@ DisplayLitmus path.litmus @@
+#	Use c2latex.sh to squish the tabs and number the lines.
+#	Note that "path.litmus" must not contain whitespace.
+# @@ RunLitmus path.litmus @@
+#	Use "herd7 -c11" to run the test, then use c2latex.sh
+#	to number the lines.  Note that "path.litmus" must not
+#	contain whitespace.
+# @@ DisplayRunLitmus path.litmus @@
+#	Why not both?  ;-)
+
+T="`mktemp -d ${TMPDIR-/tmp}/ootavacuous.sh.XXXXXX`"
+trap 'rm -rf $T' 0 2
+
+awk -v sq="'" -v dq='"' -v bs='\\' < ootavacuous.tex '
+BEGIN {
+	print "cat << " sq "___EOF___" sq;
+	print "% Automatically built, do not edit";
+	print "%";
+}
+
+/^@@ DisplayLitmus .* @@$/ {
+	path = $3;
+	print "{";
+	print bs "scriptsize";
+	print bs "begin{verbatim}";
+	print "___EOF___";
+	print "grep -v " sq "^[(/ ]" bs "*" sq " < " path " | ./c2latex.sh";
+	print "cat << " sq "___EOF___" sq;
+	print bs "end{verbatim}";
+	print "}";
+	next;
+}
+
+/^@@ DisplayRunLitmus .* @@$/ {
+	path = $3;
+	print "{";
+	print bs "scriptsize";
+	print bs "begin{verbatim}";
+	print "___EOF___";
+	print "grep -v " sq "^[(/ ]" bs "*" sq " < " path " | ./c2latex.sh";
+	print "cat << " sq "___EOF___" sq;
+	print bs "end{verbatim}";
+	print "";
+	print "Analysis by " dq bs "co{herd7 -c11 " path dq "}:";
+	print bs "begin{verbatim}";
+	print "___EOF___";
+	print "herd7 -c11 " path " | grep -v " sq "^$" sq " | ./c2latex.sh ";
+	print "cat << " sq "___EOF___" sq;
+	print bs "end{verbatim}";
+	print "}";
+	next;
+}
+
+/^@@ RunLitmus .* @@$/ {
+	path = $3;
+	print "{";
+	print bs "scriptsize";
[…]

### 2024-08-19 paper/ootavacuous.tex
paper: Remove Appendix E and references to it.
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-07-12 ootavacuous.tex
@@
 All OOTA workers owe a debt to the foundational work in the infamous
-``Causality Test Cases'',\footnote{
+``Causality Test Cases''.\footnote{
 	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
-a version of which may be found in
-Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}.
+% %%% Eventually cite P3064.
 
@@
 
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-4.litmus @@
+\caption{Causality Test Case 4}
+\label{lst:Causality Test Case 4}
+\end{listing}
+
 Later examples will usually combine the litmus test and the \co{herd7}
@@
 Other OOTA-related litmus tests may be found in
-Appendix~\ref{app:Illustrative Litmus Tests}
-and
-Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}.
+Appendix~\ref{app:Illustrative Litmus Tests}.
 
@@
 \medskip
+
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-2.litmus @@
+\caption{Causality Test Case 2}
+\label{lst:Causality Test Case 2}
+\end{listing}
 
@@
 always evaluate to a particular constant.
-JMM Causality Test Case 1 discussed in
-Appendix~\ref{app:Causality Test Case 1}
+JMM Causality Test Case 1 % %%% Cite P3064.
+% %%% discussed in
+% %%% Appendix~\ref{app:Causality Test Case 1}
 provides an example of such a global analysis.
@@
 Listing~\ref{lst:Causality Test Case 4}
-(discussed in
-Appendix~\ref{app:Causality Test Case 4})
+% %%% Cite P3064.
+% %%% (discussed in
+% %%% Appendix~\ref{app:Causality Test Case 4})
 from two threads to three.
@@
 
-\section{Litmus Tests from “Causality Test Cases"}
-\label{app:Litmus Tests from “Causality Test Cases"}
-
-These tests might be two decades old, but they are still quite
-relevant.\footnote{
-	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
-We have translated them from Java to C++, and we evaluate them using
-the \co{herd7} tool and also using manual analysis.
-Please note that differences in the semantics of the two languages
-result in changes in verdict for some litmus tests.
-
-\subsection{Causality Test Case 1}
-\label{app:Causality Test Case 1}
-
-Listing~\ref{lst:Causality Test Case 1}
-shows causality test case 1, for which the \co{r1 == r2 == 1} result
-is to be allowed.
-And indeed, this result does show up in the output from the \co{herd7} tool.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-1.litmus @@
-\caption{Causality Test Case 1}
-\label{lst:Causality Test Case 1}
-\end{listing}
-
-Compilers that can prove the value of \co{x} is always
-nonnegative can also determine that there is no semantic dependency between
-\co{P0()}'s load from \co{x} on line~8 and its store to \co{y} on line~10,
-that is, no matter which of the possible values is loaded from \co{x},
-the value 1 will always be stored to \co{y}.
-Relative to less-omniscient compilers, however,
-including those using single-thread analysis or treating atomic
-objects as volatile, there \emph{is} a semantic dependency.
-
-Even omniscient compilers will observe a semantic dependency from
-\co{P1()}'s load from \co{y} on line~14 to its store to \co{x} on line~15.
-
-Either way no OOTA cycle will occur, with omniscient compilers
-having no cycle at all because of the lack of an sdep link in \co{P0()},
-and other compilers prevented by hardware
-dependency ordering from realizing the cycle.
-
-\subsection{Causality Test Case 2}
-\label{app:Causality Test Case 2}
-
-Listing~\ref{lst:Causality Test Case 2}
-shows causality test case 2, for which the \co{r1 == r2 == r3 == 1} result
-is to be allowed.
-This result appears in the \co{herd7} output.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-2.litmus @@
-\caption{Causality Test Case 2}
-\label{lst:Causality Test Case 2}
-\end{listing}
-
-Because \co{P0()}'s loads from \co{x} on lines~8 and~9 are unordered,
-a C++ compiler that treats atomics as quasi volatile
-can act as if the source code loaded from \co{x} only
-once and assigned the returned value to both \co{r1} and \co{r2},
-merging the loads.\footnote{
-	But not in C, where relaxed atomic operations are volatile,
-	and thus are observable behavior not subject to merging.}
-The compiler could then prove that the condition on line~10
-would always true, and act as if the source code had
-unconditionally executed the relaxed store to \co{y} on line~11
-independent of the value loaded from \co{x}.
-Without this dependency there would then be no OOTA cycle.
-
-For compilers that do not merge the two loads, the hardware
-dependency ordering arising from the conditional test on line~10
-will prevent the realization of an OOTA cycle.
-
-\subsection{Causality Test Case 3}
-\label{app:Causality Test Case 3}
-
-Listing~\ref{lst:Causality Test Case 3}
-shows causality test case 3, for which the \co{r1 == r2 == r3 == 1} result
-is to be allowed.
-The \co{herd7} tool reports this result.
-
-The analysis from the previous section applies here.
-Although the addition of \co{P2()} allows sequentially consistent executions
-in which \co{P0()}'s two loads from \co{x} return different values,
-it does not change the fact that those two loads can be merged.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-3.litmus @@
-\caption{Causality Test Case 3}
-\label{lst:Causality Test Case 3}
-\end{listing}
-
-\subsection{Causality Test Case 4}
-\label{app:Causality Test Case 4}
-
-Listing~\ref{lst:Causality Test Case 4}
-shows causality test case 4, for which the \co{r1 == r2 == 1} result
-is to be forbidden.
-The \co{herd7} tool finds this OOTA result;
-the \co{S8} values given on line~3 of its output indicate
-that in the execution it describes, the variables \co{r1} and~\co{r2} were never
-assigned any specific value, so they could just as well be equal to
-one as to anything else.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-4.litmus @@
-\caption{Causality Test Case 4}
-\label{lst:Causality Test Case 4}
-\end{listing}
-
-This test case is virtually the same as the Simple OOTA cycle
-discussed in
-Section~\ref{sec:Simple OOTA Cycle}
-on
-page~\pageref{sec:Simple OOTA Cycle},
-the only difference being the final values of the variables.
-
-\subsection{Causality Test Case 5}
-\label{app:Causality Test Case 5}
-
-Listing~\ref{lst:Causality Test Case 5}
-shows causality test case 5, for which the \co{r1 == r2 == 1} and \co{r3 == 0}
-result is to be forbidden.
-The \co{herd7} tool reports this OOTA result in line~3 of its output.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-5.litmus @@
-\caption{Causality Test Case 5}
-\label{lst:Causality Test Case 5}
-\end{listing}
-
-This is the same as causality test case~4 with \co{P2()} and \co{P3()}
-added, but those routines play no role in the OOTA cycle.
-\co{P2()} has no loads and so cannot participate in an OOTA cycle in
-any case.
-
-\subsection{Causality Test Case 6}
-\label{app:Causality Test Case 6}
-
-Listing~\ref{lst:Causality Test Case 6}
-shows causality test case 6, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool reports this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-6.litmus @@
-\caption{Causality Test Case 6}
-\label{lst:Causality Test Case 6}
-\end{listing}
-
-A compiler that can prove \co{y} is always either
-zero or one can also prove that there is no semantic dependency
-from \co{P1()}'s load on line~14 to its stores on lines~16 and~19.
-Such a compiler could act as if lines~15--20 had been replaced
-by a single store to \co{x}, showing that the dependency isn't
-semantic.
-
-Less-omniscient compilers will preserve the dependency, and then
-hardware dependency ordering will prevent the OOTA cycle from
-being realized.
-
-\subsection{Causality Test Case 7}
-\label{app:Causality Test Case 7}
-
-Listing~\ref{lst:Causality Test Case 7}
-shows causality test case 7, for which the \co{r1 == r2 == r3 == 1}
-result is to be allowed.
-The \co{herd7} tool finds this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-7.litmus @@
-\caption{Causality Test Case 7}
-\label{lst:Causality Test Case 7}
-\end{listing}
-
-This test case has everything
-to do with reordering and nothing to do with OOTA cycles.
-The \co{r1 == r2 == r3 == 1} outcome happens when lines~17, 10, 11,
-15, 16 and~9 execute in that order.
-Note that this execution order respects the semantic dependencies in
-lines~10--11 and~15--16, and thus is not ruled out by hardware
-dependency ordering.
-It could have been ruled out if some of the accesses were acquire or
-release, but here they are all relaxed.
-
-\subsection{Causality Test Case 8}
-\label{app:Causality Test Case 8}
-
-Listing~\ref{lst:Causality Test Case 8}
-shows causality test case 8, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not report this result, but it does if
-\co{r2} is initialized to one before being set to
-\co{1 + r1 * r1 - r1}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-8.litmus @@
-\caption{Causality Test Case 8}
-\label{lst:Causality Test Case 8}
-\end{listing}
-
-Omniscient compilers might note that the computation on line~9
-will have the value one when the load from \co{x} returns either
-zero or one, which are the only values \co{x} and \co{y} can take.
-Such a compiler could then realize that line~9 always sets \co{r2} to one,
-independent of the value loaded in line~8.
-Then there would be no semantic dependency between \co{P0()}'s load
-and store, and thus no OOTA cycle relative to this compiler.
-
-A less-capable compiler would leave the dependency in place,
-and the resulting hardware dependency ordering would prevent
-the OOTA cycle from being realized.
-
-\subsection{Causality Test Case 9}
-\label{app:Causality Test Case 9}
-
-Listing~\ref{lst:Causality Test Case 9}
-shows causality test case 9, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not report this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-9.litmus @@
-\caption{Causality Test Case 9}
-\label{lst:Causality Test Case 9}
-\end{listing}
-
-This is the same as causality test case~8 but with \co{P2()} added;
-the store to \co{x} in \co{P2} can interfere with the formation of the
-OOTA cycle.
-
-Unlike in causality test case~8, the computation on line~9 must be
-carried out because now the load on line~8 might return two.
-Consequently \co{P0} and~\co{P1} both have semantic dependencies,
-and hardware dependency ordering will prevent the OOTA cycle from
-being realized.
-
-\subsection{Causality Test Case 9a}
-\label{app:Causality Test Case 9a}
-
-Listing~\ref{lst:Causality Test Case 9a}
-shows causality test case 9a, for which the \co{r1 == r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not report this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-9a.litmus @@
-\caption{Causality Test Case 9a}
-\label{lst:Causality Test Case 9a}
-\end{listing}
-
-This is the same as causality test case~9 except that here \co{P2()}
-sets \co{x} to zero instead of two.
-As as result the computation on line~9 is once again independent of
-the value loaded on line~8, which makes the analysis nearly the same
-as that of causality test case~8.
-
-\subsection{Causality Test Case 10}
-\label{app:Causality Test Case 10}
-
-Listing~\ref{lst:Causality Test Case 10}
-shows causality test case 10, for which the \co{r1 == r2 == 1} and \co{r3 == 0}
-result is to be forbidden.
-The \co{herd7} tool reports this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-10.litmus @@
-\caption{Causality Test Case 10}
-\label{lst:Causality Test Case 10}
-\end{listing}
-
-\co{P0()} and \co{P1()} have straightforward
-semantic dependencies between their respective loads and stores, and
-so hardware dependency ordering will prevent an OOTA cycle from being
-realized.
-\co{P2()} has no semantic dependency, and \co{P3()}'s store does not
-occur in this execution (i.e., the one described by the \co{exists} clause)
-because \co{r3} is zero.
-Thus neither of them contributes to an OOTA cycle.
-
-\subsection{Causality Test Case 11}
-\label{app:Causality Test Case 11}
-
-Listing~\ref{lst:Causality Test Case 11}
-shows causality test case 11, for which the \co{r1 == r2 == r3 == r4 == 1}
-result is to be allowed.
-The \co{herd7} tool reports this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-11.litmus @@
-\caption{Causality Test Case 11}
-\label{lst:Causality Test Case 11}
-\end{listing}
-
-Like causality test case~7, this test case has nothing to do with OOTA cycles.
-The result may be obtained by executing lines~20, 12, 13, 18, 19, 10, 11,
-and~17 in that order.
-
-\subsection{Causality Test Case 12}
-\label{app:Causality Test Case 12}
-
-Causality Test Case 12 in
-Listing~\ref{lst:Causality Test Case 12}
-uses arrays, which are not yet supported by the \co{herd7} tool.
-
-\begin{listing}[tbp]
-@@ DisplayLitmus litmus/oota-causality-12.litmus @@
-\caption{Causality Test Case 12}
-\label{lst:Causality Test Case 12}
-\end{listing}
-
-An omniscient compiler could note that only \co{P0()} accesses
-array \co{a[]} and act as if the code was simpler, but there would
-still be semantic dependencies between \co{P0()}'s and \co{P1()}s
-relaxed atomic loads and stores involving \co{x} and \co{y}.
-The resulting hardware dependency ordering will prevent an OOTA cycle
-from being realized, whether the compiler is omniscient or not.
-
-\subsection{Causality Test Case 13}
-\label{app:Causality Test Case 13}
-
-Listing~\ref{lst:Causality Test Case 13}
-shows causality test case 13, for which the \co{r1 == r2 == 1}
-result is to be forbidden.
-The \co{herd7} tool reports this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-13.litmus @@
-\caption{Causality Test Case 13}
-\label{lst:Causality Test Case 13}
-\end{listing}
-
-This is the same as causality test case~10 without the confounding
-influence of \co{P2()} and~\co{P3()}.
-
-\subsection{Causality Test Case 14}
-\label{app:Causality Test Case 14}
-
-Listing~\ref{lst:Causality Test Case 14}
-shows causality test case 14, for which the \co{r1 == r3 == 1} and \co{r2 == 0}
-result is to be forbidden.
-The \co{herd7} tool reports this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-14.litmus @@
-\caption{Causality Test Case 14}
-\label{lst:Causality Test Case 14}
-\end{listing}
-
-In the execution described by the \co{exists} clause,
-\co{P0()} has a semantic dependency from line~9 to
-line~13 and \co{P1()} has a semantic dependency from line~18 to
-line~21.
-(There's also a semantic dependency from line~17 to line~21 but it
-doesn't enter into this OOTA cycle, which involves only \co{x} and~\co{y}.)
-The \co{memory_order_seq_cst} specification on line~17 has no important effect;
-hardware dependency ordering will prevent the cycle from being realized.
-
-\subsection{Causality Test Case 15}
-\label{app:Causality Test Case 15}
-
-Listing~\ref{lst:Causality Test Case 15}
-shows causality test case 15, for which the \co{r0 == r1 == r3 == 1}
-and \co{r2 == 0}
-result is to be forbidden.
-The \co{herd7} tool reports this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-15.litmus @@
-\caption{Causality Test Case 15}
-\label{lst:Causality Test Case 15}
-\end{listing}
-
-The execution described by the \co{exists} clause
-has an OOTA cycle involving \co{u} and~\co{v}.
-The dependency in \co{P0()} from line~13 to line~19 and the dependency
-in \co{P1()} from line~24 to line~27 are both semantic.
-The resulting hardware dependency ordering prevents the cycle from
-being realized.
-
-Having no load, \co{P2()} cannot participate in an OOTA cycle.
-
-\subsection{Causality Test Case 16}
-\label{app:Causality Test Case 16}
-
-Listing~\ref{lst:Causality Test Case 16}
-shows causality test case 16, for which the \co{r1 == 2} and \co{r2 == 1}
-result is to be allowed.
-The \co{herd7} tool does not report this result, which is to be
-expected because it would violate the C++ memory model's read-write
-coherence rule regarding the modification order of an atomic object
-(6.9.2.2p17 \co{[intro.races]}).
-Because no abstract execution can obtain \co{r1 == 2} and \co{r2 == 1}, no
-valid realization will either.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-16.litmus @@
-\caption{Causality Test Case 16}
-\label{lst:Causality Test Case 16}
-\end{listing}
-
-\subsection{Causality Test Case 17}
-\label{app:Causality Test Case 17}
-
-Listing~\ref{lst:Causality Test Case 17}
-shows causality test case 17, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result, although it does
-report a different OOTA cycle involving lines 11, 12, 16, and~17
-(the \co{S17} value on line~3 of the \co{herd7} output).
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-17.litmus @@
-\caption{Causality Test Case 17}
-\label{lst:Causality Test Case 17}
-\end{listing}
-
-A compiler that treats atomics as quasi volatile could reason as follows:
-Suppose that \co{P0()}'s load from \co{x} on line~8 returns the value 42.
-In that case, the second load on line~11 can be merged with the first,
-so that it also returns the value 42.
-On the other hand, if the load on line~8 returns some other value then
-line~10 will store the value 42 to \co{x}, and then the load on line~11
-can be omitted in favor of using the value 42 stored by line~10.
-
-Either way, the value of \co{r1} will be 42.
-Therefore the loads from \co{x} cannot affect the store on line~17,
-so there is no semantic dependency.
-This allows the cycle to be realized,
-but because the dependency in \co{P0()} is not semantic,
-the cycle is not OOTA.
-
-\subsection{Causality Test Case 18}
-\label{app:Causality Test Case 18}
-
-Listing~\ref{lst:Causality Test Case 18}
-shows causality test case 18, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-18.litmus @@
-\caption{Causality Test Case 18}
-\label{lst:Causality Test Case 18}
-\end{listing}
-
-This is the same as causality test case~17, even to the alternate OOTA
-cycle reported by \co{herd7}, except that line~9 tests
-\co{r3} for equality to zero rather than non-equality to 42.
-Perhaps surprisingly, this makes a difference.
-
-A compiler that treats atomic objects as volatile or uses
-single-thread analysis cannot assume that \co{x} will necessarily be
-equal to 42 following lines~9 and~10; it might have some other value
-such as 1.
-Therefore the compiler cannot conclude that \co{r1} will always be 42
-and must leave the semantic dependency between lines~11 and~12 intact.
-Of course, the resulting OOTA cycle will not be realized, because of
-hardware dependency ordering.
-
-\subsection{Causality Test Case 19}
-\label{app:Causality Test Case 19}
-
-Listing~\ref{lst:Causality Test Case 19}
-shows causality test case 19, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result but it does report the
-OOTA cycle in line~3 of its output.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-19.litmus @@
-\caption{Causality Test Case 19}
-\label{lst:Causality Test Case 19}
-\end{listing}
-
-This is nearly the same as the Simple OOTA cycle and causality test
-case~4, with the addition of \co{P2()}.
-Just as in those examples, the OOTA cycle cannot be realized by
-compilers that treat atomic objects as volatile or that do
-single-thread analysis and treat atomic objects as quasi volatile.
-
-\subsection{Causality Test Case 20}
-\label{app:Causality Test Case 20}
-
-Listing~\ref{lst:Causality Test Case 20}
-shows causality test case 20, for which the \co{r1 == r2 == r3 == 42}
-result is to be allowed.
-The \co{herd7} tool does not report this result.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-20.litmus @@
-\caption{Causality Test Case 20}
-\label{lst:Causality Test Case 20}
-\end{listing}
-
-This is the same as causality test case~19, even to the OOTA cycle
-reported by \co{herd7}, except that line~19 tests
-\co{r3} for equality to zero rather than non-equality to 42.
-The same analysis applies.
-
-\clearpage
-
 \section{Acknowledgments}

### 2024-08-20 paper/litmus2table.sh
litmus2table.sh: Print non-zero initializations
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -6,11 +6,25 @@
 
 awk -v amp='&' -v bs='\\' -v dq='"' '
 BEGIN {
+	ininit = 0;
+	inproc = 0;
 	procnum = -1;
 	stmtno = 0;
 	memorder["memory_order_relaxed"] = "rlx";
 	memorder["memory_order_acquire"] = "acq";
 	memorder["memory_order_release"] = "rel";
+}
+
+ininit != 0 && $0 ~ /= *[[1-9]/ {
+	curinit = $0;
+	gsub(/\[/, "", curinit);
+	gsub(/]/, "", curinit);
+	gsub(/[	 ]/, "", curinit);
+	initlist = initlist " " curinit;
+}
+
+procnum == -1 && /^{$/ {
+	ininit = 1;
 }
 
 inproc != 0 && $0 != "}" {
@@ -41,10 +55,13 @@ inproc != 0 && $0 != "}" {
 }
 
 /^}$/ {
+	ininit = 0;
 	inproc = 0;
 }
 
 END {
+	if (initlist != "")
+		print "Non-zero initialization:" initlist;
 	cols = "r";
 	head = "";
 	for (i = 0; i <= procnum; i++) {

### 2024-08-20 paper/ootavacuous.tex
ootavacuous.tex: Drop troublesome atomic-increment litmus tests
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/inc.litmus @@
-\caption{Only Atomic Increment}
-\label{lst:Only Atomic Increment}
-\end{listing}
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/inc-range.litmus @@
-\caption{Only Atomic Increment, Extended}
-\label{lst:Only Atomic Increment, Extended}
-\end{listing}
-
-Let us drop the single-thread-analysis constraint for
-the moment, and suppose that the compiler is able to prove
-that the only modification of a given atomic variable is to atomically
-increment it, as is the case for \co{y} in
-Listing~\ref{lst:Only Atomic Increment}.
-Is it okay for a C++ compiler to transform this program into
-Listing~\ref{lst:Only Atomic Increment, Extended}?
-An argument in favor is that the value of \co{y} must have
-passed through the value 42 if \co{r0} and \co{r1}
-bracket that value.
-Arguments against might cite the added overhead of the invented load,
-or the possibility that the check for the value 42 is intended for statistical
-sampling.
-
 There are a number of possible reactions to these situations:

### 2024-08-20 paper/litmus2table.sh
litmus2table.sh: Handle statement blocks and maybe =rlx expressions
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -33,13 +33,14 @@ inproc != 0 && $0 != "}" {
 	gsub(/	/, "~~", stmt);
 	if (moa ~ /memory_order_/) {
 		gsub(/^.*memory_order_/, "memory_order_", moa);
-		gsub(/\);/, "", moa);
+		gsub(/\).*/, "", moa);
 		moa = memorder[moa];
 	}
 	if (stmt ~ /= atomic_load_explicit\(/) {
 		gsub(/= atomic_load_explicit\(/, "=" bs "textsubscript{" moa "} ", stmt);
-		gsub(/, memory_order_[a-z_]*\);/, "", stmt);
-	} else if (stmt != /atomic_store_explicit\(/) {
+		if (gsub(/, memory_order_[a-z_]*\);/, "", stmt) == 0)
+			gsub(/, memory_order_[a-z_]*\)/, "", stmt);
+	} else if (stmt ~ /atomic_store_explicit\(/) {
 		gsub(/atomic_store_explicit\(/, "", stmt);
 		gsub(/, memory_order_[a-z_]*\);/, "", stmt);
 		gsub(/,/, " =" bs "textsubscript{" moa "}", stmt);
@@ -66,7 +67,7 @@ END {
 	head = "";
 	for (i = 0; i <= procnum; i++) {
 		cols = cols "|l";
-		head = head "& P" i " ";
+		head = head "& " bs "multicolumn{1}{c}{P" i "} ";
 	}
 	print bs "begin{tabular}{" cols "}"
 	print head bs bs;
@@ -76,7 +77,7 @@ END {
 		tabs = "";
 		printf "%d.", ++stmtno;
 		for (j = 0; j <= procnum; j++) {
-			printf " &\n%s", tabs stmtlist[i][j];
+			printf " &\n%s", tabs bs "texttt{" stmtlist[i][j] "}";
 			tabs = tabs "	";
 		}
 		print " " bs bs;

### 2024-08-20 paper/ootavacuous.sh
ootavacuous.sh: Print litmus tests in tabular form
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -17,7 +17,7 @@
 T="`mktemp -d ${TMPDIR-/tmp}/ootavacuous.sh.XXXXXX`"
 trap 'rm -rf $T' 0 2
 
-awk -v sq="'" -v dq='"' -v bs='\\' < ootavacuous.tex '
+awk -v sq="'" -v dq='"' -v bs='\\' -v amp='&' < ootavacuous.tex '
 BEGIN {
 	print "cat << " sq "___EOF___" sq;
 	print "% Automatically built, do not edit";
@@ -26,27 +26,26 @@ BEGIN {
 
 /^@@ DisplayLitmus .* @@$/ {
 	path = $3;
+	print "% " path;
 	print "{";
 	print bs "scriptsize";
-	print bs "begin{verbatim}";
 	print "___EOF___";
-	print "grep -v " sq "^[(/ ]" bs "*" sq " < " path " | ./c2latex.sh";
+	print "sed -e " sq "s/" amp "/" bs bs amp "/g" sq " < " path " | sed -e " sq "s/	}/	" bs bs "}/" sq " | sed -e " sq "s/{/" bs bs "{/" sq "| ./litmus2table.sh";
 	print "cat << " sq "___EOF___" sq;
-	print bs "end{verbatim}";
 	print "}";
 	next;
 }
 
 /^@@ DisplayRunLitmus .* @@$/ {
 	path = $3;
+	print "% " path;
 	print "{";
 	print bs "scriptsize";
-	print bs "begin{verbatim}";
 	print "___EOF___";
-	print "grep -v " sq "^[(/ ]" bs "*" sq " < " path " | ./c2latex.sh";
+	print "sed -e " sq "s/" amp "/" bs bs amp "/g" sq " < " path " | sed -e " sq "s/	}/	" bs bs "}/" sq " | sed -e " sq "s/{/" bs bs "{/" sq "| ./litmus2table.sh";
 	print "cat << " sq "___EOF___" sq;
-	print bs "end{verbatim}";
 	print "";
+	print bs "vspace{0.1in}";
 	print "Analysis by " dq bs "co{herd7 -c11 " path dq "}:";
 	print bs "begin{verbatim}";
 	print "___EOF___";

### 2024-08-20 paper/litmus2table.sh
Move "sed" statements from ootavacuous.sh to litmus2table.sh
Way simpler!!!

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -4,6 +4,7 @@
 #
 # Outputs an abbreviated LaTeX table variant.
 
+sed -e 's/&/\\&/' | sed -e 's/	}/	\\}/' | sed -e 's/{/\\{/' |
 awk -v amp='&' -v bs='\\' -v dq='"' '
 BEGIN {
 	ininit = 0;
@@ -58,6 +59,11 @@ inproc != 0 && $0 != "}" {
 /^}$/ {
 	ininit = 0;
 	inproc = 0;
+}
+
+/^exists/ {
+	existsclause = $0;
+	gsub(/^exists[ 	]*/, "", existsclause);
 }
 
 END {

### 2024-08-20 paper/ootavacuous.sh

@@ -30,7 +30,7 @@ BEGIN {
 	print "{";
 	print bs "scriptsize";
 	print "___EOF___";
-	print "sed -e " sq "s/" amp "/" bs bs amp "/g" sq " < " path " | sed -e " sq "s/	}/	" bs bs "}/" sq " | sed -e " sq "s/{/" bs bs "{/" sq "| ./litmus2table.sh";
+	print "./litmus2table.sh < " path;
 	print "cat << " sq "___EOF___" sq;
 	print "}";
 	next;
@@ -42,7 +42,7 @@ BEGIN {
 	print "{";
 	print bs "scriptsize";
 	print "___EOF___";
-	print "sed -e " sq "s/" amp "/" bs bs amp "/g" sq " < " path " | sed -e " sq "s/	}/	" bs bs "}/" sq " | sed -e " sq "s/{/" bs bs "{/" sq "| ./litmus2table.sh";
+	print "./litmus2table.sh < " path;
 	print "cat << " sq "___EOF___" sq;
 	print "";
 	print bs "vspace{0.1in}";

### 2024-08-20 paper/litmus2table.sh
litmus2table.sh: Display exists clause
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -5,6 +5,7 @@
 # Outputs an abbreviated LaTeX table variant.
 
 sed -e 's/&/\\&/' | sed -e 's/	}/	\\}/' | sed -e 's/{/\\{/' |
+sed -e 's,\\/,$\\vee$,' | sed -e 's,/\\,$\\wedge$,' |
 awk -v amp='&' -v bs='\\' -v dq='"' '
 BEGIN {
 	ininit = 0;
@@ -88,5 +89,10 @@ END {
 		}
 		print " " bs bs;
 	}
-	print bs "end{tabular}"
+	print bs "end{tabular}";
+	if (existsclause != "") {
+		print "";
+		print bs "vspace{0.1in}";
+		print "Condition: " existsclause;
+	}
 }'

### 2024-08-21 paper/herd2result.sh
Print only result from herd7 runs next to condition
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -1,0 +1,5 @@
+#!/bin/sh
+#
+# Extracts the "Observation" result from herd7 output.
+
+grep '^Observation' | awk '{ print $3 " satisfied."; }' | tr 'A-Z' 'a-z'

### 2024-08-21 paper/litmus2table.sh

@@ -93,6 +93,6 @@ END {
 	if (existsclause != "") {
 		print "";
 		print bs "vspace{0.1in}";
-		print "Condition: " existsclause;
+		print "Condition " existsclause;
 	}
 }'

### 2024-08-21 paper/ootavacuous.sh

@@ -43,15 +43,13 @@ BEGIN {
 	print bs "scriptsize";
 	print "___EOF___";
 	print "./litmus2table.sh < " path;
+	# print "cat << " sq "___EOF___" sq;
+	# print "";
+	# print bs "vspace{0.1in}";
+	# print "Analysis by " dq bs "co{herd7 -c11 " path dq "}:";
+	# print "___EOF___";
+	print "herd7 -c11 " path " | ./herd2result.sh ";
 	print "cat << " sq "___EOF___" sq;
-	print "";
-	print bs "vspace{0.1in}";
-	print "Analysis by " dq bs "co{herd7 -c11 " path dq "}:";
-	print bs "begin{verbatim}";
-	print "___EOF___";
-	print "herd7 -c11 " path " | grep -v " sq "^$" sq " | ./c2latex.sh ";
-	print "cat << " sq "___EOF___" sq;
-	print bs "end{verbatim}";
 	print "}";
 	next;
 }

### 2024-08-22 paper/ootavacuous.tex
ootavacuous: Remove "User Influence Over Language Semantics" appendix
Instead, tell the reader how to get this information from the Linux
kernel.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 which causes it to treat variables of type \co{char} as unsigned.
-More examples of user control over language semantics are given in
-Appendix~\ref{app:User Influence Over Language Semantics}
-and by the discussion
+Many more examples of user control over language semantics
+may be found through use of the command \co{make V=1}
+in a Linux-kernel source tree and by the discussion
 in~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.
@@
 
-\section{User Influence Over Language Semantics}
-\label{app:User Influence Over Language Semantics}
-
-As noted earlier, the exact definition of a computer language is subject
-to some debate, with standards, implementations, and users all having
-some degree of
-influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.
-It is natural to dismiss user influence when compared to the text of
-standards or the code in implementations, but both of these are subject to
-change and do change over time.
-An especially easy way for users to influence the implementation is by
-means of command-line flags and switch settings.
-As an example,
-consider the following command line used to compile the Linux kernel's
-\co{kernel/rcu/tree.c} C-language source file:
-
-\begin{quote}
-{
-	\scriptsize
-	\texttt{\noindent
-	gcc -Wp,-MMD,kernel/rcu/.tree.o.d -nostdinc
-	-I./arch/x86/include \\
-	-I./arch/x86/include/generated  -I./include
-	-I./arch/x86/include/uapi \\
-	-I./arch/x86/include/generated/uapi
-	-I./include/uapi \\
-	-I./include/generated/uapi \\
-	-include ./include/linux/compiler-version.h \\
-	-include ./include/linux/kconfig.h \\
-	-include ./include/linux/compiler\_types.h \\
-	-D\_\_KERNEL\_\_
-	-fmacro-prefix-map=./=
-	-Werror -std=gnu11 -fshort-wchar \\
-	-funsigned-char -fno-common -fno-PIE
-	-fno-strict-aliasing \\
-	-mno-sse
-	-mno-mmx -mno-sse2 -mno-3dnow -mno-avx \\
-	-fcf-protection=branch
-	-fno-jump-tables -m64 -falign-jumps=1 \\
-	-falign-loops=1 -mno-80387 -mno-fp-ret-in-387 \\
-	-mpreferred-stack-boundary=3 -mskip-rax-setup \\
-	-mtune=generic -mno-red-zone -mcmodel=kernel
-	-Wno-sign-compare \\
-	-fno-asynchronous-unwind-tables
-	-mindirect-branch=thunk-extern \\
-	-mindirect-branch-register
-	-mindirect-branch-cs-prefix \\
-	-mfunction-return=thunk-extern -fno-jump-tables \\
-	-fpatchable-function-entry=16,16
-	-fno-delete-null-pointer-checks \\
-	-O2 -fno-allow-store-data-races
-	-fstack-protector-strong \\
-	-fomit-frame-pointer
-	-fno-stack-clash-protection -falign-functions=16 \\
-	-fno-strict-overflow -fno-stack-check -fconserve-stack
-	-Wall -Wundef \\
-	-Werror=implicit-function-declaration
-	-Werror=implicit-int \\
-	-Werror=return-type
-	-Werror=strict-prototypes -Wno-format-security \\
-	-Wno-trigraphs -Wno-frame-address \\
-	-Wno-address-of-packed-member
-	-Wframe-larger-than=2048 -Wno-main \\
-	-Wno-unused-but-set-variable -Wno-unused-const-variable -Wvla \\
-	-Wno-pointer-sign -Wcast-function-type -Wno-array-bounds \\
-	-Wno-alloc-size-larger-than -Wimplicit-fallthrough=5 \\
-	-Werror=date-time -Werror=incompatible-pointer-types \\
-	-Werror=designated-init -Wenum-conversion
-	-Wno-unused-but-set-variable \\
-	-Wno-unused-const-variable -Wno-restrict -Wno-packed-not-aligned \\
-	-Wno-format-overflow 
-	-Wno-format-truncation -Wno-stringop-overflow \\
-	-Wno-stringop-truncation
-	-Wno-missing-field-initializers \\
-	-Wno-type-limits
-	-Wno-shift-negative-value -Wno-maybe-uninitialized \\
-	-Wno-sign-compare 
-	-DKBUILD\_MODFILE='"kernel/rcu/tree"' \\
-	-DKBUILD\_BASENAME='"tree"' -DKBUILD\_MODNAME='"tree"' \\
-	-D\_\_KBUILD\_MODNAME=kmod\_tree \\
-	-c -o kernel/rcu/tree.o kernel/rcu/tree.c
-	}
-}
-\end{quote}
-
-We do not propose to explain all of these, and sufficiently motivated
-readers can avail themselves of the GCC documentation.
-We instead look at representative members of several categories.
-
-The \co{-funsigned-char} causes the \co{char} type to be unsigned, which
-overrides per-architecture defaults, some of which treat \co{char}
-as signed and others as unsigned.
-This choice prevents a class of bugs, and also allows the kernel to make
-reliable use of the uppermost bit of variables of type \co{char}.
-It also affects the definition of ``semantic dependency'' by changing
-the arithmetic proporties of this type.
-In theory, the standard could have specified the signedness of \co{char}
-but the variety of existing practice in the 1980s prevented this.
-Plus, optimizers were much less capable back then, so the signedness of
-\co{char} was much more of a performance concern than it is today.
-
-The \co{-mno-sse} prevents GCC from making use of the processor's SSE hardware.
-This is done for performance reasons, as it avoids the overhead of
-saving and restoring the state of this hardware when switching between
-user and kernel contexts.
-Similarly, the \co{-mcmodel=kernel} causes the kernel binary to be
-placed in the uppermost 2GB of the address space, again reducing
-the overhead of switching between user and kernel contexts.
-These are cases where the standard does not specify anything, nor
-should it.
-
-The \co{-fpatchable-function-entry=16,16} causes GCC to emit 16
-\co{nop} instructions at the beginning of each function, with
-the function's entry point being just after this string of \co{nop}
-instructions.
-The resulting buffers are used by the Linux-kernel tracing infrastructure
-which is in turn used for debugging, performance measurement, and
-monitoring.
-This is again clearly outside the scope of the standard.
-
-The \co{-fstack-protector-strong} causes GCC to emit code that provides
-some protection against some classes of attacks based on buffer overflows.
-One could rightly argue code should simply avoid ever overflowing buffers,
-but things like memory allocators and userspace memory accesses must
-use code that can be difficult to distinguish from buffer overflows.
-It is not clear that the ever-increasing variety of attacks should
-affect the standard.
-
-The \co{-fno-strict-overflow} causes GCC to act as if signed integer
-overflow is defined behavior,
-which also affects the definition of ``semantic dependency''.
-This might be a controversial choice, and another option would be to add
-a new set of signed integer types to the standard for which overflow is
-defined as wrapping, similar to the situation with unsigned integers.
-
-The \co{-Werror=strict-prototypes} causes GCC to warn if old-style
-non-ANSI function prototypes are used.
-This helps avoid certain classes of bugs.
-Warnings are by design outside the scope of the standard.
-
-To sum up, user preference can exert a nontrivial influence over
-language semantics, and in particular can affect some aspects of the
-definition of ``semantic dependency''.
-
-\clearpage
-
 \section{But What About Tooling?}

### 2024-08-31 paper/litmus2table.sh
litmus2table.sh: Handle multiple logical ands and ors
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -5,7 +5,7 @@
 # Outputs an abbreviated LaTeX table variant.
 
 sed -e 's/&/\\&/' | sed -e 's/	}/	\\}/' | sed -e 's/{/\\{/' |
-sed -e 's,\\/,$\\vee$,' | sed -e 's,/\\,$\\wedge$,' |
+sed -e 's,\\/,$\\vee$,g' | sed -e 's,/\\,$\\wedge$,g' |
 awk -v amp='&' -v bs='\\' -v dq='"' '
 BEGIN {
 	ininit = 0;

### 2024-08-31 paper/ootavacuous.tex
ootavacuous: Convert two by-hand listings to tables
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{listing}[tbp]
-\begin{verbatim}
- 1 atomic<int> x(0);
- 2 atomic<int> y(0);
- 3
- 4 void thread1()
- 5 {
- 6   int r1 = x.load(memory_order_relaxed);
- 7   y.store(r1, memory_order_relaxed);
- 8 }
- 9
-10 void thread2()
-11 {
-12   int r2 = y.load(memory_order_relaxed);
-13   x.store(r2, memory_order_relaxed);
-14 }
-\end{verbatim}
+% Constructed by hand.
+{
+\scriptsize
+\begin{tabular}{r|l|l}
+& \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
+\hline
+1. &
+\texttt{~~int r1 =\textsubscript{rlx} x} &
+	\texttt{~~int r2 =\textsubscript{rlx} y} \\
+2. &
+\texttt{~~y =\textsubscript{rlx} r1} &
+	\texttt{~~x =\textsubscript{rlx} r2} \\
+\end{tabular}
+
+\vspace{0.1in}
+Condition (0:r1=42 $\wedge$ 1:r2=42)
+sometimes satisfied.
+}
 \caption{Simple OOTA}
@@
 that there is nothing in the initial values or the executable code to
-support such an outcome:
-
-\begin{enumerate}
-\item   Line~6 loads from \co{x} into \co{r1}, claiming to read
-	the value of line~13's store rather than \co{x}'s initial value
-	and somehow obtaining 42.
-\item   Line~7 stores \co{r1}, and thus 42, to \co{y}.
-\item   Line~12 loads 42 from \co{y} to \co{r2}.
-\item   Line~13 stores 42 to \co{x}, justifying the value loaded by line~6.
-\end{enumerate}
+support such an outcome.
+First, P0's line~1 loads from \co{x} into \co{r1}, claiming to read the value
+of P1's line-2 store rather than \co{x}'s initial value and somehow
+obtaining 42.
+Second, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
+Third, P1's line~1 loads 42 from \co{y} to \co{r2}.
+Fourth and finally, P1's line~2 stores 42 to \co{x}, justifying the value
+loaded by P0's line~1.
 
@@
 \begin{listing}[tbp]
-\begin{verbatim}
- 1 atomic<int> x(0);
- 2 atomic<int> y(0);
- 3
- 4 void thread1()
- 5 {
- 6   int r1 = x.load(memory_order_relaxed);
- 7   y.store(r1, memory_order_relaxed);
- 8 }
- 9
-10 void thread2()
-11 {
-12   int r2 = y.load(memory_order_relaxed);
-13   x.store(42, memory_order_relaxed);
-14 }
-\end{verbatim}
+% Constructed by hand.
+{
+\scriptsize
+\begin{tabular}{r|l|l}
+& \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
+\hline
+1. &
+\texttt{~~int r1 =\textsubscript{rlx} x} &
+	\texttt{~~int r2 =\textsubscript{rlx} y} \\
+2. &
+\texttt{~~y =\textsubscript{rlx} r1} &
+	\texttt{~~x =\textsubscript{rlx} 42} \\
+\end{tabular}
+
+\vspace{0.1in}
+Condition (0:r1=42 $\wedge$ 1:r2=42)
+sometimes satisfied.
+}
 \caption{Simple Reordering}
@@
 lines~12 and~13, which can result in all of \co{x}, \co{y}, \co{r1},
-and \co{r2} having the value 42 as follows:
-\begin{enumerate}
-\item   Line~13 stores 42 to \co{x}.
-\item   Line~6 loads 42 from \co{x} into \co{r1}.
-\item   Line~7 stores \co{r1}, and thus 42, to \co{y}.
-\item   Line~12 loads 42 from \co{y} to \co{r2}.
-\end{enumerate}
+and \co{r2} having the value 42 via the following steps.
+First, P1's line~2 stores 42 to \co{x}.
+Second, P0's line~1 loads 42 from \co{x} into \co{r1}.
+Third, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
+Fourth and finally, P1's line~1 loads 42 from \co{y} to \co{r2}.
 Current C++ implementations can and do exhibit this reordering behavior.

### 2024-09-01 paper/ootavacuous.tex
paper/ootavacuous: Wordsmith Section 1 ("Introduction and Background")
Drop the section introducing herd7 and adjust the examples
accordindly.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Listing~\ref{lst:Simple OOTA}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
-shows a simple example where an OOTA cycle might result in all of \co{x},
-\co{y}, \co{r1}, and \co{r2} having final values of 42, despite the fact
-that there is nothing in the initial values or the executable code to
-support such an outcome.
-First, P0's line~1 loads from \co{x} into \co{r1}, claiming to read the value
-of P1's line-2 store rather than \co{x}'s initial value and somehow
-obtaining 42.
+Listing~\ref{lst:Simple OOTA} shows a simple example where an OOTA
+cycle might result in all of \co{x}, \co{y}, \co{r1}, and \co{r2}
+having final values of 42, despite the fact that there is nothing in
+the default-zero initial values or the executable code to support such
+an outcome~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
+The ``=\textsubscript{rlx}'' denotes a relaxed access, with variables
+whose names are ``\co{r}'' followed by a number denoting registers,
+and other names denoting shared variables.
+First, P0's line~1 does a relaxed atomic load from \co{x} into \co{r1},
+reading the value of P1's line-2 store rather than \co{x}'s initial
+value of zero and somehow obtaining 42.
 Second, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
@@
 
-Because nothing else in the C++ memory model rules out such OOTA cycles,
-the C++ standard explicitly prohibits them in 33.5.4p8
+Because nothing else in the C++ memory model rules out such OOTA cycles
+(as indicated by the ``sometimes satisfied'' in the figure), the C++
+standard explicitly prohibits them in 33.5.4p8
 (\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
@@
 In Listing~\ref{lst:Simple OOTA}
-there is a \emph{semantic dependency} from line~6 to line~7 and
-another from line~12 to line~13.
+there is a \emph{semantic dependency} from line~1 to line~2 in both
+P0 and P1.
 (Roughly speaking, there is a semantic dependency from a given load
@@
 	for more discussion about hardware speculation.}
-the stores in lines~7 and~13 cannot take place until the CPU
-knows what values are loaded by lines~6 and~12, respectively.
+the stores on line~2 cannot take place until the P0's and P1's CPUs
+know what values are returned by the loads on line~1.
 Thus the hardware orders these stores after their corresponding loads,
@@
 Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
-Both the C++ compiler and the CPU are within their rights to reorder
-lines~12 and~13, which can result in all of \co{x}, \co{y}, \co{r1},
+Both the C++ compiler and the CPU are within their rights to reorder P1's
+lines~1 and~2, which can result in all of \co{x}, \co{y}, \co{r1},
 and \co{r2} having the value 42 via the following steps.
@@
 Fourth and finally, P1's line~1 loads 42 from \co{y} to \co{r2}.
-Current C++ implementations can and do exhibit this reordering behavior.
-
-\medskip
+Current C++ implementations can and do exhibit this reordering behavior,
+as indicated by the ``sometimes satisfied'' in the figure.
+
+\subsubsection{Non-Trivial OOTA}
+\label{sec:Non-Trivial OOTA}
 
@@
 
+\begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-ctrl.litmus @@
+\caption{OOTA Cycle}
+\label{lst:OOTA Cycle}
+\end{listing}
+
+Listing~\ref{lst:OOTA Cycle}
+shows a code fragment that under loose C++ has an OOTA cycle, though
+the cycle is of course prohibited in full C++.
+This example differs from that of Section~\ref{sec:Simple OOTA Cycle}
+in that line~3 of both processes stores the value 42, but only if
+line~2 of a given process determines that the value loaded was 42,
+keeping in mind that variables are default-initialized to zero.
+
+This OOTA cycle is constructed in the following manner.
+First, P0's and P1's line~1 load from \co{y} and \co{x}, respectively,
+both somehow returning the value 42.
+Second, P0's and P1's line~2 determine that the values loaded were both
+42, so each process advances to its line~3.
+Third, P0's and P1's line~3 stores the value 42 that was (somehow)
+clairvoyantly loaded by the other process's line~1.
+
 The next section looks at how prior work has refined these issues.
@@
 Some of these workers appear to have abandoned this effort, but as of
-early 2024, Mark Batty is persisting with modular relaxed dependencies.
+2024, Mark Batty is persisting with modular relaxed dependencies.
+@@@ cite web site.
 
@@
 of these constraints.
-
-\subsection{Code-Analysis Tool}
-\label{sec:Code-Analysis Tool}
-
-This paper will use the \co{herd7}\footnote{
-	Available at \path{https://github.com/herd/herdtools7}.}
-tool to analyze fragments of C++ code.
-This tool carries out the moral equivalent of full state-space searches
-of concurrent code fragments.
-In some cases its output will include executions with OOTA cycles,
-on occasion reporting undefined values for the variables involved in
-the cycle.\footnote{
-	This happens only some of the time because of idiosyncrasies
-	in the algorithm used by \co{herd7}'s self-consistency solver.}
-
-\begin{listing}[tbp]
-@@ DisplayLitmus litmus/oota-ctrl.litmus @@
-\caption{OOTA Cycle}
-\label{lst:OOTA Cycle}
-\end{listing}
-
-Listing~\ref{lst:OOTA Cycle}
-shows a code fragment that under loose C++ has an OOTA cycle (although
-the cycle is of course prohibited in full C++).
-This section describes the fragment, thereby giving an overview of
-the \co{herd7} tool.
-
-The first line identifies it as a C-language litmus test and gives it
-a name, and this name identifies the litmus test's source file within
-the \path{litmus} directory in the \url{https://github.com/paulmckrcu/oota}
-repository.
-Lines~2--5 initialize variables, in this case setting the initial
-values of the global shared variables \co{x} and \co{y} to zero.
-(Variables that are not explicitly initialized are initialized
-to zero by default.)
-Lines~7--13 define the first thread, \co{P0()}, and lines~15--21
-define the second thread, \co{P1()}.\footnote{
-	The ``P'' stands for ``process'', which is \co{herd7}'s name
-	for ``thread''.}
-The arguments to both \co{P0()} and \co{P1()} specify which of
-the global shared variables each thread may access, in this case, \co{x}
-and \co{y}.
-The body of each thread contains C++ code,
-written in a slightly stilted manner to keep the load operations separate
-from the rest and because \co{herd7}'s knowledge of C++ is limited.
-
-Line~23 has a \co{locations} clause, which causes
-\co{herd7} to dump out the final values of \co{x} and \co{y}.
-Finally, line~24 specifies an \co{exists} clause, which gives a
-condition to check for the final values of the specified variables.
-The \co{0:} prefix denotes a variable local to \co{P0()} and
-the \co{1:} prefix denotes a variable local to \co{P1()}.
-The \co{/\\} is a boolean AND, and the \co{=} signs are equality comparisons.
-If a variable appears in the \co{exists} clause then the final value of
-that variable constitutes observable behavior.
-
-\begin{listing}[tbp]
-@@ RunLitmus litmus/oota-ctrl.litmus @@
-\caption{OOTA Cycle, \co{herd7} Output}
-\label{lst:OOTA Cycle, herd7 Output}
-\end{listing}
-
-Listing~\ref{lst:OOTA Cycle, herd7 Output} shows the corresponding output
-of the \co{herd7} tool.
-Lines~3 and~4 show the possible states, with line~4 showing the
-counterintuitive outcome where both threads load the value 42.
-Normally these lines would include only those variables mentioned
-in the \co{exists} clause, but because of the \co{locations} clause
-the values of \co{x} and \co{y} are also listed, which
-can be helpful for debugging.
-% @@@ As we will see, this outcome is not possible in real-world
-% @@@ implementations due to physical constraints of which \co{herd7}
-% @@@ is (by design) unaware.
-Line~9 contains \co{Sometimes} (as opposed to \co{Never} or \co{Always}),
-indicating that some executions satisfy the \co{exists} clause and
-others do not.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/simple-reordering.litmus @@
-\caption{Simple Reordering as Litmus Test}
-\label{lst:Simple Reordering as Litmus Test}
-\end{listing}
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-4.litmus @@
-\caption{Causality Test Case 4}
-\label{lst:Causality Test Case 4}
-\end{listing}
-
-Later examples will usually combine the litmus test and the \co{herd7}
-output into one listing, as shown in
-Listing~\ref{lst:Simple Reordering as Litmus Test},
-which recasts
-Listing~\ref{lst:Simple Reordering}
-into litmus-test form.
-Listing~\ref{lst:Simple OOTA}
-can also be recast as a \co{herd7} litmus test, as shown in
-Listing~\ref{lst:Causality Test Case 4}
-on page~\pageref{lst:Causality Test Case 4}.
-Other OOTA-related litmus tests may be found in
-Appendix~\ref{app:Illustrative Litmus Tests}.
 
@@
 \begin{listing}[tbp]
+@@ DisplayRunLitmus litmus/oota-causality-4.litmus @@
+\caption{Causality Test Case 4}
+\label{lst:Causality Test Case 4}
+\end{listing}
+
+\begin{listing}[tbp]
 @@ DisplayRunLitmus litmus/oota-3proc.litmus @@

### 2024-09-01 paper/ootavacuous.tex
paper/ootavacuous: Replace Appendix B with comment markers
Appendix B ("But What About Tooling?") is not referenced by any of the
rest of the paper, so leave comments commorating the sections.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\section{But What About Tooling?}
-\label{sec:But What About Tooling?}
-
-This paper focuses primarily on showing that, under commonly occurring
-constraints, OOTA cycles cannot form in real-world C++ implementations.
-
-But what about tooling?
-
-One entirely reasonable reaction is that, given the issues raised in
-Section~\ref{sec:Properties of Semantic Dependencies},
-Section~\ref{sec:C++ Compilers},
-and
-Section~\ref{sec:Outstanding Issues},
-load/store reordering is the least of the problems faced by tooling.
-Nevertheless, this appendix expands on
-Section~\ref{sec:Semantic Dependencies and Tooling}
-by looking into the possibility of:
-\begin{itemize}
-\item	Tooling focusing on only part of the language,
-\item	Changing the language to eliminate some aspects that are
-	troublesome for tooling (and potentially accepting performance
-	and energy-efficiency shortfalls), and
-\item	Changing the language to better delineate those portions that
-	tooling can easily accommodate.
-\end{itemize}
-
-But first, the next section looks at why some weakly ordered
-computer hardware appears to forbid reordering of prior relaxed loads with
-later relaxed stores,\footnote{
-	Or, in memory-order-speak, forbid all load-buffering litmus tests.}
-despite the architecture permitting such reordering.
-
-\subsection{Load/Store Ordering: Hardware View for Software Hackers}
-\label{sec:Load/Store Ordering: Hardware View for Software Hackers}
-
-Both the ARM and PowerPC architectural memory models permit reordering prior loads
-against subsequent stores, but actual tests on recent hardware fail to
-produce any evidence of such reordering.
-Nevertheless, thus far neither ARM nor PowerPC hardware architects
-have been willing to strengthen their memory models so as to forbid
-such reordering, despite a number of requests to do
-so~\cite{RichardGrisenthwaite2023TightenRelaxedJustSayNo}.
-
-This appendix offers some possible reasons for this odd juxtaposition
-of negative test results and flat refusal.
-
-One key point is that some hardware (including ARM and PowerPC) provides
-precise exceptions.
-For example, if a given load results in a segmentation violation
-exception, that exception will occur before any instructions following
-that load have committed.
-This means that a given load instruction's execution must have proceeded
-beyond the point where an exception might occur before any subsequent
-store can be permitted to commit, even if that store is completely
-unrelated to that load.
-For example, the later store cannot commit until all prior loads' address
-translations have completed successfully.
-
-However, if the load suffers a cache miss, address translation will have
-completed long before the load returns its value, meaning that a later
-store might well commit before the load completes.
-In fact, that later store might commit before the store which supplies the
-value returned by the load!
-Which explains why some hardware systems really do reorder prior loads
-and later stores.
-
-The question then becomes ``Why would weakly ordered systems fail to
-reorder prior loads with later stores?''
-
-One explanation is ECC errors.
-
-If correctable ECC errors were fixed up in hardware, then only uncorrectable
-ECC errors would be directly visible to software.
-If the only possible reaction to an uncorrectable ECC error was to
-terminate the program suffering that ECC error, it would be safe to
-allow subsequent stores to commit as soon as all prior loads' address
-translations had completed successfully.
-
-However, some kernels and applications have ways of handling even
-uncorrectable errors.
-Operating-system kernels encountering an ECC error might note that the
-affected data was being used by only one user process, and might react by
-killing that user process, taking care to account for memory shared with
-other processes.
-Some user applications might note that the corrupted data affected only
-a particular computation, and might react by restarting that computation.
-On the other hand, it is only reasonable to react to an uncorrectable ECC
-error in a load from a read-only mapped file by re-reading that data
-from that file, then restarting that load instruction.
-In such cases, the kernel and the user application would need to continue
-execution, and would thus require a precise exception.
-
-Systems that offload processing of correctable ECC errors to software
-also require precise exceptions.
-After all, the software is going to need to be able to correct the error
-and then fix up the state to make it appear that the load had returned
-the fixed-up value.
-% I was tempted to add a sentence saying that although one could imaging
-% ECC on memory but not on caches, today's caches are a bit large for this
-% to be a winning strategy.  One might nevertheless consider omitting ECC on
-% the smallest caches closest to the CPU, but this is unlikely to allow LB.
-
-ECC errors are detected near the end of their corresponding load
-instructions' execution, and so any need for precise exceptions rules out
-committing any subsequent stores until after the load has fetched
-(and possibly corrected) its value.
-
-So why are hardware architects reluctant to tighten their memory models
-to forbid reordering of earlier loads and later stores?
-
-If you would like an authoritative answer to this question, you should of
-course ask your friendly local hardware architect.
-In the meantime, here is some semi-informed speculation on this topic:
-
-\begin{itemize}
-\item	Some hardware might choose to forego ECC, for example, in order to
-	reduce cost for low-end systems or to improve energy efficiency
-	for battery-powered systems.
-\item	ECC error correction might still be done in hardware, avoiding
-	the need for precise software-visible exceptions.
-\item	Some systems might prefer to immediately shut down in response to
-	an uncorrectable ECC error, perhaps due to safety considerations.
-	This would entirely avoid the need for software-visible
-	ECC-related exceptions.\footnote{
-		One motivation for immediate shutdown is that if there
-		are uncorrectable errors, there might soon be errors
-		that change one valid bit pattern to another valid
-		bit pattern, which could result in a lack of safety.}
-\item	Someone might come up with a clever way of correcting ECC errors
-	in firmware, avoiding the need for precise software-visible
-	exceptions.
-\item	As late as early 2024, some GPGPUs have been observed reordering
-	earlier loads against later stores.
-	Other hardware architects might therefore feel the need to keep
-	this option open for their own systems.
-\end{itemize}
-
-So although it is not unreasonable to continue asking hardware vendors
-to tighten their memory models so as to prohibit reordering of earlier
-loads and later stores, it also would not be too surprising for them to
-continue to refuse.
-
-\subsection{Status Quo and Focused Tooling}
-\label{sec:Status Quo and Focused Tooling}
-
-The theoretical possibility of OOTA cycles causes some tools to have
-difficulty identifying precisely which outcomes are impossible on
-real-world C++ implementations.
-One approach is for tooling to reject programs containing instances
-of \co{memory_order_relaxed} and \co{memory_order_consume}, so that
-developers desiring their code to be analyzed by such tools would
-avoid using these \co{memory_order} values.
-The is strong precedent for this strategy, with the common prohibition
-against side effects in function arguments being but one example.
-
-However, there is a large body of existing code that uses
-\co{memory_order_relaxed}, and it would be unfortunate if such tools
-could not be applied to this code.
-
-Another approach would be to add flags to C++ implementations
-to cause \co{memory_order_relaxed} to be interpreted as either
-\co{memory_order_acquire} (for loads), \co{memory_order_release} (for
-stores), or \co{memory_order_acq_rel} (for read-modify-write operations).
-On TSO systems and on systems featuring precise ECC exceptions, this
-would allow tooling to be brought to bear without performance or
-energy-efficiency consequences beyond forgone optimizations involving
-memory reference reordering.
-
-However, this would mean that programs compiled in this way would not
-be guaranteed to run correctly on weakly ordered systems that lack
-ECC (or that lack precise exceptions).
-In addition, this change is more strict than necessary, forbidding
-optimizations that tooling could in fact analyze.
-Therefore, the next two sections look at standardizing less severe
-restrictions.
-
-\subsection{Change Relaxed to Forbid Load Buffering}
-\label{sec:Change Relaxed to Forbid Load Buffering}
-
-Tooling does not require relaxed accesses become fully acquire and/or
-release, but rather only that implementations be forbidden from reordering
-prior relaxed loads with subsequent relaxed stores, as has been suggested
-many times over the
-years~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352}.
-
-This preserves portability while enabling tooling to handle all members
-of the \co{memory_order} enumeration, but inflicts some performance and
-energy-efficiency penalties~\cite{LukeGeeson2023TightenRelaxed} in
-code not requiring these ordering
-restrictions~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
-
-\subsection{Add Load-Store Memory Order that Forbids Load Buffering}
-\label{sec:Add Load-Store Memory Order that Forbids Load Buffering}
-
-The addition of a \co{memory_order_load_store} member to the
-\co{memory_order} enumeration has been suggested starting many years
-ago~\cite{HansBoehm2019OOTArevisitedAgain,DanielLustig2018PlacedBefore}.
-This does not resolve all shortcomings in the C++ memory
-model~\cite{PaulEMcKenney2023P0124R8-LKMM,DanielLustig2018PlacedBefore},
-but it would provide a portable \co{memory_order} that minimally
-restricted compiler and hardware optimizations while still permitting
-full analyzability by current software tools.
-
-This change would also leave \co{memory_order_relaxed} in place,
-allowing minimal-overhead accesses in fastpaths.
-These fastpaths would not be analyzable by generic tools, but could
-be handled by special-case tools that analyze the binaries to verify
-that required orderings are enforced by machine-language properties
-such as dependencies.
-And a tool that checks control dependencies has in fact been
-prototyped~\cite{PaulHeidekrueger2022N4910}.
-
-This suggests a combined strategy of adding \co{memory_order} members
-as needed to extend the reach of general-purpose tooling, while also
-identifying \co{memory_order_relaxed} idioms that are checked at the
-machine level using special-purpose tooling.\footnote{
-	Kudos to Peter Sewell for clearly articulating this possibility
-	as part of an overall strategy.}
-
-\clearpage
+% \section{But What About Tooling?}
+% \label{sec:But What About Tooling?}
+% @@@ State conclusions if needed where needed.
+
+% \subsection{Load/Store Ordering: Hardware View for Software Hackers}
+% \label{sec:Load/Store Ordering: Hardware View for Software Hackers}
+% @@@ State conclusions if needed where needed.
+
+% \subsection{Status Quo and Focused Tooling}
+% \label{sec:Status Quo and Focused Tooling}
+% @@@ State conclusions if needed where needed.
+
+% \subsection{Change Relaxed to Forbid Load Buffering}
+% \label{sec:Change Relaxed to Forbid Load Buffering}
+% @@@ State conclusions if needed where needed.
+
+% \subsection{Add Load-Store Memory Order that Forbids Load Buffering}
+% \label{sec:Add Load-Store Memory Order that Forbids Load Buffering}
+% @@@ State conclusions if needed where needed.
 

### 2024-09-01 paper/ootavacuous.tex
paper/ootavacuous: Summarize "What Needs to Change" section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 probably wouldn't do anyway.
-
-We begin with a summary of what needs to change and a brief overview of
-the OOTA problem, followed by an equally brief summary of prior work in
-this area, and ending with a quick overview of the \co{herd7} tool that
-will be used to evaluate litmus tests.
-
-\subsection{What Needs to Change?}
-\label{sec:What Needs to Change?}
-
-What changes are needed to prevent OOTA from C++ implementations using
-traditional compilers and traditional hardware, including CPUs and
-GPGPUs?
-
-\begin{itemize}
-\item	To the standard:
-	No changes.
-	The indefinite prohibition of out-of-thin-air values needs
-	to remain.
-\item	To implementations:
-	No changes.
-	Future implementations must avoid optimizations that invent,
-	duplicate, or repurpose atomic loads.
-	However, such optimizations must be avoided for other reasons,
-	including preserving basic laws of arithmetic
-	(see Appendix~\ref{app:Inventing Atomic Loads} for more detail).
-\item	To user code:
-	No changes.
-	Users must avoid undefined behavior in portable code, just as
-	is the case now.
-\end{itemize}
-
-In short, no changes are needed because current implementations already
-avoid OOTA cycles.
-The point of this paper is to give a semi-formal demonstration of
-why this is the case.
+Absolutely no changes are required to either the standard or to user code.
+
+We begin with a brief overview of the OOTA problem followed by an equally
+brief summary of prior work in this area.
 

### 2024-09-01 paper/ootavacuous.tex
paper/ootavacuous: Drop table of contents and rework header
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
paper/ootavacuous: Anonymize authors for review process
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{document}
-\title{P3064R2: How to Avoid OOTA Without Really Trying}
+\title{How to Avoid OOTA Without Really Trying}
 
@@
 
-\author{
-Alan Stern\\{\small \email{stern@rowland.harvard.edu}} \and
-Paul E.~McKenney\\{\small \email{paulmck@kernel.org}} \and
-Michael Wong\\{\small \email{fraggamuffin@gmail.com}} \and
-Maged Michael\\{\small \email{maged.michael@gmail.com}}
-}
-\date{July 12, 2024 (Post-St. Louis)}
+% Blind review process.
+% \author{
+% Alan Stern\\{\small \email{stern@rowland.harvard.edu}} \and
+% Paul E.~McKenney\\{\small \email{paulmck@kernel.org}} \and
+% Michael Wong\\{\small \email{fraggamuffin@gmail.com}} \and
+% Maged Michael\\{\small \email{maged.michael@gmail.com}}
+% }
 \maketitle{}
-
-Audience: SG1
 
@@
 \end{abstract}
-
-\pagebreak
-
-\tableofcontents
-
-\pagebreak
 

### 2024-09-01 paper/ootavacuous.tex
paper/ootavacuous: Anonymize further by dropping ack section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\section{Acknowledgments}
-\label{sec:Acknowledgments}
-
-We are grateful to David Goldblatt, Jade Alglave, and Peter
-Sewell for their careful review of an early draft of this paper and
-to John Wickerson for asking Paul for a rant and taking the proffered
-rant seriously.
-We also owe David Goldblatt a debt of gratitude for his having asked an
-insightful question at the right time, his insights on combinations of
-OOTA and UB, and for his ``Deathstation 9000'' demonic CPU.
-% Alan Stern located many unclear statements and gaps in reasoning.
-Martin Uecker contributed valuable insights on backwards-propagating UB.
-Gonzalo Brito Gadeschi shared an alternative way of handling nonvolatile
-atomic operations.
-Richard Grisenthwaite patiently explained the architectural constraints
-that prevent hardware OOTA.
-Mark Batty encouraged our work on new OOTA-related litmus tests.
-
-Nonetheless, all errors and omissions in this paper are the sole property
-of the authors, and the appearance of a name in this appendix does
-not in any way constitute agreement with anything in this paper.
-
-% @@@ reviewer to do.
-% Luc Maranget: Memory models.
-% Christoph Hellwig: Why slow on BPF memory model.
-% Miguel Ojeda: LKMM and Rust.
-% Alice Ryhl: LKMM and Rust.
-% Ralf Jung: Rust.
-% Catalin Marinas: ARM and formal methods (qspinlock).
-% Segher Boessenkool:  GCC.
-% Dan Lustig: NVIDIA memory ordering.
-% Andrea Parri: Memory models, RISC-V hardware.
-% Jonas Oberhauser: Memory models.
-% Hernan Ponce de Leon: Memory models.
-% Palmer Dabbelt: RISC-V hardware.
-% Ali Sezgin: P0422R0.
-% Tony Tye: P0422R0.
-% Other memory-model maintainers.
-% Derek Williams: PowerPC hardware.
-% Hans Boehm: Skeptic, memory models, OOTA.  Suggestions for others?
-% Luke Geeson: Whatever...  Review.
-% Ori Lahav: Prior OOTA.
-% Viktor Vafeiadis:  Prior OOTA.
-% Derek Dreyer: Prior OOTA.
-% Brian Demsky: Prior OOTA.
-
-% Done as of December 18, 2023 %% as of March 18, 2024
-%% David Goldblatt
-%% Boqun Feng, Uladzislau Rezki, Neeraj Upadhyay, Joel Fernandes, and
-%	Frederic Weisbecker (RCU proteges)
-%% Michael Wong
-%% Maged Michael
-% John Wickerson: Thank you.  OK for Peter Sewell, Mark Batty, etc.
-%% Akira Yokosawa: Professional courtesy, perfbook memory models.
-%% Dan Kelley, Alexei Starovoitov, Mykola Lysenko: FYI.  OOTA dropped
-% Alan Jeffrey: Fixed-point insight P0422R0.  Reached out via LinkedIn.
-%% Jade Alglave: C++ herd model?  Memory models and variety of hardware.
-%% Alan Stern: Memory models, mathematical logic, and variety of hardware.
-% Martin Uecker (offered)
-% Peter Sewell
-% Mark Batty
-% Richard Gristenthwaite: ARM hardware.  Posed HW speculation question.
-% Peter Zijlstra.
-% Mark Rutland: ARM memory ordering.
-% Will Deacon: ARM hardware, C++, and OOTA.
-%% Greg Kroah-Hartman and Linus Torvalds: FYI, trouble being caused.
-%% Olivier Giroux: Skeptic, memory models, NVIDIA, requested.
-%% Gonzalo Brito <gonzalob@nvidia.com>
-
-% @@@ TODO:
-% Construct interleaved OOTA examples (see sheets of paper).
-% Use LKMM as check for dependencies?
-
-% Objections:
-% "Atomic operations are not observable behavior."
-%	They are for volatile atomic objects.  Plus an implementation
-%	whose analysis is confined to a thread would need to assume
-%	that any atomic store might lead to observable behavior in some
-%	other thread.  An implementation with a global view must preserve
-%	observable behavior regardless of OOTA cycles.	Plus if an OOTA
-%	cycle does not affect observable behavior, who cares?
-% "The C++ abstract machine is independent of the laws of physics."
-%	That might be, but any real implementation of that abstract
-%	machine will be governed by those laws.
-% "Real hardware can exhibit OOTA, as exemplified by audio feedback."
-%	Such hardware does not run C++.  Also, audio feedback is more
-%	closely approximated by a loop calculating sound intensity than
-%	by OOTA, especially given that you must include the audio
-%	properties of the enclosing space in analyzing the feedback loop.
-
-% Additional snark:
-% "This work reduces the problem of OOTA analysis to that of sequential
-%	program analysis.  If you expect better verification of concurrent
-%	code than of sequential code, you are in an intellectual state
-%	of sin."
-% "Forcing relaxed loads to be ordered before subsequent relaxed stores
-%	is an expensive no-op."
-% "So you don't want to learn concurrency?  Then I can only suggest
-%	that you either move further up the stack or take early
-%	retirement."  Note that the C++ committee's concurrency
-%	expertise has grown beyond recognition over the past 15 years.
-
-% One-pager:
-% Threats to validity:
-% Speed of light might not always be finite.
-% Zero-sized atoms might be discovered.
-% It might be possible to propagate information at infinite velocity
-%	despite the finite speed of light.
-% It might be possible to violate causality.
-% Hardware and compiler bugs might result in OOTA.
-% ML hallucination might convince people that OOTA is OK.
-
 \section{History}

### 2024-09-01 paper/ootavacuous.tex
paper/ootavacuous: remove History appendix
This is not referenced elsewhere and is not consistent with an
anonymous submission.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\section{History}
-\label{sec:History}
-
-P3064R2 incorporates feedback from the St.~Louis meeting, primarily
-incorporating clear statements that nothing need change to avoid OOTA.
-In other words, the title really means what it says.
-
-P3064R1 adds non-OOTA examples that show that inventing, duplicating,
-or repurposing atomic loads can violate basic laws of arithmetic.
-
 \bibliographystyle{plain}

### 2024-09-01 paper/ootavacuous.tex
paper/ootavacuous: Remove "Additional Litmus Tests" appendix
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Most of the examples in this paper involve simple semantic dependencies
-connecting a single load to a single store.
-One exception appears in
-Appendix~\ref{app:Additional Litmus Tests},
+connecting a single load to a single store,
 but this section will present a more difficult example.
@@
 
-\subsection{Additional Litmus Tests}
-\label{app:Additional Litmus Tests}
-
-This section lists a few other litmus tests of interest:
-
-\begin{itemize}
-\item	The litmus test at
-	\url{https://github.com/paulmckrcu/oota/blob/master/litmus/oota-two-source.litmus}
-	shows multiple overlapping OOTA cycles.
-\item	The litmus test at
-	\url{https://github.com/paulmckrcu/oota/blob/master/litmus/oota-non-lb.litmus}
-	shows that OOTA cycles are not confined to the load-buffering (LB)
-	pattern.
-\item	The litmus test at
-	\url{https://github.com/paulmckrcu/oota/blob/master/litmus/oota-invent-int-load.litmus}
-	shows another invented-load scenario loosely based on
-	Figure~\ref{lst:OOTA Cycle}
-	on
-	page~\pageref{lst:OOTA Cycle}.
-	See Appendix~\ref{app:Inventing Atomic Loads}
-	for more discussion on this topic.
-\end{itemize}
-
 \clearpage

### 2024-09-01 paper/ootavacuous.tex
paper/ootavacuous: Remove "Interthread Communications" appendix
Also the one reference to it.  We might want to add a citation defining
rf, rfi, and rfe, so added a comment to this effect.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 each thread and rf is the set of store-to-load ``reads-from'' links,
-whether within a thread
-(rfi) or between threads (rfe).\footnote{
-	See Appendix~\ref{app:Interthread Communications} for definitions
-	and properties of rf, rfe, and rfi.}
+whether within a thread (rfi) or between threads (rfe).
+% @@@ Consider adding citation defining rf, rfi, and rfe.
 
@@
 
-\section{Interthread Communications}
-\label{app:Interthread Communications}
-
-This section more precisely defines the ``coe'', ``fre'', and ``rfe''
-types of interthread communication (also collectively called ``links'')
-and presents information about their temporal qualities.
-
-First, naming.
-The final ``e'' in all three acronyms stands for ``external'' (that
-is, between threads) as opposed to ``internal'' communication (within
-a thread: ``coi'', ``fri'', and ``rfi'') and also as opposed to all
-communication, whether interthread or intrathread (``co'', ``fr'', and
-``rf'').
-
-Next, definitions:
-\begin{description}
-\item[coe]
-	Coherence-order (external), which connects a store to any other store
-	(in a different thread)
-	to the same object that comes later in the object's
-	modification order, i.e., that overwrites either the first store's
-	value or some later value.
-\item[fre]
-	From-read (external), which connects a load to any store
-	(in a different thread) to
-	the same object that overwrites either the value that the load
-	returned or some later value.
-\item[rfe]
-	Reads-from (external), which connects a store to any load
-	(in a different thread) from
-	the same object that returns the value stored.
-	(Note: This means that the load retrieves the information
-	written by the store, as opposed to retrieving the information
-	from a different store that happened to write the same value.)
-\end{description}
-
-\begin{figure}[tb]
-\begin{center}
-\resizebox{4in}{!}{\includegraphics{ipc}}
-\caption{ITC Diagram for coe, fre, and rfe}
-\label{fig:ITC Diagram for coe, fre, and rfe}
-\end{center}
-\end{figure}
-
-Thirdly, Figure~\ref{fig:ITC Diagram for coe, fre, and rfe} illustrates
-the coe, fre, fri, and rfe links for four threads with time advancing
-from the top to the bottom of the figure.  Start with coe: Even though
-Thread~1's store is later in global time than that of Thread~0, the
-Thread~1 store comes first in \co{x}'s modification order, then Thread~0's
-store, then those of Threads~2 and~3.
-There is thus a coe link from Thread~1's store to every other store,
-from Thread~0's store to those of Threads~2 and~3, and from Thread~2's
-store to that of Thread~3.
-This situation can occur due to the initial placement and subsequent
-movement of cache lines.
-
-Thread~3's load reads the value written by Thread~0's store, so there is
-an rfe link from that store to that load.  In addition there is an fre link from
-the load to Thread~2's earlier store and an fri link to Thread~3's later store.
-Note that the fre link goes backwards in time.
-
-Finally, the reader might desire hard evidence that coe and fre
-links really can go backwards in time.
-We provide this evidence on x86 to demonstrate that these effects
-are not confined to weakly ordered architectures.
-The machine is a dual-socket system with
-Intel(R) Xeon(R) Gold 6138 CPUs @ 2.00~GHz,
-each socket with 20~cores and each core having a pair of hardware
-threads, for a grand total of 80 hardware threads.
-The code generating this data may be found in the \path{CodeSamples/cpu}
-directory of the git archive of the book
-``Is Parallel Programming Hard, And, If So, What Can You Do About
-It?''~\cite{McKenney2018ParallelProgramming-2018-12-08a}.\footnote{
-	\co{git://git.kernel.org/pub/scm/linux/kernel/git/paulmck/perfbook.git}}
-See especially the \path{perftemporal.sh} script.
-
-\begin{figure}[tb]
-\begin{center}
-\resizebox{4in}{!}{\includegraphics{coe-sh-out}}
-\caption{On x86, coe Links Are Atemporal}
-\label{fig:On x86; coe Links Are Atemporal}
-\end{center}
-\end{figure}
-
-Figure~\ref{fig:On x86; coe Links Are Atemporal}
-shows that coe links really are atemporal, that is, they can go backwards
-in time, and quite frequently at that, even on x86.
-During a run of the test, each of the threads stores a distinct value to
-the same atomic integer variable (all at about the same time), and each
-one records timestamps just before and just after its store.
-The ``winning'' store is the one whose value is retained in the
-variable at the end of the run, overwriting all the others.
-The store-to-store latency is measured from the beginning of the
-nonwinning store that started latest to the end of the winning store.
-Every data point on the negative x-axis thus represents a group of
-runs in which a nonwinning store started after the winning store
-finished.
-This demonstrates that the winning store quite commonly is not the
-last one, even on x86.
-
-\begin{figure}[tb]
-\begin{center}
-\resizebox{5in}{!}{\includegraphics{coe-out-1}}
-\caption{On x86, coe Links Form a Coherent Partial Order}
-\label{fig:On x86, coe Links Form a Coherent Partial Order}
-\end{center}
-\end{figure}
-
-Figure~\ref{fig:On x86, coe Links Form a Coherent Partial Order} shows
-that the sequence of values read from the atomic variable by each
-thread is consistent with a number of global orders, as
-required.\footnote{
-	Note that this data was not measured on the 80-thread system but
-	rather on a 16-thread x86 laptop, to avoid an unreadable diagram
-	containing 79 bubbles.}
-Working from the far left, one thread saw the values 0, 11, and 9,
-a second thread saw 3, 11, and 9, and so on.
-Although different threads saw different sequences of values, there is no
-disagreement on the order of the values that they could see.
-For example, all threads agree that the value 9 came last.
-This is an example of single-variable sequential consistency, and it
-meets the coherence requirements of the C++ standard.
-
-\begin{figure}[tb]
-\begin{center}
-\resizebox{3in}{!}{\includegraphics{fre-sh-out}}
-\caption{On x86, fre Links Are Atemporal}
-\label{fig:On x86, fre Links Are Atemporal}
-\end{center}
-\end{figure}
-
-Figure~\ref{fig:On x86, fre Links Are Atemporal}
-plots a histogram of the elapsed time from the beginning of a load that
-returned an old value to the end of the store that provided the new value.
-Most of the data falls into negative time.
-In fact, in the most commonly occurring case the last load of an old value
-executes about 60 timestamp periods (about 30~nanoseconds) \emph{after}
-the store which overwrote that old value.
-This shows that fre links can and do go backwards in global time,
-corresponding to the fre link between Thread~3 and Thread~2 in
-Figure~\ref{fig:ITC Diagram for coe, fre, and rfe}.
-
-\begin{figure}[tb]
-\begin{center}
-\resizebox{3in}{!}{\includegraphics{rfe-sh-out}}
-\caption{On x86, rfe Links Are Temporal}
-\label{fig:On x86, rfe Links Are Temporal}
-\end{center}
-\end{figure}
-
-Figure~\ref{fig:On x86, rfe Links Are Temporal}
-plots a histogram of the elapsed time from the store of a new value
-to the first load of that new value.
-Here all of the data falls into positive time, indicating
-that rfe links always go forward in global time, as those familiar with
-computer hardware, limits on speculative execution, and the laws of
-physics would expect.\footnote{
-	Our apologies to those who might feel that this is belaboring
-	the obvious.
-	And please understand that these points have proven helpful to
-	some of those whose work has never strayed too close to
-	hardware.}
-
-\begin{figure}[tb]
-\begin{center}
-\resizebox{4in}{!}{\includegraphics{coe}}
-\resizebox{4in}{!}{\includegraphics{fre}}
-\resizebox{4in}{!}{\includegraphics{rfe}}
-\caption{Propagation Delay and Temporal Properties of coe, fre, and rfe}
-\label{fig:Propagation Delay and Temporal Properties of coe, fre, and rfe}
-\end{center}
-\end{figure}
-
-Figure~\ref{fig:Propagation Delay and Temporal Properties of coe, fre, and rfe}
-shows how propagation delay explains the temporal properties of coe,
-fre, and rfe.
-The upper portion of the figure shows an atemporal coe, in which
-CPU~0 stores 1 to \co{x} after CPU~3 stores 2 but
-the modification order of \co{x} is decided after the fact.
-In this case, the value of 2 from CPU~3's store overwrites the value
-of 1 from CPU~0's store, despite the fact that CPU~0's store
-happened later in global time.
-One way this could happen is if the cache line containing
-\co{x} arrived at CPU~0 before arriving at CPU~3.
-A more fanciful explanation is that both values arrived at an
-interconnect at the same time, and the interconnect made an
-arbitrary choice between the two.
-Either way, the end result is that an earlier store can overwrite
-a later store.
-
-The middle portion of the figure shows an atemporal fre, in which
-CPU~3's load obtains the old value of zero from \co{x} despite
-having executed long after CPU~0's store, courtesy of the fact
-that the new value of \co{x} had not yet propagated to CPU~3.
-
-The lower portion of the figure shows a temporal rfe.
-Because a load on CPU~3 cannot obtain the value stored by CPU~0 until
-after that value has propagated to CPU~3, the fact that CPU~3 obtains
-the new value implies that the load was executed after the store.
-
-And this is exactly why the C++ memory model guarantees ordering from
-rfe links but not from coe and fre links for relaxed accesses (in the
-absence of other ordering from stronger atomic memory accesses or
-\co{atomic_thread_fence(memory_order_seq_cst)}).
-
-\clearpage
-
 % \section{But What About Tooling?}

### 2024-09-01 paper/ootavacuous.tex
paper/ootavacuous: Remove "Illustrative Litmus Tests" appendix
Move a couple of examples up to reference site.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 they must consider changes to multiple variables.
-See Appendix~\ref{app:Non-Trivial Semantic Dependencies}
-for examples and additional discussion.
+% @@@  Pull anything in?  Doesn't seem necessary.
+% See Appendix~\ref{app:Non-Trivial Semantic Dependencies}
+% for examples and additional discussion.
 
@@
 Therefore we can rule out transformations like this one by insisting
-the compiler not invent (or duplicate) atomic loads.\footnote{
-	This issue is discussed more fully in
-	Appendix~\ref{app:Inventing Atomic Loads},
-	which includes other examples of invented atomic
-	loads leading to incorrect behavior, that is, behavior not
-	exhibited by the abstract machine.
+the compiler not invent (or duplicate) atomic loads.
+Please note that invented and duplicated atomic loads are also problematic
+in arithmetic:
+\begin{quote}
+\begin{verbatim}
+int r0 = x;
+int r1 = r0 * r0 + 2 * r0 + 1;
+\end{verbatim}
+\end{quote}
+For sufficiently small values of \co{x}, the abstract machine is
+guaranteed to produce a perfect square in \co{r1}.
+
+But if the compiler is permitted to invent atomic loads then the compiled
+code might not respect this guarantee, thus incorrectly deviating
+from the abstract machine.
+This issue is evident in a transformed version of the above code:
+\begin{quote}
+\begin{verbatim}
+int r0 = x;
+int invented = x;
+int r1 = r0 * r0 + 2 * invented + 1;
+\end{verbatim}
+\end{quote}
+For example, if the load into \co{r0} happened when the value of \co{x}
+was zero and the load into \co{invented} happened when the value of
+\co{x} was 10, then the value of \co{r1} will be 21, which is not
+a perfect square.\footnote{
 	See also the ``Invented Atomic Loads'' paragraph of
         Section~\ref{sec:Constraints of the Standard}.}
+
 In fact, we will require that atomic accesses be treated as
@@
 that is, behavior that the abstract machine does not exhibit.
-See
-Appendix~\ref{app:Inventing Atomic Loads}
-for more details on this example and other examples of invented
-atomic loads.
+% @@@ See
+% @@@ Appendix~\ref{app:Inventing Atomic Loads}
+% @@@ for more details on this example and other examples of invented
+% @@@ atomic loads.
 
@@
 
-\section{Illustrative Litmus Tests}
-\label{app:Illustrative Litmus Tests}
-
-These litmus tests helped illuminate important aspects of the problem
-of defining sdep and identifying OOTA cycles.
-This appendix presents these tests in roughly decreasing order of
-importance, surprise, and illumination.
-It evaluates them using the \co{herd7} tool and also using manual
-analysis.
-
-\subsection{Semantic Dependencies and \co{volatile}}
-\label{app:Semantic Dependencies and volatile}
-
-The Linux kernel uses volatile accesses to constrain the compiler,
-and it is worth looking at the example shown in
-Listing~\ref{lst:Non-Volatile Accesses and Dependencies}
-to see how this works.
-
-\begin{listing}
-\begin{verbatim}
- 1 atomic<int> x, y;
- 2
- 3 void thread1()
- 4 {
- 5   int r1 = x.load(memory_order_relaxed);
- 6   y.store(r1, memory_order_relaxed);
- 7 }
-\end{verbatim}
-\caption{Non-Volatile Accesses and Dependencies}
-\label{lst:Non-Volatile Accesses and Dependencies}
-\end{listing}
-
-Because both \co{x} and \co{y} are nonvolatile, a C++ compiler is
-free to assume that the value loaded by line~5 will be either some value
-stored to \co{x} or its initial value.
-And because there are no stores to \co{x}, the only remaining possibility
-is the default-initialized value of zero.
-Therefore, the compiler is within its rights to substitute
-the constant zero for \co{r1} on line~6, eliminating the semantic
-dependency that would otherwise extend from line~5 to line~6.
-
-\begin{listing}
-\begin{verbatim}
- 1 volatile atomic<int> x, y;
- 2
- 3 void thread1()
- 4 {
- 5   int r1 = x.load(memory_order_relaxed);
- 6   y.store(r1, memory_order_relaxed);
- 7 }
-\end{verbatim}
-\caption{Volatile Accesses and Dependencies}
-\label{lst:Volatile Accesses and Dependencies}
-\end{listing}
-
-One way to preserve this dependency is use of \co{volatile}, as shown in
-Listing~\ref{lst:Volatile Accesses and Dependencies}.
-Now the compiler is forbidden from assuming that the value loaded
-by line~5 has any relation to any stores to or initialization of \co{x}.
-
-In this case, the \co{volatile} keyword alerts the compiler to
-the possibility of interfering changes to \co{x} from outside the
-program, for example, due to \co{x} being:
-\begin{itemize}
-\item	Allocated in an MMIO region of the address space.
-\item	Modified by an unknown-to-the-compiler thread
-	or signal handler.
-\item	Subject to I/O-device DMA operations.
-\item	Modified by debugger commands.
-\item	Modified using facilities such as \path{/dev/mem}.
-\item	Modified by as-yet-unwritten dynamically linked libraries.
-\item	Modified by some other mechanism of the reader's choosing.
-\end{itemize}
-These possibilities force the compiler to preserve the semantic
-dependency from line~4 to line~5,
-and this preserved dependency helps prevent the formation of OOTA cycles
-in environments where there is no interference.\footnote{
-	When there is interference, the definition of ``OOTA cycle''
-	does not apply.}
-
-The developer is free to arrange for \co{x} to be free of any
-interference, thus obtaining predictable behavior along with the
-preserved semantic dependency.
-However, this example invalidates the rough definition of semantic
-dependency from
-Section~\ref{sec:Simple OOTA Cycle}
-because the value loaded from \co{x} will always be the initial value
-of zero, so that it is at best dubious to talk about any changes in
-the value loaded.
-
-One way forward is to assume that interference could happen when
-determining which dependencies are semantic, and then do further analysis
-using that determination.
-This forms the basis for the volatile approach presented in
-Section~\ref{sec:Global Analysis and Volatile versus Quasi Volatile}.
-
-\subsection{Non-Trivial Semantic Dependencies}
-\label{app:Non-Trivial Semantic Dependencies}
-
-Most of the examples in this paper involve simple semantic dependencies
-connecting a single load to a single store,
-but this section will present a more difficult example.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult-0.litmus @@
-\caption{OOTA Multiplication Example, Initial Value Zero}
-\label{lst:OOTA Multiplication Example, Initial Value Zero}
-\end{listing}
-
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-shows an example that builds on the multiplication-by-zero
-discussion in
-Section~\ref{sec:Semantic Dependencies and Source Code}
-on
-page~\pageref{sec:Semantic Dependencies and Source Code}.
-Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
-\co{x} and \co{y} on lines~9 and~10, respectively.
-Similarly, lines~16 and~17 store to \co{x} and \co{y}, respectively,
-the value that line~15 loads from \co{z}.
-Because all three variables are initialized to zero, and because loads
-can be ordered before any store, there is an execution in which all
-three loads return the value zero, in which case all three stores will
-of course store the value zero.
-In any related execution where the value loaded from \co{y} remains zero,
-nonzero values loaded from \co{x} cannot affect the value stored by
-line~11 to \co{z}.
-Only executions which load different (that is, nonzero) values from
-both \co{x} and \co{y} can cause the value stored to \co{z} to change
-from zero to a nonzero value.
-
-Therefore, as noted in
-Section~\ref{sec:Semantic Dependencies and Source Code},
-in an execution where the loads from both \co{x} and \co{y} return zero,
-there is no semantic dependency from the load from \co{x} to the
-store to \co{z}, nor is there a semantic dependency from the load
-from \co{y} to the store to \co{z}.
-However, there \emph{is} a semantic dependency from the \emph{combination}
-of the loads from \co{x} and \co{y} to the store to \co{z}.
-An example is given by the execution that satisfies the
-\co{exists} clause on line~20, which would be an OOTA cycle.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult-0-cond.litmus @@
-\caption{OOTA Conditional Multiplication Example, Initial Value Zero}
-\label{lst:OOTA Conditional Multiplication Example, Initial Value Zero}
-\end{listing}
-
-For further evidence that there is no semantic dependency from \co{y}
-to \co{z} in executions where the load from \co{x} is zero, see the
-equivalent\footnote{
-	But equivalent only because the atomics are all nonvolatile!}
-program shown in
-Listing~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero}.
-Here, there is not even a load from \co{y} in executions where the
-load from \co{x} returns zero, so there cannot possibly be a semantic
-dependency involving this non-existent load.
-A similar transformation would load \co{x} only in executions where the
-load from \co{y} returned nonzero, which similarly illustrates the lack
-of a semantic dependency between \co{x} and \co{z}.
-
-In both
-Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-and~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero},
-an OOTA cycle having non-zero values (such as those indicated by their
-respective \co{exists} clauses) would need to affect both the \co{x}
-and the \co{y} components of this compound semantic dependency.
-This situation poses a challenge to current work on semantic dependencies,
-which are typically restricted to a single load and store.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult-1.litmus @@
-\caption{OOTA Multiplication Example, Initial Value 1}
-\label{lst:OOTA Multiplication Example, Initial Value 1}
-\end{listing}
-
-In contrast, consider
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1},
-which differs from
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-only in the initial values (1 instead of 0) and the \co{exists}
-clause (0 instead of 1).
-Note that an execution satisfying the updated \co{exists} clause is again
-an OOTA cycle.
-
-Starting with the intuitive execution where the values loaded from \co{x}
-and \co{y} on lines~9 and~10 are the value 1 (that is, nonzero), executions
-that differ only in the value loaded from either \co{x} or \co{y} will
-now affect the value stored to \co{z} by line~11.
-There is therefore one semantic dependency from the load from \co{x}
-to the store to \co{z} and another separate semantic dependency from
-the load from \co{y} to the store to \co{z}, which matches typical
-definitions of semantic dependency.
-Yet if the initial values were instead obtained as input,
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-and
-Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
-could be thought of as being two different executions of the exact same
-program.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-mult3-0.litmus @@
-\caption{OOTA Three-Factor Multiplication Example, Initial Value Zero}
-\label{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
-\end{listing}
-
-Because the only two solutions to $x^2 = x$ are the values zero and one,
-Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
-and~\ref{lst:OOTA Multiplication Example, Initial Value 1}
-each have only two OOTA cycles.
-Note well that although software cannot distinguish the initial
-values from those of an OOTA cycle having those same values, there is
-a very real difference in the corresponding executions.
-To wit, an execution not corresponding to an OOTA cycle will read at
-least one initialization value, while an execution corresponding to an
-OOTA cycle with those same values will read only from that cycle's stores.
-On the other hand, in a similar program having initial values of (say)
-two, both executions containing OOTA cycles would be distinguishable
-from the initial values.
-
-Listing~\ref{lst:OOTA Three-Factor Multiplication Example, Initial Value Zero}
-goes one step further by multiplying the values obtained from three
-different loads.
-Because $x^3 = x$ has three solutions ($\left\{ -1, 0, 1 \right\}$),
-this execution could have up to three OOTA cycles.
-
-Much more elaborate examples can be constructed,
-which raises the question of how to determine the set of semantic
-dependencies in a given execution of a given fragment of code.
-The short answer is that when running on a real-world system, 
-any C++ program is a finite-state machine, which, unlike Turing-complete
-systems, can be analyzed, as demonstrated in
-Section~\ref{sec:Hardware Dependencies, Instruction Ordering, and the Fundamental Property}.
-
-One might ask why semantic dependencies can be so complicated when hardware
-dependencies are much more straightforward.
-The reason is that, unlike compilers, hardware:
-\begin{itemize}
-\item	Respects dependencies even when they reduce to a constant,
-\item	Optimizes much less aggressively, if at all, and
-\item	Lacks undefined behavior.
-\end{itemize}
-This situation further underscores the per-execution nature of semantic
-dependencies as well as the need for an improved definition of ``semantic
-dependency''.
-In
-Section~\ref{sec:A Definition of Semantic Dependency},
-we expand the traditional definition to include dependencies extending
-from groups of loads to a single store.
-
-\subsection{Why rfe Instead of Tried-And-True rf?}
-\label{app:Why rfe Instead of Tried-And-True rf?}
-
-Listings~\ref{lst:OOTA With an Intrathread Store-Load Link}
-and~\ref{lst:OOTA With an Intrathread Store-Load Link, Optimized}
-show the examples from
-Section~\ref{sec:OOTA: rf versus rfe}
-on
-page~\pageref{sec:OOTA: rf versus rfe}
-in the form of litmus tests.
-A compiler is allowed to transform the program in
-Listing~\ref{lst:OOTA With an Intrathread Store-Load Link}
-to the form of
-Listing~\ref{lst:OOTA With an Intrathread Store-Load Link, Optimized},
-which demonstrates that intrathread reads-from (rfi) links may be
-eliminated by the compiler and therefore should not be used in a
-definition of OOTA.
-This section provides a more involved example to drive the point home.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-3-2-proc.litmus @@
-\caption{OOTA With an Intrathread Store-Load Link}
-\label{lst:OOTA With an Intrathread Store-Load Link}
-\end{listing}
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-3-2-proc-opt.litmus @@
-\caption{OOTA With an Intrathread Store-Load Link, Optimized}
-\label{lst:OOTA With an Intrathread Store-Load Link, Optimized}
-\end{listing}
-
-The first step in this direction is
-Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4},
-which expands
-Listing~\ref{lst:Causality Test Case 4}
-% %%% Cite P3064.
-% %%% (discussed in
-% %%% Appendix~\ref{app:Causality Test Case 4})
-from two threads to three.
-The \co{herd7} tool reports OOTA values, and
-all threads have straightforward semantic dependencies from their
-loads to their stores.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-4.litmus @@
-\caption{Causality Test Case 4}
-\label{lst:Causality Test Case 4}
-\end{listing}
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-3proc.litmus @@
-\caption{Three-Process Version of JMM Causality Test Case 4}
-\label{lst:Three-Process Version of JMM Causality Test Case 4}
-\end{listing}
-
-Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
-revises Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
-so that the value stored to \co{y} is forced to be at most 17
-and the value stored to \co{z} is forced to be at least 17.
-This of course means that the value stored to \co{z} must always be
-17, as can be seen in the \co{herd7} output.
-(However, a single-thread analysis could not prove this fact.)
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-whyrfe-3.litmus @@
-\caption{Three-Process Version of JMM Causality Test Case 4 With Max and Min}
-\label{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
-\end{listing}
-
-Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
-further revises the litmus test
-to flatten \co{P0()} and \co{P1()} into a single thread, so that the
-resulting \co{P0()} has an rfi link connecting
-the store to and load from \co{y}, rather than the rfe link in
-Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}.
-As a result, even a single-thread analysis could conclude that the value
-stored to \co{z} will always be 17, by examining \co{P0()} in isolation and
-using the fact that it is valid for a C++ compiler to assume the value
-loaded from \co{y} will always be the value just stored and so to omit
-the load.\footnote{
-	This is not an option in C because in that language,
-	atomic accesses are volatile and hence must be preserved
-	in the machine code.}
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-whyrfe.litmus @@
-\caption{Why rfe Instead of Tried-And-True rf?}
-\label{lst:Why rfe Instead of Tried-And-True rf?}
-\end{listing}
-
-If a compiler makes this observation and optimizes the program by
-storing a constant 17 to \co{z} rather than going through the
-computations on lines~15--17 of
-Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?},
-it will generate an executable which could produce the all-17's result
-described by the \co{exists} clause.
-(See Listing~\ref{lst:Why rfe Instead of Tried-And-True rf When z=17?},
-where the store to \co{z} has been moved up to line~9.\footnote{
-	The load from \co{y} on line~15 is retained even though it is
-	not used, because \co{r3} appears in the \co{exists} clause and
-	hence its final value is considered observable behavior.})
-This would not be an OOTA cycle, because in either form of the program
-the store to~\co{z} in \co{P0} is not semantically dependent on the
-load from~\co{x}, so there is no cycle in (sdep $\cup$ rfe).
-A different compiler that does not make this optimization will
-generate an executable which cannot produce the all-17's result at all,
-because of hardware dependencies and consequent instruction ordering
-in the machine code.
-
-Either way, no OOTA cycle can occur.
-But what if we chose to define an OOTA cycle with (sdep $\cup$ rf) instead
-of (sdep $\cup$ rfe)?
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-whyrfe-z17.litmus @@
-\caption{Why rfe Instead of Tried-And-True rf When z=17?}
-\label{lst:Why rfe Instead of Tried-And-True rf When z=17?}
-\end{listing}
-
-In that case, it would be necessary to consider separately the dependency
-in Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
-connecting \co{P0()}'s load from \co{x} to its store to \co{y},
-and that connecting its load from \co{y} to its store to \co{z}.
-The first is a genuine semantic dependency; the second might or might
-not be, depending on the implementation.
-A compiler that uses single-thread analysis and does not omit the load
-from \co{y} would conclude that it is, because of the
-possibility that the value loaded from \co{y} might be larger than 17.
-From this point of view, the all-17's result \emph{would} be
-considered an OOTA cycle.
-
-This example demonstrates another reason why defining OOTA cycles
-in terms of (sdep $\cup$ rf) is problematic.
-Doing so can lead to classifying a cycle as OOTA even though there
-is no genuine semantic dependency in one of the participating threads.
-
-\subsection{Inventing Atomic Loads}
-\label{app:Inventing Atomic Loads}
-
-This section contains several example litmus tests
-expanding on the material presented in
-Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.
-on
-page~\pageref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.
-
-\subsubsection{Illustrative Atomic-Load Invention}
-\label{app:Illustrative Atomic-Load Invention}
-
-Consider Listing~\ref{lst:No Invented Atomic Loads}.
-At first glance, it might appear straightforward.
-\co{P0()}'s store to \co{z} clearly depends on its load from \co{y},
-and \co{P1()}'s store to \co{y} clearly depends on its load from \co{z}.
-If both dependencies are semantic, the OOTA cycle cannot be realized.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-no-invented-load.litmus @@
-\caption{No Invented Atomic Loads}
-\label{lst:No Invented Atomic Loads}
-\end{listing}
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-load-invented.litmus @@
-\caption{Atomic Loads Invented}
-\label{lst:Atomic Loads Invented}
-\end{listing}
-
-But is \co{P0()}'s dependency truly semantic?
-Imagine what might happen if a C++ compiler is permitted to
-duplicate \co{P0()}'s load from \co{x} and then make use of both loads'
-values.
-This would allow a compiler to transform
-Listing~\ref{lst:No Invented Atomic Loads}
-into
-Listing~\ref{lst:Atomic Loads Invented}.
-
-Because of the comparisons against zero on lines~9 and~10 of
-Listing~\ref{lst:No Invented Atomic Loads},
-the values of \co{r1} and \co{r2} are known to be either zero or one.
-Hence if the values of \co{r1a} and \co{r1b} on lines~9 and~10 of
-Listing~\ref{lst:Atomic Loads Invented}
-differ (say because \co{P2()} changes the value of \co{x}
-during the time between the two loads)
-then one of them must be equal to the value that would
-be obtained for \co{r2}.
-If the compiler is further permitted to make a choice at runtime
-between the two values loaded from \co{x}, it could always choose
-to use for \co{r1} the one that would be equal to \co{r2} (see line~16).
-Then the value stored to \co{z} would always be one,
-as shown on line~14, where the store has been moved before
-the load from \co{y} on line~15 to emphasize the fact that
-it does not depend on that load.
-The dependency from~\co{y} to~\co{z} would not be semantic after all,
-and the outcome \co{r1 == r2 == 1} would indeed be possible
-(although it would not be an example of OOTA because of the lack of a
-semantic dependency).
-
-This is a simple example of a more general phenomenon.
-Suppose $z$ is given by some function $f$ of $x$ and $y$.
-Under what conditions can we say that the value of $z$ doesn't depend
-on $y$?
-The standard answer is that this happens when there are values
-$z_0$ and $x_0$ such that for any $y$, we have $z_0 = f(x_0, y)$.
-But if we are allowed to choose from among multiple values of $x$,
-the situation gets more complicated.
-Then the answer would be that there is a value $z_0$ and
-a set $X$ of values for $x$ such that for any $y$, there is some
-$x \in X$ with $z_0 = f(x, y)$.
-In Listing~\ref{lst:Atomic Loads Invented},
-$f$ is the equality function and $X$ is simply the set $\{0, 1\}$;
-however, the reasoning applies in any situation where $x$ ranges
-over a finite collection of possible values.
-And as the example shows, the more general condition can hold in situations
-where the simpler condition does not, indicating that our intuitive
-notions of semantic dependency are not adequate when there can be
-multiple loads of the same variable.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-unused-load.litmus @@
-\caption{Unused Extra Atomic Load}
-\label{lst:Unused Extra Atomic Load}
-\end{listing}
-
-\subsubsection{Illustrative Atomic-Load Use Invention}
-\label{app:Illustrative Atomic-Load Use Invention}
-
-Clearly something has gone wrong if a valid (in loose C++) transformation
-like this one is capable of destroying what should be an obvious semantic
-dependency.
-One possible reaction is to declare that useful C++ compilers should never
-invent or duplicate nonvolatile relaxed atomic loads.
-But what if the initial code was as shown in
-Listing~\ref{lst:Unused Extra Atomic Load}?\footnote{
-	Unused loads can easily be generated by complex macros or
-	template metaprograms.
-	Typically compilers then remove them, but they are not obliged to.}
-In this case, the transformation to
-Listing~\ref{lst:Atomic Loads Invented}
-would not require inventing or duplicating a load,
-but instead merely inventing a use for
-the previously discarded value in \co{r0}.
-
-There are a number of possible reactions to these situations:
-
-\begin{enumerate}
-\item	Listing~\ref{lst:No Invented Atomic Loads}
-	looks like it has an OOTA cycle, but there is in fact no
-	semantic dependency.
-\item	Attempts to use real-world constraints to prevent OOTA cycles
-	are futile.
-\item	Compilers should not invent uses for values from
-	nonvolatile relaxed atomic loads.
-	If it turns out that there are useful optimizations that invent
-	such uses, then such optimizations must be applied only with
-	careful attention to the as-if rule.
-	Some might argue that inventing comparisons between a pair of
-	nonvolatile relaxed atomic loads from the same object should
-	be permitted only if the two loads could be reordered to be
-	adjacent to each other.
-\item	Any time a compiler might be tempted to invent uses for
-	values of nonvolatile relaxed atomic loads from the same
-	object, the loads should instead be merged into a single load.
-	This reaction might be attractive to those who carefully consider
-	the pointlessness of keeping two loads that return the same value
-	on the one hand or the cache-miss overhead incurred when closely
-	spaced loads return different values on the other.
-\item	Compilers that invent nonvolatile atomic loads or invent
-	new uses for nonvolatile atomic loads can adversely affect
-	observed behavior, for example, by introducing error into code
-	attempting to do statistical sampling.
-\item	What other interesting situations might come to light?
-\end{enumerate}
-
-However, before making such a choice, we should look at other situations.
-
-\subsubsection{Customary Non-Atomic-Load ``Invention''}
-\label{app:Customary Non-Atomic-Load Invention}
-
-There are situations where inventing loads from non-atomic objects
-is customary, for example, when hoisting a load out of a loop.
-In this case, compilers might hoist the load from \co{x} out of
-the loop:
-\begin{quote}
-\begin{verbatim}
-do_something_with(x);
-for (i = 0; i < limit; i++)
-    y[i] += x;
-\end{verbatim}
-\end{quote}
-A compiler might act as if the program had instead been this:
-\begin{quote}
-\begin{verbatim}
-r1 = x;
-do_something_with(r1);
-for (i = 0; i < limit; i++)
-    y[i] += r1;
-\end{verbatim}
-\end{quote}
-Note that passing \co{x} to \co{do_something_with()} is important
-when \co{x} is not atomic,
-because this allows the compiler to assume that inventing additional
-accesses to \co{x} will not result in a data race.
-Without this assumption, the above transformation might introduce a data
-race when \co{limit} is less than or equal to zero.
-Hence the use of quotation marks in the section heading:
-Here the load is not actually invented, but instead merged with an
-earlier load such that there is no synchronization between the two loads.
-
-But when \co{x} is a nonvolatile atomic, there are no worries about
-data races.
-In theory, if the \co{do_something_with()} call was not present
-the compiler could still make the transformation above with no
-qualms, especially given that the value loaded from \co{x} is discarded in
-the case where \co{limit} is less than or equal to zero.
-Except that now the data-race worry is replaced by a cache-miss worry;
-after all, optimizations are supposed to speed things up,
-not slow them down.
-A compiler might be wise to adhere to the
-same restriction that avoids data races for loads from non-atomic
-variables in order to avoid potentially expensive cache misses.
-
-In addition, compiler writers should carefully consider the issues
-raised by the following section.
-
-\subsubsection{Other Cautions for Atomic-Load Invention}
-\label{app:Other Cautions for Atomic-Load Invention}
-
-Consider the code below:
-\begin{quote}
-\begin{verbatim}
-int r0 = x;
-int r1 = r0 * r0 + 2 * r0 + 1;
-\end{verbatim}
-\end{quote}
-For sufficiently small values of \co{x}, the abstract machine is
-guaranteed to produce a perfect square in \co{r1}.
-
-But if the compiler is permitted to invent atomic loads then the compiled
-code might not respect this guarantee, thus incorrectly deviating
-from the abstract machine.
-This issue is evident in a transformed version of the above code:
-\begin{quote}
-\begin{verbatim}
-int r0 = x;
-int invented = x;
-int r1 = r0 * r0 + 2 * invented + 1;
-\end{verbatim}
-\end{quote}
-For example, if the load into \co{r0} happened when the value of \co{x}
-was zero and the load into \co{invented} happened when the value of
-\co{x} was 10, then the value of \co{r1} will be 21, which is not
-a perfect square.
-
-Similar issues can arise even when the compiler merely repurposes a
-value from an atomic load, as can be seen from this example:
-\begin{quote}
-\begin{verbatim}
-do_something(x);
-int r0 = x;
-int r1 = r0 * r0 + 2 * r0 + 1;
-\end{verbatim}
-\end{quote}
-Again, for sufficiently small values of \co{x}, the abstract machine is
-guaranteed to produce a perfect square in \co{r1}.
-So is the compiler allowed to transform this example into the following?
-\begin{quote}
-\begin{verbatim}
-repurposed = x;
-do_something(repurposed);
-int r0 = x;
-int r1 = r0 * r0 + 2 * repurposed + 1;
-\end{verbatim}
-\end{quote}
-This can result in the value of \co{r1} being something other than
-a perfect square, which is incorrect behavior.
-%
-But it would be just fine if the compiler fully merged the two loads from
-\co{x}, for example, as follows:
-\begin{quote}
-\begin{verbatim}
-int r0 = x;
-do_something(r0);
-int r1 = r0 * r0 + 2 * r0 + 1;
-\end{verbatim}
-\end{quote}
-
-Similar observations apply to atomic loads from pointers.
-For example, consider
-Listing~\ref{lst:Example Duplicated Load} on
-page~\pageref{lst:Example Duplicated Load},
-in which the abstract machine guarantees that line~17 passes two
-fields from the same structure to \co{do_something_with()}.
-This guarantee will not always be met if the compiler is permitted to
-invent atomic loads such as that contained in the commented-out line~15.
-\begin{listing}[tbp]
-\scriptsize
-\begin{verbatim}
- 1 struct foo {
- 2   int a;
- 3   int b;
- 4 };
- 5 _Atomic struct foo *globalfoop;
- 6
- 7 void bar()
- 8 {
- 9   int a;
-10   int b;
-11   struct foo *fp;
-12
-13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
-14   do_something(fp);
-15   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
-16   a = fp->a;
-17   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
-18   b = fp->b;
-19   do_something_with(a, b);
-20 }
-\end{verbatim}
-\caption{Example Merging of Duplicated Loads}
-\label{lst:Example Merging of Duplicated Loads}
-\end{listing}
-But suppose that there was an earlier load from \co{globalfoop}, as
-shown on lines~13 and~14 in
-Listing~\ref{lst:Example Merging of Duplicated Loads}.
-Then, assuming that \co{do_something()} is synchronization-free, the
-second load on line~15 of that listing could be removed without resulting
-in incorrect behavior.
-
-Please note that these examples are not just accidents of the current state
-of the C++ abstract machine.
-After all, it is imperative that users be able to rely on the laws of
-arithmetic and on atomic accesses to pointers.
-Therefore, compilers must refrain from inventing atomic loads and from repurposing
-their values.
-However, compilers are permitted to merge adjacent atomic loads from
-the same variable.
-These restrictions correspond to our notion of quasi volatile defined in
-Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
-
-\subsection{Undefined Behavior and Unwise Optimization}
-\label{app:Undefined Behavior and Unwise Optimization}
-
-Many compilers carry out optimizations based on the assumption
-that undefined behavior (UB) cannot happen,
-and if not performed carefully, these optimizations may be invalid
-because of OOTA-like interactions.
-A simple example is shown in
-Listing~\ref{lst:OOTA-Like Behavior Due To Divide-By-Zero UB}.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-div-ub.litmus @@
-\caption{OOTA-Like Behavior Due To Divide-By-Zero UB}
-\label{lst:OOTA-Like Behavior Due To Divide-By-Zero UB}
-\end{listing}
-
-On line~14 of this listing, the computation \co{1 / (r2 <= 0)}
-will yield one if \co{r2} is not positive and will result in UB
-otherwise (a divide-by-zero error).
-It may not be obvious, but the program is in fact UB-free---there are
-no abstract executions in which \co{r2} is greater than zero.
-To see why not, consider that in any such execution
-\co{P1()} would not store anything to \co{y};
-it certainly would not store a positive value.
-This means that the value loaded on line~8 would have to be \co{y}'s
-initial value of zero, and so would the value stored to \co{x} on
-line~9, and hence it would not be possible for the load on line~13
-to get a positive value for \co{r2} in the first place.
-
-Even for a compiler analyzing \co{P1()} in isolation,
-it is clear that either line~14 will store one to \co{y} or
-else the division by zero will cause UB.
-This allows the compiler to assume that the store will always
-take place, on the grounds that UB cannot happen,
-and so the compiler may optimize the comparison and division
-by replacing the whole expression with a constant \co{1}.
-There will then be no dependency from the load on line~13
-to the store and thus nothing forcing the store to execute
-after the load.
-
-When the resulting program runs on a weakly ordered architecture,
-the execution could go as follows:
-\begin{enumerate}
-\item	Line~14 executes out of order, storing the constant \co{1} to \co{y}.
-\item	Line~8 loads one from \co{y}, setting \co{r1} to one.
-\item	Line~9 stores one to \co{x}.
-\item	Line~13 loads one from \co{x}, setting \co{r2} to one.
-\end{enumerate}
-The final result would satisfy the \co{exists} clause,
-exhibiting observable behavior (\co{r2} is one at the end)
-that no abstract execution of the original program could produce.
-(This could happen even if all of the atomic objects in
-Listing~\ref{lst:OOTA-Like Behavior Due To Divide-By-Zero UB}
-were converted to volatile.)
-
-This result shows that the proposed optimization was invalid.
-To prevent the unwanted behavior, the compiler would have to take an
-extra step to force the optimized store on line~14 to be ordered after
-the load on line~13, possibly by changing the load to a load-acquire
-or changing the store to a store-release.
-Hans Boehm has proposed this store-release alternative as his
-Interpretation~B'~\cite{HansBoehm2020UBalternatives}.
-% @@@ Do we want to add a disclaimer about completeness or lack thereof
-% @@@ of this solution?
-
-In other words, the compiler would have to preserve the dependency-induced
-ordering that the UB-based optimization would otherwise destroy.
-It is tempting to consider using techniques
-% Appendix~\ref{app:Aside on Undefined Behavior}
-that prevent back-propagation of UB, but these are ineffective in this
-case because the compiler can detect and exploit UB when
-considering line~14 in isolation.
-
-\clearpage
-
 \bibliographystyle{plain}

### 2024-09-01 paper/ootavacuous.tex
paper/ootavacuous: Remove now-pointless appendix command
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
paper/ootavacuous: Replace litmus-test restatement with reference
Just one Simple OOTA suffices.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-For example, consider the Simple OOTA program, repeated here in
-simplified form:
-\begin{quote}
-\begin{verbatim}
-void thread1() {
-   int r1 = x;
-   y = r1;
-}
-
-void thread2() {
-   int r2 = y;
-   x = r2;
-}
-\end{verbatim}
-\end{quote}
+For example, consider the Simple OOTA program shown in
+Listing~\ref{lst:Simple OOTA}.
 A loose C++ compiler using global analysis and treating \co{x} and \co{y} as
@@
 \clearpage
-\appendix
+% \appendix
 

### 2024-09-02 paper/ootavacuous.sh
asplos: First step conversion to ASPLOS style
Cleanup still todo, including overly wide displays.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -86,8 +86,7 @@ do
 	fi
 done
 
-pdflatex ootavacuous-flat.tex || exit
-bibtex ootavacuous-flat.aux || exit
-pdflatex ootavacuous-flat.tex || exit
-pdflatex ootavacuous-flat.tex || exit
-cp ootavacuous-flat.pdf ootavacuous.pdf
+pdflatex asplos.tex || exit
+bibtex asplos.aux || exit
+pdflatex asplos.tex || exit
+pdflatex asplos.tex || exit

### 2024-09-02 paper/ootavacuous.sh
ootavacuous: Move from \scriptsize to \small
The ASPLOS CFP specifically warns about too-small text, so why
mess around?

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -28,7 +28,7 @@ BEGIN {
 	path = $3;
 	print "% " path;
 	print "{";
-	print bs "scriptsize";
+	print bs "small";
 	print "___EOF___";
 	print "./litmus2table.sh < " path;
 	print "cat << " sq "___EOF___" sq;
@@ -40,7 +40,7 @@ BEGIN {
 	path = $3;
 	print "% " path;
 	print "{";
-	print bs "scriptsize";
+	print bs "small";
 	print "___EOF___";
 	print "./litmus2table.sh < " path;
 	# print "cat << " sq "___EOF___" sq;
@@ -57,7 +57,7 @@ BEGIN {
 /^@@ RunLitmus .* @@$/ {
 	path = $3;
 	print "{";
-	print bs "scriptsize";
+	print bs "small";
 	print bs "begin{verbatim}";
 	print "___EOF___";
 	print "herd7 -c11 " path " | grep -v " sq "^$" sq " | ./c2latex.sh ";

### 2024-09-02 paper/ootavacuous.tex
paper/ootavacuous: Remove page break prior to references
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
paper/ootavacuous: Shrink Caausality Test Cases URL
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
asplos: First step conversion to ASPLOS style
Cleanup still todo, including overly wide displays.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Move from \scriptsize to \small
The ASPLOS CFP specifically warns about too-small text, so why
mess around?

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Fix too-long lines and listings
Signed-off-by

@@
-\documentclass[10]{article}
+% \documentclass[10]{article}
 
@@
 \usepackage{textcomp} % use symbols in TS1 encoding
-\usepackage{mathptmx,helvet,courier} % use nice, standard fonts for roman, sans and monospace respectively
+% \usepackage{mathptmx,helvet,courier} % use nice, standard fonts for roman, sans and monospace respectively
 
@@
 \usepackage{fancyhdr}
-\usepackage{epsfig}
+% \usepackage{epsfig}
 \usepackage{subfigure}
@@
 \usepackage{graphics}
-\usepackage{enumerate}
+% \usepackage{enumerate}
 \usepackage{ifthen}
@@
 % \usepackage{underscore}
-\usepackage[bookmarks=true,bookmarksnumbered=true,pdfborder={0 0 0}]{hyperref}
+% \usepackage[bookmarks=true,bookmarksnumbered=true,pdfborder={0 0 0}]{hyperref}
 
@@
 
-\usepackage[table]{xcolor}
-\usepackage{booktabs}
+% \usepackage[table]{xcolor}
+% \usepackage{booktabs}
 
@@
 
-\pagestyle{fancy}
-\rhead{}
+\pagestyle{plain}
 
@@
 % }
-\maketitle{}
 
@@
 \end{abstract}
+\maketitle{}
 
@@
 {
-\scriptsize
+\small
 \begin{tabular}{r|l|l}
@@
 standard, but it has a problem with intrathread rfi links
-as exemplified by the following code:
-\begin{quote}
-\scriptsize
-\begin{verbatim}
- 1   int r2 = atomic_load_explicit(&x, memory_order_relaxed);
- 2   atomic_store_explicit(&y, r2, memory_order_relaxed);
- 3   int r3 = atomic_load_explicit(&y, memory_order_relaxed);
- 4   atomic_store_explicit(&z, r3, memory_order_relaxed);
-\end{verbatim}
-\end{quote}
-This is an elaboration of \co{thread2()} from
+as exemplified by the following code: \\
+% \begin{quote}
+% Constructed by hand.
+{
+\small
+\begin{tabular}{r|l}
+% & \multicolumn{1}{c}{P0} \\
+% \hline
+1. &
+\texttt{~~int r2 =\textsubscript{rlx} x} \\
+2. &
+\texttt{~~y =\textsubscript{rlx} r2} \\
+3. &
+\texttt{~~int r3 =\textsubscript{rlx} y} \\
+4. &
+\texttt{~~z =\textsubscript{rlx} r3} \\
+\end{tabular}
+} \\
+% \end{quote}
+This is an elaboration of P1 from
 Listing~\ref{lst:Simple OOTA}
@@
 other threads no chance to modify \co{y} in between.
-Such an implementation might therefore behave as if the source code
-had instead been as follows:
-\begin{quote}
-\scriptsize
-\begin{verbatim}
- 1   int r2 = atomic_load_explicit(&x, memory_order_relaxed);
- 2   atomic_store_explicit(&y, r2, memory_order_relaxed);
- 3   // int r3 = atomic_load_explicit(&y, memory_order_relaxed);
- 4   atomic_store_explicit(&z, r2, memory_order_relaxed);
-\end{verbatim}
-\end{quote}
-Here line~3 has been optimized away in favor of line~4 storing the same value
-to \co{z} that was stored to \co{y} by line~2.
+Such an implementation might therefore behave as if line~3 had been
+removed from the source code and line~4's \co{r3} had been replaced by
+\co{r2}.
+This means that line~4 stores the same value to \co{z} that was stored
+to \co{y} by line~2.
 And given that the load from \co{y} no longer exists, it cannot possibly
@@
 This means:
-\begin{itemize}
-\item	The program's output must be the same as that of the abstract
-	execution.
-\item	Volatile accesses ``are evaluated strictly according to the
-	rules of the abstract machine'' (4.1.2p6.1 \co{[intro.abstract]}).
-\item	(There is a condition on the timing and interleaving of input and output,
+(1)~The program's output must be the same as that of the abstract execution;
+(2)~Volatile accesses ``are evaluated strictly according to the
+rules of the abstract machine'' (4.1.2p6.1 \co{[intro.abstract]}); and
+(3)~(There is a condition on the timing and interleaving of input and output,
 which does not matter for our purposes.)
-\end{itemize}
 We will say that the abstract execution is \emph{realized by} the
@@
 a \co{herd7} litmus test that demonstrates such a behavior.
-Without line~8, only the values~0 and~3 can be loaded into \co{r1} on
-line~13.
-With line~8, the additional value~42 can also be loaded into \co{r1}.
+Without P0's line~1, only the values~0 and~3 can be loaded into \co{r1} by
+P1's line~1.
+With P0's line~1, the additional value~42 can also be loaded into \co{r1}.
 The compiler is therefore forbidden from inventing the store of 42
@@
 litmus test that demonstrates such a behavior.
-Without line~9, if the value loaded into \co{r1} is one then the final value
+Without P0's line~1, if the value loaded into \co{r1} is one then the final value
 of \co{x} must be two.
-With line~9, the final value of \co{x} can be one even when the
+With that line, the final value of \co{x} can be one even when the
 value loaded into \co{r1} is one.
@@
 
-\clearpage
+% \clearpage
 % \appendix
@@
 \bibliographystyle{plain}
-\addcontentsline{toc}{section}{References}
+% \addcontentsline{toc}{section}{References}
 \bibliography{bib/RCU,bib/WFS,bib/hw,bib/os,bib/parallelsys,bib/patterns,bib/perfmeas,bib/refs,bib/syncrefs,bib/search,bib/swtools,bib/realtime,bib/TM,bib/standards,bib/memorymodel.bib}

### 2024-09-02 paper/ootavacuous.tex
ootavacuous: Earlier self-review and taglines
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 actions to avoid OOTA.
-In fact, for many compilers the constraints boil down to a simple
-``Don't invent or duplicate atomic accesses'', which the compiler
-probably wouldn't do anyway.
+In fact, for many compilers the constraints boil down to a simple ``Don't
+invent or duplicate atomic accesses'', which the compiler should not
+do anyway if it is to evaluate arithmetic expressions involving atomic
+accesses in accordance with the abstract machine.
 Absolutely no changes are required to either the standard or to user code.
@@
 
-This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
-in using the term \emph{full C++} to denote the standard including the
+This paper follows P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
+in using \emph{full C++} to denote the standard including the
 prohibition of OOTA mentioned above.
-Unlike that article, we will use the term \emph{loose C++}
+Unlike that article, we use \emph{loose C++}
 (rather than \emph{strict C++}, which seems too similar to
@@
 Some of these workers appear to have abandoned this effort, but as of
-2024, Mark Batty is persisting with modular relaxed dependencies.
+2024, Mark Batty is persisting with modular relaxed dependencies, which
+may have the added ability of checking compiler optimizations.
 @@@ cite web site.
@@
 \label{sec:OOTA and Semantic Dependencies}
+
+\emph{Tagline: ``Semantic dependencies are more complicated than you think.''}
+
+% @@@
+% 1: rfe, not rf.
+% 2: Rules of arithmetic (*0).
+% 3: Many-to-one sdep (*0 + ... )
+% 4: Compiler knowledge, cross-thread special case.
+% 5: Different stores.
+% 6: "If" statements.
+% 7: Matching up stores.
+% Semantic dependencies depend on execution.
+% @@@
 
@@
 
+\emph{Tagline: ``Need to look at machine code.''}
+
 This section looks more carefully at executions, in some cases revisiting
@@
 
+\emph{Tagline: Compilers are constrained in optimizing atomics, especially
+volatile atomics.}
+
 A complete C++ implementation consists of much more than just a compiler.
@@
 
+% @@@ Needed?  If so, can this be in sdep section?
+
 Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
@@
 \co{y} while \co{thread2()} does the reverse.
-A globally optimizing loose C++ compiler given that program might
-transform it to the following before translating it into machine code,
-if the compiler is sufficiently perverse:
-\begin{quote}
-\begin{verbatim}
- 1 atomic<int> x(0);
- 2 atomic<int> y(0);
- 3
- 4 void thread1()
- 5 {
- 6   int r1 = 42;
- 7   y.store(r1, memory_order_relaxed);
- 8 }
- 9
-10 void thread2()
-11 {
-12   int r2 = 42;
-13   x.store(r2, memory_order_relaxed);
-14 }
-\end{verbatim}
-\end{quote}
+A sufficiently perverse globally optimizing loose C++ compiler might
+transform that program to the following before translating it into
+machine code: \\
+% Constructed by hand.
+{
+\small
+\begin{tabular}{r|l|l}
+% & \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
+% \hline
+1. &
+\texttt{~~int r1 = 42} &
+	\texttt{~~int r2 = 42} \\
+2. &
+\texttt{~~y =\textsubscript{rlx} r1} &
+	\texttt{~~z =\textsubscript{rlx} r2} \\
+\end{tabular}
+} \\
 The loads previously on lines~6 and~12 have been replaced by constants.
@@
 \label{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}
+
+\emph{ @@@ Merge into earlier discussion of inventing atomic loads. }
 
@@
 
+\emph{Tagline: ``How to think about semantic dependencies.'' }
+
 Section~\ref{sec:Dependencies at the Hardware Level}
@@
 \end{quote}
-It's quite straightforward to show that under any implementation in which
-semantic dependencies satisfy the Fundamental Property, no OOTA cycle
-has a nontrivial realization.
-
-Indeed, suppose that abstract execution $A$ is realized by hardware
-execution $H$ and $A$ has an OOTA cycle.
+
+No implementation in which semantic dependencies satisfy the Fundamental
+Property has a nontrivial OOTA cycle.
+To see this, suppose that abstract execution $A$ is realized by hardware
+execution $H$ and $A$ does have an OOTA cycle.
 This means there are atomic stores $W_i$ in $A$,
@@
 But now we have a contradiction:
-\begin{itemize}
-\item	$W_0'$ commits after $R_{0,0}'$;
-\item	$R_{0,0}'$ commits after the store instruction $W_k'$ it reads from;
-\item	$W_k'$ commits no earlier than $W_0'$.
-\end{itemize}
+(1)~$W_0'$ commits after $R_{0,0}'$;
+(2)~$R_{0,0}'$ commits after the store instruction $W_k'$ it reads from;
+and finally
+(3)~$W_k'$ commits no earlier than $W_0'$.
 
@@
 \label{sec:A Definition of Semantic Dependency}
+
+\emph{ Tagline: ``How to think about OOTA.'' }
 

### 2024-09-02 paper/ootavacuous.tex
ootavacuous: Self-review part 1
Move compiler-related examples to the C++ Compilers section.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	``quasi-volatile'' behavior that we expect C++ implementations
-	will adhere to if they perform single-thread analysis.
+	will adhere to if they confine themselves to single-thread analysis.
 \end{abstract}
@@
 It is important to distinguish true OOTA cycles from OOTA-like
-behavior caused by simple reordering.
-An example of simple reordering is shown in
+behavior caused by simple reordering, such as that shown in
 Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
@@
 but none of these have been looked upon favorably by C++ implementers.
-Some of these workers appear to have abandoned this effort, but as of
+Many of these workers appear to have abandoned this effort, but as of
 2024, Mark Batty is persisting with modular relaxed dependencies, which
@@
 % concerns.
-This document will concentrate on examples lacking UB.
+This paper focuses on examples lacking UB.
 
@@
 then discusses properties of semantic dependences, showing that they
-are functions of executions rather than strictly of source code,
-among other things.
+are not strict functions of source code.
 
@@
 
-In order to avoid these rfi links we will substitute rfe
+To avoid this issue, we will substitute rfe
 for rf, defining an OOTA cycle---for now---as a cycle in sdep $\cup$ rfe.
@@
 and thus no impediment to the occurrence of an OOTA cycle.
-We will see examples of this destruction in
-Sections~\ref{sec:Global Optimization Can Destroy Dependencies}
-and~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies} below.
+Later, we will see examples of this destruction.
 
@@
 such as basing a dependency on the time of day or a process ID.
-
-The implementation-defined aspects can affect whether or not an
-abstract execution contains a semantic dependency.
-As an example consider the following, where the type of \co{c} is \co{char}:
-\begin{quote}
-\begin{verbatim}
-y = (c >= 0);
-\end{verbatim}
-\end{quote}
-Here \co{y} is semantically dependent on \co{c} in executions
-running on implementations in
-which \co{char} is a signed type,
-but not those for which it is unsigned.
-
-The same is true for the nondeterministic aspects of an execution.
-Consider this example, with \co{i} initially zero:
-\begin{quote}
-\begin{verbatim}
-int foo(int a, int b)
-{
-   return a / b;
-}
-
-r1 = foo(++i, ++i);
-x = r1 * z;
-\end{verbatim}
-\end{quote}
-Because early C~implementers could not come to agreement, the standard
-does not specify the order of evaluation of function arguments, so
-the value calculated for \co{r1} might be zero ($1/2$ truncated) or two
-($2/1$).
-In the former case there is no semantic dependency from \co{z} to \co{x},
-but in the latter case there is.\footnote{
-	Thanks to Peter Sewell for pointing out this possibility.}
-
-(According to the current version of the standard, conflicting side effects
-in unsequenced subexpressions constitute undefined behavior,
-although there are proposals to make them defined in both
-C++~\cite{GabrielDosReis2016P0145r3}
-and C~\cite{AlexCeleste2023N3203}.
-Nevertheless, the example above is allowed because the order of evaluation
-of arguments to a function call is ``indeterminately sequenced''
-(7.6.1.3p7 \co{[expr.call]}) rather than unsequenced, a subtle distinction.)
-
-\medskip
+In short, the standard grants C++ implementations considerable freedom,
+as detailed in
+Section~\ref{sec:C++ Compilers}.
 
@@
 allowed by the loose C++ abstract machine will ever be realized
-by the executables produced by many compilers,
+by the executables produced by many compilers.
 
@@
 
-\emph{Tagline: Compilers are constrained in optimizing atomics, especially
-volatile atomics.}
+\emph{Tagline: Compilers have considerable freedom, but are constrained
+in optimizing atomics, especially volatile atomics.}
 
@@
 
+Section~\ref{sec:Architectures Influence the Behavior of Compilers}
+show that machine architectures influence C++ compilers.
+Section~\ref{sec:Optimizations Influence the Behavior of Compilers}
+show, perhaps unsurprisingly, that optimizers influence C++ compilers.
 Section~\ref{sec:Users Influence the Behavior of Compilers}
-shows that C++ compilers can be influenced by their users as well
-as by the standard.
+shows that compilers can also be influenced by their users.
 Section~\ref{sec:Global Optimization Can Destroy Dependencies}
@@
 also defines properties of quasi-volatile atomic operations.
+
+\subsection{Architectures Influence the Behavior of Compilers}
+\label{sec:Architectures Influence the Behavior of Compilers}
+
+The implementation-defined aspects of the standard permit architectures to
+differ as to whether an abstract execution contains a semantic dependency.
+As an example consider the following, where the type of \co{c} is \co{char}:
+\begin{quote}
+\begin{verbatim}
+y = (c >= 0);
+\end{verbatim}
+\end{quote}
+Here \co{y} is semantically dependent on \co{c} in executions
+running on implementations in
+which \co{char} is a signed type,
+but not those for which it is unsigned.
+
+\subsection{Optimizations Influence the Behavior of Compilers}
+\label{sec:Optimizations Influence the Behavior of Compilers}
+
+Compiler optimizers also exploit the nondeterministic aspects of an
+execution.
+Consider this example, with \co{i} initially zero:
+\begin{quote}
+\begin{verbatim}
+int foo(int a, int b)
+{
+   return a / b;
+}
+
+r1 = foo(++i, ++i);
+x = r1 * z;
+\end{verbatim}
+\end{quote}
+Because early C~implementers could not come to agreement, the standard
+does not specify the order of evaluation of function arguments, so
+the value calculated for \co{r1} might be zero ($1/2$ truncated) or two
+($2/1$).
+In the former case there is no semantic dependency from \co{z} to \co{x},
+but in the latter case there is.\footnote{
+	Thanks to Peter Sewell for pointing out this possibility.}
+
+(According to the current version of the standard, conflicting side effects
+in unsequenced subexpressions constitute undefined behavior,
+although there are proposals to make them defined in both
+C++~\cite{GabrielDosReis2016P0145r3}
+and C~\cite{AlexCeleste2023N3203}.
+Nevertheless, the example above is allowed because the order of evaluation
+of arguments to a function call is ``indeterminately sequenced''
+(7.6.1.3p7 \co{[expr.call]}) rather than unsequenced, a subtle distinction.)
 

### 2024-09-02 paper/ootavacuous.tex
ootavacuous: Remove \medskip directives
We might want some of them back, but given the differences in LaTeX
template, most likely somewhere else.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Merge discussions of inventing atomic loads
Also add questions about large-animal text shrinkage.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\medskip
-
 We have seen several examples showing that semantic dependencies may vary
@@
 
-% @@@ Needed?  If so, can this be in sdep section?
-
 Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
@@
 
-\emph{ @@@ Merge into earlier discussion of inventing atomic loads. }
-
-Consider this code, in which the final values of \co{r1} and~\co{r2}
-are observable:
+Invented atomic loads are problematic:
+\begin{quote}
+\begin{verbatim}
+int r0 = x;
+int r1 = r0 * r0 + 2 * r0 + 1;
+\end{verbatim}
+\end{quote}
+For sufficiently small values of \co{x}, the abstract machine is
+guaranteed to produce a perfect square in \co{r1}.
+
+But if the compiler is permitted to invent atomic loads then the compiler
+might transform this code as follows:
+\begin{quote}
+\begin{verbatim}
+int r0 = x;
+int invented = x;
+int r1 = r0 * r0 + 2 * invented + 1;
+\end{verbatim}
+\end{quote}
+If the load into \co{r0} happened when the value of \co{x}
+was zero and the load into \co{invented} happened when the value of
+\co{x} was 10, then the value of \co{r1} will be 21, which is not
+a perfect square.
+This fails to provide the abstract machine's guarantees.
+
+Inventing atomic loads can also result in OOTA-like behavior.
+For example, consider this code, in which the final values of \co{r1}
+and~\co{r2} are observable:
 \begin{quote}
@@
 
-A noteworthy aspect of this transformation is that it invents an atomic load:
-The original form of the code reads \co{x} only once,
-whereas the transformed code reads it twice.
-Therefore we can rule out transformations like this one by insisting
+Therefore we should rule out transformations like these by insisting
 the compiler not invent (or duplicate) atomic loads.
-Please note that invented and duplicated atomic loads are also problematic
-in arithmetic:
-\begin{quote}
-\begin{verbatim}
-int r0 = x;
-int r1 = r0 * r0 + 2 * r0 + 1;
-\end{verbatim}
-\end{quote}
-For sufficiently small values of \co{x}, the abstract machine is
-guaranteed to produce a perfect square in \co{r1}.
-
-But if the compiler is permitted to invent atomic loads then the compiled
-code might not respect this guarantee, thus incorrectly deviating
-from the abstract machine.
-This issue is evident in a transformed version of the above code:
-\begin{quote}
-\begin{verbatim}
-int r0 = x;
-int invented = x;
-int r1 = r0 * r0 + 2 * invented + 1;
-\end{verbatim}
-\end{quote}
-For example, if the load into \co{r0} happened when the value of \co{x}
-was zero and the load into \co{invented} happened when the value of
-\co{x} was 10, then the value of \co{r1} will be 21, which is not
-a perfect square.\footnote{
-	See also the ``Invented Atomic Loads'' paragraph of
-        Section~\ref{sec:Constraints of the Standard}.}
-
 In fact, we will require that atomic accesses be treated as
@@
 	stored in a status register).
-\item	A conditional-move instruction would have as inputs the
-	condition-code bits to test, the source register value, and the
-	target register value; the output would be the value to be stored
-	in the target register.
-\item	A conditional- or computed-branch instruction would have as inputs
-	the condition-code bits to test, the address of the following
-	instruction (to be used if the condition is false) and the
-	destination address (to be used if the condition is true).
-	The output is the new address to be written to the
-	instruction-pointer register.
+% \item	A conditional-move instruction would have as inputs the
+%	condition-code bits to test, the source register value, and the
+%	target register value; the output would be the value to be stored
+%	in the target register.
+% \item	A conditional- or computed-branch instruction would have as inputs
+%	the condition-code bits to test, the address of the following
+%	instruction (to be used if the condition is false) and the
+%	destination address (to be used if the condition is true).
+%	The output is the new address to be written to the
+%	instruction-pointer register.
 \item	A memory-load instruction's input is the address to load from,
@@
 QED.
-
-\medskip
 
@@
 
-\medskip
-
 Despite these outstanding issues, our definition of semantic dependency
@@
 \label{sec:Hardware Architecture and Design}
+
+\emph{Can we summarize this section, especially given that the ASPLOS
+audience can be expected to have a greater understanding of these issues
+than the C++ standards committee?}
 
@@
 
-\medskip
-
 \begin{listing}[tbp]
@@
 \label{sec:Constraints of the Standard}
+
+\emph{Can we summarize this section with a list of permitted and prohibited
+optimizations in the discussion of quasi-volatile?}
 
@@
 
-\paragraph{Invented Atomic Loads}
-\begin{listing}[tbp]
-\scriptsize
-\begin{verbatim}
- 1 struct foo {
- 2   int a;
- 3   int b;
- 4 };
- 5 _Atomic struct foo *globalfoop;
- 6
- 7 void bar()
- 8 {
- 9   int a;
-10   int b;
-11   struct foo *fp;
-12
-13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
-14   a = fp->a;
-15   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
-16   b = fp->b;
-17   do_something_with(a, b);
-18 }
-\end{verbatim}
-\caption{Example Duplicated Load}
-\label{lst:Example Duplicated Load}
-\end{listing}
-Although one can argue that it is functionally correct to invent atomic
-loads as long as the loaded value is not used, doing so can result in
-extra cache misses for no good purpose.
-And duplicating an atomic load can also lead to incorrect results if done
-carelessly.
-To see this, consider
-Listing~\ref{lst:Example Duplicated Load}.
-Suppose that the load on line~13 were duplicated, for example, as if the
-compiler had uncommented line~15.
-Then the values of \co{a} and \co{b} passed to line~17's call to
-\co{do_something_with()} might be from different instances of the
-\co{foo} structure, something that \co{do_something_with()} probably
-isn't prepared to deal with and that represents incorrect behavior,
-that is, behavior that the abstract machine does not exhibit.
-% @@@ See
-% @@@ Appendix~\ref{app:Inventing Atomic Loads}
-% @@@ for more details on this example and other examples of invented
-% @@@ atomic loads.
-
-For these reasons, C++ compilers should avoid inventing atomic
-loads, and in fact should avoid even duplicating them.
+In contrast, compilers should not duplicate, repurpose, or otherwise
+invent atomic loads, as discussed in
+Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.
 
@@
 \label{sec:Semantic Dependencies and Tooling}
+
+\emph{Do we really need this?}
 

### 2024-09-03 paper/ootavacuous.tex
ootavacuous: Drop "rlx" subscripts from litmus tests
Woo-hoo!!!  They are are now all relaxed accesses, so no need to
differentiate any more.

Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 1. &
-\texttt{~~int r1 =\textsubscript{rlx} x} &
-	\texttt{~~int r2 =\textsubscript{rlx} y} \\
+\texttt{~~int r1 = x} &
+	\texttt{~~int r2 = y} \\
 2. &
-\texttt{~~y =\textsubscript{rlx} r1} &
-	\texttt{~~x =\textsubscript{rlx} r2} \\
+\texttt{~~y = r1} &
+	\texttt{~~x = r2} \\
 \end{tabular}
@@
 an outcome~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
-The ``=\textsubscript{rlx}'' denotes a relaxed access, with variables
-whose names are ``\co{r}'' followed by a number denoting registers,
-and other names denoting shared variables.
+Throughout this paper, variables whose names are ``\co{r}'' followed by
+a number denote registers, and other names denote shared variables.
+All accesses to shared variable are relaxed atomic accesses.
 First, P0's line~1 does a relaxed atomic load from \co{x} into \co{r1},
@@
 1. &
-\texttt{~~int r1 =\textsubscript{rlx} x} &
-	\texttt{~~int r2 =\textsubscript{rlx} y} \\
+\texttt{~~int r1 = x} &
+	\texttt{~~int r2 = y} \\
 2. &
-\texttt{~~y =\textsubscript{rlx} r1} &
-	\texttt{~~x =\textsubscript{rlx} 42} \\
+\texttt{~~y = r1} &
+	\texttt{~~x = 42} \\
 \end{tabular}
@@
 1. &
-\texttt{~~int r2 =\textsubscript{rlx} x} \\
+\texttt{~~int r2 = x} \\
 2. &
-\texttt{~~y =\textsubscript{rlx} r2} \\
+\texttt{~~y = r2} \\
 3. &
-\texttt{~~int r3 =\textsubscript{rlx} y} \\
+\texttt{~~int r3 = y} \\
 4. &
-\texttt{~~z =\textsubscript{rlx} r3} \\
+\texttt{~~z = r3} \\
 \end{tabular}
@@
 \end{quote}
-(Here and below, we have written shared-variable accesses without
-annotations, for brevity.  Please imagine they are all relaxed atomic.)
 
@@
 2. &
-\texttt{~~y =\textsubscript{rlx} r1} &
-	\texttt{~~z =\textsubscript{rlx} r2} \\
+\texttt{~~y = r1} &
+	\texttt{~~z = r2} \\
 \end{tabular}

### 2024-09-03 paper/litmus2table.sh

@@ -39,13 +39,19 @@ inproc != 0 && $0 != "}" {
 		moa = memorder[moa];
 	}
 	if (stmt ~ /= atomic_load_explicit\(/) {
-		gsub(/= atomic_load_explicit\(/, "=" bs "textsubscript{" moa "} ", stmt);
+		# Uncomment next statement to get memory-order subscripts.
+		# gsub(/= atomic_load_explicit\(/, "=" bs "textsubscript{" moa "} ", stmt);
+		gsub(/= atomic_load_explicit\(/, "= ", stmt);
+		# Comment previous statement to get memory-order subscripts.
 		if (gsub(/, memory_order_[a-z_]*\);/, "", stmt) == 0)
 			gsub(/, memory_order_[a-z_]*\)/, "", stmt);
 	} else if (stmt ~ /atomic_store_explicit\(/) {
 		gsub(/atomic_store_explicit\(/, "", stmt);
 		gsub(/, memory_order_[a-z_]*\);/, "", stmt);
-		gsub(/,/, " =" bs "textsubscript{" moa "}", stmt);
+		# Uncomment next statement to get memory-order subscripts.
+		# gsub(/,/, " =" bs "textsubscript{" moa "}", stmt);
+		gsub(/,/, " = ", stmt);
+		# Comment previous statement to get memory-order subscripts.
 	}
 	stmtlist[++stmtno][procnum] = stmt;
 	# print procnum ":" stmtno, stmt

### 2024-09-03 paper/ootavacuous.tex
ootavacuous: Fix obsolete thread0/thread1 uses for Listing 1
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 page~\pageref{lst:Simple OOTA},
-in which \co{thread1()} loads the value of \co{x} and stores it in
-\co{y} while \co{thread2()} does the reverse.
+in which P0 loads the value of \co{x} and stores it in \co{y} while P1
+does the reverse.
 A sufficiently perverse globally optimizing loose C++ compiler might

### 2024-09-03 paper/ootavacuous.tex
paper: Summarize "roadmap" paragraphs
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\section{OOTA and Semantic Dependencies}
-\label{sec:OOTA and Semantic Dependencies}
+\section{OOTA and Semantic Dependencies (sdep)}
+\label{sec:OOTA and Semantic Dependencies (sdep)}
 
@@
 
-% @@@
-% 1: rfe, not rf.
-% 2: Rules of arithmetic (*0).
-% 3: Many-to-one sdep (*0 + ... )
-% 4: Compiler knowledge, cross-thread special case.
-% 5: Different stores.
-% 6: "If" statements.
-% 7: Matching up stores.
-% Semantic dependencies depend on execution.
-% @@@
-
-Section~\ref{sec:OOTA: rf versus rfe}
-demonstrates advantages of formulating an OOTA cycle as a cycle in
-sdep $\cup$ rfe instead of the traditional sdep $\cup$ rf.
-Section~\ref{sec:Properties of Semantic Dependencies}
-then discusses properties of semantic dependences, showing that they
-are not strict functions of source code.
+This section formulates OOTA cycles as cycles in sdep $\cup$ rfe instead
+of the traditional sdep $\cup$ rf, then shows that sdep is not a strict
+function of the source code.
 
@@
 
-This section looks more carefully at executions, in some cases revisiting
-example code from
-Section~\ref{sec:Properties of Semantic Dependencies}.
-Section~\ref{sec:Abstract Executions}
-looks at executions from the viewpoint of the abstract machine, and
-Section~\ref{sec:Hardware Executions}
-looks at them from the viewpoint of the computer hardware.
-Finally,
-Section~\ref{sec:Relation Between Abstract and Hardware Executions}
-describes how to relate these two viewpoints.
+This section looks more carefully at executions from the viewpoint of
+the abstract machine and the computer hardware, and then reconciles
+these two viewpoints.
 
@@
 
-A complete C++ implementation consists of much more than just a compiler.
-Among other things, for example, it might have a collection of
+In addition to a compiler, a complete C++ implementation might include
 \co{.h} header files, a linker, runtime libraries, and a dynamic loader.
@@
 a C++ source program to directly executable machine code.
-We will therefore use the terms ``compiler'' and ``implementation''
-interchangeably.
-
-Section~\ref{sec:Architectures Influence the Behavior of Compilers}
-show that machine architectures influence C++ compilers.
-Section~\ref{sec:Optimizations Influence the Behavior of Compilers}
-show, perhaps unsurprisingly, that optimizers influence C++ compilers.
-Section~\ref{sec:Users Influence the Behavior of Compilers}
-shows that compilers can also be influenced by their users.
-Section~\ref{sec:Global Optimization Can Destroy Dependencies}
-presents an example showing that global optimizations can destroy
-semantic dependencies, and then
-Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}
-analyzes examples showing that inventing atomic loads can destroy
-semantic dependencies.
-Section~\ref{sec:Volatile and Quasi Volatile Accesses}
-then presents required properties of volatile atomic operations, and
-also defines properties of quasi-volatile atomic operations.
+We will therefore use ``compiler'' and ``implementation'' interchangeably.
+
+This section discusses influences and constraints on C++ compilers,
+then uses this information to define necessary properties of
+\emph{quasi-volatile} atomic operations.
 
@@
 differ as to whether an abstract execution contains a semantic dependency.
-As an example consider the following, where the type of \co{c} is \co{char}:
+For example, consider the following, where the type of \co{c} is \co{char}:
 \begin{quote}
@@
 
-Section~\ref{sec:Dependencies at the Hardware Level}
-discusses hardware-level dependencies involving machine instructions.
-Section~\ref{sec:Instruction Ordering}
-then examines the relationships between various instruction-ordering
-mechanisms and semantic dependencies.
-This material feeds into
-Section~\ref{sec:The Fundamental Property of Semantic Dependencies},
-which presents the fundamental property of semantic dependencies
-and related complications.
+This section examines dependencies at the machine-instruction level,
+then combines this with mapping from the previous section to define the
+fundamental property of semantic dependencies.
 
@@
 more tractable.
-
-Section~\ref{sec:For Compilers Using Single-Thread Analysis}
-focuses on compilers that restrict their analysis to a single
-thread,
-Section~\ref{sec:For Compilers Using Global Analysis}
-relaxes this single-thread restriction,
-Section~\ref{sec:Verifying the Fundamental Property}
-verifies the fundamental property of semantic dependencies,
-and
-Section~\ref{sec:Outstanding Issues}
-discusses general questions regarding our definitions.
+This insight allows us to demonstrate that OOTA cycles cannot form
+in the C++ implementations addressed by this paper.
 

### 2024-09-03 paper/ootavacuous.tex
asplos: Promote "Outstanding Issues" to section
And rename to "Issues and Refinements".

Perhaps later this is merged with "Future Directions".  But one step at
a time.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\subsection{Outstanding Issues}
-\label{sec:Outstanding Issues}
-
-Here we consider some general questions related to our definition of
-semantic dependency.
-
-\subsubsection{Relative versus Absolute Dependency}
-\label{sec:Relative versus Absolute Dependency}
-
-A drawback of the definition is that it is only relative to a specific
-compiler or implementation.
-This may strike some people as wishy-washy and avoiding the real
-problem, in that a given piece of code should either contain or not
-contain a semantic dependency, independent of the implementation used to
-run it.
-
-We can address this drawback by defining an \emph{absolute semantic
-dependency} as one that is present relative to any valid loose C++
-implementation, past, present, or future, real or imagined.
-Of course this notion has its own problems, including that it is
-extremely nonconstructive and impossible to apply in practice.
-However it may be the best we can do with our current understanding
-of computing systems.
-
-There is one thing we can definitely state:
-Programs produced by an implementation of the right sort will never
-exhibit absolute OOTA (that is, an OOTA cycle in which all the
-semantic dependencies are absolute).
-This is simply because an absolute semantic dependency is {\it
-a fortiori\/} a semantic dependency relative to the compiler in use.
-
-But in fact we have shown more than this:
-Programs will never exhibit an OOTA cycle relative to the compiler
-used to build them, even when that cycle is not absolute.
-In this sense we have gone beyond the requirement of full C++.
-
-\subsubsection{Global Analysis and Volatile versus Quasi Volatile}
-\label{sec:Global Analysis and Volatile versus Quasi Volatile}
-
-In principle we don't need to require global-analysis compilers to
-treat atomic objects as volatile; our results would still hold if they
-merely treated them as quasi volatile.
-We chose not to do this because it would violate our intuitions about
-semantic dependencies.
-
-For example, consider the Simple OOTA program shown in
-Listing~\ref{lst:Simple OOTA}.
-A loose C++ compiler using global analysis and treating \co{x} and \co{y} as
-quasi-volatile objects could omit the two loads, replacing them in the
-machine code with simple assignments ``\co{r1 = 42}'' and
-``\co{r2 = 42}''.
-This would be a valid transformation, and the resulting behavior would
-not count as OOTA according to our definition because the
-dependencies in \co{thread1} and \co{thread2} would not be semantic.
-
-To see why not, recall that a semantic dependency must have a witness,
-another execution in which the store acts differently.
-But this transformed program has no other hardware executions; every time it
-runs it will store 42 to both \co{x} and \co{y}.
-(Keep in mind also that since the atomic objects are not treated as
-volatile, they are not subject to unspecified interference when the
-program runs in a nonbenign environment.)
-
-This unintuitive behavior could not occur if the two loads were not
-omitted.
-In fact, the definition of semantic dependency might remain
-perfectly acceptable if the requirement for global-analysis compilers
-were weakened, if the compiler treated atomic objects as quasi
-volatile and in addition was not allowed to omit accesses to them.
-This is a possible topic for future research.
-
-\subsubsection{Effect of Memory Layout}
-\label{sec:Effect of Memory Layout}
-
-Part of our demonstration of the Fundamental Property of semantic
-dependencies relies on the fact, stated in
-Section~\ref{sec:Abstract Executions},
-that an abstract execution of a thread is entirely determined by the
-values obtained for its loads.
-But when we compare abstract executions of the same thread in two
-different programs, this may no longer be entirely true owing to the
-effect of differing memory layouts.
-
-Consider this simple example:
-\begin{quote}
-\begin{verbatim}
-x = (int) &x;
-\end{verbatim}
-\end{quote}
-Even though the example contains no loads at all, it may store
-different values when running in different programs because the
-object \co{x} may be allocated at differing addresses in those
-programs.
-According to our definition, this could count as a degenerate OOTA
-cycle of length one, in which the store is semantically dependent on
-an empty set of loads!
-
-To rule out such pathological counterexamples we should require that
-in a witness to a semantic dependency, the addresses of all the
-objects and functions referred to in the thread $T$ are the same as
-in the original execution.
-This is a very technical restriction but there are occasions when
-the issue might realistically arise, such as when computing a hash
-value based on an object's address.
-
-\subsubsection{Merging Quasi-Volatile Loads}
-\label{sec:Merging Quasi-Volatile Loads}
-
-The compiler is permitted to merge quasi-volatile loads.
-This can lead to surprising results because a particular load
-may be merged with an earlier load in one execution and with a
-later load in another.
-This is demonstrated in the following, which is a variant of
-the example in
-Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}:
-\begin{quote}
-\begin{verbatim}
-int r1 = (x != 0);
-int r2 = (x != 0);
-int r3 = (x != 0);
-int r4 = (y != 0);
-z = (r2 == r4);
-\end{verbatim}
-\end{quote}
-Consider an abstract execution in which \co{r1}, \co{r2}, and \co{r4}
-are zero and \co{r3} is one (because another thread changed the value
-of \co{x} between two of the loads).
-We would expect that the store to \co{z} would be semantically
-dependent on the load of \co{y} in this execution.
-
-However, a single-thread analysis compiler can translate this into the
-machine-code equivalent of:
-\begin{quote}
-\begin{verbatim}
-int r1 = (x != 0);
-int r2;
-int r3 = (x != 0);
-int r4 = (y != 0);
-if (r1 != r3) {
-   r2 = r4;
-   z = 1;
-} else {
-   r2 = r1;
-   z = (r2 == r4);
-}
-\end{verbatim}
-\end{quote}
-In effect, the \co{r2} load is merged with the \co{r1} load in
-executions where \co{r1} is equal to \co{r3} or to \co{r4}, and it is
-merged with the \co{r3} load in other executions.
-
-To demonstrate the semantic dependency in the original code,
-a suitable witness would have
-to include a hardware realization of the abstract execution in which
-\co{r1} and \co{r2} are zero and \co{r3} and \co{r4} are one.
-But there are no hardware realizations of this execution with the
-machine code indicated above!
-Since \co{r1} is different from both \co{r3} and \co{r4}, the \co{r2}
-load will be merged with the \co{r3} load and so \co{r2} will
-necessarily be one, not zero.
-Thus, relative to this compiler the example does not contain a
-semantic dependency.
-
-Although it is unexpected, we cannot say that this conclusion is
-definitely wrong, because our intuitive notions of semantic
-dependency are not clear in cases where multiple loads of the same
-variable are present.
-% Can we instead say that transforming the code might require that the
-% witness also be transformed?
-
-Despite these outstanding issues, our definition of semantic dependency
-allows us to bring to bear the real-world constraints outlined in the
-next section.
-
 \section{Real-World Constraints}
@@
 
+\section{Issues and Refinements}
+\label{sec:Issues and Refinements}
+
+Here we consider some general questions related to our definition of
+semantic dependency.
+
+\subsection{Relative versus Absolute Dependency}
+\label{sec:Relative versus Absolute Dependency}
+
+A drawback of the definition is that it is only relative to a specific
+compiler or implementation.
+This may strike some people as wishy-washy and avoiding the real
+problem, in that a given piece of code should either contain or not
+contain a semantic dependency, independent of the implementation used to
+run it.
+
+We can address this drawback by defining an \emph{absolute semantic
+dependency} as one that is present relative to any valid loose C++
+implementation, past, present, or future, real or imagined.
+Of course this notion has its own problems, including that it is
+extremely nonconstructive and impossible to apply in practice.
+However it may be the best we can do with our current understanding
+of computing systems.
+
+There is one thing we can definitely state:
+Programs produced by an implementation of the right sort will never
+exhibit absolute OOTA (that is, an OOTA cycle in which all the
+semantic dependencies are absolute).
+This is simply because an absolute semantic dependency is {\it
+a fortiori\/} a semantic dependency relative to the compiler in use.
+
+But in fact we have shown more than this:
+Programs will never exhibit an OOTA cycle relative to the compiler
+used to build them, even when that cycle is not absolute.
+In this sense we have gone beyond the requirement of full C++.
+
+\subsection{Global Analysis and Volatile versus Quasi Volatile}
+\label{sec:Global Analysis and Volatile versus Quasi Volatile}
+
+In principle we don't need to require global-analysis compilers to
+treat atomic objects as volatile; our results would still hold if they
+merely treated them as quasi volatile.
+We chose not to do this because it would violate our intuitions about
+semantic dependencies.
+
+For example, consider the Simple OOTA program shown in
+Listing~\ref{lst:Simple OOTA}.
+A loose C++ compiler using global analysis and treating \co{x} and \co{y} as
+quasi-volatile objects could omit the two loads, replacing them in the
+machine code with simple assignments ``\co{r1 = 42}'' and
+``\co{r2 = 42}''.
+This would be a valid transformation, and the resulting behavior would
+not count as OOTA according to our definition because the
+dependencies in \co{thread1} and \co{thread2} would not be semantic.
+
+To see why not, recall that a semantic dependency must have a witness,
+another execution in which the store acts differently.
+But this transformed program has no other hardware executions; every time it
+runs it will store 42 to both \co{x} and \co{y}.
+(Keep in mind also that since the atomic objects are not treated as
+volatile, they are not subject to unspecified interference when the
+program runs in a nonbenign environment.)
+
+This unintuitive behavior could not occur if the two loads were not
+omitted.
+In fact, the definition of semantic dependency might remain
+perfectly acceptable if the requirement for global-analysis compilers
+were weakened, if the compiler treated atomic objects as quasi
+volatile and in addition was not allowed to omit accesses to them.
+This is a possible topic for future research.
+
+\subsection{Effect of Memory Layout}
+\label{sec:Effect of Memory Layout}
+
+Part of our demonstration of the Fundamental Property of semantic
+dependencies relies on the fact, stated in
+Section~\ref{sec:Abstract Executions},
+that an abstract execution of a thread is entirely determined by the
+values obtained for its loads.
+But when we compare abstract executions of the same thread in two
+different programs, this may no longer be entirely true owing to the
+effect of differing memory layouts.
+
+Consider this simple example:
+\begin{quote}
+\begin{verbatim}
+x = (int) &x;
+\end{verbatim}
+\end{quote}
+Even though the example contains no loads at all, it may store
+different values when running in different programs because the
+object \co{x} may be allocated at differing addresses in those
+programs.
+According to our definition, this could count as a degenerate OOTA
+cycle of length one, in which the store is semantically dependent on
+an empty set of loads!
+
+To rule out such pathological counterexamples we should require that
+in a witness to a semantic dependency, the addresses of all the
+objects and functions referred to in the thread $T$ are the same as
+in the original execution.
+This is a very technical restriction but there are occasions when
+the issue might realistically arise, such as when computing a hash
+value based on an object's address.
+
+\subsection{Merging Quasi-Volatile Loads}
+\label{sec:Merging Quasi-Volatile Loads}
+
+The compiler is permitted to merge quasi-volatile loads.
+This can lead to surprising results because a particular load
+may be merged with an earlier load in one execution and with a
+later load in another.
+This is demonstrated in the following, which is a variant of
+the example in
+Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}:
+\begin{quote}
+\begin{verbatim}
+int r1 = (x != 0);
+int r2 = (x != 0);
+int r3 = (x != 0);
+int r4 = (y != 0);
+z = (r2 == r4);
+\end{verbatim}
+\end{quote}
+Consider an abstract execution in which \co{r1}, \co{r2}, and \co{r4}
+are zero and \co{r3} is one (because another thread changed the value
+of \co{x} between two of the loads).
+We would expect that the store to \co{z} would be semantically
+dependent on the load of \co{y} in this execution.
+
+However, a single-thread analysis compiler can translate this into the
+machine-code equivalent of:
+\begin{quote}
+\begin{verbatim}
+int r1 = (x != 0);
+int r2;
+int r3 = (x != 0);
+int r4 = (y != 0);
+if (r1 != r3) {
+   r2 = r4;
+   z = 1;
+} else {
+   r2 = r1;
+   z = (r2 == r4);
+}
+\end{verbatim}
+\end{quote}
+In effect, the \co{r2} load is merged with the \co{r1} load in
+executions where \co{r1} is equal to \co{r3} or to \co{r4}, and it is
+merged with the \co{r3} load in other executions.
+
+To demonstrate the semantic dependency in the original code,
+a suitable witness would have
+to include a hardware realization of the abstract execution in which
+\co{r1} and \co{r2} are zero and \co{r3} and \co{r4} are one.
+But there are no hardware realizations of this execution with the
+machine code indicated above!
+Since \co{r1} is different from both \co{r3} and \co{r4}, the \co{r2}
+load will be merged with the \co{r3} load and so \co{r2} will
+necessarily be one, not zero.
+Thus, relative to this compiler the example does not contain a
+semantic dependency.
+
+Although it is unexpected, we cannot say that this conclusion is
+definitely wrong, because our intuitive notions of semantic
+dependency are not clear in cases where multiple loads of the same
+variable are present.
+% Can we instead say that transforming the code might require that the
+% witness also be transformed?
+
+Despite these outstanding issues, our definition of semantic dependency
+allows us to bring to bear the real-world constraints outlined in the
+next section.
+
 \section{Future Directions}
@@
 Finally, future work might expand on
-Sections~\ref{sec:Global Analysis and Volatile versus Quasi Volatile},
-and~\ref{sec:Merging Quasi-Volatile Loads}
+Section~\ref{sec:Issues and Refinements}
 so as to more precisely delineate the limits of permissible

### 2024-09-03 paper/ootavacuous.tex
asplos: Remove Section 7.2 ("Constraints of the Standard")
Instead, provide a brutal summary at the end of Section 4.6
("Volatile and Quasi Volatile Accesses").

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Section~\ref{sec:Constraints of the Standard}
-presents examples illustrating some of these requirements.
+Applying these requirements to specific examples results in prohibiting
+invention or duplication of atomic loads and stores.
+However, in the absence of ordering constraints, it permits omitting
+redundant non-volatile atomic stores and fusing of non-volatile atomic
+accesses to adjacent objects.
 
@@
 
-\subsection{Constraints of the Standard}
-\label{sec:Constraints of the Standard}
-
-\emph{Can we summarize this section with a list of permitted and prohibited
-optimizations in the discussion of quasi-volatile?}
-
-Volatile atomic accesses constitute observable
-behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2023N4950} and
-must be executed in strict accordance with the rules of the C++ abstract
-machine.
-This means that a volatile atomic store in the source code
-corresponds directly to a machine-code store instruction, which can
-commit only after all loads whose values are used to compute the
-store's address and value have committed.
-OOTA aficionados will recognize this as a special case of ordering
-relaxed loads before relaxed stores, albeit one not requiring
-expensive memory-fence instructions on weakly ordered architectures.
-
-Although C++ nonvolatile accesses to atomic objects are not observable
-behavior, any compiler that restricts its code analysis to a
-single thread must assume (unless it can prove otherwise) that a given
-relaxed atomic store can affect observable behavior in a different
-thread that loads the value stored.
-This assumption does not constrain the compiler to anywhere near
-the extent that a volatile relaxed atomic store would, but it does add
-significant constraints over those related to nonvolatile non-atomic
-stores.
-
-For example, nonvolatile relaxed atomic accesses are subject to the
-C++ memory model.
-The ``order-preserving'' bullet point from
-Section~\ref{sec:Volatile and Quasi Volatile Accesses}
-prohibits reordering of quasi-volatile accesses when they are
-to the same object,
-but such reordering is in any case directly forbidden by the memory model
-(the four coherence rules discussed in 6.9.2.2p14--19 \co{[intro.races]}).
-The ``onto'' bullet point also largely reiterates restrictions that follow
-from the standard, in particular,
-that compilers should not invent atomic stores,
-duplicate atomic stores, or invent atomic loads.
-On the other hand, it is permissible for compilers to omit redundant
-atomic stores or fuse nonvolatile atomic accesses of adjacent objects
-under appropriate circumstances;
-these topics correspond to the ``total'' and ``one-to-one'' bullet points of
-Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
-
-\paragraph{Invented Atomic Stores}
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/invented-store.litmus @@
-\caption{Example Invented Store}
-\label{lst:Example Invented Store}
-\end{listing}
-The reason that atomic stores should not be invented is that doing so can
-introduce new (and almost certainly undesirable) behaviors that are
-forbidden by the abstract machine.
-To see this, consider Listing~\ref{lst:Example Invented Store},
-a \co{herd7} litmus test that demonstrates such a behavior.
-Without P0's line~1, only the values~0 and~3 can be loaded into \co{r1} by
-P1's line~1.
-With P0's line~1, the additional value~42 can also be loaded into \co{r1}.
-The compiler is therefore forbidden from inventing the store of 42
-unless it can prove that doing so does not negatively affect the
-program's observable behaviors.
-For example, the compiler might be able to prove that there
-are no other accesses to \co{x} at the time of the invented
-store---but it's very hard to imagine how a compiler that analyzes
-only a single thread at a time could prove such a thing.
-
-\paragraph{Duplicated Atomic Stores}
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/duplicated-store.litmus @@
-\caption{Example Duplicated Store}
-\label{lst:Example Duplicated Store}
-\end{listing}
-Duplicating atomic stores is a special case of inventing them, but it is
-worth illustrating the fact that duplicating stores can also introduce
-new and undesirable behaviors.
-To see this, consider Listing~\ref{lst:Example Duplicated Store}, a
-litmus test that demonstrates such a behavior.
-Without P0's line~1, if the value loaded into \co{r1} is one then the final value
-of \co{x} must be two.
-With that line, the final value of \co{x} can be one even when the
-value loaded into \co{r1} is one.
-The compiler is therefore forbidden from duplicating atomic stores
-unless it can prove that doing so does not negatively affect the
-program's observable behaviors.
-Again, the compiler might be able to prove that there
-are no other accesses to \co{x} at the time of the duplicated store.
-
-In contrast, compilers should not duplicate, repurpose, or otherwise
-invent atomic loads, as discussed in
-Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.
-
-\paragraph{Omitted Redundant Atomic Stores}
-In contrast, a pair of back-to-back nonvolatile atomic stores to the
-same object
-always might be executed such that no other thread accesses the object
-during the time between those two stores.
-This means that if the compiler omitted the first store,
-the user would have no way to prove this fact short of
-inspecting the assembly code.
-In cases where such omissions are undesirable, the user can resort to
-volatile atomic stores or to inline assembly\footnote{
-	For example, by placing the Linux-kernel \co{barrier()} macro
-	between the two stores.
-	This macro is an empty GCC \co{asm} that specifies the \co{memory}
-	clobber.}
-to prevent them.
-
-Omitting a redundant store cannot create an OOTA cycle,
-because any valid execution (OOTA or not)
-of the program with the first of a back-to-back pair of nonvolatile
-relaxed atomic stores omitted is a valid execution of the
-program with the store present.
-% This is awkward because a reader might reasonably extend the "Acting
-% as if" all the way to the comma.  The best I could do was "It is not
-% possible to create an OOTA cycle by acting as if a redundant store
-% was omitted from the source code", which is admittedly not all that
-% great.
-
-\paragraph{Fused Accesses of Adjacent Nonvolatile Atomic Objects}
-Suppose a pair of adjacent nonvolatile atomic objects, when combined,
-form a single machine-word aligned and machine-word sized unit.
-Suppose further that the source code contains a pair of atomic
-loads, one from each of these objects, with the last of the pair
-being a relaxed load.
-Then the compiler could generate instead
-a single atomic load of the combined pair,
-with the memory-ordering semantics of the first load.
-A similar fact applies to pairs of stores.
-
-This holds even when the atomic objects are treated as quasi volatile,
-providing an example of how the ``one-to-one'' bullet point of
-Section~\ref{sec:Volatile and Quasi Volatile Accesses} does not apply
-to them.
-
 \subsection{Semantic Dependencies and Tooling}

### 2024-09-03 paper/ootavacuous.tex
asplos: Remove Section 7 ("Real-World Constraints")
Summarize at point of references, as needed.  Move "Semantic Dependencies
and Tooling" to the "Issues and Refinements" section for further
summarization.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 ordering in strongly ordered systems and by data-dependency ordering in
-weakly ordered systems.\footnote{
-	The need to prohibit simple OOTA is one reason why compiler-based
-	value speculation optimizations require checks on such
-	speculation, and these checks must be based on actual values
-	loaded.}
+weakly ordered systems.
+The need to prohibit simple OOTA is one reason why compiler-based value
+speculation optimizations require checks on such speculation, and further
+requires that these checks must be based on the values actually loaded.
 
@@
 	make their stores visible to other CPUs until those stores
-	are no longer speculative.
-	See Section~\ref{sec:Hardware Architecture and Design}
-	for more discussion about hardware speculation.}
+	are no longer speculative.}
 the stores on line~2 cannot take place until the P0's and P1's CPUs
@@
 
-\section{Real-World Constraints}
-\label{sec:Real-World Constraints}
-
-The C++ standard imposes many constraints, and these have been considered
-in prior work.
-In real-world C++ implementations, however, the standard's abstract machine
-must be mapped onto real hardware, and this imposes additional
-constraints stemming from hardware architecture and design.
-This mapping and these constraints have been only partially accounted
-for in the standard and other work.
-Part of the reason for this is
-that accurate and executable formal descriptions of hardware memory
-models did not appear until after the standard was
-released~\cite{JadeAlglave2011ppcmem,Maranget2012TutorialARMPower}.
-But they do exist now.
-
-The following sections discuss these constraints, starting with
-hardware constraints and continuing with additional constraints imposed
-by the standard.
-An additional section discusses the OOTA-cycle implications for C++
-tooling.
-
-\subsection{Hardware Architecture and Design}
-\label{sec:Hardware Architecture and Design}
-
-\emph{Can we summarize this section, especially given that the ASPLOS
-audience can be expected to have a greater understanding of these issues
-than the C++ standards committee?}
-
-Computer systems are expected to continue increasing their use of
-speculative execution.
-Nevertheless, the main points of this paper will remain unaffected.
-To see why, keep in mind that on real systems observable effects
-must eventually be committed, but no effect can be committed
-while any of the computations that determine the effect remain
-speculative.
-Combining this with the obvious fact that a machine instruction cannot
-commit until it is no longer speculative, we can draw two conclusions:
-\begin{itemize}
-\item	A load cannot commit until the store it reads from has
-	committed.\footnote{
-	An early ARMv8 memory model did allow loads that read from an
-	earlier store in the same thread to commit before the store.
-	This apparent exception merely reflects a difference in
-	nomenclature; in the memory model a store was said to commit at
-	the time when it made its new value available to other
-	threads, whereas we say that a store commits at the time when
-	the CPU has irrevocably decided that it will take place with a
-	particular fully determined value and address.}
-\item	A store cannot commit until all the loads on which it has a
-	hardware dependency have committed.
-\end{itemize}
-
-As an example showing what can go wrong when these principles are
-violated, consider Listing~\ref{lst:Speculated Store and Non-Speculated Load}.
-Suppose that the storage for \co{x} is located immediately before that
-for \co{y[]}, and
-suppose further that hardware speculation incorrectly guesses the value
-of \co{r1} will be $-1$, so that the speculated store of 42 on line~11 uses
-the address of \co{x}.\footnote{
-	Yes, \co{y[-1]} is UB, but the CPU neither knows
-	nor cares, nor should it.}
-Then if line~13 reads from the speculative store and commits, it will load
-the value 42 into \co{r2}---an observable effect since \co{r2}
-is mentioned in the \co{exists} clause on line~16---although
-by the rules of the C++ abstract machine the value must instead be zero.
-
-\begin{listing}[tbp]
-@@ DisplayLitmus litmus/speculative-store.litmus @@
-\caption{Speculated Store and Non-Speculated Load}
-\label{lst:Speculated Store and Non-Speculated Load}
-\end{listing}
-
-Applied to Listing~\ref{lst:OOTA Cycle} on
-page~\pageref{lst:OOTA Cycle}, for another example,
-the principles dictate that if the load on line~8 reads from the store
-on line~18 and the load on line~16 reads from the store on line~10
-(as they do in the execution described by the listing's \co{exists} clause),
-then each of the stores must commit before its corresponding load.
-A further application says that
-neither line~10 nor line~12 can commit until after the load
-in line~8 (and also the conditional on line~9) has committed.
-Similarly, neither line~18 nor line~20 can commit until after line~16
-(and~17) has committed.
-Taken together, these constraints imply that the OOTA cycle in the
-listing cannot be realized, because to do so would require that
-none of the machine instructions corresponding to lines~8, 10, 16,
-and~18 could commit before the others.
-Furthermore, it's not possible to circumvent this reasoning by suggesting
-that some of those lines could commit at the same time,
-because instructions take time to execute even when executing
-speculatively.
-
-Yes, this does mean that hardware can produce OOTA cycles during
-speculative execution, but the hardware is required to prevent such
-speculative cycles from committing.
-As a special case of this requirement, if a load obtains
-a value from a speculated store that has not yet committed,
-and the store gets squelched,
-then the load must be restarted before it commits even if the load and
-the store are executed by different CPUs.
-
-The requirement applies within a single
-multithreaded core as well as between cores.
-The fact that intracore communication is faster than intercore
-communication does not magically give loads the ability to commit
-before the stores they read from, no matter what cores they execute on.
-
-\begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-causality-2.litmus @@
-\caption{Causality Test Case 2}
-\label{lst:Causality Test Case 2}
-\end{listing}
-
-Another feature we can expect of upcoming computer systems is the
-addition of new hardware instructions.
-For example, one could imagine a conditional-store instruction,
-similar to existing conditional-move instructions but carrying out
-a store to memory rather than a move to a register.
-The condition would control whether or not the store takes place.
-A compiler doing single-thread analysis could use a conditional-store
-instruction to generate code corresponding to lines~10 and~11 of
-Listing~\ref{lst:Causality Test Case 2}
-on page~\pageref{lst:Causality Test Case 2}
-without use of a conditional branch, potentially
-enabling more hardware optimizations than would otherwise be permitted.
-Nevertheless, the conditional store could not be committed until
-the instructions it depends on (lines~8 and~9) have committed.
-Any new instruction will be subject to this constraint,
-just as the instructions in current systems are.
-
-A key conclusion to draw from this discussion is that
-regardless of how advanced a computer system may be,
-it cannot commit a store having a hardware dependency on some
-set of loads until after (in global time) those loads have committed.
-
-\subsection{Semantic Dependencies and Tooling}
-\label{sec:Semantic Dependencies and Tooling}
-
-\emph{Do we really need this?}
-
-Where a C++ implementation's primary function is to correctly execute
-a C++ program, the function of tooling is often to evaluate properties
-of possible executions of that same program.
-While implementations should avoid breaking semantic dependencies,
-tools are in some cases permitted to misclassify them, as shown in
-Table~\ref{tab:Dependency Classification For Tools}.
-
-\begin{table}
-\centering
-\begin{tabular}{c|c|l}
-Actual Execution	& Tooling		& Result \\
-\hline
-sdep			& sdep			& Ordered \\
-\cline{2-3}
-			& $\neg$sdep		& False Positive (Warning?) \\
-\hline
-$\neg$sdep		& sdep			& False Negative (Warning?) \\
-\cline{2-3}
-			& $\neg$sdep		& Unordered \\
-\end{tabular}
-\caption{Dependency Classification For Tools}
-\label{tab:Dependency Classification For Tools}
-\end{table}
-
-Any misclassification will result in either a false positive or
-a false negative, except that many tools give some indication of a
-misclassification, for example, printing a warning message indicating
-that the code is too complex for it to analyze.
-Some projects would choose to restructure the code in such cases, on the
-grounds that code that is difficult for an automated tool to understand
-is also likely to be misunderstood by its all-too-human developers.
-
-One advantage that tools often have over C++ implementations is the
-ability to devote much more computational power to the problem at hand.
-In fact, some tools respond to excessive complexity by consuming all
-available CPU and memory, which can also be interpreted as a good and
-sufficient warning message.
-
-However, there is an important special case for tools that are closely
-associated with a C++ compiler.
-Such tools can simply use that compiler's classification of
-dependencies as semantic on the one hand or nonsemantic on the
-other, whether by working with the compiler's intermediate
-representations, examining binaries produced by the compiler, tracing
-the program's execution, or, in the case of dynamic tools, actually
-executing the program.
-Either way, this approach reduces the problem of analysis from the
-level of the C++ abstract machine to a much simpler lower level of
-abstraction.
-While taking this approach sacrifices significant generality, it also
-has the significant benefit of greatly reducing the cost of dependency
-analysis, potentially all the way down to zero.
-
 \section{Issues and Refinements}
@@
 
-Despite these outstanding issues, our definition of semantic dependency
-allows us to bring to bear the real-world constraints outlined in the
-next section.
+\subsection{Semantic Dependencies and Tooling}
+\label{sec:Semantic Dependencies and Tooling}
+
+\emph{Do we really need this?}
+
+Where a C++ implementation's primary function is to correctly execute
+a C++ program, the function of tooling is often to evaluate properties
+of possible executions of that same program.
+While implementations should avoid breaking semantic dependencies,
+tools are in some cases permitted to misclassify them, as shown in
+Table~\ref{tab:Dependency Classification For Tools}.
+
+\begin{table}
+\centering
+\begin{tabular}{c|c|l}
+Actual Execution	& Tooling		& Result \\
+\hline
+sdep			& sdep			& Ordered \\
+\cline{2-3}
+			& $\neg$sdep		& False Positive (Warning?) \\
+\hline
+$\neg$sdep		& sdep			& False Negative (Warning?) \\
+\cline{2-3}
+			& $\neg$sdep		& Unordered \\
+\end{tabular}
+\caption{Dependency Classification For Tools}
+\label{tab:Dependency Classification For Tools}
+\end{table}
+
+Any misclassification will result in either a false positive or
+a false negative, except that many tools give some indication of a
+misclassification, for example, printing a warning message indicating
+that the code is too complex for it to analyze.
+Some projects would choose to restructure the code in such cases, on the
+grounds that code that is difficult for an automated tool to understand
+is also likely to be misunderstood by its all-too-human developers.
+
+One advantage that tools often have over C++ implementations is the
+ability to devote much more computational power to the problem at hand.
+In fact, some tools respond to excessive complexity by consuming all
+available CPU and memory, which can also be interpreted as a good and
+sufficient warning message.
+
+However, there is an important special case for tools that are closely
+associated with a C++ compiler.
+Such tools can simply use that compiler's classification of
+dependencies as semantic on the one hand or nonsemantic on the
+other, whether by working with the compiler's intermediate
+representations, examining binaries produced by the compiler, tracing
+the program's execution, or, in the case of dynamic tools, actually
+executing the program.
+Either way, this approach reduces the problem of analysis from the
+level of the C++ abstract machine to a much simpler lower level of
+abstraction.
+While taking this approach sacrifices significant generality, it also
+has the significant benefit of greatly reducing the cost of dependency
+analysis, potentially all the way down to zero.
 

### 2024-09-03 paper/ootavacuous.tex
asplos: Cite Jay Richards's and Mark Batty's tool
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 2024, Mark Batty is persisting with modular relaxed dependencies, which
-may have the added ability of checking compiler optimizations.
-@@@ cite web site.
+may have the added ability of checking compiler
+optimizations~\cite{JayRichards2024SymbolicMrder}.
 
@@
 One advantage that tools often have over C++ implementations is the
-ability to devote much more computational power to the problem at hand.
+ability to devote much more computational power to the problem at
+hand~\cite{JayRichards2024SymbolicMrder}.
 In fact, some tools respond to excessive complexity by consuming all

### 2024-09-10 paper/ootavacuous.tex
Remove last part of section 7 (Issues and Refinements)
The last subsection of section 7 (Semantic Dependencies and Tooling) is
not appropriate for this version of our paper.  Remove it.

Also, fix a remaining dangling reference to thread1 and thread2 of
Listing 1.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 not count as OOTA according to our definition because the
-dependencies in \co{thread1} and \co{thread2} would not be semantic.
+dependencies in P0 and P1 would not be semantic.
 
@@
 % witness also be transformed?
-
-\subsection{Semantic Dependencies and Tooling}
-\label{sec:Semantic Dependencies and Tooling}
-
-\emph{Do we really need this?}
-
-Where a C++ implementation's primary function is to correctly execute
-a C++ program, the function of tooling is often to evaluate properties
-of possible executions of that same program.
-While implementations should avoid breaking semantic dependencies,
-tools are in some cases permitted to misclassify them, as shown in
-Table~\ref{tab:Dependency Classification For Tools}.
-
-\begin{table}
-\centering
-\begin{tabular}{c|c|l}
-Actual Execution	& Tooling		& Result \\
-\hline
-sdep			& sdep			& Ordered \\
-\cline{2-3}
-			& $\neg$sdep		& False Positive (Warning?) \\
-\hline
-$\neg$sdep		& sdep			& False Negative (Warning?) \\
-\cline{2-3}
-			& $\neg$sdep		& Unordered \\
-\end{tabular}
-\caption{Dependency Classification For Tools}
-\label{tab:Dependency Classification For Tools}
-\end{table}
-
-Any misclassification will result in either a false positive or
-a false negative, except that many tools give some indication of a
-misclassification, for example, printing a warning message indicating
-that the code is too complex for it to analyze.
-Some projects would choose to restructure the code in such cases, on the
-grounds that code that is difficult for an automated tool to understand
-is also likely to be misunderstood by its all-too-human developers.
-
-One advantage that tools often have over C++ implementations is the
-ability to devote much more computational power to the problem at
-hand~\cite{JayRichards2024SymbolicMrder}.
-In fact, some tools respond to excessive complexity by consuming all
-available CPU and memory, which can also be interpreted as a good and
-sufficient warning message.
-
-However, there is an important special case for tools that are closely
-associated with a C++ compiler.
-Such tools can simply use that compiler's classification of
-dependencies as semantic on the one hand or nonsemantic on the
-other, whether by working with the compiler's intermediate
-representations, examining binaries produced by the compiler, tracing
-the program's execution, or, in the case of dynamic tools, actually
-executing the program.
-Either way, this approach reduces the problem of analysis from the
-level of the C++ abstract machine to a much simpler lower level of
-abstraction.
-While taking this approach sacrifices significant generality, it also
-has the significant benefit of greatly reducing the cost of dependency
-analysis, potentially all the way down to zero.
 

### 2024-09-10 paper/ootavacuous.tex
Shrink section 8 (Future Directions)
Section 8 can be made considerably shorter by replacing verbose text
with a simple itemized list.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-This paper focuses solely on compilers, but it should be possible to
-extend these results to some classes of interpreters and JITs on the
-one hand and to link-time optimizations (LTO) on the other.
-
-This paper focuses primarily on compilers that do only local per-thread
-analysis.
-Further explorations could consider additional classes of global analysis,
-for example, analyses that demonstrate that a given expression will
-always evaluate to a particular constant.
-JMM Causality Test Case 1 % %%% Cite P3064.
-% %%% discussed in
-% %%% Appendix~\ref{app:Causality Test Case 1}
-provides an example of such a global analysis.
-
-This paper assumes that the user defines threads, and that the
-C++ implementation executes those threads unchanged.
-Future analyses might examine the possibility of ``flattening''
-optimizations that combine multiple threads into one.
-(Note that such optimizations are more difficult than they might first appear.)
-
-This paper shows that semantic dependencies are relative to specific
-executions of a program and the specific compiler in use rather than
-being determined solely by the source code.
-Future work might expand on
-Section~\ref{sec:Relative versus Absolute Dependency},
-exploring special cases in which absolute semantic dependencies are
-functions just of the source code.
-
-This paper considers only programs whose threads
-communicate using shared memory.
-Future work might include the effects of input and output or
-other operations which can vary from one execution to another,
-such as those based on the current time, process IDs, or pointer values
-(see Section~\ref{sec:Effect of Memory Layout}).
-
-Finally, future work might expand on
-Section~\ref{sec:Issues and Refinements}
-so as to more precisely delineate the limits of permissible
-behavior for quasi-volatile object accesses.
+This paper focused solely on compilers, primarily those that do only
+local per-thread anaylsis and optimization.
+It introduced the notion of quasi-volatile object accesses and gave
+a definition for semantic dependency that is relative to a particular
+C++ implementation.
+
+Further explorations could:
+\begin{itemize}
+\item	Extend these results to some classes of interpreters and JITs
+	on one hand, and to link-time optimizations (LTO) on the other.
+\item	Consider various classes of global analysis, such as
+	demonstrating that a given expression will always evaluate
+	to a constant.
+\item	Examine the possibility of ``flattening'' optimizations that
+	combine multiple threads into one.
+\item	Explore cases in which semantic dependencies must be absolute,
+	not relative to any particular compiler.
+\item	Include the effects of input and output or other operations
+	which can vary from one execution to another.
+\item	Delineate more precisely the limits of permissible behavior
+	for quasi-volatile object accesses.
+\end{itemize}
 

### 2024-09-10 paper/ootavacuous.tex
Shrink section 9 (Conclusions) and combine with section 8
Section 9 includes a lengthy restatement of the contents of the paper.
This is not necessary since the reader has just encountered all of the
material.  Replace this restatement with a two-sentence summary and a
one-sentence conclusion, merged into section 8.

A final concluding sentence may need to be added.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\section{Future Directions}
-\label{sec:Future Directions}
+\section{Conclusions and Future Directions}
+\label{sec:Conclusions and Future Directions}
 
@@
 C++ implementation.
-
+It formulated a Fundamental Property of semantic dependency and proved
+that implementations subject to some very mild restrictions on how
+they treat atomic object accesses do have this property.
+Finally, it demonstrated that satisfying the Fundamental Property
+guarantees an implementation cannot give rise to OOTA executions.
+
+The conclusion to be drawn from this work is that in this context,
+avoiding OOTA cycles requires no changes to the standard,
+to current implementions, or to user practices for portable code.
 Further explorations could:
@@
 
-\section{Conclusion}
-\label{sec:Conclusion}
-
-This paper has presented samples of code containing non-trivial
-semantic dependencies, uncovering some shortcomings in typical definitions of
-``semantic dependency''.
-It pointed out examples where semantic dependencies
-are a function of an execution rather than of the source code, and examples
-where a single semantic dependency extends from multiple loads
-to a store but not from any one of those loads.
-These complex dependencies motivated a generalization of the definition
-of ``OOTA cycle'', which is shown in
-Section~\ref{sec:Semantic Dependencies Can Be Many-To-One}.
-
-This generalization, combined with a focus on compiler-based loose C++
-implementations using single-thread analysis, and a consideration of
-the relation between source code and machine code, led to the formulation
-of the Fundamental Property of semantic dependency shown in
-Section~\ref{sec:The Fundamental Property of Semantic Dependencies}.
-This in turn led to a precise definition of ``semantic dependency'' given in
-Section~\ref{sec:For Compilers Using Single-Thread Analysis}.
-This definition was used in
-Section~\ref{sec:Verifying the Fundamental Property}
-to demonstrate that these implementations are
-incapable of producing OOTA cycles in UB-free programs,
-provided they treat all nonvolatile atomic objects as
-quasi volatile, obeying the limitations outlined in
-Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
-
-A variant definition of ``semantic dependency'' given in
-Section~\ref{sec:For Compilers Using Global Analysis}, appropriate for
-compiler-based loose C++ implementations that don't restrict
-their code transformations to those based on single-thread analysis,
-was used to show that such implementations
-are incapable of producing OOTA cycles in UB-free programs
-provided they treat all atomic objects as volatile.
-
-This work shows that avoiding OOTA cycles requires no changes to the
-standard, to current implementions, or to user practices for portable
-code.
-Future implementations must avoid optimizations that invent, duplicate,
-or repurpose atomic loads.
-More work is required to delineate any further restrictions on future
-implementations that might carry out optimizations based on global
-inter-thread analysis.
-
 % \clearpage

### 2024-09-10 paper/ootavacuous.tex
Add ghost citation
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-09-10 paper/ootavacuous.tex
@@
 semantic dependency.
+Additional information is available in a companion technical report,
+which is not cited in order to maintain anonymity.
+% P3064R2 ("How to Avoid OOTA Without Really Trying")
+% https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2024/p3064r2.pdf
 
@@
 not count as OOTA according to our definition because the
-dependencies in P0 and P1 would not be semantic.
+dependencies in \co{thread1} and \co{thread2} would not be semantic.
 
@@
 
+\subsection{Semantic Dependencies and Tooling}
+\label{sec:Semantic Dependencies and Tooling}
+
+\emph{Do we really need this?}
+
+Where a C++ implementation's primary function is to correctly execute
+a C++ program, the function of tooling is often to evaluate properties
+of possible executions of that same program.
+While implementations should avoid breaking semantic dependencies,
+tools are in some cases permitted to misclassify them, as shown in
+Table~\ref{tab:Dependency Classification For Tools}.
+
+\begin{table}
+\centering
+\begin{tabular}{c|c|l}
+Actual Execution	& Tooling		& Result \\
+\hline
+sdep			& sdep			& Ordered \\
+\cline{2-3}
+			& $\neg$sdep		& False Positive (Warning?) \\
+\hline
+$\neg$sdep		& sdep			& False Negative (Warning?) \\
+\cline{2-3}
+			& $\neg$sdep		& Unordered \\
+\end{tabular}
+\caption{Dependency Classification For Tools}
+\label{tab:Dependency Classification For Tools}
+\end{table}
+
+Any misclassification will result in either a false positive or
+a false negative, except that many tools give some indication of a
+misclassification, for example, printing a warning message indicating
+that the code is too complex for it to analyze.
+Some projects would choose to restructure the code in such cases, on the
+grounds that code that is difficult for an automated tool to understand
+is also likely to be misunderstood by its all-too-human developers.
+
+One advantage that tools often have over C++ implementations is the
+ability to devote much more computational power to the problem at
+hand~\cite{JayRichards2024SymbolicMrder}.
+In fact, some tools respond to excessive complexity by consuming all
+available CPU and memory, which can also be interpreted as a good and
+sufficient warning message.
+
+However, there is an important special case for tools that are closely
+associated with a C++ compiler.
+Such tools can simply use that compiler's classification of
+dependencies as semantic on the one hand or nonsemantic on the
+other, whether by working with the compiler's intermediate
+representations, examining binaries produced by the compiler, tracing
+the program's execution, or, in the case of dynamic tools, actually
+executing the program.
+Either way, this approach reduces the problem of analysis from the
+level of the C++ abstract machine to a much simpler lower level of
+abstraction.
+While taking this approach sacrifices significant generality, it also
+has the significant benefit of greatly reducing the cost of dependency
+analysis, potentially all the way down to zero.
+
 \section{Future Directions}

### 2024-09-10 paper/ootavacuous.tex
ootavacuous: Consolidate abstract and introduction
No point in saying the same thing twice, and especially not in the same
column of the same page.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-09-10 paper/ootavacuous.tex
@@
 \begin{abstract}
-	The out-of-thin-air (OOTA) properties
-	of \co{memory_order_relaxed} in the C++ memory model
-	have caused considerable consternation over the years.
-	Attempts to create memory models that rule out OOTA behaviors
-	have been either non-executable, complex, or unloved by C++
+	Attempts to create memory models that forbid out-of-thin-air
+	(OOTA) behaviors of C++ \co{memory_order_relaxed} accesses
+	have been either non-executable, complex, or unloved by
 	implementers.
-	But at the same time, we know of no instances of OOTA behavior
-	in real C++ implementations.
-
-	We focus on shared-memory programs and
-	C++ implementations based on traditional compilers and
-	computing hardware, including CPUs and GPGPUs.
-	This context permits us to consider the detailed relations
-	between source code and machine code required by the restricted
-	nature of volatile atomic accesses.
-	The OOTA problem and the related challenge of coming to grips
-	with semantic dependency are much more tractable at the hardware
-	level than at the source level, thanks to existing formal
-	hardware models.
-	We show that these models' constraints prevent OOTA cycles
-	from occurring in undefined-behavior-free C++ programs running
-	on compiler-based implementations, provided the cycles involve
-	only volatile atomics.
+	However, we know of no instances of OOTA behavior in real C++
+	implementations.
+
+	We focus on shared-memory programs and C++ implementations based
+	on traditional compilers and hardware, including CPUs and GPGPUs,
+	thus permitting easier analysis of semantic dependencies and
+	OOTA behaviors, thanks to existing formal hardware models.
+	We show that these models' constraints prevent OOTA cycles from
+	occurring in undefined-behavior-free C++ programs running on
+	compiler-based implementations, provided the cycles involve only
+	volatile atomics.
 	We also extend this work to nonvolatile atomics by defining
-	``quasi-volatile'' behavior that we expect C++ implementations
-	will adhere to if they confine themselves to single-thread analysis.
+	``quasi-volatile'' behavior that we C++ implementations
+	must adhere to in order to evaluate arithmetic expresssions
+	involving atomic accesses in accordance with the abstract
+	machine.
+
+	It follows that such implementations need take no special action
+	to avoid OOTA, and that no changes are required to user code or
+	to the standard.
 \end{abstract}
@@
 \label{sec:Introduction}
-
-This paper shows that compiler-based C++ implementations subject to
-reasonable constraints on how they treat accesses to atomic objects
-cannot exhibit out-of-thin-air (OOTA) cycles.
-It follows that these implementations need to take almost no special
-actions to avoid OOTA.
-In fact, for many compilers the constraints boil down to a simple ``Don't
-invent or duplicate atomic accesses'', which the compiler should not
-do anyway if it is to evaluate arithmetic expressions involving atomic
-accesses in accordance with the abstract machine.
-Absolutely no changes are required to either the standard or to user code.
 

### 2024-09-11 paper/ootavacuous.tex
Further wordsmithing of abstract
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	implementers.
-	However, we know of no instances of OOTA behavior in real C++
+	At the same time, we know of no instances of OOTA behavior in real C++
 	implementations.
@@
 	on traditional compilers and hardware, including CPUs and GPGPUs,
-	thus permitting easier analysis of semantic dependencies and
-	OOTA behaviors, thanks to existing formal hardware models.
+	thus easing analysis of semantic dependencies and OOTA behaviors
+	via existing hardware models.
 	We show that these models' constraints prevent OOTA cycles from
 	occurring in undefined-behavior-free C++ programs running on
-	compiler-based implementations, provided the cycles involve only
-	volatile atomics.
+	compiler-based implementations that use single-thread analysis,
+	provided the cycles involve only volatile atomics.
 	We also extend this work to nonvolatile atomics by defining
-	``quasi-volatile'' behavior that we C++ implementations
-	must adhere to in order to evaluate arithmetic expresssions
-	involving atomic accesses in accordance with the abstract
-	machine.
-
-	It follows that such implementations need take no special action
-	to avoid OOTA, and that no changes are required to user code or
+	``quasi-volatile'' behavior that C++ implementations must adhere
+	to in order to evaluate arithmetic expressions involving atomic
+	accesses in accordance with the standard.
+
+	It follows that enforcing OOTA avoidance requires no special
+	action from implementations and no changes to user code or
 	to the standard.

### 2024-09-12 paper/litmus2table.sh
litmus2table.sh: Convert tabs to non-breaking spaces
Ineffectually, but one step at a time...

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -6,6 +6,7 @@
 
 sed -e 's/&/\\&/' | sed -e 's/	}/	\\}/' | sed -e 's/{/\\{/' |
 sed -e 's,\\/,$\\vee$,g' | sed -e 's,/\\,$\\wedge$,g' |
+sed -e 's/	/\~/g' |
 awk -v amp='&' -v bs='\\' -v dq='"' '
 BEGIN {
 	ininit = 0;

### 2024-09-12 paper/ootavacuous.tex
ootavacuous: Compress Section 1.1 ("Brief OOTA Overview")
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 We begin with a brief overview of the OOTA problem followed by an equally
-brief summary of prior work in this area.
+brief summary of prior OOTA work.
 
@@
 
-In broad terms,
-OOTA occurs theoretically when a group of threads load from each others' stores
-and each thread's store depends on the value returned by that thread's load.
-The collection of loads and stores forms an \emph{OOTA cycle}.
-In the most extreme cases a nonsensical value can pop up ``out of thin air'';
-however, as shown by
-Listing~\ref{lst:OOTA Cycle} below,
-OOTA cycles need not involve nonsensical values.
-
-\subsubsection{Simple OOTA Cycle}
-\label{sec:Simple OOTA Cycle}
-
-\begin{listing}[tbp]
+\begin{listing}[bp]
 % Constructed by hand.
@@
 
-Listing~\ref{lst:Simple OOTA} shows a simple example where an OOTA
-cycle might result in all of \co{x}, \co{y}, \co{r1}, and \co{r2}
-having final values of 42, despite the fact that there is nothing in
-the default-zero initial values or the executable code to support such
-an outcome~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
-Throughout this paper, variables whose names are ``\co{r}'' followed by
-a number denote registers, and other names denote shared variables.
-All accesses to shared variable are relaxed atomic accesses.
+In broad terms, OOTA occurs when a group of threads load from each others'
+stores and each thread's store depends on the value returned by that
+thread's load.
+The collection of loads and stores forms an \emph{OOTA cycle}.
+In the most extreme cases a nonsensical value can pop up ``out of thin air''
+as shown in
+Listing~\ref{lst:Simple OOTA}.
+Here, all of \co{x}, \co{y}, \co{r1}, and \co{r2} might have final
+values of 42, despite their being no instances of 42 in the default-zero
+initialization or in the executable
+code~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.\footnote{
+	Throughout this paper, variables whose names are ``\co{r}''
+	followed by a number denote registers, and other names denote
+	shared variables.
+	All accesses to shared variable are relaxed atomic accesses.}
 First, P0's line~1 does a relaxed atomic load from \co{x} into \co{r1},
@@
 Third, P1's line~1 loads 42 from \co{y} to \co{r2}.
-Fourth and finally, P1's line~2 stores 42 to \co{x}, justifying the value
-loaded by P0's line~1.
+Finally, P1's line~2 stores 42 to \co{x}, justifying the value loaded
+by P0's line~1.
 
@@
 standard explicitly prohibits them in 33.5.4p8
-(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
+(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950},
+which prevents misapplication of the as-if rule in oracular C++
+implementations:
 \begin{quote}
@@
 \end{quote}
-The standard's prohibition of OOTA is of course important, for example, to
-prevent misapplication of the as-if rule in oracular C++ implementations,
-but those of us writing code in the real world must rely on actual C++
-implementations.
-And in these implementations, this prohibition is in fact enforced by TSO
-ordering in strongly ordered systems and by data-dependency ordering in
-weakly ordered systems.
-The need to prohibit simple OOTA is one reason why compiler-based value
-speculation optimizations require checks on such speculation, and further
-requires that these checks must be based on the values actually loaded.
+This paper uses \emph{full C++} to denote the standard including this
+prohibition of OOTA and \emph{loose C++} to denote a hypothetical standard
+that excludes this prohibition but is otherwise identical to full C++.
+Unqualified \emph{C++} means full C++.
 
@@
 	make their stores visible to other CPUs until those stores
-	are no longer speculative.}
+	are no longer speculative.
+	This same restriction applies to compiler-based speculation.}
 the stores on line~2 cannot take place until the P0's and P1's CPUs
@@
 
-\subsubsection{Simple Reordering}
-\label{sec:Simple Reordering}
-
-\begin{listing}[tbp]
+\begin{listing}[bp]
 % Constructed by hand.
@@
 
-It is important to distinguish true OOTA cycles from OOTA-like
-behavior caused by simple reordering, such as that shown in
+Simple reordering can cause non-OOTA, but OOTA-like, behavior, as
+exemplified by
 Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
-Both the C++ compiler and the CPU are within their rights to reorder P1's
-lines~1 and~2, which can result in all of \co{x}, \co{y}, \co{r1},
-and \co{r2} having the value 42 via the following steps.
-First, P1's line~2 stores 42 to \co{x}.
-Second, P0's line~1 loads 42 from \co{x} into \co{r1}.
-Third, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
-Fourth and finally, P1's line~1 loads 42 from \co{y} to \co{r2}.
-Current C++ implementations can and do exhibit this reordering behavior,
+Both the compiler and the CPU might reorder P1's lines~1 and~2, so that
+all of \co{x}, \co{y}, \co{r1}, and \co{r2} might have the value 42,
 as indicated by the ``sometimes satisfied'' in the figure.
-
-\subsubsection{Non-Trivial OOTA}
-\label{sec:Non-Trivial OOTA}
-
-This paper follows P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
-in using \emph{full C++} to denote the standard including the
-prohibition of OOTA mentioned above.
-Unlike that article, we use \emph{loose C++}
-(rather than \emph{strict C++}, which seems too similar to
-\emph{full C++} for comfort)
-to denote a hypothetical standard that excludes this prohibition but
-is otherwise identical to full C++.
-Unqualified \emph{C++} means full C++.
 
@@
 
-Listing~\ref{lst:OOTA Cycle}
-shows a code fragment that under loose C++ has an OOTA cycle, though
-the cycle is of course prohibited in full C++.
-This example differs from that of Section~\ref{sec:Simple OOTA Cycle}
-in that line~3 of both processes stores the value 42, but only if
-line~2 of a given process determines that the value loaded was 42,
-keeping in mind that variables are default-initialized to zero.
-
-This OOTA cycle is constructed in the following manner.
-First, P0's and P1's line~1 load from \co{y} and \co{x}, respectively,
-both somehow returning the value 42.
-Second, P0's and P1's line~2 determine that the values loaded were both
-42, so each process advances to its line~3.
-Third, P0's and P1's line~3 stores the value 42 that was (somehow)
-clairvoyantly loaded by the other process's line~1.
-
-The next section looks at how prior work has refined these issues.
+On the other hand, OOTA cycles can arise even when value 42 is mentioned,
+as shown in
+Listing~\ref{lst:OOTA Cycle}.
+Line~3 of both processes stores the value 42 only when line~2 determines
+that the value loaded was 42.
+Default-zero initialization prevents this in both full C++ and (as we will
+see) real-world loose C++, but not an oracular loose C++ implementation.
+
+The next section looks at prior work on these issues.
 
@@
 thread, i.e., the rfe relation.
-As mentioned in Section~\ref{sec:Simple OOTA Cycle}, a CPU does
+As mentioned in Section~\ref{sec:Brief OOTA Overview}, a CPU does
 not make the value of a store instruction available for other CPUs to

### 2024-09-12 paper/ootavacuous.tex
ootavacuous: Compress Section 1.2 ("Prior Work")
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Some executable C++ memory models correctly flag at least some executions
-involving OOTA cycles~\cite{JadeAlglave2014HerdingCats}.\footnote{
-	Others cleverly avoid this issue by forbidding atomic
-	stores of nonconstant values~\cite{MarkBatty2011cppmem}.}
+Some executable C++ memory models correctly flag some OOTA
+cycles~\cite{JadeAlglave2014HerdingCats}.\footnote{
+	Others avoid OOTA by forbidding atomic stores of nonconstant
+	values~\cite{MarkBatty2011cppmem}.}
 % @@@ Better herd7 C11 citation?
@@
 P0442R0 (``Out-of-Thin-Air Execution is Vacuous'')~\cite{PaulEMcKenney2016OOTA}
-provided a decision procedure for classifying behaviors as permitted
-misordering on the one hand or disallowed OOTA on the other, using
-a perturbation method based on the insight that all OOTA behaviors are
-fixed-point computations.
+provided a decision procedure for distinguishing between reordering and
+OOTA, using a perturbation method based on the insight that OOTA cycles
+are fixed-point computations.
 
@@
 theoretical problem.
-This might have been acceptable in the 1960s of some of the authors'
-youths, but it is now the year 2024.
+% This might have been acceptable in the 1960s of some of the authors'
+% youths, but it is now the year 2024.
 
@@
 cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
-but none of these have been looked upon favorably by C++ implementers.
-Many of these workers appear to have abandoned this effort, but as of
-2024, Mark Batty is persisting with modular relaxed dependencies, which
-may have the added ability of checking compiler
+but none of these have been accepted by C++ implementers.
+As of 2024, Mark Batty is persisting with modular relaxed dependencies,
+which may have the added ability of checking compiler
 optimizations~\cite{JayRichards2024SymbolicMrder}.
@@
 % concerns.
-This paper focuses on examples lacking UB.
-
-All this work focused on either identifying OOTA or seeing how C++
-implementations could avoid it.
+We focus on UB-free examples.
+
+All this work focused on either identifying OOTA or helping C++
+implementations to avoid it.
 None applied real-world hardware ordering constraints to the problem
-of avoiding OOTA cycles,
-yet doing so might help explain why no known real-world C++ implementation
-results in OOTA behavior.
-We therefore dig more deeply into OOTA cycles in the light
-of these constraints.
+of avoiding OOTA cycles.
+This paper's contribution is showing how these constraints prevent
+real-world C++ implementations from exhibiting OOTA behavior.
 

### 2024-09-12 paper/ootavacuous.tex
ootavacuous: Further rework abstract
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	on traditional compilers and hardware, including CPUs and GPGPUs,
-	thus easing analysis of semantic dependencies and OOTA behaviors
-	via existing hardware models.
+	thus enabling analysis of semantic dependencies and OOTA behaviors
+	by means of existing hardware models.
 	We show that these models' constraints prevent OOTA cycles from
 	occurring in undefined-behavior-free C++ programs running on
-	compiler-based implementations that use single-thread analysis,
-	provided the cycles involve only volatile atomics.
-	We also extend this work to nonvolatile atomics by defining
-	``quasi-volatile'' behavior that C++ implementations must adhere
-	to in order to evaluate arithmetic expressions involving atomic
-	accesses in accordance with the standard.
+	such implementations, provided the cycles involve only volatile
+	atomics.
+	For implementations doing only single-threaded analysis, we
+	accommodate nonvolatile atomics by defining ``quasi-volatile''
+	behavior that must be adhered to in order to evaluate arithmetic
+	expressions involving atomic accesses in accordance with the
+	standard.
 

### 2024-09-12 paper/ootavacuous.tex
ootavacuous: Compress Section 5.1 ("Dependencies at the Hardware Level")
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Dependencies between machine instructions can be understood in terms
-of the flow of information within a CPU.
-Each instruction has a set of inputs and is the source of a set of outputs,
-some of which flow to the inputs of later instructions.
+Dependencies between machine instructions correspond to the flow of
+information within a CPU.
+Each instruction takes a set of inputs and provides a set of outputs,
+some of which flow to inputs of later instructions.
 The inputs determine what an instruction will do.
-A few examples:
-\begin{itemize}
-\item	An \co{add} instruction would typically have two inputs
-	(the register values to be added together) and two outputs
-	(the sum to be stored in a general-purpose register and some
-	condition-code bits---e.g., Zero, Carry, and Overflow---to be
-	stored in a status register).
+
+For example, an \co{add} instruction might have two inputs (the values to
+be summed) and two outputs (the sum and some condition-code bits---e.g.,
+Zero, Carry, and Overflow).
 % \item	A conditional-move instruction would have as inputs the
@@
 %	instruction-pointer register.
-\item	A memory-load instruction's input is the address to load from,
-	and its output is the value obtained by the load, to be stored
-	in a register.
-\item	A memory-store instruction's inputs are the value to store and
-	the address at which to store it; there are no outputs.
-\end{itemize}
-Note: Hardware dependency analysis considers only the information
-flowing \emph{within} a CPU, not information flowing between the
-CPU and memory, which is handled separately as part of the action
-taken by an instruction.
+For another example, a memory-load instruction's input is the address
+to load from, and its output is the value obtained by the load.
+For a final example, a memory-store instruction's inputs are the value
+to store and the address at which to store it; there are no outputs.
 
@@
 another instruction $I$ when any of $I$'s outputs flow into $J$'s inputs,
-perhaps indirectly via some intermediate instructions.
-Tracing back the flow of information between instructions, you can
-see that any hardware dependency must ultimately emanate from a load
-instruction or the thread's initial register values (or possibly some sort of
-input instruction, but we will not consider that complication here).
-These are the only sources of truly new information in a thread.
-
-The concept is simple, but it is important because of the way
-dependencies affect instruction ordering in weakly ordered architectures.
+perhaps via intermediate instructions.
+Any hardware dependency ultimately emanates from register initial values,
+load instructions, immediate values, or I/O accesses.
+We set aside I/O, so that load instructions are the only sources
+of truly new information in a thread.
+Dependencies are important because they affect instruction ordering in
+weakly ordered architectures.
 

### 2024-09-12 paper/ootavacuous.tex
ootavacuous: Compress Section 5.2 ("Instruction Ordering")
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 of truly new information in a thread.
-Dependencies are important because they affect instruction ordering in
-weakly ordered architectures.
 
@@
 A CPU may start executing an instruction speculatively, but at some
-point it must \emph{commit} to a decision either to definitely execute the
-instruction's action with some collection of well defined inputs and outputs
-or else to abandon it.
-Thanks to the law of cause and~effect, a CPU is not able to commit an
-instruction or its outputs until the relevant inputs' sources have committed.
-The reason is obvious: An input is subject to change at any time until
-its source commits to its value,
-and the instruction and its outputs can't commit until the CPU has fully
-determined what it will do and what they will be.
-(On the other hand, the CPU need not wait for inputs that won't affect
-the instruction's results.
-For instance, a conditional-move instruction wouldn't need to wait for
-the source register input to be determined once it knew that the
-condition was definitely false.
-And, although no architectures we are aware of do this, there is nothing
-in principle to prevent a CPU from committing a \co{multiply}
-instruction as soon as either one of its inputs' sources has committed to the
-value zero.)
-
-Of course, an instruction's inputs don't all have to be outputs from
-earlier instructions; some of them may be immediate constants present
-in the instruction itself.
-In this case there is no need to wait for those inputs because
-their values are fully determined from the start.
+point it must either abandon that instruction or \emph{commit} it with
+a set of specific outputs based on well-defined inputs.
+A CPU is clearly not permitted to commit an instruction its inputs have
+been committed, unless the remaining uncommitted inputs cannot affect
+that instruction's outputs.
+For instance, a conditional-move instruction needn't wait for the source
+input to commit if it knows that the condition is definitely false.
 
@@
 committing $J$'s action and outputs.
-Thus dependencies force instructions to commit in order, even on
-weakly ordered architectures.
-
+Thus dependencies force the affected instructions to commit in order,
+even on weakly ordered architectures.
 We will say that one instruction is \emph{ordered after} another to mean that
-the CPU forces it to commit after the other one commits.
-Hence instructions are ordered after the instructions they depend on.
-(Note that this implies nothing about when a load instruction retrieves
-its value from memory; it may do so long before it or a prior
-instruction commits.)
-
-Dependencies aren't the only mechanism that can lead to
-ordering in hardware executions.
-The simplest alternative is a load reading from a store in another
-thread, i.e., the rfe relation.
+it must commit after the other one commits.\footnote{
+	Note that this implies nothing about when a load instruction
+	retrieves its value from memory; it may do so long before it
+	commits.}
+
+Dependencies aren't the only ordering mechanism.
 As mentioned in Section~\ref{sec:Brief OOTA Overview}, a CPU does
-not make the value of a store instruction available for other CPUs to
-load until the store commits or some time later.
-And after this happens, it takes some time for the information about
-the store to travel from that CPU to others, owing to the inescapable
-facts that processors have nonzero size and information cannot travel
-faster than the speed of light.
-Since the load instruction cannot commit until its output (the value it
-reads) is fully determined, it must commit after the store it reads
+not make the value of a store instruction available to other CPUs
+until after the store commits.
+It then takes additional time for the stored value to travel to other
+CPUs, owing to finite speed of light and non-zero size of computing
+hardware.
+Since a load instruction cannot commit until its output (the value
+read) is fully determined, it must commit after the store it reads
 from.
-(To be fair, this should be considered more as a \emph{result} of
-ordering than as a \emph{cause} of ordering, in that
-if a load commits before a store on another thread then it
-cannot possibly read from that store.)
-
-A similarly straightforward mechanism for ordering is conditional or
-computed branches.
+
+Conditional or computed branches also provide ordering.
 An instruction executing after such a branch cannot commit
 until the CPU has committed to whether the branch will be taken and
-if taken, where it will branch to (that is, the output value it will
-send to the instruction-pointer register);
+if taken, where it will branch to;
 until then the CPU cannot know whether the later instruction should
@@
 Therefore instructions following a conditional- or computed-branch
-instruction in a hardware execution must commit after the branch, and hence
-after the source for the branch's condition or destination input.
-(There could theoretically be an exception for a branch that
-conditionally jumps to the immediately following instruction.
-We can ignore this possibility by treating such branches as no-ops.)
+instruction must commit after the branch, and hence after the source
+for the branch's condition and/or destination
+input.\footnote{
+	One exception is a branch that conditionally jumps to the
+	next instruction.
+	We ignore this possibility by treating such branches as no-ops.}
 

### 2024-09-12 paper/ootavacuous.tex
ootavacuous: Compress Section 5.3 ("The F.P. of sdep")
I used a much lighter hand on this section than on the others due
to its importance.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-With the contents of the last few sections under our belts, we can
-formulate the Fundamental Property that we would like all semantic
-dependencies to satisfy.
-Note that this formulation makes sense only for implementations in
-which all atomic objects are treated as volatile or quasi volatile.
+We can now formulate the Fundamental Property that we want all semantic
+dependencies to satisfy in implementations that treat all atomic objects
+as volatile or quasi volatile.
 \begin{quote}
@@
 Property has a nontrivial OOTA cycle.
-To see this, suppose that abstract execution $A$ is realized by hardware
-execution $H$ and $A$ does have an OOTA cycle.
+To see this, suppose that abstract execution $A$ having an OOTA cycle
+is realized by hardware execution $H$.
 This means there are atomic stores $W_i$ in $A$,
@@
 
-If all the stores in the OOTA cycle are omitted then all the reads
-must be omitted as well.
-In effect, the entire cycle has been optimized
-out of existence by the compiler.
-Although we are unable to prove it, we conjecture that in this
-situation there must be another abstract execution which has the
-same observable effects as $A$ and is also realized by $H$, but in
-which the OOTA cycle does not occur.
+If all the stores in the OOTA cycle are omitted then all the reads must
+also be omitted, meaning that the entire cycle has been optimized away.
+We conjecture that in this situation there must be another abstract
+execution which has the same observable effects as $A$ and is also
+realized by $H$, but in which the OOTA cycle does not occur.
 Thus there would be no way to tell, merely by observing the effects of
-$H$, whether there was an OOTA cycle or not.
-For this reason we declare realizations of OOTA cycles in which
-all the accesses are omitted to be \emph{trivial}.
+$H$, whether or not there was an OOTA cycle.
+We therefore declare OOTA cycles in which all accesses are omitted to
+be \emph{trivial}.
 

### 2024-09-12 paper/ootavacuous.tex
ootavacuous: Apply Alan Stern September 12 feedback
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	atomics.
-	For implementations doing only single-threaded analysis, we
-	accommodate nonvolatile atomics by defining ``quasi-volatile''
-	behavior that must be adhered to in order to evaluate arithmetic
-	expressions involving atomic accesses in accordance with the
-	standard.
+	For implementations doing only per-thread analysis and
+	optimization, we accommodate nonvolatile atomics by defining
+	``quasi-volatile'' behavior, adherence to which avoids situations
+	in which arithmetic-expression evaluation involving atomic
+	accesses could otherwise violate the standard.
 
@@
 Here, all of \co{x}, \co{y}, \co{r1}, and \co{r2} might have final
-values of 42, despite their being no instances of 42 in the default-zero
-initialization or in the executable
+values of 42, despite there being no instances of 42 in the default-zero
+initial values or in the executable
 code~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.\footnote{
 	Throughout this paper, variables whose names are ``\co{r}''
-	followed by a number denote registers, and other names denote
+	followed by a number denote CPU registers, and other names denote
 	shared variables.
-	All accesses to shared variable are relaxed atomic accesses.}
-First, P0's line~1 does a relaxed atomic load from \co{x} into \co{r1},
+	All shared-variable accesses are relaxed atomic.}
+First, P0's line~1 does a load from \co{x} into \co{r1},
 reading the value of P1's line-2 store rather than \co{x}'s initial
@@
 	are no longer speculative.
-	This same restriction applies to compiler-based speculation.}
+	A similar restriction applies to compiler-based speculation.}
 the stores on line~2 cannot take place until the P0's and P1's CPUs
@@
 Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
-Both the compiler and the CPU might reorder P1's lines~1 and~2, so that
-all of \co{x}, \co{y}, \co{r1}, and \co{r2} might have the value 42,
+Either the compiler or the CPU might reorder P1's lines~1 and~2, so that
+all of \co{x}, \co{y}, \co{r1}, and \co{r2} might end up with the value 42,
 as indicated by the ``sometimes satisfied'' in the figure.
@@
 
-On the other hand, OOTA cycles can arise even when value 42 is mentioned,
-as shown in
+On the other hand, OOTA cycles need not involve nonsensical or
+unmotivated values, as shown in
 Listing~\ref{lst:OOTA Cycle}.
@@
 that the value loaded was 42.
-Default-zero initialization prevents this in both full C++ and (as we will
-see) real-world loose C++, but not an oracular loose C++ implementation.
-
-The next section looks at prior work on these issues.
+However, this OOTA result has not been observed on real implementations
+due to the default-zero initialization combined with semantic dependencies
+from line~1 to line~3.
 
@@
 
-Goldblatt looked at interactions between OOTA cycles and
+Goldblatt looked at interactions between OOTA and
 undefined behavior (UB)~\cite{DavidGoldblatt2019NoElegantOOTAfix}.
@@
 % concerns.
-We focus on UB-free examples.
+We consider only UB-free examples.
 
@@
 of avoiding OOTA cycles.
-This paper's contribution is showing how these constraints prevent
+This paper's contribution is to show how these constraints prevent
 real-world C++ implementations from exhibiting OOTA behavior.

### 2024-09-12 paper/ootavacuous.tex
ootavacuous: Consistently use "per-thread analysis"
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 quasi volatile.
-(For compilers that perform only single-thread analysis, not global
+(For compilers that perform only per-thread analysis, not global
 analysis, quasi volatile is sufficient.)
@@
 In this section we consider implementations whose compilers perform
-only single-thread analysis and treat atomic objects as
+only per-thread analysis and treat atomic objects as
 quasi volatile.
@@
 \begin{itemize}
-\item	uses single-thread analysis and treats atomic objects as
+\item	uses per-thread analysis and treats atomic objects as
 	quasi volatile, or
@@
 
-However, a single-thread analysis compiler can translate this into the
+However, a per-thread analysis compiler can translate this into the
 machine-code equivalent of:
@@
 This paper focused solely on compilers, primarily those that do only
-local per-thread anaylsis and optimization.
+per-thread analysis and optimization.
 It introduced the notion of quasi-volatile object accesses and gave

### 2024-09-12 paper/ootavacuous.tex
ootavacuous: Apply more Alan Stern September 12 feedback
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	Throughout this paper, variables whose names are ``\co{r}''
-	followed by a number denote CPU registers, and other names denote
-	shared variables.
+	followed by a number denote private per-process variables
+	(``registers''),
+	and other names denote shared variables.
 	All shared-variable accesses are relaxed atomic.}
-First, P0's line~1 does a load from \co{x} into \co{r1},
+First, process P0's line~1 does a load from \co{x} into \co{r1},
 reading the value of P1's line-2 store rather than \co{x}'s initial
@@
 standard explicitly prohibits them in 33.5.4p8
-(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950},
-which prevents misapplication of the as-if rule in oracular C++
-implementations:
+(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}
+(which prevents misapplication of the as-if rule in oracular C++
+implementations):
 \begin{quote}
@@
 that the value loaded was 42.
-However, this OOTA result has not been observed on real implementations
-due to the default-zero initialization combined with semantic dependencies
-from line~1 to line~3.
+This, combined with the default-zero initialization, explains why this
+OOTA result has not been observed on real implementations.
 

### 2024-09-13 paper/ootavacuous.tex
treewide: Use "threads" and "Tn" to align with C++ nomenclature
No more "processes" named "Pn".  Also fix a grammar nit while in the
area.

Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{tabular}{r|l|l}
-& \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
+& \multicolumn{1}{c}{T0} & \multicolumn{1}{c}{T1} \\
 \hline
@@
 	All shared-variable accesses are relaxed atomic.}
-First, process P0's line~1 does a load from \co{x} into \co{r1},
-reading the value of P1's line-2 store rather than \co{x}'s initial
+First, thread T0's line~1 does a load from \co{x} into \co{r1},
+reading the value of T1's line-2 store rather than \co{x}'s initial
 value of zero and somehow obtaining 42.
-Second, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
-Third, P1's line~1 loads 42 from \co{y} to \co{r2}.
-Finally, P1's line~2 stores 42 to \co{x}, justifying the value loaded
-by P0's line~1.
+Second, T0's line~2 stores \co{r1}, and thus 42, to \co{y}.
+Third, T1's line~1 loads 42 from \co{y} to \co{r2}.
+Finally, T1's line~2 stores 42 to \co{x}, justifying the value loaded
+by T0's line~1.
 
@@
 there is a \emph{semantic dependency} from line~1 to line~2 in both
-P0 and P1.
+T0 and T1.
 (Roughly speaking, there is a semantic dependency from a given load
@@
 	A similar restriction applies to compiler-based speculation.}
-the stores on line~2 cannot take place until the P0's and P1's CPUs
+the stores on line~2 cannot take place until the T0's and T1's CPUs
 know what values are returned by the loads on line~1.
@@
 Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
-Either the compiler or the CPU might reorder P1's lines~1 and~2, so that
+Either the compiler or the CPU might reorder T1's lines~1 and~2, so that
 all of \co{x}, \co{y}, \co{r1}, and \co{r2} might end up with the value 42,
@@
 \begin{tabular}{r|l}
-% & \multicolumn{1}{c}{P0} \\
+% & \multicolumn{1}{c}{T0} \\
 % \hline
@@
 % \end{quote}
-This is an elaboration of P1 from
+This is an elaboration of T1 from
 Listing~\ref{lst:Simple OOTA}
@@
 page~\pageref{lst:Simple OOTA},
-in which P0 loads the value of \co{x} and stores it in \co{y} while P1
+in which T0 loads the value of \co{x} and stores it in \co{y} while T1
 does the reverse.
@@
 \begin{tabular}{r|l|l}
-% & \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
+% & \multicolumn{1}{c}{T0} & \multicolumn{1}{c}{T1} \\
 % \hline
@@
 a set of specific outputs based on well-defined inputs.
-A CPU is clearly not permitted to commit an instruction its inputs have
-been committed, unless the remaining uncommitted inputs cannot affect
-that instruction's outputs.
+A CPU is clearly not permitted to commit an instruction until its inputs
+have been committed, unless the remaining uncommitted inputs cannot
+affect that instruction's outputs.
 For instance, a conditional-move instruction needn't wait for the source
@@
 not count as OOTA according to our definition because the
-dependencies in P0 and P1 would not be semantic.
+dependencies in T0 and T1 would not be semantic.
 

### 2024-09-13 paper/litmus2table.sh

@@ -81,7 +81,7 @@ END {
 	head = "";
 	for (i = 0; i <= procnum; i++) {
 		cols = cols "|l";
-		head = head "& " bs "multicolumn{1}{c}{P" i "} ";
+		head = head "& " bs "multicolumn{1}{c}{T" i "} ";
 	}
 	print bs "begin{tabular}{" cols "}"
 	print head bs bs;

### 2024-09-13 paper/ootavacuous.tex
ootavacuous: Untangle quasi-volatile sentence in abstract
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	atomics.
-	For implementations doing only per-thread analysis and
-	optimization, we accommodate nonvolatile atomics by defining
-	``quasi-volatile'' behavior, adherence to which avoids situations
-	in which arithmetic-expression evaluation involving atomic
-	accesses could otherwise violate the standard.
+	We accommodate nonvolatile atomics by defining ``quasi-volatile''
+	behavior that enables implementations doing per-thread analysis
+	and optimization to not only avoid OOTA, but also to avoid 
+	incorrectly evaluating arithmetic expressions involving atomic
+	accesses.
 

### 2024-09-13 paper/ootavacuous.tex
ootavacuous: Fix Listing 3 discussion
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 that the value loaded was 42.
-This, combined with the default-zero initialization, explains why this
-OOTA result has not been observed on real implementations.
+As with Listing~\ref{lst:Simple OOTA}, hardware ordering prevents the
+OOTA result, as discussed further in
+Section~\ref{sec:Hardware Dependencies, Instruction Ordering, and the Fundamental Property}.
 

### 2024-09-13 paper/ootavacuous.tex
ootavacuous: Apply September 13 Alan Stern feedback
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-All OOTA workers owe a debt to the foundational work in the infamous
+All OOTA workers owe a debt of gratitude to the foundational work led
+by William Pugh leading to the infamous Java-language
 ``Causality Test Cases''.\footnote{
@@
 
-This section formulates OOTA cycles as cycles in sdep $\cup$ rfe instead
-of the traditional sdep $\cup$ rf, then shows that sdep is not a strict
-function of the source code.
+This section analyzes the main components of OOTA cycles, namely
+semantic dependencies and reads-from links.
 
@@
 % of N4323 ("Out-of-Thin-Air Execution is Vacuous").
-where sdep is the set of semantic dependencies within
-each thread and rf is the set of store-to-load ``reads-from'' links,
-whether within a thread (rfi) or between threads (rfe).
+where sdep is the set of semantic dependencies within each thread and
+rf is the set of store-to-load ``reads-from'' links, whether internally
+within a thread (rfi) or externally between threads (rfe).
 % @@@ Consider adding citation defining rf, rfi, and rfe.

### 2024-09-13 paper/ootavacuous.tex
ootavacuous: s/Single-Thread/Per-Thread/
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Apply second round September 13 Alan Stern feedback
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	behavior that enables implementations doing per-thread analysis
-	and optimization to not only avoid OOTA, but also to avoid 
+	and optimization not only to avoid OOTA, but also to avoid 
 	incorrectly evaluating arithmetic expressions involving atomic
@@
 	Throughout this paper, variables whose names are ``\co{r}''
-	followed by a number denote private per-process variables
+	followed by a number denote private per-thread variables
 	(``registers''),
@@
 Listing~\ref{lst:OOTA Cycle}.
-Line~3 of both processes stores the value 42 only when line~2 determines
+Line~3 of both threads stores the value 42 only when line~2 determines
 that the value loaded was 42.
@@
 
-\subsection{For Compilers Using Single-Thread Analysis}
-\label{sec:For Compilers Using Single-Thread Analysis}
+\subsection{For Compilers Using Per-Thread Analysis}
+\label{sec:For Compilers Using Per-Thread Analysis}
 

### 2024-09-13 paper/ootavacuous.tex
ootavacuous: Apply third round September 13 Alan Stern feedback
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 All OOTA workers owe a debt of gratitude to the foundational work led
-by William Pugh leading to the infamous Java-language
+by William Pugh, represented by the infamous Java-language
 ``Causality Test Cases''.\footnote{
@@
 
-This section uncovers some semantic-dependency complexities.
-Section~\ref{sec:Semantic Dependencies and Source Code}
-shows that semantic dependencies are functions of executions,
-rather than being strict functions of the source code.
-Section~\ref{sec:Semantic Dependencies Can Be Many-To-One}
-shows that semantic dependencies do not necessarily extend
-from a single load to a single store, but can instead involve
-multiple loads.
-Section~\ref{sec:Semantic Dependencies Affected by Cross-Thread Optimizations}
-shows that semantic dependencies can be affected by cross-thread
-optimizations.
-Sections~\ref{sec:Semantic Dependencies Affected by if Statements}
-and~\ref{sec:Semantic Dependencies Not Affected by if Statements}
-show that \co{if} statements can have surprising effects on semantic
-dependencies, up to and including eliminating them completely.
-Finally,
-Section~\ref{sec:Semantic Dependencies and Matching Up Stores}
-demonstrates some challenges in determining which of a group of
-stores is involved in a given semantic dependency.
+This section discusses semantic-dependency complexities.
 

### 2024-09-15 paper/ootavacuous.tex
Numerous minor edits
These changes are meant to fix minor errors and improve readability
(including the indentation in Listing 3 and the spacing around the
unnumbered listing in section 4.4).  Nothing major has been altered and
we're still under 11 pages.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	A similar restriction applies to compiler-based speculation.}
-the stores on line~2 cannot take place until the T0's and T1's CPUs
+the stores on line~2 cannot take place until T0's and T1's CPUs
 know what values are returned by the loads on line~1.
@@
 \begin{listing}[tbp]
-@@ DisplayRunLitmus litmus/oota-ctrl.litmus @@
+% Constructed by litmus2table.sh and modified by hand
+{
+\small
+\begin{tabular}{r|l|l}
+& \multicolumn{1}{c}{T0} & \multicolumn{1}{c}{T1} \\
+\hline
+1. &
+\texttt{~int r0 = y} &
+	\texttt{~int r1 = x} \\
+2. &
+\texttt{~if (r0 == 42)} &
+	\texttt{~if (r1 == 42)} \\
+3. &
+\texttt{\qquad x = 42} &
+	\texttt{\qquad y = 42} \\
+4. &
+\texttt{~else} &
+	\texttt{~else} \\
+5. &
+\texttt{\qquad x = r0} &
+	\texttt{\qquad y = r1} \\
+\end{tabular}
+
+\vspace{0.1in}
+Condition (0:r0=42 $\wedge$ 1:r1=42)
+sometimes satisfied.
+}
 \caption{OOTA Cycle}
@@
 stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
-but this can require real instructions be
-executed~\cite[Section 7.1]{Maranget2012TutorialARMPower},
+but this can require executing real
+instructions~\cite[Section 7.1]{Maranget2012TutorialARMPower},
 consuming real time and real electrical power to solve a strictly
@@
 removed from the source code and line~4's \co{r3} had been replaced by
-\co{r2}.
-This means that line~4 stores the same value to \co{z} that was stored
+\co{r2},
+causing line~4 always to store the same value to \co{z} as was stored
 to \co{y} by line~2.
-And given that the load from \co{y} no longer exists, it cannot possibly
+And since the load from \co{y} has been removed, it cannot possibly
 act as a temporal constraint.
@@
 and thus no impediment to the occurrence of an OOTA cycle.
-Later, we will see examples of this destruction.
+We will see examples of this destruction later on.
 
@@
 
-This section discusses semantic-dependency complexities.
+Here we consider some of the complexities of semantic dependency.
 
@@
 \end{quote}
-This makes OOTA more complicated than a simple cycle but
+This can make OOTA more complicated than a simple cycle but
 we will continue to refer to ``OOTA cycles'' out of habit.
@@
 
-\emph{Tagline: ``Need to look at machine code.''}
-
-This section looks more carefully at executions from the viewpoint of
+\emph{Tagline: ``Need to look at both source code and machine code.''}
+
+This section looks more carefully at executions from the viewpoints of
 the abstract machine and the computer hardware, and then reconciles
@@
 pattern of loads and stores be consistent with the C++ memory model.
-We assume that programs will not indulge in any computations that
+(We assume that programs will not indulge in any computations that
 could vary spontaneously from one execution to another,
-such as basing a dependency on the time of day or a process ID.
+such as basing a dependency on the time of day or a process ID.)
 In short, the standard grants C++ implementations considerable freedom,
@@
 rules of the abstract machine'' (4.1.2p6.1 \co{[intro.abstract]}); and
-(3)~(There is a condition on the timing and interleaving of input and output,
-which does not matter for our purposes.)
+(3)~(A condition on the timing and interleaving of input and output,
+which does not matter for our purposes).
 We will say that the abstract execution is \emph{realized by} the
@@
 
-This section discusses influences and constraints on C++ compilers,
-then uses this information to define necessary properties of
-\emph{quasi-volatile} atomic operations.
+This section discusses influences and constraints on C++ compilers
+and then uses this information to define necessary properties of
+volatile and \emph{quasi-volatile} atomic accesses.
 
@@
 transform that program to the following before translating it into
-machine code: \\
+machine code:
 % Constructed by hand.
-{
+\begin{quote}
 \small
@@
 \end{tabular}
-} \\
-The loads previously on lines~6 and~12 have been replaced by constants.
+\end{quote}
+The loads previously on line~1 have been replaced by constants.
 Such a transformation complies with the loose C++ standard,
@@
 
-\subsection{Inventing Atomic Loads Can Destroy Semantic Dependencies}
-\label{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}
+\subsection{Inventing Atomic Loads Can Cause Errors and
+Destroy Semantic Dependencies}
+\label{sec:Inventing Atomic Loads Can Cause Errors and
+Destroy Semantic Dependencies}
 
@@
 \end{quote}
-For sufficiently small values of \co{x}, the abstract machine is
+For values of \co{x} small enough not to cause overflow, the abstract machine is
 guaranteed to produce a perfect square in \co{r1}.
@@
 a perfect square.
-This fails to provide the abstract machine's guarantees.
+This fails to satisfy the abstract machine's guarantee
+and therefore is an invalid transformation.
 
@@
 for \co{r1} whichever value agrees with \co{r2}, as shown on line~6.
-Then the value stored to \co{z} on line~7 will always be one,
+Then the value stored to \co{z} on line~7 will simply be one,
 with no dependence on the value loaded from \co{y}.
@@
 
-Applying these requirements to specific examples results in prohibiting
+These requirements prohibit
 invention or duplication of atomic loads and stores.
-However, in the absence of ordering constraints, it permits omitting
+However, they do permit omitting
 redundant non-volatile atomic stores and fusing of non-volatile atomic
@@
 
-This section examines dependencies at the machine-instruction level,
-then combines this with mapping from the previous section to define the
-fundamental property of semantic dependencies.
+This section examines dependencies at the level of machine instructions
+and then uses the mapping from the previous section to define the
+Fundamental Property of semantic dependencies.
 
@@
 perhaps via intermediate instructions.
-Any hardware dependency ultimately emanates from register initial values,
+Tracing back, you can see that
+any hardware dependency ultimately emanates from initial register values,
 load instructions, immediate values, or I/O accesses.
-We set aside I/O, so that load instructions are the only sources
+We are setting I/O aside, which leaves load instructions as the only sources
 of truly new information in a thread.
@@
 For instance, a conditional-move instruction needn't wait for the source
-input to commit if it knows that the condition is definitely false.
+input to commit once it knows that the condition is definitely false.
 
@@
 It then takes additional time for the stored value to travel to other
-CPUs, owing to finite speed of light and non-zero size of computing
-hardware.
+CPUs, owing to the finite speed of light and the non-zero size of
+processor hardware.
 Since a load instruction cannot commit until its output (the value
@@
 more tractable.
-This insight allows us to demonstrate that OOTA cycles cannot form
+This allows us to demonstrate that OOTA cycles cannot form
 in the C++ implementations addressed by this paper.
@@
 
-Let $W'$ and $\{R'_i\}$ be the accesses in $H$ that $W$ and the
-non-omitted $\{R_i\}$ loads map to.
+Let $W'$ and $R'_i$ be the accesses in $H$ that $W$ and the
+non-omitted $R_i$ loads map to.
 We then require that the hardware executions of $T$ in $H$ and $G$ be
@@
 
-First case: $W'$ lies in the initial period of $H$.
+First case: $W'$ lies in the initial period of $T$ in $H$.
 During the initial period of the hardware executions, $H$ and $G$
@@
 
-Second case: $W'$ lies in the common period of $H$.
+Second case: $W'$ lies in the common period of $T$ in $H$.
 Since the action of $W'$ in $H$ is different from its action in $G$,
@@
 
-Third case: $W'$ lies in the divergent part of $H$.
+Third case: $W'$ lies in the divergent part of $T$ in $H$.
 This happens when $W'$ comes after the conditional or computed branch
@@
 the example in
-Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}:
+Section~\ref{sec:Inventing Atomic Loads Can Cause Errors and
+Destroy Semantic Dependencies}:
 \begin{quote}

### 2024-09-15 paper/litmus2table.sh
litmus2table.sh: Use \\qquad to get proper indentation
Based on Alan Stern's early by-hand change.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@ -6,7 +6,7 @@
 
 sed -e 's/&/\\&/' | sed -e 's/	}/	\\}/' | sed -e 's/{/\\{/' |
 sed -e 's,\\/,$\\vee$,g' | sed -e 's,/\\,$\\wedge$,g' |
-sed -e 's/	/\~/g' |
+sed -e 's/	//' | sed -e 's/	/\\qquad /g' |
 awk -v amp='&' -v bs='\\' -v dq='"' '
 BEGIN {
 	ininit = 0;

### 2024-09-15 paper/ootavacuous.tex
litmus2table.sh: Use \\qquad to get proper indentation
Based on Alan Stern's early by-hand change.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Nit: s/not to cause/to avoid/
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{listing}[tbp]
-% Constructed by litmus2table.sh and modified by hand
-{
-\small
-\begin{tabular}{r|l|l}
-& \multicolumn{1}{c}{T0} & \multicolumn{1}{c}{T1} \\
-\hline
-1. &
-\texttt{~int r0 = y} &
-	\texttt{~int r1 = x} \\
-2. &
-\texttt{~if (r0 == 42)} &
-	\texttt{~if (r1 == 42)} \\
-3. &
-\texttt{\qquad x = 42} &
-	\texttt{\qquad y = 42} \\
-4. &
-\texttt{~else} &
-	\texttt{~else} \\
-5. &
-\texttt{\qquad x = r0} &
-	\texttt{\qquad y = r1} \\
-\end{tabular}
-
-\vspace{0.1in}
-Condition (0:r0=42 $\wedge$ 1:r1=42)
-sometimes satisfied.
-}
+@@ DisplayRunLitmus litmus/oota-ctrl.litmus @@
 \caption{OOTA Cycle}
@@
 \end{quote}
-For values of \co{x} small enough not to cause overflow, the abstract machine is
+For values of \co{x} small enough to avoid overflow, the abstract machine is
 guaranteed to produce a perfect square in \co{r1}.

### 2024-10-02 paper/ootavacuous.tex
ootavacuous: Move as-if rule caution after OOTA quote
Yes, this causes Section 1.1 to break onto page 2, but it makes
that passage flow much better.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 standard explicitly prohibits them in 33.5.4p8
-(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}
-(which prevents misapplication of the as-if rule in oracular C++
-implementations):
+(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
 \begin{quote}
@@
 \end{quote}
+This prohibition prevents misapplication of the as-if rule in oracular
+C++ implementations.
+
 This paper uses \emph{full C++} to denote the standard including this

### 2024-10-02 paper/ootavacuous.tex
ootavacuous: x<->y in 2.1 to align with Listing 1
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\emph{Tagline: ``Semantic dependencies are more complicated than you think.''}
-
 This section analyzes the main components of OOTA cycles, namely
@@
 1. &
-\texttt{~~int r2 = x} \\
+\texttt{~~int r2 = y} \\
 2. &
-\texttt{~~y = r2} \\
+\texttt{~~x = r2} \\
 3. &
-\texttt{~~int r3 = y} \\
+\texttt{~~int r3 = x} \\
 4. &
@@
 note that line~3 could well execute immediately after line~2, giving
-other threads no chance to modify \co{y} in between.
+other threads no chance to modify \co{x} in between.
 Such an implementation might therefore behave as if line~3 had been
@@
 causing line~4 always to store the same value to \co{z} as was stored
-to \co{y} by line~2.
-And since the load from \co{y} has been removed, it cannot possibly
+to \co{x} by line~2.
+And since the load from \co{x} has been removed, it cannot possibly
 act as a temporal constraint.

### 2024-10-02 paper/ootavacuous.tex
ootavacuous: Remove "|" in 2.1 example
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavactuous: Multi-variable change only sometimes
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \small
-\begin{tabular}{r|l}
+\begin{tabular}{rl}
 % & \multicolumn{1}{c}{T0} \\
@@
 single-variable perturbations;
-they must consider changes to multiple variables.
+they must at times consider changes to multiple variables.
 % @@@  Pull anything in?  Doesn't seem necessary.

### 2024-10-02 paper/ootavacuous.tex
ootavacuous: 2.2.4: Affects both value and execution
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous.tex: Remove L1 from 2.2.6 example
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 it extends to \co{y} and in others to \co{z}.
-This is an example of a load affecting not the value of
-a given store, but rather whether or not that store is executed at all.
+This is an example of a load affecting not only the value of
+a given store, but also whether or not that store is executed at all.
 
@@
 \begin{verbatim}
-if (x > 0) {
-    L1: y = 42;
-} else {
-    y = 53;
-    y = 42;
-}
+ 1 if (x > 0) {
+ 2     y = 42;
+ 3 } else {
+ 4     y = 53;
+ 5     y = 42;
+ 6 }
 \end{verbatim}
 \end{quote}
-Consider an execution in which \co{x} is greater than zero, so the
-statement labeled \co{L1} runs.
+Consider an execution in which \co{x} is greater than zero, so line~2
+runs.
 Is it semantically dependent on \co{x}?
@@
 \co{y} occurs, but 53 is written before it.
-Which of these two stores should be compared with the store in \co{L1}?
+Which of these two stores should be compared with the store on line~2?
 
@@
 they occur:
-Since \co{L1} is the first store to \co{y} in its arm of the \co{if}
+Since line~2 is the first store to \co{y} in its arm of the \co{if}
 statement, it should be matched up with the first store to \co{y} in
@@
 arm of the \co{if} statement \emph{would} agree with the store in
-\co{L1}, and so there would not be a semantic dependency.
+line~2, and so there would not be a semantic dependency.
 Once again, the decision is up to the implementation.

### 2024-10-02 paper/ootavacuous.tex
ootavacuous: Strengthen Section 2 close
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-We have seen several examples showing that semantic dependencies may vary
-according to the execution and even the implementation.
+We have seen several examples showing that semantic dependencies are more
+complicated than might first appear, varying according to the execution
+and even the implementation.
 This raises several questions, of which the first is:

### 2024-10-02 paper/ootavacuous.tex
ootavacuous: Shorten last sentence of first paragraph
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\emph{Tagline: ``Need to look at both source code and machine code.''}
-
 This section looks more carefully at executions from the viewpoints of
@@
 describing exactly what patterns of loads and stores can occur.
-Thus, leaving aside questions of asynchronous interrupts and system
+Thus, leaving aside asynchronous interrupts and system
 calls, the behavior of a CPU executing a particular thread within a
-program is entirely determined by the values obtained by the various
+program is entirely determined by the values obtained by its
 memory-load instructions.\footnote{

### 2024-10-02 paper/ootavacuous.tex
ootavacuous: Shorten last paragraph of Section 3.2
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-A computer may execute the instructions in a thread out of order.
+A computer may execute a thread's instructions out of order.
 The architecture specifies the extent to which this may happen, and it
-also specifies circumstances under which some pairs of instructions
+also specifies circumstances under which instructions
 must be executed in order.

### 2024-10-02 paper/ootavacuous.tex
ootavacuous: Remove Section 4 tagline
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Section 4.4 s/z/x/ in code example
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Add "full C++" qualifier for last of Section 5.3
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\emph{Tagline: Compilers have considerable freedom, but are constrained
-in optimizing atomics, especially volatile atomics.}
-
 In addition to a compiler, a complete C++ implementation might include
@@
 \texttt{~~y = r1} &
-	\texttt{~~z = r2} \\
+	\texttt{~~x = r2} \\
 \end{tabular}
@@
 also be omitted, meaning that the entire cycle has been optimized away.
-We conjecture that in this situation there must be another abstract
-execution which has the same observable effects as $A$ and is also
-realized by $H$, but in which the OOTA cycle does not occur.
+We conjecture that in this situation for full C++ there must be another
+abstract execution which has the same observable effects as $A$ and is
+also realized by $H$, but in which the OOTA cycle does not occur.
 Thus there would be no way to tell, merely by observing the effects of

### 2024-10-02 paper/ootavacuous.tex
ootavacuous: Flatten list at end of Section 6.3
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 either
-\begin{itemize}
-\item	uses per-thread analysis and treats atomic objects as
-	quasi volatile, or
-\item	uses global analysis and treats atomic objects as volatile,
-\end{itemize}
+(1)~uses per-thread analysis and treats atomic objects as quasi
+volatile, or
+(2)~uses global analysis and treats atomic objects as volatile,
 then programs produced by that implementation will never exhibit OOTA.

### 2024-10-02 paper/ootavacuous.tex
ootavacuous: Make first sentence of 7.1 more explicit
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Section 7.2 header s/versus/vs./
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-A drawback of the definition is that it is only relative to a specific
-compiler or implementation.
+One drawback of our definition of semantic dependency is that it is only
+relative to a specific compiler or implementation.
 This may strike some people as wishy-washy and avoiding the real
@@
 
-\subsection{Global Analysis and Volatile versus Quasi Volatile}
-\label{sec:Global Analysis and Volatile versus Quasi Volatile}
+\subsection{Global Analysis and Volatile vs.\ Quasi Volatile}
+\label{sec:Global Analysis and Volatile vs. Quasi Volatile}
 

### 2024-10-02 paper/ootavacuous.tex
ootavacuous: Separate future work from conclusions
Pull the list of future work into a new section 7.5 in order
to strengthen the end of the paper.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Move Sections 4.1 and 4.2 to Section 2.2
Also add references in the list in Sction 3.1.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+\subsubsection{Non-Deterministic Semantic Dependencies}
+\label{sec:Non-Deterministic Semantic Dependencies}
+
+Compiler optimizers also exploit the nondeterministic aspects of an
+execution.
+Consider this example, with \co{i} initially zero:
+\begin{quote}
+\begin{verbatim}
+int foo(int a, int b)
+{
+   return a / b;
+}
+
+r1 = foo(++i, ++i);
+x = r1 * z;
+\end{verbatim}
+\end{quote}
+Because early C~implementers could not come to agreement, the standard
+does not specify the order of evaluation of function arguments, so
+the value calculated for \co{r1} might be zero ($1/2$ truncated) or two
+($2/1$).
+In the former case there is no semantic dependency from \co{z} to \co{x},
+but in the latter case there is.\footnote{
+	Thanks to Peter Sewell for pointing out this possibility.}
+
+(According to the current version of the standard, conflicting side effects
+in unsequenced subexpressions constitute undefined behavior,
+although there are proposals to make them defined in both
+C++~\cite{GabrielDosReis2016P0145r3}
+and C~\cite{AlexCeleste2023N3203}.
+Nevertheless, the example above is allowed because the order of evaluation
+of arguments to a function call is ``indeterminately sequenced''
+(7.6.1.3p7 \co{[expr.call]}) rather than unsequenced, a subtle distinction.)
+
 \subsubsection{Semantic Dependencies Affected by Cross-Thread Optimizations}
@@
 
+\subsubsection{Per-Architecture Dependent Semantic Dependencies}
+\label{sec:Per-Architecture Dependent Semantic Dependencies}
+
+The implementation-defined aspects of the standard permit architectures to
+differ as to whether an abstract execution contains a semantic dependency.
+For example, consider the following, where the type of \co{c} is \co{char}:
+\begin{quote}
+\begin{verbatim}
+y = (c >= 0);
+\end{verbatim}
+\end{quote}
+Here \co{y} is semantically dependent on \co{c} in executions
+running on implementations in
+which \co{char} is a signed type,
+but not those for which it is unsigned.
+
 We have seen several examples showing that semantic dependencies are more
@@
 	bits in the various integer types
-	or whether the \co{char} type is signed.
+	or whether the \co{char} type is signed (see
+	Section~\ref{sec:Per-Architecture Dependent Semantic Dependencies}).
 \item	Some aspects of an execution are unspecified or nondeterministic,
@@
 	of most binary operators or of the arguments in a function call.
-	Implementations may choose from a set of allowed behaviors.
+	Implementations may choose from a set of allowed behaviors
+	(see
+	Section~\ref{sec:Non-Deterministic Semantic Dependencies}).
 \item	Some actions are deemed to have undefined results; the standard
@@
 volatile and \emph{quasi-volatile} atomic accesses.
-
-\subsection{Architectures Influence the Behavior of Compilers}
-\label{sec:Architectures Influence the Behavior of Compilers}
-
-The implementation-defined aspects of the standard permit architectures to
-differ as to whether an abstract execution contains a semantic dependency.
-For example, consider the following, where the type of \co{c} is \co{char}:
-\begin{quote}
-\begin{verbatim}
-y = (c >= 0);
-\end{verbatim}
-\end{quote}
-Here \co{y} is semantically dependent on \co{c} in executions
-running on implementations in
-which \co{char} is a signed type,
-but not those for which it is unsigned.
-
-\subsection{Optimizations Influence the Behavior of Compilers}
-\label{sec:Optimizations Influence the Behavior of Compilers}
-
-Compiler optimizers also exploit the nondeterministic aspects of an
-execution.
-Consider this example, with \co{i} initially zero:
-\begin{quote}
-\begin{verbatim}
-int foo(int a, int b)
-{
-   return a / b;
-}
-
-r1 = foo(++i, ++i);
-x = r1 * z;
-\end{verbatim}
-\end{quote}
-Because early C~implementers could not come to agreement, the standard
-does not specify the order of evaluation of function arguments, so
-the value calculated for \co{r1} might be zero ($1/2$ truncated) or two
-($2/1$).
-In the former case there is no semantic dependency from \co{z} to \co{x},
-but in the latter case there is.\footnote{
-	Thanks to Peter Sewell for pointing out this possibility.}
-
-(According to the current version of the standard, conflicting side effects
-in unsequenced subexpressions constitute undefined behavior,
-although there are proposals to make them defined in both
-C++~\cite{GabrielDosReis2016P0145r3}
-and C~\cite{AlexCeleste2023N3203}.
-Nevertheless, the example above is allowed because the order of evaluation
-of arguments to a function call is ``indeterminately sequenced''
-(7.6.1.3p7 \co{[expr.call]}) rather than unsequenced, a subtle distinction.)
 
@@
 
-\section{Conclusions and Future Directions}
-\label{sec:Conclusions and Future Directions}
-
-This paper focused solely on compilers, primarily those that do only
-per-thread analysis and optimization.
-It introduced the notion of quasi-volatile object accesses and gave
-a definition for semantic dependency that is relative to a particular
-C++ implementation.
-It formulated a Fundamental Property of semantic dependency and proved
-that implementations subject to some very mild restrictions on how
-they treat atomic object accesses do have this property.
-Finally, it demonstrated that satisfying the Fundamental Property
-guarantees an implementation cannot give rise to OOTA executions.
-
-The conclusion to be drawn from this work is that in this context,
-avoiding OOTA cycles requires no changes to the standard,
-to current implementions, or to user practices for portable code.
-Further explorations could:
+\subsection{Other Potential Refinements}
+\label{sec:Other Potential Refinements}
+
+Additional future work could:
 \begin{itemize}
@@
 
+\section{Summary and Conclusions}
+\label{sec:Summary and Conclusions}
+
+This paper focused solely on compilers, primarily those that do only
+per-thread analysis and optimization.
+It introduced the notion of quasi-volatile object accesses and gave
+a definition for semantic dependency that is relative to a particular
+C++ implementation.
+It formulated a Fundamental Property of semantic dependency and proved
+that implementations subject to some very mild restrictions on how
+they treat atomic object accesses do have this property.
+Finally, it demonstrated that satisfying the Fundamental Property
+guarantees an implementation cannot give rise to OOTA executions.
+
+The conclusion to be drawn from this work is that in this context,
+avoiding OOTA cycles requires no changes to the standard,
+to current implementions, or to user practices for portable code.
+
 % \clearpage

### 2024-10-03 paper/ootavacuous.tex
Move the text about spontaneously varying computations
Move the parenthetical remark in 3.1 about spontaneously varying
computations to the end of the section, where it fits better into the
surrounding context.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 pattern of loads and stores be consistent with the C++ memory model.
-(We assume that programs will not indulge in any computations that
-could vary spontaneously from one execution to another,
-such as basing a dependency on the time of day or a process ID.)
 In short, the standard grants C++ implementations considerable freedom,
@@
 in every respect during those steps, although they may diverge later.
+(We assume that programs will not indulge in any computations that
+could vary spontaneously from one execution to another,
+such as basing a dependency on the time of day or a process ID.)
 

### 2024-10-03 paper/ootavacuous.tex
Add periods after line numbers
Add periods following the line numbers for inline listings that lack them.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Convert inline listing to quote
Make the inline listing in 2.1 a regular quote, not a table, to match
the other inline listings.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 standard, but it has a problem with intrathread rfi links
-as exemplified by the following code: \\
-% \begin{quote}
-% Constructed by hand.
-{
-\small
-\begin{tabular}{rl}
-% & \multicolumn{1}{c}{T0} \\
-% \hline
-1. &
-\texttt{~~int r2 = y} \\
-2. &
-\texttt{~~x = r2} \\
-3. &
-\texttt{~~int r3 = x} \\
-4. &
-\texttt{~~z = r3} \\
-\end{tabular}
-} \\
-% \end{quote}
+as exemplified by the following code:
+\begin{quote}
+\begin{verbatim}
+1. int r2 = y;
+2. x = r2;
+3. int r3 = x;
+4. z = r3;
+\end{verbatim}
+\end{quote}
 This is an elaboration of T1 from
@@
 \begin{verbatim}
- 1 if (x > 0) {
- 2     y = 42;
- 3 } else {
- 4     y = 53;
- 5     y = 42;
- 6 }
+ 1. if (x > 0) {
+ 2.     y = 42;
+ 3. } else {
+ 4.     y = 53;
+ 5.     y = 42;
+ 6. }
 \end{verbatim}
@@
 \begin{verbatim}
- 1  int r1;
- 2  int r1a = (x != 0);
- 3  int r1b = (x != 0);
- 4  int r2 = (y != 0);
- 5  if (r1a != r1b) {
- 6      r1 = r2;
- 7      z = 1;
- 8  } else {
- 9      r1 = r1b;
-10      z = (r1 == r2);
-11  }
+ 1. int r1;
+ 2. int r1a = (x != 0);
+ 3. int r1b = (x != 0);
+ 4. int r2 = (y != 0);
+ 5. if (r1a != r1b) {
+ 6.     r1 = r2;
+ 7.     z = 1;
+ 8. } else {
+ 9.     r1 = r1b;
+10.     z = (r1 == r2);
+11. }
 \end{verbatim}

### 2024-10-03 paper/ootavacuous.tex
Improve the examples of invented loads
Explain the roles of r1a and r1b in 4.3 and explain why these results
lead us to prohibit inventing atomic loads.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
  8. } else {
- 9.     r1 = r1b;
+ 9.     r1 = r1a;
 10.     z = (r1 == r2);
@@
 \end{quote}
+Here \co{r1a} represents the original load from \co{x} while the \co{r1b}
+load is invented by the compiler.
+
 The idea is that \co{r1a}, \co{r1b}, and \co{r2} can each be only zero or one,
@@
 
-Therefore we should rule out transformations like these by insisting
+Because invented loads can cause errors and break dependencies,
+as we have just seen, we should insist that
 the compiler not invent (or duplicate) atomic loads.

### 2024-10-03 paper/ootavacuous.tex
General wordsmithing in multiple places
Various small changes to the text to make it smoother or more
explicit, or to remove unimportant information.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-A computer may execute a thread's instructions out of order.
+A CPU may execute a thread's instructions out of order.
 The architecture specifies the extent to which this may happen, and it
@@
 
-Invented atomic loads are problematic:
+Invented atomic loads are problematic.
+Consider this code:
 \begin{quote}
@@
 int r0 = x;
-int invented = x;
-int r1 = r0 * r0 + 2 * invented + 1;
+int rinvented = x;
+int r1 = r0 * r0 + 2 * rinvented + 1;
 \end{verbatim}
@@
 If the load into \co{r0} happened when the value of \co{x}
-was zero and the load into \co{invented} happened when the value of
-\co{x} was 10, then the value of \co{r1} will be 21, which is not
+was zero and the load into \co{rinvented} happened when the value of
+\co{x} was 10, then the value of \co{r1} would be 21, which is not
 a perfect square.
@@
 Inventing atomic loads can also result in OOTA-like behavior.
-For example, consider this code, in which the final values of \co{r1}
-and~\co{r2} are observable:
+For example, consider this code:
 \begin{quote}
@@
 
-Dependencies between machine instructions correspond to the flow of
+Dependencies between machine instructions are determined by the flow of
 information within a CPU.
@@
 No implementation in which semantic dependencies satisfy the Fundamental
-Property has a nontrivial OOTA cycle.
+Property can realize a nontrivial OOTA cycle.
 To see this, suppose that abstract execution $A$ having an OOTA cycle
@@
 Thus the implementation will automatically comply with full C++, even
-though it may be been designed only to comply with loose C++.
+though it may have been designed only to comply with loose C++.
 
@@
 relative to a specific compiler or implementation.
-This may strike some people as wishy-washy and avoiding the real
+This may strike some readers as wishy-washy and avoiding the real
 problem, in that a given piece of code should either contain or not

### 2024-10-03 paper/ootavacuous.tex
ootavacuous: Flatten Section 7.5 list
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Additional future work could:
-\begin{itemize}
-\item	Extend these results to some classes of interpreters and JITs
-	on one hand, and to link-time optimizations (LTO) on the other.
-\item	Consider various classes of global analysis, such as
-	demonstrating that a given expression will always evaluate
-	to a constant.
-\item	Examine the possibility of ``flattening'' optimizations that
-	combine multiple threads into one.
-\item	Explore cases in which semantic dependencies must be absolute,
-	not relative to any particular compiler.
-\item	Include the effects of input and output or other operations
-	which can vary from one execution to another.
-\item	Delineate more precisely the limits of permissible behavior
-	for quasi-volatile object accesses.
-\end{itemize}
+(1)~Extend these results to some classes of interpreters and JITs on
+one hand, and to link-time optimizations (LTO) on the other;
+(2)~Consider various classes of global analysis, such as
+demonstrating that a given expression will always evaluate to a constant;
+(3)~Examine the possibility of ``flattening'' optimizations that combine
+multiple threads into one;
+(4)~Explore cases in which semantic dependencies must be absolute,
+not relative to any particular compiler;
+(5)~Include the effects of input and output or other operations which
+can vary from one execution to another; and
+(6)~Delineate more precisely the limits of permissible behavior for
+quasi-volatile object accesses.
 

### 2024-10-03 paper/ootavacuous.tex
ootavacuous: Reorder and reword new sections 2.2.7 and 2.2.8
Which were originally 2.2.3 and 2.2.8.

Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\subsubsection{Non-Deterministic Semantic Dependencies}
-\label{sec:Non-Deterministic Semantic Dependencies}
-
-Compiler optimizers also exploit the nondeterministic aspects of an
-execution.
-Consider this example, with \co{i} initially zero:
-\begin{quote}
-\begin{verbatim}
-int foo(int a, int b)
-{
-   return a / b;
-}
-
-r1 = foo(++i, ++i);
-x = r1 * z;
-\end{verbatim}
-\end{quote}
-Because early C~implementers could not come to agreement, the standard
-does not specify the order of evaluation of function arguments, so
-the value calculated for \co{r1} might be zero ($1/2$ truncated) or two
-($2/1$).
-In the former case there is no semantic dependency from \co{z} to \co{x},
-but in the latter case there is.\footnote{
-	Thanks to Peter Sewell for pointing out this possibility.}
-
-(According to the current version of the standard, conflicting side effects
-in unsequenced subexpressions constitute undefined behavior,
-although there are proposals to make them defined in both
-C++~\cite{GabrielDosReis2016P0145r3}
-and C~\cite{AlexCeleste2023N3203}.
-Nevertheless, the example above is allowed because the order of evaluation
-of arguments to a function call is ``indeterminately sequenced''
-(7.6.1.3p7 \co{[expr.call]}) rather than unsequenced, a subtle distinction.)
-
 \subsubsection{Semantic Dependencies Affected by Cross-Thread Optimizations}
@@
 
-\subsubsection{Per-Architecture Dependent Semantic Dependencies}
-\label{sec:Per-Architecture Dependent Semantic Dependencies}
-
-The implementation-defined aspects of the standard permit architectures to
-differ as to whether an abstract execution contains a semantic dependency.
+\subsubsection{Semantic Dependencies and Free Choices}
+\label{sec:Semantic Dependencies and Free Choices}
+
+Consider the following example, with variable \co{i} initially zero:
+\begin{quote}
+\begin{verbatim}
+int foo(int a, int b)
+{
+   return a / b;
+}
+
+r1 = foo(++i, ++i);
+x = r1 * z;
+\end{verbatim}
+\end{quote}
+Because early C~implementers could not come to agreement, the standard
+does not specify the order of evaluation of function arguments, so
+the value calculated for \co{r1} might be zero ($1/2$ truncated) or two
+($2/1$).
+In the former case there is no semantic dependency from \co{z} to \co{x},
+but in the latter case there is.\footnote{
+	Thanks to Peter Sewell for pointing out this possibility.}
+
+(According to the current version of the standard, conflicting side effects
+in unsequenced subexpressions constitute undefined behavior,
+although there are proposals to make them defined in both
+C++~\cite{GabrielDosReis2016P0145r3}
+and C~\cite{AlexCeleste2023N3203}.
+Nevertheless, the example above is allowed because the order of evaluation
+of arguments to a function call is ``indeterminately sequenced''
+(7.6.1.3p7 \co{[expr.call]}) rather than unsequenced, a subtle distinction.)
+
+\subsubsection{Semantic Dependencies and Architectural Variations}
+\label{sec:Semantic Dependencies and Architectural Variations}
+
+In some cases, the standard permits architectures to differ as to whether
+an abstract execution contains a semantic dependency.
 For example, consider the following, where the type of \co{c} is \co{char}:
@@
 	or whether the \co{char} type is signed (see
-	Section~\ref{sec:Per-Architecture Dependent Semantic Dependencies}).
+	Section~\ref{sec:Semantic Dependencies and Architectural Variations}).
 \item	Some aspects of an execution are unspecified or nondeterministic,
@@
 	(see
-	Section~\ref{sec:Non-Deterministic Semantic Dependencies}).
+	Section~\ref{sec:Semantic Dependencies and Free Choices}).
 \item	Some actions are deemed to have undefined results; the standard

### 2024-10-03 paper/ootavacuous.tex
ootavacuous: Add mention of tooling to Section 7.1
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 relative to a specific compiler or implementation.
-This may strike some readers as wishy-washy and avoiding the real
-problem, in that a given piece of code should either contain or not
-contain a semantic dependency, independent of the implementation used to
-run it.
+This may strike some readers (especially those implementing tooling)
+as wishy-washy and avoiding the real problem, in that a given piece
+of code should either contain or not contain a semantic dependency,
+independent of the implementation used to run it.
 

### 2024-10-03 paper/ootavacuous.tex
Revert "ootavacuous: Add "full C++" qualifier for last of Section 5.3"
This reverts commit 14282f91a57be7b20f6e3f665a00c4532113401e.

@@
 also be omitted, meaning that the entire cycle has been optimized away.
-We conjecture that in this situation for full C++ there must be another
-abstract execution which has the same observable effects as $A$ and is
-also realized by $H$, but in which the OOTA cycle does not occur.
+We conjecture that in this situation there must be another abstract
+execution which has the same observable effects as $A$ and is also
+realized by $H$, but in which the OOTA cycle does not occur.
 Thus there would be no way to tell, merely by observing the effects of

### 2024-10-04 paper/ootavacuous.tex
ootavacuous: s/versus/vs./ in section headings
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: s/Summary and Conclusions/Summary and Conclusion/
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Generalize initial definition of "semantic dependency"
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 T0 and T1.
-(Roughly speaking, there is a semantic dependency from a given load
-to a given store when \emph{all other things being equal, a change in the
-value loaded can change the value stored or prevent the store from
-occurring at all.}
+(Roughly speaking, there is a semantic dependency from a given load to
+a given store when \emph{all other things being equal, a change in the
+value loaded can change the value stored, change the target address of
+the store, or prevent the store from occurring at all.}
 Here the dependencies are trivial because the values stored simply
@@
 
-\subsection{OOTA: rf versus rfe}
-\label{sec:OOTA: rf versus rfe}
+\subsection{OOTA: rf vs.\ rfe}
+\label{sec:OOTA: rf vs. rfe}
 
@@
 
-\subsection{Relative versus Absolute Dependency}
-\label{sec:Relative versus Absolute Dependency}
+\subsection{Relative vs.\ Absolute Dependency}
+\label{sec:Relative vs. Absolute Dependency}
 
@@
 
-\section{Summary and Conclusions}
-\label{sec:Summary and Conclusions}
+\section{Summary and Conclusion}
+\label{sec:Summary and Conclusion}
 

### 2024-10-06 paper/ootavacuous.tex
ootavacuous: Section 7.4 optimization non-recommendation
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Although it is unexpected, we cannot say that this conclusion is
-definitely wrong, because our intuitive notions of semantic
-dependency are not clear in cases where multiple loads of the same
-variable are present.
+Although it is unexpected, and while we don't recommend this sort of code
+transformation, we cannot say that this conclusion is definitely wrong,
+because our intuitive notions of semantic dependency are not clear in
+cases where multiple loads of the same variable are present.
 % Can we instead say that transforming the code might require that the

### 2024-10-06 paper/ootavacuous.tex
ootavacuous: Section 7.4 suggest less-surprising optimization
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-However, a per-thread analysis compiler can translate this into the
-machine-code equivalent of:
+Although one would expect an optimizing compiler to discard the unused
+loads into \co{r1} and \co{r3}, one could imagine a perverse compiler
+instead using per-thread analysis to translate this into the machine-code
+equivalent of:
 \begin{quote}

### 2024-10-06 paper/ootavacuous.tex
ootavacuous: Further reword Section 7.4 third paragraph
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-10-06 paper/ootavacuous.tex
@@
 
-However, a per-thread analysis compiler can translate this into the
-machine-code equivalent of:
+Although many optimizing compilers would omit the loads into \co{r1}
+and \co{r3}, or merge them with \co{r2} if they are used later, one
+could imagine a perverse compiler instead using per-thread analysis to
+translate this into the machine-code equivalent of:
 \begin{quote}

### 2024-10-09 paper/ootavacuous.tex
ootavacuous: Wordsmith Section 2.2 intro
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Add reference to Section 2.2.8 to Section 4.1
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Here we consider some of the complexities of semantic dependency.
+Here we consider some of the complexities inherent to semantic
+dependencies.
 
@@
 An example is GCC's \co{-funsigned-char} command-line argument,
-which causes it to treat variables of type \co{char} as unsigned.
+which causes it to treat variables of type \co{char} as unsigned
+(see Section~\ref{sec:Semantic Dependencies and Architectural Variations}).
 Many more examples of user control over language semantics

### 2024-10-09 paper/ootavacuous.tex
ootavacuous: Remove lingering "Tagline" notes
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Connect witnesses to perturbations
Also save a line while in the area.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\emph{Tagline: ``How to think about semantic dependencies.'' }
-
 This section examines dependencies at the level of machine instructions
@@
 
-\emph{ Tagline: ``How to think about OOTA.'' }
-
 Semantic dependency is a notoriously difficult concept to define
@@
 
+These witnesses generalize the notion of
+perturbations~\cite{PaulEMcKenney2016OOTA}.
+
 \subsection{For Compilers Using Global Analysis}
@@
 Thus the implementation will automatically comply with full C++, even
-though it may have been designed only to comply with loose C++.
+if it was designed only to comply with loose C++.
 

### 2024-10-09 paper/ootavacuous.tex
ootavacuous: Add atomic-optimization caution to summary
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-\section{Summary and Conclusion}
-\label{sec:Summary and Conclusion}
+\section{Summary and Conclusions}
+\label{sec:Summary and Conclusions}
 
@@
 
-The conclusion to be drawn from this work is that in this context,
+One conclusion to be drawn from this work is that in this context,
 avoiding OOTA cycles requires no changes to the standard,
 to current implementions, or to user practices for portable code.
+This work also demonstrates the need for caution when optimizing
+atomic operations.
 

### 2024-10-10 paper/ootavacuous.tex
ootavacuous: Revert point about generalizing perturbations
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Section 2.2.8 s/abstract execution/execution/
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 In some cases, the standard permits architectures to differ as to whether
-an abstract execution contains a semantic dependency.
+an execution contains a semantic dependency.
 For example, consider the following, where the type of \co{c} is \co{char}:
@@
 
-These witnesses generalize the notion of
-perturbations~\cite{PaulEMcKenney2016OOTA}.
-
 \subsection{For Compilers Using Global Analysis}

### 2024-10-11 paper/ootavacuous.tex
Fix wording about breaking dependencies
The example of breaking dependencies in 4.3 should be worded better.

Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 code~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.\footnote{
-	Throughout this paper, variables whose names are ``\co{r}''
-	followed by a number denote private per-thread variables
-	(``registers''),
+	Throughout this paper, variables whose names begin with ``\co{r}''
+	denote private per-thread variables (``registers''),
 	and other names denote shared variables.
@@
 
-Inventing atomic loads can also result in OOTA-like behavior.
+Inventing atomic loads can also destroy dependencies.
 For example, consider this code:
@@
 \end{quote}
-It is clear that the store to \co{z} semantically depends on the load
+It seems clear that the store to \co{z} semantically depends on the load
 from \co{y}, because the value of \co{z} will change whenever \co{y}
@@
 quasi-volatile object accesses.
-
-\section{Summary and Conclusions}
-\label{sec:Summary and Conclusions}
+This last item may well shed additional light on the dangers of
+optimizations involving non-volatile atomic operations beyond the cautions
+outlined in
+Section~\ref{sec:Inventing Atomic Loads Can Cause Errors and
+Destroy Semantic Dependencies}.
+
+\section{Summary and Conclusion}
+\label{sec:Summary and Conclusion}
 
@@
 to current implementions, or to user practices for portable code.
-This work also demonstrates the need for caution when optimizing
-atomic operations.
+% This work also demonstrates the need for caution when optimizing
+% atomic operations.
 

### 2024-10-11 paper/ootavacuous.tex
ootavacuous: Allow for non-numeric registers
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-10-11 paper/ootavacuous.tex
@@
 
-Inventing atomic loads can also destroy dependencies.
+Inventing atomic loads can also result in OOTA-like behavior.
 For example, consider this code:
@@
 \end{quote}
-It seems clear that the store to \co{z} semantically depends on the load
+It is clear that the store to \co{z} semantically depends on the load
 from \co{y}, because the value of \co{z} will change whenever \co{y}

### 2024-10-11 paper/ootavacuous.tex
ootavacuous: More rework of first footnote
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2024-10-11 paper/ootavacuous.tex
@@
 code~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.\footnote{
-	Throughout this paper, variables whose names begin with ``\co{r}''
-	denote private per-thread variables (``registers''),
-	and other names denote shared variables.
+	Throughout this paper, we use names beginning with ``\co{r}''
+	for private per-thread variables (``registers''); other names
+	denote shared variables.
 	All shared-variable accesses are relaxed atomic.}
@@
 \item	In benign environments the map must preserve the rf relation.
-	That is, if volatile load $R$ in $A$ reads from store $W$ then
+	That is, if volatile load $R$ in $A$ reads from store $W$
 	then the instruction it maps to in $H$ must read from the

### 2024-10-11 paper/ootavacuous.tex
ootavacuous: Wordsmith last sentence of Section 6 lead-in
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Yet more rework of Section 4.1's first sentence
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 The exact definition of a computer language is a subject of some debate,
-with standards, implementations, and users all having some degree of
+with standards, implementations, and users all having their own
 influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
@@
 more tractable.
-This allows us to demonstrate that OOTA cycles cannot form
-in the C++ implementations addressed by this paper.
+Using this approach, we demonstrate that OOTA cycles cannot form in the
+C++ implementations addressed by this paper.
 

### 2024-10-12 paper/ootavacuous.tex
ootavacuous: Section 8 s/One/The/
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-One conclusion to be drawn from this work is that in this context,
+The conclusion to be drawn from this work is that in this context,
 avoiding OOTA cycles requires no changes to the standard,

### 2024-10-19 paper/ootavacuous.tex
ootavacuous: More fairly minor edits
Some more edits to improve the flow of the text.  Only one of them, in
6.1, is crucial; it fixes a typo (a word that was left out).

Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Because nothing else in the C++ memory model rules out such OOTA cycles
+Because nothing in the C++ memory model rules out such OOTA cycles
 (as indicated by the ``sometimes satisfied'' in the figure), the C++
@@
 
-This paper uses \emph{full C++} to denote the standard including this
+We will use \emph{full C++} to denote the standard including this
 prohibition of OOTA and \emph{loose C++} to denote a hypothetical standard
-that excludes this prohibition but is otherwise identical to full C++.
+that excludes the prohibition but is otherwise identical to full C++.
 Unqualified \emph{C++} means full C++.
@@
 running on implementations in
-which \co{char} is a signed type,
+which \co{char} is architecturally a signed type,
 but not those for which it is unsigned.
@@
 rules of the abstract machine'' (4.1.2p6.1 \co{[intro.abstract]}); and
-(3)~(A condition on the timing and interleaving of input and output,
-which does not matter for our purposes).
+(3)~[A condition on the timing and interleaving of input and output,
+which does not matter for our purposes].
 We will say that the abstract execution is \emph{realized by} the
@@
 input.\footnote{
-	One exception is a branch that conditionally jumps to the
-	next instruction.
+	A branch that conditionally jumps to the
+	next instruction would be an exception.
 	We ignore this possibility by treating such branches as no-ops.}
@@
 abstract execution realized by a given hardware execution.
-(While it possible to argue about semantic dependencies in abstract
+(While it is possible to argue about semantic dependencies in abstract
 executions that have no hardware realizations, doing so seems pointless.)
@@
 Our third requirement for being a witness is that each load in the
-common period either must obtain the same value in $H$ and $G$, or
-must itself be one of the $R'_i$ loads, or must be ordered in $H$
+common period must either obtain the same value in $H$ and $G$, or
+itself be one of the $R'_i$ loads, or be ordered in $H$
 after one of the $R'_i$ loads (this is the remaining part of our
@@
 machine code with simple assignments ``\co{r1 = 42}'' and
-``\co{r2 = 42}''.
+``\co{r2 = 42}'', as shown in
+Section~\ref{sec:Global Optimization Can Destroy Dependencies}.
 This would be a valid transformation, and the resulting behavior would
@@
 Additional future work could:
-(1)~Extend these results to some classes of interpreters and JITs on
-one hand, and to link-time optimizations (LTO) on the other;
+(1)~Extend these results to some classes of interpreters and just-in-time
+compilers (JITs) on
+one hand, and to link-time optimization (LTO) on the other;
 (2)~Consider various classes of global analysis, such as

### 2025-01-15 paper/rebuttal.txt
Initial draft of rebuttal
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

First, we thank all the reviewers for their time, their careful reviews,
and their thoughtful feedback.

Several reviewers suggested citations that we are happy to add.

Other reviewers asked how we handle various litmus tests.  Our submission
describes how we rely on a compiler to translate the litmus test to object
code, then relate dependencies in that object code to the corresponding
source code.  Section 1.2 ("Prior Work") notes that we are not trying
to identify when OOTA cycles can form, but rather that OOTA cycles
cannot form.

Several reviewers seemed uncomfortable with informal proofs.  However,
ASPLOS has published papers with informal proofs, so we suspect that the
real discomfort was augmenting the C++ abstract machine with hardware
properties.  20 years of equivocal progress is more than sufficient to
justify trying a new approach.  We are not advocating the old approaches
be abandoned.  In fact, should someone succeed with those approaches,
we will endeavor to be the first to offer congratulations.

Others were uncomfortable with our reliance on object code, which
varies across systems.  However, these systems are underpinned by laws
of physics, namely causality, the finite speed of light, and the atomic
nature of matter.  Our definition of OOTA cycles as cycles in (rfe U
sdep) permits us to examine rfe and sdep separately.  These laws of
physics guarantee that each rfe link goes forward in time.  Because an
OOTA cycle must return to the starting time, there must be at least
one sdep link that goes backwards in time.  Correct compilers guarantee
that sdep links correspond to sequences of machine instructions, which
cannot execute backwards in time, even given speculative execution.
Thus, the OOTA cycle cannot form.

Please note that correct implementations are not required to break
non-semantic dependencies.  An unbroken non-semantic dependency will
correspond to a sequence of machine instructions that cannot go backwards
in time, and thus cannot form an OOTA cycle.

Reviewer 72F asks about three litmus tests in Section 5 of the
submission's citation [5], as do a couple of other reviewers:

o	The first example has a dependency only if the implementation
	does not hoist the i++ out of the “if” statement, and this
	is the implementation’s choice, similar to the example in
	Section 2.2.5 of the submission.

o	The second example is handled by examining f() and the functions
	that it invokes, along with the implementation’s optimization
	choices.  Note that at one extreme, f() might ignore its
	parameter and at the other extreme it might simply return
	its parameter.  Note further that f() might be a pointer to a
	function, and that function might come from a dynamically linked
	library that has not yet been written at the time the call to
	f() is compiled.  Although the submission does not contain
	this particular example, all of these issues are addressed by
	the submission’s Section 3, which discusses executions (as
	opposed to source-code syntactic dependencies).

o	An instructive elaboration of the third example is discussed in
	Section 2.2.1 of the submission.

Reviewer 72F also asks if our approach implicitly enforces the
load-to-store ordering advocated by Boehm and Demsky.  The answer is
"no".  First, although instructions commit in some order, they do not
necessarily commit in program order, which on some systems allows prior
loads to be reordered with later stores.  Even on hardware that does
commit instructions in program order, there is no guarantee that this
commit order will constrain the visibility of the corresponding memory
references, for example, due to the effects of hardware store buffers.

Reviewer 72F also asks (in two questions) if freezing of nondeterministic
choices into machine code is realistic.  The answer is "yes", but on the
per-execution basis used in our submission.  For example, specialization
optimizations might cause a given function to be compiled differently
for different combinations of arguments.  But any given execution would
correspond to a particular combination of arguments, and that execution
would be frozen into machine code.  Please note that programs that
modify their own code are outside of the standard, and thus out of our
submission's scope.

We especially thank Reviewer 72F for noting (correctly!) that Section 7.4
("Merging Quasi-Volatile Loads") poses a challenge, as this helped refine
our thinking. The refinement is to note that this challenge is not to our
technique, but rather to those implementing optimizations.  Although we
believe (one of us strongly, the rest less so) that the "optimization"
described in this section is in fact a pessimization that also degrades
ease of use, the question of whether or not this optmization is legitimate
will be decided by community processes, not by fiat and not by pure
mathematics.  Nevertheless, should an abstract-machine mathematical
treatment of OOTA appear, we would expect that this mathematics would
have a prominent seat at the community-process table.  Either way,
if this optimization is rejected the litmus test in Section 7.4 has a
semantic dependency, otherwise it does not.  And correct implementations
will reflect that decision, though again, optimizations are permitted
to refrain from breaking non-semantic dependencies.

Reviewer 72F suggested that rf links must impose restrictions.
Please note our Section 2.1 ("OOTA: rf vs. rfe") where we show that
those restrictions can be extremely loose, hence our use of rfe.
But note also that C++ modification order provides very weak guarantees,
in fact, a later-in-time store might precede an earlier-in-time store
in modification order, even on x86.  Memory models are at best loosely
connected to the passage of time.

Reviwer 72F also questions how the original C++ code can be irrelevant.
The C++ code is relevant, but only at the source-code level and to the
connections between the source-code and object-code levels.  However,
for analysis confined to the object-code level, the origial C++ code
can indeed be irrelevant.

Reviewer 72F asks what we mean by "these models are atemporal".  We
mean that typical memory models do not make use of the passage of
physical time.  The provide partial orderings, but as noted earlier,
the modification order can in fact go backwards in time.

Responses to individual reviewer feedback items is in the accompanying
PDF.

### 2025-01-15 paper/rebuttal.txt
rebuttal.txt: Self-review reducing to <800 words
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
-First, we thank all the reviewers for their time, their careful reviews,
-and their thoughtful feedback.
+First, we thank all the reviewers for their careful reviews and thoughtful
+feedback.
 
-Several reviewers suggested citations that we are happy to add.
+We are happy to add the suggested citations.
 
-Other reviewers asked how we handle various litmus tests.  Our submission
-describes how we rely on a compiler to translate the litmus test to object
-code, then relate dependencies in that object code to the corresponding
-source code.  Section 1.2 ("Prior Work") notes that we are not trying
-to identify when OOTA cycles can form, but rather that OOTA cycles
-cannot form.
+There was discomfort with informal proofs.  However, ASPLOS has published
+such papers, so we suspect that the real discomfort was augmenting the
+C++ abstract machine with hardware properties.  20 years of equivocal
+progress justifies trying a new approach.  We are not advocating the old
+approaches be abandoned.  In fact, we will endeavor to be the first to
+offer congratulations should someone succeed.
 
-Several reviewers seemed uncomfortable with informal proofs.  However,
-ASPLOS has published papers with informal proofs, so we suspect that the
-real discomfort was augmenting the C++ abstract machine with hardware
-properties.  20 years of equivocal progress is more than sufficient to
-justify trying a new approach.  We are not advocating the old approaches
-be abandoned.  In fact, should someone succeed with those approaches,
-we will endeavor to be the first to offer congratulations.
+Various litmus tests were put forward.  Our submission describes how
+we rely on a compiler to translate the litmus test to object code, then
+relate dependencies in that object code to the corresponding source code.
+Section 1.2 ("Prior Work") notes that we are not trying to identify when
+OOTA cycles can form, but rather show that OOTA cycles cannot form.
 
-Others were uncomfortable with our reliance on object code, which
-varies across systems.  However, these systems are underpinned by laws
-of physics, namely causality, the finite speed of light, and the atomic
-nature of matter.  Our definition of OOTA cycles as cycles in (rfe U
-sdep) permits us to examine rfe and sdep separately.  These laws of
-physics guarantee that each rfe link goes forward in time.  Because an
-OOTA cycle must return to the starting time, there must be at least
-one sdep link that goes backwards in time.  Correct compilers guarantee
-that sdep links correspond to sequences of machine instructions, which
-cannot execute backwards in time, even given speculative execution.
-Thus, the OOTA cycle cannot form.
+There was also discomfort with use of object code, which varies across
+systems.  However, these systems are underpinned by laws of physics,
+namely causality, the finite speed of light, and the atomic nature
+of matter.  Our definition of OOTA cycles as cycles in (rfe U sdep)
+permits us to examine rfe and sdep separately, noting that each rfe link
+goes forward in time.  Because an OOTA cycle must return to the starting
+time, there must be at least one sdep link that goes backwards in time.
+Correct compilers guarantee that sdep links correspond to sequences of
+machine instructions, which cannot execute backwards in time, even given
+speculative execution.  Thus, the OOTA cycle cannot form.
 
 Please note that correct implementations are not required to break
-non-semantic dependencies.  An unbroken non-semantic dependency will
-correspond to a sequence of machine instructions that cannot go backwards
-in time, and thus cannot form an OOTA cycle.
+non-semantic dependencies.  Regardless, a non-semantic dependency cannot
+form an OOTA cycle.
 
@@
 o	The first example has a dependency only if the implementation
-	does not hoist the i++ out of the “if” statement, and this
-	is the implementation’s choice, similar to the example in
-	Section 2.2.5 of the submission.
+	does not hoist the i++ out of the “if” statement, and this is
+	the implementation’s choice, similar to the example in Section
+	2.2.5 of the submission.  See Sections 2.2.7 and 2.2.8 for more
+	aggressive examples.
 
-o	The second example is handled by examining f() and the functions
-	that it invokes, along with the implementation’s optimization
-	choices.  Note that at one extreme, f() might ignore its
-	parameter and at the other extreme it might simply return
-	its parameter.  Note further that f() might be a pointer to a
-	function, and that function might come from a dynamically linked
-	library that has not yet been written at the time the call to
-	f() is compiled.  Although the submission does not contain
-	this particular example, all of these issues are addressed by
-	the submission’s Section 3, which discusses executions (as
-	opposed to source-code syntactic dependencies).
+o	The second example is handled by examining f() and the
+	functions that it invokes, along with the implementation’s
+	optimization choices.  This example is yet another motivation
+	for the submission’s Section 3, which discusses executions as
+	opposed to source-code syntactic dependencies.
 
@@
 
-Reviewer 72F also asks if our approach implicitly enforces the
-load-to-store ordering advocated by Boehm and Demsky.  The answer is
-"no".  First, although instructions commit in some order, they do not
-necessarily commit in program order, which on some systems allows prior
-loads to be reordered with later stores.  Even on hardware that does
-commit instructions in program order, there is no guarantee that this
-commit order will constrain the visibility of the corresponding memory
-references, for example, due to the effects of hardware store buffers.
+Reviewer 72F also asks if we implicitly enforce the load-to-store
+ordering advocated by Boehm and Demsky.  We do not.  Instructions do
+not necessarily commit in program order.  Furthermore, commit order
+might not constrain the order of the corresponding memory references.
 
@@
 correspond to a particular combination of arguments, and that execution
-would be frozen into machine code.  Please note that programs that
-modify their own code are outside of the standard, and thus out of our
-submission's scope.
+would be frozen into machine code.  Please note that self-modifying
+programs are outside of the standard, and thus out of our scope.
 
-We especially thank Reviewer 72F for noting (correctly!) that Section 7.4
-("Merging Quasi-Volatile Loads") poses a challenge, as this helped refine
-our thinking. The refinement is to note that this challenge is not to our
-technique, but rather to those implementing optimizations.  Although we
-believe (one of us strongly, the rest less so) that the "optimization"
-described in this section is in fact a pessimization that also degrades
-ease of use, the question of whether or not this optmization is legitimate
-will be decided by community processes, not by fiat and not by pure
-mathematics.  Nevertheless, should an abstract-machine mathematical
-treatment of OOTA appear, we would expect that this mathematics would
-have a prominent seat at the community-process table.  Either way,
-if this optimization is rejected the litmus test in Section 7.4 has a
-semantic dependency, otherwise it does not.  And correct implementations
-will reflect that decision, though again, optimizations are permitted
-to refrain from breaking non-semantic dependencies.
+We thank Reviewer 72F for noting that Section 7.4 ("Merging Quasi-Volatile
+Loads") poses a challenge.  However, please note that this challenge
+is not to our technique, due to our focus on object-code executions.
+Community processes will decide whether or not this section's optimization
+is legitimate, and these processes' outcomes will be implemented in
+compilers, thus being automatically handled by our technique.  The
+challenge is instead to OOTA analysis techniques not accounting for these
+community processes.  Nevertheless, any abstract-machine mathematical
+OOTA solution should have a prominent seat at the community-process table.
 
@@
 in fact, a later-in-time store might precede an earlier-in-time store
-in modification order, even on x86.  Memory models are at best loosely
-connected to the passage of time.
+in modification order, even on x86.
 
@@
 
-Reviewer 72F asks what we mean by "these models are atemporal".  We
-mean that typical memory models do not make use of the passage of
+Reviewer 72F asks what we mean by "these models are atemporal".
+We mean that typical memory models do not make use of the passage of
 physical time.  The provide partial orderings, but as noted earlier,
-the modification order can in fact go backwards in time.
+some orderings can go backwards in time.
 

### 2025-01-15 paper/rebuttal.txt
rebuttal.txt:  Apply Alan Stern feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+Responses to individual reviewer feedback items may be found in the
+accompanying PDF.
+
 We are happy to add the suggested citations.
 
-There was discomfort with informal proofs.  However, ASPLOS has published
-such papers, so we suspect that the real discomfort was augmenting the
-C++ abstract machine with hardware properties.  20 years of equivocal
-progress justifies trying a new approach.  We are not advocating the old
-approaches be abandoned.  In fact, we will endeavor to be the first to
-offer congratulations should someone succeed.
+There was discomfort with informal proofs.  However, our proofs are
+straightforward and ASPLOS has published such proofs.  We therefore
+suspect that the real discomfort was augmenting the C++ abstract machine
+with hardware properties.  20 years of equivocal progress justifies trying
+a new approach.  We are not advocating the old approaches be abandoned.
+In fact, we will endeavor to be the first to offer congratulations should
+someone succeed.
 
@@
 
-Reviewer 72F asks about three litmus tests in Section 5 of the
-submission's citation [5], as do a couple of other reviewers:
+Responses specifically requested by Reviewer 72F for the rebuttal:
+
+How do we handle the three litmus tests in Section 5 of citation [5]?
 
@@
 
-Reviewer 72F also asks if we implicitly enforce the load-to-store
-ordering advocated by Boehm and Demsky.  We do not.  Instructions do
-not necessarily commit in program order.  Furthermore, commit order
-might not constrain the order of the corresponding memory references.
+Do we implicitly enforce the load-to-store ordering advocated by Boehm
+and Demsky?  (Two questions.)  We do not.  Instructions do not necessarily
+commit in program order.  Furthermore, commit order might not constrain
+the order of the corresponding memory references.
 
-Reviewer 72F also asks (in two questions) if freezing of nondeterministic
-choices into machine code is realistic.  The answer is "yes", but on the
-per-execution basis used in our submission.  For example, specialization
-optimizations might cause a given function to be compiled differently
-for different combinations of arguments.  But any given execution would
-correspond to a particular combination of arguments, and that execution
-would be frozen into machine code.  Please note that self-modifying
-programs are outside of the standard, and thus out of our scope.
+Is freezing of nondeterministic choices into machine code realistic?
+Yes, but on the per-execution basis used in our submission.  For example,
+specialization optimizations might cause a given function to be compiled
+differently for different combinations of arguments.  But any given
+execution would correspond to a particular combination of arguments,
+and that execution would be frozen into machine code.
 
@@
 
-Reviewer 72F suggested that rf links must impose restrictions.
-Please note our Section 2.1 ("OOTA: rf vs. rfe") where we show that
-those restrictions can be extremely loose, hence our use of rfe.
-But note also that C++ modification order provides very weak guarantees,
-in fact, a later-in-time store might precede an earlier-in-time store
-in modification order, even on x86.
+Don't rf links must impose restrictions?  Yes, they do in the C++ memory
+mode, but only when they involve release-stores and acquire-loads or
+consume-loads.  These restrictions do not apply to OOTA cycles, which
+involve only relaxed accesses.  Note further that C++ modification
+order provides very weak guarantees, in fact, a later-in-time store
+might precede an earlier-in-time store in modification order, even on x86.
 
-Reviwer 72F also questions how the original C++ code can be irrelevant.
-The C++ code is relevant, but only at the source-code level and to the
-connections between the source-code and object-code levels.  However,
-for analysis confined to the object-code level, the origial C++ code
-can indeed be irrelevant.
+How the original C++ code can be irrelevant?  The C++ code is relevant,
+but only at the source-code level and to the connections between the
+source-code and object-code levels.  However, for analysis confined to
+the object-code level, the original C++ code can indeed be irrelevant.
+For example, if a program in C++ and another in Rust generated the same
+machine code, the behavior of that machine code would be independent of
+the source language.
 
-Reviewer 72F asks what we mean by "these models are atemporal".
-We mean that typical memory models do not make use of the passage of
-physical time.  The provide partial orderings, but as noted earlier,
-some orderings can go backwards in time.
-
-Responses to individual reviewer feedback items is in the accompanying
-PDF.
+What is meant by "these models are atemporal"?  We mean that typical
+memory models do not make use of the passage of physical time.
+The provide partial orderings, but as noted earlier, some orderings can
+go backwards in time.

### 2025-01-16 paper/rebuttal-mods.txt
rebuttal-mods.txt: Track possible upcoming changes
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

The "o" items assume a venue (such as a tech report) that have relaxed
page limits.

O	Detailed feedback from Review #72C.

O	s/We would be happy to change "explicitly prohibits"/"explicitly recommends against"/
	This is prior to the quote from the C++ standard.

O	s/Workers/Researcher/ in Section 1.2.

O	Add citations to Section 1.2 ("Prior Work").

O	Be less condescending towards Mark Batty.  Maybe
	s/is persisting with/is working on/.

o	Add a Section 1.3 ("Our Approach"?) that summarizes the
	rfe/sdep/non-sdep-dep approach and emphasizes proving
	OOTA cannot happen as opposed to an exact delineation
	between sdep and non-sdep.

o	Add citations to later sections that discuss similar litmus tests.

o	Consider adding a 2.2.x section that discusses the difficulties
	inherent in computing sdep for function calls, especially when
	the function will be part of a dynamic library that has not yet
	been written.

O	In section 6.1, define "witness" informally as "another execution
	in which the store acts differently".

O	On page 9, "While it possible".  [ Since fixed by Alan. ]

O	In Section 7.4 ("Merging Quasi-Volatile Loads"), change focus
	from intuitive notions to a community-process approach for
	determining what optimizations are legitimate.  Math cannot
	handle this.  We avoid the problem by looking at the compiler
	implementation, which is downstream of the community process.

	Also in Section 7.4 change "our intuitive notions" to "one's
	intuitive notions".

o	Possible future work: Relax quasi-volatile semantics to permit
	additional uses of the cmov instruction.

### 2025-01-21 paper/rebuttal-mods.txt
rebuttal-mods.txt: More updates
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 O	Detailed feedback from Review #72C.
+
+o	Early in the paper, consider explicitly stating that our approach
+	enables us to avoid the need to precisely delineate between sdep
+	and non-sdep dependencies.
 
@@
 
+o	Add diagrams illustrating what the various components of sections
+	6.1 and 6.3 refer to.
+
 O	On page 9, "While it possible".  [ Since fixed by Alan. ]
@@
 
+o	Add a section before Section 7 ("Issues and Refinements") that
+	shows how our approach handles a few well-known examples of OOTA.
+	This needs to be in greater detail than in the "Causality Test
+	Cases" discussions in P3064R2 ("How to Avoid OOTA Without Really
+	Trying").
+
 o	Possible future work: Relax quasi-volatile semantics to permit
-	additional uses of the cmov instruction.
+	load speculation, for example, use of the cmov instruction in
+	cases where the source code conditionally executes the atomic
+	load.

### 2025-01-22 paper/rebuttal-mods.txt
rebuttal-mods.txt: Alan Stern January 21 feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-o	Early in the paper, consider explicitly stating that our approach
-	enables us to avoid the need to precisely delineate between sdep
-	and non-sdep dependencies.
+o	Early in the paper, consider explicitly stating that after 20
+	years it appears that the problem of characterizing semantic
+	dependency is too difficult to solve in the context provided by
+	the C++ standard (i.e., the abstract machine) and therefore we
+	expand the context by considering actual machine-level executions.
 
@@
 O	Add citations to Section 1.2 ("Prior Work").
+
+	Add a forward reference to Section 4.2 ("Global Optimization
+	Can Destroy Dependencies") and the fact that this defeats the
+	approach of ordering relaxed loads before relaxed stores.
 
@@
 
+o	In Section 4.2 ("Global Optimization Can Destroy Dependencies"),
+	note that this optimization breaks the approach of ordering
+	relaxed loads before relaxed stores.
+
 O	In section 6.1, define "witness" informally as "another execution
-	in which the store acts differently".
+	in which the load(s) obtain a different value but all else is
+	the same as far as possible, and yet the store acts differently".
 

### 2025-02-28 paper/rebuttal-mods.txt
rebuttal-mods.txt: Distinguish between syntactic and semantic dependencies
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+o	In section 2.2 where we start to talk about semantic dependencies
+	in detail, we could mention syntactic dependencies too.  Explain
+	what they are and how they differ from semantic dependencies.
+	Then mention that some work in the literature (such as parts of
+	[5]) is not always careful to distinguish clearly between the
+	two forms.  Maybe have an entire subsection on this topic.
+
 o	In Section 4.2 ("Global Optimization Can Destroy Dependencies"),

### 2025-02-28 paper/rebuttal-mods.txt
Bibliography update and handling of trivial rebuttal items
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-O	s/We would be happy to change "explicitly prohibits"/"explicitly recommends against"/
+x	s/We would be happy to change "explicitly prohibits"/"explicitly recommends against"/
 	This is prior to the quote from the C++ standard.
 
-O	s/Workers/Researcher/ in Section 1.2.
+x	s/Workers/Researcher/ in Section 1.2.
 
@@
 
-	Add a forward reference to Section 4.2 ("Global Optimization
+x	Add a forward reference to Section 4.2 ("Global Optimization
 	Can Destroy Dependencies") and the fact that this defeats the
@@
 
-O	Be less condescending towards Mark Batty.  Maybe
+x	Be less condescending towards Mark Batty.  Maybe
 	s/is persisting with/is working on/.
@@
 
-o	Add citations to later sections that discuss similar litmus tests.
+o	Add citations to later sections that discuss litmus tests
+	that are similar to those in the literature.
 
@@
 
-o	In Section 4.2 ("Global Optimization Can Destroy Dependencies"),
+x	In Section 4.2 ("Global Optimization Can Destroy Dependencies"),
 	note that this optimization breaks the approach of ordering
@@
 
-O	In section 6.1, define "witness" informally as "another execution
+x	In section 6.1, define "witness" informally as "another execution
 	in which the load(s) obtain a different value but all else is
@@
 
-O	On page 9, "While it possible".  [ Since fixed by Alan. ]
+x	On page 9, "While it possible".  [ Since fixed by Alan. ]
 

### 2025-02-28 paper/ootavacuous.tex
ootavacuous: De-anonymize
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Bibliography update and handling of trivial rebuttal items
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-% Blind review process.
-% \author{
-% Alan Stern\\{\small \email{stern@rowland.harvard.edu}} \and
-% Paul E.~McKenney\\{\small \email{paulmck@kernel.org}} \and
-% Michael Wong\\{\small \email{fraggamuffin@gmail.com}} \and
-% Maged Michael\\{\small \email{maged.michael@gmail.com}}
-% }
+\author{
+Alan Stern (stern@rowland.harvard.edu) \and
+Paul E.~McKenney (paulmck@kernel.org) \and
+Michael Wong (fraggamuffin@gmail.com)
+}
 
@@
 (as indicated by the ``sometimes satisfied'' in the figure), the C++
-standard explicitly prohibits them in 33.5.4p8
+standard explicitly recommends against them in 33.5.4p8
 (\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
@@
 
-All OOTA workers owe a debt of gratitude to the foundational work led
+All OOTA researchers owe a debt of gratitude to the foundational work led
 by William Pugh, represented by the infamous Java-language
@@
 % youths, but it is now the year 2024.
+Furthermore, as discussed in
+Section~\ref{sec:Global Optimization Can Destroy Dependencies},
+cross-thread optimizations can defeat this ordering constraint.
 
@@
 but none of these have been accepted by C++ implementers.
-As of 2024, Mark Batty is persisting with modular relaxed dependencies,
+As of 2024, Mark Batty is working on modular relaxed dependencies,
 which may have the added ability of checking compiler
@@
 even though the resulting executable file would produce an unintuitive
-OOTA outcome every time it runs!
+OOTA outcome every time it runs, even if relaxed loads are ordered
+before relaxed stores!
 
@@
 
+We informally define a \emph{witness} to be another execution in which
+at least one of the loads obtains a different value but all else is the
+same as far as possible, and yet the store acts differently.
+
 To be a proper witness, $B$ must be an execution of some program $Q$,

### 2025-03-04 paper/ootavacuous.tex
Apply detailed feedback from Reviewer #72C
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	We accommodate nonvolatile atomics by defining ``quasi-volatile''
-	behavior that enables implementations doing per-thread analysis
-	and optimization not only to avoid OOTA, but also to avoid 
-	incorrectly evaluating arithmetic expressions involving atomic
-	accesses.
+	behavior which enables existing implementations that do per-thread
+	analysis and optimization not only to avoid OOTA, but also to
+	avoid incorrectly evaluating arithmetic expressions involving
+	atomic accesses.
 
@@
 C++ implementations.
+(The as-if rule permits those optimizations where the optimized code's
+observable behavior is the same as if it were unoptimized.)
 
@@
 OOTA, using a perturbation method based on the insight that OOTA cycles
-are fixed-point computations.
+are fixed-point computations
+In other words, an OOTA cycle must produce the same value that it
+consumes.
 
@@
 
-We have seen several examples showing that semantic dependencies are more
-complicated than might first appear, varying according to the execution
-and even the implementation.
+We have seen several examples showing that semantic dependencies can
+be more complicated than might first appear, varying according to the
+execution and even the implementation.
 This raises several questions, of which the first is:
@@
 This section looks more carefully at executions from the viewpoints of
-the abstract machine and the computer hardware, and then reconciles
+an abstract machine and the computer hardware, and then reconciles
 these two viewpoints.
@@
 abstract execution of the source code given the same input (in the
-absence of any abstract executions containing undefined behavior).\footnote{
-	This requirement is the standard's ``as-if'' rule.}
+absence of any abstract executions containing undefined behavior)
+This requirement is the standard's ``as-if'' rule.
 This means:
@@
 which does not matter for our purposes].
-We will say that the abstract execution is \emph{realized by} the
-hardware execution.
+We will say that the hardware execution realizes the abstract execution,
+or, equivalently, that the abstract execution is \emph{realized by}
+the hardware execution.
 
@@
 invent, omit, merge, or reorder these accesses, as we will discuss in
-Section~\ref{sec:Volatile and Quasi Volatile Accesses} below.
+Sections~\ref{sec:Inventing Atomic Loads Can Cause Errors and Destroy Semantic Dependencies}
+and~\ref{sec:Volatile and Quasi Volatile Accesses} below.
 

### 2025-03-05 paper/ootavacuous.tex
Add Ou and Demsky citation per Reviewer #72A
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 loads to be ordered before subsequent relaxed
-stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
+stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352,PiezhaoOu2018OOTApapl},
 but this can require executing real
@@
 This paper's contribution is to show how these constraints prevent
-real-world C++ implementations from exhibiting OOTA behavior.
+real-world C++ implementations from exhibiting OOTA behavior, and not with
+minimal cost~\cite{PiezhaoOu2018OOTApapl}, but rather with no cost at all.
 

### 2025-03-05 paper/ootavacuous.tex
ootavacuous: Clarify departure from prior OOTA work
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2025-03-04 paper/ootavacuous.tex
@@
 
-Some workers recommend avoiding OOTA by forcing prior relaxed
+Some researchers recommend avoiding OOTA by forcing prior relaxed
 loads to be ordered before subsequent relaxed
-stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
+stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352,PiezhaoOu2018OOTApapl},
 but this can require executing real
@@
 
-Other workers recommend various procedures to identify and avoid OOTA
+Other researchers recommend various procedures to identify and avoid OOTA
 cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
@@
 implementations to avoid it.
-None applied real-world hardware ordering constraints to the problem
-of avoiding OOTA cycles.
-This paper's contribution is to show how these constraints prevent
-real-world C++ implementations from exhibiting OOTA behavior.
+None applied real-world hardware ordering constraints (based on
+long-standing laws of physics) to the problem of avoiding OOTA cycles.
+
+Augmenting C++ abstractions with these constraints would of course
+limit applicability of this paper's results to C++ programs running
+on real-world computer systems.
+However, a quarter century of OOTA research based on pure C++ abstractions
+has resulted in incremental progress at best, and real-world computer
+systems comprise an exceedingly important part of the overall OOTA
+problem.
+In the wise words of Peter J.~Denning, ``Maybe we need to occasionally
+descend from the high clouds of our abstractions to the concrete earthy
+concerns of everyday life.''~\cite{PeterJDenning2025Abstraction}.
+
+Therefore, this paper's contribution is to show how these constraints
+prevent real-world C++ implementations from exhibiting OOTA behavior,
+and not with minimal cost~\cite{PiezhaoOu2018OOTApapl}, but rather with
+no cost at all.
 

### 2025-03-06 paper/ootavacuous.tex
ootavacuous: Point out that Section 7.4 depends on a community process
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 % witness also be transformed?
+In the unlikely event that this transformation proves useful, there will be
+a community process that determines whether this transformation is valid,
+that is, whether value of the use cases is worth the added conceptual load
+on developers.
+Until that time, it is impossible to say whether or not there is
+a semantic dependency from \co{y} to \co{z} without reference to a
+specific implementation.
 

### 2025-03-06 paper/ootavacuous.tex
ootavacuous: Call out further benefits of modular relaxed dependencies
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 As of 2024, Mark Batty is working on modular relaxed dependencies,
-which may have the added ability of checking compiler
-optimizations~\cite{JayRichards2024SymbolicMrder}.
+which may have the added benefit of checking compiler
+optimizations~\cite{JayRichards2024SymbolicMrder}, which is a portion
+of the OOTA problem that is out of scope for this paper.\footnote{
+	But please see Section~\ref{sec:Merging Quasi-Volatile Loads}.}
 

### 2025-03-07 paper/ootavacuous.tex
ootavacuous: Add citations per Reviewer #72E
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 loads to be ordered before subsequent relaxed
-stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352,PiezhaoOu2018OOTApapl},
+stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352,PiezhaoOu2018OOTApapl,SungHwanLee2023WeakMemoryPromising},
 but this can require executing real
@@
 Other researchers recommend various procedures to identify and avoid OOTA
-cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
-but none of these have been accepted by C++ implementers.
+cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA,MinkiCho2021PromisingSemantics,SohamChakraborty2019OOTAeventStructure,JeehoonKang2017PromisingSemantics,JeanPichonPharabod2016OOTA},
+but none of these have been accepted by C++ implementers, in part due
+to these models requiring consideration of multiple executions.
 As of 2024, Mark Batty is working on modular relaxed dependencies,
@@
 	But please see Section~\ref{sec:Merging Quasi-Volatile Loads}.}
+
+Moiseenko et al.~avoid cycles in po $\cup$
+rf~\cite{EvgeniiMoiseenko2025reexecution}, which might permit
+more optimizations than forcing prior relaxed loads to be ordered
+before subsequent relaxed stores, but which the authors characterize
+as ``(almost) a per-execution model''.
+It remains to be seen whether this is close enough to a per-execution
+model to be accepted by C++ implmenters.
+\emph{I am not fully confident of my characterization of this paper.}
+
+Jagadeesan et al.~combine preconditions and
+pomsets~\cite{RadhaJagadeesan2020OOTApomsetsPreconditions}, putting forward
+a memory model with attractive properties, but which does not support
+ARMv7 or PowerPC.
 

### 2025-03-07 paper/rebuttal-mods.txt
ootavacuous: Clarify departure from prior OOTA work
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
rebuttal-mods.txt: Update based on progress and feedback re-read
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-O	Detailed feedback from Review #72C.
+x	Detailed feedback from Review #72C.
 
@@
 O	Add citations to Section 1.2 ("Prior Work").
+	x	Ou and Demski PAPL 2018.
 
@@
 	the function will be part of a dynamic library that has not yet
-	been written.
+	been written.  (This addresses Reviewer #72F's call for handling
+	examples in Section 5 of [5].)
 
@@
 o	Add diagrams illustrating what the various components of sections
-	6.1 and 6.3 refer to.
+	6.1 and 6.3 refer to. (Reviewer #72B)
 
@@
 
-O	In Section 7.4 ("Merging Quasi-Volatile Loads"), change focus
+x	In Section 7.4 ("Merging Quasi-Volatile Loads"), change focus
 	from intuitive notions to a community-process approach for

### 2025-03-07 paper/ootavacuous.tex
ootavacuous: Add citations requested by Reviewer #72F
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 by William Pugh, represented by the infamous Java-language
-``Causality Test Cases''.\footnote{
-	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
-% %%% Eventually cite P3064.
+``Causality Test Cases''\footnote{
+	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}~\cite{JeremyManson2005JavaMemoryModel}.
+The OOTA problem has been open for more than 20~years and has been shown
+to be quite difficult~\cite{MarkBatty2015OOTAisHard}.
+% %%% @@@ Eventually cite P3064.
 
@@
 model to be accepted by C++ implmenters.
-\emph{I am not fully confident of my characterization of this paper.}
+\emph{@@@ I am not fully confident of my characterization of this paper.}
 
@@
 
-Therefore, this paper's contribution is to show how these constraints
-prevent real-world C++ implementations from exhibiting OOTA behavior,
-and not with minimal cost~\cite{PiezhaoOu2018OOTApapl}, but rather with
-no cost at all.
+Therefore, this paper's contribution is to show how these
+constraints prevent real-world C++ implementations doing per-thread
+analysis from exhibiting OOTA behavior, and not with minimal
+cost~\cite{PiezhaoOu2018OOTApapl}, but rather with no cost at all.
+\emph{@@@ Are these the right qualifiers?}
 

### 2025-03-07 paper/ootavacuous.tex
ootashort: Convert from ASPLOS to C++ working paper
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{document}
-\title{How to Avoid OOTA Without Really Trying}
+\title{DNNNNR0: How to Avoid OOTA Without Really Trying}
 
@@
 \author{
-Alan Stern (stern@rowland.harvard.edu) \and
-Paul E.~McKenney (paulmck@kernel.org) \and
-Michael Wong (fraggamuffin@gmail.com)
+Alan Stern \and
+Paul E.~McKenney \and
+Michael Wong \and
+Maged Michael
 }
@@
 	to the standard.
+
+	Audience: SG1.
 \end{abstract}
@@
 analysis from exhibiting OOTA behavior, and not with minimal
-cost~\cite{PiezhaoOu2018OOTApapl}, but rather with no cost at all.
+cost~\cite{PiezhaoOu2018OOTApapl}, but rather with no cost at
+all~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.
 \emph{@@@ Are these the right qualifiers?}
@@
 
+\section*{History}
+
+This paper is a condensation of P3064R2 ("How to Avoid OOTA Without
+Really Trying")~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.
+It also adds citations and clarifications.
+
 % \clearpage

### 2025-03-07 paper/ootavacuous.sh

@@ -86,7 +86,7 @@ do
 	fi
 done
 
-pdflatex asplos.tex || exit
-bibtex asplos.aux || exit
-pdflatex asplos.tex || exit
-pdflatex asplos.tex || exit
+pdflatex ootashort.tex || exit
+bibtex ootashort.aux || exit
+pdflatex ootashort.tex || exit
+pdflatex ootashort.tex || exit

### 2025-03-07 paper/ootavacuous.tex
ootavacuous: Add a section 1.3 ("Our Approach")
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 implementations to avoid it.
-None applied real-world hardware ordering constraints (based on
-long-standing laws of physics) to the problem of avoiding OOTA cycles.
-
-Augmenting C++ abstractions with these constraints would of course
-limit applicability of this paper's results to C++ programs running
-on real-world computer systems.
-However, a quarter century of OOTA research based on pure C++ abstractions
-has resulted in incremental progress at best, and real-world computer
-systems comprise an exceedingly important part of the overall OOTA
-problem.
+
+\subsection{Our Approach}
+\label{sec:Our Approach}
+
+Our approach applies real-world hardware constraints (which are further
+constrained by long-standing laws of physics) to the problem of avoiding
+OOTA cycles.
+We do this by augmenting the C++ abstract machine with these constraints,
+which of course limits the applicability of our results to C++ programs
+compiled and run on real-world computer systems.
+Given that a quarter century of OOTA research based on pure unaugmented
+C++ abstractions has resulted in (at best) incremental progress on the
+one hand, and that real-world computer systems comprise an exceedingly
+important part of the overall OOTA problem on the other, our results
+are valuable despite these limitations.
 In the wise words of Peter J.~Denning, ``Maybe we need to occasionally
 descend from the high clouds of our abstractions to the concrete earthy
-concerns of everyday life.''~\cite{PeterJDenning2025Abstraction}.
-
-Therefore, this paper's contribution is to show how these
+concerns of everyday life''~\cite{PeterJDenning2025Abstraction}.
+
+The key observations that enable us to apply these constraints are:
+(1)~Communicating data within a computer system takes time,
+(2)~Executing instructions that compute an output value that semantically
+depends on given input values also takes time,
+and
+(3)~Each link in an OOTA cycle consists of either data communication
+or computation of an output that semantically depends on an input.
+Because closing an OOTA cycle comprising a sequence of commmunications
+and computations requires this sequence to end at the same time as it
+starts, OOTA cycles cannot form in those C++ implementations that are
+within the scope of this paper.
+A great advantage of our approach is that we are \emph{not} required to
+precisely define the boundaries of sdep.
+After all, although compilers are forbidden from breaking semantic
+dependencies, they are by no means required to break non-semantic
+dependencies.
+
+These simple-seeming observations conceal considerable complexity,
+and so much of the remainder of this paper deals with corner cases that
+arise when carefully considering semantic dependencies, executions, C++
+compilers, and hardware properties.
+
+In short, this paper's contribution is to show how these
 constraints prevent real-world C++ implementations doing per-thread

### 2025-03-07 paper/rebuttal-mods.txt
rebuttal-mods.txt: Record progress
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-o	Early in the paper, consider explicitly stating that after 20
+x	Early in the paper, consider explicitly stating that after 20
 	years it appears that the problem of characterizing semantic
@@
 
-O	Add citations to Section 1.2 ("Prior Work").
+x	Add citations to Section 1.2 ("Prior Work").
 	x	Ou and Demski PAPL 2018.
@@
 
-o	Add a Section 1.3 ("Our Approach"?) that summarizes the
+x	Add a Section 1.3 ("Our Approach"?) that summarizes the
 	rfe/sdep/non-sdep-dep approach and emphasizes proving
@@
 
+o	Review Sections 6.1 and 6.3 for clarity per Reviewer #72F.
+
 x	On page 9, "While it possible".  [ Since fixed by Alan. ]

### 2025-03-08 paper/ootavacuous.tex
ootavacuous: Add JMM TC4 source for Listing 1
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 as shown in
-Listing~\ref{lst:Simple OOTA}.
+Listing~\ref{lst:Simple OOTA} (JMM TC4).\footnote{
+	Java Memory Model Test Case 4 from
+	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
 Here, all of \co{x}, \co{y}, \co{r1}, and \co{r2} might have final

### 2025-03-27 paper/ootavacuous.tex
ootavacuous: Distinguish between syntactic and semantic dependencies
But no separate section, not yet, anyway.  We already noted carelessness
on the part of others ("Some discussions of semantic dependencies assume
that they are strictly functions of the source code.")

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 dependencies.
+It is important to distinguish semantic dependencies from syntactic
+dependencies.
+Although every semantic dependency is also a syntactic dependency, many
+syntactic dependencies can be optimized (or ``broken''), as will be seen
+in the following sections.
+Such syntactic dependencies are not semantic dependencies.
 
@@
 are strictly functions of the source code.
-Although there are ways of making this work, many instances of
-semantic dependency must be considered functions of particular executions.
+This assumption is not always valid, in fact, we will see that many
+instances of semantic dependency must be considered functions of
+particular executions.
 Consider for example:

### 2025-03-27 paper/ootavacuous.tex
ootavacuous: Sharpen Section 7.4 community-process wording
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-Although it is unexpected, and while we don't recommend this sort of code
-transformation, we cannot say that this conclusion is definitely wrong,
-because our intuitive notions of semantic dependency are not clear in
-cases where multiple loads of the same variable are present.
+Although this is unexpected, and while we don't recommend this sort of
+code transformation, we cannot say that this conclusion is definitely
+wrong, because one's intuitive notions of semantic dependency are not
+always clear in cases involving multiple loads from the same variable.
 % Can we instead say that transforming the code might require that the
 % witness also be transformed?
-In the unlikely event that this transformation proves useful, there will be
-a community process that determines whether this transformation is valid,
-that is, whether value of the use cases is worth the added conceptual load
-on developers.
-Until that time, it is impossible to say whether or not there is
-a semantic dependency from \co{y} to \co{z} without reference to a
-specific implementation.
+In the unlikely event that this transformation proves useful, there
+will be a community process that determines whether the value of this
+transformation is worth the added conceptual load on developers.
+Until that time, pure mathematics cannot help us because it is simply
+impossible to say whether or not there is a semantic dependency from
+\co{y} to \co{z} without reference to a specific implementation.
 

### 2025-03-27 paper/ootavacuous.tex
ootavacuous: Add citations for litmus tests
The unspecified ordering for ++ might be *somewhere*...

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 exemplified by
-Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
+Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2014OOTA}.
 Either the compiler or the CPU might reorder T1's lines~1 and~2, so that
@@
 unmotivated values, as shown in
-Listing~\ref{lst:OOTA Cycle}.
+Listing~\ref{lst:OOTA Cycle}, adapted from~\cite{PaulEMcKenney2014OOTA}.
 Line~3 of both threads stores the value 42 only when line~2 determines
@@
 The OOTA problem has been open for more than 20~years and has been shown
-to be quite difficult~\cite{MarkBatty2015OOTAisHard}.
-% %%% @@@ Eventually cite P3064.
+to be quite difficult~\cite{MarkBatty2015OOTAisHard,PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.
 
@@
 rf is the set of store-to-load ``reads-from'' links, whether internally
-within a thread (rfi) or externally between threads (rfe).
-% @@@ Consider adding citation defining rf, rfi, and rfe.
+within a thread (rfi) or externally between threads (rfe).\footnote{
+	Additional background on temporal distinctions between coe, fre,
+	and rfe maybe found elsewhere~\cite[Appendix A]{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.}
 
@@
 
-Compare this example to the previous one:
+Compare this example~\cite{Howells2009membartxt} to the previous one:
 \begin{quote}
@@
 Consider the following example, with variable \co{i} initially zero:
+% @@@ Citation?
 \begin{quote}
@@
 Invented atomic loads are problematic.
-Consider this code:
+Consider this code~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}:
 \begin{quote}
@@
 Inventing atomic loads can also destroy dependencies.
-For example, consider this code:
+For example, consider this code~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}:
 \begin{quote}

### 2025-03-27 paper/ootavacuous.tex
ootavacuous: Recommend additional standard OOTA advice
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{quote}
-	Implementations should ensure that no “out-of-thin-air” values
+	Implementations should ensure that no ``out-of-thin-air'' values
 	are computed that circularly depend on their own computation.
@@
 
+\subsection{Possible Change to the Standard}
+\label{sec:Possible Change to the Standard}
+
+As noted in Section~\ref{sec:Brief OOTA Overview}, the C++
+standard explicitly recommends against them in 33.5.4p8
+(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
+\begin{quote}
+	Implementations should ensure that no ``out-of-thin-air'' values
+	are computed that circularly depend on their own computation.
+\end{quote}
+
+Sensible though this recommendation might be, it can unnecessarily raise
+anxiety levels of C++ implementers.
+After all, what exactly do they need to do in order to follow that
+recommendation?
+One way to reduce their anxiety levels would be to add something like
+the following sentence to this recommendation:
+\begin{quote}
+	Compiler-based implementations whose binaries run on conventional
+	hardware are guaranteed not to compute out-of-thin-air values as
+	long as they restrict themselves to thread-at-a-time analysis,
+	valid optimizations, and user programs that are free of undefined
+	behavior.
+\end{quote}
+
 \section{Summary and Conclusion}

### 2025-03-27 paper/ootavacuous.tex
ootavacuous: Add future work on speculative loads
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 one hand, and to link-time optimization (LTO) on the other;
-(2)~Consider various classes of global analysis, such as
+(2)~Relax quasi-volatile semantics to permit limited load speculation,
+for example, use of conditional-move instructions in cases where the
+source code conditionally consumes the results of the atomic load;
+(3)~Consider various classes of global analysis, such as
 demonstrating that a given expression will always evaluate to a constant;
-(3)~Examine the possibility of ``flattening'' optimizations that combine
+(4)~Examine the possibility of ``flattening'' optimizations that combine
 multiple threads into one;
-(4)~Explore cases in which semantic dependencies must be absolute,
+(5)~Explore cases in which semantic dependencies must be absolute,
 not relative to any particular compiler;
-(5)~Include the effects of input and output or other operations which
+(6)~Include the effects of input and output or other operations which
 can vary from one execution to another; and
-(6)~Delineate more precisely the limits of permissible behavior for
+(7)~Delineate more precisely the limits of permissible behavior for
 quasi-volatile object accesses.

### 2025-03-27 paper/rebuttal-mods.txt
rebuttal-mods.txt: Strike off completed items
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-o	Add citations to later sections that discuss litmus tests
+x	Add citations to later sections that discuss litmus tests
 	that are similar to those in the literature.
@@
 	been written.  (This addresses Reviewer #72F's call for handling
-	examples in Section 5 of [5].)
+	examples in Section 5 of [5].)  Or put this in a later section
+	demonstrating how we handle the various litmus tests.
 
-o	In section 2.2 where we start to talk about semantic dependencies
+x	In section 2.2 where we start to talk about semantic dependencies
 	in detail, we could mention syntactic dependencies too.  Explain
@@
 
+	Brainstorming for diagrams:
+
+	o	Section 1.3: Showing that an OOTA cycle cannot close.
+		Time increases downwards.
+
+	o	Section 6.1: Diagram showing source-code plane
+		and object-code plane, with mapping of accesses and
+		dependencies from one to the other.  Semantic source-level
+		dependencies are always mapped, but non-semantic
+		dependencies might or might not be mapped.
+
+		Section 6.1: Show the witness.	Or should this instead
+		be a side-by-side listing?
+
+	o	Section 6.3: Also show the witness.  Maybe only one
+		witness diagram/listing?
+
 o	Review Sections 6.1 and 6.3 for clarity per Reviewer #72F.
@@
 
-	Also in Section 7.4 change "our intuitive notions" to "one's
+x	Also in Section 7.4 change "our intuitive notions" to "one's
 	intuitive notions".
@@
 
-o	Possible future work: Relax quasi-volatile semantics to permit
+x	Possible future work: Relax quasi-volatile semantics to permit
 	load speculation, for example, use of the cmov instruction in

### 2025-04-30 paper/ootavacuous.tex
ootavacuous: Add note from search for ++ litmus tests
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Add a section on semantic dependencies and function calls
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
+\subsubsection{Semantic Dependencies and Function Calls}
+\label{sec:Semantic Dependencies and Function Calls}
+
+Boehm and Dempsky~\cite[Section 5]{Boehm:2014:OGA:2618128.2618134}
+point out that arbitrary function calls can pose challenges for any
+attempt to determine semantic dependencies based solely on analysis
+of the source code, using the following example:
+\begin{quote}
+\begin{verbatim}
+y = f(x);
+\end{verbatim}
+\end{quote}
+And it is indeed impossible to determine whether or not there is a
+semantic dependency from \co{x} to \co{y} without reference to the
+definition of \co{f()}.
+For some definitions of \co{f()}, a semantic dependency exists, and for
+others it does not.
+
+Note that \co{f} might well be a pointer to a function, in which case
+the existence of that semantic dependency might well vary at runtime as
+the value of function pointer \co{f} changes.
+
 \subsubsection{Semantic Dependencies and Free Choices}
@@
 % @@@ Citation?
+% @@@ "Outlawing ghosts: Avoiding out- of-thin-air results" has a
+% @@@    ++ example, but is concerned about hoisting that before
+% @@@    an "if" statement.
 \begin{quote}

### 2025-05-10 paper/ootavacuous.tex
ootavacuous: Add a section giving examples exercising the definition
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-As long as \co{z} is zero, changes in the
+Expanding on earlier discussions~\cite{Boehm:2014:OGA:2618128.2618134},
+as long as \co{z} is zero, changes in the
 value of \co{y} will not cause a change in the value stored to \co{x}.
@@
 
+\subsection{Exercising the Definition}
+\label{sec:Exercising the Definition}
+
+This section exercises the definition of ``semantic dependency''
+presented in the previous section on selected examples.
+These exercises use the following steps:
+
+\begin{enumerate}
+\item	Identify the potential semantic dependencies of interest.
+\item	Compile the code to assembly language, either using a
+	real compiler or conceptually.
+\item	Check for an assembly level syntactic dependency between
+	the loads and stores making of the potential semantic
+	dependency.
+	\begin{enumerate}
+	\item	If there is no assembly level syntactic dependency,
+		then there can be no semantic
+		dependency.\label{item:No assembly level syntactic dependency}
+	\item	Otherwise, if there is no execution having a witness,
+		then there can be no semantic
+		dependency.\label{item:No witness}
+	\item	Otherwise, there is a semantic
+		dependency.\label{item:No sdep}
+	\end{enumerate}
+\end{enumerate}
+
+In cases~\ref{item:No assembly level syntactic dependency}
+and~\ref{item:No witness}
+there is no semantic dependency.
+In case~\ref{item:No sdep},
+there might or might not be a semantic dependency, but
+time will definitely be consumed executing the resulting code.
+Either way, the candidate dependency cannot be part of an OOTA
+cycle.
+
+The following sections apply these steps to two of the examples in
+this paper.
+% Should we call out the relationship to the perturbation method of P0422r0?
+
+\subsubsection{Exercising Function Calls}
+\label{sec:Exercising Function Calls}
+
+Section~\ref{sec:Semantic Dependencies and Function Calls}
+provides the following example:
+\begin{quote}
+\begin{verbatim}
+y = f(x);
+\end{verbatim}
+\end{quote}
+If the compiler knows nothing about the function \co{f()} and the variable
+\co{y}, it must emit code that loads from \co{x}, passes this value to
+\co{f()}, and finally stores the return value to \co{y}.
+Is there a semantic dependency from the load of \co{x} to the store of
+\co{y}?
+To determine this, it is necessary to apply this same analysis to the
+definition of \co{f()}, treating its argument as a load and each of its
+\co{return} statements as a store.
+If and only if this analysis shows that, for a given execution, there
+is a semantic dependency within \co{f()}, then there must also be a
+semantic dependency from the load of \co{x} to the store of \co{y}.
+Of course, if the compiler cannot prove that there is no such semantic
+dependency, it must assume than there is one.
+
+If \co{f} is a pointer to a function, then the compiler must also
+generate load from \co{f}, so that there might be a semantic dependency
+from either or both of \co{f} and \co{y} to the store to \co{y}.
+This is straightforward:
+A given execution will load a pointer to a particular function from
+\co{f}, so if there is another execution that loads a pointer to a
+different function, and the two functions return different values for
+that execution's value from \co{y}, then there is a semantic dependency
+from \co{f} to \co{y}.
+
+Situations where the compiler knows about \co{f} more often result in
+inlining, in which case there is no function call, allowing the semantic
+dependencies (or lack thereof) to be determined with respect to the
+inlined code.
+
+\subsubsection{Exercising Multiplication}
+\label{sec:Exercising Multiplication}
+
+Sections~\ref{sec:Semantic Dependencies and Source Code}
+and~\ref{sec:Semantic Dependencies Can Be Many-To-One}
+provide the following example:
+\begin{quote}
+\begin{verbatim}
+x = y * z;
+\end{verbatim}
+\end{quote}
+This has potential semantic dependencies between the loads from \co{x}
+and \co{y} to the store to \co{z}.
+The corresponding assembly code can vary depending on what the compiler
+knows about the values of \co{x} and \co{y}.
+
+If it knows nothing, then it must load from both, multiply the values
+loaded, and then store the resulting product to \co{z}.
+There is an assembly-level syntactic dependency, and if we start
+with the execution in which the value 1 is loaded from both variables
+(resulting in the value 1 being stored to \co{z}),
+then the execution in which the value 2 is loaded from \co{x} witnesses
+a semantic dependency from \co{x} to \co{z}.
+Similarly, the execution in which the value 2 is loaded from \co{y}
+witnesses a semantic dependency from \co{y} to \co{z}.
+Both witnesses result in the value 2 being stored to \co{z}.
+
+But suppose we instead start with the execution where the value 0
+is loaded from \co{x} and the value 1 is loaded from \co{y}.
+In this case, there is no witness execution that varies only the value
+of \co{y}, as all such executions will result in the value 0 being stored
+to \co{z}.
+However, the execution where the value -1 is loaded from \co{x}
+witnesses a semantic dependency from \co{x} to \co{y}, resulting in
+the value -1 being stored to \co{z}.
+Furthermore, the execution where the value -1 is loaded from \co{x}
+and the value -2 is loaded from \co{y} witnesses a semantic dependency
+from the set \{\co{x}, \co{y}\} to \co{z}, resulting in the value 2 being
+stored to \co{z}.
+
+But suppose we instead start with the execution where the value 0
+is loaded from both \co{x} and \co{y}.
+There is again no witness execution that varies only the value of \co{y},
+but there is also no witness execution that varies only the value of \co{x},
+all such executions resulting in the value 0 stil being stored to \co{z}.
+However, the execution where the value 2 is loaded from both \co{x} and
+\co{y} witnesses a semantic dependency from the set \{\co{x}, \co{y}\}
+to \co{z}, resulting in the value 4 being stored to \co{z}.
+
+Alternatively, if the compiler knows that the value loaded from \co{x}
+is guaranteed to be zero, then it can omit the loads from both \co{x}
+and \co{y} (assuming that the value from \co{y} is not otherwise used),
+storing the constant 0 to \co{z}.
+There can be no semantic dependencies from these now-nonexistent loads
+to the store to \co{z}.
+And then any execution that witnessed a semantic dependency from
+either or both of \co{x} and \co{y} would indicate a bug in the
+compiler, which would have been wrong in its assumption of the
+guarantee that \co{x} was zero.
+
 \section{Issues and Refinements}

### 2025-05-10 paper/rebuttal-mods.txt
Update todo list
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-o	Consider adding a 2.2.x section that discusses the difficulties
+x	Consider adding a 2.2.x section that discusses the difficulties
 	inherent in computing sdep for function calls, especially when
@@
 
-o	Add a section before Section 7 ("Issues and Refinements") that
+x	Add a section before Section 7 ("Issues and Refinements") that
 	shows how our approach handles a few well-known examples of OOTA.

### 2025-05-11 paper/ootavacuous.tex
ootavacuous: Move history section to front
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \maketitle{}
+
+\section*{History}
+
+This paper is a condensation of P3064R2 ("How to Avoid OOTA Without
+Really Trying")~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying},
+which was criticized as being too long.
+It also adds citations and clarifications.
 
@@
 
-\section*{History}
-
-This paper is a condensation of P3064R2 ("How to Avoid OOTA Without
-Really Trying")~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.
-It also adds citations and clarifications.
-
 % \clearpage

### 2025-05-11 paper/ootavacuous.tex
ootavacuous: Page-1 wordsmithing
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2025-05-10 paper/ootavacuous.tex
@@
 	We accommodate nonvolatile atomics by defining ``quasi-volatile''
-	behavior which enables existing implementations that do per-thread
-	analysis and optimization not only to avoid OOTA, but also to
-	avoid incorrectly evaluating arithmetic expressions involving
-	atomic accesses.
+	behavior and show that existing implementations doing per-thread
+	analysis and optimization not only avoid OOTA, but also avoid
+	incorrectly evaluating arithmetic expressions involving atomic
+	accesses.
 
@@
 \maketitle{}
+
+\section*{History}
+
+This paper is a condensation of P3064R2 ("How to Avoid OOTA Without
+Really Trying")~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying},
+which was criticized as being overly long.
+It also adds citations and clarifications.
 

### 2025-05-11 paper/ootavacuous.tex
ootavacuous: Page 4 wordsmithing
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 definition of \co{f()}.
-For some definitions of \co{f()}, a semantic dependency exists, and for
-others it does not.
-
-Note that \co{f} might well be a pointer to a function, in which case
-the existence of that semantic dependency might well vary at runtime as
-the value of function pointer \co{f} changes.
+In addition, \co{f} might well be a pointer to a function, in which case
+the existence of a semantic dependency from \co{x} to \co{y} might well
+vary at runtime as the value of function pointer \co{f} changes.
 

### 2025-05-11 paper/ootavacuous.tex
ootavacuous: Wordsmith Sections 6.4 and 6.4.1
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 
-This section exercises the definition of ``semantic dependency''
-presented in the previous section on selected examples.
+This section exercises the definition of ``semantic dependency'' on
+selected examples.
 These exercises use the following steps:
@@
 If the compiler knows nothing about the function \co{f()} and the variable
-\co{y}, it must emit code that loads from \co{x}, passes this value to
+\co{x}, it must emit code that loads from \co{x}, passes this value to
 \co{f()}, and finally stores the return value to \co{y}.
@@
 semantic dependency from the load of \co{x} to the store of \co{y}.
-Of course, if the compiler cannot prove that there is no such semantic
+As always, if the compiler cannot prove that there is no such semantic
 dependency, it must assume than there is one.
 
-If \co{f} is a pointer to a function, then the compiler must also
-generate load from \co{f}, so that there might be a semantic dependency
-from either or both of \co{f} and \co{y} to the store to \co{y}.
-This is straightforward:
+If \co{f} is a pointer to a function, then the compiler must also emit
+a load from \co{f}, so that there might also be a semantic dependency
+from the load from \co{f} to the store to \co{y}.
+In the context of C++, this is straightforward:
 A given execution will load a pointer to a particular function from
@@
 different function, and the two functions return different values for
-that execution's value from \co{y}, then there is a semantic dependency
-from \co{f} to \co{y}.
-
-Situations where the compiler knows about \co{f} more often result in
+the value loaded from \co{x}, then there is a semantic dependency from
+\co{f} to \co{y}.
+Languages that permit functions to be generated at runtime must use
+more complicated analysis.
+
+Situations where the compiler knows about \co{f} often result in
 inlining, in which case there is no function call, allowing the semantic

### 2025-05-11 paper/ootavacuous.tex
ootavacuous: Sections 2.2.1/2.2.2: Use z=x*y to be consistent with y=f(x)
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{verbatim}
-x = y * z;
+z = x * y;
 \end{verbatim}
@@
 Expanding on earlier discussions~\cite{Boehm:2014:OGA:2618128.2618134},
-as long as \co{z} is zero, changes in the
-value of \co{y} will not cause a change in the value stored to \co{x}.
-As a result, the semantic dependency from \co{y} to \co{x} exists only
-in executions where \co{z} is nonzero,
+as long as \co{y} is zero, changes in the
+value of \co{x} will not cause a change in the value stored to \co{z}.
+As a result, the semantic dependency from \co{x} to \co{z} exists only
+in executions where \co{y} is nonzero,
 which shows it is a property of the execution, not just of the source code.
@@
 Suppose that in some execution of the previous example,
-both \co{y} and \co{z} are zero.
-Then changes to either \co{y} or \co{z}
-will not cause a change in the value stored to \co{x}.
+both \co{x} and \co{y} are zero.
+Then changes to either \co{x} or \co{y}
+will not cause a change in the value stored to \co{z}.
 In other words, in this execution there is no semantic dependency
-from either \co{y} or \co{z} to \co{x}.
+from either \co{x} or \co{y} to \co{z}.
 But there \emph{is} a semantic dependency from the \emph{pair}
-\{\co{y},~\co{z}\} to \co{x},
-because changes to both \co{y} and \co{z}
-can cause the value stored in \co{x} to change.
+\{\co{x},~\co{y}\} to \co{z},
+because changes to both \co{x} and \co{y}
+can cause the value stored in \co{z} to change.
 This means that, prior work~\cite{PaulEMcKenney2016OOTA}

### 2025-05-11 paper/ootavacuous.tex
ootavacuous: Wordsmith section 6.4.2
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{verbatim}
-x = y * z;
+z = x * z;
 \end{verbatim}
@@
 witnesses a semantic dependency from \co{y} to \co{z}.
-Both witnesses result in the value 2 being stored to \co{z}.
 
@@
 However, the execution where the value -1 is loaded from \co{x}
-witnesses a semantic dependency from \co{x} to \co{y}, resulting in
+witnesses a semantic dependency from \co{x} to \co{z}, resulting in
 the value -1 being stored to \co{z}.
-Furthermore, the execution where the value -1 is loaded from \co{x}
-and the value -2 is loaded from \co{y} witnesses a semantic dependency
-from the set \{\co{x}, \co{y}\} to \co{z}, resulting in the value 2 being
-stored to \co{z}.
 
@@
 but there is also no witness execution that varies only the value of \co{x},
-all such executions resulting in the value 0 stil being stored to \co{z}.
-However, the execution where the value 2 is loaded from both \co{x} and
-\co{y} witnesses a semantic dependency from the set \{\co{x}, \co{y}\}
-to \co{z}, resulting in the value 4 being stored to \co{z}.
+all such executions still resulting in the value 0 being stored to \co{z}.
+However, the execution where the value 2 is loaded from \co{x} and the
+value 3 is loaded from \co{y} witnesses a semantic dependency from the
+set \{\co{x}, \co{y}\} to \co{z}, resulting in the value 6 being stored
+to \co{z}.
 
@@
 is guaranteed to be zero, then it can omit the loads from both \co{x}
-and \co{y} (assuming that the value from \co{y} is not otherwise used),
-storing the constant 0 to \co{z}.
-There can be no semantic dependencies from these now-nonexistent loads
-to the store to \co{z}.
-And then any execution that witnessed a semantic dependency from
-either or both of \co{x} and \co{y} would indicate a bug in the
-compiler, which would have been wrong in its assumption of the
-guarantee that \co{x} was zero.
+and \co{y} (assuming that these values are not otherwise used), and
+then simply store the constant 0 to \co{z}.
+There clearly can be no semantic dependencies from these now-nonexistent
+loads to the store to \co{z}.
 

### 2025-05-11 paper/ootavacuous.tex
ootavacuous: Section 6.4.1: Assembly language level syntactic dependency
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 If and only if this analysis shows that, for a given execution, there
-is a semantic dependency within \co{f()}, then there must also be a
-semantic dependency from the load of \co{x} to the store of \co{y}.
+is an assembly level syntactic dependency within \co{f()}, then there
+must also be a semantic dependency from the load of \co{x} to the store
+of \co{y}.
 As always, if the compiler cannot prove that there is no such semantic

### 2025-05-13 paper/ootavacuous.tex
ootavacuous: Add 6.4.3 discussion section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Set up to be P3692R0
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{document}
-\title{DNNNNR0: How to Avoid OOTA Without Really Trying}
+\title{P3692R0: How to Avoid OOTA Without Really Trying}
 
@@
 
-This paper is a condensation of P3064R2 ("How to Avoid OOTA Without
-Really Trying")~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying},
+This paper is a condensation of P3064R2 (``How to Avoid OOTA Without
+Really Trying'')~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying},
 which was criticized as being overly long.
@@
 % This citation credits the formulation to Ali Sezgin, third author
-% of N4323 ("Out-of-Thin-Air Execution is Vacuous").
+% of N4323 (``Out-of-Thin-Air Execution is Vacuous'').
 where sdep is the set of semantic dependencies within each thread and
@@
 
+\subsubsection{Discussion}
+\label{sec:Discussion}
+
+Allowing the compiler to transform the source code to assembly code
+prior to analysis provides great simplifications.
+This enables us to show the main result, namely that current compiler-based
+C++ implementations do not produce OOTA cycles.
+Note well that these compilers' current analysis suffices.
+
 \section{Issues and Refinements}
@@
 which is not cited in order to maintain anonymity.
-% P3064R2 ("How to Avoid OOTA Without Really Trying")
+% P3064R2 (``How to Avoid OOTA Without Really Trying'')
 % https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2024/p3064r2.pdf

### 2025-05-22 paper/ootavacuous.tex
ootavacuous; Alan Stern May 19 2025 updates
Message ID: 87f55604-c3b6-4f4e-960d-e83026c5ea18@rowland.harvard.edu

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 C++ implementations.
-(The as-if rule permits those optimizations where the optimized code's
+(The as-if rule permits optimizations where the optimized code's
 observable behavior is the same as if it were unoptimized.)
@@
 T0 and T1.
-(Roughly speaking, there is a semantic dependency from a given load to
+Roughly speaking, there is a semantic dependency from a given load to
 a given store when \emph{all other things being equal, a change in the
@@
 the store, or prevent the store from occurring at all.}
-Here the dependencies are trivial because the values stored simply
-\emph{are} the values that were loaded.)
+(A \emph{syntactic} dependency, by contrast, exists whenever the value
+of the load appears in the source code as part of the computation
+leading up to the store, whether it affects the store or not.
+Only semantic dependencies are relevant to OOTA.)
+Here the semantic dependencies are trivial because the values stored simply
+\emph{are} the values that were loaded.
+
 Since real-world CPUs cannot store something
@@
 are fixed-point computations
-In other words, an OOTA cycle must produce the same value that it
-consumes.
+(in other words, an OOTA cycle must produce the same value that it consumes).
 
@@
 which may have the added benefit of checking compiler
-optimizations~\cite{JayRichards2024SymbolicMrder}, which is a portion
+optimizations~\cite{JayRichards2024SymbolicMrder}, a portion
 of the OOTA problem that is out of scope for this paper.\footnote{
@@
 It remains to be seen whether this is close enough to a per-execution
-model to be accepted by C++ implmenters.
+model to be accepted by C++ implementers.
 \emph{@@@ I am not fully confident of my characterization of this paper.}
@@
 
-Our approach applies real-world hardware constraints (which are further
-constrained by long-standing laws of physics) to the problem of avoiding
+Our approach applies real-world hardware constraints (including some
+imposed by long-standing laws of physics) to the problem of avoiding
 OOTA cycles.
-We do this by augmenting the C++ abstract machine with these constraints,
+We do this by relating the C++ abstract machine to hardware executions,
 which of course limits the applicability of our results to C++ programs
@@
 C++ abstractions has resulted in (at best) incremental progress on the
-one hand, and that real-world computer systems comprise an exceedingly
+one hand, and that
+compiler-based real-world computer systems comprise an exceedingly
 important part of the overall OOTA problem on the other, our results
@@
 
-The key observations that enable us to apply these constraints are:
-(1)~Communicating data within a computer system takes time,
+The key constraints mentioned above are:
+(1)~Communicating data within a computer system takes time, and
 (2)~Executing instructions that compute an output value that semantically
-depends on given input values also takes time,
-and
-(3)~Each link in an OOTA cycle consists of either data communication
-or computation of an output that semantically depends on an input.
-Because closing an OOTA cycle comprising a sequence of commmunications
+depends on given input values also takes time.
+Because each link in an OOTA cycle consists of either data communication
+or computation of an output that semantically depends on an input,
+and because closing an OOTA cycle comprising a sequence of commmunications
 and computations requires this sequence to end at the same time as it
@@
 within the scope of this paper.
-A great advantage of our approach is that we are \emph{not} required to
-precisely define the boundaries of sdep.
-After all, although compilers are forbidden from breaking semantic
-dependencies, they are by no means required to break non-semantic
-dependencies.
-
-These simple-seeming observations conceal considerable complexity,
-and so much of the remainder of this paper deals with corner cases that
+A great advantage of our approach is that we do \emph{not} need to define
+semantic dependency purely at the level of C++ source code;
+rather, we are able to utilize information already provided by existing
+hardware execution models.
+
+These simple-seeming ideas conceal considerable complexity,
+and much of the remainder of this paper deals with corner cases that
 arise when carefully considering semantic dependencies, executions, C++
@@
 In short, this paper's contribution is to show how these
-constraints prevent real-world C++ implementations doing per-thread
-analysis from exhibiting OOTA behavior, and not with minimal
-cost~\cite{PiezhaoOu2018OOTApapl}, but rather with no cost at
+constraints allow real-world C++ implementations doing per-thread
+analysis to avoid OOTA behavior easily,
+%and not with minimal cost~\cite{PiezhaoOu2018OOTApapl},
+in many cases with no cost at
 all~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.
@@
 
-\section{OOTA and Semantic Dependencies (sdep)}
-\label{sec:OOTA and Semantic Dependencies (sdep)}
+\section{OOTA, Semantic Dependencies, and Reads-From}
+\label{sec:OOTA, Semantic Dependencies, and Reads-From}
 
@@
 
-Semantic dependencies form only one type of link in an OOTA cycle.
-The other type extends from a given store to a load that returns the
+Semantic dependencies form only one type of link in an OOTA cycle;
+the other type extends from a given store to a load that returns the
 value stored.
-It is tempting to argue that the store must precede
-the load in global time and combine this with the intuitive notion that
-any real C++ implementation must consume global time when computing a
-semantic dependency.
-This combination suggests that OOTA cycles cannot occur.
-The idea has been formalized by defining an OOTA cycle as a cycle
+In principle, the time required by each of these links should prevent
+OOTA cycles from occurring.
+This idea has been formalized by defining an OOTA cycle as a cycle
 in sdep $\cup$ rf~\cite{PaulEMcKenney2014OOTA},
@@
 The inability of rfi links to act as temporal constraints is not the
-only, or even the main, weakness in the intuitive argument against
+only, or even the main, weakness in the naive argument against
 OOTA cycles.
@@
 There would then be no constraint forcing the implementation to execute
-the thread's store later in global time than the load it supposedly depends on,
+the thread's store later than the load it supposedly depends on,
 and thus no impediment to the occurrence of an OOTA cycle.
@@
 dependencies.
-It is important to distinguish semantic dependencies from syntactic
-dependencies.
-Although every semantic dependency is also a syntactic dependency, many
-syntactic dependencies can be optimized (or ``broken''), as will be seen
-in the following sections.
-Such syntactic dependencies are not semantic dependencies.
 
@@
 Some discussions of semantic dependencies assume that they
-are strictly functions of the source code.
-This assumption is not always valid, in fact, we will see that many
-instances of semantic dependency must be considered functions of
+are strictly properties of the source code.
+This assumption is not always valid; in fact, many
+instances of semantic dependency vary according to the details of
 particular executions.
@@
 the notion of a cycle involving sdep appears problematic.
-We are forced to change our definition of an OOTA cycle again;
+We are forced to change our definition of OOTA again;
 we will say that an execution is an instance of OOTA if in that execution:
@@
 
+\medskip
+
 We have seen several examples showing that semantic dependencies can
@@
 execution and even the implementation.
-This raises several questions, of which the first is:
-What exactly is an execution?
+This raises several questions, of which the first is\dots
 
@@
 abstract execution of the source code given the same input (in the
-absence of any abstract executions containing undefined behavior)
+absence of any abstract executions containing undefined behavior).
 This requirement is the standard's ``as-if'' rule.
-This means:
+It means:
 (1)~The program's output must be the same as that of the abstract execution;
@@
 which does not matter for our purposes].
-We will say that the hardware execution realizes the abstract execution,
+We will say that the hardware execution \emph{realizes} the abstract execution,
 or, equivalently, that the abstract execution is \emph{realized by}
@@
 before relaxed stores!
+(In fact, the transformed program has no loads at all, relaxed or otherwise.)
 
@@
 If all the stores in the OOTA cycle are omitted then all the reads must
-also be omitted, meaning that the entire cycle has been optimized away.
+also be omitted, meaning that the entire cycle has been optimized
+out of existence.
 We conjecture that in this situation there must be another abstract
@@
 
-We informally define a \emph{witness} to be another execution in which
-at least one of the loads obtains a different value but all else is the
-same as far as possible, and yet the store acts differently.
+Informally, a \emph{witness} is another execution in which
+at least one of the $R_i$ loads obtains a different value but all else is the
+same as far as possible, and yet the store $W$ acts differently.
 
@@
 semantic dependency.
-Additional information is available in a companion technical report,
-which is not cited in order to maintain anonymity.
-% P3064R2 (``How to Avoid OOTA Without Really Trying'')
-% https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2024/p3064r2.pdf
+Additional information is available in P3064R2 (``How to Avoid OOTA Without
+Really Trying'')~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.
 
@@
 transformation is worth the added conceptual load on developers.
-Until that time, pure mathematics cannot help us because it is simply
+Until that time, pure mathematics cannot help us simply because it is
 impossible to say whether or not there is a semantic dependency from
@@
 one hand, and to link-time optimization (LTO) on the other;
-(2)~Relax quasi-volatile semantics to permit limited load speculation,
-for example, use of conditional-move instructions in cases where the
-source code conditionally consumes the results of the atomic load;
+(2)~Relax quasi-volatile semantics to permit limited load speculation
+(for example, hoisting loads out of the bodies of conditionals);
 (3)~Consider various classes of global analysis, such as
@@
 	Compiler-based implementations whose binaries run on conventional
-	hardware are guaranteed not to compute out-of-thin-air values as
-	long as they restrict themselves to thread-at-a-time analysis,
-	valid optimizations, and user programs that are free of undefined
-	behavior.
+	hardware are guaranteed not to compute out-of-thin-air values in
+	programs that are free of undefined behavior,
+	as long as they restrict themselves to thread-at-a-time analysis
+	and do not invent or duplicate atomic accesses.
 \end{quote}
@@
 The conclusion to be drawn from this work is that in this context,
-avoiding OOTA cycles requires no changes to the standard,
-to current implementions, or to user practices for portable code.
+avoiding OOTA cycles requires no changes to the standard
+or to user practices for portable code, and only minimal changes to
+implementations.
 % This work also demonstrates the need for caution when optimizing

### 2025-06-11 paper/ootavacuous.tex
More minor updates
More minor updates to the text, mostly concerned with the introductions
of the notions of semantic and syntactic dependencies.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 C++ implementations.
-(The as-if rule permits optimizations where the optimized code's
+(The as-if rule permits optimizations for which the optimized code's
 observable behavior is the same as if it were unoptimized.)
@@
 
-In Listing~\ref{lst:Simple OOTA}
-there is a \emph{semantic dependency} from line~1 to line~2 in both
-T0 and T1.
+Unfortunately, the standard does not explain what the phrase ``depend on''
+in the quotation above really means.
+It is generally taken to mean \emph{semantic dependency},
+a somewhat vague concept.
 Roughly speaking, there is a semantic dependency from a given load to
-a given store when \emph{all other things being equal, a change in the
+a given store in the same thread when
+\emph{all other things being equal, a change in the
 value loaded can change the value stored, change the target address of
 the store, or prevent the store from occurring at all.}
+In Listing~\ref{lst:Simple OOTA}, for example,
+there is a semantic dependency from line~1 to line~2 in both T0 and T1.
+The semantic dependencies here are trivial,
+because the values stored simply \emph{are} the values that were loaded.
 (A \emph{syntactic} dependency, by contrast, exists whenever the value
@@
 leading up to the store, whether it affects the store or not.
-Only semantic dependencies are relevant to OOTA.)
-Here the semantic dependencies are trivial because the values stored simply
-\emph{are} the values that were loaded.
+\emph{@@@ Reference?}
+Only semantic dependencies contribute to OOTA.)
 
@@
 or computation of an output that semantically depends on an input,
-and because closing an OOTA cycle comprising a sequence of commmunications
-and computations requires this sequence to end at the same time as it
+and because closing such a cycle requires the sequence of links
+to end at the same time as it
 starts, OOTA cycles cannot form in those C++ implementations that are
@@
 
-These simple-seeming ideas conceal considerable complexity,
+These simple-seeming notions conceal considerable complexity,
 and much of the remainder of this paper deals with corner cases that

### 2025-06-11 paper/ootavacuous.tex
Reformulate in terms of preserving dependencies
Reformulate the paper's conception of semantic dependency relative to
a particular compiler.  In this new version, the text talks instead
about compiled code preserving syntactic dependencies.

In the original paper, perhaps the most questionable aspect was the
unspoken issue of whether the definition of semantic dependency did an
adequate job of capturing the informal notion.  In the new version,
this issue translates into an explicit Dependency Preservation thesis,
stating that semantic dependencies are always preserved.

In addition to the vir

@@
 	We accommodate nonvolatile atomics by defining ``quasi-volatile''
-	behavior and show that existing implementations doing per-thread
-	analysis and optimization not only avoid OOTA, but also avoid
+	behavior and show that for existing implementations doing per-thread
+	analysis and optimization, this behavior not only avoids OOTA
+	but also is necessary to avoid
 	incorrectly evaluating arithmetic expressions involving atomic
@@
 	It follows that enforcing OOTA avoidance requires no special
-	action from implementations and no changes to user code or
-	to the standard.
+	action from implementations and no changes to user code.
 
@@
 because the values stored simply \emph{are} the values that were loaded.
-(A \emph{syntactic} dependency, by contrast, exists whenever the value
+
+A \emph{syntactic} dependency, by contrast, exists whenever the value
 of the load appears in the source code as part of the computation
@@
 \emph{@@@ Reference?}
-Only semantic dependencies contribute to OOTA.)
+Although we will not give a precise definition of syntactic dependency
+here, doing so presents no fundamental difficulties.
+In fact, the C++ standard covers much of the same ground in its definition
+of ``carries a dependency'' in 6.9.2.2p7 (\co{[Data races]}).
+Of course, all semantic dependencies are also syntactic.
+Note, however, that only semantic dependencies contribute to OOTA.
 
@@
 	A similar restriction applies to compiler-based speculation.}
-the stores on line~2 cannot take place until T0's and T1's CPUs
+the stores on line~2 of Listing~\ref{lst:Simple OOTA}
+cannot take place until T0's and T1's CPUs
 know what values are returned by the loads on line~1.
@@
 The primary difficulty lies in the fact that the code transformations
-performed by optimizing compilers can destroy dependencies, including
-semantic ones (depending on one's definition).
-That is, even when the potential for a dependency exists in the
+performed by optimizing compilers can destroy syntactic dependencies,
+possibly even semantic ones (depending on one's definition).
+That is, even when a syntactic dependency exists in the
 source code for a thread, there might be no dependency in the machine
@@
 
-\section{A Definition of Semantic Dependency}
-\label{sec:A Definition of Semantic Dependency}
+\section{Preservation of Dependencies}
+\label{sec:Preservation of Dependencies}
 
@@
 the variables involved.
-In this section we will stick our necks out by offering just such a
-definition.
-No doubt many people will object to it for various reasons, but we
-nevertheless hope it will help move the discussion forward.
+In this section we will bypass the difficulty by instead defining
+when an execution of compiled code \emph{preserves a syntactic dependency}.
+
+In addition, we propose the following Dependency Preservation (DP) thesis:
+\begin{quote}
+	If a syntactic dependency is semantic then it is preserved
+	by all compilers for which the definition below applies.
+\end{quote}
+As with the Church-Turing thesis in computability theory,
+our DP thesis is not susceptible of formal proof because
+the main concept it deals with (semantic dependency)
+does not have a formal definition.
+Nevertheless, we hope that readers will agree our definition of
+dependency preservation captures the informal notion of
+semantic dependency sufficiently well to justify the thesis.
+It will be all we need to rule out the possibility of OOTA.
 
@@
 Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
-The key insight is that this allows us to consider semantic
+The key insight is that this allows us to consider
 dependencies at the level of the machine code, where they are much
 more tractable.
-Using this approach, we demonstrate that OOTA cycles cannot form in the
-C++ implementations addressed by this paper.
 
@@
 
-We begin by recognizing that semantic dependencies are relative to a
-particular execution and a particular implementation,
+We begin by recognizing that preservation of dependencies is relative to a
+particular execution,
 as described in Section~\ref{sec:Properties of Semantic Dependencies}.
-The same source code may or may not contain a semantic dependency,
-according to the details of the execution in question and the
-machine code produced by the compiler.
-For this reason we will characterize semantic dependencies in a given
+A syntactic dependency present in the source code may or may not be
+preserved in the machine code produced by a compiler,
+according to the details of the execution in question.
+For this reason we will characterize preservation of dependencies in a given
 abstract execution realized by a given hardware execution.
-(While it is possible to argue about semantic dependencies in abstract
+(While it is possible to argue about preservation of dependencies in abstract
 executions that have no hardware realizations, doing so seems pointless.)
@@
 \ldots,~$R_n\}$ in $T$ be a set of loads from atomic objects on which
-$W$ might or might not depend.
+$W$ has a syntactic dependency.
 We can dispose of one case immediately: If $W$ is omitted in $H$ then
-the issue of semantic dependency is moot.
+the issue of dependency preservation is moot.
 You can give either answer since it will have no effect.
@@
 \begin{quote}
-There is a semantic dependency from $\{R_i\}$ to $W$ in $A$ and $H$,
-relative to the compiler used to produce $H$, if there is another
-abstract execution $B$ realized by hardware execution $G$ under the
-same compiler that together \emph{witness the semantic dependency}.
+The dependency from $\{R_i\}$ to $W$ in $A$ is preserved in $H$
+if there is another
+abstract execution $B$ realized by hardware execution $G$
+(of machine code produced by the same compiler as $H$)
+that together \emph{witness the dependency}.
 \end{quote}
@@
 
-In this context our definition of semantic dependency is essentially
+In this context our definition of dependency preservation is essentially
 the same as before.
@@
 
-Of course we want to check that our definition of semantic dependency
-satisfies the Fundamental Property of
+Of course we want to check that whenever a dependency is preserved,
+it satisfies the Fundamental Property of
 Section~\ref{sec:The Fundamental Property of Semantic Dependencies}.
@@
 there is an abstract execution $B$ with hardware realization $G$
-witnessing the semantic dependency of the store $W$ on the loads
-$\{R_i\}$ in $A$ and $H$.
+witnessing that the dependency of the store $W$ on the loads
+$\{R_i\}$ in $A$ is preserved in $H$.
 We must show that some $R_i$ is not omitted and $R'_i$ is ordered
@@
 both.
-This contradicts the fact that $B$ and $G$ witness the semantic
-dependency.
+This contradicts the fact that $B$ and $G$ witness the preservation
+of the dependency.
 
@@
 And since the load must lie in the common period of $H$, by the
-definition of semantic dependency it must either be one of the
+definition above it must either be one of the
 $R'_i$ or be ordered after one of them.
@@
 
+\medskip
+
 A corollary of this result is that if an implementation's compiler
@@
 (2)~uses global analysis and treats atomic objects as volatile,
-then programs produced by that implementation will never exhibit OOTA.
+then programs produced by that implementation will never exhibit
+an OOTA cycle in which the semantic dependencies are preserved.
+And since by the DP thesis, programs produced by
+the implementation always preserve semantic dependencies,
+these programs will never exhibit any OOTA cycles at all.
 Thus the implementation will automatically comply with full C++, even
@@
 
-Here we consider some general questions related to our definition of
-semantic dependency.
+Here we consider some general questions related to semantic dependency
+and our definition of dependency preservation.
 Additional information is available in P3064R2 (``How to Avoid OOTA Without
@@
 
-\subsection{Relative vs.\ Absolute Dependency}
-\label{sec:Relative vs. Absolute Dependency}
-
-One drawback of our definition of semantic dependency is that it is only
-relative to a specific compiler or implementation.
-This may strike some readers (especially those implementing tooling)
-as wishy-washy and avoiding the real problem, in that a given piece
-of code should either contain or not contain a semantic dependency,
-independent of the implementation used to run it.
-
-We can address this drawback by defining an \emph{absolute semantic
-dependency} as one that is present relative to any valid loose C++
+\subsection{Defining Semantic Dependency}
+\label{sec:Defining Semantic Dependency}
+
+Some readers may object that our approach does not include any
+rigorous definition of semantic dependency itself.
+One could address this drawback by turning our DP thesis around
+and using it instead as a \emph{definition} of semantic dependency.
+In other words, a syntactic dependency could be considered to be semantic
+if it is preserved by any valid loose C++
 implementation, past, present, or future, real or imagined.
@@
 
-There is one thing we can definitely state:
-Programs produced by an implementation of the right sort will never
-exhibit absolute OOTA (that is, an OOTA cycle in which all the
-semantic dependencies are absolute).
-This is simply because an absolute semantic dependency is {\it
-a fortiori\/} a semantic dependency relative to the compiler in use.
-
-But in fact we have shown more than this:
-Programs will never exhibit an OOTA cycle relative to the compiler
-used to build them, even when that cycle is not absolute.
-In this sense we have gone beyond the requirement of full C++.
+Either way, regardless of how one wants to define semantic dependency,
+we have shown that as long as the DP thesis continues to hold,
+programs produced by an implementation of the right sort will never
+exihibit OOTA.
 
@@
 
-In principle we don't need to require global-analysis compilers to
-treat atomic objects as volatile; our results would still hold if they
-merely treated them as quasi volatile.
-We chose not to do this because it would violate our intuitions about
-semantic dependencies.
-
-For example, consider the Simple OOTA program shown in
+One can ask if we really need to require global-analysis compilers to
+treat atomic objects as volatile; would our results still hold if they
+merely treated them as quasi volatile?
+Answer: The proof that preserved dependencies obey the Fundamental Property
+would go through as before,
+but the DP thesis would no longer be valid.
+Here's an example showing how it would fail.
+
+Consider the Simple OOTA program shown in
 Listing~\ref{lst:Simple OOTA}.
@@
 Section~\ref{sec:Global Optimization Can Destroy Dependencies}.
-This would be a valid transformation, and the resulting behavior would
-not count as OOTA according to our definition because the
-dependencies in T0 and T1 would not be semantic.
-
-To see why not, recall that a semantic dependency must have a witness,
+This would be a valid transformation,
+but it would destroy the dependencies in T0 and T1 rather than preserving them,
+even though those dependencies are considered to be semantic.
+
+This fact is intuitively obvious, but let us describe in detail
+exactly how the preservation definition would fail.
+Recall that according to the definition, in order to be preserved in an execution
+a syntactic dependency must have a witness,
 another execution in which the store acts differently.
@@
 This unintuitive behavior could not occur if the two loads were not
-omitted.
-In fact, the definition of semantic dependency might remain
-perfectly acceptable if the requirement for global-analysis compilers
-were weakened, if the compiler treated atomic objects as quasi
-volatile and in addition was not allowed to omit accesses to them.
+omitted in the transformed program.
+In fact, the definition of dependency preservation might remain
+strong enough to justify the DP thesis
+if the requirement for global-analysis compilers
+were weakened, if the compiler were allowed to treat atomic objects as
+quasi volatile except that it was not allowed to omit accesses to them.
 This is a possible topic for future research.
@@
 
-Part of our demonstration of the Fundamental Property of semantic
+Part of our demonstration of the Fundamental Property for preserved
 dependencies relies on the fact, stated in
@@
 To rule out such pathological counterexamples we should require that
-in a witness to a semantic dependency, the addresses of all the
+in a witness to a preserved dependency, the addresses of all the
 objects and functions referred to in the thread $T$ are the same as
@@
 
-To demonstrate the semantic dependency in the original code,
+To demonstrate that the dependency in the original code is preserved,
 a suitable witness would have
-to include a hardware realization of the abstract execution in which
+to include a hardware realization of an abstract execution in which
 \co{r1} and \co{r2} are zero and \co{r3} and \co{r4} are one.
-But there are no hardware realizations of this execution with the
+But there are no such hardware realizations with the
 machine code indicated above!
@@
 necessarily be one, not zero.
-Thus, relative to this compiler the example does not contain a
-semantic dependency.
-
-Although this is unexpected, and while we don't recommend this sort of
+Thus the dependency is not preserved, which leads one to question
+whether it really was semantic.
+
+Although this result is unexpected, and while we don't recommend this sort of
 code transformation, we cannot say that this conclusion is definitely
@@
 multiple threads into one;
-(5)~Explore cases in which semantic dependencies must be absolute,
-not relative to any particular compiler;
+(5)~Explore more detailed justifications for the Dependency Preservation thesis;
 (6)~Include the effects of input and output or other operations which
@@
 It introduced the notion of quasi-volatile object accesses and gave
-a definition for semantic dependency that is relative to a particular
-C++ implementation.
-It formulated a Fundamental Property of semantic dependency and proved
+a definition for dependency preservation by a C++ implementation.
+It formulated a Fundamental Property of semantic dependency along with
+a Dependency Preservation thesis, and using the thesis proved
 that implementations subject to some very mild restrictions on how

### 2025-06-13 paper/ootavacuous.tex
ootavacuous: Add presenting author's email address
(I was chided for having left this out by working-group chairs who
need to be able to get in touch with the author who will be presenting.
I am assuming that Alan Stern won't be in Sofia next week, so that would
be me.)

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Add Alan email, standard non-request, D-number paper
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{document}
-\title{P3692R0: How to Avoid OOTA Without Really Trying}
+\title{D3692R1: How to Avoid OOTA Without Really Trying}
 
@@
 \author{
-Alan Stern \and
-Paul E.~McKenney \and
+Alan Stern (stern@rowland.harvard.edu) \and
+Paul E.~McKenney (paulmckrcu@gmail.com) \and
 Michael Wong \and
@@
 	action from implementations and no changes to user code.
+	As a consequence, we are not asking for any normative changes
+	to the standard.
 

### 2025-06-14 paper/ootavacuous.tex
ootavacuous: Clarify what we are asking of the standard
Namely, a small non-normative change.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Add D3692P1 item to history section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	action from implementations and no changes to user code.
-	As a consequence, we are not asking for any normative changes
-	to the standard.
+	As a consequence, we are asking for only for a small non-normative
+	change to the standard.
 
@@
 
-This paper is a condensation of P3064R2 (``How to Avoid OOTA Without
-Really Trying'')~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying},
+\begin{itemize}
+\item	D3692R1 adds contact information, defines ``preserved
+	dependencies'' that are a subset of syntactic and a superset of
+	semantic dependencies, and makes numerous changes in the name
+	of wordsmithing and clarification.
+\item	This paper is a condensation of P3064R2 (``How to Avoid OOTA
+	Without Really
+	Trying'')~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying},
 	which was criticized as being overly long.
 	It also adds citations and clarifications.
+\end{itemize}
 

### 2025-06-14 paper/ootavacuous.tex
ootavacuous: Add for shadowing of the newish Section 1.3
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Remove "references" question from section 1.1
On further thought, it does not seem like we really need a reference
for "syntactic dependency".  Of course, if there proves to be a nice
convenient reference, why not?

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 We begin with a brief overview of the OOTA problem followed by an equally
-brief summary of prior OOTA work.
+brief summary of prior OOTA work and of our approach.
 
@@
 leading up to the store, whether it affects the store or not.
-\emph{@@@ Reference?}
 Although we will not give a precise definition of syntactic dependency

### 2025-06-14 paper/ootavacuous.tex
ootavacuous: Note that "carries a dependency" is deprecated
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Remove statement of lack of confidence from section 1.2
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 here, doing so presents no fundamental difficulties.
-In fact, the C++ standard covers much of the same ground in its definition
+In fact, the C++ standard covers much of the same ground in its
+(now deprecated) definition
 of ``carries a dependency'' in 6.9.2.2p7 (\co{[Data races]}).
@@
 model to be accepted by C++ implementers.
-\emph{@@@ I am not fully confident of my characterization of this paper.}
 

### 2025-06-14 paper/ootavacuous.tex
ootavacuous: Clarify non-semantic dependency in Section 2.2.2
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Cite P3064R2 Appendix B for gcc command line
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 both \co{x} and \co{y} are zero.
-Then changes to either \co{x} or \co{y}
+Then a change to exactly one of \co{x} or \co{y}
 will not cause a change in the value stored to \co{z}.
@@
 may be found through use of the command \co{make V=1}
-in a Linux-kernel source tree and by the discussion
+in a Linux-kernel source
+tree~\cite[Appendix B]{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.
+and by the discussion
 in~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.

### 2025-06-14 paper/ootavacuous.tex
ootavacuous: Add Section 7.1 forward reference to 7.4
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Of course this notion has its own problems, including that it is
-extremely nonconstructive and impossible to apply in practice.
+extremely nonconstructive and impossible to apply in practice,
+see for example
+Section~\ref{sec:Merging Quasi-Volatile Loads}.
 However it may be the best we can do with our current understanding

### 2025-06-14 paper/ootavacuous.tex
ootavacuous: Section 7.4: Drop "simply", just "because"
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Section 7.6: Expand "them" to "OOTA cycles"
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 transformation is worth the added conceptual load on developers.
-Until that time, pure mathematics cannot help us simply because it is
+Until that time, pure mathematics cannot help us because it is
 impossible to say whether or not there is a semantic dependency from
@@
 As noted in Section~\ref{sec:Brief OOTA Overview}, the C++
-standard explicitly recommends against them in 33.5.4p8
+standard recommends against OOTA cycles in 33.5.4p8
 (\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:

### 2025-06-17 paper/ootavacuous.tex
ootavacuous: Section 7.6: Wordsmith away redundancy
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Section 7.6 further wordsmithing
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Add offlist discussions to history
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Comment out rhetorical question for Sofia meeting
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
ootavacuous: Add footnote for Section 7.4 litmus test
Also add the missing "repurpose" to the proposed standard addition.

Sign

@@
 	which was criticized as being overly long.
+	Voluminous off-list discussions are reflected.
 	It also adds citations and clarifications.
@@
 all~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.
-\emph{@@@ Are these the right qualifiers?}
+% \emph{@@@ Are these the right qualifiers?}
 
@@
 impossible to say whether or not there is a semantic dependency from
-\co{y} to \co{z} without reference to a specific implementation.
+\co{y} to \co{z} without reference to a specific implementation.\footnote{
+	At the 2025 C++ Standards Committee meeting in Sofia, it was
+	pointed out that the added ``\co{if (r1 != r3)}'' can be argued
+	to be repurposing the first and third relaxed atomic load from
+	\co{x}.
+	At first glance, this is a very reasonable argument that would
+	unambigously rule out this optimization.}
 
@@
 recommendation?
-One way to reduce their anxiety levels would be to add something like
-the following sentence to this recommendation:
+We therefore advocate adding the following non-normative sentence to this
+recommendation:
 \begin{quote}
@@
 	as long as they restrict themselves to thread-at-a-time analysis
-	and do not invent or duplicate atomic accesses.
+	and do not invent, duplicate, or repurpose atomic accesses.
 \end{quote}

### 2025-07-08 paper/ootavacuous.tex
ootavacuous: State benefits to full-up OOTA research in "Our Approach"
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 Michael Wong \and
-Maged Michael
+Maged Michael \and
+Gonzalo Brito
 }
@@
 all~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.
+In addition, for those researchers working on identifying and avoiding
+OOTA cycles, our results accommodate the needs of almost all C++
+implementers, thus removing a key obstacle to the application of
+high-complexity methods to this more general problem.
 % \emph{@@@ Are these the right qualifiers?}

### 2025-07-08 paper/ootavacuous.tex
ootavacuosu: Rework change to standard in terms of volatile atomics
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 	as long as they restrict themselves to thread-at-a-time analysis
-	and do not invent, duplicate, or repurpose atomic accesses.
+	and and treat non-volatile atomic accesses as if they were
+	volatile, except that they may merge or omit such accesses when
+	permitted by the as-if rule.
 \end{quote}

### 2025-07-08 paper/ootavacuous.tex
ootavacuous: Add reordering of non-volatile atomics on different objects
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

--- 2025-07-08 paper/ootavacuous.tex
@@
 	as long as they restrict themselves to thread-at-a-time analysis
-	and do not invent, duplicate, or repurpose atomic accesses.
+	and and treat non-volatile atomic accesses as if they were
+	volatile, except that, when permitted by the as-if rule, they
+	may omit accesses, merge accesses to the same object, or reorder
+	accesses to different objects.
 \end{quote}

### 2025-07-10 paper/ootavacuous.tex
ootavacuous: Restate benefits to full-up OOTA research in "Our Approach"
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 all~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.
-In addition, for those researchers working on identifying and avoiding
-OOTA cycles, our results accommodate the needs of almost all C++
-implementers, thus removing a key obstacle to the application of
-high-complexity methods to this more general problem.
+In addition, our results simplify the job of researchers seeking
+practical impact by allowing them ignore C++ implementers' need for
+extreme simplicity and performance.
+This in turn permits such researchers to focus on advanced topics such
+as formal-verification tooling and evaluation of optimizations.
 % \emph{@@@ Are these the right qualifiers?}

### 2025-07-10 paper/ootavacuous.tex
ootavacuous: Update history and move from D3692R1 to P3692R2
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@
 \begin{document}
-\title{D3692R1: How to Avoid OOTA Without Really Trying}
+\title{P3692R2: How to Avoid OOTA Without Really Trying}
 
@@
 \begin{itemize}
-\item	D3692R1 adds contact information, defines ``preserved
+\item	P3692R2 applies feedback from the 2025 Sophia meeting,
+	including reworking the requested change to the standard
+	in terms of volatile semantics (courtesy of Olivier Giroux),
+	states this paper's benefits to other OOTA researchers, and
+	updates the author list.
+\item	P3692R1 adds contact information, defines ``preserved
 	dependencies'' that are a subset of syntactic and a superset of

### 2025-07-11 paper/ootavacuous.tex

@@
 all~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.
-In addition, our results simplify the job of researchers seeking
-practical impact by allowing them ignore C++ implementers' need for
+In addition, our results simplify the job of OOTA researchers seeking
+practical impact by allowing them to ignore C++ implementers' need for
 extreme simplicity and performance.
